### 2019-05-17 paper/move-op.tex
Paper setup

\documentclass[sigconf]{acmart}

\usepackage{isabelle,isabellesym}
\isabellestyle{it}
\renewenvironment{isabelle}{%
  \bigbreak\noindent\hspace{\parindent}%
  \begin{minipage}{\textwidth-\parindent}% use minipage to prevent page breaks
  \begin{isabellebody}%
  \begin{tabbing}%
}{%
  \end{tabbing}%
  \end{isabellebody}%
  \end{minipage}%
  \bigbreak%
}
\renewcommand{\isanewline}{\\}

\setcopyright{rightsretained}
\copyrightyear{2019}
\acmYear{2019}
\acmDOI{}
\acmConference[]{}{}{}
\acmBooktitle{}
\acmPrice{}
\acmISBN{}

\begin{document}
\title{A highly-available move operation for replicated trees}

\author{Martin Kleppmann}
\email{mk428@cl.cam.ac.uk}
\orcid{0000-0001-7252-6958}
\affiliation{%
  \institution{Department of Computer Science and Technology}
  \streetaddress{15 JJ Thomson Avenue}
  \city{Cambridge}
  \state{}
  \postcode{CB3 0FD}
  \country{United Kingdom}
}

\author{Dominic P.\ Mulligan}
\email{Dominic.Mulligan@arm.com}
\orcid{0000-0003-4643-3541}
\affiliation{%
  \institution{Arm Research}
  \streetaddress{110 Fulbourn Road}
  \city{Cambridge}
  \state{}
  \postcode{CB1 9NJ}
  \country{United Kingdom}
}

\author{Victor B.\ F.\ Gomes}
\email{vb358@cl.cam.ac.uk}
\orcid{0000-0002-2954-4648}
\affiliation{%
  \institution{Department of Computer Science and Technology}
  \streetaddress{15 JJ Thomson Avenue}
  \city{Cambridge}
  \state{}
  \postcode{CB3 0FD}
  \country{United Kingdom}
}

\author{Alastair R.\ Beresford}
\email{arb33@cl.cam.ac.uk}
\orcid{0000-0003-0818-6535}
\affiliation{%
  \institution{Department of Computer Science and Technology}
  \streetaddress{15 JJ Thomson Avenue}
  \city{Cambridge}
  \state{}
  \postcode{CB3 0FD}
  \country{United Kingdom}
}

\renewcommand{\shortauthors}{Kleppmann, Mulligan, Gomes, and Beresford}
\maketitle

\begin{abstract}
\end{abstract}

\section{Introduction}

CRDTs rock \cite{Shapiro:2011wy,Shapiro:2011un}.

\begin{acks}
The authors wish to acknowledge the support of The Boeing Company and the EPSRC ``REMS: Rigorous Engineering for Mainstream Systems'' programme grant (EP/K008528).
\end{acks}

\bibliographystyle{ACM-Reference-Format}
\bibliography{references}{}
\end{document}

### 2019-05-22 paper/move-op.tex
Typeset Isabelle code in paper

@@
 
+\usepackage{lineno}
 \usepackage{isabelle,isabellesym}
-\isabellestyle{it}
+\isabellestyle{tt}
 \renewenvironment{isabelle}{%
@@
 }
-\renewcommand{\isanewline}{\\}
+\renewcommand{\isacartoucheopen}{}
+\renewcommand{\isacartoucheclose}{}
 
@@
 
+\begin{figure*}
+\raggedright
+\begin{isabellebody}
+\internallinenumbers
+\isacommand{datatype}\isamarkupfalse%
+\ {\isacharparenleft}{\isacharprime}t{\isacharcomma}\ {\isacharprime}n{\isacharcomma}\ {\isacharprime}m{\isacharparenright}\ operation\isanewline
+\ \ {\isacharequal}\ Move\ {\isacharparenleft}move{\isacharunderscore}time{\isacharcolon}\ {\isacharprime}t{\isacharparenright}\isanewline
+\ \ \ \ \ \ \ \ \ {\isacharparenleft}move{\isacharunderscore}parent{\isacharcolon}\ {\isacharprime}n{\isacharparenright}\isanewline
+\ \ \ \ \ \ \ \ \ {\isacharparenleft}move{\isacharunderscore}meta{\isacharcolon}\ {\isacharprime}m{\isacharparenright}\isanewline
+\ \ \ \ \ \ \ \ \ {\isacharparenleft}move{\isacharunderscore}child{\isacharcolon}\ {\isacharprime}n{\isacharparenright}\isanewline
+\isanewline
+\isacommand{datatype}\isamarkupfalse%
+\ {\isacharparenleft}{\isacharprime}t{\isacharcomma}\ {\isacharprime}n{\isacharcomma}\ {\isacharprime}m{\isacharparenright}\ log{\isacharunderscore}op\isanewline
+\ \ {\isacharequal}\ LogMove\ {\isacharparenleft}log{\isacharunderscore}time{\isacharcolon}\ {\isacharprime}t{\isacharparenright}\isanewline
+\ \ \ \ \ \ \ \ \ \ \ \ {\isacharparenleft}old{\isacharunderscore}parent{\isacharcolon}\ {\isacartoucheopen}{\isacharparenleft}{\isacharprime}n\ {\isasymtimes}\ {\isacharprime}m{\isacharparenright}\ option{\isacartoucheclose}{\isacharparenright}\isanewline
+\ \ \ \ \ \ \ \ \ \ \ \ {\isacharparenleft}new{\isacharunderscore}parent{\isacharcolon}\ {\isacharprime}n{\isacharparenright}\isanewline
+\ \ \ \ \ \ \ \ \ \ \ \ {\isacharparenleft}log{\isacharunderscore}meta{\isacharcolon}\ {\isacharprime}m{\isacharparenright}\isanewline
+\ \ \ \ \ \ \ \ \ \ \ \ {\isacharparenleft}log{\isacharunderscore}child{\isacharcolon}\ {\isacharprime}n{\isacharparenright}\isanewline
+\isanewline
+\isacommand{type{\isacharunderscore}synonym}\isamarkupfalse%
+\ {\isacharparenleft}{\isacharprime}t{\isacharcomma}\ {\isacharprime}n{\isacharcomma}\ {\isacharprime}m{\isacharparenright}\ state\ {\isacharequal}\ {\isacartoucheopen}{\isacharparenleft}{\isacharprime}t{\isacharcomma}\ {\isacharprime}n{\isacharcomma}\ {\isacharprime}m{\isacharparenright}\ log{\isacharunderscore}op\ list\ {\isasymtimes}\ {\isacharparenleft}{\isacharprime}n\ {\isasymtimes}\ {\isacharprime}m\ {\isasymtimes}\ {\isacharprime}n{\isacharparenright}\ set{\isacartoucheclose}\isanewline
+\isanewline
+\isacommand{definition}\isamarkupfalse%
+\ get{\isacharunderscore}parent\ {\isacharcolon}{\isacharcolon}\ {\isacartoucheopen}{\isacharparenleft}{\isacharprime}n\ {\isasymtimes}\ {\isacharprime}m\ {\isasymtimes}\ {\isacharprime}n{\isacharparenright}\ set\ {\isasymRightarrow}\ {\isacharprime}n\ {\isasymRightarrow}\ {\isacharparenleft}{\isacharprime}n\ {\isasymtimes}\ {\isacharprime}m{\isacharparenright}\ option{\isacartoucheclose}\ \isakeyword{where}\isanewline
+\ \ {\isacartoucheopen}get{\isacharunderscore}parent\ tree\ child\ {\isasymequiv}\isanewline
+\ \ \ \ \ if\ {\isasymexists}{\isacharbang}parent{\isachardot}\ {\isasymexists}{\isacharbang}meta{\isachardot}\ {\isacharparenleft}parent{\isacharcomma}\ meta{\isacharcomma}\ child{\isacharparenright}\ {\isasymin}\ tree\ then\isanewline
+\ \ \ \ \ \ \ Some\ {\isacharparenleft}THE\ {\isacharparenleft}parent{\isacharcomma}\ meta{\isacharparenright}{\isachardot}\ {\isacharparenleft}parent{\isacharcomma}\ meta{\isacharcomma}\ child{\isacharparenright}\ {\isasymin}\ tree{\isacharparenright}\isanewline
+\ \ \ \ \ else\ None{\isacartoucheclose}\isanewline
+\isanewline
+\isacommand{inductive}\isamarkupfalse%
+\ ancestor\ {\isacharcolon}{\isacharcolon}\ {\isacartoucheopen}{\isacharparenleft}{\isacharprime}n\ {\isasymtimes}\ {\isacharprime}m\ {\isasymtimes}\ {\isacharprime}n{\isacharparenright}\ set\ {\isasymRightarrow}\ {\isacharprime}n\ {\isasymRightarrow}\ {\isacharprime}n\ {\isasymRightarrow}\ bool{\isacartoucheclose}\ \isakeyword{where}\isanewline
+\ \ {\isacartoucheopen}{\isasymlbrakk}{\isacharparenleft}parent{\isacharcomma}\ meta{\isacharcomma}\ child{\isacharparenright}\ {\isasymin}\ tree{\isasymrbrakk}\ {\isasymLongrightarrow}\ ancestor\ tree\ parent\ child{\isacartoucheclose}\ {\isacharbar}\isanewline
+\ \ {\isacartoucheopen}{\isasymlbrakk}{\isacharparenleft}parent{\isacharcomma}\ meta{\isacharcomma}\ child{\isacharparenright}\ {\isasymin}\ tree{\isacharsemicolon}\ ancestor\ tree\ anc\ parent{\isasymrbrakk}\ {\isasymLongrightarrow}\ ancestor\ tree\ anc\ child{\isacartoucheclose}\isanewline
+\isanewline
+\isacommand{fun}\isamarkupfalse%
+\ do{\isacharunderscore}op\ {\isacharcolon}{\isacharcolon}\ {\isacartoucheopen}{\isacharparenleft}{\isacharprime}t{\isacharcomma}\ {\isacharprime}n{\isacharcomma}\ {\isacharprime}m{\isacharparenright}\ operation\ {\isasymtimes}\ {\isacharparenleft}{\isacharprime}n\ {\isasymtimes}\ {\isacharprime}m\ {\isasymtimes}\ {\isacharprime}n{\isacharparenright}\ set\ {\isasymRightarrow}\isanewline
+\ \ \ \ \ \ \ \ \ \ \ \ \ \ {\isacharparenleft}{\isacharprime}t{\isacharcomma}\ {\isacharprime}n{\isacharcomma}\ {\isacharprime}m{\isacharparenright}\ log{\isacharunderscore}op\ {\isasymtimes}\ {\isacharparenleft}{\isacharprime}n\ {\isasymtimes}\ {\isacharprime}m\ {\isasymtimes}\ {\isacharprime}n{\isacharparenright}\ set{\isacartoucheclose}\ \isakeyword{where}\isanewline
+\ \ {\isacartoucheopen}do{\isacharunderscore}op\ {\isacharparenleft}Move\ t\ newp\ m\ c{\isacharcomma}\ tree{\isacharparenright}\ {\isacharequal}\isanewline
+\ \ \ \ \ {\isacharparenleft}LogMove\ t\ {\isacharparenleft}get{\isacharunderscore}parent\ tree\ c{\isacharparenright}\ newp\ m\ c{\isacharcomma}\isanewline
+\ \ \ \ \ \ if\ ancestor\ tree\ c\ newp\ {\isasymor}\ c\ {\isacharequal}\ newp\ then\ tree\isanewline
+\ \ \ \ \ \ else\ {\isacharbraceleft}{\isacharparenleft}p{\isacharprime}{\isacharcomma}\ m{\isacharprime}{\isacharcomma}\ c{\isacharprime}{\isacharparenright}\ {\isasymin}\ tree{\isachardot}\ c{\isacharprime}\ {\isasymnoteq}\ c{\isacharbraceright}\ {\isasymunion}\ {\isacharbraceleft}{\isacharparenleft}newp{\isacharcomma}\ m{\isacharcomma}\ c{\isacharparenright}{\isacharbraceright}{\isacharparenright}{\isacartoucheclose}\isanewline
+\isanewline
+\isacommand{fun}\isamarkupfalse%
+\ undo{\isacharunderscore}op\ {\isacharcolon}{\isacharcolon}\ {\isacartoucheopen}{\isacharparenleft}{\isacharprime}t{\isacharcomma}\ {\isacharprime}n{\isacharcomma}\ {\isacharprime}m{\isacharparenright}\ log{\isacharunderscore}op\ {\isasymtimes}\ {\isacharparenleft}{\isacharprime}n\ {\isasymtimes}\ {\isacharprime}m\ {\isasymtimes}\ {\isacharprime}n{\isacharparenright}\ set\ {\isasymRightarrow}\ {\isacharparenleft}{\isacharprime}n\ {\isasymtimes}\ {\isacharprime}m\ {\isasymtimes}\ {\isacharprime}n{\isacharparenright}\ set{\isacartoucheclose}\ \isakeyword{where}\isanewline
+\ \ {\isacartoucheopen}undo{\isacharunderscore}op\ {\isacharparenleft}LogMove\ t\ None\ newp\ m\ c{\isacharcomma}\ tree{\isacharparenright}\ {\isacharequal}\ {\isacharbraceleft}{\isacharparenleft}p{\isacharprime}{\isacharcomma}\ m{\isacharprime}{\isacharcomma}\ c{\isacharprime}{\isacharparenright}\ {\isasymin}\ tree{\isachardot}\ c{\isacharprime}\ {\isasymnoteq}\ c{\isacharbraceright}{\isacartoucheclose}\ {\isacharbar}\isanewline
+\ \ {\isacartoucheopen}undo{\isacharunderscore}op\ {\isacharparenleft}LogMove\ t\ {\isacharparenleft}Some\ {\isacharparenleft}oldp{\isacharcomma}\ oldm{\isacharparenright}{\isacharparenright}\ newp\ m\ c{\isacharcomma}\ tree{\isacharparenright}\ {\isacharequal}\isanewline
+\ \ \ \ \ {\isacharbraceleft}{\isacharparenleft}p{\isacharprime}{\isacharcomma}\ m{\isacharprime}{\isacharcomma}\ c{\isacharprime}{\isacharparenright}\ {\isasymin}\ tree{\isachardot}\ c{\isacharprime}\ {\isasymnoteq}\ c{\isacharbraceright}\ {\isasymunion}\ {\isacharbraceleft}{\isacharparenleft}oldp{\isacharcomma}\ oldm{\isacharcomma}\ c{\isacharparenright}{\isacharbraceright}{\isacartoucheclose}\isanewline
+\isanewline
+\isacommand{fun}\isamarkupfalse%
+\ redo{\isacharunderscore}op\ {\isacharcolon}{\isacharcolon}\ {\isacartoucheopen}{\isacharparenleft}{\isacharprime}t{\isacharcomma}\ {\isacharprime}n{\isacharcomma}\ {\isacharprime}m{\isacharparenright}\ log{\isacharunderscore}op\ {\isasymRightarrow}\ {\isacharparenleft}{\isacharprime}t{\isacharcomma}\ {\isacharprime}n{\isacharcomma}\ {\isacharprime}m{\isacharparenright}\ state\ {\isasymRightarrow}\ {\isacharparenleft}{\isacharprime}t{\isacharcomma}\ {\isacharprime}n{\isacharcomma}\ {\isacharprime}m{\isacharparenright}\ state{\isacartoucheclose}\ \isakeyword{where}\isanewline
+\ \ {\isacartoucheopen}redo{\isacharunderscore}op\ {\isacharparenleft}LogMove\ t\ {\isacharunderscore}\ p\ m\ c{\isacharparenright}\ {\isacharparenleft}ops{\isacharcomma}\ tree{\isacharparenright}\ {\isacharequal}\isanewline
+\ \ \ \ \ {\isacharparenleft}let\ {\isacharparenleft}op{\isadigit{2}}{\isacharcomma}\ tree{\isadigit{2}}{\isacharparenright}\ {\isacharequal}\ do{\isacharunderscore}op\ {\isacharparenleft}Move\ t\ p\ m\ c{\isacharcomma}\ tree{\isacharparenright}\isanewline
+\ \ \ \ \ \ in\ {\isacharparenleft}op{\isadigit{2}}\ {\isacharhash}\ ops{\isacharcomma}\ tree{\isadigit{2}}{\isacharparenright}{\isacharparenright}{\isacartoucheclose}\isanewline
+\isanewline
+\isacommand{fun}\isamarkupfalse%
+\ interp{\isacharunderscore}op\ {\isacharcolon}{\isacharcolon}\ {\isacartoucheopen}{\isacharparenleft}{\isacharprime}t{\isacharcolon}{\isacharcolon}{\isacharbraceleft}linorder{\isacharbraceright}{\isacharcomma}\ {\isacharprime}n{\isacharcomma}\ {\isacharprime}m{\isacharparenright}\ operation\ {\isasymRightarrow}\isanewline
+\ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ {\isacharparenleft}{\isacharprime}t{\isacharcomma}\ {\isacharprime}n{\isacharcomma}\ {\isacharprime}m{\isacharparenright}\ state\ {\isasymRightarrow}\ {\isacharparenleft}{\isacharprime}t{\isacharcomma}\ {\isacharprime}n{\isacharcomma}\ {\isacharprime}m{\isacharparenright}\ state{\isacartoucheclose}\ \isakeyword{where}\isanewline
+\ \ {\isacartoucheopen}interp{\isacharunderscore}op\ op{\isadigit{1}}\ {\isacharparenleft}{\isacharbrackleft}{\isacharbrackright}{\isacharcomma}\ tree{\isadigit{1}}{\isacharparenright}\ {\isacharequal}\isanewline
+\ \ \ \ \ {\isacharparenleft}let\ {\isacharparenleft}op{\isadigit{2}}{\isacharcomma}\ tree{\isadigit{2}}{\isacharparenright}\ {\isacharequal}\ do{\isacharunderscore}op\ {\isacharparenleft}op{\isadigit{1}}{\isacharcomma}\ tree{\isadigit{1}}{\isacharparenright}\isanewline
+\ \ \ \ \ \ in\ {\isacharparenleft}{\isacharbrackleft}op{\isadigit{2}}{\isacharbrackright}{\isacharcomma}\ tree{\isadigit{2}}{\isacharparenright}{\isacharparenright}{\isacartoucheclose}\ {\isacharbar}\isanewline
+\ \ {\isacartoucheopen}interp{\isacharunderscore}op\ op{\isadigit{1}}\ {\isacharparenleft}logop\ {\isacharhash}\ ops{\isacharcomma}\ tree{\isadigit{1}}{\isacharparenright}\ {\isacharequal}\isanewline
+\ \ \ \ \ {\isacharparenleft}if\ move{\isacharunderscore}time\ op{\isadigit{1}}\ {\isacharless}\ log{\isacharunderscore}time\ logop\isanewline
+\ \ \ \ \ \ then\ redo{\isacharunderscore}op\ logop\ {\isacharparenleft}interp{\isacharunderscore}op\ op{\isadigit{1}}\ {\isacharparenleft}ops{\isacharcomma}\ undo{\isacharunderscore}op\ {\isacharparenleft}logop{\isacharcomma}\ tree{\isadigit{1}}{\isacharparenright}{\isacharparenright}{\isacharparenright}\isanewline
+\ \ \ \ \ \ else\ let\ {\isacharparenleft}op{\isadigit{2}}{\isacharcomma}\ tree{\isadigit{2}}{\isacharparenright}\ {\isacharequal}\ do{\isacharunderscore}op\ {\isacharparenleft}op{\isadigit{1}}{\isacharcomma}\ tree{\isadigit{1}}{\isacharparenright}\ in\ {\isacharparenleft}op{\isadigit{2}}\ {\isacharhash}\ logop\ {\isacharhash}\ ops{\isacharcomma}\ tree{\isadigit{2}}{\isacharparenright}{\isacharparenright}{\isacartoucheclose}\isanewline
+\isanewline
+\isacommand{definition}\isamarkupfalse%
+\ interp{\isacharunderscore}ops\ {\isacharcolon}{\isacharcolon}\ {\isacartoucheopen}{\isacharparenleft}{\isacharprime}t{\isacharcolon}{\isacharcolon}{\isacharbraceleft}linorder{\isacharbraceright}{\isacharcomma}\ {\isacharprime}n{\isacharcomma}\ {\isacharprime}m{\isacharparenright}\ operation\ list\ {\isasymRightarrow}\ {\isacharparenleft}{\isacharprime}t{\isacharcomma}\ {\isacharprime}n{\isacharcomma}\ {\isacharprime}m{\isacharparenright}\ state{\isacartoucheclose}\ \isakeyword{where}\isanewline
+\ \ {\isacartoucheopen}interp{\isacharunderscore}ops\ ops\ {\isasymequiv}\ foldl\ {\isacharparenleft}{\isasymlambda}state\ oper{\isachardot}\ interp{\isacharunderscore}op\ oper\ state{\isacharparenright}\ {\isacharparenleft}{\isacharbrackleft}{\isacharbrackright}{\isacharcomma}\ {\isacharbraceleft}{\isacharbraceright}{\isacharparenright}\ ops{\isacartoucheclose}\isanewline
+\isanewline
+\isacommand{definition}\isamarkupfalse%
+\ unique{\isacharunderscore}parent\ {\isacharcolon}{\isacharcolon}\ {\isacartoucheopen}{\isacharparenleft}{\isacharprime}n\ {\isasymtimes}\ {\isacharprime}m\ {\isasymtimes}\ {\isacharprime}n{\isacharparenright}\ set\ {\isasymRightarrow}\ bool{\isacartoucheclose}\ \isakeyword{where}\isanewline
+\ \ {\isacartoucheopen}unique{\isacharunderscore}parent\ tree\ {\isasymequiv}\ {\isacharparenleft}{\isasymforall}p{\isadigit{1}}\ p{\isadigit{2}}\ m{\isadigit{1}}\ m{\isadigit{2}}\ c{\isachardot}\ {\isacharparenleft}p{\isadigit{1}}{\isacharcomma}\ m{\isadigit{1}}{\isacharcomma}\ c{\isacharparenright}\ {\isasymin}\ tree\ {\isasymand}\ {\isacharparenleft}p{\isadigit{2}}{\isacharcomma}\ m{\isadigit{2}}{\isacharcomma}\ c{\isacharparenright}\ {\isasymin}\ tree\ {\isasymlongrightarrow}\ p{\isadigit{1}}\ {\isacharequal}\ p{\isadigit{2}}\ {\isasymand}\ m{\isadigit{1}}\ {\isacharequal}\ m{\isadigit{2}}{\isacharparenright}{\isacartoucheclose}%
+\end{isabellebody}
+\caption{The move operation algorithm, implemented in the Isabelle/HOL language.}
+\label{fig:code}
+\end{figure*}
+
 \begin{acks}

### 2019-06-04 Test.hs
code-gen: some more fiddling
Added Haskell code generation.  Run with:

  > ghci -XScopedTypeVariables Test.hs

@@ -1,0 +1,65 @@
+module Ancestor(Set, ancestor) where {
+
+import Prelude ((==), (/=), (<), (<=), (>=), (>), (+), (-), (*), (/), (**),
+  (>>=), (>>), (=<<), (&&), (||), (^), (^^), (.), ($), ($!), (++), (!!), Eq,
+  error, id, return, not, fst, snd, map, filter, concat, concatMap, reverse,
+  zip, null, takeWhile, dropWhile, all, any, Integer, negate, abs, divMod,
+  String, Bool(True, False), Maybe(Nothing, Just));
+import qualified Prelude;
+
+data Set a = Set [a] | Coset [a];
+
+bex :: forall a. Set a -> (a -> Bool) -> Bool;
+bex (Set xs) p = any p xs;
+
+fold :: forall a b. (a -> b -> b) -> [a] -> b -> b;
+fold f (x : xs) s = fold f xs (f x s);
+fold f [] s = s;
+
+membera :: forall a. (Eq a) => [a] -> a -> Bool;
+membera [] y = False;
+membera (x : xs) y = x == y || membera xs y;
+
+member :: forall a. (Eq a) => a -> Set a -> Bool;
+member x (Coset xs) = not (membera xs x);
+member x (Set xs) = membera xs x;
+
+removeAll :: forall a. (Eq a) => a -> [a] -> [a];
+removeAll x [] = [];
+removeAll x (y : xs) = (if x == y then removeAll x xs else y : removeAll x xs);
+
+inserta :: forall a. (Eq a) => a -> [a] -> [a];
+inserta x xs = (if membera xs x then xs else x : xs);
+
+insert :: forall a. (Eq a) => a -> Set a -> Set a;
+insert x (Coset xs) = Coset (removeAll x xs);
+insert x (Set xs) = Set (inserta x xs);
+
+sup_set :: forall a. (Eq a) => Set a -> Set a -> Set a;
+sup_set (Coset xs) a = Coset (filter (\ x -> not (member x a)) xs);
+sup_set (Set xs) a = fold insert xs a;
+
+bot_set :: forall a. Set a;
+bot_set = Set [];
+
+sup_seta :: forall a. (Eq a) => Set (Set a) -> Set a;
+sup_seta (Set xs) = fold sup_set xs bot_set;
+
+image :: forall a b. (a -> b) -> Set a -> Set b;
+image f (Set xs) = Set (map f xs);
+
[…]

### 2019-06-05 paper/move-op.tex

@@
 
-CRDTs rock \cite{Shapiro:2011wy,Shapiro:2011un}.
+For many systems and applications, the data model that describes the system state is a tree.
+For example:
+\begin{itemize}
+    \item Most filesystems, e.g.
+on Windows and Unix-like OSes, are organised as a tree: folders or directories are branch nodes, and files are leaf nodes.\footnote{Although symbolic links appear to allow non-tree graph structures, such as cyclic references, the underlying inode structure remains a tree: a symbolic link is not a reference to another inode, but rather a special file containing the path of another file or directory, which is resolved to an inode when the symlink is accessed.
+        If POSIX hardlinks are allowed, the filesystem is a restricted DAG rather than a tree, because the same file inode can be referenced from multiple places in the directory tree.
+        However, hardlinks are not supported by Windows, nor by popular file sync products such as Dropbox, Google Drive or Microsoft OneDrive, so we do not consider hardlinks further in this paper.}
+        In Unix-style filesystems, the nodes of the tree are known as \emph{inodes}.
+    \item Rich text editors maintain a tree of textual structures such as paragraphs, bulleted and numbered lists, figures, sections, and so on.
+        These can be nested inside each other: for example, a paragraph may appear inside a bullet point of a bulleted list, which may in turn appear inside a numbered point of an enumeration.
+        Rich text documents are typically represented in memory using the Document Object Model (DOM), and written to file in HTML, XML, or JSON format.
+    \item Vector graphics and presentation software represents images using graphical objects such as text boxes, rectangles, ellipses, lines, and so on.
+        These objects are contained within nodes representing pages of a document, or slides in a presentation.
+        Multiple objects may be combined in a \emph{group} so that they can be moved or manipulated as a unit; this corresponds to making these objects children of a common parent node.
+        Multiple objects and groups may in turn be combined into higher-level groups, forming a tree.
+        Moreover, the order of nodes determines visibility (i.e.\ objects that are ``further back'' may be obscured by objects that are ``closer to the front'').
+    \item Personal note-taking and task-management tools such as org-mode for Emacs, or Workflowy, explicitly present the user with a tree structure that they can inspect and manipulate.
+\end{itemize}
+
+In this paper we consider applications that use such a tree data model, and that replicate this tree across multiple nodes.
+Moreover, we focus on \emph{optimistic} replication~\cite{Saito:2005jw}, that is, systems in which any replica can autonomously make changes to the data, without waiting for communication or coordination with any other replicas.
+Such systems have the advantage that they can continue processing read and write requests even in the presence of arbitrary network partitions; in other words, they are \emph{available} and \emph{partition-tolerant} in the sense of the CAP theorem~\cite{Gilbert:2002il}.
+
+CRDTs~\cite{Shapiro:2011wy,Shapiro:2011un}
+
+\section{Why a move operation is hard}
+
+Errors in Google Drive
+
+% - No Synchronisation or consensus required. Available in the sense of CAP
+% - Rely only on uniqueness of node IDs, e.g. inode numbers
+% - Composing JSON/XML tree by using list datatype
+% - Location is Lww register. Concurrent move does not duplicate
+% - Isabelle: combines the conciseness of pseudocode with the precision of mathematical notation.
+% - Unlike NFS, can work offline
 

### 2019-06-05 Test.hs
code-gen: fiddling...

@@ -1,65 +1,0 @@
-module Ancestor(Set, ancestor) where {
-
-import Prelude ((==), (/=), (<), (<=), (>=), (>), (+), (-), (*), (/), (**),
-  (>>=), (>>), (=<<), (&&), (||), (^), (^^), (.), ($), ($!), (++), (!!), Eq,
-  error, id, return, not, fst, snd, map, filter, concat, concatMap, reverse,
-  zip, null, takeWhile, dropWhile, all, any, Integer, negate, abs, divMod,
-  String, Bool(True, False), Maybe(Nothing, Just));
-import qualified Prelude;
-
-data Set a = Set [a] | Coset [a];
-
-bex :: forall a. Set a -> (a -> Bool) -> Bool;
-bex (Set xs) p = any p xs;
-
-fold :: forall a b. (a -> b -> b) -> [a] -> b -> b;
-fold f (x : xs) s = fold f xs (f x s);
-fold f [] s = s;
-
-membera :: forall a. (Eq a) => [a] -> a -> Bool;
-membera [] y = False;
-membera (x : xs) y = x == y || membera xs y;
-
-member :: forall a. (Eq a) => a -> Set a -> Bool;
-member x (Coset xs) = not (membera xs x);
-member x (Set xs) = membera xs x;
-
-removeAll :: forall a. (Eq a) => a -> [a] -> [a];
-removeAll x [] = [];
-removeAll x (y : xs) = (if x == y then removeAll x xs else y : removeAll x xs);
-
-inserta :: forall a. (Eq a) => a -> [a] -> [a];
-inserta x xs = (if membera xs x then xs else x : xs);
-
-insert :: forall a. (Eq a) => a -> Set a -> Set a;
-insert x (Coset xs) = Coset (removeAll x xs);
-insert x (Set xs) = Set (inserta x xs);
-
-sup_set :: forall a. (Eq a) => Set a -> Set a -> Set a;
-sup_set (Coset xs) a = Coset (filter (\ x -> not (member x a)) xs);
-sup_set (Set xs) a = fold insert xs a;
-
-bot_set :: forall a. Set a;
-bot_set = Set [];
-
-sup_seta :: forall a. (Eq a) => Set (Set a) -> Set a;
-sup_seta (Set xs) = fold sup_set xs bot_set;
-
-image :: forall a b. (a -> b) -> Set a -> Set b;
-image f (Set xs) = Set (map f xs);
-
[…]

### 2019-06-05 paper/move-op.tex
paper: rest of introduction

--- 2019-05-22 paper/move-op.tex
@@
 
-\renewcommand{\shortauthors}{Kleppmann, Mulligan, Gomes, and Beresford}
+%\renewcommand{\shortauthors}{Kleppmann, Mulligan, Gomes, and Beresford}
+
+\begin{abstract}
+    TODO abstract.
+\end{abstract}
 \maketitle
 
-\begin{abstract}
-\end{abstract}
+\section{Introduction}\label{sec:intro}
 
-\section{Introduction}
+For many systems and applications, the data model that describes the system state is a tree.
+For example:
+\begin{itemize}
+    \item Most \textbf{filesystems}, e.g.
+on Windows and Unix-like OSes, are organised as a tree: folders or directories are branch nodes, and files are leaf nodes.\footnote{In filesystems that support hardlinks, the structure is---strictly speaking---a restricted DAG in which a file inode can be referenced from multiple places in the directory tree, but directories cannot have more than one parent directory.
+        However, for the purposes of this paper, we will treat a filesystem as a tree; given such a tree, it is easy to add hardlinks and symlinks as features on top.}
+        In Unix-style filesystems, the nodes of the tree are known as \emph{inodes}.
+    \item \textbf{Rich text editors} maintain a tree of textual structures such as paragraphs, bulleted and numbered lists, figures, sections, and so on.
+        These can be nested inside each other: for example, a paragraph may appear inside a bullet point of a bulleted list, which may in turn appear inside a numbered point of an enumeration.
+        Rich text documents are typically represented in memory using the Document Object Model (DOM), and written to file in HTML, XML, or JSON format.
+    \item \textbf{Vector graphics} and \textbf{presentation software} represents images using graphical objects such as text boxes, rectangles, ellipses, lines, and so on.
+        These objects are contained within nodes representing pages of a document, or slides of a presentation.
+        Multiple objects may be combined into a \emph{group} so that they can be manipulated as a unit; this corresponds to making these objects children of a common parent node.
+        Multiple objects and groups may in turn be combined into higher-level groups, forming a tree.
+        %Moreover, the order of nodes determines visibility (i.e.\ objects that are ``further back'' may be obscured by objects that are ``closer to the front'').
+    \item Personal \textbf{note-taking} and \textbf{task-management tools} such as Org-mode for Emacs~\cite{OrgMode} or OmniOutliner~\cite{OmniOutliner} explicitly present the user with a tree structure that they can inspect and manipulate.
+\end{itemize}
 
-CRDTs rock \cite{Shapiro:2011wy,Shapiro:2011un}.
+In this paper we consider applications that use such a tree data model, and that replicate this tree across multiple nodes.
+Moreover, we focus on \emph{optimistic replication}~\cite{Saito:2005jw}, that is, systems in which any replica can autonomously make changes to the data, without waiting for communication or coordination with any other replicas.
+Such systems have the advantage that they can continue processing read and write requests even in the presence of arbitrary network partitions; in other words, they are \emph{available} and \emph{partition-tolerant} in the sense of the CAP theorem~\cite{Gilbert:2002il}.
+This approach is desirable since it enables disconnected operation in a mobile computing context, and high availability in geo-replicated settings.
+
+As the tree structure may be concurrently modified on different replicas, the state of these replicas may temporarily diverge.
+In this paper we show how the replicas can nevertheless achieve \emph{strong eventual consistency}~\cite{Shapiro:2011un}: as the replicas communicate, we guarantee that they will converge towards a consistent state.
+Our algorithm for achieving convergence is an example of a \emph{Conflict-free Replicated Data Type} or CRDT~\cite{Shapiro:2011wy,Shapiro:2011un}.
+
+In this paper we use an abstract model of a tree that can support any of the applications listed above, including distributed filesystems.
+We allow replicas to manipulate this tree in any way: by creating new nodes, deleting nodes, or moving subtrees to a new location within the tree.
+While there are many existing systems that support creating and deleting nodes (see Section~\ref{sec:relwork}), the key innovation of our algorithm is the support for moving subtrees.
+We explain in Section~\ref{sec:move-is-hard} why this move operation is so challenging.
+
+Our contributions in this work are as follows:
+\begin{itemize}
+    \item We define a Conflict-free Replicated Data Type for trees that supports a \emph{move} operation, without requiring any coordination between replicas such as locking or consensus.
+        As discussed in Section~\ref{sec:move-is-hard}, this has previously been thought to be impossible to achieve \cite{Najafzadeh:2017vk,Najafzadeh:2018bw}.
+    \item We formalise our algorithm using Isabelle/HOL~\cite{DBLP:conf/tphol/WenzelPN08}, an interactive proof assistant, and produce a computer-checked proof of correctness of our algorithm.
+        In particular, we prove that arbitrary concurrent modifications to the tree can be merged such that all replicas converge towards a consistent state (strong eventual consistency).
+    \item To demonstrate the practical viability of the algorithm we extract a formally verified, executable implementation from the Isabelle/HOL definitions, and perform tests on this implementation.
+    \item We perform experiments with replicated filesystem products such as Dropbox, Google Drive, and Microsoft OneDrive, and show that some of them exhibit problems that would be prevented by our algorithm.
+\end{itemize}
+
+\section{Why a move operation is hard}\label{sec:move-is-hard}
+
+Errors in Google Drive
+
+% - No Synchronisation or consensus required. Available in the sense of CAP
+% - Rely only on uniqueness of node IDs, e.g. inode numbers
+% - Composing JSON/XML tree by using list datatype
+% - Location is Lww register. Concurrent move does not duplicate
+% - Isabelle: combines the conciseness of pseudocode with the precision of mathematical notation.
+% - Unlike NFS, can work offline 
+% - Garbage collection: discard causally stable prefix
 
@@
 
+\section{Related work}\label{sec:relwork}
+
+% Related work:
+% OT implementation of move operation: https://hal.inria.fr/inria-00107740/
+% Ficus (does not discuss move): https://pdos.csail.mit.edu/6.824/papers/ficus.pdf
+% Roam: https://ieeexplore.ieee.org/abstract/document/795151
+
 \begin{acks}

### 2019-06-07 paper/move-op.tex
paper: add figures

@@
 \documentclass[sigconf]{acmart}
+
+\usepackage{tikz}
+\usetikzlibrary{arrows.meta}
 
@@
 
-In this paper we consider applications that use such a tree data model, and that replicate this tree across multiple nodes.
+In this paper we consider applications that use such a tree data model, and that replicate this tree across multiple replicas.
 Moreover, we focus on \emph{optimistic replication}~\cite{Saito:2005jw}, that is, systems in which any replica can autonomously make changes to the data, without waiting for communication or coordination with any other replicas.
@@
     \item To demonstrate the practical viability of the algorithm we extract a formally verified, executable implementation from the Isabelle/HOL definitions, and perform tests on this implementation.
-    \item We perform experiments with replicated filesystem products such as Dropbox, Google Drive, and Microsoft OneDrive, and show that some of them exhibit problems that would be prevented by our algorithm.
+    \item We perform experiments with replicated filesystem products such as Dropbox and Google Drive, and show that they exhibit problems that would be prevented by our algorithm.
 \end{itemize}
+
+\begin{figure*}
+\centering
+\begin{tikzpicture}
+  \tikzstyle{time}=[thick,->,gray]
+  \tikzstyle{network}=[thick,dashed,blue,-{Stealth[length=3mm]}]
+  \node [anchor=east] at (-1.8,3) {Replica 1:};
+  \node [anchor=east] at (-1.8,0) {Replica 2:};
+  \node [rectangle,draw] (start1) at (0,3) {
+      \begin{tikzpicture}[level distance=7mm]
+          \tikzstyle{level 1}=[sibling distance=10mm]
+          \tikzstyle{level 2}=[sibling distance=6mm]
+          \node {$\mathsf{root}$}
+              child {node {$A$} child {node {$a_1$}} child {node {$a_2$}} child {node {$a_3$}}}
+              child {node {$B$}}
+              child {node {$C$}};
+      \end{tikzpicture}
+  };
+  \node [rectangle,draw] (start2) at (0,0) {
+      \begin{tikzpicture}[level distance=7mm]
+          \tikzstyle{level 1}=[sibling distance=10mm]
+          \tikzstyle{level 2}=[sibling distance=6mm]
+          \node {$\mathsf{root}$}
+              child {node {$A$} child {node {$a_1$}} child {node {$a_2$}} child {node {$a_3$}}}
+              child {node {$B$}}
+              child {node {$C$}};
+      \end{tikzpicture}
+  };
+  \node [rectangle,draw] (change1) at (5,3) {
+      \begin{tikzpicture}[level distance=7mm]
+          \tikzstyle{level 1}=[sibling distance=10mm]
+          \tikzstyle{level 3}=[sibling distance=6mm]
+          \node {$\mathsf{root}$}
+              child {node {$B$} child {node {$A$} child {node {$a_1$}} child {node {$a_2$}} child {node {$a_3$}}}}
+              child {node {$C$}};
+      \end{tikzpicture}
+  };
+  \node [rectangle,draw] (change2) at (5,0) {
+      \begin{tikzpicture}[level distance=7mm]
+          \tikzstyle{level 1}=[sibling distance=10mm]
+          \tikzstyle{level 3}=[sibling distance=6mm]
+          \node {$\mathsf{root}$}
+              child {node {$B$}}
+              child {node {$C$} child {node {$A$} child {node {$a_1$}} child {node {$a_2$}} child {node {$a_3$}}}};
+      \end{tikzpicture}
+  };
+  \node [rectangle,draw,inner sep=3mm] (merge1) at (8.2,3) {?};
+  \node [rectangle,draw,inner sep=3mm] (merge2) at (8.2,0) {?};
+  \draw [time] (start1.east) -- node [above,text width=2.5cm,text centered,inner ysep=5pt] {Move $A$ to be a child of $B$} (change1.west);
+  \draw [time] (change1.east) -- (merge1.west);
+  \draw [time] (start2.east) -- node [above,text width=2.5cm,text centered,inner ysep=5pt] {Move $A$ to be a child of $C$} (change2.west);
+  \draw [time] (change2.east) -- (merge2.west);
+  \draw [network] (6.5,0) to [out=90,in=270] (7.2,3);
+  \draw [network] (6.5,3) to [out=270,in=90] (7.2,0);
+  \node [rotate=90,blue,font=\footnotesize] at (7.5,1.5) {network communication};
+  %%%
+  \node at (9.85,4.04) {(\emph{a})};
+  \node at (14.15,4.04) {(\emph{b})};
+  \node at (9.85,1.04) {(\emph{c})};
+  \node at (14.15,1.04) {(\emph{d})};
+  \node [rectangle,draw] at (10.9,3) {
+      \begin{tikzpicture}[level distance=7mm]
+          \tikzstyle{level 1}=[sibling distance=12mm]
+          \tikzstyle{level 3}=[sibling distance=4mm]
+          \node {$\mathsf{root}$}
+              child {node {$B$} child {node {$A$} child {node {$a_1$}} child {node {$a_2$}} child {node {$a_3$}}}}
+              child {node {$C$} child {node {$A$} child {node {$a_1$}} child {node {$a_2$}} child {node {$a_3$}}}};
+      \end{tikzpicture}
+  };
+  \node [rectangle,draw] at (13.5,3) {
+      \begin{tikzpicture}[level distance=7mm]
+          \tikzstyle{level 1}=[sibling distance=12mm]
+          \node at (0,1.4) {$\mathsf{root}$} child {node (b) {$B$}} child {node (c) {$C$}};
+          \tikzstyle{level 1}=[sibling distance=6mm]
+          \node (a) at (0,0) {$A$} child {node {$a_1$}} child {node {$a_2$}} child {node {$a_3$}};
+          \draw (b) -- (a);
+          \draw (c) -- (a);
+      \end{tikzpicture}
+  };
+  \node [rectangle,draw] at (10.7,0) {
+      \begin{tikzpicture}[level distance=7mm]
+          \tikzstyle{level 1}=[sibling distance=10mm]
+          \tikzstyle{level 3}=[sibling distance=6mm]
+          \node {$\mathsf{root}$}
+              child {node {$B$} child {node {$A$} child {node {$a_1$}} child {node {$a_2$}} child {node {$a_3$}}}}
+              child {node {$C$}};
+      \end{tikzpicture}
+  };
+  \node [rectangle,draw] at (13.3,0) {
+      \begin{tikzpicture}[level distance=7mm]
+          \tikzstyle{level 1}=[sibling distance=10mm]
+          \tikzstyle{level 3}=[sibling distance=6mm]
+          \node {$\mathsf{root}$}
+              child {node {$B$}}
+              child {node {$C$} child {node {$A$} child {node {$a_1$}} child {node {$a_2$}} child {node {$a_3$}}}};
+      \end{tikzpicture}
+  };
+\end{tikzpicture}
+\caption{Replica 1 moves $A$ to be a child of $B$, while concurrently replica 2 moves the same node $A$ to be a child of $C$.
+Boxes (\emph{a}) to (\emph{d}) show possible outcomes after the replicas have communicated and merged their states.}
+\label{fig:move-same-item}
+\Description{}
+\end{figure*}
+
+\begin{figure*}
+\centering
+\begin{tikzpicture}
+  \tikzstyle{time}=[thick,->,gray]
+  \tikzstyle{network}=[thick,dashed,blue,-{Stealth[length=3mm]}]
+  \node [anchor=east] at (-1.2,3) {Replica 1:};
+  \node [anchor=east] at (-1.2,0) {Replica 2:};
+  \node [rectangle,draw] (start1) at (0,3) {
+      \begin{tikzpicture}[level distance=7mm]
+      \node {$\mathsf{root}$} child {node {$A$} child {node {$C$}}} child {node {$B$}};
+      \end{tikzpicture}
+  };
+  \node [rectangle,draw] (start2) at (0,0) {
+      \begin{tikzpicture}[level distance=7mm]
+      \node {$\mathsf{root}$} child {node {$A$} child {node {$C$}}} child {node {$B$}};
+      \end{tikzpicture}
+  };
+  \node [rectangle,draw] (change1) at (4.5,3) {
+      \begin{tikzpicture}[level distance=7mm]
+      \node {$\mathsf{root}$} child {node {$A$} child {node {$B$}} child {node {$C$}}};
+      \end{tikzpicture}
+  };
+  \node [rectangle,draw] (change2) at (4.5,0) {
+      \begin{tikzpicture}[level distance=7mm]
+      \node {$\mathsf{root}$} child {node {$B$} child {node {$A$} child {node {$C$}}}};
+      \end{tikzpicture}
+  };
+  \node [rectangle,draw,inner sep=3mm] (merge1) at (8,3) {?};
+  \node [rectangle,draw,inner sep=3mm] (merge2) at (8,0) {?};
+  \draw [time] (start1.east) -- node [above,text width=2.5cm,text centered,inner ysep=5pt] {Move $B$ to be a child of $A$} (change1.west);
+  \draw [time] (change1.east) -- (merge1.west);
+  \draw [time] (start2.east) -- node [above,text width=2.5cm,text centered,inner ysep=5pt] {Move $A$ to be a child of $B$} (change2.west);
+  \draw [time] (change2.east) -- (merge2.west);
+  \draw [network] (6.0,0) to [out=90,in=270] (6.7,3);
+  \draw [network] (6.0,3) to [out=270,in=90] (6.7,0);
+  \node [rotate=90,blue,font=\footnotesize] at (7.0,1.5) {network communication};
+  %%%%%
+  \node [rectangle,draw] at (10.95,3) {
+      \begin{tikzpicture}[level distance=7mm]
+      \useasboundingbox (-1.3,-1.3) rectangle (1,1.2);
+      \node at (-1.1,1.0) {(\emph{a})};
+      \node at (0,1) {$\mathsf{root}$};
+      \node (a1) {$A$} child {node (b1) {$B$}} child {node {$C$}};
+      \draw (b1.south west) .. controls (-2,-2) and (0,1.5) .. (a1.north);
+      \end{tikzpicture}
+  };
+  \node [rectangle,draw] at (13.8,3) {
+      \begin{tikzpicture}[level distance=7mm]
+      \useasboundingbox (-1.15,-2.3) rectangle (1.15,0.2);
+      \node at (0.95,0.0) {(\emph{b})};
+      \node {$\mathsf{root}$} child {node (a2) {$A$} child {node {$B$}}} child {node {$B'$} child {node {$A'$} child {node (c2) {$C$}}}};
+      \draw (a2) -- (c2);
+      \end{tikzpicture}
+  };
+  \node [rectangle,draw] (start) at (10.95,0) {
+      \begin{tikzpicture}[level distance=7mm]
+      \useasboundingbox (-1.15,-2.3) rectangle (1.15,0.2);
+      \node at (-0.95,0.0) {(\emph{c})};
+      \node {$\mathsf{root}$} child {node {$A$} child {node {$B$}} child {node {$C$}}};
+      \end{tikzpicture}
+  };
+  \node [rectangle,draw] (right) at (13.8,0) {
+      \begin{tikzpicture}[level distance=7mm]
+      \useasboundingbox (-1.15,-2.3) rectangle (1.15,0.2);
+      \node at (0.95,0.0) {(\emph{d})};
+      \node {$\mathsf{root}$} child {node {$B$} child {node {$A$} child {node {$C$}}}};
+      \end{tikzpicture}
+  };
+\end{tikzpicture}
+\caption{Initially, $A$ and $B$ are siblings.
+$B$ is moved to be a child of $A$, while concurrently $A$ is moved to be a child of $B$.
+Boxes (\emph{a}) to (\emph{d}) show possible outcomes after the replicas have communicated and merged their states.}\label{fig:move-cycle}
+\Description{Visualisation of trees with concurrent move operations}
+\end{figure*}
+
+\begin{figure*}
+  \centering
+  \includegraphics[width=\textwidth,keepaspectratio=true]{gdrive-error.png}
+  \caption{Error message produced by Google Drive Backup and Sync on Mac OS as a result of performing the operations described in Section~\ref{sec:move-cycle}.}
+  \label{fig:gdrive-error}
+  \Description{Screenshot showing the errors ``Can't remove files with changes pending. Try again after sync is complete.'' and ``An unknown error occurred.''}
+\end{figure*}
 
@@
 
+Applications that rely on a tree data model often need to move a node from one location to another location within the tree, whereby all children of the moved node move along with it.
+For example:
+\begin{itemize}
+    \item In a filesystem, any file or directory can be moved to a different parent directory.
+        Moreover, renaming a file is equivalent to moving the file to a new name without changing its parent directory.
+    \item In a rich text editor, a paragraph node can be turned into a bullet point.
+        This corresponds to creating a new bullet list node, creating a new bullet point node within it, and then moving the paragraph node to be a child of the new bullet point node.
+    \item In presentation software, grouping two graphical objects corresponds to creating a new group node, and then moving the two objects to be children of the new group node.
+\end{itemize}
+As these operations are so common, it is not obvious why a move operation should be difficult in a replicated setting.
+In this section we demonstrate some of those problems by example, before proceeding to our solution in Section~\ref{sec:algorithm}.
+
+\subsection{Concurrent moves of the same node}\label{sec:move-same-item}
+
+The first difficulty arises when a node is moved to a new location on one replica, and concurrently the same node is moved to a different location on a different replica.
+
+\subsection{Moving a node to be a descendant of itself}\label{sec:move-cycle}
+
 Errors in Google Drive
+
+\section{The replicated move operation}\label{sec:algorithm}
 
@@
 \label{fig:code}
+\Description{The move operation algorithm, implemented in the Isabelle/HOL language.}
 \end{figure*}

### 2019-06-10 paper/move-op.tex
paper: more background

@@
   \node [rotate=90,blue,font=\footnotesize] at (7.5,1.5) {network communication};
+  \path [draw,dotted] (-3.1,1.5) -- (8.7,1.5);
   %%%
-  \node at (9.85,4.04) {(\emph{a})};
-  \node at (14.15,4.04) {(\emph{b})};
-  \node at (9.85,1.04) {(\emph{c})};
-  \node at (14.15,1.04) {(\emph{d})};
+  \node [fill=red!15] at (9.83,4.10) {(\emph{a})};
+  \node [fill=red!15] at (14.18,4.10) {(\emph{b})};
+  \node [fill=red!15] at (9.84,1.10) {(\emph{c})};
+  \node [fill=red!15] at (14.15,1.10) {(\emph{d})};
   \node [rectangle,draw] at (10.9,3) {
       \begin{tikzpicture}[level distance=7mm]
+          \tikzstyle{every node}=[text height=5pt,text depth=0pt]
           \tikzstyle{level 1}=[sibling distance=12mm]
@@
               child {node {$B$} child {node {$A$} child {node {$a_1$}} child {node {$a_2$}} child {node {$a_3$}}}}
-              child {node {$C$} child {node {$A$} child {node {$a_1$}} child {node {$a_2$}} child {node {$a_3$}}}};
+              child {node {$C$} child {node {$A'$} child {node {$a_1'$}} child {node {$a_2'$}} child {node {$a_3'$}}}};
       \end{tikzpicture}
@@
   \node [rotate=90,blue,font=\footnotesize] at (7.0,1.5) {network communication};
+  \path [draw,dotted] (-2.5,1.6) -- (8.5,1.6);
   %%%%%
@@
       \useasboundingbox (-1.3,-1.3) rectangle (1,1.2);
-      \node at (-1.1,1.0) {(\emph{a})};
+      \node [fill=red!15] at (-1.12,1.05) {(\emph{a})};
       \node at (0,1) {$\mathsf{root}$};
@@
       \useasboundingbox (-1.15,-2.3) rectangle (1.15,0.2);
-      \node at (0.95,0.0) {(\emph{b})};
+      \node [fill=red!15] at (0.97,0.05) {(\emph{b})};
       \node {$\mathsf{root}$} child {node (a2) {$A$} child {node {$B$}}} child {node {$B'$} child {node {$A'$} child {node (c2) {$C$}}}};
@@
       \begin{tikzpicture}[level distance=7mm]
-      \useasboundingbox (-1.15,-2.3) rectangle (1.15,0.2);
-      \node at (-0.95,0.0) {(\emph{c})};
+      \useasboundingbox (-1.17,-2.3) rectangle (1.15,0.2);
+      \node [fill=red!15] at (-1.00,0.05) {(\emph{c})};
       \node {$\mathsf{root}$} child {node {$A$} child {node {$B$}} child {node {$C$}}};
@@
       \useasboundingbox (-1.15,-2.3) rectangle (1.15,0.2);
-      \node at (0.95,0.0) {(\emph{d})};
+      \node [fill=red!15] at (0.96,0.05) {(\emph{d})};
       \node {$\mathsf{root}$} child {node {$B$} child {node {$A$} child {node {$C$}}}};
@@
 \end{tikzpicture}
-\caption{Initially, $A$ and $B$ are siblings.
-$B$ is moved to be a child of $A$, while concurrently $A$ is moved to be a child of $B$.
+\caption{Initially, nodes $A$ and $B$ are siblings.
+Replica 1 moves $B$ to be a child of $A$, while concurrently replica 2 moves $A$ to be a child of $B$.
 Boxes (\emph{a}) to (\emph{d}) show possible outcomes after the replicas have communicated and merged their states.}\label{fig:move-cycle}
@@
     \item In a filesystem, any file or directory can be moved to a different parent directory.
-        Moreover, renaming a file is equivalent to moving the file to a new name without changing its parent directory.
+        Moreover, renaming a file or directory is equivalent to moving it to a new name without changing its parent directory.
     \item In a rich text editor, a paragraph node can be turned into a bullet point.
-        This corresponds to creating a new bullet list node, creating a new bullet point node within it, and then moving the paragraph node to be a child of the new bullet point node.
+        In the DOM tree this corresponds to creating a new bullet list node, creating a new bullet point node within it, and then moving the paragraph node to be a child of the new bullet point node.
     \item In presentation software, grouping two graphical objects corresponds to creating a new group node, and then moving the two objects to be children of the new group node.
@@
 As these operations are so common, it is not obvious why a move operation should be difficult in a replicated setting.
-In this section we demonstrate some of those problems by example, before proceeding to our solution in Section~\ref{sec:algorithm}.
+In this section we demonstrate by example some of the problems that arise with replicated trees, before proceeding to our solution in Section~\ref{sec:algorithm}.
 
@@
 The first difficulty arises when a node is moved to a new location on one replica, and concurrently the same node is moved to a different location on a different replica.
+This scenario is illustrated in Figure~\ref{fig:move-same-item}, where replica 1 moves node $A$ to be a child of $B$, while concurrently replica 2 moves $A$ to be a child of $C$.
+After the replicas communicate these operations to each other, what should the merged state of the tree be?
+
+If a move operation is implemented by deleting the moved subtree from its old location, and then re-creating it at the new location, the merged state will be as shown in Figure~\ref{fig:move-same-item}a: the concurrent moves will duplicate the moved subtree, since each concurrent move independently recreates the subtree in each destination location.
+We believe that this duplication is undesirable, since subsequent edits to nodes in the duplicated subtree will apply to only one of the copies.
+Two users who believe they are collaborating on the same file may in fact be editing two different copies, which will then become inconsistent with each other.
+
+Rather than duplicating the moved subtree, another possible resolution is for the destination locations of both moves to refer to the same node, as shown in Figure~\ref{fig:move-same-item}b.
+However, the result is a DAG, not a tree.
+POSIX filesystems to not allow this outcome, since they allow hardlinks only to files, but not to directories.
+
+In our opinion, the only reasonable outcomes are those shown in Figure~\ref{fig:move-same-item}c and~\ref{fig:move-same-item}d: the moved subtree appears either in replica 1's destination location or in replica 2's destination location, but not in both.
+Which one of these two is picked is arbitrary, due to the symmetry between the two replicas.
+For example, the ``winning'' location could be picked based on a timestamp in the operations, similarly to the ``last writer wins'' conflict resolution method of Thomas's write rule (TODO citation).
+All replicas pick the same operation as winner, and effectively ignore the conflicting operation.
+
+We tested this scenario with file sync products Dropbox and Google Drive by concurrently moving the same directory to two different destination directories.\footnote{Experiment setup: we installed the official Mac OS clients for Dropbox and Google Drive on two computers, and configured them to each sync a directory on the local filesystem with the corresponding cloud service.
+To test concurrent operations, we turned off the network interface on both computers, performed a move operation on the local filesystem of each computer, then reconnected both computers to the Internet and waited for them to sync.}
+Dropbox exhibited the undesirable duplication behaviour of Figure~\ref{fig:move-same-item}a, while the outcome on Google Drive was as in Figure~\ref{fig:move-same-item}c/d.
 
 \subsection{Moving a node to be a descendant of itself}\label{sec:move-cycle}
+
+On a filesystem, the destination directory of a move operation must not be a subdirectory of the directory being moved.
+For example, if \texttt{b} is a subdirectory of \texttt{a}, then the Unix shell command \texttt{mv a a/b/} will fail with an error.
+This restriction is required because allowing this operation would introduce a cycle into the directory graph, and so the filesystem would no longer be a tree.
+The same restriction is required in any other tree structure that supports a move operation.
+
+In an unreplicated tree it is easy to prevent cycles being introduced: if the node being moved is an ancestor of the destination node, the move operation must be rejected.
+However, in a replicated setting a new problem arises: different replicas may perform operations that individually are safe, but whose combination leads to a cycle.
+
+One example of such a scenario is illustrated in Figure~\ref{fig:move-cycle}.
+Here, replica 1 moves $B$ to be a child of $A$, while concurrently replica 2 moves $A$ to be a child of $B$.
+As each replica propagates its operation to the other replica, a careless implementation might end up in the state shown in Figure~\ref{fig:move-cycle}a, in which $A$ and $B$ have become locked in a cycle and detached from the tree.
 

### 2019-06-10 paper/move-op.tex
Finish section 2

@@
     \item We define a Conflict-free Replicated Data Type for trees that supports a \emph{move} operation, without requiring any coordination between replicas such as locking or consensus.
-        As discussed in Section~\ref{sec:move-is-hard}, this has previously been thought to be impossible to achieve \cite{Najafzadeh:2017vk,Najafzadeh:2018bw}.
+        As discussed in Section~\ref{sec:impossibility}, this has previously been thought to be impossible to achieve \cite{Najafzadeh:2017vk,Najafzadeh:2018bw}.
     \item We formalise our algorithm using Isabelle/HOL~\cite{DBLP:conf/tphol/WenzelPN08}, an interactive proof assistant, and produce a computer-checked proof of correctness of our algorithm.
@@
       \node [fill=red!15] at (0.97,0.05) {(\emph{b})};
-      \node {$\mathsf{root}$} child {node (a2) {$A$} child {node {$B$}}} child {node {$B'$} child {node {$A'$} child {node (c2) {$C$}}}};
-      \draw (a2) -- (c2);
+      \tikzstyle{level 1}=[sibling distance=12mm]
+      \tikzstyle{level 2}=[sibling distance=6mm]
+      \node {$\mathsf{root}$}
+          child {node {$A$} child {node {$B$}} child {node {$C$}}}
+          child {node {$B'$} child {node {$A'$} child {node {$C'$}}}};
       \end{tikzpicture}
@@
   \includegraphics[width=\textwidth,keepaspectratio=true]{gdrive-error.png}
-  \caption{Error message produced by Google Drive Backup and Sync on Mac OS as a result of performing the operations described in Section~\ref{sec:move-cycle}.}
+  \caption{Error message produced by Google Drive Backup and Sync on Mac OS as a result of performing the operations shown in Figure~\ref{fig:move-cycle}.}
   \label{fig:gdrive-error}
@@
 Which one of these two is picked is arbitrary, due to the symmetry between the two replicas.
-For example, the ``winning'' location could be picked based on a timestamp in the operations, similarly to the ``last writer wins'' conflict resolution method of Thomas's write rule (TODO citation).
+For example, the ``winning'' location could be picked based on a timestamp in the operations, similarly to the ``last writer wins'' conflict resolution method of Thomas's write rule~\cite{Johnson:1975we}.
 All replicas pick the same operation as winner, and effectively ignore the conflicting operation.
@@
 
-In an unreplicated tree it is easy to prevent cycles being introduced: if the node being moved is an ancestor of the destination node, the move operation must be rejected.
+In an unreplicated tree it is easy to prevent cycles being introduced: if the node being moved is an ancestor of the destination node, then the move operation must be rejected.
 However, in a replicated setting a new problem arises: different replicas may perform operations that individually are safe, but whose combination leads to a cycle.
@@
 
-Errors in Google Drive
+Another possible resolution is shown in Figure~\ref{fig:move-cycle}b: the nodes involved in the concurrent moves (and their children) could be duplicated, so that both ``$A$ as a child of $B$'' and ``$B$ as a child of $A$'' can exist in the tree.
+However, such duplication is undesirable for the same reasons as in Section~\ref{sec:move-same-item}.
+
+In our opinion, the best way of handling the conflicting operations of Figure~\ref{fig:move-cycle} is to choose either Figure~\ref{fig:move-cycle}c or~\ref{fig:move-cycle}d: that is, either the result of applying replica 1's operation and ignoring replica 2's operation, or vice versa.
+Like in Section~\ref{sec:move-same-item}, the winning operation can be picked based on an operation timestamp.
+
+As before, we tested this scenario with Google Drive and Dropbox.
+In Google Drive, one replica was able to successfully sync with the server, while the other replica displayed the ``unknown error'' message shown in Figure~\ref{fig:gdrive-error}.
+The replica in error state refused to sync the conflicting directory, and its filesystem state remained inconsistent with the other replica.
+This error state persisted until the directories on the erroring replica were manually moved to match the state of the other replica.
+
+On the other hand, Dropbox exhibits the duplication behaviour shown in Figure~\ref{fig:move-cycle}b.
+
+\subsection{The impossibility of a highly-available move operation}\label{sec:impossibility}
+
+Najafzadeh et al.~\cite{Najafzadeh:2017vk,Najafzadeh:2018bw} have previously implemented a replicated filesystem with a move operation, and analysed the case of concurrent move operations introducing a cycle, as in Section~\ref{sec:move-cycle}.
+Using the CISE ('Cause I'm Strong Enough) proof tool~\cite{DBLP:conf/popl/GotsmanYFNS16,Najafzadeh:2016fi} the authors confirm that it is not sufficient for the replica that generates a move operation to check whether the operation introduces a cycle: like in Figure~\ref{fig:move-cycle}, two concurrent operations may be safe individually, but introduce a cycle when combined.
+
+Najafzadeh et al.
+propose two solutions to this problem: either to duplicate tree nodes, as in Figure~\ref{fig:move-cycle}b, or to execute a synchronous locking protocol that prevents two move operations concurrently modifying the same part of the tree.
+The downside of a locking protocol is that the move operation is no longer highly available in the presence of network partitions, since it must wait for communication with other replicas.
+
+While these solutions are valid, the authors go on to claim that these are the only possible options, asserting that ``no file system can support an unsynchronised move without anomalies, such as loss or duplication''~\cite{Najafzadeh:2018bw}.
+We revise that claim in this paper: our algorithm does not perform any locking, coordination or synchronisation among replicas, but it nevertheless ensures that the tree invariants are always satisified (in particular, it never introduces cycles), and it never duplicates any tree nodes.
+To our knowledge, our algorithm is the first to provide all of these properties simultaneously.
+We give a precise specification of our algorithm's consistency properties in Section~\ref{sec:proof}.
 
@@
 
+\section{Proof of correctness}\label{sec:proof}
+
 \section{Related work}\label{sec:relwork}

### 2019-06-10 paper/move-op.tex
paper: start describing algorithm

@@
 In an unreplicated tree it is easy to prevent cycles being introduced: if the node being moved is an ancestor of the destination node, then the move operation must be rejected.
-However, in a replicated setting a new problem arises: different replicas may perform operations that individually are safe, but whose combination leads to a cycle.
+However, a new problem arises in a replicated setting: different replicas may perform operations that individually are safe, but whose combination leads to a cycle.
 
@@
 Najafzadeh et al.
-propose two solutions to this problem: either to duplicate tree nodes, as in Figure~\ref{fig:move-cycle}b, or to execute a synchronous locking protocol that prevents two move operations concurrently modifying the same part of the tree.
+propose two solutions to this problem: either to duplicate tree nodes, as in Figure~\ref{fig:move-cycle}b, or to execute a synchronous locking protocol that prevents two move operations from concurrently modifying the same part of the tree.
 The downside of a locking protocol is that the move operation is no longer highly available in the presence of network partitions, since it must wait for communication with other replicas.
 
-While these solutions are valid, the authors go on to claim that these are the only possible options, asserting that ``no file system can support an unsynchronised move without anomalies, such as loss or duplication''~\cite{Najafzadeh:2018bw}.
-We revise that claim in this paper: our algorithm does not perform any locking, coordination or synchronisation among replicas, but it nevertheless ensures that the tree invariants are always satisified (in particular, it never introduces cycles), and it never duplicates any tree nodes.
+While these solutions are valid, the authors go on to claim that ``no file system can support an unsynchronised move without anomalies, such as loss or duplication''~\cite{Najafzadeh:2018bw}.
+We revisit that claim in this paper: our algorithm does not perform any locking, coordination or synchronisation among replicas, but it nevertheless ensures that the tree invariants are always satisified (in particular, it never introduces cycles), and it never duplicates or loses any tree nodes.
 To our knowledge, our algorithm is the first to provide all of these properties simultaneously.
 We give a precise specification of our algorithm's consistency properties in Section~\ref{sec:proof}.
-
-\section{The replicated move operation}\label{sec:algorithm}
-
-% - No Synchronisation or consensus required. Available in the sense of CAP
-% - Rely only on uniqueness of node IDs, e.g. inode numbers
-% - Composing JSON/XML tree by using list datatype
-% - Location is Lww register. Concurrent move does not duplicate
-% - Isabelle: combines the conciseness of pseudocode with the precision of mathematical notation.
-% - Unlike NFS, can work offline 
-% - Garbage collection: discard causally stable prefix
 
@@
 \isacommand{fun}\isamarkupfalse%
-\ interp{\isacharunderscore}op\ {\isacharcolon}{\isacharcolon}\ {\isacartoucheopen}{\isacharparenleft}{\isacharprime}t{\isacharcolon}{\isacharcolon}{\isacharbraceleft}linorder{\isacharbraceright}{\isacharcomma}\ {\isacharprime}n{\isacharcomma}\ {\isacharprime}m{\isacharparenright}\ operation\ {\isasymRightarrow}\isanewline
+\ apply{\isacharunderscore}op\ {\isacharcolon}{\isacharcolon}\ {\isacartoucheopen}{\isacharparenleft}{\isacharprime}t{\isacharcolon}{\isacharcolon}{\isacharbraceleft}linorder{\isacharbraceright}{\isacharcomma}\ {\isacharprime}n{\isacharcomma}\ {\isacharprime}m{\isacharparenright}\ operation\ {\isasymRightarrow}\isanewline
 \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ {\isacharparenleft}{\isacharprime}t{\isacharcomma}\ {\isacharprime}n{\isacharcomma}\ {\isacharprime}m{\isacharparenright}\ state\ {\isasymRightarrow}\ {\isacharparenleft}{\isacharprime}t{\isacharcomma}\ {\isacharprime}n{\isacharcomma}\ {\isacharprime}m{\isacharparenright}\ state{\isacartoucheclose}\ \isakeyword{where}\isanewline
-\ \ {\isacartoucheopen}interp{\isacharunderscore}op\ op{\isadigit{1}}\ {\isacharparenleft}{\isacharbrackleft}{\isacharbrackright}{\isacharcomma}\ tree{\isadigit{1}}{\isacharparenright}\ {\isacharequal}\isanewline
+\ \ {\isacartoucheopen}apply{\isacharunderscore}op\ op{\isadigit{1}}\ {\isacharparenleft}{\isacharbrackleft}{\isacharbrackright}{\isacharcomma}\ tree{\isadigit{1}}{\isacharparenright}\ {\isacharequal}\isanewline
 \ \ \ \ \ {\isacharparenleft}let\ {\isacharparenleft}op{\isadigit{2}}{\isacharcomma}\ tree{\isadigit{2}}{\isacharparenright}\ {\isacharequal}\ do{\isacharunderscore}op\ {\isacharparenleft}op{\isadigit{1}}{\isacharcomma}\ tree{\isadigit{1}}{\isacharparenright}\isanewline
 \ \ \ \ \ \ in\ {\isacharparenleft}{\isacharbrackleft}op{\isadigit{2}}{\isacharbrackright}{\isacharcomma}\ tree{\isadigit{2}}{\isacharparenright}{\isacharparenright}{\isacartoucheclose}\ {\isacharbar}\isanewline
-\ \ {\isacartoucheopen}interp{\isacharunderscore}op\ op{\isadigit{1}}\ {\isacharparenleft}logop\ {\isacharhash}\ ops{\isacharcomma}\ tree{\isadigit{1}}{\isacharparenright}\ {\isacharequal}\isanewline
+\ \ {\isacartoucheopen}apply{\isacharunderscore}op\ op{\isadigit{1}}\ {\isacharparenleft}logop\ {\isacharhash}\ ops{\isacharcomma}\ tree{\isadigit{1}}{\isacharparenright}\ {\isacharequal}\isanewline
 \ \ \ \ \ {\isacharparenleft}if\ move{\isacharunderscore}time\ op{\isadigit{1}}\ {\isacharless}\ log{\isacharunderscore}time\ logop\isanewline
-\ \ \ \ \ \ then\ redo{\isacharunderscore}op\ logop\ {\isacharparenleft}interp{\isacharunderscore}op\ op{\isadigit{1}}\ {\isacharparenleft}ops{\isacharcomma}\ undo{\isacharunderscore}op\ {\isacharparenleft}logop{\isacharcomma}\ tree{\isadigit{1}}{\isacharparenright}{\isacharparenright}{\isacharparenright}\isanewline
+\ \ \ \ \ \ then\ redo{\isacharunderscore}op\ logop\ {\isacharparenleft}apply{\isacharunderscore}op\ op{\isadigit{1}}\ {\isacharparenleft}ops{\isacharcomma}\ undo{\isacharunderscore}op\ {\isacharparenleft}logop{\isacharcomma}\ tree{\isadigit{1}}{\isacharparenright}{\isacharparenright}{\isacharparenright}\isanewline
 \ \ \ \ \ \ else\ let\ {\isacharparenleft}op{\isadigit{2}}{\isacharcomma}\ tree{\isadigit{2}}{\isacharparenright}\ {\isacharequal}\ do{\isacharunderscore}op\ {\isacharparenleft}op{\isadigit{1}}{\isacharcomma}\ tree{\isadigit{1}}{\isacharparenright}\ in\ {\isacharparenleft}op{\isadigit{2}}\ {\isacharhash}\ logop\ {\isacharhash}\ ops{\isacharcomma}\ tree{\isadigit{2}}{\isacharparenright}{\isacharparenright}{\isacartoucheclose}\isanewline
@@
 \isacommand{definition}\isamarkupfalse%
-\ interp{\isacharunderscore}ops\ {\isacharcolon}{\isacharcolon}\ {\isacartoucheopen}{\isacharparenleft}{\isacharprime}t{\isacharcolon}{\isacharcolon}{\isacharbraceleft}linorder{\isacharbraceright}{\isacharcomma}\ {\isacharprime}n{\isacharcomma}\ {\isacharprime}m{\isacharparenright}\ operation\ list\ {\isasymRightarrow}\ {\isacharparenleft}{\isacharprime}t{\isacharcomma}\ {\isacharprime}n{\isacharcomma}\ {\isacharprime}m{\isacharparenright}\ state{\isacartoucheclose}\ \isakeyword{where}\isanewline
-\ \ {\isacartoucheopen}interp{\isacharunderscore}ops\ ops\ {\isasymequiv}\ foldl\ {\isacharparenleft}{\isasymlambda}state\ oper{\isachardot}\ interp{\isacharunderscore}op\ oper\ state{\isacharparenright}\ {\isacharparenleft}{\isacharbrackleft}{\isacharbrackright}{\isacharcomma}\ {\isacharbraceleft}{\isacharbraceright}{\isacharparenright}\ ops{\isacartoucheclose}\isanewline
+\ apply{\isacharunderscore}ops\ {\isacharcolon}{\isacharcolon}\ {\isacartoucheopen}{\isacharparenleft}{\isacharprime}t{\isacharcolon}{\isacharcolon}{\isacharbraceleft}linorder{\isacharbraceright}{\isacharcomma}\ {\isacharprime}n{\isacharcomma}\ {\isacharprime}m{\isacharparenright}\ operation\ list\ {\isasymRightarrow}\ {\isacharparenleft}{\isacharprime}t{\isacharcomma}\ {\isacharprime}n{\isacharcomma}\ {\isacharprime}m{\isacharparenright}\ state{\isacartoucheclose}\ \isakeyword{where}\isanewline
+\ \ {\isacartoucheopen}apply{\isacharunderscore}ops\ ops\ {\isasymequiv}\ foldl\ {\isacharparenleft}{\isasymlambda}state\ oper{\isachardot}\ apply{\isacharunderscore}op\ oper\ state{\isacharparenright}\ {\isacharparenleft}{\isacharbrackleft}{\isacharbrackright}{\isacharcomma}\ {\isacharbraceleft}{\isacharbraceright}{\isacharparenright}\ ops{\isacartoucheclose}\isanewline
 \isanewline
@@
 
+\section{The replicated move operation}\label{sec:algorithm}
+
+We now introduce our algorithm for a replicated tree data structure that supports a move operation.
+We model each replica as a state machine that transitions from one state to the next by applying an operation.
+The algorithm is executed independently on each replica with no shared memory between replicas.
+
+When the user makes a change to the tree, that change is expressed by generating an operation on their local replica.
+We assume the existence of a messaging middleware that copies operations between replicas, so that eventually every operation is applied on every replica, provided that network partitions are of a finite duration.
+This middleware may deliver operations in any order.
+Operations received from a remote replica are applied in exactly the same way as operations generated by a local user.
+
+The key consistency property of our algorithm is \emph{convergence}: that is, whenever any two replicas have applied the same set of operations, then they must be in the same state---even if the operations were applied in a different order on different replicas.
+We prove this in Section~\ref{sec:proof} by showing that with our algorithm, applying operations is commutative.
+
+Figure~\ref{fig:code} gives the full source code for our algorithm, implemented using the Isabelle/HOL language~\cite{DBLP:conf/tphol/WenzelPN08}.
+We choose this language because it combines the conciseness of pseudocode with the precision of mathematical notation.
+It supports formal reasoning, allowing us to prove the correctness of the algorithm (Section~\ref{sec:proof}), and it also allows executable code to be exported to common functional programming languages such as Scala, Haskell, OCaml, or Standard ML (Section~\ref{sec:evaluation}).
+
+In this section we walk through the code in Figure~\ref{fig:code} step by step, explaining features of the Isabelle/HOL syntax as we encounter them.
+For additional background on the language, standard documentation is available~\cite{DBLP:books/sp/NipkowK14}.
+
+\subsection{Operations and trees}\label{sec:ops-trees}
+
+We define only one kind of operation: \isa{Move t p m c} (Figure~\ref{fig:code}, lines 1--5).
+A move operation is a record consisting of four fields: a timestamp \isa{t} of type \isa{'t}, a parent node \isa{p} of type \isa{'n}, a metadata field \isa{m} of type \isa{'m}, and a child node \isa{c} of type \isa{'n}.
+Here, \isa{'t}, \isa{'n} and \isa{'m} are \emph{type variables} that can be instantiated with arbitrary datatypes; our only requirement is that timestamps \isa{'t} are globally unique and have a total order.
+For example, Lamport timestamps~\cite{Lamport:1978jq} meet these requirements.
+
+The meaning of an operation \isa{Move t p m c} is that at time \isa{t}, the child node \isa{c} is moved to be a child of parent node \isa{p}.
+The operation does not specify the old location of \isa{c}; the algorithm simply ``steals'' \isa{c} from wherever it is currently located in the tree, and moves it to \isa{p}.
+If \isa{c} does not currently exist in the tree, it is created as a child of \isa{p}.
+For this reason we do not need a separate operation type to create tree nodes: simply performing a move with a fresh child node \isa{c} is sufficient to create \isa{c}.
+Moreover, we do not define a remove operation: instead, we can designate a particular tree node as the ``trash'', and perform deletion of a node \isa{c} by moving \isa{c} to be a child of the trash node.
+
+The metadata field \isa{m} in a move operation allows additional information to be associated with the parent-child relationship of \isa{p} and \isa{c}.
+For example, in a filesystem, the parent and child would be the inodes of a directory and a file within it, respectively, and the metadata would contain the filename of the child.
+Thus, a file with inode \isa{c} can be renamed by performing a \isa{Move t p m c}, where the new parent directory \isa{p} is the unchanged inode of the existing parent directory, but the metadata \isa{m} contains the new filename.
+
+When users want to make changes to the tree on their local replica, they generate new \isa{Move t p m c} operations for these changes, and apply these operations to their local replica.
+The timestamp \isa{t} of each operation must be globally unique; the other fields \isa{p}, \isa{m} and \isa{c} can be chosen freely.
+
+We can now represent the tree as a set of \isa{(parent, meta, child)} triples, denoted in Isabelle as \isa{('n {\isasymtimes} 'm {\isasymtimes} 'n) set}.
+When we have \isa{(p, m, c) {\isasymin} tree}, that means \isa{c} is a child of \isa{p} in the tree, with associated metadata \isa{m}.
+Given a \isa{tree}, we can move the child \isa{c} to a new parent \isa{p}, with associated metadata \isa{m}, as follows:
+\begin{quote}
+\isa{\{(p', m', c') {\isasymin} tree. c' {\isasymnoteq} c\} {\isasymunion} \{(p, m, c)\}}
+\end{quote}
+That is, we remove any existing parent-child relationship for \isa{c} from the set \isa{tree}, and then add \isa{\{(p, m, c)\}} to represent the new parent-child relationship.
+This expression appears on lines 31 and 36 of Figure~\ref{fig:code}, as we shall explain shortly.
+
+\subsection{Replica state and operation log}
+
+In order to correctly apply move operations, a replica needs to maintain not only the current state of the tree, but also an \emph{operation log}.
+The log is a linked list of \isa{LogMove} records in descending timestamp order.
+A \isa{LogMove t oldp p m c} record (lines 7--12) is similar to a record \isa{Move t p m c}; the only difference is that the \isa{LogMove} has an additional field \isa{oldp} of type \isa{('n {\isasymtimes} 'm) option}.
+This \isa{option} type means the field can either take the value \isa{None} (equivalent to null), or a pair of an inode and a metadata field.
+
+When a replica applies a \isa{Move} operation to its tree, it also records a corresponding \isa{LogMove} operation in its log.
+The \isa{t}, \isa{p}, \isa{m} and \isa{c} fields are taken directly from the \isa{Move} record, while the \isa{oldp} field is filled in based on the state of the tree before the move.
+If \isa{c} did not exist in the tree prior to the move, \isa{oldp} is set to \isa{None}.
+Otherwise, \isa{oldp} records the previous parent and metadata of \isa{c}, that is: if there exist \isa{p'}, \isa{m'} such that \isa{(p', m', c) {\isasymin} tree}, then \isa{oldp} is set to \isa{Some (p', m')}.
+This logic is performed by the \isa{get\_parent} function (lines 16--20).
+
+In the first line of \isa{get\_parent}, the expression between \isa{::} and \textbf{where} is the type signature of the function, in this case:
+\begin{quote}
+\isa{('n {\isasymtimes} 'm {\isasymtimes} 'n) set {\isasymRightarrow} 'n {\isasymRightarrow} ('n {\isasymtimes} 'm) option}
+\end{quote}
+This signature denotes a function that takes two arguments: a tree \isa{('n {\isasymtimes} 'm {\isasymtimes} 'n) set} and a node \isa{'n}.
+It then returns a \isa{('n {\isasymtimes} 'm) option}.
+\isa{{\isasymexists}!x} means ``there exists a unique value \isa{x} such that\dots''.
+
+In line 14 we define the datatype for the state of a replica: a pair \isa{(log, tree)} where \isa{log} is a list of \isa{LogMove} records, and \isa{tree} is a set of \isa{(parent, meta, child)} triples as before.
+
+\subsection{Applying operations}
+
+% - No Synchronisation or consensus required. Available in the sense of CAP
+% - Rely only on uniqueness of node IDs, e.g. inode numbers
+% - Composing JSON/XML tree by using list datatype
+% - Location is Lww register. Concurrent move does not duplicate
+% - Unlike NFS, can work offline 
+% - Garbage collection: discard causally stable prefix
+
 \section{Proof of correctness}\label{sec:proof}
+
+\section{Evaluation}\label{sec:evaluation}
 

### 2019-06-11 paper/move-op.tex
paper: rest of section 3

@@
 
-\subsection{Replica state and operation log}
+\subsection{Replica state and operation log}\label{sec:state-log}
 
 In order to correctly apply move operations, a replica needs to maintain not only the current state of the tree, but also an \emph{operation log}.
-The log is a linked list of \isa{LogMove} records in descending timestamp order.
+The log is a list of \isa{LogMove} records in descending timestamp order.
 A \isa{LogMove t oldp p m c} record (lines 7--12) is similar to a record \isa{Move t p m c}; the only difference is that the \isa{LogMove} has an additional field \isa{oldp} of type \isa{('n {\isasymtimes} 'm) option}.
@@
 
-\subsection{Applying operations}
+\subsection{Preventing cycles}\label{sec:prevent-cycles}
+
+Recall from Section~\ref{sec:move-cycle} that in order to prevent a cycle being introduced, the node being moved must not be an ancestor of the destination node.
+To implement this in Figure~\ref{fig:code}, we first define the \isa{ancestor} relation in lines 22--24.
+This relation is the transitive closure of the parent-child relation of a tree: if \isa{(p, m, c) {\isasymin} tree} then \isa{p} is an ancestor of \isa{c} (line 23); moreover, if \isa{(p, m, c) {\isasymin} tree} and \isa{a} is an ancestor of \isa{p}, then \isa{a} is also an ancestor of \isa{c} (line 24).
+The \textbf{inductive} keyword indicates that this recursive definition is iterated until a fixed point is reached.
+
+The \isa{do\_op} function (lines 26--31) now performs the actual work of applying a move operation.
+This function takes as argument a pair consisting of a \isa{Move} operation and the current tree, and it returns a pair consisting of a \isa{LogMove} operation (which will be added to the log) and an updated tree.
+In line 29, the \isa{LogMove} record is constructed as described in Section~\ref{sec:state-log}, obtaining the prior parent and metadata of \isa{c} using the \isa{get\_parent} function.
+
+Line 30 performs the check that ensures no cycles are introduced: if \isa{ancestor tree c newp}, i.e.\ if the node being moved \isa{c} is an ancestor of the new parent \isa{newp}, then the tree is returned unmodified---in other words, the operation is ignored.
+Similarly, if \isa{c = newp}, the operation is also ignored.
+Otherwise (line 31), the tree is updated by removing \isa{c} from its existing parent, if any, and adding the new parent-child relationship \isa{(newp, m, c)} to the tree.
+
+\subsection{Applying operations in any order}\label{sec:applying}
+
+The \isa{do\_op} function is sufficient for applying operations if all replicas apply operations in the same order.
+However, in an optimistic replication setting, each replica may apply the operations in a different order, and we need to ensure that the replica state nevertheless converges towards a consistent state.
+This goal is accomplished by the \isa{undo\_op}, \isa{redo\_op}, and \isa{apply\_op} functions (lines 33--51).
+
+The intuition behind these functions is as follows: when a replica needs to apply an operation with timestamp \isa{t}, it first undoes the effect of any operations with a timestamp greater than \isa{t}, then performs the new operation, and finally re-applies the previously undone operations.
+The result of this procedure is that the state of the tree is as if the operations had all been applied in order of increasing timestamp, even though in fact they might have been applied in any order.
+
+The \isa{apply\_op} function (lines 43--51) takes two arguments: a \isa{Move} operation to apply and the current replica state; and it returns the new replica state.
+The expression \isa{'t::\{linorder\}} in the type signature indicates that the timestamps \isa{'t} belong to the type class \isa{linorder}, which means that timestamps can be compared with the \isa{<} operator defining a total order.
+This comparison occurs on line 49.
+
+Recall that the replica state includes the operation log (line 14), and we use this log to perform the undo-do-redo cycle.
+Lines 45--47 handle the case where the log is empty: in this case, we simply perform the operation using \isa{do\_op}, and return the new tree along with a log containing a single \isa{LogMove} record.
+
+If the log is nonempty (line 48), we take \isa{logop} to be the first element of the log, and \isa{ops} to be the rest.
+(The hash character in \isa{logop {\isacharhash} ops} is the \emph{cons} operator that adds one element to the head of a list.)
+If the timestamp of \isa{logop} is greater than the timestamp of the new operation (line 49) we first undo \isa{logop} with \isa{undo\_op}, then recursively apply the new operation to the remaining log, and finally reapply \isa{logop} with \isa{redo\_op} (line 50).
+Otherwise we perform the operation using \isa{do\_op}, and add the corresponding \isa{LogMove} record as the head of the log (line 51).
+
+This logic ensures that the log is maintained in descending timestamp order, with the greatest timestamp at the head.
+\isa{undo\_op} (lines 33--36) inverts the effect of a previous move operation by restoring the prior parent and metadata that were recorded in the \isa{LogMove}'s additional field.
+\isa{redo\_op} (lines 38--41) uses \isa{do\_op} to perform an operation again and recomputes the \isa{LogMove} record (which might have changed due to the effect of the new operation with a lower timestamp).
 

### 2019-06-12 paper/move-op.tex

@@
 \renewenvironment{isabelle}{%
-  \bigbreak\noindent\hspace{\parindent}%
-  \begin{minipage}{\textwidth-\parindent}% use minipage to prevent page breaks
+  \bigbreak\noindent%
+  \renewcommand{\isanewline}{\\}%
+  \begin{minipage}{\columnwidth}% use minipage to prevent page breaks
   \begin{isabellebody}%
@@
 \ unique{\isacharunderscore}parent\ {\isacharcolon}{\isacharcolon}\ {\isacartoucheopen}{\isacharparenleft}{\isacharprime}n\ {\isasymtimes}\ {\isacharprime}m\ {\isasymtimes}\ {\isacharprime}n{\isacharparenright}\ set\ {\isasymRightarrow}\ bool{\isacartoucheclose}\ \isakeyword{where}\isanewline
-\ \ {\isacartoucheopen}unique{\isacharunderscore}parent\ tree\ {\isasymequiv}\ {\isacharparenleft}{\isasymforall}p{\isadigit{1}}\ p{\isadigit{2}}\ m{\isadigit{1}}\ m{\isadigit{2}}\ c{\isachardot}\ {\isacharparenleft}p{\isadigit{1}}{\isacharcomma}\ m{\isadigit{1}}{\isacharcomma}\ c{\isacharparenright}\ {\isasymin}\ tree\ {\isasymand}\ {\isacharparenleft}p{\isadigit{2}}{\isacharcomma}\ m{\isadigit{2}}{\isacharcomma}\ c{\isacharparenright}\ {\isasymin}\ tree\ {\isasymlongrightarrow}\ p{\isadigit{1}}\ {\isacharequal}\ p{\isadigit{2}}\ {\isasymand}\ m{\isadigit{1}}\ {\isacharequal}\ m{\isadigit{2}}{\isacharparenright}{\isacartoucheclose}%
+\ \ {\isacartoucheopen}unique{\isacharunderscore}parent\ tree\ {\isasymequiv}\ {\isacharparenleft}{\isasymforall}p{\isadigit{1}}\ p{\isadigit{2}}\ m{\isadigit{1}}\ m{\isadigit{2}}\ c{\isachardot}\ {\isacharparenleft}p{\isadigit{1}}{\isacharcomma}\ m{\isadigit{1}}{\isacharcomma}\ c{\isacharparenright}\ {\isasymin}\ tree\ {\isasymand}\ {\isacharparenleft}p{\isadigit{2}}{\isacharcomma}\ m{\isadigit{2}}{\isacharcomma}\ c{\isacharparenright}\ {\isasymin}\ tree\ {\isasymlongrightarrow}\ p{\isadigit{1}}\ {\isacharequal}\ p{\isadigit{2}}\ {\isasymand}\ m{\isadigit{1}}\ {\isacharequal}\ m{\isadigit{2}}{\isacharparenright}{\isacartoucheclose}\isanewline%
+\isanewline
+\isacommand{definition}\isamarkupfalse%
+\ acyclic\ {\isacharcolon}{\isacharcolon}\ {\isacartoucheopen}{\isacharparenleft}{\isacharprime}n\ {\isasymtimes}\ {\isacharprime}m\ {\isasymtimes}\ {\isacharprime}n{\isacharparenright}\ set\ {\isasymRightarrow}\ bool{\isacartoucheclose}\ \isakeyword{where}\isanewline
+\ \ {\isacartoucheopen}acyclic\ tree\ {\isasymequiv}\ {\isacharparenleft}{\isasymnexists}n{\isachardot}\ ancestor\ tree\ n\ n{\isacharparenright}{\isacartoucheclose}
 \end{isabellebody}
@@
 
+We now discuss the correctness properties of the algorithm from Section~\ref{sec:algorithm}.
+All theorems stated here have been formally proved and mechanically checked using Isabelle.
+For space reasons we give only a brief sketch of the proofs in this paper, but the full details can be found in the Isabelle files of our open source proof development.\footnote{\url{https://github.com/trvedata/move-op}}
+
+\subsection{Tree invariants}\label{sec:tree-invariants}
+
+A tree is an acyclic graph in which every node has exactly one parent, except for the root, which has no parent.
+To prove that our algorithm always maintains a tree structure, no matter which operations are applied, we need to prove that these invariants hold.
+
+In fact, we slightly generalise this property: we allow more than one root to exist, so the graph is actually a forest.
+This allows an application to move nodes between multiple trees, if desired.
+For example, the ``trash'' node used for deletion (Section~\ref{sec:ops-trees}) can be separate from the main tree.
+If the
+
+\begin{isabelle}
+\isacommand{theorem}\isamarkupfalse%
+\ apply{\isacharunderscore}ops{\isacharunderscore}unique{\isacharunderscore}parent{\isacharcolon}\isanewline
+\ \ \isakeyword{assumes}\ {\isacartoucheopen}apply{\isacharunderscore}ops\ ops\ {\isacharequal}\ {\isacharparenleft}log{\isacharcomma}\ tree{\isacharparenright}{\isacartoucheclose}\isanewline
+\ \ \isakeyword{shows}\ {\isacartoucheopen}unique{\isacharunderscore}parent\ tree{\isacartoucheclose}
+\end{isabelle}
+
+\begin{isabelle}
+\isacommand{theorem}\isamarkupfalse%
+\ apply{\isacharunderscore}ops{\isacharunderscore}acyclic{\isacharcolon}\isanewline
+\ \ \isakeyword{assumes}\ {\isacartoucheopen}apply{\isacharunderscore}ops\ ops\ {\isacharequal}\ {\isacharparenleft}log{\isacharcomma}\ tree{\isacharparenright}{\isacartoucheclose}\isanewline
+\ \ \isakeyword{shows}\ {\isacartoucheopen}acyclic\ tree{\isacartoucheclose}
+\end{isabelle}
+
 \section{Evaluation}\label{sec:evaluation}

### 2019-06-12 generated.scala
Generated code

@@ -1,0 +1,1009 @@
+object Uint32 {
+
+def less(x: Int, y: Int) : Boolean =
+  if (x < 0) y < 0 && x < y
+  else y < 0 || x < y
+
+def less_eq(x: Int, y: Int) : Boolean =
+  if (x < 0) y < 0 && x <= y
+  else y < 0 || x <= y
+
+def set_bit(x: Int, n: BigInt, b: Boolean) : Int =
+  if (b)
+    x | (1 << n.intValue)
+  else
+    x & (1 << n.intValue).unary_~
+
+def shiftl(x: Int, n: BigInt) : Int = x << n.intValue
+
+def shiftr(x: Int, n: BigInt) : Int = x >>> n.intValue
+
+def shiftr_signed(x: Int, n: BigInt) : Int = x >> n.intValue
+
+def test_bit(x: Int, n: BigInt) : Boolean =
+  (x & (1 << n.intValue)) != 0
+
+} /* object Uint32 */
+
+
+object DiffArray {
+
+  import scala.collection.mutable.ArraySeq
+
+  protected abstract sealed class DiffArray_D[A]
+  final case class Current[A] (a:ArraySeq[A]) extends DiffArray_D[A]
+  final case class Upd[A] (i:Int, v:A, n:DiffArray_D[A]) extends DiffArray_D[A]
+
+  object DiffArray_Realizer {
+    def realize[A](a:DiffArray_D[A]) : ArraySeq[A] = a match {
+      case Current(a) => ArraySeq.empty ++ a
+      case Upd(j,v,n) => {val a = realize(n); a.update(j, v); a}
+    }
+  }
+
+  class T[A] (var d:DiffArray_D[A]) {
+
+    def realize () = { val a=DiffArray_Realizer.realize(d); d = Current(a); a }
+    override def toString() = realize().toSeq.toString
+
+    override def equals(obj:Any) =
+      if (obj.isInstanceOf[T[A]]) obj.asInstanceOf[T[A]].realize().equals(realize())
+      else false
+
+  }
+
+
+  def array_of_list[A](l : List[A]) : T[A] = new T(Current(ArraySeq.empty ++ l))
+    def new_array[A](v:A, sz : BigInt) = new T[A](Current[A](ArraySeq.fill[A](sz.intValue)(v)))
+
+    private def length[A](a:DiffArray_D[A]) : BigInt = a match {
[…]

### 2019-06-12 evaluation/src/main/scala/Main.scala
Set up Scala project with generated code

@@ -1,0 +1,6 @@
+object Main extends App {
+  val initState = (Nil, generated.hm_empty[String, (String, String)].apply(()))
+  val operation = generated.Move((generated.zero_int, "a"), "parent", "meta", "child")
+  val state = generated.example_apply_op(operation)(initState)
+  println(state)
+}

### 2019-06-12 evaluation/src/main/scala/Move_Code.scala

@@ -1,0 +1,1123 @@
+object Uint32 {
+
+def less(x: Int, y: Int) : Boolean =
+  if (x < 0) y < 0 && x < y
+  else y < 0 || x < y
+
+def less_eq(x: Int, y: Int) : Boolean =
+  if (x < 0) y < 0 && x <= y
+  else y < 0 || x <= y
+
+def set_bit(x: Int, n: BigInt, b: Boolean) : Int =
+  if (b)
+    x | (1 << n.intValue)
+  else
+    x & (1 << n.intValue).unary_~
+
+def shiftl(x: Int, n: BigInt) : Int = x << n.intValue
+
+def shiftr(x: Int, n: BigInt) : Int = x >>> n.intValue
+
+def shiftr_signed(x: Int, n: BigInt) : Int = x >> n.intValue
+
+def test_bit(x: Int, n: BigInt) : Boolean =
+  (x & (1 << n.intValue)) != 0
+
+} /* object Uint32 */
+
+
+object DiffArray {
+
+  import scala.collection.mutable.ArraySeq
+
+  protected abstract sealed class DiffArray_D[A]
+  final case class Current[A] (a:ArraySeq[A]) extends DiffArray_D[A]
+  final case class Upd[A] (i:Int, v:A, n:DiffArray_D[A]) extends DiffArray_D[A]
+
+  object DiffArray_Realizer {
+    def realize[A](a:DiffArray_D[A]) : ArraySeq[A] = a match {
+      case Current(a) => ArraySeq.empty ++ a
+      case Upd(j,v,n) => {val a = realize(n); a.update(j, v); a}
+    }
+  }
+
+  class T[A] (var d:DiffArray_D[A]) {
+
+    def realize () = { val a=DiffArray_Realizer.realize(d); d = Current(a); a }
+    override def toString() = realize().toSeq.toString
+
+    override def equals(obj:Any) =
+      if (obj.isInstanceOf[T[A]]) obj.asInstanceOf[T[A]].realize().equals(realize())
+      else false
+
+  }
+
+
+  def array_of_list[A](l : List[A]) : T[A] = new T(Current(ArraySeq.empty ++ l))
+    def new_array[A](v:A, sz : BigInt) = new T[A](Current[A](ArraySeq.fill[A](sz.intValue)(v)))
+
+    private def length[A](a:DiffArray_D[A]) : BigInt = a match {
+    case Current(a) => a.length
+    case Upd(_,_,n) => length(n)
+  }
+
[…]

### 2019-06-12 evaluation/src/main/scala/Main.scala
Fix generating integer values

@@ -1,6 +1,8 @@
 object Main extends App {
+  def int(value: Int): generated.int = generated.int_of_integer(BigInt(value))
+
   val initState = (Nil, generated.hm_empty[String, (String, String)].apply(()))
-  val operation = generated.Move((generated.zero_int, "a"), "parent", "meta", "child")
+  val operation = generated.Move((int(0), "a"), "parent", "meta", "child")
   val state = generated.example_apply_op(operation)(initState)
   println(state)
 }

### 2019-06-12 evaluation/src/main/scala/Main.scala
Setup for profiling Scala code

@@ -1,8 +1,19 @@
+import java.util.Random
+
 object Main extends App {
   def int(value: Int): generated.int = generated.int_of_integer(BigInt(value))
 
-  val initState = (Nil, generated.hm_empty[String, (String, String)].apply(()))
-  val operation = generated.Move((int(0), "a"), "parent", "meta", "child")
-  val state = generated.example_apply_op(operation)(initState)
-  println(state)
+  var state: (List[generated.log_op[(generated.int, String),String,String]], generated.hashmap[String,(String, String)])
+    = (Nil, generated.hm_empty[String, (String, String)].apply(()))
+  val random = new Random()
+  val startTime = System.nanoTime()
+
+  for (i <- 1 to 1000) {
+    val parent = "%d".format(random.nextInt(1000))
+    val child  = "%d".format(random.nextInt(1000))
+    val operation = generated.Move((int(i), "a"), parent, "", child)
+    state = generated.example_apply_op(operation)(state)
+  }
+
+  println("elapsed time: %d ms" format ((System.nanoTime() - startTime) / 1000000))
 }

### 2019-06-13 evaluation/src/main/scala/TestClient.scala
Set up Akka with gRPC

@@ -1,0 +1,40 @@
+import examplerpc.{ MoveService, MoveServiceClient }
+import akka.actor.ActorSystem
+import akka.grpc.GrpcClientSettings
+import akka.stream.{ ActorMaterializer, Materializer }
+import scala.concurrent.{ ExecutionContext, Future }
+import scala.concurrent.duration._
+import scala.util.{ Failure, Success }
+
+object TestClient {
+
+  def main(args: Array[String]): Unit = {
+    implicit val sys: ActorSystem = ActorSystem("TestClient")
+    implicit val mat: Materializer = ActorMaterializer()
+    implicit val ec: ExecutionContext = sys.dispatcher
+
+    val config = GrpcClientSettings.connectToServiceAt("localhost", 8080)
+      .withDeadline(1.second)
+      .withTls(false)
+    val client: MoveService = MoveServiceClient(config)
+
+    runSingleRequestReplyExample()
+
+    sys.scheduler.schedule(1.second, 1.second) {
+      runSingleRequestReplyExample()
+    }
+
+    def runSingleRequestReplyExample(): Unit = {
+      sys.log.info("Performing request")
+      val timestamp = examplerpc.LamportTS(0, "a")
+      val operation = examplerpc.Move(Some(timestamp), "parent", "meta", "child")
+      val reply = client.sendMove(operation)
+      reply.onComplete {
+        case Success(msg) =>
+          println(s"got single reply: $msg")
+        case Failure(e) =>
+          println(s"Error response: $e")
+      }
+    }
+  }
+}

### 2019-06-13 evaluation/src/main/scala/TestServer.scala

@@ -1,0 +1,39 @@
+import akka.actor.ActorSystem
+import akka.http.scaladsl.UseHttp2.Always
+import akka.http.scaladsl.model.{ HttpRequest, HttpResponse }
+import akka.http.scaladsl.{ Http, HttpConnectionContext }
+import akka.stream.{ ActorMaterializer, Materializer }
+import com.typesafe.config.ConfigFactory
+import examplerpc.MoveServiceHandler
+import scala.concurrent.{ ExecutionContext, Future }
+
+object TestServer {
+
+  def main(args: Array[String]): Unit = {
+    val conf = ConfigFactory.parseString("akka.http.server.preview.enable-http2 = on")
+      .withFallback(ConfigFactory.defaultApplication())
+    val system = ActorSystem("TestServer", conf)
+    new TestServer(system).run()
+    // ActorSystem threads will keep the app alive until `system.terminate()` is called
+  }
+}
+
+class TestServer(system: ActorSystem) {
+
+  def run(): Future[Http.ServerBinding] = {
+    implicit val sys: ActorSystem = system
+    implicit val mat: Materializer = ActorMaterializer()
+    implicit val ec: ExecutionContext = sys.dispatcher
+
+    val service: HttpRequest => Future[HttpResponse] = MoveServiceHandler(new TestService(system))
+
+    val binding = Http().bindAndHandleAsync(service, interface = "0.0.0.0", port = 8080,
+      connectionContext = HttpConnectionContext(http2 = Always))
+
+    binding.foreach { binding =>
+      println(s"gRPC server bound to: ${binding.localAddress}")
+    }
+
+    binding
+  }
+}

### 2019-06-13 evaluation/src/main/scala/TestService.scala

@@ -1,0 +1,19 @@
+import akka.actor.ActorSystem
+import akka.pattern.ask
+import akka.util.Timeout
+import scala.concurrent.Future
+import scala.concurrent.duration._
+
+class TestService(system: ActorSystem) extends examplerpc.MoveService {
+
+  val treeActor = system.actorOf(TreeActor.props("a"), "treeActor")
+
+  override def sendMove(move: examplerpc.Move): Future[examplerpc.LamportTS] = {
+    println(s"received: ${move}")
+
+    implicit val timeout: Timeout = 3.seconds
+    import system.dispatcher
+
+    (treeActor ? move).mapTo[examplerpc.LamportTS]
+  }
+}

### 2019-06-13 evaluation/src/main/scala/TreeActor.scala

@@ -1,0 +1,29 @@
+import akka.actor.{ Actor, Props }
+import akka.event.Logging
+
+object TreeActor {
+  case class RequestMove(parent: String, metadata: String, child: String)
+
+  def props(actorId: String) = Props(new TreeActor(actorId))
+}
+
+class TreeActor(actorId: String) extends Actor {
+  val log = Logging(context.system, this)
+
+  def int(value: Long): generated.int = generated.int_of_integer(BigInt(value))
+
+  var state: (List[generated.log_op[(generated.int, String),String,String]], generated.hashmap[String,(String, String)])
+    = (Nil, generated.hm_empty[String, (String, String)].apply(()))
+
+  def receive = {
+    case examplerpc.Move(timestamp, parent, metadata, child) => {
+      val ts = (int(timestamp.get.counter), timestamp.get.replica)
+      val operation = generated.Move(ts, parent, metadata, child)
+      state = generated.example_apply_op(operation)(state)
+      log.info("Applied %s".format(operation))
+      sender() ! timestamp.get
+    }
+
+    case TreeActor.RequestMove(parent, metadata, child) => ()
+  }
+}

### 2019-06-14 evaluation/src/main/scala/Move_Code.scala
Export apply_ops as well

@@ -631,6 +631,16 @@ final case class RBT_HM[B, A](a: rbt[Int, assoc_list[B, A]]) extends
 
 def eq[A : equal](a: A, b: A): Boolean = equal[A](a, b)
 
+def integer_of_nat(x0: nat): BigInt = x0 match {
+  case Nat(x) => x
+}
+
+def plus_nat(m: nat, n: nat): nat = Nat(integer_of_nat(m) + integer_of_nat(n))
+
+def one_nat: nat = Nat(BigInt(1))
+
+def Suc(n: nat): nat = plus_nat(n, one_nat)
+
 def empty[A : linorder, B]: rbt[A, B] = RBT[A, B](Empty[A, B]())
 
 def map_of[A : equal, B](x0: List[(A, B)], k: A): Option[B] = (x0, k) match {
@@ -785,6 +795,153 @@ def balance[A, B](x0: rbta[A, B], s: A, t: B, x3: rbta[A, B]): rbta[A, B] =
                      s, t, Branch[A, B](B(), va, vb, vc, vd))
 }
 
+def paint[A, B](c: color, x1: rbta[A, B]): rbta[A, B] = (c, x1) match {
+  case (c, Empty()) => Empty[A, B]()
+  case (c, Branch(uu, l, k, v, r)) => Branch[A, B](c, l, k, v, r)
+}
+
+def balance_right[A, B](a: rbta[A, B], k: A, x: B, xa3: rbta[A, B]): rbta[A, B]
+  =
+  (a, k, x, xa3) match {
+  case (a, k, x, Branch(R(), b, s, y, c)) =>
+    Branch[A, B](R(), a, k, x, Branch[A, B](B(), b, s, y, c))
+  case (Branch(B(), a, k, x, b), s, y, Empty()) =>
+    balance[A, B](Branch[A, B](R(), a, k, x, b), s, y, Empty[A, B]())
+  case (Branch(B(), a, k, x, b), s, y, Branch(B(), va, vb, vc, vd)) =>
+    balance[A, B](Branch[A, B](R(), a, k, x, b), s, y,
+                   Branch[A, B](B(), va, vb, vc, vd))
+  case (Branch(R(), a, k, x, Branch(B(), b, s, y, c)), t, z, Empty()) =>
+    Branch[A, B](R(), balance[A, B](paint[A, B](R(), a), k, x, b), s, y,
[…]

### 2019-06-14 paper/move-op.tex

@@
 As the tree structure may be concurrently modified on different replicas, the state of these replicas may temporarily diverge.
-In this paper we show how the replicas can nevertheless achieve \emph{strong eventual consistency}~\cite{Shapiro:2011un}: as the replicas communicate, we guarantee that they will converge towards a consistent state.
+In this paper we show how the replicas can nevertheless achieve \emph{strong eventual consistency}~\cite{Shapiro:2011un,Gomes:2017gy}: as the replicas communicate, we guarantee that they will converge towards a consistent state.
 Our algorithm for achieving convergence is an example of a \emph{Conflict-free Replicated Data Type} or CRDT~\cite{Shapiro:2011wy,Shapiro:2011un}.
@@
 
+TODO
+
 \section{Related work}\label{sec:relwork}
 
-% Related work:
-% OT implementation of move operation: https://hal.inria.fr/inria-00107740/
-% Ficus (does not discuss move): https://pdos.csail.mit.edu/6.824/papers/ficus.pdf
-% Roam: https://ieeexplore.ieee.org/abstract/document/795151
+Many replicated data systems use optimistic replication~\cite{Saito:2005jw}, which allows the state of replicas to temporarily diverge, in order to achieve better performance and better availability in the presence of faults than strongly consistent systems~\cite{Bailis:2014th,Gilbert:2002il}.
+As a consequence, these systems require a mechanism for merging or reconciling conflicting updates that were made concurrently on different replicas.
+For example, version control systems such as Git~\cite{Chacon:2014kr} leave any conflicts for the user to resolve manually.
+Some systems, such as Dynamo~\cite{DeCandia:2007ui} and Bayou~\cite{Terry:1995dn}, rely on the application programmer to provide explicit conflict resolution logic; however, such logic is difficult to get right \cite{Bailis:2013jc,Burckhardt:2014hy,Gomes:2017gy}.
+Hence, we focus on systems that automatically ensure that all replicas converge towards a consistent state, without requiring custom application logic---a consistency model known as \emph{strong eventual consistency}~\cite{Shapiro:2011un,Gomes:2017gy}.
+
+\subsection{Conflict-free Replicated Data Types}
+
+Our algorithm is an example of an operation-based Conflict-free Replicated Data Type or CRDT~\cite{Shapiro:2011wy,Shapiro:2011un,Burckhardt:2014ft}.
+All CRDTs share the property that concurrent changes on different replicas can be merged in any order; any two replicas that have seen the same set of updates are guaranteed to be in the same state, regardless of the order in which they processed these updates.
+CRDTs have been defined for various abstract datatypes, such as registers and counters \cite{Shapiro:2011wy,Shapiro:2011un}, maps \cite{Baquero:2016iv,Kleppmann:2016ve}, sets \cite{Bieniusa:2012wu,Bieniusa:2012gt}, lists (sequences) \cite{Roh:2011dw,Nedelec:2013ky}, and text \cite{Preguica:2009fz,Weiss:2010hx}.
+For tree data structures, several approaches have been proposed:
+\begin{itemize}
+    \item Martin et al.~\cite{Martin:2010ih} define a CRDT for XML data, and Kleppmann and Beresford~\cite{Kleppmann:2016ve} define a CRDT for JSON.
+        However, these algorithms only deal with insertion and deletion of tree nodes, and do not support a move operation.
+    \item As discussed in Section~\ref{sec:impossibility}, Najafzadeh et al.~\cite{Najafzadeh:2017vk,Najafzadeh:2018bw} propose two alternative implementations for a replicated filesystem: a CRDT in which conflicting moves are handled by duplicating tree nodes (as in Figures~\ref{fig:move-same-item}b and~\ref{fig:move-cycle}b), and a centralised implementation in which move operations must obtain a lock before executing (not a CRDT since it relies on synchronous coordination).
+    \item Ahmed-Nacer et al.~\cite{AhmedNacer:2012us} survey approaches to handling conflicts on trees without providing concrete algorithms.
+    \item Tao et al.~\cite{Tao:2015gd} propose handling conflicting move operations by allowing the same object to appear in more than one location; thus, their datatype is strictly a DAG, not a tree.
+        Some conflicts are handled by duplicating tree nodes.
+    \item In our previous unpublished work \cite{ExtendedVersion} we develop the foundations for the algorithm described in this paper.
+\end{itemize}
+
+Besides CRDTs, another family of algorithms for concurrent modification of data structures is called \emph{Operational Transformation}~\cite{Sun:1998vf}.
+Several authors have defined collaboratively editable tree structures using the Operational Transformation approach \cite{Jungnickel:2016cb,Ignat:2003jy,Davis:2002iv}, but they only consider insertion and deletion of tree nodes, and do not support a move operation.
+
+Molli et al.~\cite{Molli:2003cd} define an Operational Transformation tree structure with a move operation.
+However, it requires that all communication between replicas is performed via total order broadcast, which in turn requires the use of a sequencer server or a consensus algorithm~\cite{Chandra:1996cp}.
+Our algorithm has better availability characteristics in the presence of network partitions because it allows messages to be delivered in any order.
+
+\subsection{Distributed filesystems}
+
+TODO: Coda~\cite{Kumar:1995wf}, Ficus~\cite{Reiher:1994wh}, Rumor~\cite{Guy:1999gy}, Roam~\cite{Ratner:1999fh}
+
+\section{Conclusions}
+
+TODO
 

### 2019-06-14 paper/move-op.tex
paper: some small changes, typos fixed

@@
         As discussed in Section~\ref{sec:impossibility}, this has previously been thought to be impossible to achieve \cite{Najafzadeh:2017vk,Najafzadeh:2018bw}.
-    \item We formalise our algorithm using Isabelle/HOL~\cite{DBLP:conf/tphol/WenzelPN08}, an interactive proof assistant, and produce a computer-checked proof of correctness of our algorithm.
-        In particular, we prove that arbitrary concurrent modifications to the tree can be merged such that all replicas converge towards a consistent state (strong eventual consistency).
-    \item To demonstrate the practical viability of the algorithm we extract a formally verified, executable implementation from the Isabelle/HOL definitions, and perform tests on this implementation.
+    \item We formalise our abstract algorithm using Isabelle/HOL~\cite{DBLP:conf/tphol/WenzelPN08}, an interactive proof assistant based on higher-order logic, and obtain a computer-checked proof of correctness of our algorithm.
+        In particular, we prove that arbitrary concurrent modifications to the tree can be merged such that all replicas converge towards a consistent state (strong eventual consistency) whilst maintaining a tree-like structure.
+    \item To demonstrate the practical viability of our approach, we refine our abstract algorithm to an efficient implementation within Isabelle/HOL and prove the equivalence of the two.
+We thereafter extract a formally verified, executable Scala implementation from the Isabelle/HOL definitions, and perform tests on this implementation.
     \item We perform experiments with replicated filesystem products such as Dropbox and Google Drive, and show that they exhibit problems that would be prevented by our algorithm.
@@
 However, the result is a DAG, not a tree.
-POSIX filesystems to not allow this outcome, since they allow hardlinks only to files, but not to directories.
+POSIX-compliant filesystems do not allow this outcome, since they allow hardlinks only to files, but not to directories.
 
@@
 In Google Drive, one replica was able to successfully sync with the server, while the other replica displayed the ``unknown error'' message shown in Figure~\ref{fig:gdrive-error}.
-The replica in error state refused to sync the conflicting directory, and its filesystem state remained inconsistent with the other replica.
+The replica in an error state refused to sync the conflicting directory, and its filesystem state remained inconsistent with the other replica.
 This error state persisted until the directories on the erroring replica were manually moved to match the state of the other replica.

### 2019-06-14 paper/move-op.tex
Add abstract

@@
 \begin{abstract}
-    TODO abstract.
+    Replicated tree data structures are found in distributed filesystems, such as Google Drive and Dropbox, and applications with a JSON or XML data model.
+    These systems need to support a \emph{move} operation that allows a subtree to be moved to a new location within the tree.
+    Such a move operation is easy to implement in a centralised system, e.g.\ with a designated primary replica, but it is difficult in settings where different replicas can concurrently perform arbitrary move operations.
+    For example, we find that Google Drive and Dropbox fail to correctly synchronise files when users on different devices concurrently perform certain move operations.
+    In this paper we present an algorithm that handles arbitrary concurrent modifications on trees, while ensuring that the tree structure remains valid (in particular, no cycles are introduced), and guaranteeing that all replicas eventually converge towards the same consistent state.
+    We formally prove the correctness of our algorithm using the Isabelle/HOL interactive proof assistant, and we evaluate the performance of our formally verified implementation in a geo-replicated setting to demonstrate its viability in practice.
 \end{abstract}

### 2019-06-15 paper/move-op.tex
paper: add sections 3.5 and 3.6

@@
 \acmISBN{}
+
+\hyphenation{da-ta-cen-ter da-ta-cen-ters time-stamp time-stamps time-stamped}
 
@@
         Multiple objects and groups may in turn be combined into higher-level groups, forming a tree.
-        %Moreover, the order of nodes determines visibility (i.e.\ objects that are ``further back'' may be obscured by objects that are ``closer to the front'').
     \item Personal \textbf{note-taking} and \textbf{task-management tools} such as Org-mode for Emacs~\cite{OrgMode} or OmniOutliner~\cite{OmniOutliner} explicitly present the user with a tree structure that they can inspect and manipulate.
@@
     \item We formalise our abstract algorithm using Isabelle/HOL~\cite{DBLP:conf/tphol/WenzelPN08}, an interactive proof assistant based on higher-order logic, and obtain a computer-checked proof of correctness of our algorithm.
-        In particular, we prove that arbitrary concurrent modifications to the tree can be merged such that all replicas converge towards a consistent state (strong eventual consistency) whilst maintaining a tree-like structure.
+        In particular, we prove that arbitrary concurrent modifications to the tree can be merged such that all replicas converge towards a consistent state (strong eventual consistency) while keeping the data in a tree structure.
     \item To demonstrate the practical viability of our approach, we refine our abstract algorithm to an efficient implementation within Isabelle/HOL and prove the equivalence of the two.
@@
 
+\subsection{Handling conflicts}\label{sec:conflicts}
+
+The consequence of performing the undo-do-redo cycle is that, effectively, the state of the tree is as if all operations had been applied using \isa{do\_op} in increasing timestamp order, regardless of the order in which they were actually applied.
+This provides a clear approach to the handling of conflicts:
+\begin{itemize}
+    \item If two operations concurrently move the same node, then the ``winning'' operation is the one with the higher timestamp.
+        This is the case because, if the operations are applied in increasing timestamp order, the node being moved is first moved to the parent indicated by the operation with the lower timestamp, and then moved again to the parent indicated by the operation with the higher timestamp.
+        Since move operations do not specify the old location of a node, but only the new location, this sequential execution of concurrent operations is well-defined.
+    \item If two operations would introduce a cycle when combined, as in Section~\ref{sec:move-cycle}, then the operation with the greater timestamp is ignored.
+        This is the case because \isa{do\_op} checks for cycles based on the tree created by all operations with a lower timestamp.
+        The lower of two conflicting operations will take effect, since that operation by itself is safe.
+        When the higher-timestamped operation is applied, \isa{do\_op} detects that it would introduce a cycle, and therefore ignores the operation.
+\end{itemize}
+
+Note that the safety of an operation (whether or not it would introduce a cycle) may change as subsequent operations with lower timestamps are applied.
+For example, an operation may initially be regarded as safe, and then be reclassified as unsafe after applying a conflicting operation with a lower timestamp.
+The opposite is also possible: an operation previously regarded as unsafe may become safe through the application of a lower-timestamped operation that removes the risk of introducing a cycle.
+For this reason, the operation log must include all operations, even those that have no effect on the tree.
+
+One final type of conflict that we have not discussed so far is multiple child nodes with the same parent and the same metadata.
+For example, in a filesystem, two users could concurrently create files with the same name in the same directory.
+Our algorithm does not take any measures to prevent such conflict, but simply retains both child nodes.
+In practice, the filename collision would be resolved by making the filenames distict, e.g.\ by appending a replica identifier to the filenames.
+
+\subsection{Algorithm extensions}\label{sec:extensions}
+
+The algorithm as specified in Figure~\ref{fig:code} retains operations in the log indefinitely, so the memory/storage use grows without bound.
+However, in practice it is easy to truncate the log, because \isa{apply\_op} only examines the prefix of operations whose timestamp is greater than that of the operation being applied.
+Thus, once it is known that all future operations will have a timestamp greater than $t$, then all operations with timestamp $t$ or less can be discarded from the log.
+In this case, we say that $t$ is \emph{causally stable}~\cite{Baquero:2014ed}.
+
+We can determine causal stability in a system where the set of replicas is known, where each replica generates operations with monotonically increasing timestamps, and where the communication link between any pair of replicas is FIFO (messages are received in the order in which they are sent, as implemented e.g.
+by TCP).
+In this case, each replica can keep track of the most recent timestamp it has seen from all replicas (including its own), and the minimum of these timestamps is the causally stable threshold.
+
+Another useful extension of the tree algorithm is to allow children within the same branch node to have an ordering.
+For example, in a graphics application, the order of nodes determines visibility (objects that are ``further back'' may be obscured by objects that are ``closer to the front'').
+This can be implemented by maintaining a separate list CRDT for each branch node, e.g.\ using RGA~\cite{Roh:2011dw}, Treedoc~\cite{Preguica:2009fz}, Logoot~\cite{Weiss:2010hx}, or LSEQ~\cite{Nedelec:2013ky}.
+These algorithms assign a unique identifier to each element of the list, and this unique identifier can be included in the metadata field of move operations in order to determine the order of children of the same parent node.
+
+This approach to determining ordering also easily supports reordering of child nodes within a parent, e.g.\ to implement the ``bring to front'' and ``send to back'' feature of applications like PowerPoint.
+To move a node to a different position in a list, we use the list CRDT to generate a new identifier at the desired position in the sequence.
+Then we perform a move operation in which the parent node is unchanged, but the metadata is updated to include this new identifier.
+
 % - No Synchronisation or consensus required. Available in the sense of CAP
@@
         Some conflicts are handled by duplicating tree nodes.
+        Tao et al.\ also perform experiments with Dropbox, Google Drive, and OneDrive, similar to our experiments discussed in Section~\ref{sec:move-is-hard}.
     \item In our previous unpublished work \cite{ExtendedVersion} we develop the foundations for the algorithm described in this paper.

### 2019-06-17 paper/move-op.tex
paper: starting work on Section 5

@@
 
-TODO
+\subsection{Extracting efficient algorithms from HOL}
+
+Isabelle/HOL's can extract executable Haskell, OCaml, Scala, and Standard ML code from HOL definitions using a code generation mechanism.
+However,  care must be taken when using this mechanism, as:
+\begin{enumerate}
+\item
+The mechanism is \emph{partial} in the sense that not all HOL definitions can be realised in an executable form.
+Definitions making essential use of choice principles (such as our \isa{get\_parent} function) are obvious examples of definitions that will cause problems.
+However, \emph{inductively defined} relations characterised by a series of introduction rules, such as our \isa{ancestor} relation are also problematic for the code generator as making them executable necessarily requires Prolog-like proof search and mode inference.
+\item
+Unsophisticated use of the code generator often leads to inefficient target language code.
+This is especially true when using HOL's \isa{set} type, a type of potentially infinite mathematical sets.
+Whilst convenient for theorem proving purposes, this type often leads to gross inefficiency in the extracted code.
+\end{enumerate}
+
+Because of these problems, we produce
 

### 2019-06-17 paper/move-op.tex
paper: more on Section 5

@@
 
-Isabelle/HOL's can extract executable Haskell, OCaml, Scala, and Standard ML code from HOL definitions using a code generation mechanism.
+Isabelle/HOL can generate executable Haskell, OCaml, Scala, and Standard ML code from HOL definitions using a code generation mechanism~\cite{XXX}.
 However,  care must be taken when using this mechanism, as:
@@
 
-Because of these problems, we produce
+As a result, prior to evaluating our algorithms in a real-world setting, we first produce a series of refined definitions that are optimised for efficient execution rather than ease of theorem proving.
+Within these definitions, we eschew the HOL \isa{set} type, and instead use a more efficient hash-map implementation provided by the Isabelle Collections Framework~\cite{XXX}.
+Using an explicit \emph{simulation relation}, we show the correctness of these efficient algorithms taking our abstract algorithms as ground truth of the correct behaviour.
+Moreover, we show that our correctness theorems, such as \isa{apply\_ops\_acyclic}, easily transfer to the more efficient implementations.
 

### 2019-06-17 paper/move-op.tex
paper: Section 5, again

@@
 Our algorithm does not take any measures to prevent such conflict, but simply retains both child nodes.
-In practice, the filename collision would be resolved by making the filenames distict, e.g.\ by appending a replica identifier to the filenames.
+In practice, the filename collision would be resolved by making the filenames distinct, e.g.\ by appending a replica identifier to the filenames.
 
@@
 
-Isabelle/HOL can generate executable Haskell, OCaml, Scala, and Standard ML code from HOL definitions using a code generation mechanism~\cite{XXX}.
+Isabelle/HOL can generate executable Haskell, OCaml, Scala, and Standard ML code from HOL type and constant definitions using a sophisticated code generation mechanism~\cite{XXX}.
 However,  care must be taken when using this mechanism, as:
@@
 Within these definitions, we eschew the HOL \isa{set} type, and instead use a more efficient hash-map implementation provided by the Isabelle Collections Framework~\cite{XXX}.
+Write $('a::\{hashable\}, 'b)\ \isa{hm}$ for the type of hash-maps with key type $\isa{'a}$ and binding type $\isa{'b}$.
+Note that keys have an explicit $\isa{hashable}$ constraint, meaning that this hash-map type may only be used with types of keys with an associated hashing function.
+With this type, we represent trees by the type $(\isa{'n}::\{hashable\}, \isa{'m} \times \isa{'n})\ \isa{hm}$.
+Intuitively, this type can be thought of as a mapping from \emph{child} nodes to \emph{parent} nodes, represented by the type-variable $\isa{'n}$, and any associated metadata (represented by the type variable $\isa{'m}$.
+
+We introduce an explicit \emph{simulation relation} between our two representations of trees.
+Specifically, we say that a hash-map $\isa{t}$ of type $(\isa{'n}, \isa{'m} \times \isa{'n})\ \isa{hm}$ simulates a set $T$ of type $(\isa{'n} \times \isa{'m} \times \isa{'n})\ \isa{set}$:
+\begin{displaymath}
+\isa{hm.lookup}\ c t = \isa{Some}\ (m, p) \leftrightarrow (p, m, c) \in T \quad\text{for all $c$, $m$, and $p$}
+\end{displaymath}
+where $hm.lookup$ is the (potentially failing) function that searches for a binding for key $\isa{c}$ in hash-map $t$.
+We write $t \preceq T$ to assert that the hash-map $t$ and set $T$ are contained within the simulation relation, defined above.
+
 Using an explicit \emph{simulation relation}, we show the correctness of these efficient algorithms taking our abstract algorithms as ground truth of the correct behaviour.

### 2019-06-17 paper/move-op.tex
paper: first draft of Section 5 almost complete

@@
 With this type, we represent trees by the type $(\isa{'n}::\{hashable\}, \isa{'m} \times \isa{'n})\ \isa{hm}$.
-Intuitively, this type can be thought of as a mapping from \emph{child} nodes to \emph{parent} nodes, represented by the type-variable $\isa{'n}$, and any associated metadata (represented by the type variable $\isa{'m}$.
+Intuitively, this type can be thought of as a mapping from \emph{child} nodes to \emph{parent} nodes, represented by the type-variable $\isa{'n}$, and any associated metadata (represented by the type variable $\isa{'m}$).
 
 We introduce an explicit \emph{simulation relation} between our two representations of trees.
-Specifically, we say that a hash-map $\isa{t}$ of type $(\isa{'n}, \isa{'m} \times \isa{'n})\ \isa{hm}$ simulates a set $T$ of type $(\isa{'n} \times \isa{'m} \times \isa{'n})\ \isa{set}$:
+Specifically, we say that a hash-map $\isa{t}$ of type $(\isa{'n}, \isa{'m} \times \isa{'n})\ \isa{hm}$ simulates a set $T$ of type $(\isa{'n} \times \isa{'m} \times \isa{'n})\ \isa{set}$ whenever
 \begin{displaymath}
-\isa{hm.lookup}\ c t = \isa{Some}\ (m, p) \leftrightarrow (p, m, c) \in T \quad\text{for all $c$, $m$, and $p$}
+\isa{hm.lookup}\ c\ t = \isa{Some}\ (m, p) \quad\text{iff}\quad (p, m, c) \in T \quad\text{forall $c$, $m$, and $p$}
 \end{displaymath}
 where $hm.lookup$ is the (potentially failing) function that searches for a binding for key $\isa{c}$ in hash-map $t$.
-We write $t \preceq T$ to assert that the hash-map $t$ and set $T$ are contained within the simulation relation, defined above.
+We write $t \preceq T$ to assert that the hash-map $t$ and set $T$ are contained within the simulation relation, as defined above.
+Note that many useful corollaries follow immediately from this definition, for example we have $t \preceq T$ implies $\isa{unique\_parent}\ \isa{T}$.
 
-Using an explicit \emph{simulation relation}, we show the correctness of these efficient algorithms taking our abstract algorithms as ground truth of the correct behaviour.
-Moreover, we show that our correctness theorems, such as \isa{apply\_ops\_acyclic}, easily transfer to the more efficient implementations.
+Our general strategy henceforth is to produce a new set of definitions---mirroring the structure of our existing definitions---wherein we replace all occurrences of the $(\isa{'n} \times \isa{'m} \times \isa{'n})\ \isa{set}$ type, and any associated set operations, with the $(\isa{'n}, \isa{'m} \times \isa{'n})\ \isa{hm}$ type and associated hash-map operations.
+We adopt the convention that any definition using hash-maps will be referred to as an ``efficient'' definition, and any definition making use of HOL's set type will be referred to as an ``abstract'' definition, for clarity.
+
+For most definitions this is an almost mechanical process, and we therefore elide further description here.
+However, \isa{ancestor} is slightly different to our other definitions, as it is an inductively defined relation rather than a recursive function, and we therefore need further work to make this efficiently executable.
+Specifically, we can characterise this relation recursively with the following theorem, which we prove inside Isabelle/HOL:
+
+% DPM: add efficient_ancestor_simp here
+
+Note that the theorem \isa{efficient\_ancestor\_simp} is tagged with a $\isa{code}$ attribute, which is a hint that this theorem should be used as a transformation during the code generation process.
+Isabelle/HOL takes this theorem as the definition, for code generation, of the \isa{efficient\_ancestor} relation and is able to directly translate this to a recursive function definition in each target backend.
+
+Next, we demonstrate the correctness of our new efficient definitions against our older, abstract definitions.
+The correctness properties that we prove are best described using an example, as they all follow a similar pattern, and we therefore focus on the lemma \isa{efficient\_do\_op\_simulates} which states the following property:
+
+% DPM: add efficient_do_op_simulates here
+
+Note that we show that the \isa{efficient\_do\_op} and the \isa{do\_op} functions maintain a ``lock-step'' property, mapping related trees $t \preceq T$ to related trees $u \preceq U$ and producing identical logs of operations, whenever they are applied to the same input operation, \isa{oper}.
+This pattern of lock-step simulation is repeated bottom-up for every pair of efficient-abstract algorithm pairs, until we reach the final correctness theorem for \isa{apply\_ops}:
+
+%DPM: add efficient_apply_ops_simulates here
+
+Note that this property correctly captures our intuitive understanding of what it means for \isa{efficient\_apply\_ops} to be correct with respect to \isa{apply\_ops}.
+Specifically, the theorem states that if both algorithms are applied to the same set of operations then they produce identical logs and also produce trees that contain the same set of parent-metadata-child records, i.e.
+the two trees are extensionally equivalent, despite having very different concrete representations.
+We can further check that this theorem correctly captures the correctness of our \isa{apply\_ops} definition by recasting the various tree invariant and convergence theorems in terms of \isa{efficient\_apply\_ops}.
+For example:
+
+% DPM: add efficient_apply_ops_commutes and efficient_apply_ops_acyclic here
+
+Both follow as corollaries of the theorems \isa{apply\_ops\_commutes} and \text{apply\_ops\_acyclic}, respectively.
+Note that the statement of \isa{efficient\_apply\_ops\_commutes} differs slightly from the statement of \isa{apply\_ops\_commutes}---the latter theorem asserts that identical sets are obtained whilst the former asserts that simply reading any key binding in $t$ gives the same result as reading in $u$.
+Whilst this statement appears to be weaker than the statement of \text{apply\_ops\_acyclic}, it is correct, as the resulting hash-maps may not have an identical representation in-memory, depending on the particular ordering of key-value insertions, despite containing the same key-value bindings.
 

### 2019-06-17 paper/move-op.tex

@@
 With this type, we represent trees by the type $(\isa{'n}::\{hashable\}, \isa{'m} \times \isa{'n})\ \isa{hm}$.
-Intuitively, this type can be thought of as a mapping from \emph{child} nodes to \emph{parent} nodes, represented by the type-variable $\isa{'n}$, and any associated metadata (represented by the type variable $\isa{'m}$).
+Intuitively, this type can be thought of as a mapping from a \emph{child} node to its \emph{parent} node and associated metadata, if any such mapping exists.
 
@@
 \end{displaymath}
-where $hm.lookup$ is the (potentially failing) function that searches for a binding for key $\isa{c}$ in hash-map $t$.
+where $hm.lookup$ is the function that searches for a binding for key $\isa{c}$ in hash-map $t$, returning $\isa{Some}\ (m, p)$ when a binding from $c$ to $(m, p)$ exists in the hash-map, or $\isa{None}$ otherwise.
 We write $t \preceq T$ to assert that the hash-map $t$ and set $T$ are contained within the simulation relation, as defined above.

### 2019-06-17 paper/move-op.tex
paper: add section 4 on correctness proof

@@
 \renewenvironment{isabelle}{%
-  \bigbreak\noindent%
+  \medbreak\noindent%
   \renewcommand{\isanewline}{\\}%
@@
   \end{minipage}%
-  \bigbreak%
+  \medbreak%
 }
@@
 
-We tested this scenario with file sync products Dropbox and Google Drive by concurrently moving the same directory to two different destination directories.\footnote{Experiment setup: we installed the official Mac OS clients for Dropbox and Google Drive on two computers, and configured them to each sync a directory on the local filesystem with the corresponding cloud service.
+We tested this scenario with file sync products Dropbox and Google Drive by concurrently moving the same directory to two different destination directories.\footnote{Experiment setup: we installed the official Mac OS clients for Dropbox and Google Drive on two computers, logged into the same Dropbox/Google accounts, and configured them to each sync a directory on the local filesystem with the corresponding cloud service.
 To test concurrent operations, we turned off the network interface on both computers, performed a move operation on the local filesystem of each computer, then reconnected both computers to the Internet and waited for them to sync.}
@@
 All theorems stated here have been formally proved and mechanically checked using Isabelle.
-For space reasons we give only a brief sketch of the proofs in this paper, but the full details can be found in the Isabelle files of our open source proof development.\footnote{\url{https://github.com/trvedata/move-op}}
+For space reasons we give only a brief sketch of the proofs in this paper, but the full details can be found in the open source Isabelle files of our proof development.\footnote{\url{https://github.com/trvedata/move-op}}
+
+To reason about the state of a replica we define the function \isa{apply\_ops} on lines 53--54 of Figure~\ref{fig:code}.
+It takes a list of operations \isa{ops} and returns the state of a replica after it has applied all the operations in \isa{ops}.
+The \isa{apply\_ops} function works by starting in the initial state \isa{([], \{\})} consisting of the empty operation log \isa{[]} and the empty tree \isa{\{\}}, and then applying the operations one by one to the state using the \isa{apply\_op} function (introduced in Section~\ref{sec:applying}).
+The \isa{foldl} function from Isabelle's standard library performs the iteration over the list of operations.
 
@@
 For example, the ``trash'' node used for deletion (Section~\ref{sec:ops-trees}) can be separate from the main tree.
-If the
 
+\subsubsection{Each node's parent is unique}
+
+The first invariant we prove is that each tree node has either no parent (if it is the root of a tree) or exactly one parent (if it is a non-root node).
+We state this theorem in Isabelle as follows, where \isa{apply\_ops\_unique\_parent} is the name we give to this theorem:
 \begin{isabelle}
@@
 \end{isabelle}
+\noindent That is, we consider any list of operations \isa{ops} and define \isa{(log, tree)} to be the replica state after \isa{ops} have been applied.
+We then prove that \isa{unique\_parent tree} holds, where the \isa{unique\_parent} predicate is defined on lines 56--57 of Figure~\ref{fig:code}: it states that whenever the tree contains a triple whose third element is the child node \isa{c}, then the first and second elements of the triple (the parent node and the metadata) are uniquely defined.
 
+Because we make no assumptions about \isa{ops}, this theorem holds for any replica state that can be reached by applying any number of operations.
+Proving \isa{apply\_ops\_unique\_parent} is easy because each of the \isa{do\_op}, \isa{undo\_op} and \isa{redo\_op} functions individually maintains the \isa{unique\_parent tree} invariant, and thus \isa{apply\_op} also maintains the invariant.
+
+\subsubsection{The graph contains no cycles}
+
+The second invariant we require is that the graph must not contain any cycles.
+We state this theorem in Isabelle as follows:
 \begin{isabelle}
@@
 \end{isabelle}
+\noindent The \isa{acyclic} predicate is defined on lines 59--60 of Figure~\ref{fig:code}, using the \isa{ancestor} relation (the transitive closure of the graph's edges): a graph contains no cycles if no node is an ancestor of itself.
+
+Proof sketch: Assume \isa{do\_op (op1, tree1) = (op2, tree2)} for some \isa{op1}, \isa{op2}, \isa{tree1} and \isa{tree2}.
+Further assume \isa{acyclic tree1}.
+We can then prove by contradiction that \isa{tree2} is also acylic.
+Assume to the contrary that \isa{tree2} contains a cycle.
+If \isa{do\_op} left the tree unchanged (line 30), we immediately have a contradiction.
+Otherwise we can assume \isa{{\isasymnot}(ancestor tree c newp {\isasymor} c = newp)} and we have:
+\medbreak\noindent
+\isa{tree2 = \{(p', m', c') {\isasymin} tree1. c' {\isasymnoteq} c\} {\isasymunion} \{(newp, m, c)\}}
+\medbreak\noindent
+That is, \isa{do\_op} adds one edge \isa{\{(newp, m, c)\}} to the graph, and perhaps removes another edge.
+Since \isa{tree1} is acyclic, the added edge must lie on the cycle in \isa{tree2}.
+Moreover, since only one edge from \isa{newp} to \isa{c} was added, \isa{tree1} must contain a path from \isa{c} to \isa{newp} (which is then closed by the new edge to form a cycle), i.e.\ \isa{c} must be an ancestor of \isa{newp} in \isa{tree1}.
+However, in that case, the predicate \isa{ancestor tree1 c newp} on line 30 would have been true, which contradicts the earlier assumption.
+Hence, the \isa{do\_op} function preserves the acyclicity invariant of the tree.
+
+To prove that \isa{apply\_op} preserves the acyclicity invariant, we need to show that for any tree produced by \isa{apply\_ops}, there exists a sequence of operations such that applying those operations using \isa{do\_op} invocations alone (not using \isa{undo\_op} or \isa{redo\_op}) results in the same tree.
+(This is the case when operations are applied in order of increasing timestamp.)
+Then we can use the result of the previous paragraph to prove that any tree produced by \isa{apply\_ops} must be acyclic.
+
+\subsection{Commutativity of apply\_op}
+
+As discussed in Section~\ref{sec:applying}, we require that when replicas apply the same set of operations, they converge towards the same state, regardless of the order in which the operations are applied.
+We formalise this in Isabelle as follows:
+\begin{isabelle}
+\isacommand{theorem}\isamarkupfalse%
+\ apply{\isacharunderscore}ops{\isacharunderscore}commutes{\isacharcolon}\isanewline
+\ \ \isakeyword{assumes}\ {\isacartoucheopen}set\ ops{\isadigit{1}}\ {\isacharequal}\ set\ ops{\isadigit{2}}{\isacartoucheclose}\isanewline
+\ \ \ \ \isakeyword{and}\ {\isacartoucheopen}distinct\ {\isacharparenleft}map\ move{\isacharunderscore}time\ ops{\isadigit{1}}{\isacharparenright}{\isacartoucheclose}\isanewline
+\ \ \ \ \isakeyword{and}\ {\isacartoucheopen}distinct\ {\isacharparenleft}map\ move{\isacharunderscore}time\ ops{\isadigit{2}}{\isacharparenright}{\isacartoucheclose}\isanewline
+\ \ \isakeyword{shows}\ {\isacartoucheopen}apply{\isacharunderscore}ops\ ops{\isadigit{1}}\ {\isacharequal}\ apply{\isacharunderscore}ops\ ops{\isadigit{2}}{\isacartoucheclose}
+\end{isabelle}
+
+The predicate \isa{distinct} takes a list as argument, and returns true if all elements of the list are distinct (i.e.\ no value occurs more than once in the list).
+The assumption \isa{distinct (map move\_time ops1)} therefore states that in the list of operations \isa{ops1}, there are no two operations with the same timestamp.
+
+The function \isa{set} takes a list and turns it into an unordered set with the same elements.
+Thus, the assumption \isa{set ops1 = set ops2} means that the lists \isa{ops1} and \isa{ops2} contain the same elements, but perhaps in a different order---in other words, \isa{ops1} is a permutation of \isa{ops2}.
+Under these assumptions, \isa{apply\_ops\_commutes} proves that applying the list of operations \isa{ops1} results in the same replica state as applying the list of operations \isa{ops2}.
+
+To prove this theorem we must first show that \isa{undo\_op} performs the inverse of \isa{do\_op}, that is:
+\begin{isabelle}
+\isacommand{lemma}\isamarkupfalse%
+\ do{\isacharunderscore}undo{\isacharunderscore}op{\isacharunderscore}inv{\isacharcolon}\isanewline
+\ \ \isakeyword{assumes}\ {\isacartoucheopen}unique{\isacharunderscore}parent\ tree{\isacartoucheclose}\isanewline
+\ \ \isakeyword{shows}\ {\isacartoucheopen}undo{\isacharunderscore}op\ {\isacharparenleft}do{\isacharunderscore}op\ {\isacharparenleft}oper{\isacharcomma}\ tree{\isacharparenright}{\isacharparenright}\ {\isacharequal}\ tree{\isacartoucheclose}
+\end{isabelle}
+
+\isa{apply\_ops\_commutes} then follows from the following lemma, which proves that two calls of \isa{apply\_op} commute, provided that the timestamps in the operations and in the log are all distinct:
+\begin{isabelle}
+\isacommand{lemma}\isamarkupfalse%
+\ apply{\isacharunderscore}op{\isacharunderscore}commute{\isadigit{2}}{\isacharcolon}\isanewline
+\ \ \isakeyword{assumes}\ {\isacartoucheopen}unique{\isacharunderscore}parent\ tree{\isacartoucheclose}\isanewline
+\ \ \ \ \isakeyword{and}\ {\isacartoucheopen}distinct\ {\isacharparenleft}{\isacharparenleft}map\ log{\isacharunderscore}time\ log{\isacharparenright}\ {\isacharat}\isanewline
+\ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ {\isacharparenleft}map\ move{\isacharunderscore}time\ {\isacharbrackleft}oper{\isadigit{1}}{\isacharcomma}\ oper{\isadigit{2}}{\isacharbrackright}{\isacharparenright}{\isacharparenright}{\isacartoucheclose}\isanewline
+\ \ \isakeyword{shows}\ {\isacartoucheopen}apply{\isacharunderscore}op\ oper{\isadigit{2}}\ {\isacharparenleft}apply{\isacharunderscore}op\ oper{\isadigit{1}}\ {\isacharparenleft}log{\isacharcomma}\ tree{\isacharparenright}{\isacharparenright}\ {\isacharequal}\isanewline
+\ \ \ \ \ \ \ \ \ apply{\isacharunderscore}op\ oper{\isadigit{1}}\ {\isacharparenleft}apply{\isacharunderscore}op\ oper{\isadigit{2}}\ {\isacharparenleft}log{\isacharcomma}\ tree{\isacharparenright}{\isacharparenright}{\isacartoucheclose}
+\end{isabelle}
+
+\noindent Let \isa{t1 = move\_time oper1} and \isa{t2 = move\_time oper2} be the timestamps of the two operations \isa{oper1} and \isa{oper2}.
+Assume \isa{t1 < t2} without loss of generality.
+We can now prove that lemma by induction over \isa{log}.
+
+For the base case, where \isa{log} is empty, \isa{apply\_op oper2 (apply\_op oper1 (log, tree))} evaluates to first applying \isa{oper1} using \isa{do\_op}, then applying \isa{oper2} using \isa{do\_op}.
+On the other hand, \isa{apply\_op oper1 (apply\_op oper2 (log, tree))} first applies \isa{oper2} using \isa{do\_op}, then undoes \isa{oper2} again to yield the original \isa{tree} (per \isa{do\_undo\_op\_inv}), then applies \isa{oper1} using \isa{do\_op}, and finally reapplies \isa{oper2} using \isa{redo\_op}.
+The final states are the same on both sides of the equality.
+
+For the inductive step, let the log be \isa{logop {\isacharhash} ops} (where \isa{\isacharhash} is the \emph{cons} operator), and let \isa{t3 = log\_time logop} be the timestamp of the operation at the head of the log.
+We now consider three cases:
+\begin{enumerate}
+    \item \isa{t3 < t1 < t2}.
+        Then neither applying \isa{oper1} nor applying \isa{oper2} will undo \isa{logop}.
+        Thus, the two operations are applied in the same way as the base case.
+    \item \isa{t1 < t3 < t2}.
+        Then \isa{apply\_op oper2 (apply\_op oper1 (logop {\isacharhash} ops, tree))} first undoes \isa{logop}, then applies \isa{oper1} using \isa{do\_op}, then reapplies \isa{logop} using \isa{redo\_op}, and finally applies \isa{oper2} using \isa{do\_op}.
+        On the other hand, \isa{apply\_op oper1 (apply\_op oper2 (logop {\isacharhash} ops, tree))} first applies \isa{oper2} using \isa{do\_op}, then undoes it again, then undoes \isa{logop}, applies \isa{oper1} and reapplies \isa{logop} as before, and finally reapplies \isa{oper2} using \isa{redo\_op}.
+        The end result is the same.
+    \item \isa{t1 < t2 < t3}.
+        Then both applying \isa{oper1} and applying \isa{oper2} first undoes \isa{logop}, yielding \isa{tree' = undo\_op (logop, tree)}.
+        From the inductive hypothesis we have
+        \begin{quote}
+        \isa{apply\_op oper2 (apply\_op oper1 (ops, tree')) =\\
+             apply\_op oper1 (apply\_op oper2 (ops, tree'))}
+        \end{quote}
+        We now apply \isa{oper1} and \isa{oper2} as in the base case, and finally reapply \isa{logop} to both sides using \isa{redo\_op}, resulting in the same state on both sides.
+\end{enumerate}
 

### 2019-06-17 paper/move-op.tex
add isabelle lemmas

--- 2019-06-17 paper/move-op.tex
@@
 % DPM: add efficient_ancestor_simp here
+\begin{isabelle}
+\isacommand{theorem}\isamarkupfalse%
+\ efficient{\isacharunderscore}ancestor{\isacharunderscore}simp\ {\isacharbrackleft}code{\isacharbrackright}{\isacharcolon}\isanewline
+\ \ \isakeyword{shows}\ {\isacartoucheopen}efficient{\isacharunderscore}ancestor\ t\ p\ c\ {\isasymlongleftrightarrow}\isanewline
+\ \ \ \ \ \ \ \ \ \ {\isacharparenleft}case\ hm{\isachardot}lookup\ c\ t\ of\isanewline
+\ \ \ \ \ \ \ \ \ \ \ \ \ \ None\ {\isasymRightarrow}\ False\isanewline
+\ \ \ \ \ \ \ \ \ \ \ \ {\isacharbar}\ Some\ {\isacharparenleft}m{\isacharcomma}\ a{\isacharparenright}\ {\isasymRightarrow}\isanewline
+\ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ a\ {\isacharequal}\ p\ {\isasymor}\ efficient{\isacharunderscore}ancestor\ t\ p\ a{\isacharparenright}{\isacartoucheclose}
+\end{isabelle}
 
@@
 % DPM: add efficient_do_op_simulates here
+\begin{isabelle}
+\isacommand{lemma}\isamarkupfalse%
+\ efficient{\isacharunderscore}do{\isacharunderscore}op{\isacharunderscore}simulates{\isacharcolon}\isanewline
+\ \ \isakeyword{assumes}\ {\isadigit{1}}{\isacharcolon}\ {\isacartoucheopen}t\ {\isasympreceq}\ T{\isacartoucheclose}\isanewline
+\ \ \ \ \isakeyword{and}\ {\isadigit{2}}{\isacharcolon}\ {\isacartoucheopen}efficient{\isacharunderscore}do{\isacharunderscore}op\ {\isacharparenleft}oper{\isacharcomma}\ t{\isacharparenright}\ {\isacharequal}\ {\isacharparenleft}log{\isadigit{1}}{\isacharcomma}\ u{\isacharparenright}{\isacartoucheclose}\isanewline
+\ \ \ \ \isakeyword{and}\ {\isadigit{3}}{\isacharcolon}\ {\isacartoucheopen}do{\isacharunderscore}op\ {\isacharparenleft}oper{\isacharcomma}\ T{\isacharparenright}\ {\isacharequal}\ {\isacharparenleft}log{\isadigit{2}}{\isacharcomma}\ U{\isacharparenright}{\isacartoucheclose}\isanewline
+\ \ \isakeyword{shows}\ {\isacartoucheopen}log{\isadigit{1}}\ {\isacharequal}\ log{\isadigit{2}}\ {\isasymand}\ u\ {\isasympreceq}\ U{\isacartoucheclose}
+\end{isabelle}
 
@@
 %DPM: add efficient_apply_ops_simulates here
+\begin{isabelle}
+\isacommand{lemma}\isamarkupfalse%
+\ efficient{\isacharunderscore}apply{\isacharunderscore}ops{\isacharunderscore}simulates{\isacharcolon}\isanewline
+\ \ \isakeyword{assumes}\ {\isadigit{1}}{\isacharcolon}\ {\isacartoucheopen}efficient{\isacharunderscore}apply{\isacharunderscore}ops\ opers\ {\isacharequal}\ {\isacharparenleft}log{\isadigit{1}}{\isacharcomma}\ u{\isacharparenright}{\isacartoucheclose}\isanewline
+\ \ \ \ \isakeyword{and}\ {\isadigit{2}}{\isacharcolon}\ {\isacartoucheopen}apply{\isacharunderscore}ops\ opers\ {\isacharequal}\ {\isacharparenleft}log{\isadigit{2}}{\isacharcomma}\ U{\isacharparenright}{\isacartoucheclose}\isanewline
+\ \ \isakeyword{shows}\ {\isacartoucheopen}log{\isadigit{1}}\ {\isacharequal}\ log{\isadigit{2}}\ {\isasymand}\ u\ {\isasympreceq}\ U{\isacartoucheclose}
+\end{isabelle}
 
@@
 % DPM: add efficient_apply_ops_commutes and efficient_apply_ops_acyclic here
+\begin{isabelle}
+\isacommand{theorem}\isamarkupfalse%
+\ efficient{\isacharunderscore}apply{\isacharunderscore}ops{\isacharunderscore}commutes{\isacharcolon}\isanewline
+\ \ \isakeyword{assumes}\ {\isadigit{1}}{\isacharcolon}\ {\isacartoucheopen}set\ ops{\isadigit{1}}\ {\isacharequal}\ set\ ops{\isadigit{2}}{\isacartoucheclose}\isanewline
+\ \ \ \ \isakeyword{and}\ {\isadigit{2}}{\isacharcolon}\ {\isacartoucheopen}distinct\ {\isacharparenleft}map\ move{\isacharunderscore}time\ ops{\isadigit{1}}{\isacharparenright}{\isacartoucheclose}\isanewline
+\ \ \ \ \isakeyword{and}\ {\isadigit{3}}{\isacharcolon}\ {\isacartoucheopen}distinct\ {\isacharparenleft}map\ move{\isacharunderscore}time\ ops{\isadigit{2}}{\isacharparenright}{\isacartoucheclose}\isanewline
+\ \ \ \ \isakeyword{and}\ {\isadigit{4}}{\isacharcolon}\ {\isacartoucheopen}efficient{\isacharunderscore}apply{\isacharunderscore}ops\ ops{\isadigit{1}}\ {\isacharequal}\ {\isacharparenleft}log{\isadigit{1}}{\isacharcomma}\ t{\isacharparenright}{\isacartoucheclose}\isanewline
+\ \ \ \ \isakeyword{and}\ {\isadigit{5}}{\isacharcolon}\ {\isacartoucheopen}efficient{\isacharunderscore}apply{\isacharunderscore}ops\ ops{\isadigit{2}}\ {\isacharequal}\ {\isacharparenleft}log{\isadigit{2}}{\isacharcomma}\ u{\isacharparenright}{\isacartoucheclose}\isanewline
+\ \ \isakeyword{shows}\ {\isacartoucheopen}log{\isadigit{1}}\ {\isacharequal}\ log{\isadigit{2}}\ {\isasymand}\ hm{\isachardot}lookup\ c\ t\ {\isacharequal}\ hm{\isachardot}lookup\ c\ u{\isacartoucheclose}\isanewline
+\isanewline
+\isacommand{theorem}\isamarkupfalse%
+\ efficient{\isacharunderscore}apply{\isacharunderscore}ops{\isacharunderscore}acyclic{\isacharcolon}\isanewline
+\ \ \isakeyword{assumes}\ {\isadigit{1}}{\isacharcolon}\ {\isacartoucheopen}efficient{\isacharunderscore}apply{\isacharunderscore}ops\ ops\ {\isacharequal}\ {\isacharparenleft}log{\isacharcomma}\ t{\isacharparenright}{\isacartoucheclose}\isanewline
+\ \ \isakeyword{shows}\ {\isacartoucheopen}{\isasymnexists}n{\isachardot}\ efficient{\isacharunderscore}ancestor\ t\ n\ n{\isacartoucheclose}
+\end{isabelle}
 

### 2019-06-17 paper/move-op.tex
paper: make apply_op_commute2 inductive step more explicit

@@
 Assume \isa{t1 < t2} without loss of generality.
-We can now prove that lemma by induction over \isa{log}.
+We can now prove lemma \isa{apply\_op\_commute2} by induction over \isa{log}.
 
-For the base case, where \isa{log} is empty, \isa{apply\_op oper2 (apply\_op oper1 (log, tree))} evaluates to first applying \isa{oper1} using \isa{do\_op}, then applying \isa{oper2} using \isa{do\_op}.
-On the other hand, \isa{apply\_op oper1 (apply\_op oper2 (log, tree))} first applies \isa{oper2} using \isa{do\_op}, then undoes \isa{oper2} again to yield the original \isa{tree} (per \isa{do\_undo\_op\_inv}), then applies \isa{oper1} using \isa{do\_op}, and finally reapplies \isa{oper2} using \isa{redo\_op}.
+For the base case, where \isa{log} is empty, \isa{apply\_op oper2 (apply\_op oper1 ([], tree))} evaluates to first applying \isa{oper1} using \isa{do\_op}, then applying \isa{oper2} using \isa{do\_op}.
+On the other hand, \isa{apply\_op oper1 (apply\_op oper2 ([], tree))} first applies \isa{oper2} using \isa{do\_op}, then undoes \isa{oper2} again to yield the original \isa{tree} (per \isa{do\_undo\_op\_inv}), then applies \isa{oper1} using \isa{do\_op}, and finally reapplies \isa{oper2} using \isa{redo\_op}.
 The final states are the same on both sides of the equality.
 
-For the inductive step, let the log be \isa{logop {\isacharhash} ops} (where \isa{\isacharhash} is the \emph{cons} operator), and let \isa{t3 = log\_time logop} be the timestamp of the operation at the head of the log.
+For the inductive step, let the log be \isa{logop {\isacharhash} ops} (where \isa{\isacharhash} is the \emph{cons} operator).
+We now need to show that
+\begin{isabelle}
+apply\_op oper2 (apply\_op oper1 (logop {\isacharhash} ops, tree)) =\\
+apply\_op oper1 (apply\_op oper2 (logop {\isacharhash} ops, tree)).
+\end{isabelle}
+Let \isa{t3 = log\_time logop} be the timestamp of the operation at the head of the log.
 We now consider three cases:
@@
 
+\subsection{Performance experiments}
+
+TODO\dots
+
 \section{Related work}\label{sec:relwork}

### 2019-06-17 paper/move-op.tex
paper: integration with OOPSLA framework

@@
 
-\subsection{Commutativity of apply\_op}
+\subsection{Convergence}
 
@@
 Under these assumptions, \isa{apply\_ops\_commutes} proves that applying the list of operations \isa{ops1} results in the same replica state as applying the list of operations \isa{ops2}.
+
+\subsubsection{Commutativity of apply\_op}
 
@@
 \end{enumerate}
+
+\subsubsection{Strong eventual consistency}
+
+Gomes et al.~\cite{Gomes:2017gy} define a framework in Isabelle/HOL for proving the strong eventual consistency properties of CRDTs.
+Based on the commutativity proof above we integrate our tree datatype with this framework, and thus prove that our move operation on trees does indeed guarantee strong eventual consistency.
+The details of this proof are given in the open source Isabelle source files.\footnote{\url{https://github.com/trvedata/move-op}}
 

### 2019-06-17 paper/move-op.tex
paper: added missing references in Section 5

@@
 
-Isabelle/HOL can generate executable Haskell, OCaml, Scala, and Standard ML code from HOL type and constant definitions using a sophisticated code generation mechanism~\cite{XXX}.
+Isabelle/HOL can generate executable Haskell, OCaml, Scala, and Standard ML code from HOL type and constant definitions using a sophisticated code generation mechanism~\cite{DBLP:conf/flops/HaftmannN10}.
 However,  care must be taken when using this mechanism, as:
@@
 As a result, prior to evaluating our algorithms in a real-world setting, we first produce a series of refined definitions that are optimised for efficient execution rather than ease of theorem proving.
-Within these definitions, we eschew the HOL \isa{set} type, and instead use a more efficient hash-map implementation provided by the Isabelle Collections Framework~\cite{XXX}.
+Within these definitions, we eschew the HOL \isa{set} type, and instead use a more efficient hash-map implementation provided by the Isabelle Collections Framework~\cite{DBLP:conf/itp/LammichL10}.
 Write $('a::\{hashable\}, 'b)\ \isa{hm}$ for the type of hash-maps with key type $\isa{'a}$ and binding type $\isa{'b}$.

### 2019-06-17 paper/move-op.tex
paper: tightening up Section 5
Rewording some sentences.  Removing flab.

@@
 \item
-The mechanism is \emph{partial} in the sense that not all HOL definitions can be realised in an executable form.
-Definitions making essential use of choice principles (such as our \isa{get\_parent} function) are obvious examples of definitions that will cause problems.
-However, \emph{inductively defined} relations characterised by a series of introduction rules, such as our \isa{ancestor} relation are also problematic for the code generator as making them executable necessarily requires Prolog-like proof search and mode inference.
+The mechanism is \emph{partial}: not all HOL definitions can be realised in executable form.
+Definitions making essential use of choice principles (e.g.~our \isa{get\_parent}) are obvious examples of definitions that will cause problems.
+However, \emph{inductively defined} relations characterised by a series of introduction rules, such as our \isa{ancestor} relation, are also problematic for the code generator as making them executable necessarily requires Prolog-like proof search and mode inference.
 \item
-Unsophisticated use of the code generator often leads to inefficient target language code.
+Na\"{i}ve use of the code generator often leads to inefficient target language code.
 This is especially true when using HOL's \isa{set} type, a type of potentially infinite mathematical sets.
@@
 
-As a result, prior to evaluating our algorithms in a real-world setting, we first produce a series of refined definitions that are optimised for efficient execution rather than ease of theorem proving.
-Within these definitions, we eschew the HOL \isa{set} type, and instead use a more efficient hash-map implementation provided by the Isabelle Collections Framework~\cite{DBLP:conf/itp/LammichL10}.
+We must therefore first produce a series of refined definitions that are optimised for efficient execution rather than ease of theorem proving, before trying to evaluate our algorithms.
+Within these definitions, we eschew the HOL \isa{set} type, and instead use a more efficient hash-map implementation provided by the Isabelle Collections Framework~\cite{DBLP:conf/itp/LammichL10}, a library of verified, efficient container datatypes and associated algorithms.
 Write $('a::\{hashable\}, 'b)\ \isa{hm}$ for the type of hash-maps with key type $\isa{'a}$ and binding type $\isa{'b}$.
-Note that keys have an explicit $\isa{hashable}$ constraint, meaning that this hash-map type may only be used with types of keys with an associated hashing function.
-With this type, we represent trees by the type $(\isa{'n}::\{hashable\}, \isa{'m} \times \isa{'n})\ \isa{hm}$.
-Intuitively, this type can be thought of as a mapping from a \emph{child} node to its \emph{parent} node and associated metadata, if any such mapping exists.
+Note that keys have an explicit $\isa{hashable}$ constraint, meaning that this hash-map type may only be used with kay types with an associated hashing function.
+With this type, we represent trees by the type $(\isa{'n}::\{hashable\}, \isa{'m} \times \isa{'n})\ \isa{hm}$, mapping from a \emph{child} node to its \emph{parent} node and associated metadata, if any such mapping exists.
 
 We introduce an explicit \emph{simulation relation} between our two representations of trees.
-Specifically, we say that a hash-map $\isa{t}$ of type $(\isa{'n}, \isa{'m} \times \isa{'n})\ \isa{hm}$ simulates a set $T$ of type $(\isa{'n} \times \isa{'m} \times \isa{'n})\ \isa{set}$ whenever
+Specifically, we say that a hash-map $\isa{t}$ of type $(\isa{'n}, \isa{'m} \times \isa{'n})\ \isa{hm}$ \emph{simulates} a set $T$ of type $(\isa{'n} \times \isa{'m} \times \isa{'n})\ \isa{set}$ whenever
 \begin{displaymath}
@@
 \end{displaymath}
-where $hm.lookup$ is the function that searches for a binding for key $\isa{c}$ in hash-map $t$, returning $\isa{Some}\ (m, p)$ when a binding from $c$ to $(m, p)$ exists in the hash-map, or $\isa{None}$ otherwise.
-We write $t \preceq T$ to assert that the hash-map $t$ and set $T$ are contained within the simulation relation, as defined above.
-Note that many useful corollaries follow immediately from this definition, for example we have $t \preceq T$ implies $\isa{unique\_parent}\ \isa{T}$.
+where $hm.lookup$ searches for a binding for key $\isa{c}$ in hash-map $t$, returning $\isa{Some}\ (m, p)$ when a binding from $c$ to $(m, p)$ exists in the hash-map, or $\isa{None}$ otherwise.
+We write $t \preceq T$ to assert that the hash-map $t$ and set $T$ are contained within the simulation relation, defined above.
+Note some useful corollaries follow immediately from this definition: e,g,~we have $t \preceq T$ implies $\isa{unique\_parent}\ \isa{T}$.
 
 Our general strategy henceforth is to produce a new set of definitions---mirroring the structure of our existing definitions---wherein we replace all occurrences of the $(\isa{'n} \times \isa{'m} \times \isa{'n})\ \isa{set}$ type, and any associated set operations, with the $(\isa{'n}, \isa{'m} \times \isa{'n})\ \isa{hm}$ type and associated hash-map operations.
-We adopt the convention that any definition using hash-maps will be referred to as an ``efficient'' definition, and any definition making use of HOL's set type will be referred to as an ``abstract'' definition, for clarity.
-
-For most definitions this is an almost mechanical process, and we therefore elide further description here.
-However, \isa{ancestor} is slightly different to our other definitions, as it is an inductively defined relation rather than a recursive function, and we therefore need further work to make this efficiently executable.
-Specifically, we can characterise this relation recursively with the following theorem, which we prove inside Isabelle/HOL:
+We adopt the convention that any definition using hash-maps will be referred to as an ``efficient'' definition, and any definition making use of HOL's set type will be referred to as an ``abstract'' definition, for the sake of clarity.
+For most definitions this is a mechanical process, and we elide further description here.
+However \isa{ancestor} is slightly different, as it is an inductively defined relation rather than a recursive function, and we therefore need further work to make this efficiently executable.
+We characterise this relation recursively with the following theorem:
 
@@
 
-Note that the theorem \isa{efficient\_ancestor\_simp} is tagged with a $\isa{code}$ attribute, which is a hint that this theorem should be used as a transformation during the code generation process.
+Note that the theorem \isa{efficient\_ancestor\_simp} is tagged with a $\isa{code}$ ``attribute'', which is a hint that this theorem should be used as a transformation during the code generation process.
 Isabelle/HOL takes this theorem as the definition, for code generation, of the \isa{efficient\_ancestor} relation and is able to directly translate this to a recursive function definition in each target backend.
@@
 Next, we demonstrate the correctness of our new efficient definitions against our older, abstract definitions.
-The correctness properties that we prove are best described using an example, as they all follow a similar pattern, and we therefore focus on the lemma \isa{efficient\_do\_op\_simulates} which states the following property:
+The correctness properties that we prove are best described using an example, as they all follow a similar pattern.
+We therefore focus on the lemma \isa{efficient\_do\_op\_simulates} which states:
 
@@
 
-Note that we show that the \isa{efficient\_do\_op} and the \isa{do\_op} functions maintain a ``lock-step'' property, mapping related trees $t \preceq T$ to related trees $u \preceq U$ and producing identical logs of operations, whenever they are applied to the same input operation, \isa{oper}.
+Note that we show that the \isa{efficient\_do\_op} and the \isa{do\_op} functions maintain a ``lock-step'' property, mapping related trees $t \preceq T$ to related trees $u \preceq U$ and producing identical logs of operations whenever they are applied to the same input operation, \isa{oper}.
 This pattern of lock-step simulation is repeated bottom-up for every pair of efficient-abstract algorithm pairs, until we reach the final correctness theorem for \isa{apply\_ops}:
@@
 
-Note that this property correctly captures our intuitive understanding of what it means for \isa{efficient\_apply\_ops} to be correct with respect to \isa{apply\_ops}.
-Specifically, the theorem states that if both algorithms are applied to the same set of operations then they produce identical logs and also produce trees that contain the same set of parent-metadata-child records, i.e.
+This property captures our intuitive understanding of what it means for \isa{efficient\_apply\_ops} to be correct with respect to \isa{apply\_ops}: if both algorithms are applied to the same set of operations then they produce identical logs and also produce trees that contain the same set of parent-metadata-child records---i.e.
 the two trees are extensionally equivalent, despite having very different concrete representations.
 We can further check that this theorem correctly captures the correctness of our \isa{apply\_ops} definition by recasting the various tree invariant and convergence theorems in terms of \isa{efficient\_apply\_ops}.
-For example:
+For example, we have the following corollaries of the theorems \isa{apply\_ops\_commutes} and \isa{apply\_ops\_acyclic}, respectively\footnote{The ``efficient'' analogue of \isa{apply\_ops\_unique\_parent} is trivially true, as the output tree is stored as a hash-map, and therefore too trivial to state as a theorem.}:
 
@@
 
-Both follow as corollaries of the theorems \isa{apply\_ops\_commutes} and \text{apply\_ops\_acyclic}, respectively.
-Note that the statement of \isa{efficient\_apply\_ops\_commutes} differs slightly from the statement of \isa{apply\_ops\_commutes}---the latter theorem asserts that identical sets are obtained whilst the former asserts that simply reading any key binding in $t$ gives the same result as reading in $u$.
-Whilst this statement appears to be weaker than the statement of \text{apply\_ops\_acyclic}, it is correct, as the resulting hash-maps may not have an identical representation in-memory, depending on the particular ordering of key-value insertions, despite containing the same key-value bindings.
+Note the statement of \isa{efficient\_apply\_ops\_commutes} differs in an essential way from that of \isa{apply\_ops\_commutes}---the latter asserts that identical sets are obtained from convergence, whilst the former asserts that simply reading any key binding in $t$ gives the same result as reading in $u$.
+Whilst this statement appears weaker than the statement of \isa{apply\_ops\_acyclic}, it is correct, as the resulting hash-maps may not have an identical representation in-memory---depending on the particular ordering of key-value insertions---despite containing the same key-value bindings.
 

### 2019-06-17 paper/move-op.tex
paper: minor changes to description of get_parent
paper: more minor rewording as reading through...

@@
         As discussed in Section~\ref{sec:impossibility}, this has previously been thought to be impossible to achieve \cite{Najafzadeh:2017vk,Najafzadeh:2018bw}.
-    \item We formalise our abstract algorithm using Isabelle/HOL~\cite{DBLP:conf/tphol/WenzelPN08}, an interactive proof assistant based on higher-order logic, and obtain a computer-checked proof of correctness of our algorithm.
+    \item We formalise an abstract algorithm using Isabelle/HOL~\cite{DBLP:conf/tphol/WenzelPN08}, an interactive proof assistant based on higher-order logic, and obtain a computer-checked proof of correctness of the algorithm.
         In particular, we prove that arbitrary concurrent modifications to the tree can be merged such that all replicas converge towards a consistent state (strong eventual consistency) while keeping the data in a tree structure.
@@
 A move operation is a record consisting of four fields: a timestamp \isa{t} of type \isa{'t}, a parent node \isa{p} of type \isa{'n}, a metadata field \isa{m} of type \isa{'m}, and a child node \isa{c} of type \isa{'n}.
-Here, \isa{'t}, \isa{'n} and \isa{'m} are \emph{type variables} that can be instantiated with arbitrary datatypes; our only requirement is that timestamps \isa{'t} are globally unique and have a total order.
-For example, Lamport timestamps~\cite{Lamport:1978jq} meet these requirements.
+Here, \isa{'t}, \isa{'n} and \isa{'m} are \emph{type variables} that can be \emph{substituted} for arbitrary types; our only requirement is that timestamps \isa{'t} are globally unique and have an associated total order.
+Concretely, Lamport timestamps~\cite{Lamport:1978jq} would meet these requirements.
 
@@
 A \isa{LogMove t oldp p m c} record (lines 7--12) is similar to a record \isa{Move t p m c}; the only difference is that the \isa{LogMove} has an additional field \isa{oldp} of type \isa{('n {\isasymtimes} 'm) option}.
-This \isa{option} type means the field can either take the value \isa{None} (equivalent to null), or a pair of an inode and a metadata field.
+This \isa{option} type means the field can either take the value \isa{None} (intuitively similar to null), or a pair of an inode and a metadata field.
 
@@
 It then returns a \isa{('n {\isasymtimes} 'm) option}.
-\isa{{\isasymexists}!x} means ``there exists a unique value \isa{x} such that\dots''.
+\isa{{\isasymexists}!x} means ``there exists a unique value \isa{x} such that\dots'', whilst $\isa{THE}\ \isa{x}. \ldots$ means ``\emph{choose} the unique value \isa{x} such that\ldots''.
 
@@
 This relation is the transitive closure of the parent-child relation of a tree: if \isa{(p, m, c) {\isasymin} tree} then \isa{p} is an ancestor of \isa{c} (line 23); moreover, if \isa{(p, m, c) {\isasymin} tree} and \isa{a} is an ancestor of \isa{p}, then \isa{a} is also an ancestor of \isa{c} (line 24).
-The \textbf{inductive} keyword indicates that this recursive definition is iterated until a fixed point is reached.
+The \textbf{inductive} keyword indicates that this recursive definition is iterated until a fixed point (the least) is reached.
 
@@
 The \isa{apply\_op} function (lines 43--51) takes two arguments: a \isa{Move} operation to apply and the current replica state; and it returns the new replica state.
-The expression \isa{'t::\{linorder\}} in the type signature indicates that the timestamps \isa{'t} belong to the type class \isa{linorder}, which means that timestamps can be compared with the \isa{<} operator defining a total order.
+The constraint \isa{'t::\{linorder\}} in the type signature indicates that timestamps \isa{'t} are instances of \isa{linorder} type class, and timestamps can therefore be compared with the \isa{<} operator defining a total order.
 This comparison occurs on line 49.

### 2019-06-17 paper/move-op.tex
Some minor clarifications or tidying up.

--- 2019-06-17 paper/move-op.tex
@@
 
-In this paper we consider applications that use such a tree data model, and that replicate this tree across multiple replicas.
+In this paper we consider applications that use such a tree data model, and that replicate this tree across multiple nodes.
 Moreover, we focus on \emph{optimistic replication}~\cite{Saito:2005jw}, that is, systems in which any replica can autonomously make changes to the data, without waiting for communication or coordination with any other replicas.
@@
 In an unreplicated tree it is easy to prevent cycles being introduced: if the node being moved is an ancestor of the destination node, then the move operation must be rejected.
-However, a new problem arises in a replicated setting: different replicas may perform operations that individually are safe, but whose combination leads to a cycle.
+However, a new problem arises in a replicated setting: different replicas may perform operations that are individually safe, but whose combination leads to a cycle.
 
@@
 If \isa{c} did not exist in the tree prior to the move, \isa{oldp} is set to \isa{None}.
-Otherwise, \isa{oldp} records the previous parent and metadata of \isa{c}, that is: if there exist \isa{p'}, \isa{m'} such that \isa{(p', m', c) {\isasymin} tree}, then \isa{oldp} is set to \isa{Some (p', m')}.
+Otherwise, \isa{oldp} records the previous parent and metadata of \isa{c}, that is: if there exists a \isa{p'} and \isa{m'} such that \isa{(p', m', c) {\isasymin} tree}, then \isa{oldp} is set to \isa{Some (p', m')}.
 This logic is performed by the \isa{get\_parent} function (lines 16--20).
@@
 
-Line 30 performs the check that ensures no cycles are introduced: if \isa{ancestor tree c newp}, i.e.\ if the node being moved \isa{c} is an ancestor of the new parent \isa{newp}, then the tree is returned unmodified---in other words, the operation is ignored.
+Line 30 performs the check that ensures no cycles are introduced: if \isa{ancestor tree c newp}, i.e.\ if the node \isa{c} is being moved, and \isa{c} is an ancestor of the new parent \isa{newp}, then the tree is returned unmodified---in other words, the operation is ignored.
 Similarly, if \isa{c = newp}, the operation is also ignored.
@@
 Our general strategy henceforth is to produce a new set of definitions---mirroring the structure of our existing definitions---wherein we replace all occurrences of the $(\isa{'n} \times \isa{'m} \times \isa{'n})\ \isa{set}$ type, and any associated set operations, with the $(\isa{'n}, \isa{'m} \times \isa{'n})\ \isa{hm}$ type and associated hash-map operations.
-We adopt the convention that any definition using hash-maps will be referred to as an ``efficient'' definition, and any definition making use of HOL's set type will be referred to as an ``abstract'' definition, for the sake of clarity.
+For the sake of clarity, we adopt the convention that any definition using hash-maps will be referred to as an ``efficient'' definition, and any definition making use of HOL's set type will be referred to as an ``abstract'' definition.
 For most definitions this is a mechanical process, and we elide further description here.

### 2019-06-18 paper/move-op.tex
paper: more tightening, a few lines saved here and there
Excess language flab cut down.

@@
 Definitions making essential use of choice principles (e.g.~our \isa{get\_parent}) are obvious examples of definitions that will cause problems.
-However, \emph{inductively defined} relations characterised by a series of introduction rules, such as our \isa{ancestor} relation, are also problematic for the code generator as making them executable necessarily requires Prolog-like proof search and mode inference.
+However, \emph{inductively defined} relations characterised by a series of introduction rules, such as our \isa{ancestor} relation, are also problematic as their execution necessarily requires Prolog-like proof search.
 \item
-Na\"{i}ve use of the code generator often leads to inefficient target language code.
-This is especially true when using HOL's \isa{set} type, a type of potentially infinite mathematical sets.
-Whilst convenient for theorem proving purposes, this type often leads to gross inefficiency in the extracted code.
+Na\"{i}ve use often leads to inefficient target language code, especially true when using HOL's \isa{set} type, a type of potentially infinite sets.
+Whilst convenient for theorem proving this type often leads to inefficiency in generated code.
 \end{enumerate}
 
-We must therefore first produce a series of refined definitions that are optimised for efficient execution rather than ease of theorem proving, before trying to evaluate our algorithms.
-Within these definitions, we eschew the HOL \isa{set} type, and instead use a more efficient hash-map implementation provided by the Isabelle Collections Framework~\cite{DBLP:conf/itp/LammichL10}, a library of verified, efficient container datatypes and associated algorithms.
+We therefore produce a series of definitions that are optimised for execution rather than theorem proving, before trying to evaluate our algorithms.
+Notably, we eschew the HOL \isa{set} type, and instead use a more efficient hash-map implementation provided by the Isabelle Collections Framework~\cite{DBLP:conf/itp/LammichL10}, a library of verified, efficient container datatypes and associated algorithms.
+
 Write $('a::\{hashable\}, 'b)\ \isa{hm}$ for the type of hash-maps with key type $\isa{'a}$ and binding type $\isa{'b}$.
@@
 
-We introduce an explicit \emph{simulation relation} between our two representations of trees.
-Specifically, we say that a hash-map $\isa{t}$ of type $(\isa{'n}, \isa{'m} \times \isa{'n})\ \isa{hm}$ \emph{simulates} a set $T$ of type $(\isa{'n} \times \isa{'m} \times \isa{'n})\ \isa{set}$ whenever
+We use a \emph{simulation relation} between our two representations of trees, and say that a hash-map $\isa{t}$ of type $(\isa{'n}, \isa{'m} \times \isa{'n})\ \isa{hm}$ \emph{simulates} a set $T$ of type $(\isa{'n} \times \isa{'m} \times \isa{'n})\ \isa{set}$ whenever
 \begin{displaymath}
@@
 where $hm.lookup$ searches for a binding for key $\isa{c}$ in hash-map $t$, returning $\isa{Some}\ (m, p)$ when a binding from $c$ to $(m, p)$ exists in the hash-map, or $\isa{None}$ otherwise.
-We write $t \preceq T$ to assert that the hash-map $t$ and set $T$ are contained within the simulation relation, defined above.
-Note some useful corollaries follow immediately from this definition: e,g,~we have $t \preceq T$ implies $\isa{unique\_parent}\ \isa{T}$.
+Write $t \preceq T$ to assert that $t$ and $T$ are contained within the simulation relation.
+Note some corollaries follow immediately: e,g,~we have $t \preceq T$ implies $\isa{unique\_parent}\ \isa{T}$.
 
-Our general strategy henceforth is to produce a new set of definitions---mirroring the structure of our existing definitions---wherein we replace all occurrences of the $(\isa{'n} \times \isa{'m} \times \isa{'n})\ \isa{set}$ type, and any associated set operations, with the $(\isa{'n}, \isa{'m} \times \isa{'n})\ \isa{hm}$ type and associated hash-map operations.
-For the sake of clarity, we adopt the convention that any definition using hash-maps will be referred to as an ``efficient'' definition, and any definition making use of HOL's set type will be referred to as an ``abstract'' definition.
-For most definitions this is a mechanical process, and we elide further description here.
-However \isa{ancestor} is slightly different, as it is an inductively defined relation rather than a recursive function, and we therefore need further work to make this efficiently executable.
-We characterise this relation recursively with the following theorem:
+Our approach is to produce a new set of definitions---mirroring the structure of our existing ones---wherein we replace all occurrences of the $(\isa{'n} \times \isa{'m} \times \isa{'n})\ \isa{set}$ type, and any associated set operations, with the $(\isa{'n}, \isa{'m} \times \isa{'n})\ \isa{hm}$ type and associated hash-map operations.
+For clarity, we refer to any definition using hash-maps as an ``efficient'' definition, and any definition making use of the \isa{set} type as an ``abstract'' definition.
+For most definitions this is a mechanical process, and we elide further discussion of the required changes, referring the interested reader to our Isabelle/HOL source code.
+However \isa{ancestor} is different to other definitions as it is an inductively defined relation rather than a recursive function, and we need further work to make this efficiently executable.
+We characterise this relation with the following:
 
@@
 
-Note that the theorem \isa{efficient\_ancestor\_simp} is tagged with a $\isa{code}$ ``attribute'', which is a hint that this theorem should be used as a transformation during the code generation process.
-Isabelle/HOL takes this theorem as the definition, for code generation, of the \isa{efficient\_ancestor} relation and is able to directly translate this to a recursive function definition in each target backend.
+Note that \isa{efficient\_ancestor\_simp} is tagged with a $\isa{code}$ ``attribute'', a hint that this theorem should be used as a transformation during code generation.
+Isabelle/HOL takes this theorem as the definition, for code generation, of the \isa{efficient\_ancestor} relation and is able to directly translate this to a recursive function definition in each target backend---this transformation is ``semantics preserving'', as the characterisation was proved as a theorem within HOL.
 
-Next, we demonstrate the correctness of our new efficient definitions against our older, abstract definitions.
-The correctness properties that we prove are best described using an example, as they all follow a similar pattern.
-We therefore focus on the lemma \isa{efficient\_do\_op\_simulates} which states:
+Next, we demonstrate the correctness of our efficient definitions against our abstract definitions, and the properties that we prove are best described using an example, as they all follow a similar pattern.
+We use the lemma \isa{efficient\_do\_op\_simulates} as our exemplar:
 
@@
 
-Note that we show that the \isa{efficient\_do\_op} and the \isa{do\_op} functions maintain a ``lock-step'' property, mapping related trees $t \preceq T$ to related trees $u \preceq U$ and producing identical logs of operations whenever they are applied to the same input operation, \isa{oper}.
-This pattern of lock-step simulation is repeated bottom-up for every pair of efficient-abstract algorithm pairs, until we reach the final correctness theorem for \isa{apply\_ops}:
+We show that the \isa{efficient\_do\_op} and the \isa{do\_op} functions are in ``lock-step'', mapping related trees $t \preceq T$ to related trees $u \preceq U$ and producing identical logs of operations whenever they are applied to the same input operation, \isa{oper}.
+This lock-step is repeated bottom-up for every efficient-abstract algorithm pairs, until we reach the final correctness theorem for \isa{apply\_ops}:
 
@@
 
-This property captures our intuitive understanding of what it means for \isa{efficient\_apply\_ops} to be correct with respect to \isa{apply\_ops}: if both algorithms are applied to the same set of operations then they produce identical logs and also produce trees that contain the same set of parent-metadata-child records---i.e.
+Intuitively, this states that if both algorithms are applied to the same set of operations then they produce identical logs and also produce trees that contain the same set of parent-metadata-child records---i.e.
 the two trees are extensionally equivalent, despite having very different concrete representations.
-We can further check that this theorem correctly captures the correctness of our \isa{apply\_ops} definition by recasting the various tree invariant and convergence theorems in terms of \isa{efficient\_apply\_ops}.
-For example, we have the following corollaries of the theorems \isa{apply\_ops\_commutes} and \isa{apply\_ops\_acyclic}, respectively\footnote{The ``efficient'' analogue of \isa{apply\_ops\_unique\_parent} is trivially true, as the output tree is stored as a hash-map, and therefore too trivial to state as a theorem.}:
+Moreover, we have the following corollaries of the theorems \isa{apply\_ops\_commutes} and \isa{apply\_ops\_acyclic}, respectively, further establishing the correctness of our efficient algorithms:\footnote{The ``efficient'' analogue of \isa{apply\_ops\_unique\_parent} is trivially true, as the output tree is stored as a hash-map, and therefore too trivial to state as a theorem.}:
 
@@
 
-Note the statement of \isa{efficient\_apply\_ops\_commutes} differs in an essential way from that of \isa{apply\_ops\_commutes}---the latter asserts that identical sets are obtained from convergence, whilst the former asserts that simply reading any key binding in $t$ gives the same result as reading in $u$.
-Whilst this statement appears weaker than the statement of \isa{apply\_ops\_acyclic}, it is correct, as the resulting hash-maps may not have an identical representation in-memory---depending on the particular ordering of key-value insertions---despite containing the same key-value bindings.
+Note the statement of \isa{efficient\_apply\_ops\_commutes} differs from that of \isa{apply\_ops\_commutes}---the latter asserts identical sets are obtained from convergence, whilst the former asserts that reading any key binding in $t$ gives the same result as a reading in $u$.
+Whilst this statement appears weaker than that of \isa{apply\_ops\_acyclic}, it is correct, and the strongest statement one can hope for: the resulting hash-maps may contain the same key-value bindings but may not be structurally equal, depending on the ordering of operations.
 

### 2019-06-18 evaluation/src/main/scala/TestClient.scala
evaluation: add dropwizard metrics

@@ -2,38 +2,48 @@ import examplerpc.{ MoveService, MoveServiceClient }
 import akka.actor.ActorSystem
 import akka.grpc.GrpcClientSettings
 import akka.stream.{ ActorMaterializer, Materializer }
+import com.codahale.metrics.{ ConsoleReporter, MetricRegistry }
+import java.util.concurrent.TimeUnit.SECONDS
 import scala.concurrent.{ ExecutionContext, Future }
 import scala.concurrent.duration._
 import scala.util.{ Failure, Success }
 
 object TestClient {
 
+  val metrics  = new MetricRegistry()
+  val requests = metrics.timer("TestClient.requests")
+  val errors   = metrics.meter("TestClient.errors")
+
   def main(args: Array[String]): Unit = {
     implicit val sys: ActorSystem = ActorSystem("TestClient")
     implicit val mat: Materializer = ActorMaterializer()
     implicit val ec: ExecutionContext = sys.dispatcher
 
-    val config = GrpcClientSettings.connectToServiceAt("localhost", 8080)
-      .withDeadline(1.second)
+    ConsoleReporter.forRegistry(metrics).build().start(20, SECONDS)
+
+    val config = GrpcClientSettings.connectToServiceAt("52.37.96.68", 8080)
+      .withDeadline(10 seconds)
       .withTls(false)
     val client: MoveService = MoveServiceClient(config)
 
     runSingleRequestReplyExample()
 
-    sys.scheduler.schedule(1.second, 1.second) {
+    sys.scheduler.schedule(10 millis, 10 millis) {
       runSingleRequestReplyExample()
     }
 
     def runSingleRequestReplyExample(): Unit = {
-      sys.log.info("Performing request")
+      val timer = requests.time()
       val timestamp = examplerpc.LamportTS(0, "a")
       val operation = examplerpc.Move(Some(timestamp), "parent", "meta", "child")
       val reply = client.sendMove(operation)
       reply.onComplete {
         case Success(msg) =>
-          println(s"got single reply: $msg")
+          timer.stop()
         case Failure(e) =>
-          println(s"Error response: $e")
+          sys.log.info(s"Error response: $e")
+          timer.stop()
+          errors.mark()
       }
     }
   }

### 2019-06-18 evaluation/src/main/scala/TestServer.scala

@@ -2,30 +2,37 @@ import akka.actor.ActorSystem
 import akka.http.scaladsl.UseHttp2.Always
 import akka.http.scaladsl.model.{ HttpRequest, HttpResponse }
 import akka.http.scaladsl.{ Http, HttpConnectionContext }
+import akka.pattern.ask
 import akka.stream.{ ActorMaterializer, Materializer }
+import akka.util.Timeout
+import com.codahale.metrics.{ ConsoleReporter, MetricRegistry }
 import com.typesafe.config.ConfigFactory
 import examplerpc.MoveServiceHandler
 import scala.concurrent.{ ExecutionContext, Future }
+import scala.concurrent.duration._
 
 object TestServer {
+  val metrics = new MetricRegistry()
 
   def main(args: Array[String]): Unit = {
+    ConsoleReporter.forRegistry(metrics).build().start(20, SECONDS)
+
     val conf = ConfigFactory.parseString("akka.http.server.preview.enable-http2 = on")
       .withFallback(ConfigFactory.defaultApplication())
     val system = ActorSystem("TestServer", conf)
-    new TestServer(system).run()
+    new TestServer(system, metrics).run()
     // ActorSystem threads will keep the app alive until `system.terminate()` is called
   }
 }
 
-class TestServer(system: ActorSystem) {
+class TestServer(system: ActorSystem, metrics: MetricRegistry) {
 
   def run(): Future[Http.ServerBinding] = {
     implicit val sys: ActorSystem = system
     implicit val mat: Materializer = ActorMaterializer()
     implicit val ec: ExecutionContext = sys.dispatcher
 
-    val service: HttpRequest => Future[HttpResponse] = MoveServiceHandler(new TestService(system))
+    val service: HttpRequest => Future[HttpResponse] = MoveServiceHandler(new TestService(system, metrics))
 
     val binding = Http().bindAndHandleAsync(service, interface = "0.0.0.0", port = 8080,
       connectionContext = HttpConnectionContext(http2 = Always))
@@ -35,5 +42,24 @@ class TestServer(system: ActorSystem) {
     }
 
     binding
+  }
+}
+
+class TestService(system: ActorSystem, metrics: MetricRegistry) extends examplerpc.MoveService {
+  val requests  = metrics.timer("TestService.requests")
+  val treeActor = system.actorOf(TreeActor.props("a"), "treeActor")
+
[…]

### 2019-06-18 evaluation/src/main/scala/TestService.scala

@@ -1,19 +1,0 @@
-import akka.actor.ActorSystem
-import akka.pattern.ask
-import akka.util.Timeout
-import scala.concurrent.Future
-import scala.concurrent.duration._
-
-class TestService(system: ActorSystem) extends examplerpc.MoveService {
-
-  val treeActor = system.actorOf(TreeActor.props("a"), "treeActor")
-
-  override def sendMove(move: examplerpc.Move): Future[examplerpc.LamportTS] = {
-    println(s"received: ${move}")
-
-    implicit val timeout: Timeout = 3.seconds
-    import system.dispatcher
-
-    (treeActor ? move).mapTo[examplerpc.LamportTS]
-  }
-}

### 2019-06-18 evaluation/src/main/scala/TreeActor.scala

@@ -20,7 +20,7 @@ class TreeActor(actorId: String) extends Actor {
       val ts = (int(timestamp.get.counter), timestamp.get.replica)
       val operation = generated.Move(ts, parent, metadata, child)
       state = generated.example_apply_op(operation)(state)
-      log.info("Applied %s".format(operation))
+      //log.info("Applied %s".format(operation))
       sender() ! timestamp.get
     }
 

### 2019-06-18 paper/move-op.tex
paper: better reference for Coda
paper: tidy up discussion of hardlinks

@@
 \begin{itemize}
-    \item Most \textbf{filesystems}, e.g.
-on Windows and Unix-like OSes, are organised as a tree: folders or directories are branch nodes, and files are leaf nodes.\footnote{In filesystems that support hardlinks, the structure is---strictly speaking---a restricted DAG in which a file inode can be referenced from multiple places in the directory tree, but directories cannot have more than one parent directory.
-        However, for the purposes of this paper, we will treat a filesystem as a tree; given such a tree, it is easy to add hardlinks and symlinks as features on top.}
+    \item The \textbf{filesystem} on most operating systems is organised as a tree: folders or directories are branch nodes, and files are leaf nodes.
         In Unix-style filesystems, the nodes of the tree are known as \emph{inodes}.
+        Strictly speaking, filesystems that support hardlinks form a DAG, but we will focus on trees for now and show in Section~\ref{sec:hardlinks} how to add support for links.
     \item \textbf{Rich text editors} maintain a tree of textual structures such as paragraphs, bulleted and numbered lists, figures, sections, and so on.
@@
 
-The algorithm as specified in Figure~\ref{fig:code} retains operations in the log indefinitely, so the memory/storage use grows without bound.
+\subsubsection{Hardlinks and symlinks}\label{sec:hardlinks}
+
+Unix filesystems support hardlinks, which allow the same file inode to be referenced from multiple locations in a tree.
+Our tree data structure can easily be extended to support this: rather than placing file data directly in the leaf nodes of the tree, the leaf node must reference the file inode.
+Thus, references to the same file inode can appear in multiple leaf nodes of the tree.
+Symlinks are also easy to support, since they are just leaf nodes containing a path (not a reference to an inode).
+
+\subsubsection{Log truncation}\label{sec:log-truncation}
+
+The algorithm as specified in Figure~\ref{fig:code} retains operations in the log indefinitely, so the memory use grows without bound.
 However, in practice it is easy to truncate the log, because \isa{apply\_op} only examines the prefix of operations whose timestamp is greater than that of the operation being applied.
@@
 In this case, each replica can keep track of the most recent timestamp it has seen from all replicas (including its own), and the minimum of these timestamps is the causally stable threshold.
+
+\subsubsection{Ordering of sibling nodes}\label{sec:ordering}
 
@@
 
-\noindent Let \isa{t1 = move\_time oper1} and \isa{t2 = move\_time oper2} be the timestamps of the two operations \isa{oper1} and \isa{oper2}.
+\noindent Here the \isa{\isacharat} operator concatenates two lists.
+Let \isa{t1 = move\_time oper1} and \isa{t2 = move\_time oper2} be the timestamps of the two operations \isa{oper1} and \isa{oper2}.
 Assume \isa{t1 < t2} without loss of generality.
@@
 
-TODO: Coda~\cite{Kumar:1995wf}, Ficus~\cite{Reiher:1994wh}, Rumor~\cite{Guy:1999gy}, Roam~\cite{Ratner:1999fh}
+TODO: Coda~\cite{Kumar:1993ey}, Ficus~\cite{Reiher:1994wh}, Rumor~\cite{Guy:1999gy}, Roam~\cite{Ratner:1999fh}
 
@@
 TODO
+
+% Disconnected operation
 

### 2019-06-19 paper/move-op.tex
Paper edits from Alastair's feedback

@@
 All replicas pick the same operation as winner, and effectively ignore the conflicting operation.
+(The \emph{timestamp} in this context need not come from a physical clock; it could also be logical, such as a Lamport timestamp~\cite{Lamport:1978jq}.)
 
@@
 
+\subsubsection{File merging}\label{sec:files}
+
+In a distributed filesystem, replication and conflict resolution is required not only for the directory structure, but also for the contents of individual files.
+This can be accomplished by using CRDTs for file contents as well.
+For example, we have explored CRDT algorithms for applications using a JSON data model~\cite{Kleppmann:2016ve} and implemented them in a library called Automerge~\cite{Automerge}.
+We discuss distributed filesystems further in Section~\ref{sec:filesystems}.
+
 % - No Synchronisation or consensus required. Available in the sense of CAP
@@
 
-\subsection{Distributed filesystems}
+\subsection{Distributed filesystems}\label{sec:filesystems}
 

### 2019-06-19 evaluation/src/main/scala/TestClient.scala
evaluation: first version of a replica

@@ -21,7 +21,7 @@ object TestClient {
 
     ConsoleReporter.forRegistry(metrics).build().start(20, SECONDS)
 
-    val config = GrpcClientSettings.connectToServiceAt("52.37.96.68", 8080)
+    val config = GrpcClientSettings.connectToServiceAt("18.138.235.108", 8080)
       .withDeadline(10 seconds)
       .withTls(false)
     val client: MoveService = MoveServiceClient(config)

### 2019-06-19 evaluation/src/main/scala/TestReplica.scala

@@ -1,0 +1,119 @@
+import akka.actor.{ ActorRef, ActorSystem }
+import akka.grpc.GrpcClientSettings
+import akka.http.scaladsl.UseHttp2.Always
+import akka.http.scaladsl.model.{ HttpRequest, HttpResponse }
+import akka.http.scaladsl.{ Http, HttpConnectionContext }
+import akka.pattern.ask
+import akka.stream.{ ActorMaterializer, Materializer }
+import akka.util.Timeout
+import com.codahale.metrics.{ ConsoleReporter, MetricRegistry, Meter, Timer }
+import com.typesafe.config.ConfigFactory
+import examplerpc.{ MoveService, MoveServiceClient, MoveServiceHandler }
+import java.util.concurrent.TimeUnit.SECONDS
+import java.util.Random
+import scala.concurrent.{ ExecutionContext, Future }
+import scala.concurrent.duration._
+import scala.util.{ Failure, Success }
+
+object TestReplica {
+
+  // TCP port number for gRPC
+  val PORT = 8080
+
+  // Time interval between generated operations (= 1 / operation rate)
+  val OPERATION_INTERVAL = 1000 millis
+
+  def main(args: Array[String]): Unit = {
+    if (args.length < 2) {
+      throw new Exception("Usage: TestReplica own-ip remote-ip1 [remote-ip2 ...]")
+    }
+
+    val conf = ConfigFactory.parseString("akka.http.server.preview.enable-http2 = on")
+      .withFallback(ConfigFactory.defaultApplication())
+
+    implicit val metrics = new MetricRegistry()
+    implicit val sys: ActorSystem = ActorSystem("TestReplica", conf)
+    implicit val mat: Materializer = ActorMaterializer()
+    implicit val ec: ExecutionContext = sys.dispatcher
+
+    ConsoleReporter.forRegistry(metrics).build().start(20, SECONDS)
+
+    val treeActor = sys.actorOf(TreeActor.props(args(0), metrics), "treeActor")
+    val service: HttpRequest => Future[HttpResponse] = MoveServiceHandler(new ReplicaService(treeActor, sys, metrics))
+
+    val binding = Http().bindAndHandleAsync(service, interface = "0.0.0.0", port = PORT,
+      connectionContext = HttpConnectionContext(http2 = Always))
+
+    binding.foreach { binding =>
+      println(s"gRPC server bound to: ${binding.localAddress}")
+    }
+
+    val generator = new LoadGenerator(treeActor, args.drop(1))
[…]

### 2019-06-19 evaluation/src/main/scala/TestServer.scala

@@ -47,7 +47,7 @@ class TestServer(system: ActorSystem, metrics: MetricRegistry) {
 
 class TestService(system: ActorSystem, metrics: MetricRegistry) extends examplerpc.MoveService {
   val requests  = metrics.timer("TestService.requests")
-  val treeActor = system.actorOf(TreeActor.props("a"), "treeActor")
+  val treeActor = system.actorOf(TreeActor.props("a", metrics), "treeActor")
 
   override def sendMove(move: examplerpc.Move): Future[examplerpc.LamportTS] = {
     //println(s"received: ${move}")

### 2019-06-19 evaluation/src/main/scala/TreeActor.scala

@@ -1,29 +1,62 @@
 import akka.actor.{ Actor, Props }
 import akka.event.Logging
+import com.codahale.metrics.MetricRegistry
 
 object TreeActor {
   case class RequestMove(parent: String, metadata: String, child: String)
 
-  def props(actorId: String) = Props(new TreeActor(actorId))
+  def props(actorId: String, metrics: MetricRegistry) = Props(new TreeActor(actorId, metrics))
 }
 
-class TreeActor(actorId: String) extends Actor {
+class TreeActor(actorId: String, metrics: MetricRegistry) extends Actor {
   val log = Logging(context.system, this)
+  val localTimer  = metrics.timer("TreeActor.local")
+  val remoteTimer = metrics.timer("TreeActor.remote")
 
-  def int(value: Long): generated.int = generated.int_of_integer(BigInt(value))
+  var counter: Long = 0
 
   var state: (List[generated.log_op[(generated.int, String),String,String]], generated.hashmap[String,(String, String)])
     = (Nil, generated.hm_empty[String, (String, String)].apply(()))
 
+  def int(value: Long): generated.int = generated.int_of_integer(BigInt(value))
+
+  def applyOp(op: examplerpc.Move) {
+    val opCounter = op.timestamp.get.counter
+    val timestamp = (int(opCounter), op.timestamp.get.replica)
+    val operation = generated.Move(timestamp, op.parent, op.metadata, op.child)
+    log.info("Applied %s".format(operation))
+    state = generated.example_apply_op(operation)(state)
+    if (opCounter > counter) counter = opCounter
+  }
+
+  def generateOp(parent: String, metadata: String, child: String): examplerpc.Move = {
+    val timestamp = examplerpc.LamportTS(counter + 1, actorId)
+    val operation = examplerpc.Move(Some(timestamp), parent, metadata, child)
+    applyOp(operation)
+    operation
+  }
+
   def receive = {
-    case examplerpc.Move(timestamp, parent, metadata, child) => {
-      val ts = (int(timestamp.get.counter), timestamp.get.replica)
-      val operation = generated.Move(ts, parent, metadata, child)
-      state = generated.example_apply_op(operation)(state)
-      //log.info("Applied %s".format(operation))
-      sender() ! timestamp.get
+    // Remote request
+    case op: examplerpc.Move => {
+      val timer = remoteTimer.time()
+      try {
+        applyOp(op)
+        sender() ! op.timestamp.get
+      } finally {
[…]

### 2019-06-19 evaluation/src/main/scala/TestReplica.scala
evaluation: tweaks for running experiments

@@ -50,7 +50,10 @@ object TestReplica {
 
     val generator = new LoadGenerator(treeActor, args.drop(1))
     generator.run()
-    // ActorSystem threads will keep the app alive until `sys.terminate()` is called
+
+    sys.scheduler.scheduleOnce(10 minutes) {
+      sys.terminate()
+    }
   }
 }
 
@@ -64,7 +67,7 @@ class LoadGenerator(treeActor: ActorRef, remoteReplicas: Array[String])
     val requests = metrics.timer(s"LoadGenerator($remoteIp).requests")
     val errors   = metrics.meter(s"LoadGenerator($remoteIp).errors")
     val config   = GrpcClientSettings.connectToServiceAt(remoteIp, TestReplica.PORT)
-      .withDeadline(10 seconds)
+      .withDeadline(20 seconds)
       .withTls(false)
     sendRequest(remoteIp, MoveServiceClient(config), requests, errors)(_)
   }
@@ -103,17 +106,9 @@ class LoadGenerator(treeActor: ActorRef, remoteReplicas: Array[String])
 }
 
 class ReplicaService(treeActor: ActorRef, sys: ActorSystem, metrics: MetricRegistry) extends examplerpc.MoveService {
-  val requests = metrics.timer("TestReplica.requests")
-
   override def sendMove(move: examplerpc.Move): Future[examplerpc.LamportTS] = {
     implicit val timeout: Timeout = 3 seconds
     import sys.dispatcher
-
-    val timer = requests.time()
-    try {
-      (treeActor ? move).mapTo[examplerpc.LamportTS]
-    } finally {
-      timer.stop()
-    }
+    (treeActor ? move).mapTo[examplerpc.LamportTS]
   }
 }

### 2019-06-19 evaluation/src/main/scala/TreeActor.scala

@@ -24,7 +24,7 @@ class TreeActor(actorId: String, metrics: MetricRegistry) extends Actor {
     val opCounter = op.timestamp.get.counter
     val timestamp = (int(opCounter), op.timestamp.get.replica)
     val operation = generated.Move(timestamp, op.parent, op.metadata, op.child)
-    log.info("Applied %s".format(operation))
+    //log.info("Applied %s".format(operation))
     state = generated.example_apply_op(operation)(state)
     if (opCounter > counter) counter = opCounter
   }

### 2019-06-19 paper/move-op.tex
Updates to related work.

--- 2019-06-18 paper/move-op.tex
@@
 Definitions making essential use of choice principles (e.g.~our \isa{get\_parent}) are obvious examples of definitions that will cause problems.
-However, \emph{inductively defined} relations characterised by a series of introduction rules, such as our \isa{ancestor} relation, are also problematic as their execution necessarily requires Prolog-like proof search.
+However, \emph{inductively defined} relations characterised by a series of introduction rules, such as our \isa{ancestor} relation, are also problematic for the code generator as making them executable necessarily requires Prolog-like proof search and mode inference.
 \item
-Na\"{i}ve use often leads to inefficient target language code, especially true when using HOL's \isa{set} type, a type of potentially infinite sets.
-Whilst convenient for theorem proving this type often leads to inefficiency in generated code.
+Na\"{i}ve use of the code generator often leads to inefficient target language code.
+This is especially true when using HOL's \isa{set} type, a type of potentially infinite mathematical sets.
+Whilst convenient for theorem proving purposes, this type often leads to gross inefficiency in the extracted code.
 \end{enumerate}
 
-We therefore produce a series of definitions that are optimised for execution rather than theorem proving, before trying to evaluate our algorithms.
-Notably, we eschew the HOL \isa{set} type, and instead use a more efficient hash-map implementation provided by the Isabelle Collections Framework~\cite{DBLP:conf/itp/LammichL10}, a library of verified, efficient container datatypes and associated algorithms.
-
+We must therefore first produce a series of refined definitions that are optimised for efficient execution rather than ease of theorem proving, before trying to evaluate our algorithms.
+Within these definitions, we eschew the HOL \isa{set} type, and instead use a more efficient hash-map implementation provided by the Isabelle Collections Framework~\cite{DBLP:conf/itp/LammichL10}, a library of verified, efficient container datatypes and associated algorithms.
 Write $('a::\{hashable\}, 'b)\ \isa{hm}$ for the type of hash-maps with key type $\isa{'a}$ and binding type $\isa{'b}$.
@@
 
-We use a \emph{simulation relation} between our two representations of trees, and say that a hash-map $\isa{t}$ of type $(\isa{'n}, \isa{'m} \times \isa{'n})\ \isa{hm}$ \emph{simulates} a set $T$ of type $(\isa{'n} \times \isa{'m} \times \isa{'n})\ \isa{set}$ whenever
+We introduce an explicit \emph{simulation relation} between our two representations of trees.
+Specifically, we say that a hash-map $\isa{t}$ of type $(\isa{'n}, \isa{'m} \times \isa{'n})\ \isa{hm}$ \emph{simulates} a set $T$ of type $(\isa{'n} \times \isa{'m} \times \isa{'n})\ \isa{set}$ whenever
 \begin{displaymath}
@@
 where $hm.lookup$ searches for a binding for key $\isa{c}$ in hash-map $t$, returning $\isa{Some}\ (m, p)$ when a binding from $c$ to $(m, p)$ exists in the hash-map, or $\isa{None}$ otherwise.
-Write $t \preceq T$ to assert that $t$ and $T$ are contained within the simulation relation.
-Note some corollaries follow immediately: e,g,~we have $t \preceq T$ implies $\isa{unique\_parent}\ \isa{T}$.
+We write $t \preceq T$ to assert that the hash-map $t$ and set $T$ are contained within the simulation relation, defined above.
+Note some useful corollaries follow immediately from this definition: e,g,~we have $t \preceq T$ implies $\isa{unique\_parent}\ \isa{T}$.
 
-Our approach is to produce a new set of definitions---mirroring the structure of our existing ones---wherein we replace all occurrences of the $(\isa{'n} \times \isa{'m} \times \isa{'n})\ \isa{set}$ type, and any associated set operations, with the $(\isa{'n}, \isa{'m} \times \isa{'n})\ \isa{hm}$ type and associated hash-map operations.
-For clarity, we refer to any definition using hash-maps as an ``efficient'' definition, and any definition making use of the \isa{set} type as an ``abstract'' definition.
-For most definitions this is a mechanical process, and we elide further discussion of the required changes, referring the interested reader to our Isabelle/HOL source code.
-However \isa{ancestor} is different to other definitions as it is an inductively defined relation rather than a recursive function, and we need further work to make this efficiently executable.
-We characterise this relation with the following:
+Our general strategy henceforth is to produce a new set of definitions---mirroring the structure of our existing definitions---wherein we replace all occurrences of the $(\isa{'n} \times \isa{'m} \times \isa{'n})\ \isa{set}$ type, and any associated set operations, with the $(\isa{'n}, \isa{'m} \times \isa{'n})\ \isa{hm}$ type and associated hash-map operations.
+For the sake of clarity, we adopt the convention that any definition using hash-maps will be referred to as an ``efficient'' definition, and any definition making use of HOL's set type will be referred to as an ``abstract'' definition.
+For most definitions this is a mechanical process, and we elide further description here.
+However \isa{ancestor} is slightly different, as it is an inductively defined relation rather than a recursive function, and we therefore need further work to make this efficiently executable.
+We characterise this relation recursively with the following theorem:
 
@@
 
-Note that \isa{efficient\_ancestor\_simp} is tagged with a $\isa{code}$ ``attribute'', a hint that this theorem should be used as a transformation during code generation.
-Isabelle/HOL takes this theorem as the definition, for code generation, of the \isa{efficient\_ancestor} relation and is able to directly translate this to a recursive function definition in each target backend---this transformation is ``semantics preserving'', as the characterisation was proved as a theorem within HOL.
+Note that the theorem \isa{efficient\_ancestor\_simp} is tagged with a $\isa{code}$ ``attribute'', which is a hint that this theorem should be used as a transformation during the code generation process.
+Isabelle/HOL takes this theorem as the definition, for code generation, of the \isa{efficient\_ancestor} relation and is able to directly translate this to a recursive function definition in each target backend.
 
-Next, we demonstrate the correctness of our efficient definitions against our abstract definitions, and the properties that we prove are best described using an example, as they all follow a similar pattern.
-We use the lemma \isa{efficient\_do\_op\_simulates} as our exemplar:
+Next, we demonstrate the correctness of our new efficient definitions against our older, abstract definitions.
+The correctness properties that we prove are best described using an example, as they all follow a similar pattern.
+We therefore focus on the lemma \isa{efficient\_do\_op\_simulates} which states:
 
@@
 
-We show that the \isa{efficient\_do\_op} and the \isa{do\_op} functions are in ``lock-step'', mapping related trees $t \preceq T$ to related trees $u \preceq U$ and producing identical logs of operations whenever they are applied to the same input operation, \isa{oper}.
-This lock-step is repeated bottom-up for every efficient-abstract algorithm pairs, until we reach the final correctness theorem for \isa{apply\_ops}:
+Note that we show that the \isa{efficient\_do\_op} and the \isa{do\_op} functions maintain a ``lock-step'' property, mapping related trees $t \preceq T$ to related trees $u \preceq U$ and producing identical logs of operations whenever they are applied to the same input operation, \isa{oper}.
+This pattern of lock-step simulation is repeated bottom-up for every pair of efficient-abstract algorithm pairs, until we reach the final correctness theorem for \isa{apply\_ops}:
 
@@
 
-Intuitively, this states that if both algorithms are applied to the same set of operations then they produce identical logs and also produce trees that contain the same set of parent-metadata-child records---i.e.
+This property captures our intuitive understanding of what it means for \isa{efficient\_apply\_ops} to be correct with respect to \isa{apply\_ops}: if both algorithms are applied to the same set of operations then they produce identical logs and also produce trees that contain the same set of parent-metadata-child records---i.e.
 the two trees are extensionally equivalent, despite having very different concrete representations.
-Moreover, we have the following corollaries of the theorems \isa{apply\_ops\_commutes} and \isa{apply\_ops\_acyclic}, respectively, further establishing the correctness of our efficient algorithms:\footnote{The ``efficient'' analogue of \isa{apply\_ops\_unique\_parent} is trivially true, as the output tree is stored as a hash-map, and therefore too trivial to state as a theorem.}:
+We can further check that this theorem correctly captures the correctness of our \isa{apply\_ops} definition by recasting the various tree invariant and convergence theorems in terms of \isa{efficient\_apply\_ops}.
+For example, we have the following corollaries of the theorems \isa{apply\_ops\_commutes} and \isa{apply\_ops\_acyclic}, respectively\footnote{The ``efficient'' analogue of \isa{apply\_ops\_unique\_parent} is trivially true, as the output tree is stored as a hash-map, and therefore too trivial to state as a theorem.}:
 
@@
 
-Note the statement of \isa{efficient\_apply\_ops\_commutes} differs from that of \isa{apply\_ops\_commutes}---the latter asserts identical sets are obtained from convergence, whilst the former asserts that reading any key binding in $t$ gives the same result as a reading in $u$.
-Whilst this statement appears weaker than that of \isa{apply\_ops\_acyclic}, it is correct, and the strongest statement one can hope for: the resulting hash-maps may contain the same key-value bindings but may not be structurally equal, depending on the ordering of operations.
+Note the statement of \isa{efficient\_apply\_ops\_commutes} differs in an essential way from that of \isa{apply\_ops\_commutes}---the latter asserts that identical sets are obtained from convergence, whilst the former asserts that simply reading any key binding in $t$ gives the same result as reading in $u$.
+Whilst this statement appears weaker than the statement of \isa{apply\_ops\_acyclic}, it is correct, as the resulting hash-maps may not have an identical representation in-memory---depending on the particular ordering of key-value insertions---despite containing the same key-value bindings.
 
@@
 
-TODO: Coda~\cite{Kumar:1995wf}, Ficus~\cite{Reiher:1994wh}, Rumor~\cite{Guy:1999gy}, Roam~\cite{Ratner:1999fh}
+Unison is a file synchronisation tool with a formal specification~\cite{PierceVouillon:UnisonSpecTR} which allows two replicas, such as laptops, desktops or servers, to synchronise the state of a directory.
+Unison permits offline updates to both replicas. 
+The (possibly divergent) state of two replicas are resynchronised by running the Unison tool on the two replicas. 
+Unison works by recording a summary record of the state of the filesystem on each replica,  inspecting this summary whenever the unison program is invoked, and sharing any update information with the other replica in order to determine the changes made to the filesystems on each replica since they were last synchronised.
+This approach has the benefit that no special kernel extension or daemon process is required; instead Unison is an application program which inspects the current state of the filesystem whenever it is run.
+When presented with the move operations described in Figure~\ref{fig:move-same-item}, Unison duplicates the files, resulting in the outcome shown in Figure~\ref{fig:move-same-item}~(a). 
+Unison is unable to automatically synchronise the move operations shown in Figure~\ref{fig:move-cycle} and instead asks the user to confirm whether changes made to directories A and B should be propagated or deleted.
+As a result, depending on user choice, both replicas can end up in one of four states, including those shown in Figure~\ref{fig:move-cycle}~(b), (c) or (d) or with an empty root directory on both replicas.
+In all cases, Unison successfully ensures both replicas are in the same state after synchronisation.
+
+Coda is a distributed file system which allows clients to cache or \emph{hoard} copies of files otherwise stored in a shared data repository~\cite{kistler1992coda}. 
+Clients can edit copies of data held in the cache while offline, during which time a kernel module keeps track of all updates. 
+When the client comes back online, the client attempts to resynchronise changes with the repository. 
+This design means that conflicts may arise if multiple clients edit the same files or directories while one or more clients are offline. 
+One proposed solution to this is to support application-specific resolvers which "encapsulate the knowledge needed for file resolution"~\cite{Kumar:1995wf}, an approach is similar to that found in Bayou~\cite{Terry:1995dn}.
+
+Ficus~\cite{Reiher:1994wh}, Rumor~\cite{Guy:1999gy}, Roam~\cite{Ratner:1999fh}
 

### 2019-06-19 paper/move-op.tex

--- 2019-06-19 paper/move-op.tex
@@
 
-TODO: Coda~\cite{Kumar:1993ey}, Ficus~\cite{Reiher:1994wh}, Rumor~\cite{Guy:1999gy}, Roam~\cite{Ratner:1999fh}
+Unison is a file synchronisation tool with a formal specification~\cite{PierceVouillon:UnisonSpecTR} which allows two replicas, such as laptops, desktops or servers, to synchronise the state of a directory.
+Unison permits offline updates to both replicas. 
+The (possibly divergent) state of two replicas are resynchronised by running the Unison tool on the two replicas. 
+Unison works by recording a summary record of the state of the filesystem on each replica,  inspecting this summary whenever the unison program is invoked, and sharing any update information with the other replica in order to determine the changes made to the filesystems on each replica since they were last synchronised.
+This approach has the benefit that no special kernel extension or daemon process is required; instead Unison is an application program which inspects the current state of the filesystem whenever it is run.
+When presented with the move operations described in Figure~\ref{fig:move-same-item}, Unison duplicates the files, resulting in the outcome shown in Figure~\ref{fig:move-same-item}~(a). 
+Unison is unable to automatically synchronise the move operations shown in Figure~\ref{fig:move-cycle} and instead asks the user to confirm whether changes made to directories A and B should be propagated or deleted.
+As a result, depending on user choice, both replicas can end up in one of four states, including those shown in Figure~\ref{fig:move-cycle}~(b), (c) or (d) or with an empty root directory on both replicas.
+In all cases, Unison successfully ensures both replicas are in the same state after synchronisation.
+
+Coda is a distributed file system which allows clients to cache or \emph{hoard} copies of files otherwise stored in a shared data repository~\cite{kistler1992coda}. 
+Clients can edit copies of data held in the cache while offline, during which time a kernel module keeps track of all updates. 
+When the client comes back online, the client attempts to resynchronise changes with the repository. 
+This design means that conflicts may arise if multiple clients edit the same files or directories while one or more clients are offline. 
+One proposed solution to this is to support application-specific resolvers which "encapsulate the knowledge needed for file resolution"~\cite{Kumar:1995wf}, an approach is similar to that found in Bayou~\cite{Terry:1995dn}.
+
+Ficus~\cite{Reiher:1994wh}, Rumor~\cite{Guy:1999gy}, Roam~\cite{Ratner:1999fh}
 

### 2019-06-19 evaluation/src/main/scala/Main.scala
evaluation: profiling of remote operation application
Result: 88% of the time is spent in hm_update; 85% spent in
rbt_insert_with_key (rebalancing of red-black tree?). Doesn't look like
it will be easy to speed this up.

Only 8% of the time is spent in hm_lookup, in contrast to applying local
operations, for which hm_lookup is the dominant cost (68%). But since
local operations are vastly faster anyway, hm_lookup doesn't really
matter.

@@ -1,19 +1,32 @@
 import java.util.Random
 
 object Main extends App {
+  val random = new Random()
+
   def int(value: Int): generated.int = generated.int_of_integer(BigInt(value))
+
+  def operations(num: Int, actorId: String) = {
+    (0 to num).map { i =>
+      val parent = "%d".format(random.nextInt(1000))
+      val child  = "%d".format(random.nextInt(1000))
+      generated.Move((int(i), actorId), parent, "", child)
+    }
+  }
 
   var state: (List[generated.log_op[(generated.int, String),String,String]], generated.hashmap[String,(String, String)])
     = (Nil, generated.hm_empty[String, (String, String)].apply(()))
-  val random = new Random()
-  val startTime = System.nanoTime()
 
-  for (i <- 1 to 1000) {
-    val parent = "%d".format(random.nextInt(1000))
-    val child  = "%d".format(random.nextInt(1000))
-    val operation = generated.Move((int(i), "a"), parent, "", child)
-    state = generated.example_apply_op(operation)(state)
+  println("applying local ops...")
+  val start1 = System.nanoTime()
+  for (op <- operations(1000, "a")) {
+    state = generated.example_apply_op(op)(state)
   }
+  println("elapsed time: %d ms" format ((System.nanoTime() - start1) / 1000000))
 
-  println("elapsed time: %d ms" format ((System.nanoTime() - startTime) / 1000000))
+  println("applying remote ops...")
+  val start2 = System.nanoTime()
+  for (op <- operations(1000, "b")) {
+    state = generated.example_apply_op(op)(state)
+  }
+  println("elapsed time: %d ms" format ((System.nanoTime() - start2) / 1000000))
 }

### 2019-06-20 paper/move-op.tex
evaluation: add experiment data
paper: add discussion of experiments

@@
 
+\begin{figure}
+  \includegraphics[width=\columnwidth,keepaspectratio=true]{plots.png}
+  \caption{Median execution time for applying an operation to the replica state, on an Amazon EC2 t2.small instance, at varying operation throughput rates.
+Error bars indicate the minimum and 95th percentile execution times.}
+  \label{fig:plots}
+  \Description{Plot in which the x axis goes from 0 to 30 move operations per second, and the y axis goes from 0 to 12 milliseconds.
+Local operations take a near-constant time less than 1ms; the time for remote operations gradually increases to 2ms until 24 operations/sec, at which point it suddenly shoots up off the top of the scale.}
+\end{figure}
+
+\begin{table}
+  \caption{Median response times of a RPC request between replicas in different regions.}
+  \label{tab:rpc-times}
+  \begin{tabular}{r|rrr}
+    \toprule
+               & California & Ireland & Singapore \\
+    \midrule
+    California &            & 395 ms  & 491 ms    \\
+    Ireland    & 395 ms     &         & 547 ms    \\
+    Singapore  & 491 ms     & 547 ms  &           \\
+    \bottomrule
+\end{tabular}
+\end{table}
+
 \subsection{Performance experiments}
 
-TODO\dots
+\subsubsection{Experimental setup and results}
+
+To evaluate the viability of our algorithm in practice, we used Isabelle/HOL's code generation facility to extract executable Scala code from our formally verified definitions.
+We wrapped this implementation in a network service using the Akka~\cite{Akka} actor framework, using gRPC~\cite{gRPC} for network communication.
+We deployed three replicas of this service on Amazon EC2 t2.small instances in different regions worldwide: one in Northern California (region us-west-1), one in Ireland (eu-west-1), and one in Singapore (ap-southeast-1).
+The network delays for communication between these regions are substantial; see Table~\ref{tab:rpc-times}.
+
+We perform a synthetic workload in which each replica generates move operations at the same rate; for each move operation the generating replica chooses parent and child nodes uniformly at random from a set of 1,000 tree nodes.
+We use Lamport timestamps~\cite{Lamport:1978jq} as operation timestamps, and string values to identify tree nodes.
+When a replica generates an operation, it immediately applies that operation to the local replica state, and then asynchronously sends the operation to the other two replicas via gRPC requests.
+When a replica receives an operation from a remote replica, it also applies that operation to its own replica state.
+Thus, of the operations applied by each replica, approximately one third are locally generated, and two thirds are received from the other two replicas.
+
+We varied the rate at which replicas generated operations, and measured the execution time of our algorithm for applying each operation.
+Figure~\ref{fig:plots} shows the median execution times for local and remote operations respectively.
+The execution time for local operations is near-constant (median times mostly between 0.2~ms and 0.6~ms), while the execution time for remote operations increases approximately linearly from 0.2~ms to 2.3~ms over the range of 3 to 24 operations/sec.
+Above 25 operations/sec the execution time for remote operations suddenly increases to over 30~ms (off the scale of Figure~\ref{fig:plots}).
+We do not plot the exact values for execution times above 12~ms because they are less meaningful, as explained below.
+
+\subsubsection{Interpretation of results}
+
+A locally generated operation always has a timestamp that is higher than any existing operation at the generating replica at that time (by the definition of Lamport timestamps), and thus applying that operation requires only a \isa{do\_op}, and no \isa{undo\_op} or \isa{redo\_op}, explaining the near-constant execution time.
+On the other hand, for remote operations, some number of undos and redos needs to be performed, depending on the degree of concurrency.
+As the rate of operations increases, the time between successive operations becomes smaller than the network delay between replicas; thus, there are more operations ``in flight'' at the same time, and more calls to \isa{undo\_op} and \isa{redo\_op} are required to put operations into timestamp order.
+
+The time taken to apply a remote operation is proportional to the number of undos and redos required.
+Above 25 operations/sec the CPU utilisation of our virtual machines grows to the point that incoming network requests from other replicas start queueing before they are processed.
+When a remote operation spends time queueing, more concurrent operations have been applied by the time it is processed, and thus more undos and redos are required to apply the incoming operation, which in turn increases the CPU utilization and further increases queueing.
+This self-reinforcing effect explains the sudden dramatic increase in Figure~\ref{fig:plots}.
+
+Above 25 operations/sec the queueing delay increases so much that some RPC requests time out, and thus not all operations are received by all replicas.
+As the remote operation execution times above this threshold depend more on the network queueing properties than our algorithm, we discard these datapoints.
+
+\subsubsection{Practical relevance of results}
+
+The Scala implementation generated from our Isabelle definitions is optimised for correctness, not for performance; we expect that a more optimised implementation could achieve much higher throughput than our 24 operations/sec.
+Nevertheless, we believe that our evaluation sends a positive signal about the practical applicability of our algorithm.
+
+Note that when a user is interacting with a system, they only have to wait for local operations to execute; any remote operations can be applied in the background without affecting user interaction.
+Since our algorithm need not perform any undos or redos for local operations, these user interactions are consistently fast.
+
+The alternative to our optimistic replication approach would be to use a central server (master replica) or consensus protocol that processes all operations in a serialisable order.
+However, performing an operation on such a system requires waiting for synchronous communication with a server potentially located on a different continent, taking hundreds of milliseconds (Table~\ref{tab:rpc-times}); by contrast, a local operation taking a fraction of a millisecond is orders of magnitude faster.
+Moreover, a local operation requires no waiting for network communication, and thus it supports disconnected operation: mobile devices can read and modify the replicated tree even while disconnected from the Internet, which is not the case with centralised systems.
 

### 2019-06-20 paper/move-op.tex
Related work updates.

@@
 
-Coda is a distributed file system which allows clients to cache or \emph{hoard} copies of files otherwise stored in a shared data repository~\cite{kistler1992coda}. 
+Coda is a distributed client-server file system which allows clients to cache or \emph{hoard} copies of files otherwise stored in a shared data repository~\cite{kistler1992coda}. 
 Clients can edit copies of data held in the cache while offline, during which time a kernel module keeps track of all updates. 
@@
 This design means that conflicts may arise if multiple clients edit the same files or directories while one or more clients are offline. 
-One proposed solution to this is to support application-specific resolvers which "encapsulate the knowledge needed for file resolution"~\cite{Kumar:1995wf}, an approach is similar to that found in Bayou~\cite{Terry:1995dn}.
-
-Ficus~\cite{Reiher:1994wh}, Rumor~\cite{Guy:1999gy}, Roam~\cite{Ratner:1999fh}
+One proposed solution to resolving conflicts in files is to support application-specific resolvers which ``encapsulate the knowledge needed for file resolution''~\cite{Kumar:1995wf}, an approach is similar to that found in Bayou~\cite{Terry:1995dn}.
+Concurrent edits on directory entries across different replicas in Coda have been considered, but the authors note that they do ``not address transparent resolution of cross-directory [move operations] in [their] current implementation.''~\cite{kumar1993log}.
+Furthermore, while the authors consider a number of conflicts associated with directory move operations, they do not highlight the potential for the creation of cycles as seen in Figure~\ref{fig:move-cycle} (a).
+
+The Ficus replicated file system~\cite{Reiher:1994wh} is a in-kernel SunOS-based replicated peer-to-peer file system. 
+Ficus supported updates to replicas during periods of network partition and claims ``conflicting updates to directories are detected an automatically repaired''~\cite{guy1990implementation}. 
+Unfortunately we were unable to find a precise definition of the algorithm used in any of the available publications. 
+Rumor~\cite{Guy:1999gy} is the successor to Ficus. 
+An important distinction is between Rumor and previous work is that it operates at the application layer and needs to be invoked periodically by the user or a deamon process.
+Rumor is therefore very similar in operation and design to Unison described earlier.
+
+%The user manual (in Archive) is quite helpful for Rumor: https://web.archive.org/web/20170705140958/ftp://ftp.cs.ucla.edu/pub/rumor/rumor_users_manual.ps.
+Note that this also makes the claim that "Unix directories are an important example of a kind of file for which Rumor automatically resolves conflicts"
+
+TODO: Roam~\cite{Ratner:1999fh}
 

### 2019-06-20 paper/move-op.tex
paper: editing, removing lines with a few words
Some small changes to experiment write-up and code-gen sections.

@@
 
-\subsection{Extracting efficient algorithms from HOL}
+\subsection{Extracting efficient algorithms from HOL} \label{subsect.extracting}
 
@@
 
-Write $('a::\{hashable\}, 'b)\ \isa{hm}$ for the type of hash-maps with key type $\isa{'a}$ and binding type $\isa{'b}$.
-Note that keys have an explicit $\isa{hashable}$ constraint, meaning that this hash-map type may only be used with kay types with an associated hashing function.
+Write $('a::\{hashable\}, 'b)\ \isa{hm}$ for a hash-map type with key type $\isa{'a}$ and binding type $\isa{'b}$.
+Note that keys have a $\isa{hashable}$ constraint, meaning that the type may only be used with key types with a defined hashing function.
 With this type, we represent trees by the type $(\isa{'n}::\{hashable\}, \isa{'m} \times \isa{'n})\ \isa{hm}$, mapping from a \emph{child} node to its \emph{parent} node and associated metadata, if any such mapping exists.
@@
 
-Intuitively, this states that if both algorithms are applied to the same set of operations then they produce identical logs and also produce trees that contain the same set of parent-metadata-child records---i.e.
+That is, if both algorithms are applied to the same set of operations then they produce identical logs and also produce trees that contain the same set of key-value bindings---i.e.
 the two trees are extensionally equivalent, despite having very different concrete representations.
@@
 
-To evaluate the viability of our algorithm in practice, we used Isabelle/HOL's code generation facility to extract executable Scala code from our formally verified definitions.
-We wrapped this implementation in a network service using the Akka~\cite{Akka} actor framework, using gRPC~\cite{gRPC} for network communication.
-We deployed three replicas of this service on Amazon EC2 t2.small instances in different regions worldwide: one in Northern California (region us-west-1), one in Ireland (eu-west-1), and one in Singapore (ap-southeast-1).
+We used Isabelle/HOL's code generation mechanism to extract executable Scala code from our HOL definitions, as discussed in Subsection~\ref{subsect.extracting}, and wrapped this implementation in a network service using the Akka~\cite{Akka} actor framework, using gRPC~\cite{gRPC} for network communication.
+We deployed three replicas of this service on Amazon EC2 \texttt{t2.small} instances in different regions worldwide: one in Northern California (region \texttt{us-west-1}), one in Ireland (\texttt{eu-west-1}), and one in Singapore (\texttt{ap-southeast-1}).
 The network delays for communication between these regions are substantial; see Table~\ref{tab:rpc-times}.

### 2019-06-20 paper/move-op.tex
paper: more editing throughout
Editing throughout the paper.

@@
 
-For many systems and applications, the data model that describes the system state is a tree.
+Many systems use a tree-structured data model.
 For example:
@@
         Multiple objects and groups may in turn be combined into higher-level groups, forming a tree.
-    \item Personal \textbf{note-taking} and \textbf{task-management tools} such as Org-mode for Emacs~\cite{OrgMode} or OmniOutliner~\cite{OmniOutliner} explicitly present the user with a tree structure that they can inspect and manipulate.
+    \item \textbf{Note-taking} and \textbf{task-management tools} such as Org-mode for Emacs~\cite{OrgMode} or OmniOutliner~\cite{OmniOutliner} present the user with a tree structure that they can inspect and manipulate.
 \end{itemize}
@@
 We explain in Section~\ref{sec:move-is-hard} why this move operation is so challenging.
-
-Our contributions in this work are as follows:
+In summary, our contributions are:
 \begin{itemize}
@@
 
-The first difficulty arises when a node is moved to a new location on one replica, and concurrently the same node is moved to a different location on a different replica.
+The first difficulty arises when a node is concurrently moved into different locations on different replicas.
 This scenario is illustrated in Figure~\ref{fig:move-same-item}, where replica 1 moves node $A$ to be a child of $B$, while concurrently replica 2 moves $A$ to be a child of $C$.
@@
 
-In an unreplicated tree it is easy to prevent cycles being introduced: if the node being moved is an ancestor of the destination node, then the move operation must be rejected.
+In an unreplicated tree it is easy to prevent cycles being introduced: if the node being moved is an ancestor of the destination node, then the operation is rejected.
 However, a new problem arises in a replicated setting: different replicas may perform operations that are individually safe, but whose combination leads to a cycle.
@@
 In this section we walk through the code in Figure~\ref{fig:code} step by step, explaining features of the Isabelle/HOL syntax as we encounter them.
-For additional background on the language, standard documentation is available~\cite{DBLP:books/sp/NipkowK14}.
+Additional background documentation is available~\cite{DBLP:books/sp/NipkowK14}.
 
@@
 A move operation is a record consisting of four fields: a timestamp \isa{t} of type \isa{'t}, a parent node \isa{p} of type \isa{'n}, a metadata field \isa{m} of type \isa{'m}, and a child node \isa{c} of type \isa{'n}.
-Here, \isa{'t}, \isa{'n} and \isa{'m} are \emph{type variables} that can be \emph{substituted} for arbitrary types; our only requirement is that timestamps \isa{'t} are globally unique and have an associated total order.
-Concretely, Lamport timestamps~\cite{Lamport:1978jq} would meet these requirements.
+Here, \isa{'t}, \isa{'n} and \isa{'m} are \emph{type variables} that can be \emph{substituted} for arbitrary types; our only requirement is that timestamps \isa{'t} are globally unique and have an associated total order (e.g.
+Lamport timestamps~\cite{Lamport:1978jq} would suffice).
 
@@
 For this reason we do not need a separate operation type to create tree nodes: simply performing a move with a fresh child node \isa{c} is sufficient to create \isa{c}.
-Moreover, we do not define a remove operation: instead, we can designate a particular tree node as the ``trash'', and perform deletion of a node \isa{c} by moving \isa{c} to be a child of the trash node.
+Nor do we define a remove operation: instead, we can designate a particular tree node as the ``trash'', and perform deletion of a node \isa{c} by moving \isa{c} to be a child of the trash node.
 
@@
 Otherwise, \isa{oldp} records the previous parent and metadata of \isa{c}, that is: if there exists a \isa{p'} and \isa{m'} such that \isa{(p', m', c) {\isasymin} tree}, then \isa{oldp} is set to \isa{Some (p', m')}.
-This logic is performed by the \isa{get\_parent} function (lines 16--20).
+The \isa{get\_parent} function implements this (lines 16--20).
 
@@
 Otherwise we perform the operation using \isa{do\_op}, and add the corresponding \isa{LogMove} record as the head of the log (line 51).
-
 This logic ensures that the log is maintained in descending timestamp order, with the greatest timestamp at the head.
@@
 We then prove that \isa{unique\_parent tree} holds, where the \isa{unique\_parent} predicate is defined on lines 56--57 of Figure~\ref{fig:code}: it states that whenever the tree contains a triple whose third element is the child node \isa{c}, then the first and second elements of the triple (the parent node and the metadata) are uniquely defined.
-
-Because we make no assumptions about \isa{ops}, this theorem holds for any replica state that can be reached by applying any number of operations.
+As we make no assumptions about \isa{ops}, this theorem holds for any replica state that can be reached by applying any number of operations.
 Proving \isa{apply\_ops\_unique\_parent} is easy because each of the \isa{do\_op}, \isa{undo\_op} and \isa{redo\_op} functions individually maintains the \isa{unique\_parent tree} invariant, and thus \isa{apply\_op} also maintains the invariant.
@@
 
-To prove that \isa{apply\_op} preserves the acyclicity invariant, we need to show that for any tree produced by \isa{apply\_ops}, there exists a sequence of operations such that applying those operations using \isa{do\_op} invocations alone (not using \isa{undo\_op} or \isa{redo\_op}) results in the same tree.
+To prove that \isa{apply\_op} preserves acyclicity, we need to show that for any tree produced by \isa{apply\_ops}, there exists a sequence of operations such that applying those operations using \isa{do\_op} invocations alone (not using \isa{undo\_op} or \isa{redo\_op}) results in the same tree.
 (This is the case when operations are applied in order of increasing timestamp.)

### 2019-06-20 paper/move-op.tex
paper: editing and condensing

--- 2019-06-20 paper/move-op.tex
@@
     \item To demonstrate the practical viability of our approach, we refine our abstract algorithm to an efficient implementation within Isabelle/HOL and prove the equivalence of the two.
-We thereafter extract a formally verified, executable Scala implementation from the Isabelle/HOL definitions, and perform tests on this implementation.
+        We then extract a formally verified, executable Scala implementation from the Isabelle/HOL definitions and evaluate its performance with replicas across three continents.
     \item We perform experiments with replicated filesystem products such as Dropbox and Google Drive, and show that they exhibit problems that would be prevented by our algorithm.
@@
         Moreover, renaming a file or directory is equivalent to moving it to a new name without changing its parent directory.
-    \item In a rich text editor, a paragraph node can be turned into a bullet point.
-        In the DOM tree this corresponds to creating a new bullet list node, creating a new bullet point node within it, and then moving the paragraph node to be a child of the new bullet point node.
+    \item In a rich text editor, a paragraph can be turned into a bullet point.
+        In the DOM tree this corresponds to creating a new bullet list node, creating a new bullet point node within it, and then moving the paragraph node inside the bullet point.
     \item In presentation software, grouping two graphical objects corresponds to creating a new group node, and then moving the two objects to be children of the new group node.
@@
 
-The first difficulty arises when a node is moved to a new location on one replica, and concurrently the same node is moved to a different location on a different replica.
+The first issue arises when a node is moved to a new location on one replica, and concurrently the same node is moved to a different location on another replica.
 This scenario is illustrated in Figure~\ref{fig:move-same-item}, where replica 1 moves node $A$ to be a child of $B$, while concurrently replica 2 moves $A$ to be a child of $C$.
-After the replicas communicate these operations to each other, what should the merged state of the tree be?
-
-If a move operation is implemented by deleting the moved subtree from its old location, and then re-creating it at the new location, the merged state will be as shown in Figure~\ref{fig:move-same-item}a: the concurrent moves will duplicate the moved subtree, since each concurrent move independently recreates the subtree in each destination location.
+After the replicas communicate, what should the merged state of the tree be?
+
+If a move operation is implemented by deleting the moved subtree from its old location, and then re-creating it at the new location, the merged state will be as shown in Figure~\ref{fig:move-same-item}a: the concurrent moves will duplicate the moved subtree, since each move independently recreates the subtree in each destination location.
 We believe that this duplication is undesirable, since subsequent edits to nodes in the duplicated subtree will apply to only one of the copies.
@@
 This error state persisted until the directories on the erroring replica were manually moved to match the state of the other replica.
-
-On the other hand, Dropbox exhibits the duplication behaviour shown in Figure~\ref{fig:move-cycle}b.
+On the other hand, Dropbox exhibited the duplication behaviour shown in Figure~\ref{fig:move-cycle}b.
 
@@
 propose two solutions to this problem: either to duplicate tree nodes, as in Figure~\ref{fig:move-cycle}b, or to execute a synchronous locking protocol that prevents two move operations from concurrently modifying the same part of the tree.
-The downside of a locking protocol is that the move operation is no longer highly available in the presence of network partitions, since it must wait for communication with other replicas.
+The downside of a locking protocol is that the move operation is no longer highly available in the presence of network partitions, since it must wait for synchronous communication with other replicas.
 
@@
 This middleware may deliver operations in any order.
-Operations received from a remote replica are applied in exactly the same way as operations generated by a local user.
+Operations received from a remote replica are applied using the same algorithm as operations generated by a local user.
 
 The key consistency property of our algorithm is \emph{convergence}: that is, whenever any two replicas have applied the same set of operations, then they must be in the same state---even if the operations were applied in a different order on different replicas.
-We prove this in Section~\ref{sec:proof} by showing that with our algorithm, applying operations is commutative.
+We prove this in Section~\ref{sec:proof} by showing that applying operations is commutative.
 
@@
 We choose this language because it combines the conciseness of pseudocode with the precision of mathematical notation.
-It supports formal reasoning, allowing us to prove the correctness of the algorithm (Section~\ref{sec:proof}), and it also allows executable code to be exported to common functional programming languages such as Scala, Haskell, OCaml, or Standard ML (Section~\ref{sec:evaluation}).
+It supports formal reasoning, allowing us to prove the correctness of the algorithm (Section~\ref{sec:proof}), and it also allows executable code to be exported to common functional programming languages such as Scala, Haskell, or OCaml.
 
@@
 A move operation is a record consisting of four fields: a timestamp \isa{t} of type \isa{'t}, a parent node \isa{p} of type \isa{'n}, a metadata field \isa{m} of type \isa{'m}, and a child node \isa{c} of type \isa{'n}.
-Here, \isa{'t}, \isa{'n} and \isa{'m} are \emph{type variables} that can be \emph{substituted} for arbitrary types; our only requirement is that timestamps \isa{'t} are globally unique and have an associated total order.
-Concretely, Lamport timestamps~\cite{Lamport:1978jq} would meet these requirements.
+Here, \isa{'t}, \isa{'n} and \isa{'m} are \emph{type variables} that can be substituted for arbitrary types; our only requirement is that timestamps \isa{'t} are globally unique and have an associated total order.
+Lamport timestamps~\cite{Lamport:1978jq} meet these requirements, for example.
 
@@
 For this reason we do not need a separate operation type to create tree nodes: simply performing a move with a fresh child node \isa{c} is sufficient to create \isa{c}.
-Moreover, we do not define a remove operation: instead, we can designate a particular tree node as the ``trash'', and perform deletion of a node \isa{c} by moving \isa{c} to be a child of the trash node.
+Moreover, we do not need to define a remove operation: instead, we can designate a particular tree node as the ``trash'', and perform deletion of a node \isa{c} by moving \isa{c} to be a child of the trash node.
 
@@
 The \isa{t}, \isa{p}, \isa{m} and \isa{c} fields are taken directly from the \isa{Move} record, while the \isa{oldp} field is filled in based on the state of the tree before the move.
-If \isa{c} did not exist in the tree prior to the move, \isa{oldp} is set to \isa{None}.
-Otherwise, \isa{oldp} records the previous parent and metadata of \isa{c}, that is: if there exists a \isa{p'} and \isa{m'} such that \isa{(p', m', c) {\isasymin} tree}, then \isa{oldp} is set to \isa{Some (p', m')}.
+If \isa{c} did not exist in the tree, \isa{oldp} is set to \isa{None}.
+Otherwise, \isa{oldp} records the previous parent and metadata of \isa{c}, that is: if there exist \isa{p'} and \isa{m'} such that \isa{(p', m', c) {\isasymin} tree}, then \isa{oldp} is set to \isa{Some (p', m')}.
 This logic is performed by the \isa{get\_parent} function (lines 16--20).
@@
 It then returns a \isa{('n {\isasymtimes} 'm) option}.
-\isa{{\isasymexists}!x} means ``there exists a unique value \isa{x} such that\dots'', whilst $\isa{THE}\ \isa{x}. \ldots$ means ``\emph{choose} the unique value \isa{x} such that\ldots''.
+\isa{{\isasymexists}!x} means ``there exists a unique value \isa{x} such that\dots'', while \isa{THE x}\dots means ``\emph{choose} the unique value \isa{x} such that\ldots''.
 
@@
 This relation is the transitive closure of the parent-child relation of a tree: if \isa{(p, m, c) {\isasymin} tree} then \isa{p} is an ancestor of \isa{c} (line 23); moreover, if \isa{(p, m, c) {\isasymin} tree} and \isa{a} is an ancestor of \isa{p}, then \isa{a} is also an ancestor of \isa{c} (line 24).
-The \textbf{inductive} keyword indicates that this recursive definition is iterated until a fixed point (the least) is reached.
+The \textbf{inductive} keyword indicates that this recursive definition is iterated until the least fixed point is reached.
 
@@
 Line 30 performs the check that ensures no cycles are introduced: if \isa{ancestor tree c newp}, i.e.\ if the node \isa{c} is being moved, and \isa{c} is an ancestor of the new parent \isa{newp}, then the tree is returned unmodified---in other words, the operation is ignored.
-Similarly, if \isa{c = newp}, the operation is also ignored.
+Similarly, the operation is also ignored if \isa{c = newp}.
 Otherwise (line 31), the tree is updated by removing \isa{c} from its existing parent, if any, and adding the new parent-child relationship \isa{(newp, m, c)} to the tree.
@@
 The \isa{apply\_op} function (lines 43--51) takes two arguments: a \isa{Move} operation to apply and the current replica state; and it returns the new replica state.
-The constraint \isa{'t::\{linorder\}} in the type signature indicates that timestamps \isa{'t} are instances of \isa{linorder} type class, and timestamps can therefore be compared with the \isa{<} operator defining a total order.
+The constraint \isa{'t::\{linorder\}} in the type signature indicates that timestamps \isa{'t} are instances of \isa{linorder} type class, and they can therefore be compared with the \isa{<} operator defining a total order.
 This comparison occurs on line 49.
@@
 \isa{undo\_op} (lines 33--36) inverts the effect of a previous move operation by restoring the prior parent and metadata that were recorded in the \isa{LogMove}'s additional field.
-\isa{redo\_op} (lines 38--41) uses \isa{do\_op} to perform an operation again and recomputes the \isa{LogMove} record (which might have changed due to the effect of the new operation with a lower timestamp).
+\isa{redo\_op} (lines 38--41) uses \isa{do\_op} to perform an operation again and recomputes the \isa{LogMove} record (which might have changed due to the effect of the new operation).
 
@@
 
-The consequence of performing the undo-do-redo cycle is that, effectively, the state of the tree is as if all operations had been applied using \isa{do\_op} in increasing timestamp order, regardless of the order in which they were actually applied.
+Due to the undo-do-redo cycle, the state of the tree is as if all operations had been applied using \isa{do\_op} in increasing timestamp order, regardless of the order in which they were actually applied.
 This provides a clear approach to the handling of conflicts:
@@
     \item If two operations concurrently move the same node, then the ``winning'' operation is the one with the higher timestamp.
-        This is the case because, if the operations are applied in increasing timestamp order, the node being moved is first moved to the parent indicated by the operation with the lower timestamp, and then moved again to the parent indicated by the operation with the higher timestamp.
+        Effectively, the operation with the lower timestamp moves the node first, and then the operation with the higher timestamp moves it again.
         Since move operations do not specify the old location of a node, but only the new location, this sequential execution of concurrent operations is well-defined.
@@
 For example, an operation may initially be regarded as safe, and then be reclassified as unsafe after applying a conflicting operation with a lower timestamp.
-The opposite is also possible: an operation previously regarded as unsafe may become safe through the application of a lower-timestamped operation that removes the risk of introducing a cycle.
+The opposite is also possible: an operation previously regarded as unsafe may become safe through the application of an operation that removes the risk of introducing a cycle.
 For this reason, the operation log must include all operations, even those that have no effect on the tree.
@@
 This can be implemented by maintaining a separate list CRDT for each branch node, e.g.\ using RGA~\cite{Roh:2011dw}, Treedoc~\cite{Preguica:2009fz}, Logoot~\cite{Weiss:2010hx}, or LSEQ~\cite{Nedelec:2013ky}.
-These algorithms assign a unique identifier to each element of the list, and this unique identifier can be included in the metadata field of move operations in order to determine the order of children of the same parent node.
+These algorithms assign a unique identifier to each element of the list, and this unique identifier can be included in the metadata field of move operations in order to determine the order sibling nodes.
 
@@
 To move a node to a different position in a list, we use the list CRDT to generate a new identifier at the desired position in the sequence.
-Then we perform a move operation in which the parent node is unchanged, but the metadata is updated to include this new identifier.
+Then we perform a move operation in which the parent node is unchanged, but the metadata is changed to this new identifier.
 
@@
 This can be accomplished by using CRDTs for file contents as well.
-For example, we have explored CRDT algorithms for applications using a JSON data model~\cite{Kleppmann:2016ve} and implemented them in a library called Automerge~\cite{Automerge}.
+For example, we have explored CRDT algorithms for applications using a JSON data model~\cite{Kleppmann:2016ve,LocalFirst} and implemented them in a library called Automerge~\cite{Automerge}.
 We discuss distributed filesystems further in Section~\ref{sec:filesystems}.
-
-% - No Synchronisation or consensus required. Available in the sense of CAP
-% - Rely only on uniqueness of node IDs, e.g. inode numbers
-% - Composing JSON/XML tree by using list datatype
-% - Location is Lww register. Concurrent move does not duplicate
-% - Unlike NFS, can work offline 
-% - Garbage collection: discard causally stable prefix
 
@@
 All theorems stated here have been formally proved and mechanically checked using Isabelle.
-For space reasons we give only a brief sketch of the proofs in this paper, but the full details can be found in the open source Isabelle files of our proof development.\footnote{\url{https://github.com/trvedata/move-op}}
+For space reasons we give only a brief sketch of the proofs in this paper, but the full details can be found in our open source Isabelle files~\cite{SourceFiles}.
 
@@
 It takes a list of operations \isa{ops} and returns the state of a replica after it has applied all the operations in \isa{ops}.
-The \isa{apply\_ops} function works by starting in the initial state \isa{([], \{\})} consisting of the empty operation log \isa{[]} and the empty tree \isa{\{\}}, and then applying the operations one by one to the state using the \isa{apply\_op} function (introduced in Section~\ref{sec:applying}).
-The \isa{foldl} function from Isabelle's standard library performs the iteration over the list of operations.
+The \isa{apply\_ops} function works by starting in the initial state \isa{([], \{\})} consisting of the empty operation log \isa{[]} and the tree represented by the empty set \isa{\{\}}, and then applying the operations one by one to the state using the \isa{apply\_op} function (introduced in Section~\ref{sec:applying}).
+The \isa{foldl} function from the Isabelle/HOL standard library performs the iteration over the list of operations.
 
@@
 
-\subsubsection{Each node's parent is unique}
+\subsubsection{Each node's parent is unique}\label{sec:unique-parent}
 
@@
 Because we make no assumptions about \isa{ops}, this theorem holds for any replica state that can be reached by applying any number of operations.
-Proving \isa{apply\_ops\_unique\_parent} is easy because each of the \isa{do\_op}, \isa{undo\_op} and \isa{redo\_op} functions individually maintains the \isa{unique\_parent tree} invariant, and thus \isa{apply\_op} also maintains the invariant.
-
-\subsubsection{The graph contains no cycles}
+Proving it is easy because each of the \isa{do\_op}, \isa{undo\_op} and \isa{redo\_op} functions individually maintains the invariant \isa{unique\_parent tree}, and thus \isa{apply\_op} also maintains the invariant.
+
+\subsubsection{The graph contains no cycles}\label{sec:no-cycles}
 
 The second invariant we require is that the graph must not contain any cycles.
-We state this theorem in Isabelle as follows:
+In Isabelle:
 \begin{isabelle}
@@
 
-To prove that \isa{apply\_op} preserves the acyclicity invariant, we need to show that for any tree produced by \isa{apply\_ops}, there exists a sequence of operations such that applying those operations using \isa{do\_op} invocations alone (not using \isa{undo\_op} or \isa{redo\_op}) results in the same tree.
+To prove that \isa{apply\_op} preserves the acyclicity invariant, we first show that for any tree produced by \isa{apply\_ops}, there exists a sequence of operations such that applying those operations using \isa{do\_op} invocations alone (not using \isa{undo\_op} or \isa{redo\_op}) results in the same tree.
 (This is the case when operations are applied in order of increasing timestamp.)
-Then we can use the result of the previous paragraph to prove that any tree produced by \isa{apply\_ops} must be acyclic.
-
-\subsection{Convergence}
+Then, by the result of the previous paragraph, any tree produced by \isa{apply\_ops} must be acyclic.
+
+\subsection{Convergence}\label{sec:convergence}
 
@@
 
-\subsubsection{Commutativity of apply\_op}
+\subsubsection{Commutativity of apply\_op}\label{sec:commutativity}
 
@@
 Based on the commutativity proof above we integrate our tree datatype with this framework, and thus prove that our move operation on trees does indeed guarantee strong eventual consistency.
-The details of this proof are given in the open source Isabelle source files.\footnote{\url{https://github.com/trvedata/move-op}}
-
-\section{Evaluation}\label{sec:evaluation}
-
-\subsection{Extracting efficient algorithms from HOL}
-
-Isabelle/HOL can generate executable Haskell, OCaml, Scala, and Standard ML code from HOL type and constant definitions using a sophisticated code generation mechanism~\cite{DBLP:conf/flops/HaftmannN10}.
-However,  care must be taken when using this mechanism, as:
-\begin{enumerate}
-\item
-The mechanism is \emph{partial}: not all HOL definitions can be realised in executable form.
-Definitions making essential use of choice principles (e.g.~our \isa{get\_parent}) are obvious examples of definitions that will cause problems.
-However, \emph{inductively defined} relations characterised by a series of introduction rules, such as our \isa{ancestor} relation, are also problematic as their execution necessarily requires Prolog-like proof search.
-\item
-Na\"{i}ve use often leads to inefficient target language code, especially true when using HOL's \isa{set} type, a type of potentially infinite sets.
-Whilst convenient for theorem proving this type often leads to inefficiency in generated code.
-\end{enumerate}
-
-We therefore produce a series of definitions that are optimised for execution rather than theorem proving, before trying to evaluate our algorithms.
-Notably, we eschew the HOL \isa{set} type, and instead use a more efficient hash-map implementation provided by the Isabelle Collections Framework~\cite{DBLP:conf/itp/LammichL10}, a library of verified, efficient container datatypes and associated algorithms.
-
-Write $('a::\{hashable\}, 'b)\ \isa{hm}$ for the type of hash-maps with key type $\isa{'a}$ and binding type $\isa{'b}$.
-Note that keys have an explicit $\isa{hashable}$ constraint, meaning that this hash-map type may only be used with kay types with an associated hashing function.
-With this type, we represent trees by the type $(\isa{'n}::\{hashable\}, \isa{'m} \times \isa{'n})\ \isa{hm}$, mapping from a \emph{child} node to its \emph{parent} node and associated metadata, if any such mapping exists.
+The details of this proof are given in our open source Isabelle source files~\cite{SourceFiles}.
+
+\subsection{Extracting efficient algorithms from HOL}\label{sec:code-gen}
+
+Isabelle/HOL can generate executable Haskell, OCaml, Scala, and Standard ML code from HOL definitions using a sophisticated code generation mechanism~\cite{DBLP:conf/flops/HaftmannN10}.
+However, not all definitions can be realised in executable form: for example, the use of choice principles (like in \isa{get\_parent}) and inductively defined relations (e.g.\ \isa{ancestor}) cause problems.
+Moreover, HOL's \isa{set} type allows infinite sets.
+Whilst these constructs are convenient for theorem proving, they do not translate well to efficient code.
+
+We therefore produce variants of the definitions in Figure~\ref{fig:code} that are designed for execution rather than theorem proving.
+Rather than representing the tree as a set of (parent, metadata, child) triples, these definitions use a hash-map in which the keys are child nodes, and the values are (metadata, parent) pairs, written in Isabelle as \isa{('n::\{hashable\}, 'm {\isasymtimes} 'n) hm}.
+The \isa{hashable} type class on keys means that keys must have an associated hashing function.
+In effect, this hash-map is an index over the set of triples, using the fact that the parent and metadata for a given child are unique (Section~\ref{sec:unique-parent}).
+We use the hash-map implementation from the Isabelle Collections Framework~\cite{DBLP:conf/itp/LammichL10}.
 
@@
 
+\section{Evaluation}\label{sec:evaluation}
+
 \subsection{Performance experiments}
@@
 We perform a synthetic workload in which each replica generates move operations at the same rate; for each move operation the generating replica chooses parent and child nodes uniformly at random from a set of 1,000 tree nodes.
+For a given operation rate we run the system for 10 minutes to reach a steady state.
 We use Lamport timestamps~\cite{Lamport:1978jq} as operation timestamps, and string values to identify tree nodes.
@@
 
-Coda is a distributed client-server file system which allows clients to cache or \emph{hoard} copies of files otherwise stored in a shared data repository~\cite{kistler1992coda}. 
+Coda is a distributed file system which allows clients to cache or \emph{hoard} copies of files otherwise stored in a shared data repository~\cite{kistler1992coda}. 
 Clients can edit copies of data held in the cache while offline, during which time a kernel module keeps track of all updates. 
@@
 This design means that conflicts may arise if multiple clients edit the same files or directories while one or more clients are offline. 
-One proposed solution to resolving conflicts in files is to support application-specific resolvers which ``encapsulate the knowledge needed for file resolution''~\cite{Kumar:1995wf}, an approach is similar to that found in Bayou~\cite{Terry:1995dn}.
-Concurrent edits on directory entries across different replicas in Coda have been considered, but the authors note that they do ``not address transparent resolution of cross-directory [move operations] in [their] current implementation.''~\cite{kumar1993log}.
-Furthermore, while the authors consider a number of conflicts associated with directory move operations, they do not highlight the potential for the creation of cycles as seen in Figure~\ref{fig:move-cycle} (a).
-
-The Ficus replicated file system~\cite{Reiher:1994wh} is a in-kernel SunOS-based replicated peer-to-peer file system. 
-Ficus supported updates to replicas during periods of network partition and claims ``conflicting updates to directories are detected an automatically repaired''~\cite{guy1990implementation}. 
-Unfortunately we were unable to find a precise definition of the algorithm used in any of the available publications. 
-Rumor~\cite{Guy:1999gy} is the successor to Ficus. 
-An important distinction is between Rumor and previous work is that it operates at the application layer and needs to be invoked periodically by the user or a deamon process.
-Rumor is therefore very similar in operation and design to Unison described earlier.
-
-%The user manual (in Archive) is quite helpful for Rumor: https://web.archive.org/web/20170705140958/ftp://ftp.cs.ucla.edu/pub/rumor/rumor_users_manual.ps.
-Note that this also makes the claim that "Unix directories are an important example of a kind of file for which Rumor automatically resolves conflicts"
-
-TODO: Roam~\cite{Ratner:1999fh}
+One proposed solution to this is to support application-specific resolvers which "encapsulate the knowledge needed for file resolution"~\cite{Kumar:1995wf}, an approach is similar to that found in Bayou~\cite{Terry:1995dn}.
+
+Ficus~\cite{Reiher:1994wh}, Rumor~\cite{Guy:1999gy}, Roam~\cite{Ratner:1999fh}
 
@@
 % Disconnected operation
+% - No Synchronisation or consensus required. Available in the sense of CAP
+% - Rely only on uniqueness of node IDs, e.g. inode numbers
+% - Composing JSON/XML tree by using list datatype
+% - Location is Lww register. Concurrent move does not duplicate
+% - Unlike NFS, can work offline 
+% - Garbage collection: discard causally stable prefix
 

### 2019-06-20 paper/move-op.tex

@@
 
-The first issue arises when a node is moved to a new location on one replica, and concurrently the same node is moved to a different location on another replica.
+The first difficulty arises when a node is concurrently moved into different locations on different replicas.
 This scenario is illustrated in Figure~\ref{fig:move-same-item}, where replica 1 moves node $A$ to be a child of $B$, while concurrently replica 2 moves $A$ to be a child of $C$.
@@
 A move operation is a record consisting of four fields: a timestamp \isa{t} of type \isa{'t}, a parent node \isa{p} of type \isa{'n}, a metadata field \isa{m} of type \isa{'m}, and a child node \isa{c} of type \isa{'n}.
-Here, \isa{'t}, \isa{'n} and \isa{'m} are \emph{type variables} that can be substituted for arbitrary types; our only requirement is that timestamps \isa{'t} are globally unique and have an associated total order.
-Lamport timestamps~\cite{Lamport:1978jq} meet these requirements, for example.
+Here, \isa{'t}, \isa{'n} and \isa{'m} are \emph{type variables} that can be substituted for arbitrary types; our only requirement is that timestamps \isa{'t} are globally unique and have an associated total order (e.g.
+Lamport timestamps~\cite{Lamport:1978jq} would suffice).
 
@@
 For this reason we do not need a separate operation type to create tree nodes: simply performing a move with a fresh child node \isa{c} is sufficient to create \isa{c}.
-Moreover, we do not need to define a remove operation: instead, we can designate a particular tree node as the ``trash'', and perform deletion of a node \isa{c} by moving \isa{c} to be a child of the trash node.
+Nor do we need to define a remove operation: instead, we can designate a particular tree node as the ``trash'', and perform deletion of a node \isa{c} by moving \isa{c} to be a child of the trash node.
 
@@
 Otherwise, \isa{oldp} records the previous parent and metadata of \isa{c}, that is: if there exist \isa{p'} and \isa{m'} such that \isa{(p', m', c) {\isasymin} tree}, then \isa{oldp} is set to \isa{Some (p', m')}.
-This logic is performed by the \isa{get\_parent} function (lines 16--20).
+The \isa{get\_parent} function implements this (lines 16--20).
 
@@
 
-Because we make no assumptions about \isa{ops}, this theorem holds for any replica state that can be reached by applying any number of operations.
+As we make no assumptions about \isa{ops}, this theorem holds for any replica state that can be reached by applying any number of operations.
 Proving it is easy because each of the \isa{do\_op}, \isa{undo\_op} and \isa{redo\_op} functions individually maintains the invariant \isa{unique\_parent tree}, and thus \isa{apply\_op} also maintains the invariant.
@@
 
-To prove that \isa{apply\_op} preserves the acyclicity invariant, we first show that for any tree produced by \isa{apply\_ops}, there exists a sequence of operations such that applying those operations using \isa{do\_op} invocations alone (not using \isa{undo\_op} or \isa{redo\_op}) results in the same tree.
+To prove that \isa{apply\_op} preserves acyclicity, we first show that for any tree produced by \isa{apply\_ops}, there exists a sequence of operations such that applying those operations using \isa{do\_op} invocations alone (not using \isa{undo\_op} or \isa{redo\_op}) results in the same tree.
 (This is the case when operations are applied in order of increasing timestamp.)
@@
 Based on the commutativity proof above we integrate our tree datatype with this framework, and thus prove that our move operation on trees does indeed guarantee strong eventual consistency.
-The details of this proof are given in our open source Isabelle source files~\cite{SourceFiles}.
-
-\subsection{Extracting efficient algorithms from HOL}\label{sec:code-gen}
+The details of this proof are given in our open source Isabelle proof files~\cite{SourceFiles}.
+
+\subsection{Extracting efficient algorithms from HOL} \label{subsect.extracting}
 

### 2019-06-20 paper/move-op.tex
paper: renamed "efficient" algorithms to "executable"
Rewording in explanation of practical results to make it sound more
positive, also.

--- 2019-06-20 paper/move-op.tex
@@
 
-\subsection{Extracting efficient algorithms from HOL} \label{subsect.extracting}
-
-Isabelle/HOL can generate executable Haskell, OCaml, Scala, and Standard ML code from HOL type and constant definitions using a sophisticated code generation mechanism~\cite{DBLP:conf/flops/HaftmannN10}.
-However,  care must be taken when using this mechanism, as:
+\subsection{Making the HOL definitions executable} \label{subsect.extracting}
+
+Isabelle/HOL's code generation mechanism~\cite{DBLP:conf/flops/HaftmannN10} can produce executable code from HOL definitions.
+However,  care must be taken:
 \begin{enumerate}
@@
 
-We therefore produce a series of definitions that are optimised for execution rather than theorem proving, before trying to evaluate our algorithms.
+We therefore produce definitions optimised for execution rather than proof, prior to practical evaluation.
 Notably, we eschew the HOL \isa{set} type, and instead use a more efficient hash-map implementation provided by the Isabelle Collections Framework~\cite{DBLP:conf/itp/LammichL10}, a library of verified, efficient container datatypes and associated algorithms.
@@
 
-Our approach is to produce a new set of definitions---mirroring the structure of our existing ones---wherein we replace all occurrences of the $(\isa{'n} \times \isa{'m} \times \isa{'n})\ \isa{set}$ type, and any associated set operations, with the $(\isa{'n}, \isa{'m} \times \isa{'n})\ \isa{hm}$ type and associated hash-map operations.
-For clarity, we refer to any definition using hash-maps as an ``efficient'' definition, and any definition making use of the \isa{set} type as an ``abstract'' definition.
+We produce a new set of definitions---mirroring the structure of our existing ones---wherein we replace all occurrences of the $(\isa{'n} \times \isa{'m} \times \isa{'n})\ \isa{set}$ type, and any associated set operations, with the $(\isa{'n}, \isa{'m} \times \isa{'n})\ \isa{hm}$ type and associated hash-map operations.
+For clarity, we refer to any definition using hash-maps as an ``executable'' definition, and any definition making use of the \isa{set} type as an ``abstract'' definition.
 For most definitions this is a mechanical process, and we elide further discussion of the required changes, referring the interested reader to our Isabelle/HOL source code.
-However \isa{ancestor} is different to other definitions as it is an inductively defined relation rather than a recursive function, and we need further work to make this efficiently executable.
+However \isa{ancestor} is different to other definitions as it is an inductively defined relation rather than a recursive function, and we need further work to make this executable.
 We characterise this relation with the following:
@@
 \isacommand{theorem}\isamarkupfalse%
-\ efficient{\isacharunderscore}ancestor{\isacharunderscore}simp\ {\isacharbrackleft}code{\isacharbrackright}{\isacharcolon}\isanewline
-\ \ \isakeyword{shows}\ {\isacartoucheopen}efficient{\isacharunderscore}ancestor\ t\ p\ c\ {\isasymlongleftrightarrow}\isanewline
+\ executable{\isacharunderscore}ancestor{\isacharunderscore}simp\ {\isacharbrackleft}code{\isacharbrackright}{\isacharcolon}\isanewline
+\ \ \isakeyword{shows}\ {\isacartoucheopen}executable{\isacharunderscore}ancestor\ t\ p\ c\ {\isasymlongleftrightarrow}\isanewline
 \ \ \ \ \ \ \ \ \ \ {\isacharparenleft}case\ hm{\isachardot}lookup\ c\ t\ of\isanewline
@@
 \ \ \ \ \ \ \ \ \ \ \ \ {\isacharbar}\ Some\ {\isacharparenleft}m{\isacharcomma}\ a{\isacharparenright}\ {\isasymRightarrow}\isanewline
-\ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ a\ {\isacharequal}\ p\ {\isasymor}\ efficient{\isacharunderscore}ancestor\ t\ p\ a{\isacharparenright}{\isacartoucheclose}
+\ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ a\ {\isacharequal}\ p\ {\isasymor}\ executable{\isacharunderscore}ancestor\ t\ p\ a{\isacharparenright}{\isacartoucheclose}
 \end{isabelle}
 
-Note that \isa{efficient\_ancestor\_simp} is tagged with a $\isa{code}$ ``attribute'', a hint that this theorem should be used as a transformation during code generation.
-Isabelle/HOL takes this theorem as the definition, for code generation, of the \isa{efficient\_ancestor} relation and is able to directly translate this to a recursive function definition in each target backend---this transformation is ``semantics preserving'', as the characterisation was proved as a theorem within HOL.
-
-Next, we demonstrate the correctness of our efficient definitions against our abstract definitions, and the properties that we prove are best described using an example, as they all follow a similar pattern.
-We use the lemma \isa{efficient\_do\_op\_simulates} as our exemplar:
+Note that \isa{executable\_ancestor\_simp} is tagged with a $\isa{code}$ ``attribute'', a hint that this theorem should be used as a transformation during code generation.
+Isabelle/HOL takes this theorem as the definition, for code generation, of the \isa{executable\_ancestor} relation and is able to translate this to a recursive function definition in each target backend---this transformation is ``semantics preserving'', as the characterisation was proved as a theorem within HOL.
+
+We demonstrate the correctness of our executable definitions against our abstract definitions.
+The properties proved are best described using an example, as they all follow a similar pattern.
+We use the lemma \isa{executable\_do\_op\_simulates} as our exemplar:
 
@@
 \isacommand{lemma}\isamarkupfalse%
-\ efficient{\isacharunderscore}do{\isacharunderscore}op{\isacharunderscore}simulates{\isacharcolon}\isanewline
+\ executable{\isacharunderscore}do{\isacharunderscore}op{\isacharunderscore}simulates{\isacharcolon}\isanewline
 \ \ \isakeyword{assumes}\ {\isadigit{1}}{\isacharcolon}\ {\isacartoucheopen}t\ {\isasympreceq}\ T{\isacartoucheclose}\isanewline
-\ \ \ \ \isakeyword{and}\ {\isadigit{2}}{\isacharcolon}\ {\isacartoucheopen}efficient{\isacharunderscore}do{\isacharunderscore}op\ {\isacharparenleft}oper{\isacharcomma}\ t{\isacharparenright}\ {\isacharequal}\ {\isacharparenleft}log{\isadigit{1}}{\isacharcomma}\ u{\isacharparenright}{\isacartoucheclose}\isanewline
+\ \ \ \ \isakeyword{and}\ {\isadigit{2}}{\isacharcolon}\ {\isacartoucheopen}executable{\isacharunderscore}do{\isacharunderscore}op\ {\isacharparenleft}oper{\isacharcomma}\ t{\isacharparenright}\ {\isacharequal}\ {\isacharparenleft}log{\isadigit{1}}{\isacharcomma}\ u{\isacharparenright}{\isacartoucheclose}\isanewline
 \ \ \ \ \isakeyword{and}\ {\isadigit{3}}{\isacharcolon}\ {\isacartoucheopen}do{\isacharunderscore}op\ {\isacharparenleft}oper{\isacharcomma}\ T{\isacharparenright}\ {\isacharequal}\ {\isacharparenleft}log{\isadigit{2}}{\isacharcomma}\ U{\isacharparenright}{\isacartoucheclose}\isanewline
@@
 
-We show that the \isa{efficient\_do\_op} and the \isa{do\_op} functions are in ``lock-step'', mapping related trees $t \preceq T$ to related trees $u \preceq U$ and producing identical logs of operations whenever they are applied to the same input operation, \isa{oper}.
-This lock-step is repeated bottom-up for every efficient-abstract algorithm pairs, until we reach the final correctness theorem for \isa{apply\_ops}:
+We show that the \isa{executable\_do\_op} and the \isa{do\_op} functions are in ``lock-step'', mapping related trees $t \preceq T$ to related trees $u \preceq U$ and producing identical logs of operations whenever they are applied to the same input operation, \isa{oper}.
+This lock-step is repeated bottom-up for every executable-abstract algorithm pairs, until we reach the final correctness theorem for \isa{apply\_ops}:
 
@@
 \isacommand{lemma}\isamarkupfalse%
-\ efficient{\isacharunderscore}apply{\isacharunderscore}ops{\isacharunderscore}simulates{\isacharcolon}\isanewline
-\ \ \isakeyword{assumes}\ {\isadigit{1}}{\isacharcolon}\ {\isacartoucheopen}efficient{\isacharunderscore}apply{\isacharunderscore}ops\ opers\ {\isacharequal}\ {\isacharparenleft}log{\isadigit{1}}{\isacharcomma}\ u{\isacharparenright}{\isacartoucheclose}\isanewline
+\ executable{\isacharunderscore}apply{\isacharunderscore}ops{\isacharunderscore}simulates{\isacharcolon}\isanewline
+\ \ \isakeyword{assumes}\ {\isadigit{1}}{\isacharcolon}\ {\isacartoucheopen}executable{\isacharunderscore}apply{\isacharunderscore}ops\ opers\ {\isacharequal}\ {\isacharparenleft}log{\isadigit{1}}{\isacharcomma}\ u{\isacharparenright}{\isacartoucheclose}\isanewline
 \ \ \ \ \isakeyword{and}\ {\isadigit{2}}{\isacharcolon}\ {\isacartoucheopen}apply{\isacharunderscore}ops\ opers\ {\isacharequal}\ {\isacharparenleft}log{\isadigit{2}}{\isacharcomma}\ U{\isacharparenright}{\isacartoucheclose}\isanewline
@@
 the two trees are extensionally equivalent, despite having very different concrete representations.
-Moreover, we have the following corollaries of the theorems \isa{apply\_ops\_commutes} and \isa{apply\_ops\_acyclic}, respectively, further establishing the correctness of our efficient algorithms:\footnote{The ``efficient'' analogue of \isa{apply\_ops\_unique\_parent} is trivially true, as the output tree is stored as a hash-map, and therefore too trivial to state as a theorem.}:
+Moreover, we have the following corollaries of \isa{apply\_ops\_commutes} and \isa{apply\_ops\_acyclic}, respectively, further establishing the correctness of our executable algorithms:\footnote{The ``executable'' analogue of \isa{apply\_ops\_unique\_parent} is trivially true, as the output tree is stored as a hash-map, and therefore too trivial to state as a theorem.}:
 
@@
 \isacommand{theorem}\isamarkupfalse%
-\ efficient{\isacharunderscore}apply{\isacharunderscore}ops{\isacharunderscore}commutes{\isacharcolon}\isanewline
+\ executable{\isacharunderscore}apply{\isacharunderscore}ops{\isacharunderscore}commutes{\isacharcolon}\isanewline
 \ \ \isakeyword{assumes}\ {\isadigit{1}}{\isacharcolon}\ {\isacartoucheopen}set\ ops{\isadigit{1}}\ {\isacharequal}\ set\ ops{\isadigit{2}}{\isacartoucheclose}\isanewline
@@
 \ \ \ \ \isakeyword{and}\ {\isadigit{3}}{\isacharcolon}\ {\isacartoucheopen}distinct\ {\isacharparenleft}map\ move{\isacharunderscore}time\ ops{\isadigit{2}}{\isacharparenright}{\isacartoucheclose}\isanewline
-\ \ \ \ \isakeyword{and}\ {\isadigit{4}}{\isacharcolon}\ {\isacartoucheopen}efficient{\isacharunderscore}apply{\isacharunderscore}ops\ ops{\isadigit{1}}\ {\isacharequal}\ {\isacharparenleft}log{\isadigit{1}}{\isacharcomma}\ t{\isacharparenright}{\isacartoucheclose}\isanewline
-\ \ \ \ \isakeyword{and}\ {\isadigit{5}}{\isacharcolon}\ {\isacartoucheopen}efficient{\isacharunderscore}apply{\isacharunderscore}ops\ ops{\isadigit{2}}\ {\isacharequal}\ {\isacharparenleft}log{\isadigit{2}}{\isacharcomma}\ u{\isacharparenright}{\isacartoucheclose}\isanewline
+\ \ \ \ \isakeyword{and}\ {\isadigit{4}}{\isacharcolon}\ {\isacartoucheopen}executable{\isacharunderscore}apply{\isacharunderscore}ops\ ops{\isadigit{1}}\ {\isacharequal}\ {\isacharparenleft}log{\isadigit{1}}{\isacharcomma}\ t{\isacharparenright}{\isacartoucheclose}\isanewline
+\ \ \ \ \isakeyword{and}\ {\isadigit{5}}{\isacharcolon}\ {\isacartoucheopen}executable{\isacharunderscore}apply{\isacharunderscore}ops\ ops{\isadigit{2}}\ {\isacharequal}\ {\isacharparenleft}log{\isadigit{2}}{\isacharcomma}\ u{\isacharparenright}{\isacartoucheclose}\isanewline
 \ \ \isakeyword{shows}\ {\isacartoucheopen}log{\isadigit{1}}\ {\isacharequal}\ log{\isadigit{2}}\ {\isasymand}\ hm{\isachardot}lookup\ c\ t\ {\isacharequal}\ hm{\isachardot}lookup\ c\ u{\isacartoucheclose}\isanewline
@@
 \isacommand{theorem}\isamarkupfalse%
-\ efficient{\isacharunderscore}apply{\isacharunderscore}ops{\isacharunderscore}acyclic{\isacharcolon}\isanewline
-\ \ \isakeyword{assumes}\ {\isadigit{1}}{\isacharcolon}\ {\isacartoucheopen}efficient{\isacharunderscore}apply{\isacharunderscore}ops\ ops\ {\isacharequal}\ {\isacharparenleft}log{\isacharcomma}\ t{\isacharparenright}{\isacartoucheclose}\isanewline
-\ \ \isakeyword{shows}\ {\isacartoucheopen}{\isasymnexists}n{\isachardot}\ efficient{\isacharunderscore}ancestor\ t\ n\ n{\isacartoucheclose}
+\ executable{\isacharunderscore}apply{\isacharunderscore}ops{\isacharunderscore}acyclic{\isacharcolon}\isanewline
+\ \ \isakeyword{assumes}\ {\isadigit{1}}{\isacharcolon}\ {\isacartoucheopen}executable{\isacharunderscore}apply{\isacharunderscore}ops\ ops\ {\isacharequal}\ {\isacharparenleft}log{\isacharcomma}\ t{\isacharparenright}{\isacartoucheclose}\isanewline
+\ \ \isakeyword{shows}\ {\isacartoucheopen}{\isasymnexists}n{\isachardot}\ executable{\isacharunderscore}ancestor\ t\ n\ n{\isacartoucheclose}
 \end{isabelle}
 
-Note the statement of \isa{efficient\_apply\_ops\_commutes} differs from that of \isa{apply\_ops\_commutes}---the latter asserts identical sets are obtained from convergence, whilst the former asserts that reading any key binding in $t$ gives the same result as a reading in $u$.
-Whilst this statement appears weaker than that of \isa{apply\_ops\_acyclic}, it is correct, and the strongest statement one can hope for: the resulting hash-maps may contain the same key-value bindings but may not be structurally equal, depending on the ordering of operations.
+Note the statement of \isa{executable\_apply\_ops\_commutes} differs from \isa{apply\_ops\_commutes}---the latter asserts identical sets are obtained from convergence, whilst the former asserts that reading any key binding in $t$ gives the same result as a reading in $u$.
+Whilst this is weaker than that of \isa{apply\_ops\_acyclic}, it is the strongest statement one can hope for: the resulting hash-maps may contain the same key-value bindings but may not be structurally equal, depending on the ordering of operations.
 
@@
 
-We used Isabelle/HOL's code generation mechanism to extract executable Scala code from our HOL definitions, as discussed in Subsection~\ref{subsect.extracting}, and wrapped this implementation in a network service using the Akka~\cite{Akka} actor framework, using gRPC~\cite{gRPC} for network communication.
+We used Isabelle/HOL's code generation mechanism to extract executable Scala code from our HOL definitions, as discussed in \S\ref{subsect.extracting}, and wrapped this implementation in a network service using the Akka~\cite{Akka} actor framework, using gRPC~\cite{gRPC} for network communication.
 We deployed three replicas of this service on Amazon EC2 \texttt{t2.small} instances in different regions worldwide: one in Northern California (region \texttt{us-west-1}), one in Ireland (\texttt{eu-west-1}), and one in Singapore (\texttt{ap-southeast-1}).
@@
 We use Lamport timestamps~\cite{Lamport:1978jq} as operation timestamps, and string values to identify tree nodes.
-When a replica generates an operation, it immediately applies that operation to the local replica state, and then asynchronously sends the operation to the other two replicas via gRPC requests.
+When a replica generates an operation, it immediately applies that operation to its local state, and then asynchronously sends the operation to the other two replicas via gRPC requests.
 When a replica receives an operation from a remote replica, it also applies that operation to its own replica state.
@@
 
-A locally generated operation always has a timestamp that is higher than any existing operation at the generating replica at that time (by the definition of Lamport timestamps), and thus applying that operation requires only a \isa{do\_op}, and no \isa{undo\_op} or \isa{redo\_op}, explaining the near-constant execution time.
-On the other hand, for remote operations, some number of undos and redos needs to be performed, depending on the degree of concurrency.
-As the rate of operations increases, the time between successive operations becomes smaller than the network delay between replicas; thus, there are more operations ``in flight'' at the same time, and more calls to \isa{undo\_op} and \isa{redo\_op} are required to put operations into timestamp order.
+A locally generated operation always has a timestamp greater than any existing operation at the generating replica at that time (from the use of Lamport timestamps), and thus applying that operation requires only a \isa{do\_op}, and no \isa{undo\_op} or \isa{redo\_op}, hence the near-constant execution time.
+However, for remote operations, a number of undos and redos must be performed, depending on the degree of concurrency.
+As the rate of operations increases, the time between successive operations becomes smaller than the network delay between replicas; thus, there are more operations ``in flight'' at the same time, and more calls to \isa{undo\_op} and \isa{redo\_op} are required order operations by timestamp.
 
@@
 
-The Scala implementation generated from our Isabelle definitions is optimised for correctness, not for performance; we expect that a more optimised implementation could achieve much higher throughput than our 24 operations/sec.
-Nevertheless, we believe that our evaluation sends a positive signal about the practical applicability of our algorithm.
-
-Note that when a user is interacting with a system, they only have to wait for local operations to execute; any remote operations can be applied in the background without affecting user interaction.
+A hand-optimised implementation of our algorithm could achieve higher throughput than our 24 operations/sec: despite \S\ref{subsect.extracting}, the code generated from our HOL definitions contains many optimisation opportunities due to its extraction from a theorem prover.
+Nevertheless, our evaluation sends a positive signal about the practical applicability of our algorithm, and the generated code could serve as an executable ``golden model'' for validating other implementations.
+
+Note that when a user is interacting with a system they only have to wait for local operations to execute; any remote operations can be applied in the background without affecting user interaction.
 Since our algorithm need not perform any undos or redos for local operations, these user interactions are consistently fast.

### 2019-06-20 paper/move-op.tex

--- 2019-06-20 paper/move-op.tex
@@
 
-\subsection{Extracting efficient algorithms from HOL} \label{subsect.extracting}
+\subsection{Making the HOL definitions executable} \label{subsect.extracting}
 

### 2019-06-20 paper/move-op.tex
paper: minor editing, removing single-word lines

@@
 
-We perform a synthetic workload in which each replica generates move operations at the same rate; for each move operation the generating replica chooses parent and child nodes uniformly at random from a set of 1,000 tree nodes.
+We use a synthetic workload wherein each replica generates move operations at the same rate; for each move operation the generating replica chooses parent and child nodes uniformly at random from a set of 1,000 tree nodes.
 For a given operation rate we run the system for 10 minutes to reach a steady state.
 We use Lamport timestamps~\cite{Lamport:1978jq} as operation timestamps, and string values to identify tree nodes.
-When a replica generates an operation, it immediately applies that operation to its local state, and then asynchronously sends the operation to the other two replicas via gRPC requests.
+When a replica generates an operation, it immediately applies that operation to its local state, and asynchronously sends the operation to the other two replicas via gRPC requests.
 When a replica receives an operation from a remote replica, it also applies that operation to its own replica state.

### 2019-06-20 evaluation/src/main/scala/Move_Code.scala
scala: added new extracted Scala file

@@ -1152,20 +1152,30 @@ def map_option[A, B](f: A => B, x1: Option[A]): Option[B] = (f, x1) match {
   case (f, Some(x2)) => Some[B](f(x2))
 }
 
-def efficient_ancestor[A : equal : hashable,
-                        B](t: hashmap[A, (B, A)], p: A, c: A):
+def executable_undo_op[A, B : equal : hashable,
+                        C](x0: (log_op[A, B, C], hashmap[B, (C, B)])):
+      hashmap[B, (C, B)]
+  =
+  x0 match {
+  case (LogMove(t, Some((oldp, oldm)), newp, m, c), tree) =>
+    hm_update[B, (C, B)](c, (oldm, oldp), tree)
+  case (LogMove(t, None, newp, m, c), tree) => hm_delete[B, (C, B)](c, tree)
+}
+
+def executable_ancestor[A : equal : hashable,
+                         B](t: hashmap[A, (B, A)], p: A, c: A):
       Boolean
   =
   (hm_lookup[A, (B, A)](c, t) match {
      case None => false
      case Some(a) => {
                        val (_, aa): (B, A) = a;
-                       eq[A](aa, p) || efficient_ancestor[A, B](t, p, aa)
+                       eq[A](aa, p) || executable_ancestor[A, B](t, p, aa)
                      }
    })
 
-def efficient_do_op[A, B : equal : hashable,
-                     C](x0: (operation[A, B, C], hashmap[B, (C, B)])):
+def executable_do_op[A, B : equal : hashable,
+                      C](x0: (operation[A, B, C], hashmap[B, (C, B)])):
       (log_op[A, B, C], hashmap[B, (C, B)])
   =
   x0 match {
@@ -1176,57 +1186,47 @@ def efficient_do_op[A, B : equal : hashable,
   (snd[C, B](x), fst[C, B](x))),
  hm_lookup[B, (C, B)](c, tree)),
                   newp, m, c),
-      (if (efficient_ancestor[B, C](tree, c, newp) || eq[B](c, newp)) tree
+      (if (executable_ancestor[B, C](tree, c, newp) || eq[B](c, newp)) tree
         else hm_update[B, (C, B)](c, (m, newp), tree)))
 }
 
-def efficient_undo_op[A, B : equal : hashable,
-                       C](x0: (log_op[A, B, C], hashmap[B, (C, B)])):
-      hashmap[B, (C, B)]
-  =
[…]

### 2019-06-20 paper/move-op.tex
paper: Section --> \S throughout
Some other minor save-spacing edits here and there.

@@
         In Unix-style filesystems, the nodes of the tree are known as \emph{inodes}.
-        Strictly speaking, filesystems that support hardlinks form a DAG, but we will focus on trees for now and show in Section~\ref{sec:hardlinks} how to add support for links.
+        Strictly speaking, filesystems that support hardlinks form a DAG, but we will focus on trees for now and show in \S\ref{sec:hardlinks} how to add support for links.
     \item \textbf{Rich text editors} maintain a tree of textual structures such as paragraphs, bulleted and numbered lists, figures, sections, and so on.
@@
 We allow replicas to manipulate this tree in any way: by creating new nodes, deleting nodes, or moving subtrees to a new location within the tree.
-While there are many existing systems that support creating and deleting nodes (see Section~\ref{sec:relwork}), the key innovation of our algorithm is the support for moving subtrees.
-We explain in Section~\ref{sec:move-is-hard} why this move operation is so challenging.
+While there are many existing systems that support creating and deleting nodes (see \S\ref{sec:relwork}), the key innovation of our algorithm is the support for moving subtrees.
+We explain in \S\ref{sec:move-is-hard} why this move operation is so challenging.
 In summary, our contributions are:
@@
     \item We define a Conflict-free Replicated Data Type for trees that supports a \emph{move} operation, without requiring any coordination between replicas such as locking or consensus.
-        As discussed in Section~\ref{sec:impossibility}, this has previously been thought to be impossible to achieve \cite{Najafzadeh:2017vk,Najafzadeh:2018bw}.
+        As discussed in \S\ref{sec:impossibility}, this has previously been thought to be impossible to achieve \cite{Najafzadeh:2017vk,Najafzadeh:2018bw}.
     \item We formalise an abstract algorithm using Isabelle/HOL~\cite{DBLP:conf/tphol/WenzelPN08}, an interactive proof assistant based on higher-order logic, and obtain a computer-checked proof of correctness of the algorithm.
@@
 As these operations are so common, it is not obvious why a move operation should be difficult in a replicated setting.
-In this section we demonstrate by example some of the problems that arise with replicated trees, before proceeding to our solution in Section~\ref{sec:algorithm}.
+In this section we demonstrate by example some of the problems that arise with replicated trees, before proceeding to our solution in \S\ref{sec:algorithm}.
 
@@
 Another possible resolution is shown in Figure~\ref{fig:move-cycle}b: the nodes involved in the concurrent moves (and their children) could be duplicated, so that both ``$A$ as a child of $B$'' and ``$B$ as a child of $A$'' can exist in the tree.
-However, such duplication is undesirable for the same reasons as in Section~\ref{sec:move-same-item}.
+However, such duplication is undesirable for the same reasons as in \S\ref{sec:move-same-item}.
 
 In our opinion, the best way of handling the conflicting operations of Figure~\ref{fig:move-cycle} is to choose either Figure~\ref{fig:move-cycle}c or~\ref{fig:move-cycle}d: that is, either the result of applying replica 1's operation and ignoring replica 2's operation, or vice versa.
-Like in Section~\ref{sec:move-same-item}, the winning operation can be picked based on an operation timestamp.
+Like in \S\ref{sec:move-same-item}, the winning operation can be picked based on an operation timestamp.
 
@@
 
-Najafzadeh et al.~\cite{Najafzadeh:2017vk,Najafzadeh:2018bw} have previously implemented a replicated filesystem with a move operation, and analysed the case of concurrent move operations introducing a cycle, as in Section~\ref{sec:move-cycle}.
+Najafzadeh et al.~\cite{Najafzadeh:2017vk,Najafzadeh:2018bw} have previously implemented a replicated filesystem with a move operation, and analysed the case of concurrent move operations introducing a cycle, as in \S\ref{sec:move-cycle}.
 Using the CISE ('Cause I'm Strong Enough) proof tool~\cite{DBLP:conf/popl/GotsmanYFNS16,Najafzadeh:2016fi} the authors confirm that it is not sufficient for the replica that generates a move operation to check whether the operation introduces a cycle: like in Figure~\ref{fig:move-cycle}, two concurrent operations may be safe individually, but introduce a cycle when combined.
@@
 To our knowledge, our algorithm is the first to provide all of these properties simultaneously.
-We give a precise specification of our algorithm's consistency properties in Section~\ref{sec:proof}.
+We give a precise specification of our algorithm's consistency properties in \S\ref{sec:proof}.
 
@@
 The key consistency property of our algorithm is \emph{convergence}: that is, whenever any two replicas have applied the same set of operations, then they must be in the same state---even if the operations were applied in a different order on different replicas.
-We prove this in Section~\ref{sec:proof} by showing that applying operations is commutative.
+We prove this in \S\ref{sec:proof} by showing that applying operations is commutative.
 
@@
 We choose this language because it combines the conciseness of pseudocode with the precision of mathematical notation.
-It supports formal reasoning, allowing us to prove the correctness of the algorithm (Section~\ref{sec:proof}), and it also allows executable code to be exported to common functional programming languages such as Scala, Haskell, or OCaml.
+It supports formal reasoning, allowing us to prove the correctness of the algorithm (\S\ref{sec:proof}), and it also allows executable code to be exported to common functional programming languages such as Scala, Haskell, or OCaml.
 
@@
 
-Recall from Section~\ref{sec:move-cycle} that in order to prevent a cycle being introduced, the node being moved must not be an ancestor of the destination node.
+Recall from \S\ref{sec:move-cycle} that in order to prevent a cycle being introduced, the node being moved must not be an ancestor of the destination node.
 To implement this in Figure~\ref{fig:code}, we first define the \isa{ancestor} relation in lines 22--24.
@@
 This function takes as argument a pair consisting of a \isa{Move} operation and the current tree, and it returns a pair consisting of a \isa{LogMove} operation (which will be added to the log) and an updated tree.
-In line 29, the \isa{LogMove} record is constructed as described in Section~\ref{sec:state-log}, obtaining the prior parent and metadata of \isa{c} using the \isa{get\_parent} function.
+In line 29, the \isa{LogMove} record is constructed as described in \S\ref{sec:state-log}, obtaining the prior parent and metadata of \isa{c} using the \isa{get\_parent} function.
 
@@
         Since move operations do not specify the old location of a node, but only the new location, this sequential execution of concurrent operations is well-defined.
-    \item If two operations would introduce a cycle when combined, as in Section~\ref{sec:move-cycle}, then the operation with the greater timestamp is ignored.
+    \item If two operations would introduce a cycle when combined, as in \S\ref{sec:move-cycle}, then the operation with the greater timestamp is ignored.
         This is the case because \isa{do\_op} checks for cycles based on the tree created by all operations with a lower timestamp.
@@
 For example, we have explored CRDT algorithms for applications using a JSON data model~\cite{Kleppmann:2016ve,LocalFirst} and implemented them in a library called Automerge~\cite{Automerge}.
-We discuss distributed filesystems further in Section~\ref{sec:filesystems}.
+We discuss distributed filesystems further in \S\ref{sec:filesystems}.
 
@@
 
-We now discuss the correctness properties of the algorithm from Section~\ref{sec:algorithm}.
+We now discuss the correctness properties of the algorithm from \S\ref{sec:algorithm}.
 All theorems stated here have been formally proved and mechanically checked using Isabelle.
@@
 It takes a list of operations \isa{ops} and returns the state of a replica after it has applied all the operations in \isa{ops}.
-The \isa{apply\_ops} function works by starting in the initial state \isa{([], \{\})} consisting of the empty operation log \isa{[]} and the tree represented by the empty set \isa{\{\}}, and then applying the operations one by one to the state using the \isa{apply\_op} function (introduced in Section~\ref{sec:applying}).
+The \isa{apply\_ops} function works by starting in the initial state \isa{([], \{\})} consisting of the empty operation log \isa{[]} and the tree represented by the empty set \isa{\{\}}, and then applying the operations one by one to the state using the \isa{apply\_op} function (introduced in \S\ref{sec:applying}).
 The \isa{foldl} function from the Isabelle/HOL standard library performs the iteration over the list of operations.
@@
 This allows an application to move nodes between multiple trees, if desired.
-For example, the ``trash'' node used for deletion (Section~\ref{sec:ops-trees}) can be separate from the main tree.
+For example, the ``trash'' node used for deletion (\S\ref{sec:ops-trees}) can be separate from the main tree.
 
@@
 
-As discussed in Section~\ref{sec:applying}, we require that when replicas apply the same set of operations, they converge towards the same state, regardless of the order in which the operations are applied.
+As discussed in \S\ref{sec:applying}, we require that when replicas apply the same set of operations, they converge towards the same state, regardless of the order in which the operations are applied.
 We formalise this in Isabelle as follows:
@@
 Whilst these constructs are convenient for theorem proving, they do not translate well to efficient code.
-
 We therefore produce variants of the definitions in Figure~\ref{fig:code} that are designed for execution rather than theorem proving.
@@
 The \isa{hashable} type class on keys means that keys must have an associated hashing function.
-In effect, this hash-map is an index over the set of triples, using the fact that the parent and metadata for a given child are unique (Section~\ref{sec:unique-parent}).
+In effect, this hash-map is an index over the set of triples, using the fact that the parent and metadata for a given child are unique (\S\ref{sec:unique-parent}).
 We use the hash-map implementation from the Isabelle Collections Framework~\cite{DBLP:conf/itp/LammichL10}.
@@
 
-That is, if both algorithms are applied to the same set of operations then they produce identical logs and also produce trees that contain the same set of key-value bindings---i.e.
-the two trees are extensionally equivalent, despite having very different concrete representations.
-Moreover, we have the following corollaries of \isa{apply\_ops\_commutes} and \isa{apply\_ops\_acyclic}, respectively, further establishing the correctness of our executable algorithms:\footnote{The ``executable'' analogue of \isa{apply\_ops\_unique\_parent} is trivially true, as the output tree is stored as a hash-map, and therefore too trivial to state as a theorem.}:
+That is, if both algorithms are applied to the same operations then they produce identical logs and trees that contain the same set of bindings---i.e.
+the two trees are extensionally equivalent, despite having in-memory representations.
+Moreover, we have corollaries of \isa{apply\_ops\_commutes} and \isa{apply\_ops\_acyclic}, respectively, further establishing the correctness of our executable algorithms:\footnote{The ``executable'' analogue of \isa{apply\_ops\_unique\_parent} is trivially true, as the output tree is stored as a hash-map, and therefore too trivial to state as a theorem.}:
 
@@
         However, these algorithms only deal with insertion and deletion of tree nodes, and do not support a move operation.
-    \item As discussed in Section~\ref{sec:impossibility}, Najafzadeh et al.~\cite{Najafzadeh:2017vk,Najafzadeh:2018bw} propose two alternative implementations for a replicated filesystem: a CRDT in which conflicting moves are handled by duplicating tree nodes (as in Figures~\ref{fig:move-same-item}b and~\ref{fig:move-cycle}b), and a centralised implementation in which move operations must obtain a lock before executing (not a CRDT since it relies on synchronous coordination).
+    \item As discussed in \S\ref{sec:impossibility}, Najafzadeh et al.~\cite{Najafzadeh:2017vk,Najafzadeh:2018bw} propose two alternative implementations for a replicated filesystem: a CRDT in which conflicting moves are handled by duplicating tree nodes (as in Figures~\ref{fig:move-same-item}b and~\ref{fig:move-cycle}b), and a centralised implementation in which move operations must obtain a lock before executing (not a CRDT since it relies on synchronous coordination).
     \item Ahmed-Nacer et al.~\cite{AhmedNacer:2012us} survey approaches to handling conflicts on trees without providing concrete algorithms.
@@
         Some conflicts are handled by duplicating tree nodes.
-        Tao et al.\ also perform experiments with Dropbox, Google Drive, and OneDrive, similar to our experiments discussed in Section~\ref{sec:move-is-hard}.
+        Tao et al.\ also perform experiments with Dropbox, Google Drive, and OneDrive, similar to our experiments discussed in \S\ref{sec:move-is-hard}.
     \item In our previous unpublished work \cite{ExtendedVersion} we develop the foundations for the algorithm described in this paper.

### 2019-06-21 paper/move-op.tex
paper: more editing, space saving
More short lines rewritten away.  Down to ~1 paragraph above limit.

@@
 \begin{itemize}
-    \item The \textbf{filesystem} on most operating systems is organised as a tree: folders or directories are branch nodes, and files are leaf nodes.
-        In Unix-style filesystems, the nodes of the tree are known as \emph{inodes}.
-        Strictly speaking, filesystems that support hardlinks form a DAG, but we will focus on trees for now and show in \S\ref{sec:hardlinks} how to add support for links.
-    \item \textbf{Rich text editors} maintain a tree of textual structures such as paragraphs, bulleted and numbered lists, figures, sections, and so on.
-        These can be nested inside each other: for example, a paragraph may appear inside a bullet point of a bulleted list, which may in turn appear inside a numbered point of an enumeration.
+    \item Most \textbf{filesystems} are organised as a tree with directories as branch nodes and files as leaf nodes.
+        % In Unix-style filesystems, the nodes of the tree are known as \emph{inodes}.
+        Strictly speaking, filesystems with hardlinks form a DAG; we will focus on trees now and show in \S\ref{sec:hardlinks} how to add support for links.
+    \item \textbf{Rich text editors} maintain a tree of textual structures such as paragraphs, lists, figures, sections, and so on.
+        These can be nested: for example, a paragraph may appear inside a bulleted list item, which may in turn appear inside a numbered point of an enumeration.
         Rich text documents are typically represented in memory using the Document Object Model (DOM), and written to file in HTML, XML, or JSON format.
-    \item \textbf{Vector graphics} and \textbf{presentation software} represents images using graphical objects such as text boxes, rectangles, ellipses, lines, and so on.
+    \item \textbf{Vector graphics} and \textbf{presentation software} represents images using graphical objects such as text boxes, rectangles, lines, and so on.
         These objects are contained within nodes representing pages of a document, or slides of a presentation.
-        Multiple objects may be combined into a \emph{group} so that they can be manipulated as a unit; this corresponds to making these objects children of a common parent node.
-        Multiple objects and groups may in turn be combined into higher-level groups, forming a tree.
+        Objects may be combined into a \emph{group} and manipulated as a unit; this corresponds to making these objects children of a common parent node.
+        Multiple objects and groups may be combined further into higher-level groups, forming a tree.
     \item \textbf{Note-taking} and \textbf{task-management tools} such as Org-mode for Emacs~\cite{OrgMode} or OmniOutliner~\cite{OmniOutliner} present the user with a tree structure that they can inspect and manipulate.
@@
         As discussed in \S\ref{sec:impossibility}, this has previously been thought to be impossible to achieve \cite{Najafzadeh:2017vk,Najafzadeh:2018bw}.
-    \item We formalise an abstract algorithm using Isabelle/HOL~\cite{DBLP:conf/tphol/WenzelPN08}, an interactive proof assistant based on higher-order logic, and obtain a computer-checked proof of correctness of the algorithm.
+    \item We formalise an abstract algorithm using Isabelle/HOL~\cite{DBLP:conf/tphol/WenzelPN08} and obtain a computer-checked proof of correctness of the algorithm.
         In particular, we prove that arbitrary concurrent modifications to the tree can be merged such that all replicas converge towards a consistent state (strong eventual consistency) while keeping the data in a tree structure.
@@
 For this reason we do not need a separate operation type to create tree nodes: simply performing a move with a fresh child node \isa{c} is sufficient to create \isa{c}.
-Nor do we need to define a remove operation: instead, we can designate a particular tree node as the ``trash'', and perform deletion of a node \isa{c} by moving \isa{c} to be a child of the trash node.
+Nor do we need to define a remove operation: instead, we can designate a tree node as the ``trash'', and perform deletion of a node \isa{c} by moving \isa{c} to be a child of the trash node.
 
@@
 
-The intuition behind these functions is as follows: when a replica needs to apply an operation with timestamp \isa{t}, it first undoes the effect of any operations with a timestamp greater than \isa{t}, then performs the new operation, and finally re-applies the previously undone operations.
+Intuitively, when a replica needs to apply an operation with timestamp \isa{t}, it first undoes the effect of any operations with a timestamp greater than \isa{t}, then performs the new operation, and finally re-applies the previously undone operations.
 The result of this procedure is that the state of the tree is as if the operations had all been applied in order of increasing timestamp, even though in fact they might have been applied in any order.
@@
 For example, in a filesystem, two users could concurrently create files with the same name in the same directory.
-Our algorithm does not take any measures to prevent such conflict, but simply retains both child nodes.
-In practice, the filename collision would be resolved by making the filenames distinct, e.g.\ by appending a replica identifier to the filenames.
+Our algorithm does not prevent such a conflict but simply retains both child nodes.
+In practice, the collision would be resolved by making the filenames distinct, e.g.\ by appending a replica identifier to the filenames.
 
@@
 A tree is an acyclic graph in which every node has exactly one parent, except for the root, which has no parent.
-To prove that our algorithm always maintains a tree structure, no matter which operations are applied, we need to prove that these invariants hold.
-
-In fact, we slightly generalise this property: we allow more than one root to exist, so the graph is actually a forest.
-This allows an application to move nodes between multiple trees, if desired.
+To prove that our algorithm maintains a tree structure, no matter which operations are applied, we prove that these invariants hold.
+In fact, we slightly generalise this property and allow more than one root to exist, so the graph represents a forest, allowing an application to move nodes between multiple trees, if desired.
 For example, the ``trash'' node used for deletion (\S\ref{sec:ops-trees}) can be separate from the main tree.
@@
 
-To prove that \isa{apply\_op} preserves acyclicity, we first show that for any tree produced by \isa{apply\_ops}, there exists a sequence of operations such that applying those operations using \isa{do\_op} invocations alone (not using \isa{undo\_op} or \isa{redo\_op}) results in the same tree.
+To prove \isa{apply\_op} preserves acyclicity we show that trees obtained from \isa{apply\_ops} can be built from repeated use of \isa{do\_op} invocations alone, without \isa{undo\_op} or \isa{redo\_op}.
 (This is the case when operations are applied in order of increasing timestamp.)
-Then, by the result of the previous paragraph, any tree produced by \isa{apply\_ops} must be acyclic.
+Then, from our previous results, any tree produced by \isa{apply\_ops} is acyclic.
 
@@
 Gomes et al.~\cite{Gomes:2017gy} define a framework in Isabelle/HOL for proving the strong eventual consistency properties of CRDTs.
-Based on the commutativity proof above we integrate our tree datatype with this framework, and thus prove that our move operation on trees does indeed guarantee strong eventual consistency.
-The details of this proof are given in our open source Isabelle proof files~\cite{SourceFiles}.
+Using our commutativity proof above we integrate our tree datatype with this framework, and thus prove that our move operation on trees does indeed guarantee strong eventual consistency.
+We refer the reader to our Isabelle theory files~\cite{SourceFiles}.
 
@@
 The alternative to our optimistic replication approach would be to use a central server (master replica) or consensus protocol that processes all operations in a serialisable order.
-However, performing an operation on such a system requires waiting for synchronous communication with a server potentially located on a different continent, taking hundreds of milliseconds (Table~\ref{tab:rpc-times}); by contrast, a local operation taking a fraction of a millisecond is orders of magnitude faster.
+Then, performing an operation requires waiting for synchronous communication with a server potentially located on a different continent, taking hundreds of milliseconds (Table~\ref{tab:rpc-times}); by contrast, a local operation taking a fraction of a millisecond is orders of magnitude faster.
 Moreover, a local operation requires no waiting for network communication, and thus it supports disconnected operation: mobile devices can read and modify the replicated tree even while disconnected from the Internet, which is not the case with centralised systems.
@@
         However, these algorithms only deal with insertion and deletion of tree nodes, and do not support a move operation.
-    \item As discussed in \S\ref{sec:impossibility}, Najafzadeh et al.~\cite{Najafzadeh:2017vk,Najafzadeh:2018bw} propose two alternative implementations for a replicated filesystem: a CRDT in which conflicting moves are handled by duplicating tree nodes (as in Figures~\ref{fig:move-same-item}b and~\ref{fig:move-cycle}b), and a centralised implementation in which move operations must obtain a lock before executing (not a CRDT since it relies on synchronous coordination).
+    \item As discussed in \S\ref{sec:impossibility}, Najafzadeh et al.~\cite{Najafzadeh:2017vk,Najafzadeh:2018bw} propose two implementations for a replicated filesystem: a CRDT in which conflicting moves are handled by duplicating tree nodes (as in Figures~\ref{fig:move-same-item}b and~\ref{fig:move-cycle}b), and a centralised implementation in which move operations must obtain a lock before executing (not a CRDT since it relies on synchronous coordination).
     \item Ahmed-Nacer et al.~\cite{AhmedNacer:2012us} survey approaches to handling conflicts on trees without providing concrete algorithms.

### 2019-06-21 paper/move-op.tex
Paper editing

@@
 \begin{itemize}
-    \item Most \textbf{filesystems} are organised as a tree with directories as branch nodes and files as leaf nodes.
-        % In Unix-style filesystems, the nodes of the tree are known as \emph{inodes}.
-        Strictly speaking, filesystems with hardlinks form a DAG; we will focus on trees now and show in \S\ref{sec:hardlinks} how to add support for links.
+    \item The \textbf{filesystem} on most operating systems is a tree: directories are branch nodes, and files are leaf nodes.
+        In Unix systems the nodes of the tree are known as \emph{inodes}.
+        Strictly speaking, filesystems with hardlinks form a DAG; we will focus on trees now and show in \S\ref{sec:hardlinks} how to support links.
     \item \textbf{Rich text editors} maintain a tree of textual structures such as paragraphs, lists, figures, sections, and so on.
@@
         As discussed in \S\ref{sec:impossibility}, this has previously been thought to be impossible to achieve \cite{Najafzadeh:2017vk,Najafzadeh:2018bw}.
-    \item We formalise an abstract algorithm using Isabelle/HOL~\cite{DBLP:conf/tphol/WenzelPN08} and obtain a computer-checked proof of correctness of the algorithm.
+    \item We formalise an abstract algorithm using Isabelle/HOL~\cite{DBLP:conf/tphol/WenzelPN08}, an interactive proof assistant based on higher-order logic, and obtain a computer-checked proof of correctness of the algorithm.
         In particular, we prove that arbitrary concurrent modifications to the tree can be merged such that all replicas converge towards a consistent state (strong eventual consistency) while keeping the data in a tree structure.
-    \item To demonstrate the practical viability of our approach, we refine our abstract algorithm to an efficient implementation within Isabelle/HOL and prove the equivalence of the two.
-        We then extract a formally verified, executable Scala implementation from the Isabelle/HOL definitions and evaluate its performance with replicas across three continents.
+    \item To demonstrate the practical viability of our approach, we refine our abstract algorithm to an executable implementation within Isabelle/HOL and prove the equivalence of the two.
+        We then extract a formally verified Scala implementation from the Isabelle/HOL definitions and evaluate its performance with replicas across three continents.
     \item We perform experiments with replicated filesystem products such as Dropbox and Google Drive, and show that they exhibit problems that would be prevented by our algorithm.
@@
 Recall from \S\ref{sec:move-cycle} that in order to prevent a cycle being introduced, the node being moved must not be an ancestor of the destination node.
-To implement this in Figure~\ref{fig:code}, we first define the \isa{ancestor} relation in lines 22--24.
-This relation is the transitive closure of the parent-child relation of a tree: if \isa{(p, m, c) {\isasymin} tree} then \isa{p} is an ancestor of \isa{c} (line 23); moreover, if \isa{(p, m, c) {\isasymin} tree} and \isa{a} is an ancestor of \isa{p}, then \isa{a} is also an ancestor of \isa{c} (line 24).
+To implement this we first define the \isa{ancestor} relation in lines 22--24.
+It is the transitive closure of a tree's parent-child relation: if \isa{(p, m, c) {\isasymin} tree} then \isa{p} is an ancestor of \isa{c} (line 23); moreover, if \isa{a} is an ancestor of \isa{p} and \isa{(p, m, c) {\isasymin} tree}, then \isa{a} is also an ancestor of \isa{c} (line 24).
 The \textbf{inductive} keyword indicates that this recursive definition is iterated until the least fixed point is reached.
@@
 
-Intuitively, when a replica needs to apply an operation with timestamp \isa{t}, it first undoes the effect of any operations with a timestamp greater than \isa{t}, then performs the new operation, and finally re-applies the previously undone operations.
-The result of this procedure is that the state of the tree is as if the operations had all been applied in order of increasing timestamp, even though in fact they might have been applied in any order.
+When a replica needs to apply an operation with timestamp \isa{t}, it first undoes the effect of any operations with a timestamp greater than \isa{t}, then performs the new operation, and finally re-applies the undone operations.
+As a result, the state of the tree is as if the operations had all been applied in order of increasing timestamp, even though in fact they might have been applied in any order.
 
@@
 Otherwise we perform the operation using \isa{do\_op}, and add the corresponding \isa{LogMove} record as the head of the log (line 51).
+
 This logic ensures that the log is maintained in descending timestamp order, with the greatest timestamp at the head.
@@
 For example, in a filesystem, two users could concurrently create files with the same name in the same directory.
-Our algorithm does not prevent such a conflict but simply retains both child nodes.
+Our algorithm does not prevent such a conflict, but simply retains both child nodes.
 In practice, the collision would be resolved by making the filenames distinct, e.g.\ by appending a replica identifier to the filenames.
@@
 All theorems stated here have been formally proved and mechanically checked using Isabelle.
-For space reasons we give only a brief sketch of the proofs in this paper, but the full details can be found in our open source Isabelle files~\cite{SourceFiles}.
+For space reasons we give only a brief sketch of the proofs in this paper, but the full details can be found in our open source Isabelle theory files~\cite{SourceFiles}.
 
@@
 As we make no assumptions about \isa{ops}, this theorem holds for any replica state that can be reached by applying any number of operations.
-Proving it is easy because each of the \isa{do\_op}, \isa{undo\_op} and \isa{redo\_op} functions individually maintains the invariant \isa{unique\_parent tree}, and thus \isa{apply\_op} also maintains the invariant.
+Proving it is easy because the \isa{do\_op}, \isa{undo\_op} and \isa{redo\_op} functions individually maintain the invariant \isa{unique\_parent tree}, and thus \isa{apply\_op} also maintains the invariant.
 
@@
 
-To prove \isa{apply\_op} preserves acyclicity we show that trees obtained from \isa{apply\_ops} can be built from repeated use of \isa{do\_op} invocations alone, without \isa{undo\_op} or \isa{redo\_op}.
+Next, we show that for any tree produced by \isa{apply\_ops}, there exists a sequence of operations such that applying those operations using \isa{do\_op} invocations alone (not using \isa{undo\_op} or \isa{redo\_op}) results in the same tree.
 (This is the case when operations are applied in order of increasing timestamp.)
-Then, from our previous results, any tree produced by \isa{apply\_ops} is acyclic.
+Then, from the result of the previous paragraph, any tree produced by \isa{apply\_ops} is acyclic.
 
@@
 Using our commutativity proof above we integrate our tree datatype with this framework, and thus prove that our move operation on trees does indeed guarantee strong eventual consistency.
-We refer the reader to our Isabelle theory files~\cite{SourceFiles}.
+The details appear in our Isabelle theory files~\cite{SourceFiles}.
 
@@
 
-We use a synthetic workload wherein each replica generates move operations at the same rate; for each move operation the generating replica chooses parent and child nodes uniformly at random from a set of 1,000 tree nodes.
+We use a synthetic workload in which each replica generates move operations at the same rate; for each move operation the generating replica chooses parent and child nodes uniformly at random from a set of 1,000 tree nodes.
 For a given operation rate we run the system for 10 minutes to reach a steady state.
@@
 
-A locally generated operation always has a timestamp greater than any existing operation at the generating replica at that time (from the use of Lamport timestamps), and thus applying that operation requires only a \isa{do\_op}, and no \isa{undo\_op} or \isa{redo\_op}, hence the near-constant execution time.
+A locally generated operation always has a timestamp greater than any existing operation at the generating replica at that time (by definition of Lamport timestamps).
+Applying that operation requires only a \isa{do\_op}, and no \isa{undo\_op} or \isa{redo\_op}, hence the near-constant execution time.
 However, for remote operations, a number of undos and redos must be performed, depending on the degree of concurrency.
-As the rate of operations increases, the time between successive operations becomes smaller than the network delay between replicas; thus, there are more operations ``in flight'' at the same time, and more calls to \isa{undo\_op} and \isa{redo\_op} are required order operations by timestamp.
+As the rate of operations increases, the time between successive operations becomes smaller than the network delay between replicas; thus, there are more operations ``in flight'' at the same time, and more calls to \isa{undo\_op} and \isa{redo\_op} are required to order operations by timestamp.
 
@@
 
-A hand-optimised implementation of our algorithm could achieve higher throughput than our 24 operations/sec: despite \S\ref{subsect.extracting}, the code generated from our HOL definitions contains many optimisation opportunities due to its extraction from a theorem prover.
-Nevertheless, our evaluation sends a positive signal about the practical applicability of our algorithm, and the generated code could serve as an executable ``golden model'' for validating other implementations.
-
-Note that when a user is interacting with a system they only have to wait for local operations to execute; any remote operations can be applied in the background without affecting user interaction.
+A hand-optimised implementation of our algorithm could achieve higher throughput than our 24 operations/sec, since code generated from HOL definitions contains inefficiencies.
+Nevertheless, our evaluation sends a positive signal about the practical applicability of our algorithm.
+The generated code could also serve as a reference implementation against which other implementations could be tested.
+
+Note that when a user is interacting with a system, they only have to wait for local operations to execute; any remote operations can be applied in the background without affecting user interaction.
 Since our algorithm need not perform any undos or redos for local operations, these user interactions are consistently fast.

### 2019-06-21 paper/move-op.tex
paper: condense code generation section

@@
 Moreover, HOL's \isa{set} type allows infinite sets.
-Whilst these constructs are convenient for theorem proving, they do not translate well to efficient code.
+Whilst these constructs are convenient for theorem proving, they do not translate well to executable code.
+
 We therefore produce variants of the definitions in Figure~\ref{fig:code} that are designed for execution rather than theorem proving.
 Rather than representing the tree as a set of (parent, metadata, child) triples, these definitions use a hash-map in which the keys are child nodes, and the values are (metadata, parent) pairs, written in Isabelle as \isa{('n::\{hashable\}, 'm {\isasymtimes} 'n) hm}.
-The \isa{hashable} type class on keys means that keys must have an associated hashing function.
+The \isa{hashable} type class means that keys must have a hashing function.
 In effect, this hash-map is an index over the set of triples, using the fact that the parent and metadata for a given child are unique (\S\ref{sec:unique-parent}).
-We use the hash-map implementation from the Isabelle Collections Framework~\cite{DBLP:conf/itp/LammichL10}.
-
-We use a \emph{simulation relation} between our two representations of trees, and say that a hash-map $\isa{t}$ of type $(\isa{'n}, \isa{'m} \times \isa{'n})\ \isa{hm}$ \emph{simulates} a set $T$ of type $(\isa{'n} \times \isa{'m} \times \isa{'n})\ \isa{set}$ whenever
-\begin{displaymath}
-\isa{hm.lookup}\ c\ t = \isa{Some}\ (m, p) \quad\text{iff}\quad (p, m, c) \in T \quad\text{forall $c$, $m$, and $p$}
-\end{displaymath}
-where $hm.lookup$ searches for a binding for key $\isa{c}$ in hash-map $t$, returning $\isa{Some}\ (m, p)$ when a binding from $c$ to $(m, p)$ exists in the hash-map, or $\isa{None}$ otherwise.
-Write $t \preceq T$ to assert that $t$ and $T$ are contained within the simulation relation.
-Note some corollaries follow immediately: e,g,~we have $t \preceq T$ implies $\isa{unique\_parent}\ \isa{T}$.
-
-We produce a new set of definitions---mirroring the structure of our existing ones---wherein we replace all occurrences of the $(\isa{'n} \times \isa{'m} \times \isa{'n})\ \isa{set}$ type, and any associated set operations, with the $(\isa{'n}, \isa{'m} \times \isa{'n})\ \isa{hm}$ type and associated hash-map operations.
-For clarity, we refer to any definition using hash-maps as an ``executable'' definition, and any definition making use of the \isa{set} type as an ``abstract'' definition.
-For most definitions this is a mechanical process, and we elide further discussion of the required changes, referring the interested reader to our Isabelle/HOL source code.
-However \isa{ancestor} is different to other definitions as it is an inductively defined relation rather than a recursive function, and we need further work to make this executable.
-We characterise this relation with the following:
-
-% DPM: add efficient_ancestor_simp here
+The hash-map implementation is from the Isabelle Collections Framework~\cite{DBLP:conf/itp/LammichL10}.
+
+We say a hash-map \isa{t} of type \isa{('n, 'm {\isasymtimes} 'n) hm} \emph{simulates} a set \isa{T} of type \isa{('n {\isasymtimes} 'm {\isasymtimes} 'n) set} when they have the same entries:
 \begin{isabelle}
-\isacommand{theorem}\isamarkupfalse%
-\ executable{\isacharunderscore}ancestor{\isacharunderscore}simp\ {\isacharbrackleft}code{\isacharbrackright}{\isacharcolon}\isanewline
-\ \ \isakeyword{shows}\ {\isacartoucheopen}executable{\isacharunderscore}ancestor\ t\ p\ c\ {\isasymlongleftrightarrow}\isanewline
-\ \ \ \ \ \ \ \ \ \ {\isacharparenleft}case\ hm{\isachardot}lookup\ c\ t\ of\isanewline
-\ \ \ \ \ \ \ \ \ \ \ \ \ \ None\ {\isasymRightarrow}\ False\isanewline
-\ \ \ \ \ \ \ \ \ \ \ \ {\isacharbar}\ Some\ {\isacharparenleft}m{\isacharcomma}\ a{\isacharparenright}\ {\isasymRightarrow}\isanewline
-\ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ a\ {\isacharequal}\ p\ {\isasymor}\ executable{\isacharunderscore}ancestor\ t\ p\ a{\isacharparenright}{\isacartoucheclose}
+\isacommand{definition}\isamarkupfalse%
+\ simulates\ \isakeyword{where}\ {\isacartoucheopen}simulates\ t\ T\ {\isasymequiv}\isanewline
+\ \ {\isacharparenleft}{\isasymforall}p\ m\ c{\isachardot}\ hm{\isachardot}lookup\ c\ t\ {\isacharequal}\ Some\ {\isacharparenleft}m{\isacharcomma}\ p{\isacharparenright}\ {\isasymlongleftrightarrow}\ {\isacharparenleft}p{\isacharcomma}\ m{\isacharcomma}\ c{\isacharparenright}\ {\isasymin}\ T{\isacharparenright}{\isacartoucheclose}
 \end{isabelle}
-
-Note that \isa{executable\_ancestor\_simp} is tagged with a $\isa{code}$ ``attribute'', a hint that this theorem should be used as a transformation during code generation.
-Isabelle/HOL takes this theorem as the definition, for code generation, of the \isa{executable\_ancestor} relation and is able to translate this to a recursive function definition in each target backend---this transformation is ``semantics preserving'', as the characterisation was proved as a theorem within HOL.
-
-We demonstrate the correctness of our executable definitions against our abstract definitions.
-The properties proved are best described using an example, as they all follow a similar pattern.
-We use the lemma \isa{executable\_do\_op\_simulates} as our exemplar:
-
-% DPM: add efficient_do_op_simulates here
+\noindent where \isa{hm.lookup c t} looks up the key \isa{c} in the hash-map \isa{t}.
+We can now prove the equivalence of the set-based and the hash-map-based versions of each function.
+For example, \isa{executable\_do\_op} is the equivalent of the \isa{do\_op} function based on a hash-map:
 \begin{isabelle}
@@
 \ executable{\isacharunderscore}do{\isacharunderscore}op{\isacharunderscore}simulates{\isacharcolon}\isanewline
-\ \ \isakeyword{assumes}\ {\isadigit{1}}{\isacharcolon}\ {\isacartoucheopen}t\ {\isasympreceq}\ T{\isacartoucheclose}\isanewline
-\ \ \ \ \isakeyword{and}\ {\isadigit{2}}{\isacharcolon}\ {\isacartoucheopen}executable{\isacharunderscore}do{\isacharunderscore}op\ {\isacharparenleft}oper{\isacharcomma}\ t{\isacharparenright}\ {\isacharequal}\ {\isacharparenleft}log{\isadigit{1}}{\isacharcomma}\ u{\isacharparenright}{\isacartoucheclose}\isanewline
-\ \ \ \ \isakeyword{and}\ {\isadigit{3}}{\isacharcolon}\ {\isacartoucheopen}do{\isacharunderscore}op\ {\isacharparenleft}oper{\isacharcomma}\ T{\isacharparenright}\ {\isacharequal}\ {\isacharparenleft}log{\isadigit{2}}{\isacharcomma}\ U{\isacharparenright}{\isacartoucheclose}\isanewline
-\ \ \isakeyword{shows}\ {\isacartoucheopen}log{\isadigit{1}}\ {\isacharequal}\ log{\isadigit{2}}\ {\isasymand}\ u\ {\isasympreceq}\ U{\isacartoucheclose}
+\ \ \isakeyword{assumes}\ {\isacartoucheopen}simulates\ t\ T{\isacartoucheclose}\isanewline
+\ \ \ \ \isakeyword{and}\ {\isacartoucheopen}executable{\isacharunderscore}do{\isacharunderscore}op\ {\isacharparenleft}oper{\isacharcomma}\ t{\isacharparenright}\ {\isacharequal}\ {\isacharparenleft}log{\isadigit{1}}{\isacharcomma}\ u{\isacharparenright}{\isacartoucheclose}\isanewline
+\ \ \ \ \isakeyword{and}\ {\isacartoucheopen}do{\isacharunderscore}op\ {\isacharparenleft}oper{\isacharcomma}\ T{\isacharparenright}\ {\isacharequal}\ {\isacharparenleft}log{\isadigit{2}}{\isacharcomma}\ U{\isacharparenright}{\isacartoucheclose}\isanewline
+\ \ \isakeyword{shows}\ {\isacartoucheopen}log{\isadigit{1}}\ {\isacharequal}\ log{\isadigit{2}}\ {\isasymand}\ simulates\ u\ U{\isacartoucheclose}
 \end{isabelle}
 
-We show that the \isa{executable\_do\_op} and the \isa{do\_op} functions are in ``lock-step'', mapping related trees $t \preceq T$ to related trees $u \preceq U$ and producing identical logs of operations whenever they are applied to the same input operation, \isa{oper}.
-This lock-step is repeated bottom-up for every executable-abstract algorithm pairs, until we reach the final correctness theorem for \isa{apply\_ops}:
-
-%DPM: add efficient_apply_ops_simulates here
+Similarly, we can show this simulation relation holds for each function in our algorithm, all the way to \isa{apply\_ops}:
 \begin{isabelle}
@@
 \ executable{\isacharunderscore}apply{\isacharunderscore}ops{\isacharunderscore}simulates{\isacharcolon}\isanewline
-\ \ \isakeyword{assumes}\ {\isadigit{1}}{\isacharcolon}\ {\isacartoucheopen}executable{\isacharunderscore}apply{\isacharunderscore}ops\ opers\ {\isacharequal}\ {\isacharparenleft}log{\isadigit{1}}{\isacharcomma}\ u{\isacharparenright}{\isacartoucheclose}\isanewline
-\ \ \ \ \isakeyword{and}\ {\isadigit{2}}{\isacharcolon}\ {\isacartoucheopen}apply{\isacharunderscore}ops\ opers\ {\isacharequal}\ {\isacharparenleft}log{\isadigit{2}}{\isacharcomma}\ U{\isacharparenright}{\isacartoucheclose}\isanewline
-\ \ \isakeyword{shows}\ {\isacartoucheopen}log{\isadigit{1}}\ {\isacharequal}\ log{\isadigit{2}}\ {\isasymand}\ u\ {\isasympreceq}\ U{\isacartoucheclose}
+\ \ \isakeyword{assumes}\ {\isacartoucheopen}executable{\isacharunderscore}apply{\isacharunderscore}ops\ ops\ {\isacharequal}\ {\isacharparenleft}log{\isadigit{1}}{\isacharcomma}\ t{\isacharparenright}{\isacartoucheclose}\isanewline
+\ \ \ \ \isakeyword{and}\ {\isacartoucheopen}apply{\isacharunderscore}ops\ ops\ {\isacharequal}\ {\isacharparenleft}log{\isadigit{2}}{\isacharcomma}\ T{\isacharparenright}{\isacartoucheclose}\isanewline
+\ \ \isakeyword{shows}\ {\isacartoucheopen}log{\isadigit{1}}\ {\isacharequal}\ log{\isadigit{2}}\ {\isasymand}\ simulates\ t\ T{\isacartoucheclose}
 \end{isabelle}
 
-That is, if both algorithms are applied to the same operations then they produce identical logs and trees that contain the same set of bindings---i.e.
-the two trees are extensionally equivalent, despite having in-memory representations.
-Moreover, we have corollaries of \isa{apply\_ops\_commutes} and \isa{apply\_ops\_acyclic}, respectively, further establishing the correctness of our executable algorithms:\footnote{The ``executable'' analogue of \isa{apply\_ops\_unique\_parent} is trivially true, as the output tree is stored as a hash-map, and therefore too trivial to state as a theorem.}:
-
-% DPM: add efficient_apply_ops_commutes and efficient_apply_ops_acyclic here
-\begin{isabelle}
-\isacommand{theorem}\isamarkupfalse%
-\ executable{\isacharunderscore}apply{\isacharunderscore}ops{\isacharunderscore}commutes{\isacharcolon}\isanewline
-\ \ \isakeyword{assumes}\ {\isadigit{1}}{\isacharcolon}\ {\isacartoucheopen}set\ ops{\isadigit{1}}\ {\isacharequal}\ set\ ops{\isadigit{2}}{\isacartoucheclose}\isanewline
-\ \ \ \ \isakeyword{and}\ {\isadigit{2}}{\isacharcolon}\ {\isacartoucheopen}distinct\ {\isacharparenleft}map\ move{\isacharunderscore}time\ ops{\isadigit{1}}{\isacharparenright}{\isacartoucheclose}\isanewline
-\ \ \ \ \isakeyword{and}\ {\isadigit{3}}{\isacharcolon}\ {\isacartoucheopen}distinct\ {\isacharparenleft}map\ move{\isacharunderscore}time\ ops{\isadigit{2}}{\isacharparenright}{\isacartoucheclose}\isanewline
-\ \ \ \ \isakeyword{and}\ {\isadigit{4}}{\isacharcolon}\ {\isacartoucheopen}executable{\isacharunderscore}apply{\isacharunderscore}ops\ ops{\isadigit{1}}\ {\isacharequal}\ {\isacharparenleft}log{\isadigit{1}}{\isacharcomma}\ t{\isacharparenright}{\isacartoucheclose}\isanewline
-\ \ \ \ \isakeyword{and}\ {\isadigit{5}}{\isacharcolon}\ {\isacartoucheopen}executable{\isacharunderscore}apply{\isacharunderscore}ops\ ops{\isadigit{2}}\ {\isacharequal}\ {\isacharparenleft}log{\isadigit{2}}{\isacharcomma}\ u{\isacharparenright}{\isacartoucheclose}\isanewline
-\ \ \isakeyword{shows}\ {\isacartoucheopen}log{\isadigit{1}}\ {\isacharequal}\ log{\isadigit{2}}\ {\isasymand}\ hm{\isachardot}lookup\ c\ t\ {\isacharequal}\ hm{\isachardot}lookup\ c\ u{\isacartoucheclose}\isanewline
-\isanewline
-\isacommand{theorem}\isamarkupfalse%
-\ executable{\isacharunderscore}apply{\isacharunderscore}ops{\isacharunderscore}acyclic{\isacharcolon}\isanewline
-\ \ \isakeyword{assumes}\ {\isadigit{1}}{\isacharcolon}\ {\isacartoucheopen}executable{\isacharunderscore}apply{\isacharunderscore}ops\ ops\ {\isacharequal}\ {\isacharparenleft}log{\isacharcomma}\ t{\isacharparenright}{\isacartoucheclose}\isanewline
-\ \ \isakeyword{shows}\ {\isacartoucheopen}{\isasymnexists}n{\isachardot}\ executable{\isacharunderscore}ancestor\ t\ n\ n{\isacartoucheclose}
-\end{isabelle}
-
-Note the statement of \isa{executable\_apply\_ops\_commutes} differs from \isa{apply\_ops\_commutes}---the latter asserts identical sets are obtained from convergence, whilst the former asserts that reading any key binding in $t$ gives the same result as a reading in $u$.
-Whilst this is weaker than that of \isa{apply\_ops\_acyclic}, it is the strongest statement one can hope for: the resulting hash-maps may contain the same key-value bindings but may not be structurally equal, depending on the ordering of operations.
+That is, if \isa{executable\_apply\_ops} and \isa{apply\_ops} are applied to the same list of operations then they produce identical logs, and also produce trees that contain the same set of key-value bindings---i.e.
+the two trees are extensionally equivalent, despite having very different in-memory representations.
+Moreover, we can prove corollaries of \isa{apply\_ops\_commutes} and \isa{apply\_ops\_acyclic}, respectively, for the hash-map-based implementation.
+%The analogue of \isa{apply\_ops\_unique\_parent} holds trivially since by construction, the hash-map contains at most one (metadata, parent) pair for a given child.
 

### 2019-06-21 paper/move-op.tex
paper: edit sections 5 and 6.1

@@
 
-\subsection{Performance experiments}
-
-\subsubsection{Experimental setup and results}
+\subsection{Experimental setup and results}
 
@@
 
-\subsubsection{Interpretation of results}
+\subsection{Interpretation of results}
 
@@
 However, for remote operations, a number of undos and redos must be performed, depending on the degree of concurrency.
-As the rate of operations increases, the time between successive operations becomes smaller than the network delay between replicas; thus, there are more operations ``in flight'' at the same time, and more calls to \isa{undo\_op} and \isa{redo\_op} are required to order operations by timestamp.
+As the rate of operations increases, the time interval between successive operations becomes smaller than the network delay between replicas; thus, there are more operations ``in flight'' at the same time, and more calls to \isa{undo\_op} and \isa{redo\_op} are required to order operations by timestamp.
 
@@
 A hand-optimised implementation of our algorithm could achieve higher throughput than our 24 operations/sec, since code generated from HOL definitions contains inefficiencies.
+The generated code could serve as a reference implementation against which other implementations could be tested.
+
 Nevertheless, our evaluation sends a positive signal about the practical applicability of our algorithm.
-The generated code could also serve as a reference implementation against which other implementations could be tested.
-
 Note that when a user is interacting with a system, they only have to wait for local operations to execute; any remote operations can be applied in the background without affecting user interaction.
@@
 The alternative to our optimistic replication approach would be to use a central server (master replica) or consensus protocol that processes all operations in a serialisable order.
-Then, performing an operation requires waiting for synchronous communication with a server potentially located on a different continent, taking hundreds of milliseconds (Table~\ref{tab:rpc-times}); by contrast, a local operation taking a fraction of a millisecond is orders of magnitude faster.
+Performing an operation then requires waiting for synchronous communication with a server potentially located on a different continent, taking hundreds of milliseconds (Table~\ref{tab:rpc-times}); by contrast, a local operation taking a fraction of a millisecond is orders of magnitude faster.
 Moreover, a local operation requires no waiting for network communication, and thus it supports disconnected operation: mobile devices can read and modify the replicated tree even while disconnected from the Internet, which is not the case with centralised systems.
@@
 
-Besides CRDTs, another family of algorithms for concurrent modification of data structures is called \emph{Operational Transformation}~\cite{Sun:1998vf}.
-Several authors have defined collaboratively editable tree structures using the Operational Transformation approach \cite{Jungnickel:2016cb,Ignat:2003jy,Davis:2002iv}, but they only consider insertion and deletion of tree nodes, and do not support a move operation.
-
-Molli et al.~\cite{Molli:2003cd} define an Operational Transformation tree structure with a move operation.
+Besides CRDTs, another family of algorithms for concurrent modification of data structures is \emph{Operational Transformation} (OT)~\cite{Sun:1998vf}.
+Several authors have defined concurrent tree structures using OT \cite{Jungnickel:2016cb,Ignat:2003jy,Davis:2002iv}, but they only consider insertion and deletion of nodes, and do not support a move operation.
+
+Molli et al.~\cite{Molli:2003cd} define an OT tree structure with a move operation.
 However, it requires that all communication between replicas is performed via total order broadcast, which in turn requires the use of a sequencer server or a consensus algorithm~\cite{Chandra:1996cp}.

### 2019-06-21 paper/move-op.tex
Updated version of distributed filesystems.

--- 2019-06-21 paper/move-op.tex
@@
 
+Coda is a distributed client-server file system which allows clients to cache or \emph{hoard} copies of files otherwise stored in a shared data repository~\cite{kistler1992coda}. 
+Clients can edit copies of data held in the cache while offline, during which time a kernel module keeps track of all updates. 
+When the client comes back online, the client attempts to resynchronise changes with the repository. 
+This design means that conflicts may arise if multiple clients edit the same files or directories while one or more clients are offline. 
+One proposed solution to resolving conflicts in files is to support application-specific resolvers which ``encapsulate the knowledge needed for file resolution''~\cite{Kumar:1995wf}, an approach similar to that found in Bayou~\cite{Terry:1995dn}.
+Concurrent edits on directory entries across different replicas in Coda have been considered, but the authors note that they do ``not address transparent resolution of cross-directory [move operations] in [their] current implementation''~\cite{kumar1993log}.
+Furthermore, while the authors consider a number of conflicts associated with directory move operations, they do not highlight the potential for the creation of cycles as seen in Figure~\ref{fig:move-cycle} (a).
+
+The Ficus replicated file system~\cite{Reiher:1994wh} was a in-kernel SunOS-based replicated peer-to-peer file system. 
+Ficus supported updates to replicas during periods of network partition and claims ``conflicting updates to directories are detected and automatically repaired''~\cite{guy1990implementation}. 
+Unfortunately we were unable to find a precise definition of the algorithm used in any of the available publications. 
+Rumor~\cite{Guy:1999gy} is the successor to Ficus. 
+An important distinction between Rumor and previous work is that it operates at the application layer and needs to be invoked periodically by the user or by a deamon process. 
+The original version of Rumor was unable to scale beyond 20 replicas, however an extension called Roam~\cite{Ratner:1999fh} allowed an order of magnitude better scaling. 
+We obtained the user manual and source code for Rumor from \url{archive.org}, however after modest effort, we were unable to get the software running.
+%The user manual (in Archive) is quite helpful for Rumor: https://web.archive.org/web/20170705140958/ftp://ftp.cs.ucla.edu/pub/rumor/rumor_users_manual.ps.
+Note that this also makes the claim that "Unix directories are an important example of a kind of file for which Rumor automatically resolves conflicts"
+
 Unison is a file synchronisation tool with a formal specification~\cite{PierceVouillon:UnisonSpecTR} which allows two replicas, such as laptops, desktops or servers, to synchronise the state of a directory.
 Unison permits offline updates to both replicas.
-The (possibly divergent) state of two replicas are resynchronised by running the Unison tool on the two replicas. 
-Unison works by recording a summary record of the state of the filesystem on each replica,  inspecting this summary whenever the unison program is invoked, and sharing any update information with the other replica in order to determine the changes made to the filesystems on each replica since they were last synchronised.
-This approach has the benefit that no special kernel extension or daemon process is required; instead Unison is an application program which inspects the current state of the filesystem whenever it is run.
+Like Rumor, Unison also works at the application layer and supports the resynchronisation of two (possibly divergent) replicas by executing a copy of Unison on both nodes. 
+Unison records a summary record of the state of the filesystem on each replica whenever it is run.
+By inspecting this summary whenever the Unison program is invoked, and by sharing any update information with the other replica, both replicas can determine the changes made to the filesystems on each replica since they were last synchronised.
+This approach has the benefit that no special kernel extension or daemon process is required; instead the user invokes the Unison program whenever synchronisation is needed.
 When presented with the move operations described in Figure~\ref{fig:move-same-item}, Unison duplicates the files, resulting in the outcome shown in Figure~\ref{fig:move-same-item}~(a). 
@@
 
-Coda is a distributed client-server file system which allows clients to cache or \emph{hoard} copies of files otherwise stored in a shared data repository~\cite{kistler1992coda}. 
-Clients can edit copies of data held in the cache while offline, during which time a kernel module keeps track of all updates. 
-When the client comes back online, the client attempts to resynchronise changes with the repository. 
-This design means that conflicts may arise if multiple clients edit the same files or directories while one or more clients are offline. 
-One proposed solution to resolving conflicts in files is to support application-specific resolvers which ``encapsulate the knowledge needed for file resolution''~\cite{Kumar:1995wf}, an approach is similar to that found in Bayou~\cite{Terry:1995dn}.
-Concurrent edits on directory entries across different replicas in Coda have been considered, but the authors note that they do ``not address transparent resolution of cross-directory [move operations] in [their] current implementation.''~\cite{kumar1993log}.
-Furthermore, while the authors consider a number of conflicts associated with directory move operations, they do not highlight the potential for the creation of cycles as seen in Figure~\ref{fig:move-cycle} (a).
-
-The Ficus replicated file system~\cite{Reiher:1994wh} is a in-kernel SunOS-based replicated peer-to-peer file system. 
-Ficus supported updates to replicas during periods of network partition and claims ``conflicting updates to directories are detected an automatically repaired''~\cite{guy1990implementation}. 
-Unfortunately we were unable to find a precise definition of the algorithm used in any of the available publications. 
-Rumor~\cite{Guy:1999gy} is the successor to Ficus. 
-An important distinction is between Rumor and previous work is that it operates at the application layer and needs to be invoked periodically by the user or a deamon process.
-Rumor is therefore very similar in operation and design to Unison described earlier.
-
-%The user manual (in Archive) is quite helpful for Rumor: https://web.archive.org/web/20170705140958/ftp://ftp.cs.ucla.edu/pub/rumor/rumor_users_manual.ps.
-Note that this also makes the claim that "Unix directories are an important example of a kind of file for which Rumor automatically resolves conflicts"
-
-TODO: Roam~\cite{Ratner:1999fh}
-
 \section{Conclusions}
 
-TODO
+TODO.
 

### 2019-06-21 paper/move-op.tex
A conclusion.

@@
 
-TODO.
+In this paper we have defined a novel algorithm that handles arbitrary concurrent modifications on a tree data structure in a peer-to-peer  setting.
+Our approach ensures that all replicas converge to the same consistent state and the data structure remains a tree; in other words, no cycles are introduced.
+Updates made to a local replica appear immediately, while operations from remote replicas can be applied in the background without affecting user interaction.
+This approach means user interaction is consistently fast, even in the face of unbounded communication delays between replicas.
+
+We formally proved the correctness of our algorithm in the Isabelle/HOL theorem prover and evaluated the performance of our formally verified implementation between three replicas located in California, Ireland and Singapore.
+Our implementation is able to apply local updates in under 1~ms, orders of magnitude faster than is possible with a central server (master replica) or consensus protocol operating over these distances.
+The current implementation is able to sustain up to 24 remote updates per second. 
+Unfortunately rates above 24 remote updates per second quickly generate unbounded delays and hence failure since the implementation is unable to keep up with the number of undo and redos required by the algorithm.
+We hypothesise this limitation is because the current implementation uses Isabelle/HOL's code generation mechanism to extract executable Scala code which is not optimised. 
+We plan to develop a significantly more scaleable implementation as part of future work.
 

### 2019-06-21 paper/move-op.tex
paper: tighten up section 6.2

@@
 
-Coda is a distributed client-server file system which allows clients to cache or \emph{hoard} copies of files otherwise stored in a shared data repository~\cite{kistler1992coda}. 
-Clients can edit copies of data held in the cache while offline, during which time a kernel module keeps track of all updates. 
-When the client comes back online, the client attempts to resynchronise changes with the repository. 
-This design means that conflicts may arise if multiple clients edit the same files or directories while one or more clients are offline. 
-One proposed solution to resolving conflicts in files is to support application-specific resolvers which ``encapsulate the knowledge needed for file resolution''~\cite{Kumar:1995wf}, an approach similar to that found in Bayou~\cite{Terry:1995dn}.
-Concurrent edits on directory entries across different replicas in Coda have been considered, but the authors note that they do ``not address transparent resolution of cross-directory [move operations] in [their] current implementation''~\cite{kumar1993log}.
-Furthermore, while the authors consider a number of conflicts associated with directory move operations, they do not highlight the potential for the creation of cycles as seen in Figure~\ref{fig:move-cycle} (a).
-
-The Ficus replicated file system~\cite{Reiher:1994wh} was a in-kernel SunOS-based replicated peer-to-peer file system. 
-Ficus supported updates to replicas during periods of network partition and claims ``conflicting updates to directories are detected and automatically repaired''~\cite{guy1990implementation}. 
+Coda is a client-server filesystem that allows clients to locally cache copies of files stored in a server-side data repository~\cite{kistler1992coda}. 
+Clients can edit data in the cache while offline, during which time a kernel module keeps track of all updates. 
+When the client comes back online it attempts to resynchronise changes with the server. 
+To resolve conflicts due to concurrent updates, Coda uses application-specific resolvers~\cite{Kumar:1995wf}, similarly to Bayou's approach~\cite{Terry:1995dn}.
+Concurrent renaming and move operations have been considered, but the authors note that they do ``not address transparent resolution of cross-directory renames [i.e.\ move operations] in [their] current implementation''~\cite{kumar1993log}.
+Furthermore, while the authors consider a number of conflicts associated with directory move operations, they do not highlight the potential for the creation of cycles.
+
+Ficus~\cite{Reiher:1994wh} is an in-kernel SunOS-based replicated peer-to-peer filesystem. 
+Ficus supports updates to replicas during periods of network partition and claims ``conflicting updates to directories are detected and automatically repaired''~\cite{guy1990implementation}. 
 Unfortunately we were unable to find a precise definition of the algorithm used in any of the available publications.
-Rumor~\cite{Guy:1999gy} is the successor to Ficus. 
-An important distinction between Rumor and previous work is that it operates at the application layer and needs to be invoked periodically by the user or by a deamon process. 
-The original version of Rumor was unable to scale beyond 20 replicas, however an extension called Roam~\cite{Ratner:1999fh} allowed an order of magnitude better scaling. 
-We obtained the user manual and source code for Rumor from \url{archive.org}, however after modest effort, we were unable to get the software running.
-%The user manual (in Archive) is quite helpful for Rumor: https://web.archive.org/web/20170705140958/ftp://ftp.cs.ucla.edu/pub/rumor/rumor_users_manual.ps.
-Note that this also makes the claim that "Unix directories are an important example of a kind of file for which Rumor automatically resolves conflicts"
-
-Unison is a file synchronisation tool with a formal specification~\cite{PierceVouillon:UnisonSpecTR} which allows two replicas, such as laptops, desktops or servers, to synchronise the state of a directory.
-Unison permits offline updates to both replicas.
-Like Rumor, Unison also works at the application layer and supports the resynchronisation of two (possibly divergent) replicas by executing a copy of Unison on both nodes. 
-Unison records a summary record of the state of the filesystem on each replica whenever it is run.
-By inspecting this summary whenever the Unison program is invoked, and by sharing any update information with the other replica, both replicas can determine the changes made to the filesystems on each replica since they were last synchronised.
-This approach has the benefit that no special kernel extension or daemon process is required; instead the user invokes the Unison program whenever synchronisation is needed.
-When presented with the move operations described in Figure~\ref{fig:move-same-item}, Unison duplicates the files, resulting in the outcome shown in Figure~\ref{fig:move-same-item}~(a). 
-Unison is unable to automatically synchronise the move operations shown in Figure~\ref{fig:move-cycle} and instead asks the user to confirm whether changes made to directories A and B should be propagated or deleted.
-As a result, depending on user choice, both replicas can end up in one of four states, including those shown in Figure~\ref{fig:move-cycle}~(b), (c) or (d) or with an empty root directory on both replicas.
-In all cases, Unison successfully ensures both replicas are in the same state after synchronisation.
+
+Rumor~\cite{Guy:1999gy,RumorManual} is the successor to Ficus.
+While previous work uses the kernel filesystem interface, Rumor is a userspace process that is invoked periodically by the user or by a deamon; when run, it compares the state of the replicas.
+The original version of Rumor was unable to scale beyond 20 replicas, but an extension called Roam~\cite{Ratner:1999fh} allowed better scaling. 
+In an attempt to test Rumor's conflict handling we obtained the source code from archive.org~\cite{RumorSource}; however, we were unable to get it running after modest effort.
+%The user manual \cite{RumorManual} makes the claim that "Unix directories are an important example of a kind of file for which Rumor automatically resolves conflicts"
+
+Unison is a file synchronisation tool with a formal specification~\cite{PierceVouillon:UnisonSpecTR} that allows two replicas to synchronise the state of a directory.
+It permits offline updates to both replicas.
+Like Rumor, Unison is a userspace process that works by comparing replica states.
+Whenever it is run, Unison records a summary of the filesystem state on each replica, and it uses this summary to determine the changes made since the last synchronisation. 
+When presented with the move operations described in Figure~\ref{fig:move-same-item}, Unison duplicates the files, resulting in the outcome shown in Figure~\ref{fig:move-same-item}a. 
+Unison is unable to automatically synchronise the move operations shown in Figure~\ref{fig:move-cycle} and instead asks the user to choose one of four possible resolutions: those shown in Figure~\ref{fig:move-cycle}b, \ref{fig:move-cycle}c, \ref{fig:move-cycle}d, or to delete both directories.
 

### 2019-06-21 paper/move-op.tex
paper: edit conclusions

@@
 
+Many distributed filesystems, such as NFS, rely on synchronous interaction with a server.
+While this has the advantage of not requiring conflict resolution, it rules out users working while offline.
+
 Coda is a client-server filesystem that allows clients to locally cache copies of files stored in a server-side data repository~\cite{kistler1992coda}. 
@@
 
-In this paper we have defined a novel algorithm that handles arbitrary concurrent modifications on a tree data structure in a peer-to-peer  setting.
-Our approach ensures that all replicas converge to the same consistent state and the data structure remains a tree; in other words, no cycles are introduced.
-Updates made to a local replica appear immediately, while operations from remote replicas can be applied in the background without affecting user interaction.
-This approach means user interaction is consistently fast, even in the face of unbounded communication delays between replicas.
-
-We formally proved the correctness of our algorithm in the Isabelle/HOL theorem prover and evaluated the performance of our formally verified implementation between three replicas located in California, Ireland and Singapore.
-Our implementation is able to apply local updates in under 1~ms, orders of magnitude faster than is possible with a central server (master replica) or consensus protocol operating over these distances.
-The current implementation is able to sustain up to 24 remote updates per second. 
-Unfortunately rates above 24 remote updates per second quickly generate unbounded delays and hence failure since the implementation is unable to keep up with the number of undo and redos required by the algorithm.
-We hypothesise this limitation is because the current implementation uses Isabelle/HOL's code generation mechanism to extract executable Scala code which is not optimised. 
-We plan to develop a significantly more scaleable implementation as part of future work.
-
-% Disconnected operation
-% - No Synchronisation or consensus required. Available in the sense of CAP
-% - Rely only on uniqueness of node IDs, e.g. inode numbers
-% - Composing JSON/XML tree by using list datatype
-% - Location is Lww register. Concurrent move does not duplicate
-% - Unlike NFS, can work offline 
-% - Garbage collection: discard causally stable prefix
+In this paper we have defined a novel algorithm that handles arbitrary concurrent modifications of a tree data structure---adding, moving, and removing nodes---in a peer-to-peer setting.
+It is applicable to distributed filesystems and many other applications that use a tree-structured data model.
+Our approach ensures that all replicas converge to the same consistent state without requiring any manual conflict resolution, and without needing application developers to implement conflict handling logic.
+Updates made to a local replica take effect immediately, while operations from remote replicas can be applied in the background.
+This approach means that user interaction is consistently fast, even in the face of unbounded communication delays between replicas, e.g.\ during disconnected operation of mobile devices.
+
+While existing systems exhibit anomalies (such as duplicating nodes or introducing cycles) or error messages (like in Figure~\ref{fig:gdrive-error}) in the face of concurrent move operations, our algorithm is free from such problems.
+We formally verified the correctness of our algorithm using the Isabelle/HOL proof assistant; our theorems show that replicas can apply operations in any order, and that the result is always a valid tree (nodes have at most one parent, and the graph does not contain any cycles).
+
+We also extracted a formally verified Scala implementation from our Isabelle/HOL definitions, and evaluated its performance across three replicas located in California, Ireland and Singapore.
+Our implementation is able to apply local updates in under 1~ms, orders of magnitude faster than is possible with a central server or consensus protocol operating over these distances.
+The current implementation is able to sustain up to 24 updates per second; we plan to develop a more scalable implementation in future work.
 

### 2019-06-22 paper/move-op.tex
paper: spellcheck

@@
 While these solutions are valid, the authors go on to claim that ``no file system can support an unsynchronised move without anomalies, such as loss or duplication''~\cite{Najafzadeh:2018bw}.
-We revisit that claim in this paper: our algorithm does not perform any locking, coordination or synchronisation among replicas, but it nevertheless ensures that the tree invariants are always satisified (in particular, it never introduces cycles), and it never duplicates or loses any tree nodes.
+We revisit that claim in this paper: our algorithm does not perform any locking, coordination or synchronisation among replicas, but it nevertheless ensures that the tree invariants are always satisfied (in particular, it never introduces cycles), and it never duplicates or loses any tree nodes.
 To our knowledge, our algorithm is the first to provide all of these properties simultaneously.
@@
 Further assume \isa{acyclic tree1}.
-We can then prove by contradiction that \isa{tree2} is also acylic.
+We can then prove by contradiction that \isa{tree2} is also acyclic.
 Assume to the contrary that \isa{tree2} contains a cycle.
@@
 Above 25 operations/sec the CPU utilisation of our virtual machines grows to the point that incoming network requests from other replicas start queueing before they are processed.
-When a remote operation spends time queueing, more concurrent operations have been applied by the time it is processed, and thus more undos and redos are required to apply the incoming operation, which in turn increases the CPU utilization and further increases queueing.
+When a remote operation spends time queueing, more concurrent operations have been applied by the time it is processed, and thus more undos and redos are required to apply the incoming operation, which in turn increases the CPU utilisation and further increases queueing.
 This self-reinforcing effect explains the sudden dramatic increase in Figure~\ref{fig:plots}.
@@
 Rumor~\cite{Guy:1999gy,RumorManual} is the successor to Ficus.
-While previous work uses the kernel filesystem interface, Rumor is a userspace process that is invoked periodically by the user or by a deamon; when run, it compares the state of the replicas.
+While previous work uses the kernel filesystem interface, Rumor is a userspace process that is invoked periodically by the user or by a daemon; when run, it compares the state of the replicas.
 The original version of Rumor was unable to scale beyond 20 replicas, but an extension called Roam~\cite{Ratner:1999fh} allowed better scaling. 

### 2019-06-22 paper/move-op.tex
paper: last few edits, ready to submit

@@
 Two users who believe they are collaborating on the same file may in fact be editing two different copies, which will then become inconsistent with each other.
+In the rich text editor and presentation software examples, such duplication is also undesirable.
 
@@
 
-When the user makes a change to the tree, that change is expressed by generating an operation on their local replica.
-We assume the existence of a messaging middleware that copies operations between replicas, so that eventually every operation is applied on every replica, provided that network partitions are of a finite duration.
-This middleware may deliver operations in any order.
-Operations received from a remote replica are applied using the same algorithm as operations generated by a local user.
+When the user wants to make a change to the tree, they generate an operation object and apply it on their local replica.
+Every operation is also asynchronously sent over the network to all other replicas, and applied on every remote replica using the same algorithm as for local operations.
+The network may deliver operations in any order; thus, the communication can be performed peer-to-peer, and it does not require any central server or consensus protocol.
+We only assume that eventually every operation is applied on every replica, provided that network partitions are of a finite duration.
 
@@
 It then returns a \isa{('n {\isasymtimes} 'm) option}.
-\isa{{\isasymexists}!x} means ``there exists a unique value \isa{x} such that\dots'', while \isa{THE x}\dots means ``\emph{choose} the unique value \isa{x} such that\ldots''.
+The operator \isa{{\isasymexists}!x} means ``there exists a unique value \isa{x} such that\dots'', while \isa{THE x}\dots means ``\emph{choose} the unique value \isa{x} such that\ldots''.
 
@@
 \begin{itemize}
-    \item If two operations concurrently move the same node, then the ``winning'' operation is the one with the higher timestamp.
-        Effectively, the operation with the lower timestamp moves the node first, and then the operation with the higher timestamp moves it again.
+    \item If two operations concurrently move the same node, then the ``winning'' operation is the one with the greater timestamp.
+        Effectively, the operation with the lower timestamp moves the node first, and then the operation with the greater timestamp moves it again, so the final parent is determined by the latter.
         Since move operations do not specify the old location of a node, but only the new location, this sequential execution of concurrent operations is well-defined.
@@
 The algorithm as specified in Figure~\ref{fig:code} retains operations in the log indefinitely, so the memory use grows without bound.
-However, in practice it is easy to truncate the log, because \isa{apply\_op} only examines the prefix of operations whose timestamp is greater than that of the operation being applied.
-Thus, once it is known that all future operations will have a timestamp greater than $t$, then all operations with timestamp $t$ or less can be discarded from the log.
+However, in practice it is easy to truncate the log, because \isa{apply\_op} only examines the log prefix of operations whose timestamp is greater than that of the operation being applied.
+Thus, once it is known that all future operations will have a timestamp greater than $t$, then operations with timestamp $t$ or less can be discarded from the log.
 In this case, we say that $t$ is \emph{causally stable}~\cite{Baquero:2014ed}.
 
-We can determine causal stability in a system where the set of replicas is known, where each replica generates operations with monotonically increasing timestamps, and where the communication link between any pair of replicas is FIFO (messages are received in the order in which they are sent, as implemented e.g.
-by TCP).
-In this case, each replica can keep track of the most recent timestamp it has seen from all replicas (including its own), and the minimum of these timestamps is the causally stable threshold.
+We can determine causal stability in a system where the set of replicas is known, where each replica generates operations with monotonically increasing timestamps, and where the communication link between any pair of replicas is FIFO (messages are received in the order in which they are sent, as implemented e.g.\ by TCP).
+In this case, we can keep track of the most recent timestamp we have seen from each replica (including our own), and the minimum of these timestamps is the causally stable threshold.
 
@@
 
-To reason about the state of a replica we define the function \isa{apply\_ops} on lines 53--54 of Figure~\ref{fig:code}.
+To reason about the state of a replica we first define the function \isa{apply\_ops} on lines 53--54 of Figure~\ref{fig:code}.
 It takes a list of operations \isa{ops} and returns the state of a replica after it has applied all the operations in \isa{ops}.
@@
 
-\isa{apply\_ops\_commutes} then follows from the following lemma, which proves that two calls of \isa{apply\_op} commute, provided that the timestamps in the operations and in the log are all distinct:
+\noindent \isa{apply\_ops\_commutes} then follows from the following lemma, which proves that two calls of \isa{apply\_op} commute, provided that the timestamps in the operations and in the log are all distinct:
 \begin{isabelle}
@@
         Then neither applying \isa{oper1} nor applying \isa{oper2} will undo \isa{logop}.
-        Thus, the two operations are applied in the same way as the base case.
+        Thus, the two operations are applied in the same way as in the base case.
     \item \isa{t1 < t3 < t2}.
@@
 \end{isabelle}
-\noindent where \isa{hm.lookup c t} looks up the key \isa{c} in the hash-map \isa{t}.
+\noindent where \isa{hm.lookup c t} looks up the key \isa{c} in the hash-map \isa{t}, returning \isa{Some x} if \isa{c} maps to the value \isa{x}, and returning \isa{None} if \isa{c} does not appear in the hash-map.
 We can now prove the equivalence of the set-based and the hash-map-based versions of each function.
@@
 However, it requires that all communication between replicas is performed via total order broadcast, which in turn requires the use of a sequencer server or a consensus algorithm~\cite{Chandra:1996cp}.
-Our algorithm has better availability characteristics in the presence of network partitions because it allows messages to be delivered in any order.
+Our algorithm has better availability characteristics in the presence of network partitions because it allows messages to be delivered in any order, e.g.\ via peer-to-peer protocols.
 
@@
 
-Unison is a file synchronisation tool with a formal specification~\cite{PierceVouillon:UnisonSpecTR} that allows two replicas to synchronise the state of a directory.
+Unison is a file synchronisation tool with a formal specification that allows two replicas to synchronise the state of a directory~\cite{PierceVouillon:UnisonSpecTR}.
 It permits offline updates to both replicas.
@@
 
-In this paper we have defined a novel algorithm that handles arbitrary concurrent modifications of a tree data structure---adding, moving, and removing nodes---in a peer-to-peer setting.
+In this paper we have defined a novel algorithm that handles arbitrary concurrent modifications of a tree data structure---adding, moving, and removing nodes---in a peer-to-peer replication setting.
 It is applicable to distributed filesystems and many other applications that use a tree-structured data model.
 Our approach ensures that all replicas converge to the same consistent state without requiring any manual conflict resolution, and without needing application developers to implement conflict handling logic.
-Updates made to a local replica take effect immediately, while operations from remote replicas can be applied in the background.
+Updates made to a local replica take effect immediately, while operations from remote replicas can be propagated and applied in the background.
 This approach means that user interaction is consistently fast, even in the face of unbounded communication delays between replicas, e.g.\ during disconnected operation of mobile devices.
@@
 
-We also extracted a formally verified Scala implementation from our Isabelle/HOL definitions, and evaluated its performance across three replicas located in California, Ireland and Singapore.
-Our implementation is able to apply local updates in under 1~ms, orders of magnitude faster than is possible with a central server or consensus protocol operating over these distances.
-The current implementation is able to sustain up to 24 updates per second; we plan to develop a more scalable implementation in future work.
+We also extracted a formally verified Scala implementation from our Isabelle/HOL definitions, and evaluated its performance across three replicas in California, Ireland and Singapore.
+Our implementation applies local updates in under 1~ms, orders of magnitude faster than is possible with a central server or consensus protocol operating over these distances.
+It is able to sustain 24 updates per second; we plan to develop a more scalable implementation in future work.
 

### 2019-10-29 paper/move-op.tex
More references

@@
         Rich text documents are typically represented in memory using the Document Object Model (DOM), and written to file in HTML, XML, or JSON format.
-    \item \textbf{Vector graphics} and \textbf{presentation software} represents images using graphical objects such as text boxes, rectangles, lines, and so on.
+    \item \textbf{Vector graphics} and \textbf{presentation software} represents images using graphical objects such as text boxes, rectangles, lines, and so on~\cite{Wallace:2019vf}.
         These objects are contained within nodes representing pages of a document, or slides of a presentation.
@@
 
+Hughes et al.~\cite{Hughes:2016fp} test Dropbox and Google Drive against a formal specification, but they do not consider moving files, and thus do not find the issue described in \S~\ref{sec:move-cycle}.
+
+Bj{\o}rner~\cite{Bjorner:2007hp} discusses the development of the Distributed File System Replication (DFS-R) component of Windows Server, during which a model checker found an issue with concurrent moves similar to Figure~\ref{fig:move-cycle}a.
+Bj{\o}rner outlines several possible solutions, but notes that model-checking their algorithm was not feasible due to state space explosion.
+Our use of proof by induction, rather than model-checking, allows us to verify the correctness of our algorithm in unbounded executions.
+
 \section{Conclusions}

### 2019-10-29 paper/move-op.tex
Formatting for EuroSys
Condense pages 1-6

@@
-\documentclass[sigconf]{acmart}
+\documentclass[sigplan,anonymous]{acmart}
+
+% EuroSys recommended formatting
+\renewcommand\footnotetextcopyrightpermission[1]{}
+\pagestyle{plain}
 
@@
 \begin{abstract}
-    Replicated tree data structures are found in distributed filesystems, such as Google Drive and Dropbox, and applications with a JSON or XML data model.
+    Replicated tree data structures are a fundamental building block of distributed filesystems, such as Google Drive and Dropbox, and collaborative applications with a JSON or XML data model.
     These systems need to support a \emph{move} operation that allows a subtree to be moved to a new location within the tree.
-    Such a move operation is easy to implement in a centralised system, e.g.\ with a designated primary replica, but it is difficult in settings where different replicas can concurrently perform arbitrary move operations.
-    For example, we find that Google Drive and Dropbox fail to correctly synchronise files when users on different devices concurrently perform certain move operations.
-    In this paper we present an algorithm that handles arbitrary concurrent modifications on trees, while ensuring that the tree structure remains valid (in particular, no cycles are introduced), and guaranteeing that all replicas eventually converge towards the same consistent state.
-    We formally prove the correctness of our algorithm using the Isabelle/HOL interactive proof assistant, and we evaluate the performance of our formally verified implementation in a geo-replicated setting to demonstrate its viability in practice.
+    However, such a move operation is difficult to implement correctly if different replicas can concurrently perform arbitrary move operations, and we demonstrate bugs in Google Drive and Dropbox that arise with concurrent moves.
+    In this paper we present an algorithm that handles arbitrary concurrent modifications on trees, while ensuring that the tree structure remains valid (in particular, no cycles are introduced), and guaranteeing that all replicas converge towards the same consistent state.
+    We formally prove the correctness of our algorithm using the Isabelle/HOL proof assistant, and we evaluate the performance of our formally verified implementation in a geo-replicated setting.
 \end{abstract}
+
+\settopmatter{printfolios=true} % Recommended EuroSys formatting
 \maketitle
@@
         In Unix systems the nodes of the tree are known as \emph{inodes}.
-        Strictly speaking, filesystems with hardlinks form a DAG; we will focus on trees now and show in \S\ref{sec:hardlinks} how to support links.
-    \item \textbf{Rich text editors} maintain a tree of textual structures such as paragraphs, lists, figures, sections, and so on.
-        These can be nested: for example, a paragraph may appear inside a bulleted list item, which may in turn appear inside a numbered point of an enumeration.
-        Rich text documents are typically represented in memory using the Document Object Model (DOM), and written to file in HTML, XML, or JSON format.
-    \item \textbf{Vector graphics} and \textbf{presentation software} represents images using graphical objects such as text boxes, rectangles, lines, and so on~\cite{Wallace:2019vf}.
+        Strictly speaking, filesystems with hardlinks form a DAG; we will focus on trees for now and show in \S\ref{sec:hardlinks} how to support links.
+    \item \textbf{Rich text editors} maintain a HTML or XML structure: a tree of textual elements such as paragraphs, lists, figures, sections, and so on.
+        These can be nested: for example, a paragraph may appear inside a bulleted list item, which may in turn appear inside an enumeration.
+    \item \textbf{Vector graphics} and \textbf{presentation software} represent images using graphical objects such as text boxes, rectangles, lines, and so on~\cite{Wallace:2019vf}.
         These objects are contained within nodes representing pages of a document, or slides of a presentation.
@@
 Such systems have the advantage that they can continue processing read and write requests even in the presence of arbitrary network partitions; in other words, they are \emph{available} and \emph{partition-tolerant} in the sense of the CAP theorem~\cite{Gilbert:2002il}.
-This approach is desirable since it enables disconnected operation in a mobile computing context, and high availability in geo-replicated settings.
-
-As the tree structure may be concurrently modified on different replicas, the state of these replicas may temporarily diverge.
-In this paper we show how the replicas can nevertheless achieve \emph{strong eventual consistency}~\cite{Shapiro:2011un,Gomes:2017gy}: as the replicas communicate, we guarantee that they will converge towards a consistent state.
+This approach is desirable since it enables disconnected operation on mobile devices, and high availability in geo-replicated settings.
+
+As the tree structure is concurrently modified on different replicas, the state of these replicas temporarily diverges.
+In this paper we show how we can nevertheless achieve \emph{strong eventual consistency}~\cite{Shapiro:2011un,Gomes:2017gy}: as the replicas communicate, we guarantee that they converge to a consistent state.
 Our algorithm for achieving convergence is an example of a \emph{Conflict-free Replicated Data Type} or CRDT~\cite{Shapiro:2011wy,Shapiro:2011un}.
@@
 We explain in \S\ref{sec:move-is-hard} why this move operation is so challenging.
-In summary, our contributions are:
+Our contributions are:
 \begin{itemize}
@@
         As discussed in \S\ref{sec:impossibility}, this has previously been thought to be impossible to achieve \cite{Najafzadeh:2017vk,Najafzadeh:2018bw}.
-    \item We formalise an abstract algorithm using Isabelle/HOL~\cite{DBLP:conf/tphol/WenzelPN08}, an interactive proof assistant based on higher-order logic, and obtain a computer-checked proof of correctness of the algorithm.
-        In particular, we prove that arbitrary concurrent modifications to the tree can be merged such that all replicas converge towards a consistent state (strong eventual consistency) while keeping the data in a tree structure.
-    \item To demonstrate the practical viability of our approach, we refine our abstract algorithm to an executable implementation within Isabelle/HOL and prove the equivalence of the two.
-        We then extract a formally verified Scala implementation from the Isabelle/HOL definitions and evaluate its performance with replicas across three continents.
-    \item We perform experiments with replicated filesystem products such as Dropbox and Google Drive, and show that they exhibit problems that would be prevented by our algorithm.
+    \item We formalise the algorithm using Isabelle/HOL \cite{DBLP:conf/tphol/WenzelPN08}, a proof assistant based on higher-order logic, and obtain a computer-checked proof of correctness.
+        In particular, we prove that arbitrary concurrent modifications to the tree can be merged such that all replicas converge to a consistent state (strong eventual consistency) while keeping the data in a tree structure.
+    \item To demonstrate the practical viability of our approach, we refine the algorithm to an executable implementation within Isabelle/HOL and prove the equivalence of the two.
+        We then extract a formally verified Scala implementation from Isabelle and evaluate its performance with replicas across three continents.
+    \item We perform experiments with Dropbox and Google Drive, and show that they exhibit problems that would be prevented by our algorithm.
 \end{itemize}
@@
       \useasboundingbox (-1.3,-1.3) rectangle (1,1.2);
-      \node [fill=red!15] at (-1.12,1.05) {(\emph{a})};
+      \node [fill=red!15] at (-1.10,1.03) {(\emph{a})};
       \node at (0,1) {$\mathsf{root}$};
@@
       \useasboundingbox (-1.15,-2.3) rectangle (1.15,0.2);
-      \node [fill=red!15] at (0.97,0.05) {(\emph{b})};
+      \node [fill=red!15] at (0.94,0.03) {(\emph{b})};
       \tikzstyle{level 1}=[sibling distance=12mm]
@@
       \useasboundingbox (-1.17,-2.3) rectangle (1.15,0.2);
-      \node [fill=red!15] at (-1.00,0.05) {(\emph{c})};
+      \node [fill=red!15] at (-0.98,0.03) {(\emph{c})};
       \node {$\mathsf{root}$} child {node {$A$} child {node {$B$}} child {node {$C$}}};
@@
       \useasboundingbox (-1.15,-2.3) rectangle (1.15,0.2);
-      \node [fill=red!15] at (0.96,0.05) {(\emph{d})};
+      \node [fill=red!15] at (0.93,0.03) {(\emph{d})};
       \node {$\mathsf{root}$} child {node {$B$} child {node {$A$} child {node {$C$}}}};
@@
   \includegraphics[width=\textwidth,keepaspectratio=true]{gdrive-error.png}
-  \caption{Error message produced by Google Drive Backup and Sync on Mac OS as a result of performing the operations shown in Figure~\ref{fig:move-cycle}.}
+  \caption{Error message produced by Google Drive Backup and Sync on Mac OS as a result of performing the operations shown in Figure~\ref{fig:move-cycle}, indicating a bug in the underlying sync implementation.}
   \label{fig:gdrive-error}
@@
 
-Applications that rely on a tree data model often need to move a node from one location to another location within the tree, whereby all children of the moved node move along with it.
-For example:
+Applications that rely on a tree data model often need to move a node from one location to another location within the tree, such that all of its children move along with it:
 \begin{itemize}
@@
     \item In a rich text editor, a paragraph can be turned into a bullet point.
-        In the DOM tree this corresponds to creating a new bullet list node, creating a new bullet point node within it, and then moving the paragraph node inside the bullet point.
-    \item In presentation software, grouping two graphical objects corresponds to creating a new group node, and then moving the two objects to be children of the new group node.
+        In the XML tree this corresponds to creating new list and bullet point nodes, and then moving the paragraph node inside the bullet point.
+    \item In presentation software, grouping two graphical objects corresponds to creating a new group node, and then moving the two objects into the new group node.
 \end{itemize}
 As these operations are so common, it is not obvious why a move operation should be difficult in a replicated setting.
-In this section we demonstrate by example some of the problems that arise with replicated trees, before proceeding to our solution in \S\ref{sec:algorithm}.
+In this section we demonstrate some problems that arise with replicated trees, before proceeding to our solution in \S\ref{sec:algorithm}.
 
@@
 
-The first difficulty arises when a node is concurrently moved into different locations on different replicas.
+The first difficulty arises when the same node is concurrently moved into different locations on different replicas.
 This scenario is illustrated in Figure~\ref{fig:move-same-item}, where replica 1 moves node $A$ to be a child of $B$, while concurrently replica 2 moves $A$ to be a child of $C$.
@@
 
-Rather than duplicating the moved subtree, another possible resolution is for the destination locations of both moves to refer to the same node, as shown in Figure~\ref{fig:move-same-item}b.
+Another possible resolution is for the destination locations of both moves to refer to the same node, as shown in Figure~\ref{fig:move-same-item}b.
 However, the result is a DAG, not a tree.
-POSIX-compliant filesystems do not allow this outcome, since they allow hardlinks only to files, but not to directories.
+POSIX filesystems do not allow this outcome, since they do not allow hardlinks to directories.
 
@@
 Which one of these two is picked is arbitrary, due to the symmetry between the two replicas.
-For example, the ``winning'' location could be picked based on a timestamp in the operations, similarly to the ``last writer wins'' conflict resolution method of Thomas's write rule~\cite{Johnson:1975we}.
-All replicas pick the same operation as winner, and effectively ignore the conflicting operation.
+The ``winning'' location could be picked based on a timestamp in the operations, similarly to the ``last writer wins'' conflict resolution method of Thomas's write rule~\cite{Johnson:1975we}.
 (The \emph{timestamp} in this context need not come from a physical clock; it could also be logical, such as a Lamport timestamp~\cite{Lamport:1978jq}.)
 
-We tested this scenario with file sync products Dropbox and Google Drive by concurrently moving the same directory to two different destination directories.\footnote{Experiment setup: we installed the official Mac OS clients for Dropbox and Google Drive on two computers, logged into the same Dropbox/Google accounts, and configured them to each sync a directory on the local filesystem with the corresponding cloud service.
-To test concurrent operations, we turned off the network interface on both computers, performed a move operation on the local filesystem of each computer, then reconnected both computers to the Internet and waited for them to sync.}
+We tested this scenario with file sync products Dropbox and Google Drive by concurrently moving the same directory to two different destination directories.\footnote{Experiment setup: we installed the official Mac OS clients for Dropbox and Google Drive on two computers, logged into the same Dropbox/Google accounts, and configured them to sync a directory on the local filesystem.
+To test concurrent operations, we disconnected both computers from the Internet, performed a move operation on the local filesystem of each computer, then reconnected and waited for them to sync.}
 Dropbox exhibited the undesirable duplication behaviour of Figure~\ref{fig:move-same-item}a, while the outcome on Google Drive was as in Figure~\ref{fig:move-same-item}c/d.
@@
 
-In an unreplicated tree it is easy to prevent cycles being introduced: if the node being moved is an ancestor of the destination node, then the operation is rejected.
-However, a new problem arises in a replicated setting: different replicas may perform operations that are individually safe, but whose combination leads to a cycle.
-
-One example of such a scenario is illustrated in Figure~\ref{fig:move-cycle}.
+In an unreplicated tree it is easy to prevent cycles being introduced: if the node being moved is an ancestor of the destination node, the operation is rejected.
+However, in a replicated setting, different replicas may perform operations that are individually safe, but whose combination leads to a cycle.
+One such example is illustrated in Figure~\ref{fig:move-cycle}.
 Here, replica 1 moves $B$ to be a child of $A$, while concurrently replica 2 moves $A$ to be a child of $B$.
-As each replica propagates its operation to the other replica, a careless implementation might end up in the state shown in Figure~\ref{fig:move-cycle}a, in which $A$ and $B$ have become locked in a cycle and detached from the tree.
+As each replica propagates its operation to the other replica, a careless implementation might end up in the state shown in Figure~\ref{fig:move-cycle}a, in which $A$ and $B$ have formed a cycle, detached from the tree.
 
@@
 In our opinion, the best way of handling the conflicting operations of Figure~\ref{fig:move-cycle} is to choose either Figure~\ref{fig:move-cycle}c or~\ref{fig:move-cycle}d: that is, either the result of applying replica 1's operation and ignoring replica 2's operation, or vice versa.
-Like in \S\ref{sec:move-same-item}, the winning operation can be picked based on an operation timestamp.
+Like in \S\ref{sec:move-same-item}, the winning operation can be picked based on a timestamp.
 
@@
 In Google Drive, one replica was able to successfully sync with the server, while the other replica displayed the ``unknown error'' message shown in Figure~\ref{fig:gdrive-error}.
-The replica in an error state refused to sync the conflicting directory, and its filesystem state remained inconsistent with the other replica.
+The replica in an error state refused to sync the conflicting directory, and its filesystem state remained permanently inconsistent with the other replica.
 This error state persisted until the directories on the erroring replica were manually moved to match the state of the other replica.
@@
 
-\subsection{The impossibility of a highly-available move operation}\label{sec:impossibility}
-
-Najafzadeh et al.~\cite{Najafzadeh:2017vk,Najafzadeh:2018bw} have previously implemented a replicated filesystem with a move operation, and analysed the case of concurrent move operations introducing a cycle, as in \S\ref{sec:move-cycle}.
-Using the CISE ('Cause I'm Strong Enough) proof tool~\cite{DBLP:conf/popl/GotsmanYFNS16,Najafzadeh:2016fi} the authors confirm that it is not sufficient for the replica that generates a move operation to check whether the operation introduces a cycle: like in Figure~\ref{fig:move-cycle}, two concurrent operations may be safe individually, but introduce a cycle when combined.
+\subsection{A highly-available move operation: impossible?}\label{sec:impossibility}
+
+Najafzadeh et al.~\cite{Najafzadeh:2017vk,Najafzadeh:2018bw} previously implemented a replicated filesystem with a move operation, and analysed the case of concurrent move operations introducing a cycle.
+Using the CISE proof tool~\cite{DBLP:conf/popl/GotsmanYFNS16,Najafzadeh:2016fi} the authors confirm that it is not sufficient for the replica that generates a move operation to check whether the operation introduces a cycle: like in Figure~\ref{fig:move-cycle}, two concurrent operations may be safe individually, but introduce a cycle when combined.
 
@@
 propose two solutions to this problem: either to duplicate tree nodes, as in Figure~\ref{fig:move-cycle}b, or to execute a synchronous locking protocol that prevents two move operations from concurrently modifying the same part of the tree.
-The downside of a locking protocol is that the move operation is no longer highly available in the presence of network partitions, since it must wait for synchronous communication with other replicas.
+The downside of a locking protocol is that the move operation is no longer highly available in the presence of network partitions, since it must wait for synchronous communication with other replicas or a lock server.
 
 While these solutions are valid, the authors go on to claim that ``no file system can support an unsynchronised move without anomalies, such as loss or duplication''~\cite{Najafzadeh:2018bw}.
-We revisit that claim in this paper: our algorithm does not perform any locking, coordination or synchronisation among replicas, but it nevertheless ensures that the tree invariants are always satisfied (in particular, it never introduces cycles), and it never duplicates or loses any tree nodes.
+We refute that claim in this paper: our algorithm does not perform any locking, coordination or synchronisation among replicas, but it nevertheless ensures that the tree invariants are always satisfied (in particular, it never introduces cycles), and it never duplicates or loses any tree nodes.
 To our knowledge, our algorithm is the first to provide all of these properties simultaneously.
@@
 
-We now introduce our algorithm for a replicated tree data structure that supports a move operation.
+We now introduce our algorithm for a replicated tree that supports a move operation.
 We model each replica as a state machine that transitions from one state to the next by applying an operation.
@@
 
-When the user wants to make a change to the tree, they generate an operation object and apply it on their local replica.
+When the user wants to make a change to the tree, they generate an operation and apply it on their local replica.
 Every operation is also asynchronously sent over the network to all other replicas, and applied on every remote replica using the same algorithm as for local operations.
@@
 
-Figure~\ref{fig:code} gives the full source code for our algorithm, implemented using the Isabelle/HOL language~\cite{DBLP:conf/tphol/WenzelPN08}.
+Figure~\ref{fig:code} gives the full source code for our algorithm in the Isabelle/HOL language~\cite{DBLP:conf/tphol/WenzelPN08}.
 We choose this language because it combines the conciseness of pseudocode with the precision of mathematical notation.
-It supports formal reasoning, allowing us to prove the correctness of the algorithm (\S\ref{sec:proof}), and it also allows executable code to be exported to common functional programming languages such as Scala, Haskell, or OCaml.
-
-In this section we walk through the code in Figure~\ref{fig:code} step by step, explaining features of the Isabelle/HOL syntax as we encounter them.
+It supports formal reasoning, allowing us to prove the correctness of the algorithm (\S\ref{sec:proof}), and it can be exported to Scala, Haskell, or OCaml.
+
+In this section we walk through the code in Figure~\ref{fig:code} step by step, explaining the Isabelle/HOL syntax as we encounter it.
 Additional background documentation is available~\cite{DBLP:books/sp/NipkowK14}.
@@
 
-We define only one kind of operation: \isa{Move t p m c} (Figure~\ref{fig:code}, lines 1--5).
-A move operation is a record consisting of four fields: a timestamp \isa{t} of type \isa{'t}, a parent node \isa{p} of type \isa{'n}, a metadata field \isa{m} of type \isa{'m}, and a child node \isa{c} of type \isa{'n}.
-Here, \isa{'t}, \isa{'n} and \isa{'m} are \emph{type variables} that can be substituted for arbitrary types; our only requirement is that timestamps \isa{'t} are globally unique and have an associated total order (e.g.
-Lamport timestamps~\cite{Lamport:1978jq} would suffice).
-
-The meaning of an operation \isa{Move t p m c} is that at time \isa{t}, the child node \isa{c} is moved to be a child of parent node \isa{p}.
-The operation does not specify the old location of \isa{c}; the algorithm simply ``steals'' \isa{c} from wherever it is currently located in the tree, and moves it to \isa{p}.
+We define one kind of operation: \isa{Move t p m c} (line 1--5).
+A move operation is a 4-tuple consisting of a timestamp \isa{t} of type \isa{'t}, a parent node ID \isa{p} of type \isa{'n}, a metadata field \isa{m} of type \isa{'m}, and a child node ID \isa{c} of type \isa{'n}.
+Here, \isa{'t}, \isa{'n} and \isa{'m} are \emph{type variables} that can be replaced with arbitrary types; we only require that node identifiers \isa{'n} are globally unique (e.g.\ UUIDs); timestamps \isa{'t} need to be globally unique and totally ordered (e.g.
+Lamport timestamps~\cite{Lamport:1978jq}).
+
+The meaning of an operation \isa{Move t p m c} is that at time \isa{t}, the node with ID \isa{c} is moved to be a child of the parent node with ID \isa{p}.
+The operation does not specify the old location of \isa{c}; the algorithm simply removes \isa{c} from wherever it is currently located in the tree, and moves it to \isa{p}.
 If \isa{c} does not currently exist in the tree, it is created as a child of \isa{p}.
-For this reason we do not need a separate operation type to create tree nodes: simply performing a move with a fresh child node \isa{c} is sufficient to create \isa{c}.
-Nor do we need to define a remove operation: instead, we can designate a tree node as the ``trash'', and perform deletion of a node \isa{c} by moving \isa{c} to be a child of the trash node.
+For this reason we do not need a separate operation type to create tree nodes: simply performing a move with a fresh child node ID \isa{c} is sufficient to create \isa{c}.
+To delete \isa{c}, we move \isa{c} to be a child of the ``trash'' (a designated node ID that does not exist in the tree), so \isa{c} is removed from the tree.
 
@@
 
-When users want to make changes to the tree on their local replica, they generate new \isa{Move t p m c} operations for these changes, and apply these operations to their local replica.
-The timestamp \isa{t} of each operation must be globally unique; the other fields \isa{p}, \isa{m} and \isa{c} can be chosen freely.
+When users want to make changes to the tree on their local replica, they generate new \isa{Move t p m c} operations for these changes, and apply these operations to their local replica using the algorithm described in the rest of this section.
 
@@
 When we have \isa{(p, m, c) {\isasymin} tree}, that means \isa{c} is a child of \isa{p} in the tree, with associated metadata \isa{m}.
-Given a \isa{tree}, we can move the child \isa{c} to a new parent \isa{p}, with associated metadata \isa{m}, as follows:
-\begin{quote}
-\isa{\{(p', m', c') {\isasymin} tree. c' {\isasymnoteq} c\} {\isasymunion} \{(p, m, c)\}}
-\end{quote}
+Given a \isa{tree}, we can construct a new \isa{tree'} in which the child \isa{c} is moved to a new parent \isa{p}, with associated metadata \isa{m}, as follows:
+\begin{isabelle}
+tree' = \{(p', m', c') {\isasymin} tree. c' {\isasymnoteq} c\} {\isasymunion} \{(p, m, c)\}
+\end{isabelle}
 That is, we remove any existing parent-child relationship for \isa{c} from the set \isa{tree}, and then add \isa{\{(p, m, c)\}} to represent the new parent-child relationship.
@@
 The log is a list of \isa{LogMove} records in descending timestamp order.
-A \isa{LogMove t oldp p m c} record (lines 7--12) is similar to a record \isa{Move t p m c}; the only difference is that the \isa{LogMove} has an additional field \isa{oldp} of type \isa{('n {\isasymtimes} 'm) option}.
-This \isa{option} type means the field can either take the value \isa{None} (intuitively similar to null), or a pair of an inode and a metadata field.
+\isa{LogMove t oldp p m c} (lines 7--12) is similar to \isa{Move t p m c}; the difference is that \isa{LogMove} has an additional field \isa{oldp} of type \isa{('n {\isasymtimes} 'm) option}.
+This \isa{option} type means the field can either take the value \isa{None} (similar to null), or a pair of a node ID and a metadata field.
 
@@
 If \isa{c} did not exist in the tree, \isa{oldp} is set to \isa{None}.
-Otherwise, \isa{oldp} records the previous parent and metadata of \isa{c}, that is: if there exist \isa{p'} and \isa{m'} such that \isa{(p', m', c) {\isasymin} tree}, then \isa{oldp} is set to \isa{Some (p', m')}.
+Otherwise, \isa{oldp} records the previous parent and metadata of \isa{c}: if there exist \isa{p'} and \isa{m'} such that \isa{(p', m', c) {\isasymin} tree}, then \isa{oldp} is set to \isa{Some (p', m')}.
 The \isa{get\_parent} function implements this (lines 16--20).
@@
 \end{quote}
-This signature denotes a function that takes two arguments: a tree \isa{('n {\isasymtimes} 'm {\isasymtimes} 'n) set} and a node \isa{'n}.
+This signature denotes a function that takes two arguments: a tree \isa{('n {\isasymtimes} 'm {\isasymtimes} 'n) set} and a node ID \isa{'n}.
 It then returns a \isa{('n {\isasymtimes} 'm) option}.
-The operator \isa{{\isasymexists}!x} means ``there exists a unique value \isa{x} such that\dots'', while \isa{THE x}\dots means ``\emph{choose} the unique value \isa{x} such that\ldots''.
+The operator \isa{{\isasymexists}!x} means ``there exists a unique value \isa{x} such that\dots'', while \isa{THE x} means ``\emph{choose} the unique value \isa{x} such that\ldots''.
 
@@
 Lines 45--47 handle the case where the log is empty: in this case, we simply perform the operation using \isa{do\_op}, and return the new tree along with a log containing a single \isa{LogMove} record.
-
 If the log is nonempty (line 48), we take \isa{logop} to be the first element of the log, and \isa{ops} to be the rest.
@@
 The authors wish to acknowledge the support of The Boeing Company and the EPSRC ``REMS: Rigorous Engineering for Mainstream Systems'' programme grant (EP/K008528).
+% AWS credits from the AWS Educate program
 \end{acks}

### 2019-10-30 paper/move-op.tex
Squeeze rest of paper, and add more related work

@@
 
-\hyphenation{da-ta-cen-ter da-ta-cen-ters time-stamp time-stamps time-stamped}
+\hyphenation{da-ta-cen-ter da-ta-cen-ters time-stamp time-stamps time-stamped hard-links}
 
@@
         In Unix systems the nodes of the tree are known as \emph{inodes}.
-        Strictly speaking, filesystems with hardlinks form a DAG; we will focus on trees for now and show in \S\ref{sec:hardlinks} how to support links.
+        Strictly speaking, filesystems with hardlinks form a DAG; we will focus on trees for now and show in \S\ref{sec:extensions} how to support links.
     \item \textbf{Rich text editors} maintain a HTML or XML structure: a tree of textual elements such as paragraphs, lists, figures, sections, and so on.
@@
 A move operation is a 4-tuple consisting of a timestamp \isa{t} of type \isa{'t}, a parent node ID \isa{p} of type \isa{'n}, a metadata field \isa{m} of type \isa{'m}, and a child node ID \isa{c} of type \isa{'n}.
-Here, \isa{'t}, \isa{'n} and \isa{'m} are \emph{type variables} that can be replaced with arbitrary types; we only require that node identifiers \isa{'n} are globally unique (e.g.\ UUIDs); timestamps \isa{'t} need to be globally unique and totally ordered (e.g.
+Here, \isa{'t}, \isa{'n} and \isa{'m} are \emph{type variables} that can be replaced with arbitrary types; we only require that node identifiers \isa{'n} are globally unique (e.g.\ UUIDs~\cite{Leach:2005hm}); timestamps \isa{'t} need to be globally unique and totally ordered (e.g.
 Lamport timestamps~\cite{Lamport:1978jq}).
@@
 \begin{itemize}
-    \item If two operations concurrently move the same node, then the ``winning'' operation is the one with the greater timestamp.
-        Effectively, the operation with the lower timestamp moves the node first, and then the operation with the greater timestamp moves it again, so the final parent is determined by the latter.
+    \item If two operations concurrently move the same node, the operation with the lower timestamp moves the node first, and then the operation with the greater timestamp moves it again, so the final parent is determined by the latter.
         Since move operations do not specify the old location of a node, but only the new location, this sequential execution of concurrent operations is well-defined.
@@
 The opposite is also possible: an operation previously regarded as unsafe may become safe through the application of an operation that removes the risk of introducing a cycle.
-For this reason, the operation log must include all operations, even those that have no effect on the tree.
+For this reason, the operation log must include all operations, even those that were ignored.
 
@@
 
-\subsubsection{Hardlinks and symlinks}\label{sec:hardlinks}
-
+\noindent\textbf{Hardlinks and symlinks.}
 Unix filesystems support hardlinks, which allow the same file inode to be referenced from multiple locations in a tree.
@@
 
-\subsubsection{Log truncation}\label{sec:log-truncation}
-
+\smallbreak\noindent\textbf{Log truncation.}
 The algorithm as specified in Figure~\ref{fig:code} retains operations in the log indefinitely, so the memory use grows without bound.
@@
 
+A similar approach can be used to garbage-collect any tree nodes in the trash (\S\ref{sec:ops-trees}).
+Initially, trashed nodes must be retained because a concurrent move operation may move them back out of the trash.
+However, once the operation that moved a node to the trash is causally stable, we know that no future operations will refer to this node, and so the trashed node and its descendants can be discarded.
+
 We can determine causal stability in a system where the set of replicas is known, where each replica generates operations with monotonically increasing timestamps, and where the communication link between any pair of replicas is FIFO (messages are received in the order in which they are sent, as implemented e.g.\ by TCP).
@@
 
-\subsubsection{Ordering of sibling nodes}\label{sec:ordering}
-
-Another useful extension of the tree algorithm is to allow children within the same branch node to have an ordering.
+\smallbreak\noindent\textbf{Ordering of sibling nodes.}
+Another useful extension of the tree algorithm is to allow children of the same parent node to have an explicit ordering.
 For example, in a graphics application, the order of nodes determines visibility (objects that are ``further back'' may be obscured by objects that are ``closer to the front'').
-This can be implemented by maintaining a separate list CRDT for each branch node, e.g.\ using RGA~\cite{Roh:2011dw}, Treedoc~\cite{Preguica:2009fz}, Logoot~\cite{Weiss:2010hx}, or LSEQ~\cite{Nedelec:2013ky}.
-These algorithms assign a unique identifier to each element of the list, and this unique identifier can be included in the metadata field of move operations in order to determine the order sibling nodes.
+This can be implemented by maintaining an additional list CRDT for each branch node, e.g.\ using RGA~\cite{Roh:2011dw}, Treedoc~\cite{Preguica:2009fz}, Logoot~\cite{Weiss:2010hx}, or LSEQ~\cite{Nedelec:2013ky}.
+These algorithms assign a unique ID to each element of the list, and this ID can be included in the metadata field of move operations in order to determine the order of sibling nodes.
 
 This approach to determining ordering also easily supports reordering of child nodes within a parent, e.g.\ to implement the ``bring to front'' and ``send to back'' feature of applications like PowerPoint.
-To move a node to a different position in a list, we use the list CRDT to generate a new identifier at the desired position in the sequence.
-Then we perform a move operation in which the parent node is unchanged, but the metadata is changed to this new identifier.
-
-\subsubsection{File merging}\label{sec:files}
-
+To move a node to a different position in a list, we use the list CRDT to generate a new ID at the desired position in the sequence.
+Then we perform a move operation in which the parent node is unchanged, but the metadata is changed to this new ID.
+
+\smallbreak\noindent\textbf{File merging.}
 In a distributed filesystem, replication and conflict resolution is required not only for the directory structure, but also for the contents of individual files.
 This can be accomplished by using CRDTs for file contents as well.
-For example, we have explored CRDT algorithms for applications using a JSON data model~\cite{Kleppmann:2016ve,LocalFirst} and implemented them in a library called Automerge~\cite{Automerge}.
 We discuss distributed filesystems further in \S\ref{sec:filesystems}.
@@
 All theorems stated here have been formally proved and mechanically checked using Isabelle.
-For space reasons we give only a brief sketch of the proofs in this paper, but the full details can be found in our open source Isabelle theory files~\cite{SourceFiles}.
+For space reasons we give only a brief sketch of the proofs in this paper; the Isabelle files containing the full details are open source [URL omitted for blind review].
+
+% TODO \cite{SourceFiles}
 
@@
 To prove that our algorithm maintains a tree structure, no matter which operations are applied, we prove that these invariants hold.
-In fact, we slightly generalise this property and allow more than one root to exist, so the graph represents a forest, allowing an application to move nodes between multiple trees, if desired.
+In fact, we slightly generalise this property and allow more than one root to exist, so the graph represents a forest, allowing an application to move nodes between different trees, if desired.
 For example, the ``trash'' node used for deletion (\S\ref{sec:ops-trees}) can be separate from the main tree.
@@
 \noindent That is, we consider any list of operations \isa{ops} and define \isa{(log, tree)} to be the replica state after \isa{ops} have been applied.
-We then prove that \isa{unique\_parent tree} holds, where the \isa{unique\_parent} predicate is defined on lines 56--57 of Figure~\ref{fig:code}: it states that whenever the tree contains a triple whose third element is the child node \isa{c}, then the first and second elements of the triple (the parent node and the metadata) are uniquely defined.
+We then prove that \isa{unique\_parent tree} holds, where the \isa{unique\_parent} predicate is defined on lines 56--57 of Figure~\ref{fig:code}: whenever the tree contains a triple whose third element is the child node \isa{c}, then the first and second elements of the triple (the parent node and the metadata) are uniquely defined.
 
@@
 Otherwise we can assume \isa{{\isasymnot}(ancestor tree c newp {\isasymor} c = newp)} and we have:
-\medbreak\noindent
-\isa{tree2 = \{(p', m', c') {\isasymin} tree1. c' {\isasymnoteq} c\} {\isasymunion} \{(newp, m, c)\}}
-\medbreak\noindent
+\begin{isabelle}
+tree2 = \{(p', m', c') {\isasymin} tree1. c' {\isasymnoteq} c\} {\isasymunion} \{(newp, m, c)\}
+\end{isabelle}
 That is, \isa{do\_op} adds one edge \isa{\{(newp, m, c)\}} to the graph, and perhaps removes another edge.
@@
         From the inductive hypothesis we have
-        \begin{quote}
-        \isa{apply\_op oper2 (apply\_op oper1 (ops, tree')) =\\
-             apply\_op oper1 (apply\_op oper2 (ops, tree'))}
-        \end{quote}
+        \begin{isabelle}
+        apply\_op oper2 (apply\_op oper1 (ops, tree')) =\\
+        apply\_op oper1 (apply\_op oper2 (ops, tree'))
+        \end{isabelle}
         We now apply \isa{oper1} and \isa{oper2} as in the base case, and finally reapply \isa{logop} to both sides using \isa{redo\_op}, resulting in the same state on both sides.
@@
 Using our commutativity proof above we integrate our tree datatype with this framework, and thus prove that our move operation on trees does indeed guarantee strong eventual consistency.
-The details appear in our Isabelle theory files~\cite{SourceFiles}.
+The details appear in our Isabelle theory files [URL omitted for blind review]. % \cite{SourceFiles}.
 
@@
 
-We say a hash-map \isa{t} of type \isa{('n, 'm {\isasymtimes} 'n) hm} \emph{simulates} a set \isa{T} of type \isa{('n {\isasymtimes} 'm {\isasymtimes} 'n) set} when they have the same entries:
+A hash-map \isa{t} of type \isa{('n, 'm {\isasymtimes} 'n) hm} \emph{simulates} a set \isa{T} of type \isa{('n {\isasymtimes} 'm {\isasymtimes} 'n) set} when their entries are the same:
 \begin{isabelle}
@@
 \ simulates\ \isakeyword{where}\ {\isacartoucheopen}simulates\ t\ T\ {\isasymequiv}\isanewline
-\ \ {\isacharparenleft}{\isasymforall}p\ m\ c{\isachardot}\ hm{\isachardot}lookup\ c\ t\ {\isacharequal}\ Some\ {\isacharparenleft}m{\isacharcomma}\ p{\isacharparenright}\ {\isasymlongleftrightarrow}\ {\isacharparenleft}p{\isacharcomma}\ m{\isacharcomma}\ c{\isacharparenright}\ {\isasymin}\ T{\isacharparenright}{\isacartoucheclose}
+{\isacharparenleft}{\isasymforall}p\ m\ c{\isachardot}\ hm{\isachardot}lookup\ c\ t\ {\isacharequal}\ Some\ {\isacharparenleft}m{\isacharcomma}\ p{\isacharparenright}\ {\isasymlongleftrightarrow}\ {\isacharparenleft}p{\isacharcomma}\ m{\isacharcomma}\ c{\isacharparenright}\ {\isasymin}\ T{\isacharparenright}{\isacartoucheclose}
 \end{isabelle}
@@
 
-That is, if \isa{executable\_apply\_ops} and \isa{apply\_ops} are applied to the same list of operations then they produce identical logs, and also produce trees that contain the same set of key-value bindings---i.e.
-the two trees are extensionally equivalent, despite having very different in-memory representations.
-Moreover, we can prove corollaries of \isa{apply\_ops\_commutes} and \isa{apply\_ops\_acyclic}, respectively, for the hash-map-based implementation.
+That is, if \isa{executable\_apply\_ops} and \isa{apply\_ops} are applied to the same list of operations, they produce identical logs, and also produce trees that contain the same set of key-value bindings---i.e.
+the trees are extensionally equivalent, despite having very different in-memory representations.
+We can also prove corollaries of \isa{apply\_ops\_commutes} and \isa{apply\_ops\_acyclic} for the hash-map-based implementation.
 %The analogue of \isa{apply\_ops\_unique\_parent} holds trivially since by construction, the hash-map contains at most one (metadata, parent) pair for a given child.
 
-\begin{figure}
-  \includegraphics[width=\columnwidth,keepaspectratio=true]{plots.png}
-  \caption{Median execution time for applying an operation to the replica state, on an Amazon EC2 t2.small instance, at varying operation throughput rates.
-Error bars indicate the minimum and 95th percentile execution times.}
-  \label{fig:plots}
-  \Description{Plot in which the x axis goes from 0 to 30 move operations per second, and the y axis goes from 0 to 12 milliseconds.
-Local operations take a near-constant time less than 1ms; the time for remote operations gradually increases to 2ms until 24 operations/sec, at which point it suddenly shoots up off the top of the scale.}
-\end{figure}
+\section{Evaluation}\label{sec:evaluation}
+
+\subsection{Experimental setup and results}
+
+We used Isabelle/HOL's code generation mechanism to extract executable Scala code from our HOL definitions, as discussed in \S\ref{subsect.extracting}, and wrapped this implementation in a network service using the Akka~\cite{Akka} actor framework, using gRPC~\cite{gRPC} for network communication.
+We deployed three replicas of this service on Amazon EC2 \texttt{t2.small} instances in different regions worldwide: one in Northern California (region \texttt{us-west-1}), one in Ireland (\texttt{eu-west-1}), and one in Singapore (\texttt{ap-southeast-1}).
+The network delays between these regions are substantial; see Table~\ref{tab:rpc-times}.
+
+We use a synthetic workload in which each replica generates move operations at the same rate; for each move operation the generating replica chooses parent and child nodes uniformly at random from a set of 1,000 tree nodes.
+For a given operation rate we run the system for 10 minutes to reach a steady state.
+We use Lamport timestamps~\cite{Lamport:1978jq} as operation timestamps, and string values to identify tree nodes.
+When a replica generates an operation, it immediately applies that operation to its local state, and asynchronously sends the operation to the other two replicas via gRPC requests.
+When a replica receives an operation from a remote replica, it also applies that operation to its own replica state.
+Thus, of the operations applied by each replica, approximately one third are locally generated, and two thirds are received from the other two replicas.
+
+We varied the rate at which replicas generated operations, and measured the execution time of our algorithm for applying each operation.
+Figure~\ref{fig:plots} shows the median execution times for local and remote operations respectively.
+The execution time for local operations is near-constant (median times mostly between 0.2~ms and 0.6~ms), while the execution time for remote operations increases approximately linearly from 0.2~ms to 2.3~ms over the range of 3 to 24 operations/sec.
+Above 25 operations/sec the execution time for remote operations suddenly increases to over 30~ms (off the scale in Figure~\ref{fig:plots}).
+We do not plot the exact execution times above 12~ms because they are less meaningful, as explained below.
 
@@
 
-\section{Evaluation}\label{sec:evaluation}
-
-\subsection{Experimental setup and results}
-
-We used Isabelle/HOL's code generation mechanism to extract executable Scala code from our HOL definitions, as discussed in \S\ref{subsect.extracting}, and wrapped this implementation in a network service using the Akka~\cite{Akka} actor framework, using gRPC~\cite{gRPC} for network communication.
-We deployed three replicas of this service on Amazon EC2 \texttt{t2.small} instances in different regions worldwide: one in Northern California (region \texttt{us-west-1}), one in Ireland (\texttt{eu-west-1}), and one in Singapore (\texttt{ap-southeast-1}).
-The network delays for communication between these regions are substantial; see Table~\ref{tab:rpc-times}.
-
-We use a synthetic workload in which each replica generates move operations at the same rate; for each move operation the generating replica chooses parent and child nodes uniformly at random from a set of 1,000 tree nodes.
-For a given operation rate we run the system for 10 minutes to reach a steady state.
-We use Lamport timestamps~\cite{Lamport:1978jq} as operation timestamps, and string values to identify tree nodes.
-When a replica generates an operation, it immediately applies that operation to its local state, and asynchronously sends the operation to the other two replicas via gRPC requests.
-When a replica receives an operation from a remote replica, it also applies that operation to its own replica state.
-Thus, of the operations applied by each replica, approximately one third are locally generated, and two thirds are received from the other two replicas.
-
-We varied the rate at which replicas generated operations, and measured the execution time of our algorithm for applying each operation.
-Figure~\ref{fig:plots} shows the median execution times for local and remote operations respectively.
-The execution time for local operations is near-constant (median times mostly between 0.2~ms and 0.6~ms), while the execution time for remote operations increases approximately linearly from 0.2~ms to 2.3~ms over the range of 3 to 24 operations/sec.
-Above 25 operations/sec the execution time for remote operations suddenly increases to over 30~ms (off the scale of Figure~\ref{fig:plots}).
-We do not plot the exact values for execution times above 12~ms because they are less meaningful, as explained below.
+\begin{figure}
+  \includegraphics[width=\columnwidth,keepaspectratio=true]{plots.png}
+  \caption{Median execution time for applying an operation to the replica state, on an Amazon EC2 t2.small instance, at varying operation throughput rates.
+Error bars indicate the minimum and 95th percentile execution times.}
+  \label{fig:plots}
+  \Description{Plot in which the x axis goes from 0 to 30 move operations per second, and the y axis goes from 0 to 12 milliseconds.
+Local operations take a near-constant time less than 1ms; the time for remote operations gradually increases to 2ms until 24 operations/sec, at which point it suddenly shoots up off the top of the scale.}
+\end{figure}
 
@@
 Nevertheless, our evaluation sends a positive signal about the practical applicability of our algorithm.
-Note that when a user is interacting with a system, they only have to wait for local operations to execute; any remote operations can be applied in the background without affecting user interaction.
+When a user is interacting with a system, they only have to wait for local operations to execute; any remote operations can be applied in the background without affecting user interaction.
 Since our algorithm need not perform any undos or redos for local operations, these user interactions are consistently fast.
@@
 The alternative to our optimistic replication approach would be to use a central server (master replica) or consensus protocol that processes all operations in a serialisable order.
-Performing an operation then requires waiting for synchronous communication with a server potentially located on a different continent, taking hundreds of milliseconds (Table~\ref{tab:rpc-times}); by contrast, a local operation taking a fraction of a millisecond is orders of magnitude faster.
+Performing an operation then requires waiting for communication with a server potentially located on a different continent, taking hundreds of milliseconds (Table~\ref{tab:rpc-times}); by contrast, a local operation taking a fraction of a millisecond is orders of magnitude faster.
 Moreover, a local operation requires no waiting for network communication, and thus it supports disconnected operation: mobile devices can read and modify the replicated tree even while disconnected from the Internet, which is not the case with centralised systems.
@@
 
-Many replicated data systems use optimistic replication~\cite{Saito:2005jw}, which allows the state of replicas to temporarily diverge, in order to achieve better performance and better availability in the presence of faults than strongly consistent systems~\cite{Bailis:2014th,Gilbert:2002il}.
+Many replicated data systems use optimistic replication~\cite{Saito:2005jw}, which allows the state of replicas to temporarily diverge, in order to achieve better performance and availability in the presence of faults than strongly consistent systems~\cite{Bailis:2014th,Gilbert:2002il}.
 As a consequence, these systems require a mechanism for merging or reconciling conflicting updates that were made concurrently on different replicas.
-For example, version control systems such as Git~\cite{Chacon:2014kr} leave any conflicts for the user to resolve manually.
-Some systems, such as Dynamo~\cite{DeCandia:2007ui} and Bayou~\cite{Terry:1995dn}, rely on the application programmer to provide explicit conflict resolution logic; however, such logic is difficult to get right \cite{Bailis:2013jc,Burckhardt:2014hy,Gomes:2017gy}.
+For example, version control systems such as Git~\cite{Chacon:2014kr} leave conflicts for the user to resolve manually.
+Systems such as Dynamo~\cite{DeCandia:2007ui} and Bayou~\cite{Terry:1995dn} rely on the application programmer to provide explicit conflict resolution logic; however, such logic is difficult to get right \cite{Bailis:2013jc,Burckhardt:2014hy,Gomes:2017gy}.
 Hence, we focus on systems that automatically ensure that all replicas converge towards a consistent state, without requiring custom application logic---a consistency model known as \emph{strong eventual consistency}~\cite{Shapiro:2011un,Gomes:2017gy}.
@@
 CRDTs have been defined for various abstract datatypes, such as registers and counters \cite{Shapiro:2011wy,Shapiro:2011un}, maps \cite{Baquero:2016iv,Kleppmann:2016ve}, sets \cite{Bieniusa:2012wu,Bieniusa:2012gt}, lists (sequences) \cite{Roh:2011dw,Nedelec:2013ky}, and text \cite{Preguica:2009fz,Weiss:2010hx}.
-For tree data structures, several approaches have been proposed:
+For trees, several approaches have been proposed:
 \begin{itemize}
     \item Martin et al.~\cite{Martin:2010ih} define a CRDT for XML data, and Kleppmann and Beresford~\cite{Kleppmann:2016ve} define a CRDT for JSON.
-        However, these algorithms only deal with insertion and deletion of tree nodes, and do not support a move operation.
+        However, these algorithms only deal with insertion and deletion of tree nodes, and do not support moves.
     \item As discussed in \S\ref{sec:impossibility}, Najafzadeh et al.~\cite{Najafzadeh:2017vk,Najafzadeh:2018bw} propose two implementations for a replicated filesystem: a CRDT in which conflicting moves are handled by duplicating tree nodes (as in Figures~\ref{fig:move-same-item}b and~\ref{fig:move-cycle}b), and a centralised implementation in which move operations must obtain a lock before executing (not a CRDT since it relies on synchronous coordination).
-    \item Ahmed-Nacer et al.~\cite{AhmedNacer:2012us} survey approaches to handling conflicts on trees without providing concrete algorithms.
+    \item Ahmed-Nacer et al.~\cite{AhmedNacer:2012us} outline approaches to handling conflicts on trees, but do not provide algorithms.
     \item Tao et al.~\cite{Tao:2015gd} propose handling conflicting move operations by allowing the same object to appear in more than one location; thus, their datatype is strictly a DAG, not a tree.
@@
         Tao et al.\ also perform experiments with Dropbox, Google Drive, and OneDrive, similar to our experiments discussed in \S\ref{sec:move-is-hard}.
-    \item In our previous unpublished work \cite{ExtendedVersion} we develop the foundations for the algorithm described in this paper.
 \end{itemize}
@@
 Besides CRDTs, another family of algorithms for concurrent modification of data structures is \emph{Operational Transformation} (OT)~\cite{Sun:1998vf}.
-Several authors have defined concurrent tree structures using OT \cite{Jungnickel:2016cb,Ignat:2003jy,Davis:2002iv}, but they only consider insertion and deletion of nodes, and do not support a move operation.
+Several authors have defined concurrent tree structures using OT \cite{Jungnickel:2016cb,Ignat:2003jy,Davis:2002iv}, but they only consider insertion and deletion of nodes, and do not support moves.
 
@@
 
+\subsection{Totally ordered operation log}
+
+Ordering operations by a timestamp, and undoing/redoing them as necessary so that they take effect in ascending timestamp order, is a well-known approach.
+It is used for example in systems like Bayou~\cite{Terry:1995dn}, Jefferson's \emph{Time Warp} mechanism~\cite{Jefferson:1985em}, Shapiro et al.'s \emph{capricious total order}~\cite{Shapiro:2016hp}, and Burckhardt's \emph{standard conflict resolution}~\cite[\S~4.3.3]{Burckhardt:2014hy}.
+
+However, to our knowledge, this approach has not previously been applied to the problem of replicated trees, nor are we aware of previous mechanised proofs (using Isabelle or other tools) that formalise this approach.
+
 \section{Conclusions}
@@
 Updates made to a local replica take effect immediately, while operations from remote replicas can be propagated and applied in the background.
-This approach means that user interaction is consistently fast, even in the face of unbounded communication delays between replicas, e.g.\ during disconnected operation of mobile devices.
-
-While existing systems exhibit anomalies (such as duplicating nodes or introducing cycles) or error messages (like in Figure~\ref{fig:gdrive-error}) in the face of concurrent move operations, our algorithm is free from such problems.
+This approach means that user interaction is consistently fast, even in the face of unbounded communication delays or during disconnected operation of mobile devices.
+
+While existing systems exhibit anomalies (such as duplicating nodes or introducing cycles) or bugs and internal errors (like in Figure~\ref{fig:gdrive-error}) in the face of concurrent move operations, our algorithm is free from such problems.
 We formally verified the correctness of our algorithm using the Isabelle/HOL proof assistant; our theorems show that replicas can apply operations in any order, and that the result is always a valid tree (nodes have at most one parent, and the graph does not contain any cycles).

### 2019-10-31 paper/move-op.tex
paper: minor changes after a read-through of Martin's editing
Some small language/presentation changes here-and-there.

@@
 
-When the user wants to make a change to the tree, they generate an operation and apply it on their local replica.
-Every operation is also asynchronously sent over the network to all other replicas, and applied on every remote replica using the same algorithm as for local operations.
+When the user wants to make a change to the tree, they generate an operation and apply it to their local replica.
+Every operation is also asynchronously sent over the network to all other replicas, and applied to every remote replica using the same algorithm as for local operations.
 The network may deliver operations in any order; thus, the communication can be performed peer-to-peer, and it does not require any central server or consensus protocol.
@@
 
-We can now represent the tree as a set of \isa{(parent, meta, child)} triples, denoted in Isabelle as \isa{('n {\isasymtimes} 'm {\isasymtimes} 'n) set}.
+We can now represent the tree as a set of \isa{(parent, meta, child)} triples, denoted in Isabelle/HOL as \isa{('n {\isasymtimes} 'm {\isasymtimes} 'n) set}.
 When we have \isa{(p, m, c) {\isasymin} tree}, that means \isa{c} is a child of \isa{p} in the tree, with associated metadata \isa{m}.
@@
 The \isa{apply\_op} function (lines 43--51) takes two arguments: a \isa{Move} operation to apply and the current replica state; and it returns the new replica state.
-The constraint \isa{'t::\{linorder\}} in the type signature indicates that timestamps \isa{'t} are instances of \isa{linorder} type class, and they can therefore be compared with the \isa{<} operator defining a total order.
+The constraint \isa{'t::\{linorder\}} in the type signature indicates that timestamps \isa{'t} are instances of \isa{linorder} type class, and they can therefore be compared with the \isa{<} operator defining a linear (or total) order.
 This comparison occurs on line 49.
@@
 If the log is nonempty (line 48), we take \isa{logop} to be the first element of the log, and \isa{ops} to be the rest.
-(The hash character in \isa{logop {\isacharhash} ops} is the \emph{cons} operator that adds one element to the head of a list.)
+(The hash character in \isa{logop {\isacharhash} ops} is the \emph{list cons} operator that adds one element to the head of a list.)
 If the timestamp of \isa{logop} is greater than the timestamp of the new operation (line 49) we first undo \isa{logop} with \isa{undo\_op}, then recursively apply the new operation to the remaining log, and finally reapply \isa{logop} with \isa{redo\_op} (line 50).
@@
 
-This approach to determining ordering also easily supports reordering of child nodes within a parent, e.g.\ to implement the ``bring to front'' and ``send to back'' feature of applications like PowerPoint.
+This approach to determining ordering also easily supports reordering of child nodes within a parent, e.g.\ to implement the ``bring to front'' and ``send to back'' feature of applications like Microsoft PowerPoint.
 To move a node to a different position in a list, we use the list CRDT to generate a new ID at the desired position in the sequence.
@@
 The first invariant we prove is that each tree node has either no parent (if it is the root of a tree) or exactly one parent (if it is a non-root node).
-We state this theorem in Isabelle as follows, where \isa{apply\_ops\_unique\_parent} is the name we give to this theorem:
+We state this theorem in Isabelle/HOL as follows, where \isa{apply\_ops\_unique\_parent} is the name we give to this theorem:
 \begin{isabelle}
@@
 The second invariant we require is that the graph must not contain any cycles.
-In Isabelle:
+In Isabelle/HOL:
 \begin{isabelle}
@@
 
-Proof sketch: Assume \isa{do\_op (op1, tree1) = (op2, tree2)} for some \isa{op1}, \isa{op2}, \isa{tree1} and \isa{tree2}.
+\emph{Proof sketch}: Assume \isa{do\_op (op1, tree1) = (op2, tree2)} for some \isa{op1}, \isa{op2}, \isa{tree1} and \isa{tree2}.
 Further assume \isa{acyclic tree1}.
@@
 As discussed in \S\ref{sec:applying}, we require that when replicas apply the same set of operations, they converge towards the same state, regardless of the order in which the operations are applied.
-We formalise this in Isabelle as follows:
+We formalise this in Isabelle/HOL as follows:
 \begin{isabelle}
@@
 
-For the inductive step, let the log be \isa{logop {\isacharhash} ops} (where \isa{\isacharhash} is the \emph{cons} operator).
-We now need to show that
+For the inductive step, let the log be \isa{logop {\isacharhash} ops} (where \isa{\isacharhash} is the \emph{list cons} operator).
+We now need to show that:
 \begin{isabelle}
@@
         Then both applying \isa{oper1} and applying \isa{oper2} first undoes \isa{logop}, yielding \isa{tree' = undo\_op (logop, tree)}.
-        From the inductive hypothesis we have
+        From the inductive hypothesis we have:
         \begin{isabelle}
@@
 We therefore produce variants of the definitions in Figure~\ref{fig:code} that are designed for execution rather than theorem proving.
-Rather than representing the tree as a set of (parent, metadata, child) triples, these definitions use a hash-map in which the keys are child nodes, and the values are (metadata, parent) pairs, written in Isabelle as \isa{('n::\{hashable\}, 'm {\isasymtimes} 'n) hm}.
+Rather than representing the tree as a set of (parent, metadata, child) triples, these definitions use a hash-map in which the keys are child nodes, and the values are (metadata, parent) pairs, written in Isabelle/HOL as \isa{('n::\{hashable\}, 'm {\isasymtimes} 'n) hm}.
 The \isa{hashable} type class means that keys must have a hashing function.
@@
     \midrule
-    California &            & 395 ms  & 491 ms    \\
-    Ireland    & 395 ms     &         & 547 ms    \\
-    Singapore  & 491 ms     & 547 ms  &           \\
+    California &      ---      & 395 ms  & 491 ms    \\
+    Ireland    & 395 ms     &    ---     & 547 ms    \\
+    Singapore  & 491 ms     & 547 ms  &     ---      \\
     \bottomrule
@@
 The original version of Rumor was unable to scale beyond 20 replicas, but an extension called Roam~\cite{Ratner:1999fh} allowed better scaling. 
-In an attempt to test Rumor's conflict handling we obtained the source code from archive.org~\cite{RumorSource}; however, we were unable to get it running after modest effort.
+In an attempt to test Rumor's conflict handling we obtained the source code from \emph{archive.org}~\cite{RumorSource}; however, we were unable to get it running after modest effort.
 %The user manual \cite{RumorManual} makes the claim that "Unix directories are an important example of a kind of file for which Rumor automatically resolves conflicts"

### 2019-10-31 evaluation/src/main/scala/TestReplica.scala
Rewrite evaluation: use plain TCP and threads instead of gRPC and Akka

@@ -1,120 +1,260 @@
-import akka.NotUsed
-import akka.actor.{ ActorRef, ActorSystem }
-import akka.grpc.GrpcClientSettings
-import akka.http.scaladsl.UseHttp2.Always
-import akka.http.scaladsl.model.{ HttpRequest, HttpResponse }
-import akka.http.scaladsl.{ Http, HttpConnectionContext }
-import akka.pattern.ask
-import akka.stream.{ ActorMaterializer, Materializer }
-import akka.stream.scaladsl.{ BroadcastHub, Keep, Sink, Source }
-import akka.util.Timeout
 import com.codahale.metrics.{ ConsoleReporter, MetricRegistry, Timer }
-import com.typesafe.config.ConfigFactory
-import examplerpc.{ MoveService, MoveServiceClient, MoveServiceHandler }
-import java.util.concurrent.TimeUnit.SECONDS
-import java.util.concurrent.ConcurrentHashMap
+import java.io.{ ByteArrayInputStream, ByteArrayOutputStream, DataInputStream, DataOutputStream, InputStream, OutputStream }
+import java.net.{ ServerSocket, Socket }
+import java.util.concurrent.{ArrayBlockingQueue, ConcurrentHashMap, TimeUnit}
 import java.util.Random
-import scala.collection.JavaConversions._
-import scala.concurrent.{ ExecutionContext, Future }
-import scala.concurrent.duration._
-import scala.util.{ Failure, Success }
 
 object TestReplica {
 
-  // TCP port number for gRPC
+  // TCP port number for communication between replicas
   val PORT = 8080
 
   // Time interval between generated operations (= 1 / operation rate)
-  val OPERATION_INTERVAL = 1000 millis
+  val OPERATION_INTERVAL = TimeUnit.MILLISECONDS.toNanos(1000)
+
+  // How long to run the test before shutting down
+  val RUN_DURATION = TimeUnit.MINUTES.toNanos(10)
 
   def main(args: Array[String]): Unit = {
     if (args.length < 1) {
-      throw new Exception("Usage: TestReplica own-ip remote-ip1 [remote-ip2 ...]")
+      throw new Exception("Usage: TestReplica replica-id remote-ip1 [remote-ip2 ...]")
     }
-
-    val conf = ConfigFactory.parseString("akka.http.server.preview.enable-http2 = on")
-      .withFallback(ConfigFactory.defaultApplication())
 
[…]

### 2019-10-31 evaluation/src/main/scala/TreeActor.scala

@@ -1,60 +1,0 @@
-import akka.actor.{ Actor, Props }
-import akka.event.Logging
-import com.codahale.metrics.MetricRegistry
-
-object TreeActor {
-  case class RequestMove(parent: Long, metadata: String, child: Long)
-
-  def props(actorId: Long, metrics: MetricRegistry) = Props(new TreeActor(actorId, metrics))
-}
-
-class TreeActor(actorId: Long, metrics: MetricRegistry) extends Actor {
-  val log = Logging(context.system, this)
-  val localTimer  = metrics.timer("TreeActor.local")
-  val remoteTimer = metrics.timer("TreeActor.remote")
-
-  var counter: Long = 0
-
-  var state: (List[generated.log_op[(BigInt, BigInt), BigInt, String]], generated.hashmap[BigInt, (String, BigInt)])
-    = (Nil, generated.hm_empty[BigInt, (String, BigInt)].apply(()))
-
-  def applyOp(op: examplerpc.Move) {
-    val opCounter = op.timestamp.get.counter
-    val timestamp = (BigInt(opCounter), BigInt(op.timestamp.get.replica))
-    val operation = generated.Move(timestamp, BigInt(op.parent), op.metadata, BigInt(op.child))
-    //log.info("Applied %s".format(operation))
-    state = generated.integer_apply_op(operation)(state)
-    if (opCounter > counter) counter = opCounter
-  }
-
-  def generateOp(parent: Long, metadata: String, child: Long): examplerpc.Move = {
-    val timestamp = examplerpc.LamportTS(counter + 1, actorId)
-    val operation = examplerpc.Move(Some(timestamp), parent, metadata, child)
-    applyOp(operation)
-    operation
-  }
-
-  def receive = {
-    // Remote request
-    case op: examplerpc.Move => {
-      val timer = remoteTimer.time()
-      try {
-        applyOp(op)
-        sender() ! op.timestamp.get
-      } finally {
-        timer.stop()
-      }
-    }
-
-    // Local request
-    case TreeActor.RequestMove(parent, metadata, child) => {
-      val timer = localTimer.time()
-      try {
-        val operation = generateOp(parent, metadata, child)
-        sender() ! operation
-      } finally {
-        timer.stop()
-      }
-    }
-  }
-}

### 2019-10-31 evaluation/run-all.sh
Test automation

@@ -1,0 +1,23 @@
+#!/bin/bash
+set -e
+
+US_WEST_1=18.144.1.184
+EU_WEST_1=34.255.190.108
+AP_SOUTHEAST_1=18.140.64.126
+
+run_batch () {
+    INTERVAL=$1
+
+    ssh -i ~/.ec2/martin-aws-us-west-1.pem ubuntu@$US_WEST_1 /home/ubuntu/move-op/evaluation/run-test.sh $INTERVAL 1 $EU_WEST_1 $AP_SOUTHEAST_1 &
+    PROC1=$!
+    ssh -i ~/.ec2/martin-aws-eu-west-1.pem ubuntu@$EU_WEST_1 /home/ubuntu/move-op/evaluation/run-test.sh $INTERVAL 2 $US_WEST_1 $AP_SOUTHEAST_1 &
+    PROC2=$!
+    ssh -i ~/.ec2/martin-aws-ap-southeast-1.pem ubuntu@$AP_SOUTHEAST_1 /home/ubuntu/move-op/evaluation/run-test.sh $INTERVAL 3 $US_WEST_1 $EU_WEST_1 &
+    PROC3=$!
+
+    wait $PROC1
+    wait $PROC2
+    wait $PROC3
+}
+
+run_batch 100

### 2019-10-31 evaluation/run-test.sh

@@ -1,0 +1,21 @@
+#!/bin/bash
+set -e
+
+INTERVAL="${1?Please specify interval}"
+REPLICA_ID="${2?Please specify replica ID}"
+REMOTE1="${3?Please specify remote IP 1}"
+REMOTE2="${4?Please specify remote IP 2}"
+LOGFILE="data/logs/interval_$INTERVAL.log"
+
+cd "$(dirname "$0")"
+
+sed -i "s/\\(val OPERATION_INTERVAL =.*\\)([0-9]*)/\\1($INTERVAL)/" src/main/scala/TestReplica.scala
+
+mkdir -p data/logs
+
+if [ -f "$LOGFILE" ]; then
+    echo "log file already exists"
+    exit 1
+fi
+
+sbt "runMain TestReplica $REPLICA_ID $REMOTE1 $REMOTE2" | tee "$LOGFILE"

### 2019-10-31 evaluation/src/main/scala/TestReplica.scala

@@ -221,7 +221,6 @@ class ReplicaThread(replicaId: Long, metrics: MetricRegistry) extends Runnable {
       val move = Protocol.Move(counter + 1, replicaId, random.nextInt(1000), random.nextInt(1000))
       this.applyMove(move)
       for (client <- clients) client.send(move)
-      println(s"Generated: ${move}")
     } finally {
       timer.stop()
     }

### 2019-11-02 evaluation/src/main/scala/TestReplica.scala
Adjust backpressure config

@@ -15,6 +15,9 @@ object TestReplica {
   // How long to run the test before shutting down
   val RUN_DURATION = TimeUnit.MINUTES.toNanos(10)
 
+  // Backpressure kicks in if more than this number of requests in flight
+  val MAX_PENDING_REQUESTS = 50
+
   def main(args: Array[String]): Unit = {
     if (args.length < 1) {
       throw new Exception("Usage: TestReplica replica-id remote-ip1 [remote-ip2 ...]")
@@ -125,7 +128,6 @@ abstract class Connection(socket: Socket, recvFrameSize: Int) extends Runnable {
 class ClientThread(val remoteIp: String, metrics: MetricRegistry)
     extends Connection(new Socket(remoteIp, TestReplica.PORT), 2 * 8) {
 
-  val MAX_PENDING_REQUESTS = 10 // backpressure kicks in if more than this pending
   val timer = metrics.timer(s"ClientThread($remoteIp).requests")
   val requests = new ConcurrentHashMap[Protocol.Ack, Timer.Context]()
 
@@ -143,7 +145,7 @@ class ClientThread(val remoteIp: String, metrics: MetricRegistry)
   // Returns false if we're happy to accept more requests, and true if we need
   // to hold off on enqueueing more requests for now.
   def backpressure: Boolean = {
-    requests.size() >= MAX_PENDING_REQUESTS
+    requests.size() >= TestReplica.MAX_PENDING_REQUESTS
   }
 }
 

### 2019-11-02 evaluation/data/processing_times.sh
Scripts for running experiments and data analysis

@@ -1,0 +1,15 @@
+#!/bin/bash
+
+echo ",,us_west_1,,,,,,,eu_west_1,,,,,,,ap_southeast_1" >> processing_times.csv
+echo -n "interval,ops/sec," >> processing_times.csv
+echo -n "local min,local median,local p95,remote min,remote median,remote p95," >> processing_times.csv
+echo -n "local min,local median,local p95,remote min,remote median,remote p95," >> processing_times.csv
+echo    "local min,local median,local p95,remote min,remote median,remote p95"  >> processing_times.csv
+
+for interval in 100 50 20 14 10 7 6 5 2; do
+    us_times="$(cat logs/interval_${interval}_us_west_1.log.gz      | gunzip | awk -f processing_times.awk | tail -n 2 | head -n 1 | tr '\n' ',')"
+    eu_times="$(cat logs/interval_${interval}_eu_west_1.log.gz      | gunzip | awk -f processing_times.awk | tail -n 2 | head -n 1 | tr '\n' ',')"
+    ap_times="$(cat logs/interval_${interval}_ap_southeast_1.log.gz | gunzip | awk -f processing_times.awk | tail -n 2 | head -n 1 | tr '\n' ',')"
+    op_rate="$(echo "${us_times%%,*} + ${eu_times%%,*} + ${ap_times%%,*}" | bc)"
+    echo "${interval},${op_rate},${us_times#*,}${eu_times#*,}${ap_times#*,}" >> processing_times.csv
+done

### 2019-11-02 evaluation/run-all.sh

@@ -9,15 +9,17 @@ run_batch () {
     INTERVAL=$1
 
     ssh -i ~/.ec2/martin-aws-us-west-1.pem ubuntu@$US_WEST_1 /home/ubuntu/move-op/evaluation/run-test.sh $INTERVAL 1 $EU_WEST_1 $AP_SOUTHEAST_1 &
-    PROC1=$!
     ssh -i ~/.ec2/martin-aws-eu-west-1.pem ubuntu@$EU_WEST_1 /home/ubuntu/move-op/evaluation/run-test.sh $INTERVAL 2 $US_WEST_1 $AP_SOUTHEAST_1 &
-    PROC2=$!
     ssh -i ~/.ec2/martin-aws-ap-southeast-1.pem ubuntu@$AP_SOUTHEAST_1 /home/ubuntu/move-op/evaluation/run-test.sh $INTERVAL 3 $US_WEST_1 $EU_WEST_1 &
-    PROC3=$!
+    wait
 
-    wait $PROC1
-    wait $PROC2
-    wait $PROC3
+    scp -i ~/.ec2/martin-aws-us-west-1.pem ubuntu@$US_WEST_1:/home/ubuntu/move-op/evaluation/data/logs/interval_$INTERVAL.log data/logs/interval_${INTERVAL}_us_west_1.log
+    scp -i ~/.ec2/martin-aws-eu-west-1.pem ubuntu@$EU_WEST_1:/home/ubuntu/move-op/evaluation/data/logs/interval_$INTERVAL.log data/logs/interval_${INTERVAL}_eu_west_1.log
+    scp -i ~/.ec2/martin-aws-ap-southeast-1.pem ubuntu@$AP_SOUTHEAST_1:/home/ubuntu/move-op/evaluation/data/logs/interval_$INTERVAL.log data/logs/interval_${INTERVAL}_ap_southeast_1.log
+
+    gzip data/logs/interval_${INTERVAL}_us_west_1.log data/logs/interval_${INTERVAL}_eu_west_1.log data/logs/interval_${INTERVAL}_ap_southeast_1.log
 }
 
-run_batch 100
+cd "$(dirname "$0")"
+mkdir -p data/logs
+run_batch 14

### 2019-11-02 evaluation/run-test.sh

@@ -1,6 +1,9 @@
 #!/bin/bash
 set -e
 
+# This script executes on an EC2 VM. It runs a replica that generates
+# operations with a given inter-operation interval.
+
 INTERVAL="${1?Please specify interval}"
 REPLICA_ID="${2?Please specify replica ID}"
 REMOTE1="${3?Please specify remote IP 1}"

### 2019-11-02 paper/move-op.tex
Logs and data from new set of experiments

@@
     \toprule
-               & California & Ireland & Singapore \\
+                    & to California & to Ireland & to Singapore \\
     \midrule
-    California &      ---      & 395 ms  & 491 ms    \\
-    Ireland    & 395 ms     &    ---     & 547 ms    \\
-    Singapore  & 491 ms     & 547 ms  &     ---      \\
+    from California & ---           & 139 ms     & 175 ms       \\
+    from Ireland    & 145 ms        & ---        & 182 ms       \\
+    from Singapore  & 176 ms        & 178 ms     & ---          \\
     \bottomrule
@@
 \begin{figure}
-  \includegraphics[width=\columnwidth,keepaspectratio=true]{plots.png}
+  \includegraphics[width=\columnwidth,keepaspectratio=true]{plot.pdf}
   \caption{Median execution time for applying an operation to the replica state, on an Amazon EC2 t2.small instance, at varying operation throughput rates.

### 2019-11-02 evaluation/run-all.sh
VM setup instructions

@@ -1,9 +1,22 @@
 #!/bin/bash
 set -e
 
+# To run the experiments, log in to the AWS EC2 console and start up a c5.large
+# instance running Ubuntu 18.04 in each of the three regions us-west-1, eu-west-1,
+# and ap-southeast-1. Then fill in their IP addresses here:
 US_WEST_1=18.144.1.184
 EU_WEST_1=34.255.190.108
 AP_SOUTHEAST_1=18.140.64.126
+
+# Set up the security groups such that you can log in to the VMs by SSH (TCP port 22),
+# and that they can all connect to each other on TCP port 8080.
+# Then, to set up the VMs, log in to each by SSH and run the following:
+
+# sudo apt-get update && sudo apt-get upgrade -y && sudo apt-get install -y apt-transport-https gnupg wget unzip
+# echo "deb https://dl.bintray.com/sbt/debian /" | sudo tee -a /etc/apt/sources.list.d/sbt.list
+# sudo apt-key adv --keyserver hkp://keyserver.ubuntu.com:80 --recv 2EE0EA64E40A89B84B2DF73499E82A75642AC823
+# sudo apt-get update && sudo apt-get install -y openjdk-8-jdk-headless sbt
+# git clone https://github.com/trvedata/move-op.git && cd move-op/evaluation && sbt compile
 
 run_batch () {
     INTERVAL=$1

### 2019-11-02 paper/move-op.tex
Tweaks for better line breaks in the paper

@@
 
-We can now represent the tree as a set of \isa{(parent, meta, child)} triples, denoted in Isabelle/HOL as \isa{('n {\isasymtimes} 'm {\isasymtimes} 'n) set}.
+We can represent the tree as a set of \isa{(parent, meta, child)} triples, denoted in Isabelle/HOL as \isa{('n {\isasymtimes} 'm {\isasymtimes} 'n) set}.
 When we have \isa{(p, m, c) {\isasymin} tree}, that means \isa{c} is a child of \isa{p} in the tree, with associated metadata \isa{m}.
@@
 We therefore produce variants of the definitions in Figure~\ref{fig:code} that are designed for execution rather than theorem proving.
-Rather than representing the tree as a set of (parent, metadata, child) triples, these definitions use a hash-map in which the keys are child nodes, and the values are (metadata, parent) pairs, written in Isabelle/HOL as \isa{('n::\{hashable\}, 'm {\isasymtimes} 'n) hm}.
+Rather than representing the tree as a set of (parent, meta, child) triples, these definitions use a hash-map in which the keys are child nodes, and the values are (meta, parent) pairs, written in Isabelle/HOL as \isa{('n::\{hashable\}, 'm {\isasymtimes} 'n) hm}.
 The \isa{hashable} type class means that keys must have a hashing function.

### 2019-11-03 paper/move-op.tex
Update paper with new evaluation results

@@
     However, such a move operation is difficult to implement correctly if different replicas can concurrently perform arbitrary move operations, and we demonstrate bugs in Google Drive and Dropbox that arise with concurrent moves.
-    In this paper we present an algorithm that handles arbitrary concurrent modifications on trees, while ensuring that the tree structure remains valid (in particular, no cycles are introduced), and guaranteeing that all replicas converge towards the same consistent state.
-    We formally prove the correctness of our algorithm using the Isabelle/HOL proof assistant, and we evaluate the performance of our formally verified implementation in a geo-replicated setting.
+    In this paper we present a CRDT algorithm that handles arbitrary concurrent modifications on trees, while ensuring that the tree structure remains valid (in particular, no cycles are introduced), and guaranteeing that all replicas converge towards the same consistent state.
+    We formally prove the correctness of our algorithm using the Isabelle/HOL proof assistant.
+    Our formally verified implementation achieves a throughput of 600 operations/sec in a geo-replicated setting.
 \end{abstract}
@@
 
-We used Isabelle/HOL's code generation mechanism to extract executable Scala code from our HOL definitions, as discussed in \S\ref{subsect.extracting}, and wrapped this implementation in a network service using the Akka~\cite{Akka} actor framework, using gRPC~\cite{gRPC} for network communication.
-We deployed three replicas of this service on Amazon EC2 \texttt{t2.small} instances in different regions worldwide: one in Northern California (region \texttt{us-west-1}), one in Ireland (\texttt{eu-west-1}), and one in Singapore (\texttt{ap-southeast-1}).
-The network delays between these regions are substantial; see Table~\ref{tab:rpc-times}.
-
-We use a synthetic workload in which each replica generates move operations at the same rate; for each move operation the generating replica chooses parent and child nodes uniformly at random from a set of 1,000 tree nodes.
-For a given operation rate we run the system for 10 minutes to reach a steady state.
-We use Lamport timestamps~\cite{Lamport:1978jq} as operation timestamps, and string values to identify tree nodes.
-When a replica generates an operation, it immediately applies that operation to its local state, and asynchronously sends the operation to the other two replicas via gRPC requests.
+We used Isabelle/HOL's code generation mechanism to extract executable Scala code from our HOL definitions, as discussed in \S\ref{subsect.extracting}, and wrapped this implementation in a simple network service.
+We deployed three replicas of this service on Amazon EC2 \texttt{c5.large} instances in different regions worldwide: in Northern California (region \texttt{us-west-1}), Ireland (\texttt{eu-west-1}), and Singapore (\texttt{ap-southeast-1}).
+The network delays between these regions are substantial, as shown in Table~\ref{tab:rpc-times}.
+
+We use a synthetic workload in which each replica starts with an empty tree, and then generates move operations at a fixed rate; for each move operation the generating replica chooses parent and child nodes uniformly at random from a set of 1,000 tree nodes.
+A tree node is created by the first operation that refers to it.
+We use Lamport timestamps~\cite{Lamport:1978jq} as operation timestamps, and 64-bit integers as tree node identifiers.
+When a replica generates an operation, it immediately applies that operation to its local state, and asynchronously sends the operation to the other two replicas via TCP connections.
 When a replica receives an operation from a remote replica, it also applies that operation to its own replica state.
@@
 
-We varied the rate at which replicas generated operations, and measured the execution time of our algorithm for applying each operation.
+For a given operation rate we ran the system for 10 minutes to reach a steady state.
+We repeated the experiment with different operation rates, and measured the execution time of our algorithm for applying each operation.
 Figure~\ref{fig:plots} shows the median execution times for local and remote operations respectively.
-The execution time for local operations is near-constant (median times mostly between 0.2~ms and 0.6~ms), while the execution time for remote operations increases approximately linearly from 0.2~ms to 2.3~ms over the range of 3 to 24 operations/sec.
-Above 25 operations/sec the execution time for remote operations suddenly increases to over 30~ms (off the scale in Figure~\ref{fig:plots}).
-We do not plot the exact execution times above 12~ms because they are less meaningful, as explained below.
+The execution time for local operations is near-constant (median times between 50~{\textmu}s and 70~{\textmu}s), while the execution time for remote operations increases approximately linearly from 0.14~ms to 1.5~ms over the range of 30 to 600 operations/sec.
+At around 600 operations/sec each single-threaded replica is fully utilising one CPU core, and increasing the rate at which operations are generated does not increase throughput any further.
 
 \begin{table}
-  \caption{Median response times of a RPC request between replicas in different regions.}
+  \caption{Median response times of an RPC request between replicas in different regions.}
   \label{tab:rpc-times}
@@
   \includegraphics[width=\columnwidth,keepaspectratio=true]{plot.pdf}
-  \caption{Median execution time for applying an operation to the replica state, on an Amazon EC2 t2.small instance, at varying operation throughput rates.
+  \caption{Median execution time for applying an operation to the replica state, on an Amazon EC2 c5.large instance, at varying operation throughput rates.
 Error bars indicate the minimum and 95th percentile execution times.}
   \label{fig:plots}
-  \Description{Plot in which the x axis goes from 0 to 30 move operations per second, and the y axis goes from 0 to 12 milliseconds.
-Local operations take a near-constant time less than 1ms; the time for remote operations gradually increases to 2ms until 24 operations/sec, at which point it suddenly shoots up off the top of the scale.}
+  \Description{Plot in which the x axis goes from 0 to 600 move operations per second, and the y axis goes from 0 to 2 milliseconds.
+Local operations take a near-constant time around 50 microseconds; the time for remote operations increases linearly from 0.2ms to 1.4ms.}
 \end{figure}
@@
 Applying that operation requires only a \isa{do\_op}, and no \isa{undo\_op} or \isa{redo\_op}, hence the near-constant execution time.
+
 However, for remote operations, a number of undos and redos must be performed, depending on the degree of concurrency.
-As the rate of operations increases, the time interval between successive operations becomes smaller than the network delay between replicas; thus, there are more operations ``in flight'' at the same time, and more calls to \isa{undo\_op} and \isa{redo\_op} are required to order operations by timestamp.
-
-The time taken to apply a remote operation is proportional to the number of undos and redos required.
-Above 25 operations/sec the CPU utilisation of our virtual machines grows to the point that incoming network requests from other replicas start queueing before they are processed.
-When a remote operation spends time queueing, more concurrent operations have been applied by the time it is processed, and thus more undos and redos are required to apply the incoming operation, which in turn increases the CPU utilisation and further increases queueing.
-This self-reinforcing effect explains the sudden dramatic increase in Figure~\ref{fig:plots}.
-
-Above 25 operations/sec the queueing delay increases so much that some RPC requests time out, and thus not all operations are received by all replicas.
-As the remote operation execution times above this threshold depend more on the network queueing properties than our algorithm, we discard these datapoints.
-
-\subsubsection{Practical relevance of results}
-
-A hand-optimised implementation of our algorithm could achieve higher throughput than our 24 operations/sec, since code generated from HOL definitions contains inefficiencies.
+As the time interval between successive operations is smaller than the network delay between replicas, there are multiple operations ``in flight'' at the same time, and more calls to \isa{undo\_op} and \isa{redo\_op} are required to order operations by timestamp.
+The time taken to apply a remote operation is proportional to the number of undos and redos required, and hence proportional to the operation rate.
+
+A hand-optimised implementation of our algorithm could achieve higher throughput than our 600 operations/sec, since code generated from HOL definitions contains inefficiencies.
 The generated code could serve as a reference implementation against which other implementations could be tested.
@@
 The alternative to our optimistic replication approach would be to use a central server (master replica) or consensus protocol that processes all operations in a serialisable order.
-Performing an operation then requires waiting for communication with a server potentially located on a different continent, taking hundreds of milliseconds (Table~\ref{tab:rpc-times}); by contrast, a local operation taking a fraction of a millisecond is orders of magnitude faster.
+Performing an operation then requires waiting for communication with a server potentially located on a different continent, taking over 100~ms (Table~\ref{tab:rpc-times}); by contrast, a local operation taking around 50~{\textmu}s is at least three orders of magnitude faster.
 Moreover, a local operation requires no waiting for network communication, and thus it supports disconnected operation: mobile devices can read and modify the replicated tree even while disconnected from the Internet, which is not the case with centralised systems.
@@
 We also extracted a formally verified Scala implementation from our Isabelle/HOL definitions, and evaluated its performance across three replicas in California, Ireland and Singapore.
-Our implementation applies local updates in under 1~ms, orders of magnitude faster than is possible with a central server or consensus protocol operating over these distances.
-It is able to sustain 24 updates per second; we plan to develop a more scalable implementation in future work.
+Our implementation applies local updates in approximately 50~{\textmu}s, orders of magnitude faster than is possible with a central server or consensus protocol operating over these distances.
+It is able to sustain 600 updates per second; a more scalable implementation could be created by optimising the automatically extracted Scala code.
 

### 2019-11-03 paper/move-op.tex
paper: small addition in Section 3.2
paper: small addition in conclusions
Added discussion of proof vs. model checking in conclusions.

@@
 Due to the undo-do-redo cycle, the state of the tree is as if all operations had been applied using \isa{do\_op} in increasing timestamp order, regardless of the order in which they were actually applied.
-This provides a clear approach to the handling of conflicts:
+This provides a clear and consistent approach to the handling of conflicts:
 \begin{itemize}
@@
 We formally verified the correctness of our algorithm using the Isabelle/HOL proof assistant; our theorems show that replicas can apply operations in any order, and that the result is always a valid tree (nodes have at most one parent, and the graph does not contain any cycles).
+Moreover, these results are \emph{strong} in the sense that they apply to unbounded executions and an arbitrary number of replicas: an inherent advantage of using proof rather than model checking.
 

### 2019-11-03 paper/move-op.tex
paper: some rewording of additions in conclusion

--- 2019-11-03 paper/move-op.tex
@@
 We formally verified the correctness of our algorithm using the Isabelle/HOL proof assistant; our theorems show that replicas can apply operations in any order, and that the result is always a valid tree (nodes have at most one parent, and the graph does not contain any cycles).
+Moreover, these results apply to unbounded executions and an arbitrary number of replicas---an inherent advantage of using a proof assistant rather than other formal approaches such as model checking.
 

### 2019-11-03 paper/move-op.tex
paper: further small addition
paper: a little on future work

@@
 
-Gomes et al.~\cite{Gomes:2017gy} define a framework in Isabelle/HOL for proving the strong eventual consistency properties of CRDTs.
+Gomes et al.~\cite{Gomes:2017gy} define a framework in Isabelle/HOL for proving the strong eventual consistency properties of CRDTs above an axiomatic network model.
 Using our commutativity proof above we integrate our tree datatype with this framework, and thus prove that our move operation on trees does indeed guarantee strong eventual consistency.
@@
 
+As mentioned in Section~\ref{sec:intro}, the work presented in this paper is especially interesting due to ubiquity of a tree-structured data model.
+In future work we hope to build novel peer-to-peer distributed systems, such as a distributed file-system, that make essential use of our algorithm.
+
 \begin{acks}

### 2019-11-03 paper/move-op.tex
paper: reword last paragraph a little

@@
 
-As mentioned in Section~\ref{sec:intro}, the work presented in this paper is especially interesting due to ubiquity of a tree-structured data model.
-In future work we hope to build novel peer-to-peer distributed systems, such as a distributed file-system, that make essential use of our algorithm.
+As mentioned in the introduction, the work presented in this paper is especially interesting due to the ubiquity of the tree-structured data model across a range of different computing systems.
+In future work we hope to build novel peer-to-peer distributed systems, such as a distributed file-system, that make use of our algorithm.
 

### 2019-11-03 paper/move-op.tex
paper: final fiddling
More rewording of closing paragraph.

@@
 
-As mentioned in the introduction, the work presented in this paper is especially interesting due to the ubiquity of the tree-structured data model across a range of different computing systems.
+As mentioned in the introduction, our work is especially interesting due to the ubiquity of the tree-structured data model across a range of different computing systems.
 In future work we hope to build novel peer-to-peer distributed systems, such as a distributed file-system, that make use of our algorithm.

### 2019-11-04 paper/move-op.tex
Update acknowledgements
Mention some stats about Isabelle proof in the evaluation section

--- 2019-11-03 paper/move-op.tex
@@
 
+\subsection{Evaluation of formal proof}
+
+The formalisation of our algorithm, and the proofs of its properties as described in \S\ref{sec:proof}, have been formally checked by Isabelle/HOL.
+Our proofs contain no unproven assumptions (i.e. no occurrences of the \textbf{sorry} keyword).
+Checking all of the proofs takes 3.5 minutes on a 2018 MacBook Pro.
+
+Besides the 60 lines of definitions given in Figure~\ref{fig:code}, our Isabelle/HOL formalisation consists of a further 2,166 lines of code.
+Of this, we use 203 lines to prove that every node has a unique parent, 443 lines to prove that the tree contains no cycles, 450 lines to prove that move operations commute and replicas converge, 327 lines to prove the strong eventual consistency of our algorithm using the framework of Gomes et al.~\cite{Gomes:2017gy}, and 743 lines to define the executable variant of our algorithm (\S\ref{subsect.extracting}) and prove its equivalence to the abstract definitions of Figure~\ref{fig:code}.
+
 \section{Related work}\label{sec:relwork}
@@
 
-As mentioned in Section~\ref{sec:intro}, the work presented in this paper is especially interesting due to ubiquity of a tree-structured data model.
-In future work we hope to build novel peer-to-peer distributed systems, such as a distributed file-system, that make essential use of our algorithm.
+As mentioned in the introduction, our work is especially interesting due to the ubiquity of tree data models across many different computing systems.
+In future work we hope to build novel peer-to-peer distributed systems, such as a distributed filesystem, based on our algorithm.
 
@@
 The authors wish to acknowledge the support of The Boeing Company and the EPSRC ``REMS: Rigorous Engineering for Mainstream Systems'' programme grant (EP/K008528).
-% AWS credits from the AWS Educate program
+Martin Kleppmann is supported by a Leverhulme Trust Early Career Fellowship and by the Isaac Newton Trust.
+Our evaluation was conducted using AWS credits from the AWS Educate program.
+Thank you to Marc Shapiro for feedback on a draft of this paper.
 \end{acks}

### 2020-02-10 paper/list-move.tex
Start paper on list reordering

\documentclass[sigplan,10pt]{acmart}

\usepackage{tikz}

\setcopyright{rightsretained}
\copyrightyear{2020}
\acmYear{2020}
\acmDOI{}
\acmConference[]{}{}{}
\acmBooktitle{}
\acmPrice{}
\acmISBN{}

\hyphenation{da-ta-cen-ter da-ta-cen-ters time-stamp time-stamps time-stamped hard-links}

\begin{document}
\title{Reordering Elements in List CRDTs}

\author{Martin Kleppmann}
\email{mk428@cl.cam.ac.uk}
\orcid{0000-0001-7252-6958}
\affiliation{%
  \institution{Department of Computer Science and Technology}
  \streetaddress{15 JJ Thomson Avenue}
  \city{Cambridge}
  \state{}
  \postcode{CB3 0FD}
  \country{United Kingdom}
}

\begin{abstract}
\end{abstract}

\settopmatter{printfolios=true} % Recommended EuroSys formatting
\maketitle

\section{Introduction}

Conflict-free Replicated Data Types (CRDTs) allow multiple replicas to concurrently modify some shared data object, while ensuring that all replicas eventually converge towards the same, consistent state~\cite{Shapiro:2011un}.
CRDT algorithms have been developed for various different datatypes, and one of the most important datatypes is the \emph{list} (also called \emph{sequence} or \emph{array} datatype).
A list is a collection of elements in a total order, which is chosen by the user.

Replicated lists can be used to implement a variety of important applications, such as:
\begin{itemize}
    \item text editors (text is a list of characters);
    \item graphics applications (a list of graphical objects, where the order determines visibility: objects ``further down'' may be occluded or filtered by the objects ``higher up'', as illustrated in Fig.~\ref{fig:smiley});
    \item to-do lists and task trackers (a list of tasks, where the order may reflect the relative priority of tasks as chosen by the user).
\end{itemize}

\begin{figure*}
  \centering
  \includegraphics{smiley.pdf}
  \caption{Graphics software often places objects in a list that determines the z-index; objects higher in the list occlude lower ones. %
    Here, the eyeballs are placed above the yellow face, and the pupils in turn are placed above the eyeballs.}
  \label{fig:smiley}
\end{figure*}

Many list CRDTs have been developed, such as WOOT~\cite{Oster:2006wj}, Treedoc~\cite{Preguica:2009fz}, RGA~\cite{Roh:2011dw}, Causal Trees/Timestamped Insertion Trees~\cite{Grishchenko:2014eh,Attiya:2016kh}, Logoot \cite{Weiss:2009ht,Weiss:2010hx}, and LSEQ \cite{Nedelec:2013ky,Nedelec:2016eo}.
Moreover, most of the field of Operational Transformation algorithms is dedicated to algorithms for collaboratively editable text, i.e.\ lists~\cite{Ellis:1989ue,Nichols:1995fd,Ressel:1996wx,Sun:1998vf,Oster:2006tr}.

All of the aforementioned algorithms allow replicas to \emph{insert} or \emph{delete} elements anywhere in the list.
However, none of them have explicit support for \emph{reordering} or \emph{moving} elements from one position to another position in the list.
This is a surprising omission because reordering is a commonly required operation: in many to-do list applications, a user can drag and drop list elements to reorder them, and graphics software allows users to reorder the object list with commands such as ``bring to front'' (which moves an object to the top of the list) and ``send to back'' (which moves an object to the bottom).

In this paper we introduce an algorithm that allows an existing list CRDT to be extended with a reorder operation.
The algorithm is generic in the sense that it can be implemented on top of any of the aforementioned list CRDTs.

\section{A Generic Reordering Algorithm}

\begin{acks}
Martin Kleppmann is supported by a Leverhulme Trust Early Career Fellowship and by the Isaac Newton Trust.
\end{acks}

\bibliographystyle{ACM-Reference-Format}
\bibliography{references}{}
\end{document}

### 2020-02-10 paper/list-move.tex
Discuss duplication problem with concurrent moves

@@
 \usepackage{tikz}
+\usetikzlibrary{arrows.meta}
 
@@
   \caption{Graphics software often places objects in a list that determines the z-index; objects higher in the list occlude lower ones. %
-    Here, the eyeballs are placed above the yellow face, and the pupils in turn are placed above the eyeballs.}
+    Here, the eyeballs and mouth are placed above the yellow face, and the pupils in turn are placed above the eyeballs.}
   \label{fig:smiley}
@@
 
+\section{Semantics of Concurrent Moves}
+
+The simplest way of moving a list element is to delete it from its existing position, and to re-insert it at the new position.
+However, if two replicas concurrently move the same element, the result is that the element is duplicated: the two deletions of the same element behave the same as one deletion, and the two insertions independently re-insert the element twice.
+
+\begin{figure*}
+    \centering
+    \vspace{3em}
+    \begin{tikzpicture}[auto,scale=1.0]
+        \tikzstyle{item}=[rectangle,draw,inner ysep=4pt,text height=9pt,text depth=2pt,minimum width=2.2cm]
+        \tikzstyle{network}=[thick,dashed,blue,-{Stealth[length=3mm]}]
+        \path [draw,dotted] (0,2) -- (17.5,2);
+        \node at (1, 4) {Replica $A$:};
+        \node at (1, 0) {Replica $B$:};
+        \matrix [row sep={19pt,between origins},nodes=item] at (4, 4) {
+            \node [minimum width=0.5cm] {1.}; & \node (top1) {buy milk}; \\
+            \node [minimum width=0.5cm] {2.}; & \node {water plants}; \\
+            \node [minimum width=0.5cm,fill=red!10] {3.}; & \node [fill=red!10] (joe1) {phone joe}; \\
+        };
+        \matrix [row sep={19pt,between origins},nodes=item] at (4, 0) {
+            \node [minimum width=0.5cm] {1.}; & \node (top2) {buy milk}; \\
+            \node [minimum width=0.5cm] {2.}; & \node {water plants}; \\
+            \node [minimum width=0.5cm,fill=red!10] {3.}; & \node [fill=red!10] (joe2) {phone joe}; \\
+        };
+        \draw [->,red,thick] (joe1.east) to [in=0, out=0] node [right,near end] {move} (top1.north east);
+        \draw [->,red,thick] (joe2.east) to [in=0, out=0] node [right,near end] {move} (top2.north east);
+        \draw [->,thick] (6, 0) -- (8.5, 0);
+        \draw [->,thick] (6, 4) -- (8.5, 4);
+        \matrix [row sep={19pt,between origins},nodes=item] at (10, 4) {
+            \node [minimum width=0.5cm,fill=red!10] {1.}; & \node [fill=red!10] {phone joe}; \\
+            \node [minimum width=0.5cm] {2.}; & \node {buy milk}; \\
+            \node [minimum width=0.5cm] {3.}; & \node {water plants}; \\
+        };
+        \matrix [row sep={19pt,between origins},nodes=item] at (10, 0) {
+            \node [minimum width=0.5cm,fill=red!10] {1.}; & \node [fill=red!10] {phone joe}; \\
+            \node [minimum width=0.5cm] {2.}; & \node {buy milk}; \\
+            \node [minimum width=0.5cm] {3.}; & \node {water plants}; \\
+        };
+        \draw [->,thick] (11.5, 0) -- (14.5, 0);
+        \draw [->,thick] (11.5, 4) -- (14.5, 4);
+        \draw [network] (12.5, 0) to [out=90,in=270] (13.1, 4);
+        \draw [network] (12.5, 4) to [out=270,in=90] (13.1, 0);
+        \node [rotate=90,blue,font=\footnotesize] at (13.5, 2) {network communication};
+        \matrix [row sep={19pt,between origins},nodes=item] at (16, 4) {
+            \node [minimum width=0.5cm,fill=red!10] {1.}; & \node [fill=red!10] {phone joe}; \\
+            \node [minimum width=0.5cm,fill=red!10] {2.}; & \node [fill=red!10] {phone joe}; \\
+            \node [minimum width=0.5cm] {3.}; & \node {buy milk}; \\
+            \node [minimum width=0.5cm] {4.}; & \node {water plants}; \\
+        };
+        \matrix [row sep={19pt,between origins},nodes=item] at (16, 0) {
+            \node [minimum width=0.5cm,fill=red!10] {1.}; & \node [fill=red!10] {phone joe}; \\
+            \node [minimum width=0.5cm,fill=red!10] {2.}; & \node [fill=red!10] {phone joe}; \\
+            \node [minimum width=0.5cm] {3.}; & \node {buy milk}; \\
+            \node [minimum width=0.5cm] {4.}; & \node {water plants}; \\
+        };
+    \end{tikzpicture}
+    \caption{If reordering is implemented by deleting and re-inserting, concurrent moves of the same item result in the anomaly shown here: the moved item is duplicated.}
+    \label{fig:duplication}
+\end{figure*}
+
+\begin{figure*}
+    \centering
+    \vspace{3em}
+    \begin{tikzpicture}[auto,scale=1.0]
+        \tikzstyle{item}=[rectangle,draw,inner ysep=4pt,text height=9pt,text depth=2pt,minimum width=2.2cm]
+        \tikzstyle{network}=[thick,dashed,blue,-{Stealth[length=3mm]}]
+        \path [draw,dotted] (0,1.5) -- (17.5,1.5);
+        \node at (1, 3) {Replica $A$:};
+        \node at (1, 0) {Replica $B$:};
+        \matrix [row sep={19pt,between origins},nodes=item] at (4, 3) {
+            \node [minimum width=0.5cm] {1.}; & \node (top1) {buy milk}; \\
+            \node [minimum width=0.5cm] {2.}; & \node {water plants}; \\
+            \node [minimum width=0.5cm,fill=red!10] {3.}; & \node [fill=red!10] (joe1) {phone joe}; \\
+        };
+        \matrix [row sep={19pt,between origins},nodes=item] at (4, 0) {
+            \node [minimum width=0.5cm] {1.}; & \node (top2) {buy milk}; \\
+            \node [minimum width=0.5cm] {2.}; & \node {water plants}; \\
+            \node [minimum width=0.5cm,fill=red!10] {3.}; & \node [fill=red!10] (joe2) {phone joe}; \\
+        };
+        \draw [->,red,thick] (joe1.east) to [in=0, out=0] node [right,near end] {move} (top1.north east);
+        \draw [->,red,thick] (joe2.east) to [in=0, out=0] node [right,at end,inner xsep=8pt] {move} (top2.south east);
+        \draw [->,thick] (6, 0) -- (8.5, 0);
+        \draw [->,thick] (6, 3) -- (8.5, 3);
+        \matrix [row sep={19pt,between origins},nodes=item] at (10, 3) {
+            \node [minimum width=0.5cm,fill=red!10] {1.}; & \node [fill=red!10] {phone joe}; \\
+            \node [minimum width=0.5cm] {2.}; & \node {buy milk}; \\
+            \node [minimum width=0.5cm] {3.}; & \node {water plants}; \\
+        };
+        \matrix [row sep={19pt,between origins},nodes=item] at (10, 0) {
+            \node [minimum width=0.5cm] {1.}; & \node {buy milk}; \\
+            \node [minimum width=0.5cm,fill=red!10] {2.}; & \node [fill=red!10] {phone joe}; \\
+            \node [minimum width=0.5cm] {3.}; & \node {water plants}; \\
+        };
+        \draw [->,thick] (11.5, 0) -- (14.5, 0);
+        \draw [->,thick] (11.5, 3) -- (14.5, 3);
+        \draw [network] (12.5, 0) to [out=90,in=270] (13.1, 3);
+        \draw [network] (12.5, 3) to [out=270,in=90] (13.1, 0);
+        \node [rotate=90,blue,font=\footnotesize] at (13.5, 1.5) {network communication};
+        \matrix [row sep={19pt,between origins},nodes=item] at (16, 3) {
+            \node [minimum width=0.5cm,fill=red!10] {1.}; & \node [fill=red!10] {phone joe}; \\
+            \node [minimum width=0.5cm] {2.}; & \node {buy milk}; \\
+            \node [minimum width=0.5cm] {3.}; & \node {water plants}; \\
+        };
+        \matrix [row sep={19pt,between origins},nodes=item] at (16, 0) {
+            \node [minimum width=0.5cm,fill=red!10] {1.}; & \node [fill=red!10] {phone joe}; \\
+            \node [minimum width=0.5cm] {2.}; & \node {buy milk}; \\
+            \node [minimum width=0.5cm] {3.}; & \node {water plants}; \\
+        };
+    \end{tikzpicture}
+    \caption{When the same item is concurrently moved to different positions on different replicas, one of those positions should become the ``winning'' position as the replicas converge.
+In this example, replica $A$'s move operation wins.}
+    \label{fig:concurrent}
+\end{figure*}
+
+We argue that such duplication in the face of concurrency is undesirable, as it is generally not the behaviour expected by users.
+For example, say a user has replicas of their to-do list on their laptop and their smartphone.
+They reorder items on their laptop to match new priorities, and also reorder some items on their phone while the phone is offline, as illustrated in Fig.~\ref{fig:duplication}.
+Later, when the phone is online again and synchronises with the laptop, any items that were moved on both devices will be duplicated.
+
+This behaviour is especially confusing if the same item was moved to the same position (e.g.\ to the top of the list) on both devices, in which case the duplicated items are likely to be adjacent in the final list, as in Fig.~\ref{fig:duplication} (depending on any other insertions that might have taken place).
+However, even if the concurrent move operations had different destination positions, duplicating the moved item at both destinations is unlikely to be desirable behaviour in most applications.
+
+Rather, we argue that the best semantics in this situation is for the system to pick one of the destination positions as the ``winner''; as the replicas communicate, they all converge towards a list in which the moved element is located at the winning position, and nowhere else.
+This approach is illustrated in Fig.~\ref{fig:concurrent}.
+Any deterministic method of picking the winner is suitable (e.g.\ based on a priority given to each replica, or based on the timestamp of the operations).
+
 \section{A Generic Reordering Algorithm}

### 2020-02-10 paper/list-move.tex
Initial description of algorithm

@@
 \begin{abstract}
+    List CRDTs (Conflict-free Replicated Data Types) allow multiple users to concurrently insert and delete items in a shared list.
+    However, existing algorithms behave poorly when users concurrently reorder items in the list, or move items to a new position.
+    We describe an extension to list CRDTs that improves their support for reordering.
 \end{abstract}
@@
 
-\section{Semantics of Concurrent Moves}
+\section{Semantics of Concurrent Moves}\label{sec:semantics}
 
@@
 
-Rather, we argue that the best semantics in this situation is for the system to pick one of the destination positions as the ``winner''; as the replicas communicate, they all converge towards a list in which the moved element is located at the winning position, and nowhere else.
+Rather, we argue that the best semantics in this situation is for the system to pick one of the destination positions as the ``winner''; as the replicas communicate, they all converge towards a list in which the moved element is located only at the winning position, and nowhere else.
 This approach is illustrated in Fig.~\ref{fig:concurrent}.
@@
 
-\section{A Generic Reordering Algorithm}
+\section{A Generic Reordering Algorithm}\label{sec:algorithm}
+
+Intuitively, if we want to pick one winner from among several concurrent moves for the same element, we can use a long-established CRDT: a last-writer wins (LWW) register~\cite{Shapiro:2011wy,Johnson:1975we}.
+We need one such register for each list element, containing the position of that element.
+Moving an element to a new location is then merely a matter of updating the value of that element's register.
+
+In general, we could also use a multi-value (MV) register~\cite{Shapiro:2011wy}, which would allow an element to exist at multiple positions in the list simultaneously after having experienced conflicting moves.
+This is not the same as the duplication described in Section~\ref{sec:semantics}, since the multiple positions still belong to the same element, and a subsequent move would collapse them back into a single position again.
+The MV register option can be used if the LWW behaviour (discarding all but one of the concurrent moves of the same element) is unacceptable.
+
+Regardless of whether a LWW or MV register is used, this approach requires a stable way of referencing some position in the list (i.e.\ the value held by the register).
+An integer index is not suitable, because an insertion or deletion at any position would require changing all of the indices following that position.
+Fortunately, all known list CRDTs already incorporate schemes for stable position identifiers.
+The details vary: Treedoc uses a path through a binary tree~\cite{Preguica:2009fz}, Logoot uses a list of (integer, replicaID) pairs~\cite{Weiss:2009ht}, RGA uses a so-called s4vector~\cite{Roh:2011dw}, Causal Trees/Timestamped Insertion Trees use a timestamp~\cite{Grishchenko:2014eh,Attiya:2016kh}, and so on.
+But all of these approaches have in common that they can uniquely refer to a particular position in a list, in a way that is not affected by operations on other list positions.
+
+Assume now that $p_1, p_2, \dots$ are position identifiers according to one of the aforementioned schemes.
+We can now model the state of the list as a set of $(\mathit{value}, \mathit{register})$ pairs contained in an add-wins set CRDT:
+\begin{align*}
+\mathit{state} = \mathrm{AWSet}(\{ & (v_1, \mathrm{LWWRegister}(p_1)),\\
+    & (v_2, \mathrm{LWWRegister}(p_2)), \;\dots\})
+\end{align*}
+In some schemes (such as Treedoc, Logoot and LSEQ), the position identifiers themselves encode their relative ordering~\cite{Interleaving}, and thus no further state is required.
+In others (such as RGA and CT), the list CRDT requires an additional data structure that determines the order in which position identifiers appear in the list.
+
+In order to iterate over the current state of the list, we need to query the above set by the position identifiers stored within the registers.
+We can do this efficiently by using a database index, which allows us to find all the tuples within a set that contain a certain value (or range of values).
+
+Note that when the position of a list element is changed, the old position identifier no longer has any value associated with it, and thus it should be regarded as nonexistent (like a tombstone in some CRDTs).
+It is also possible for multiple list elements to have the same position identifier, and thus for several list elements to appear at the same position.
+This situation is similar to having a multi-value register of values at each list position.
+(Note this situation is different from our earlier use of a MV register, which was about one element having multiple positions, not one position having multiple elements.)
+
+We can avoid having multiple list elements with the same position identifier by always creating a fresh position identifier as the destination of a move operation.
+All of the aforementioned list CRDTs have methods of creating new position identifiers that are guaranteed to be globally unique.
 

### 2020-02-18 paper/list-move.tex
Add section on range moves

@@
 
+\usepackage[normalem]{ulem} % \sout{} for strikethrough text
 \usepackage{tikz}
-\usetikzlibrary{arrows.meta}
+\usetikzlibrary{arrows.meta,snakes}
 
@@
 \affiliation{%
-  \institution{Department of Computer Science and Technology}
+  \institution{University of Cambridge}
   \streetaddress{15 JJ Thomson Avenue}
@@
 \begin{abstract}
-    List CRDTs (Conflict-free Replicated Data Types) allow multiple users to concurrently insert and delete items in a shared list.
-    However, existing algorithms behave poorly when users concurrently reorder items in the list, or move items to a new position.
-    We describe an extension to list CRDTs that improves their support for reordering.
+Conflict-free Replicated Data Types (CRDTs) for lists allow multiple users to concurrently insert and delete elements in a shared list object.
+However, existing algorithms behave poorly when users concurrently reorder elements in the list (i.e.\ move list elements to a new position).
+We demonstrate the need for such a move operation, and describe an algorithm that extends a list CRDT with an explicit move operation.
+Our algorithm can be used in conjunction with any existing list CRDT algorithm.
+In addition to moving a single list element, we also discuss the open problem of moving ranges of elements.
 \end{abstract}
+
+\begin{CCSXML}
+<ccs2012>
+   <concept>
+       <concept_id>10003752.10003809.10010172</concept_id>
+       <concept_desc>Theory of computation~Distributed algorithms</concept_desc>
+       <concept_significance>500</concept_significance>
+   </concept>
+   <concept>
+       <concept_id>10003120.10003130.10003131.10003570</concept_id>
+       <concept_desc>Human-centered computing~Computer supported cooperative work</concept_desc>
+       <concept_significance>300</concept_significance>
+   </concept>
+   <concept>
+       <concept_id>10002951.10003227.10003233</concept_id>
+       <concept_desc>Information systems~Collaborative and social computing systems and tools</concept_desc>
+       <concept_significance>300</concept_significance>
+   </concept>
+   <concept>
+       <concept_id>10010520.10010521.10010537.10010540</concept_id>
+       <concept_desc>Computer systems organization~Peer-to-peer architectures</concept_desc>
+       <concept_significance>300</concept_significance>
+   </concept>
+ </ccs2012>
+\end{CCSXML}
+
+\ccsdesc[500]{Theory of computation~Distributed algorithms}
+\ccsdesc[300]{Human-centered computing~Computer supported cooperative work}
+\ccsdesc[300]{Information systems~Collaborative and social computing systems and tools}
+\ccsdesc[300]{Computer systems organization~Peer-to-peer architectures}
+
+\keywords{conflict-free replicated data types, distributed consistency, collaborative editing, optimistic replication}
 
@@
 \begin{itemize}
-    \item text editors (text is a list of characters);
-    \item graphics applications (a list of graphical objects, where the order determines visibility: objects ``further down'' may be occluded or filtered by the objects ``higher up'', as illustrated in Fig.~\ref{fig:smiley});
+    \item collaborative text editors (text is a list of characters);
+    \item collaborative graphics applications (a list of graphical objects, where the order determines visibility: objects ``further down'' may be occluded or filtered by the objects ``higher up'', as illustrated in Figure~\ref{fig:smiley});
     \item to-do lists and task trackers (a list of tasks, where the order may reflect the relative priority of tasks as chosen by the user).
@@
 However, none of them have explicit support for \emph{reordering} or \emph{moving} elements from one position to another position in the list.
-This is a surprising omission because reordering is a commonly required operation: in many to-do list applications, a user can drag and drop list elements to reorder them, and graphics software allows users to reorder the object list with commands such as ``bring to front'' (which moves an object to the top of the list) and ``send to back'' (which moves an object to the bottom).
+This is a surprising omission because reordering is a commonly required operation: in many to-do list applications, a user can drag and drop list elements to reorder them, and graphics software allows users to reorder the object list with commands such as ``bring to front'' (which moves an object to the top of the list, so that it occludes other objects) and ``send to back'' (which moves an object to the bottom, so that it is occluded by other objects).
 
@@
 
+\definecolor{insertcolor}{rgb}{0, 0.5, 0}
+\definecolor{deletecolor}{rgb}{1, 0, 0}
+\definecolor{movecolor}{rgb}{0.5, 0, 1}
+\definecolor{networkcolor}{rgb}{0, 0.5, 1}
+
 \begin{figure*}
@@
         \tikzstyle{item}=[rectangle,draw,inner ysep=4pt,text height=9pt,text depth=2pt,minimum width=2.2cm]
-        \tikzstyle{network}=[thick,dashed,blue,-{Stealth[length=3mm]}]
+        \tikzstyle{network}=[thick,dashed,networkcolor,-{Stealth[length=3mm]}]
         \path [draw,dotted] (0,2) -- (17.5,2);
@@
             \node [minimum width=0.5cm] {2.}; & \node {water plants}; \\
-            \node [minimum width=0.5cm,fill=red!10] {3.}; & \node [fill=red!10] (joe1) {phone joe}; \\
+            \node [minimum width=0.5cm,fill=movecolor!10] {3.}; & \node [fill=movecolor!10] (joe1) {phone joe}; \\
         };
@@
             \node [minimum width=0.5cm] {2.}; & \node {water plants}; \\
-            \node [minimum width=0.5cm,fill=red!10] {3.}; & \node [fill=red!10] (joe2) {phone joe}; \\
+            \node [minimum width=0.5cm,fill=movecolor!10] {3.}; & \node [fill=movecolor!10] (joe2) {phone joe}; \\
         };
-        \draw [->,red,thick] (joe1.east) to [in=0, out=0] node [right,near end] {move} (top1.north east);
-        \draw [->,red,thick] (joe2.east) to [in=0, out=0] node [right,near end] {move} (top2.north east);
+        \draw [->,movecolor,thick] (joe1.east) to [in=0, out=0] node [right,near end] {move} (top1.north east);
+        \draw [->,movecolor,thick] (joe2.east) to [in=0, out=0] node [right,near end] {move} (top2.north east);
         \draw [->,thick] (6, 0) -- (8.5, 0);
@@
         \matrix [row sep={19pt,between origins},nodes=item] at (10, 4) {
-            \node [minimum width=0.5cm,fill=red!10] {1.}; & \node [fill=red!10] {phone joe}; \\
+            \node [minimum width=0.5cm,fill=movecolor!10] {1.}; & \node [fill=movecolor!10] {phone joe}; \\
             \node [minimum width=0.5cm] {2.}; & \node {buy milk}; \\
@@
         \matrix [row sep={19pt,between origins},nodes=item] at (10, 0) {
-            \node [minimum width=0.5cm,fill=red!10] {1.}; & \node [fill=red!10] {phone joe}; \\
+            \node [minimum width=0.5cm,fill=movecolor!10] {1.}; & \node [fill=movecolor!10] {phone joe}; \\
             \node [minimum width=0.5cm] {2.}; & \node {buy milk}; \\
@@
         \draw [network] (12.5, 4) to [out=270,in=90] (13.1, 0);
-        \node [rotate=90,blue,font=\footnotesize] at (13.5, 2) {network communication};
+        \node [rotate=90,networkcolor,font=\footnotesize] at (13.5, 2) {network communication};
         \matrix [row sep={19pt,between origins},nodes=item] at (16, 4) {
-            \node [minimum width=0.5cm,fill=red!10] {1.}; & \node [fill=red!10] {phone joe}; \\
-            \node [minimum width=0.5cm,fill=red!10] {2.}; & \node [fill=red!10] {phone joe}; \\
+            \node [minimum width=0.5cm,fill=movecolor!10] {1.}; & \node [fill=movecolor!10] {phone joe}; \\
+            \node [minimum width=0.5cm,fill=movecolor!10] {2.}; & \node [fill=movecolor!10] {phone joe}; \\
             \node [minimum width=0.5cm] {3.}; & \node {buy milk}; \\
@@
         \matrix [row sep={19pt,between origins},nodes=item] at (16, 0) {
-            \node [minimum width=0.5cm,fill=red!10] {1.}; & \node [fill=red!10] {phone joe}; \\
-            \node [minimum width=0.5cm,fill=red!10] {2.}; & \node [fill=red!10] {phone joe}; \\
+            \node [minimum width=0.5cm,fill=movecolor!10] {1.}; & \node [fill=movecolor!10] {phone joe}; \\
+            \node [minimum width=0.5cm,fill=movecolor!10] {2.}; & \node [fill=movecolor!10] {phone joe}; \\
             \node [minimum width=0.5cm] {3.}; & \node {buy milk}; \\
@@
     \end{tikzpicture}
-    \caption{If reordering is implemented by deleting and re-inserting, concurrent moves of the same item result in the anomaly shown here: the moved item is duplicated.}
+    \caption{If reordering is implemented by deleting and re-inserting, concurrent moves of the same element result in the anomaly shown here: the moved element is duplicated.}
     \label{fig:duplication}
@@
         \tikzstyle{item}=[rectangle,draw,inner ysep=4pt,text height=9pt,text depth=2pt,minimum width=2.2cm]
-        \tikzstyle{network}=[thick,dashed,blue,-{Stealth[length=3mm]}]
+        \tikzstyle{network}=[thick,dashed,networkcolor,-{Stealth[length=3mm]}]
         \path [draw,dotted] (0,1.5) -- (17.5,1.5);
@@
             \node [minimum width=0.5cm] {2.}; & \node {water plants}; \\
-            \node [minimum width=0.5cm,fill=red!10] {3.}; & \node [fill=red!10] (joe2) {phone joe}; \\
+            \node [minimum width=0.5cm,fill=movecolor!10] {3.}; & \node [fill=movecolor!10] (joe2) {phone joe}; \\
         };
-        \draw [->,red,thick] (joe1.east) to [in=0, out=0] node [right,near end] {move} (top1.north east);
-        \draw [->,red,thick] (joe2.east) to [in=0, out=0] node [right,at end,inner xsep=8pt] {move} (top2.south east);
+        \draw [->,movecolor,thick] (joe1.east) to [in=0, out=0] node [right,near end] {move} (top1.north east);
+        \draw [->,movecolor,thick] (joe2.east) to [in=0, out=0] node [right,at end,inner xsep=8pt] {move} (top2.south east);
         \draw [->,thick] (6, 0) -- (8.5, 0);
@@
         \matrix [row sep={19pt,between origins},nodes=item] at (10, 3) {
-            \node [minimum width=0.5cm,fill=red!10] {1.}; & \node [fill=red!10] {phone joe}; \\
+            \node [minimum width=0.5cm,fill=movecolor!10] {1.}; & \node [fill=movecolor!10] {phone joe}; \\
             \node [minimum width=0.5cm] {2.}; & \node {buy milk}; \\
@@
             \node [minimum width=0.5cm] {1.}; & \node {buy milk}; \\
-            \node [minimum width=0.5cm,fill=red!10] {2.}; & \node [fill=red!10] {phone joe}; \\
+            \node [minimum width=0.5cm,fill=movecolor!10] {2.}; & \node [fill=movecolor!10] {phone joe}; \\
             \node [minimum width=0.5cm] {3.}; & \node {water plants}; \\
@@
         \draw [network] (12.5, 3) to [out=270,in=90] (13.1, 0);
-        \node [rotate=90,blue,font=\footnotesize] at (13.5, 1.5) {network communication};
+        \node [rotate=90,networkcolor,font=\footnotesize] at (13.5, 1.5) {network communication};
         \matrix [row sep={19pt,between origins},nodes=item] at (16, 3) {
-            \node [minimum width=0.5cm,fill=red!10] {1.}; & \node [fill=red!10] {phone joe}; \\
+            \node [minimum width=0.5cm,fill=movecolor!10] {1.}; & \node [fill=movecolor!10] {phone joe}; \\
             \node [minimum width=0.5cm] {2.}; & \node {buy milk}; \\
@@
         \matrix [row sep={19pt,between origins},nodes=item] at (16, 0) {
-            \node [minimum width=0.5cm,fill=red!10] {1.}; & \node [fill=red!10] {phone joe}; \\
+            \node [minimum width=0.5cm,fill=movecolor!10] {1.}; & \node [fill=movecolor!10] {phone joe}; \\
             \node [minimum width=0.5cm] {2.}; & \node {buy milk}; \\
@@
     \end{tikzpicture}
-    \caption{When the same item is concurrently moved to different positions on different replicas, one of those positions should become the ``winning'' position as the replicas converge.
+    \caption{When the same element is concurrently moved to different positions on different replicas, one of those positions should become the ``winning'' position as the replicas converge.
 In this example, replica $A$'s move operation wins.}
@@
 For example, say a user has replicas of their to-do list on their laptop and their smartphone.
-They reorder items on their laptop to match new priorities, and also reorder some items on their phone while the phone is offline, as illustrated in Fig.~\ref{fig:duplication}.
+They reorder items on their laptop to match new priorities, and also reorder some items on their phone while the phone is offline, as illustrated in Figure~\ref{fig:duplication}.
 Later, when the phone is online again and synchronises with the laptop, any items that were moved on both devices will be duplicated.
 
-This behaviour is especially confusing if the same item was moved to the same position (e.g.\ to the top of the list) on both devices, in which case the duplicated items are likely to be adjacent in the final list, as in Fig.~\ref{fig:duplication} (depending on any other insertions that might have taken place).
-However, even if the concurrent move operations had different destination positions, duplicating the moved item at both destinations is unlikely to be desirable behaviour in most applications.
+This behaviour is especially confusing if the same item was moved to the same position (e.g.\ to the top of the list) on both devices, in which case the duplicated items are likely to be adjacent in the final list, as in Figure~\ref{fig:duplication} (depending on any other insertions that might have taken place).
+However, even if the concurrent move operations had different destination positions, duplicating the moved element at both destinations is unlikely to be desirable behaviour in most applications.
 
 Rather, we argue that the best semantics in this situation is for the system to pick one of the destination positions as the ``winner''; as the replicas communicate, they all converge towards a list in which the moved element is located only at the winning position, and nowhere else.
-This approach is illustrated in Fig.~\ref{fig:concurrent}.
+This approach is illustrated in Figure~\ref{fig:concurrent}.
 Any deterministic method of picking the winner is suitable (e.g.\ based on a priority given to each replica, or based on the timestamp of the operations).
@@
 
-Assume now that $p_1, p_2, \dots$ are position identifiers according to one of the aforementioned schemes.
+Assume now that $p_1, p_2, \dots$ are position identifiers according to one of the aforementioned schemes, and $v_1, v_2, \dots$ are the values of list elements (e.g.\ to-do list items).
 We can now model the state of the list as a set of $(\mathit{value}, \mathit{register})$ pairs contained in an add-wins set CRDT:
@@
 
-In order to iterate over the current state of the list, we need to query the above set by the position identifiers stored within the registers.
-We can do this efficiently by using a database index, which allows us to find all the tuples within a set that contain a certain value (or range of values).
+In order to obtain the elements in their list order we need to query the above set by the position identifiers stored within the registers.
+We can do this efficiently by treating the set of pairs as a table in a relational database and using a database index on the position-register column.
+The index lets us efficiently find all the list elements with a particular position ID (or range of position IDs).
 
@@
 It is also possible for multiple list elements to have the same position identifier, and thus for several list elements to appear at the same position.
-This situation is similar to having a multi-value register of values at each list position.
+This is similar to having a multi-value register of values at each list position.
 (Note this situation is different from our earlier use of a MV register, which was about one element having multiple positions, not one position having multiple elements.)
@@
 We can avoid having multiple list elements with the same position identifier by always creating a fresh position identifier as the destination of a move operation.
-All of the aforementioned list CRDTs have methods of creating new position identifiers that are guaranteed to be globally unique.
+All of the aforementioned list CRDTs have methods of creating new, globally unique position identifiers for any location within a list.
+
+This algorithm is a CRDT because its state is a straightforward composition of existing CRDTs.
+
+\begin{figure*}
+    \begin{tikzpicture}
+        \tikzstyle{state}=[matrix,anchor=west,column sep={18pt,between origins},row sep={18pt,between origins}]
+        \tikzstyle{char}=[draw,anchor=base,minimum width=18pt,text height=8pt,text depth=3pt,minimum height=18pt]
+        \tikzstyle{mov}=[fill=movecolor!10]
+        \tikzstyle{ins}=[fill=insertcolor!10]
+        \tikzstyle{del}=[fill=deletecolor!10]
+        \tikzstyle{network}=[thick,dashed,networkcolor,-{Stealth[length=3mm]}]
+        \path [draw,dotted] (0,1.5) -- (17.5,1.5);
+        \node at (0, 4) [anchor=west] {Replica $A$:};
+        \node at (0, 1) [anchor=west] {Replica $B$:};
+        \node (before1) at (0,3) [state] {
+            \node [char] {$\bullet$}; &
+            \node [char] {}; &
+            \node [char] {\texttt{B}}; &
+            \node [char] {\texttt{a}}; &
+            \node [char] {\texttt{c}}; &
+            \node [char] {\texttt{o}}; &
+            \node [char] {\texttt{n}}; &
+            \node [char] {\verb|\n|}; &
+            \node {}; &
+            \node {}; &
+            \node {}; &
+            \node {}; \\
+            \node [char] {$\bullet$}; &
+            \node [char] {}; &
+            \node [char,ins] (ins-begin) {\color{insertcolor}{\texttt{S}}}; &
+            \node [char,ins] {\color{insertcolor}{\texttt{o}}}; &
+            \node [char,ins] {\color{insertcolor}{\texttt{y}}}; &
+            \node [char,ins] {}; &
+            \node [char,ins] (ins-end) {\color{insertcolor}{\texttt{m}}}; &
+            \node [char,del] (del) {\color{deletecolor}{\renewcommand{\ULthickness}{1pt}\sout{~\texttt{M}~}}}; &
+            \node [char] {\texttt{i}}; &
+            \node [char] {\texttt{l}}; &
+            \node [char] {\texttt{k}}; &
+            \node [char] {\verb|\n|}; &
+            \node {}; \\
+        };
+        \draw [snake=brace,raise snake=3pt,gap around snake=1pt,insertcolor,thick]
+            (ins-end.south east) -- node [midway,below,inner ysep=8pt] {insert} (ins-begin.south west);
+        \draw [snake=brace,raise snake=3pt,gap around snake=1pt,deletecolor,thick]
+            (del.south east) -- node [midway,below,inner ysep=8pt] {delete} (del.south west);
+        \node (before2) at (0,0) [state] {
+            \node [char] {$\bullet$}; &
+            \node [char] {}; &
+            \node [char] {\texttt{B}}; &
+            \node [char] {\texttt{a}}; &
+            \node [char] {\texttt{c}}; &
+            \node [char] {\texttt{o}}; &
+            \node [char] {\texttt{n}}; &
+            \node [char] (dest) {\verb|\n|}; \\
+            \node [char,mov] {$\bullet$}; &
+            \node [char,mov] {}; &
+            \node [char,mov] {\texttt{M}}; &
+            \node [char,mov] {\texttt{i}}; &
+            \node [char,mov] {\texttt{l}}; &
+            \node [char,mov] {\texttt{k}}; &
+            \node [char,mov] (src) {\verb|\n|}; &
+            \node {}; \\
+        };
+        \draw [->,movecolor,thick] (src.east) .. controls +(0:2cm) and +(0:1cm) .. node [right,near end,inner xsep=7pt] {move} (dest.north east);
+        \draw [->,thick] (6.2, 0) -- (10.0, 0);
+        \draw [->,thick] (8.0, 3) -- (10.0, 3);
+        \draw [network] (8.5, 0) to [out=90,in=270] (9.1, 3);
+        \draw [network] (8.5, 3) to [out=270,in=90] (9.1, 0);
+        \node [rotate=90,networkcolor,font=\footnotesize] at (9.5, 1.5) {network communication};
+        \node (after1) at (10,3) [state] {
+            \node [char,mov] {$\bullet$}; &
+            \node [char,mov] {}; &
+            \node [char,mov] {\texttt{M}}; &
+            \node [char,mov] {\texttt{i}}; &
+            \node [char,mov] {\texttt{l}}; &
+            \node [char,mov] {\texttt{k}}; &
+            \node [char,mov] {\verb|\n|}; &
+            \node {}; \\
+            \node [char] {$\bullet$}; &
+            \node [char] {}; &
+            \node [char] {\texttt{B}}; &
+            \node [char] {\texttt{a}}; &
+            \node [char] {\texttt{c}}; &
+            \node [char] {\texttt{o}}; &
+            \node [char] {\texttt{n}}; &
+            \node [char] {\verb|\n|}; \\
+            \node [char,ins] {\color{insertcolor}{\texttt{S}}}; &
+            \node [char,ins] {\color{insertcolor}{\texttt{o}}}; &
+            \node [char,ins] {\color{insertcolor}{\texttt{y}}}; &
+            \node [char,ins] {}; &
+            \node [char,ins] {\color{insertcolor}{\texttt{m}}}; &
+            \node {}; &
+            \node {}; &
+            \node {}; \\
+        };
+        \node (after2) at (10,0) [state] {
+            \node [char,mov] {$\bullet$}; &
+            \node [char,mov] {}; &
+            \node [char,mov] {\texttt{M}}; &
+            \node [char,mov] {\texttt{i}}; &
+            \node [char,mov] {\texttt{l}}; &
+            \node [char,mov] {\texttt{k}}; &
+            \node [char,mov] {\verb|\n|}; &
+            \node {}; \\
+            \node [char] {$\bullet$}; &
+            \node [char] {}; &
+            \node [char] {\texttt{B}}; &
+            \node [char] {\texttt{a}}; &
+            \node [char] {\texttt{c}}; &
+            \node [char] {\texttt{o}}; &
+            \node [char] {\texttt{n}}; &
+            \node [char] {\verb|\n|}; \\
+            \node [char,ins] {\color{insertcolor}{\texttt{S}}}; &
+            \node [char,ins] {\color{insertcolor}{\texttt{o}}}; &
+            \node [char,ins] {\color{insertcolor}{\texttt{y}}}; &
+            \node [char,ins] {}; &
+            \node [char,ins] {\color{insertcolor}{\texttt{m}}}; &
+            \node {}; &
+            \node {}; &
+            \node {}; \\
+        };
+    \end{tikzpicture}
+    \caption{Replica $B$ moves a range of characters (here, ``$\bullet$ Milk''), while concurrently replica $A$ makes an edit inside the same range.
+If the move is performed character-by-character, $A$'s edit remains at the old position, while its surrounding context has moved to a new position, leading to an anomalous outcome.}
+    \label{fig:range-bad}
+\end{figure*}
+
+\begin{figure*}
+    \vspace{2em}
+    \begin{tikzpicture}
+        \tikzstyle{state}=[matrix,anchor=west,column sep={18pt,between origins},row sep={18pt,between origins}]
+        \tikzstyle{char}=[draw,anchor=base,minimum width=18pt,text height=8pt,text depth=3pt,minimum height=18pt]
+        \tikzstyle{mov}=[fill=movecolor!10]
+        \tikzstyle{ins}=[fill=insertcolor!10]
+        \tikzstyle{del}=[fill=deletecolor!10]
+        \tikzstyle{network}=[thick,dashed,networkcolor,-{Stealth[length=3mm]}]
+        \path [draw,dotted] (0,1.5) -- (17.5,1.5);
+        \node at (0, 4) [anchor=west] {Replica $A$:};
+        \node at (0, 1) [anchor=west] {Replica $B$:};
+        \node (before1) at (0,3) [state] {
+            \node [char] {$\bullet$}; &
+            \node [char] {}; &
+            \node [char] {\texttt{B}}; &
+            \node [char] {\texttt{a}}; &
+            \node [char] {\texttt{c}}; &
+            \node [char] {\texttt{o}}; &
+            \node [char] {\texttt{n}}; &
+            \node [char] {\verb|\n|}; &
+            \node {}; &
+            \node {}; &
+            \node {}; &
+            \node {}; \\
+            \node [char] {$\bullet$}; &
+            \node [char] {}; &
+            \node [char,ins] (ins-begin) {\color{insertcolor}{\texttt{S}}}; &
+            \node [char,ins] {\color{insertcolor}{\texttt{o}}}; &
+            \node [char,ins] {\color{insertcolor}{\texttt{y}}}; &
+            \node [char,ins] {}; &
+            \node [char,ins] (ins-end) {\color{insertcolor}{\texttt{m}}}; &
+            \node [char,del] {\color{deletecolor}{\renewcommand{\ULthickness}{1pt}\sout{~\texttt{M}~}}}; &
+            \node [char] {\texttt{i}}; &
+            \node [char] {\texttt{l}}; &
+            \node [char] {\texttt{k}}; &
+            \node [char] {\verb|\n|}; &
+            \node {}; \\
+        };
+        \draw [snake=brace,raise snake=3pt,gap around snake=1pt,insertcolor,thick]
+            (ins-end.south east) -- node [midway,below,inner ysep=8pt] {insert} (ins-begin.south west);
+        \draw [snake=brace,raise snake=3pt,gap around snake=1pt,deletecolor,thick]
+            (del.south east) -- node [midway,below,inner ysep=8pt] {delete} (del.south west);
+        \node (before2) at (0,0) [state] {
+            \node [char] {$\bullet$}; &
+            \node [char] {}; &
+            \node [char] {\texttt{B}}; &
+            \node [char] {\texttt{a}}; &
+            \node [char] {\texttt{c}}; &
+            \node [char] {\texttt{o}}; &
+            \node [char] {\texttt{n}}; &
+            \node [char] (dest) {\verb|\n|}; \\
+            \node [char,mov] {$\bullet$}; &
+            \node [char,mov] {}; &
+            \node [char,mov] {\texttt{M}}; &
+            \node [char,mov] {\texttt{i}}; &
+            \node [char,mov] {\texttt{l}}; &
+            \node [char,mov] {\texttt{k}}; &
+            \node [char,mov] (src) {\verb|\n|}; &
+            \node {}; \\
+        };
+        \draw [->,movecolor,thick] (src.east) .. controls +(0:2cm) and +(0:1cm) .. node [right,near end,inner xsep=7pt] {move} (dest.north east);
+        \draw [->,thick] (6.2, 0) -- (10.0, 0);
+        \draw [->,thick] (8.0, 3) -- (10.0, 3);
+        \draw [network] (8.5, 0) to [out=90,in=270] (9.1, 3);
+        \draw [network] (8.5, 3) to [out=270,in=90] (9.1, 0);
+        \node [rotate=90,networkcolor,font=\footnotesize] at (9.5, 1.5) {network communication};
+        \node (after1) at (10,3) [state] {
+            \node [char,mov] {$\bullet$}; &
+            \node [char,mov] {}; &
+            \node [char,ins] {\texttt{S}}; &
+            \node [char,ins] {\texttt{o}}; &
+            \node [char,ins] {\texttt{y}}; &
+            \node [char,ins] {}; &
+            \node [char,ins] {\texttt{m}}; &
+            \node [char,mov] {\texttt{i}}; &
+            \node [char,mov] {\texttt{l}}; &
+            \node [char,mov] {\texttt{k}}; &
+            \node [char,mov] {\verb|\n|}; &
+            \node {}; \\
+            \node [char] {$\bullet$}; &
+            \node [char] {}; &
+            \node [char] {\texttt{B}}; &
+            \node [char] {\texttt{a}}; &
+            \node [char] {\texttt{c}}; &
+            \node [char] {\texttt{o}}; &
+            \node [char] {\texttt{n}}; &
+            \node [char] {\verb|\n|}; &
+            \node {}; &
+            \node {}; &
+            \node {}; &
+            \node {}; \\
+        };
+        \node (after2) at (10,0) [state] {
+            \node [char,mov] {$\bullet$}; &
+            \node [char,mov] {}; &
+            \node [char,ins] {\texttt{S}}; &
+            \node [char,ins] {\texttt{o}}; &
+            \node [char,ins] {\texttt{y}}; &
+            \node [char,ins] {}; &
+            \node [char,ins] {\texttt{m}}; &
+            \node [char,mov] {\texttt{i}}; &
+            \node [char,mov] {\texttt{l}}; &
+            \node [char,mov] {\texttt{k}}; &
+            \node [char,mov] {\verb|\n|}; &
+            \node {}; \\
+            \node [char] {$\bullet$}; &
+            \node [char] {}; &
+            \node [char] {\texttt{B}}; &
+            \node [char] {\texttt{a}}; &
+            \node [char] {\texttt{c}}; &
+            \node [char] {\texttt{o}}; &
+            \node [char] {\texttt{n}}; &
+            \node [char] {\verb|\n|}; &
+            \node {}; &
+            \node {}; &
+            \node {}; &
+            \node {}; \\
+        };
+    \end{tikzpicture}
+    \caption{The desired outcome of the scenario in Figure~\ref{fig:range-bad}.}
+    \label{fig:range-good}
+\end{figure*}
+
+\section{Moving Ranges of Elements}\label{sec:ranges}
+
+The algorithm of Section~\ref{sec:algorithm} allows a single list element to be moved at a time.
+In some applications, a further challenge arises: we may want to move not just one element, but move a whole consecutive range of elements in one operation.
+For example, in a text document, a user may want to move a section of text (e.g.\ a paragraph, or the text of a bullet point) to a different position in the text.
+
+We might try to move a range of elements by moving each element individually.
+However, this approach has a problem, which is illustrated in Figure~\ref{fig:range-bad}.
+In this example, replica $B$ wants to swap the order of two bullet points in the text of a shopping list by moving the character range ``$\bullet$ Milk'' (including the bullet point character $\bullet$), while concurrently replica $A$ changes the text ``Milk'' to ``Soy milk''.
+While the move is successful, the character insertions and deletion of $A$'s edit remain attached to the old position of ``Milk'' (at the end of the document, after ``Bacon$\backslash$n'').
+This is the case because CRDTs for lists and text perform insertions and deletions at particular position IDs, which remain unchanged even as elements are moved to new position IDs.
+
+In contrast, the desired outcome of this scenario is shown in Figure~\ref{fig:range-good}: we want $A$'s edit to apply in the context in which it was originally made, even if that context has since moved to a new position in the document.
+In order to enable this semantics, it is not sufficient to move each element individually: we need to capture the fact that a certain range of elements is moved, so that any concurrent edits that fall within that range can be redirected to the new position of that range.
+
+At present, there is no known algorithm for performing such moves of ranges of elements.
+Various tricky edge-cases would need to be handled by such an algorithm, such as concurrent moves of partially overlapping ranges, and moves whose destination position falls within another range that is itself being moved.
+Part of the solution may be to give each list element a unique identity that is independent from its position in the document, so that the position of a list element can be changed without changing its identity; insertions and deletions can then reference the \emph{identities} of their context, rather than their position IDs.
+
+We will briefly discuss one more approach for moving ranges, and explain why it is not sufficient.
+In the example of Figures~\ref{fig:range-bad} and~\ref{fig:range-good}, if the application needs to support reordering of bullet points, we might try to simplify the problem by using a two-level hierarchy instead of a flat sequence of characters.
+That is, we can create a top-level list object in which each list element contains the entire content of a single bullet point, and use a separate text CRDT object for the content of each bullet point.
+In this design, reordering bullet points requires only a single-element move operation.
+
+However, this approach has a different problem: when the user presses the enter key in the middle of the text of a bullet point, the usual behaviour in most editors is to split it into two bullet points.
+Conversely, hitting the delete key while the cursor is positioned at the end of a bullet point's text usually joins it with the following bullet point's text.
+If each bullet point is represented as a separate text object, these splitting and joining operations require moving a range of characters from one text object to another.
+
+Thus, representing each bullet point as a single list element for the sake of reordering does not actually obviate the need for moving ranges of characters.
+As before, implementing move as delete-and-reinsert has the effect of duplicating text if two users concurrently split or join the same bullet point.
+Thus, it seems that support for moving ranges of characters is an important feature for collaborative document editors.
+
+% There is also a more specific open problem: fixing a typo in the middle of a paragraph while concuurrently the entire paragraph is deleted.
+
+% When moving a range, take the identity timestamp of all but the last element and move them to the new location.
+% For the last element, leave the identity timestamp in place, but move the value to the new location.
+% Insertions and deletions use the identity timestamp to locate their position.
+% Thus, if they fall within the moved range, their identity timestamp will be located at the new location.
+
+% The move operation itself also has a timestamp.
+% When performing the move, for all elements but the last one in the range, we also pick up any intervening elements.
+% (Can we say anything about the timestamps of those elements relative to the move operation's timestamp?)
+
+% When moving two different ranges to the same destination position, need to ensure they end up in a deterministic order.
+% Use positionID timestamps for this.
+
+% When moving two partially overlapping ranges... how to ensure we don’t end up matching half the document?
+
+% xdst ... lots of text ... xstart ... ystart ... xend ... yend ... ydst
+
+% If x is moved to xdst first, then y second, the range of y would appear to include “lots of text”
+
+% How to ensure the destination does not fall within the range being moved?
+
+\section{Conclusions}
+
+This short paper has made the case for list CRDTs to support an explicit move operation, allowing list elements to be reordered without duplicating them in the case of concurrent moves.
+We have outlined an algorithm for single-element moves that can be retrofitted to any existing list CRDT that uses stable position IDs to refer to locations within the list (which is the case for all currently known list CRDTs).
+We have also demonstrated the need for moving a contiguous range of list elements to a new position; finding an algorithm that adds this feature to a CRDTs is at present an open problem.
+We hope that future research will succeed in developing an algorithm for moving ranges of elements.
 
@@
 
+\balance
 \bibliographystyle{ACM-Reference-Format}

### 2020-02-20 paper/list-move.tex
Move figure, add related work

@@
     <concept>
+        <concept_id>10011007.10010940.10010992.10010993.10010996</concept_id>
+        <concept_desc>Software and its engineering~Consistency</concept_desc>
+        <concept_significance>500</concept_significance>
+    </concept>
+    <concept>
         <concept_id>10003120.10003130.10003131.10003570</concept_id>
@@
 
+\begin{CCSXML}
+<ccs2012>
+ </ccs2012>
+\end{CCSXML}
+
 \ccsdesc[500]{Theory of computation~Distributed algorithms}
+\ccsdesc[500]{Software and its engineering~Consistency}
 \ccsdesc[300]{Human-centered computing~Computer supported cooperative work}
@@
 
-\section{A Generic Reordering Algorithm}\label{sec:algorithm}
-
-Intuitively, if we want to pick one winner from among several concurrent moves for the same element, we can use a long-established CRDT: a last-writer wins (LWW) register~\cite{Shapiro:2011wy,Johnson:1975we}.
-We need one such register for each list element, containing the position of that element.
-Moving an element to a new location is then merely a matter of updating the value of that element's register.
-
-In general, we could also use a multi-value (MV) register~\cite{Shapiro:2011wy}, which would allow an element to exist at multiple positions in the list simultaneously after having experienced conflicting moves.
-This is not the same as the duplication described in Section~\ref{sec:semantics}, since the multiple positions still belong to the same element, and a subsequent move would collapse them back into a single position again.
-The MV register option can be used if the LWW behaviour (discarding all but one of the concurrent moves of the same element) is unacceptable.
-
-Regardless of whether a LWW or MV register is used, this approach requires a stable way of referencing some position in the list (i.e.\ the value held by the register).
-An integer index is not suitable, because an insertion or deletion at any position would require changing all of the indices following that position.
-Fortunately, all known list CRDTs already incorporate schemes for stable position identifiers.
-The details vary: Treedoc uses a path through a binary tree~\cite{Preguica:2009fz}, Logoot uses a list of (integer, replicaID) pairs~\cite{Weiss:2009ht}, RGA uses a so-called s4vector~\cite{Roh:2011dw}, Causal Trees/Timestamped Insertion Trees use a timestamp~\cite{Grishchenko:2014eh,Attiya:2016kh}, and so on.
-But all of these approaches have in common that they can uniquely refer to a particular position in a list, in a way that is not affected by operations on other list positions.
-
-Assume now that $p_1, p_2, \dots$ are position identifiers according to one of the aforementioned schemes, and $v_1, v_2, \dots$ are the values of list elements (e.g.\ to-do list items).
-We can now model the state of the list as a set of $(\mathit{value}, \mathit{register})$ pairs contained in an add-wins set CRDT:
-\begin{align*}
-\mathit{state} = \mathrm{AWSet}(\{ & (v_1, \mathrm{LWWRegister}(p_1)),\\
-    & (v_2, \mathrm{LWWRegister}(p_2)), \;\dots\})
-\end{align*}
-In some schemes (such as Treedoc, Logoot and LSEQ), the position identifiers themselves encode their relative ordering~\cite{Interleaving}, and thus no further state is required.
-In others (such as RGA and CT), the list CRDT requires an additional data structure that determines the order in which position identifiers appear in the list.
-
-In order to obtain the elements in their list order we need to query the above set by the position identifiers stored within the registers.
-We can do this efficiently by treating the set of pairs as a table in a relational database and using a database index on the position-register column.
-The index lets us efficiently find all the list elements with a particular position ID (or range of position IDs).
-
-Note that when the position of a list element is changed, the old position identifier no longer has any value associated with it, and thus it should be regarded as nonexistent (like a tombstone in some CRDTs).
-It is also possible for multiple list elements to have the same position identifier, and thus for several list elements to appear at the same position.
-This is similar to having a multi-value register of values at each list position.
-(Note this situation is different from our earlier use of a MV register, which was about one element having multiple positions, not one position having multiple elements.)
-
-We can avoid having multiple list elements with the same position identifier by always creating a fresh position identifier as the destination of a move operation.
-All of the aforementioned list CRDTs have methods of creating new, globally unique position identifiers for any location within a list.
-
-This algorithm is a CRDT because its state is a straightforward composition of existing CRDTs.
-
 \begin{figure*}
@@
 
+\section{A Generic Reordering Algorithm}\label{sec:algorithm}
+
+Intuitively, if we want to pick one winner from among several concurrent moves for the same element, we can use a long-established CRDT: a last-writer wins (LWW) register~\cite{Shapiro:2011wy,Johnson:1975we}.
+We need one such register for each list element, containing the position of that element.
+Moving an element to a new location is then merely a matter of updating the value of that element's register.
+
+In general, we could also use a multi-value (MV) register~\cite{Shapiro:2011wy}, which would allow an element to exist at multiple positions in the list simultaneously after having experienced conflicting moves.
+This is not the same as the duplication described in Section~\ref{sec:semantics}, since the multiple positions still belong to the same element, and a subsequent move would collapse them back into a single position again.
+The MV register option can be used if the LWW behaviour (discarding all but one of the concurrent moves of the same element) is unacceptable.
+
+Regardless of whether a LWW or MV register is used, this approach requires a stable way of referencing some position in the list (i.e.\ the value held by the register).
+An integer index is not suitable, because an insertion or deletion at any position would require changing all of the indices following that position.
+Fortunately, all known list CRDTs already incorporate schemes for stable position identifiers.
+The details vary: Treedoc uses a path through a binary tree~\cite{Preguica:2009fz}, Logoot uses a list of (integer, replicaID) pairs~\cite{Weiss:2009ht}, RGA uses a so-called s4vector~\cite{Roh:2011dw}, Causal Trees/Timestamped Insertion Trees use a timestamp~\cite{Grishchenko:2014eh,Attiya:2016kh}, and so on.
+But all of these approaches have in common that they can uniquely refer to a particular position in a list, in a way that is not affected by operations on other list positions.
+
+Assume now that $p_1, p_2, \dots$ are position identifiers according to one of the aforementioned schemes, and $v_1, v_2, \dots$ are the values of list elements (e.g.\ to-do list items).
+We can now model the state of the list as a set of $(\mathit{value}, \mathit{register})$ pairs contained in an add-wins set CRDT:
+\begin{align*}
+\mathit{state} = \mathrm{AWSet}(\{ & (v_1, \mathrm{LWWRegister}(p_1)),\\
+    & (v_2, \mathrm{LWWRegister}(p_2)), \;\dots\})
+\end{align*}
+In some schemes (such as Treedoc, Logoot and LSEQ), the position identifiers themselves encode their relative ordering~\cite{Interleaving}, and thus no further state is required.
+In others (such as RGA and CT), the list CRDT requires an additional data structure that determines the order in which position identifiers appear in the list.
+
+In order to obtain the elements in their list order we need to query the above set by the position identifiers stored within the registers.
+We can do this efficiently by treating the set of pairs as a table in a relational database and using a database index on the position-register column.
+The index lets us efficiently find all the list elements with a particular position ID (or range of position IDs).
+
+Note that when the position of a list element is changed, the old position identifier no longer has any value associated with it, and thus it should be regarded as nonexistent (like a tombstone in some CRDTs).
+It is also possible for multiple list elements to have the same position identifier, and thus for several list elements to appear at the same position.
+This is similar to having a multi-value register of values at each list position.
+(Note this situation is different from our earlier use of a MV register, which was about one element having multiple positions, not one position having multiple elements.)
+
+We can avoid having multiple list elements with the same position identifier by always creating a fresh position identifier as the destination of a move operation.
+All of the aforementioned list CRDTs have methods of creating new, globally unique position identifiers for any location within a list.
+
+This algorithm is a CRDT because its state is a straightforward composition of existing CRDTs.
+
 \begin{figure*}
-    \vspace{2em}
     \begin{tikzpicture}
@@
 The algorithm of Section~\ref{sec:algorithm} allows a single list element to be moved at a time.
-In some applications, a further challenge arises: we may want to move not just one element, but move a whole consecutive range of elements in one operation.
-For example, in a text document, a user may want to move a section of text (e.g.\ a paragraph, or the text of a bullet point) to a different position in the text.
+In some applications, a further challenge arises: we may want to move not just one element, but move an entire consecutive range of elements in one operation.
+For example, in a text document, a user may want to move a section of text (e.g.\ a paragraph, or the text of a bullet point) to a different position in the text~\cite{Lord:2019bo}.
 
@@
 In this example, replica $B$ wants to swap the order of two bullet points in the text of a shopping list by moving the character range ``$\bullet$ Milk'' (including the bullet point character $\bullet$), while concurrently replica $A$ changes the text ``Milk'' to ``Soy milk''.
-While the move is successful, the character insertions and deletion of $A$'s edit remain attached to the old position of ``Milk'' (at the end of the document, after ``Bacon$\backslash$n'').
+While the move is successful, the character insertions and deletion of $A$'s edit remain attached to the old position of ``Milk'' (near the end of the document, after ``Bacon$\backslash$n'').
 This is the case because CRDTs for lists and text perform insertions and deletions at particular position IDs, which remain unchanged even as elements are moved to new position IDs.
@@
 
+% bounds of range are given as positions (so that moving a range containing a bound does not move
+% the bound somewhere else), but insertions/deletions reference element identity
+
+% Inserted item may be relocated if its insertion position is within a range that was moved. This
+% implies the insertion position cannot be given as a position ID in the operation, since a new
+% position ID would have to be generated in the effector. However, if insertion contains only
+% identity timestamp, as in RGA, then it’s fine?
+
+% But then if a range’s bounds are specified as position IDs, and an insertion contains only an
+% identity timestamp, where do the position IDs come from?
+
+% Maybe an insert op should contain two timestamps — one for identity, one for position?
+
+\section{Related work}
+
+Our previous unpublished work \cite{ExtendedVersion} discusses a single-element move operation for lists in the context of a larger specification framework for CRDTs.
+However, this work does not provide an efficient implementation of this operation.
+
+Apart from this work, to our knowledge Ahmed-Nacer et al.~\cite{AhmedNacer:2013eq} provide the only published list CRDT that includes a single-element move operation.
+However, it is designed to duplicate an element when moved concurrently~-- semantics we consider undesirable.
+Moreover, we believe the algorithm to be more complicated than necessary.
+
+All other list CRDTs~\cite{Oster:2006wj,Preguica:2009fz,Roh:2011dw,Grishchenko:2014eh,Attiya:2016kh,Weiss:2009ht,Weiss:2010hx,Nedelec:2013ky,Nedelec:2016eo} and Operational Transformation algorithms \cite{Ellis:1989ue,Nichols:1995fd,Ressel:1996wx,Sun:1998vf,Oster:2006tr} provide only insertion and deletion.
+
+Lord~\cite{Lord:2019bo} discusses the problem of moving ranges of elements, but does not present a solution.
+
 \section{Conclusions}

### 2020-02-24 paper/list-move.tex
More consistently use "move" instead of "reorder"

@@
 \begin{document}
-\title{Reordering Elements in List CRDTs}
+\title{Moving Elements in List CRDTs}
 
@@
 Conflict-free Replicated Data Types (CRDTs) for lists allow multiple users to concurrently insert and delete elements in a shared list object.
-However, existing algorithms behave poorly when users concurrently reorder elements in the list (i.e.\ move list elements to a new position).
+However, existing algorithms behave poorly when users concurrently move list elements to a new position (i.e.\ reorder the list elements).
 We demonstrate the need for such a move operation, and describe an algorithm that extends a list CRDT with an explicit move operation.
@@
 All of the aforementioned algorithms allow replicas to \emph{insert} or \emph{delete} elements anywhere in the list.
-However, none of them have explicit support for \emph{reordering} or \emph{moving} elements from one position to another position in the list.
-This is a surprising omission because reordering is a commonly required operation: in many to-do list applications, a user can drag and drop list elements to reorder them, and graphics software allows users to reorder the object list with commands such as ``bring to front'' (which moves an object to the top of the list, so that it occludes other objects) and ``send to back'' (which moves an object to the bottom, so that it is occluded by other objects).
+However, none of them have explicit support for \emph{moving} elements from one position to another position in the list (\emph{reordering}).
+This is a surprising omission because moving is a commonly required operation: in many to-do list applications, a user can drag and drop list elements to reorder them, and graphics software allows users to reorder the object list with commands such as ``bring to front'' (which moves an object to the top of the list, so that it occludes other objects) and ``send to back'' (which moves an object to the bottom, so that it is occluded by other objects).
 
-In this paper we introduce an algorithm that allows an existing list CRDT to be extended with a reorder operation.
+In this paper we introduce an algorithm that allows an existing list CRDT to be extended with a move operation.
 The algorithm is generic in the sense that it can be implemented on top of any of the aforementioned list CRDTs.
@@
     \end{tikzpicture}
-    \caption{If reordering is implemented by deleting and re-inserting, concurrent moves of the same element result in the anomaly shown here: the moved element is duplicated.}
+    \caption{If moving is implemented by deleting and re-inserting, concurrent moves of the same element result in the anomaly shown here: the moved element is duplicated.}
     \label{fig:duplication}
@@
 For example, say a user has replicas of their to-do list on their laptop and their smartphone.
-They reorder items on their laptop to match new priorities, and also reorder some items on their phone while the phone is offline, as illustrated in Figure~\ref{fig:duplication}.
+They move items on their laptop to match new priorities, and also move some items on their phone while the phone is offline, as illustrated in Figure~\ref{fig:duplication}.
 Later, when the phone is online again and synchronises with the laptop, any items that were moved on both devices will be duplicated.
@@
 
-\section{A Generic Reordering Algorithm}\label{sec:algorithm}
+\section{A Generic Moving Algorithm}\label{sec:algorithm}
 
@@
 That is, we can create a top-level list object in which each list element contains the entire content of a single bullet point, and use a separate text CRDT object for the content of each bullet point.
-In this design, reordering bullet points requires only a single-element move operation.
+In this design, moving bullet points requires only a single-element move operation.
 
@@
 
-Thus, representing each bullet point as a single list element for the sake of reordering does not actually obviate the need for moving ranges of characters.
+Thus, representing each bullet point as a single list element for the sake of moving does not actually obviate the need for moving ranges of characters.
 As before, implementing move as delete-and-reinsert has the effect of duplicating text if two users concurrently split or join the same bullet point.

### 2020-02-26 evaluation/src/main/scala/TestReplica.scala
Add a leader-based execution model to evaluation

@@ -18,6 +18,16 @@ object TestReplica {
   // Backpressure kicks in if more than this number of requests in flight
   val MAX_PENDING_REQUESTS = 50
 
+  // If false, runs in CRDT mode. If true, uses leader-based replication.
+  val USE_LEADER = false
+
+  def startDaemon[T <: Runnable](runnable: T): T = {
+    val thread = new Thread(runnable)
+    thread.setDaemon(true)
+    thread.start()
+    return runnable
+  }
+
   def main(args: Array[String]): Unit = {
     if (args.length < 1) {
       throw new Exception("Usage: TestReplica replica-id remote-ip1 [remote-ip2 ...]")
@@ -26,22 +36,24 @@ object TestReplica {
     implicit val metrics = new MetricRegistry()
     ConsoleReporter.forRegistry(metrics).build().start(20, TimeUnit.SECONDS)
 
-    val replica = new ReplicaThread(args(0).toLong, metrics)
-    val replicaThread = new Thread(replica)
-    replicaThread.start()
+    val replicaId = args(0).toLong
+    val replica = new ReplicaThread(replicaId, metrics)
+    new Thread(replica).start()
 
-    val server = new Thread(new AcceptThread(replica))
-    server.setDaemon(true)
-    server.start()
+    startDaemon(new AcceptThread(replica))
 
     TimeUnit.SECONDS.sleep(10) // time for other servers to come up
 
-    for (remoteIp <- args.drop(1)) {
-      val client = new ClientThread(remoteIp, metrics)
-      val clientThread = new Thread(client)
-      clientThread.setDaemon(true)
-      clientThread.start()
-      replica.addClient(client)
+    if (USE_LEADER) {
+      // Leader mode: connect to only one IP, namely the leader
+      if (replicaId > 0) {
+        replica.addClient(startDaemon(new FollowerThread(args(1), replica, metrics)))
+      }
+    } else {
+      // CRDT mode: connect to all of the remote IPs
+      for (remoteIp <- args.drop(1)) {
+        replica.addClient(startDaemon(new CRDTClientThread(remoteIp, metrics)))
+      }
     }
   }
 }
@@ -123,10 +135,15 @@ abstract class Connection(socket: Socket, recvFrameSize: Int) extends Runnable {
   }
 }
 
+trait ClientThread {
+  def send(move: Protocol.Move)
+  def backpressure: Boolean
+}
+
 // Thread that handles the client side of a connection. It sends Move requests
 // to the server, and waits for Ack responses in reply.
[…]

### 2020-02-26 evaluation/src/main/scala/TestReplica.scala
Fix oversight

@@ -184,7 +184,7 @@ class FollowerThread(val leaderIp: String, replica: ReplicaThread, metrics: Metr
     val move = Protocol.decodeMove(bytes)
     val requestId = Protocol.Ack(move.time, move.replica)
     requests.remove(requestId).stop()
-    replica.request(move, this)
+    replica.request(move, null)
   }
 
   // Returns false if we're happy to accept more requests, and true if we need
@@ -260,7 +260,7 @@ class ReplicaThread(replicaId: Long, metrics: MetricRegistry) extends Runnable {
     try {
       applyMove(move)
       if (TestReplica.USE_LEADER) {
-        sender.send(move)
+        if (sender != null) sender.send(move)
       } else {
         sender.send(Protocol.Ack(move.time, move.replica))
       }

### 2020-02-26 evaluation/run-all.sh
Document evaluation procedure

@@ -1,12 +1,16 @@
 #!/bin/bash
 set -e
 
+# Run this script with one argument: the interval between generated move
+# operations, in milliseconds, on each replica.
+INTERVAL="${1?Please specify interval between operations (in milliseconds)}"
+
 # To run the experiments, log in to the AWS EC2 console and start up a c5.large
 # instance running Ubuntu 18.04 in each of the three regions us-west-1, eu-west-1,
 # and ap-southeast-1. Then fill in their IP addresses here:
-US_WEST_1=18.144.1.184
-EU_WEST_1=34.255.190.108
-AP_SOUTHEAST_1=18.140.64.126
+US_WEST_1=13.52.240.53
+EU_WEST_1=52.48.156.251
+AP_SOUTHEAST_1=54.169.188.185
 
 # Set up the security groups such that you can log in to the VMs by SSH (TCP port 22),
 # and that they can all connect to each other on TCP port 8080.
@@ -18,21 +22,21 @@ AP_SOUTHEAST_1=18.140.64.126
 # sudo apt-get update && sudo apt-get install -y openjdk-8-jdk-headless sbt
 # git clone https://github.com/trvedata/move-op.git && cd move-op/evaluation && sbt compile
 
-run_batch () {
-    INTERVAL=$1
-
-    ssh -i ~/.ec2/martin-aws-us-west-1.pem ubuntu@$US_WEST_1 /home/ubuntu/move-op/evaluation/run-test.sh $INTERVAL 1 $EU_WEST_1 $AP_SOUTHEAST_1 &
-    ssh -i ~/.ec2/martin-aws-eu-west-1.pem ubuntu@$EU_WEST_1 /home/ubuntu/move-op/evaluation/run-test.sh $INTERVAL 2 $US_WEST_1 $AP_SOUTHEAST_1 &
-    ssh -i ~/.ec2/martin-aws-ap-southeast-1.pem ubuntu@$AP_SOUTHEAST_1 /home/ubuntu/move-op/evaluation/run-test.sh $INTERVAL 3 $US_WEST_1 $EU_WEST_1 &
-    wait
-
-    scp -i ~/.ec2/martin-aws-us-west-1.pem ubuntu@$US_WEST_1:/home/ubuntu/move-op/evaluation/data/logs/interval_$INTERVAL.log data/logs/interval_${INTERVAL}_us_west_1.log
[…]

### 2020-02-26 evaluation/data/processing_times.sh
milliseconds -> microseconds

@@ -6,7 +6,7 @@ echo -n "local min,local median,local p95,remote min,remote median,remote p95," 
 echo -n "local min,local median,local p95,remote min,remote median,remote p95," >> processing_times.csv
 echo    "local min,local median,local p95,remote min,remote median,remote p95"  >> processing_times.csv
 
-for interval in 100 50 20 14 10 7 6 5 2; do
+for interval in 100000 50000 20000 14000 10000 7000 6000 5000 2000; do
     us_times="$(cat logs/interval_${interval}_us_west_1.log.gz      | gunzip | awk -f processing_times.awk | tail -n 2 | head -n 1 | tr '\n' ',')"
     eu_times="$(cat logs/interval_${interval}_eu_west_1.log.gz      | gunzip | awk -f processing_times.awk | tail -n 2 | head -n 1 | tr '\n' ',')"
     ap_times="$(cat logs/interval_${interval}_ap_southeast_1.log.gz | gunzip | awk -f processing_times.awk | tail -n 2 | head -n 1 | tr '\n' ',')"

### 2020-02-26 evaluation/run-all.sh

@@ -2,8 +2,8 @@
 set -e
 
 # Run this script with one argument: the interval between generated move
-# operations, in milliseconds, on each replica.
-INTERVAL="${1?Please specify interval between operations (in milliseconds)}"
+# operations, in microseconds, on each replica.
+INTERVAL="${1?Please specify interval between operations (in microseconds)}"
 
 # To run the experiments, log in to the AWS EC2 console and start up a c5.large
 # instance running Ubuntu 18.04 in each of the three regions us-west-1, eu-west-1,

### 2020-02-26 evaluation/src/main/scala/TestReplica.scala

@@ -10,7 +10,7 @@ object TestReplica {
   val PORT = 8080
 
   // Time interval between generated operations (= 1 / operation rate)
-  val OPERATION_INTERVAL = TimeUnit.MILLISECONDS.toNanos(1000)
+  val OPERATION_INTERVAL = TimeUnit.MICROSECONDS.toNanos(1000000)
 
   // How long to run the test before shutting down
   val RUN_DURATION = TimeUnit.MINUTES.toNanos(10)

### 2020-02-26 evaluation/run-all.sh
Allow test run to use leader-based mode

@@ -5,6 +5,9 @@ set -e
 # operations, in microseconds, on each replica.
 INTERVAL="${1?Please specify interval between operations (in microseconds)}"
 
+# If false, runs in CRDT mode. If true, runs in leader-based mode.
+USE_LEADER=false
+
 # To run the experiments, log in to the AWS EC2 console and start up a c5.large
 # instance running Ubuntu 18.04 in each of the three regions us-west-1, eu-west-1,
 # and ap-southeast-1. Then fill in their IP addresses here:
@@ -27,16 +30,25 @@ AP_SOUTHEAST_1=54.169.188.185
 # into evaluation/data/logs/*.log.gz. Those logs are then analysed by the
 # script evaluation/data/processing_times.sh.
 
+if [ "$USE_LEADER" = "true"]; then
+    LOGDIR="data/logs-leader"
+    # Define us-west-1 to be the leader
+    ssh -i ~/.ec2/martin-aws-us-west-1.pem ubuntu@$US_WEST_1 /home/ubuntu/move-op/evaluation/run-test.sh $INTERVAL $USE_LEADER 0 &
+    ssh -i ~/.ec2/martin-aws-eu-west-1.pem ubuntu@$EU_WEST_1 /home/ubuntu/move-op/evaluation/run-test.sh $INTERVAL $USE_LEADER 1 $US_WEST_1 &
+    ssh -i ~/.ec2/martin-aws-ap-southeast-1.pem ubuntu@$AP_SOUTHEAST_1 /home/ubuntu/move-op/evaluation/run-test.sh $INTERVAL $USE_LEADER 2 $US_WEST_1 &
+else
+    LOGDIR="data/logs"
+    ssh -i ~/.ec2/martin-aws-us-west-1.pem ubuntu@$US_WEST_1 /home/ubuntu/move-op/evaluation/run-test.sh $INTERVAL $USE_LEADER 1 $EU_WEST_1 $AP_SOUTHEAST_1 &
+    ssh -i ~/.ec2/martin-aws-eu-west-1.pem ubuntu@$EU_WEST_1 /home/ubuntu/move-op/evaluation/run-test.sh $INTERVAL $USE_LEADER 2 $US_WEST_1 $AP_SOUTHEAST_1 &
[…]

### 2020-02-26 evaluation/run-test.sh

@@ -5,16 +5,26 @@ set -e
 # operations with a given inter-operation interval.
 
 INTERVAL="${1?Please specify interval}"
-REPLICA_ID="${2?Please specify replica ID}"
-REMOTE1="${3?Please specify remote IP 1}"
-REMOTE2="${4?Please specify remote IP 2}"
-LOGFILE="data/logs/interval_$INTERVAL.log"
+USE_LEADER="${2?Please specify whether to use leader}"
+REPLICA_ID="${3?Please specify replica ID}"
+REMOTE1="$4"
+REMOTE2="$5"
+
+if [ "$USE_LEADER" = "true"]; then
+    LOGDIR="data/logs-leader"
+else
+    LOGDIR="data/logs"
+fi
+
+LOGFILE="$LOGDIR/interval_$INTERVAL.log"
 
 cd "$(dirname "$0")"
 
-sed -i "s/\\(val OPERATION_INTERVAL =.*\\)([0-9]*)/\\1($INTERVAL)/" src/main/scala/TestReplica.scala
+sed -i~ -e "s/\\(val OPERATION_INTERVAL =.*\\)([0-9]*)/\\1($INTERVAL)/" \
+    -e "s/\\(val USE_LEADER = \\).*/\\1$USE_LEADER/" \
+    src/main/scala/TestReplica.scala
 
-mkdir -p data/logs
+mkdir -p "$LOGDIR"
 
 if [ -f "$LOGFILE" ]; then
     echo "log file already exists"

### 2020-02-26 evaluation/run-all.sh
Fix typos

@@ -30,7 +30,7 @@ AP_SOUTHEAST_1=54.169.188.185
 # into evaluation/data/logs/*.log.gz. Those logs are then analysed by the
 # script evaluation/data/processing_times.sh.
 
-if [ "$USE_LEADER" = "true"]; then
+if [ "$USE_LEADER" = "true" ]; then
     LOGDIR="data/logs-leader"
     # Define us-west-1 to be the leader
     ssh -i ~/.ec2/martin-aws-us-west-1.pem ubuntu@$US_WEST_1 /home/ubuntu/move-op/evaluation/run-test.sh $INTERVAL $USE_LEADER 0 &

### 2020-02-26 evaluation/run-test.sh

@@ -10,7 +10,7 @@ REPLICA_ID="${3?Please specify replica ID}"
 REMOTE1="$4"
 REMOTE2="$5"
 
-if [ "$USE_LEADER" = "true"]; then
+if [ "$USE_LEADER" = "true" ]; then
     LOGDIR="data/logs-leader"
 else
     LOGDIR="data/logs"

### 2020-02-26 evaluation/run-all.sh
Fix leader-based experiment

@@ -36,6 +36,7 @@ if [ "$USE_LEADER" = "true" ]; then
     ssh -i ~/.ec2/martin-aws-us-west-1.pem ubuntu@$US_WEST_1 /home/ubuntu/move-op/evaluation/run-test.sh $INTERVAL $USE_LEADER 0 &
     ssh -i ~/.ec2/martin-aws-eu-west-1.pem ubuntu@$EU_WEST_1 /home/ubuntu/move-op/evaluation/run-test.sh $INTERVAL $USE_LEADER 1 $US_WEST_1 &
     ssh -i ~/.ec2/martin-aws-ap-southeast-1.pem ubuntu@$AP_SOUTHEAST_1 /home/ubuntu/move-op/evaluation/run-test.sh $INTERVAL $USE_LEADER 2 $US_WEST_1 &
+    wait
 else
     LOGDIR="data/logs"
     ssh -i ~/.ec2/martin-aws-us-west-1.pem ubuntu@$US_WEST_1 /home/ubuntu/move-op/evaluation/run-test.sh $INTERVAL $USE_LEADER 1 $EU_WEST_1 $AP_SOUTHEAST_1 &

### 2020-02-26 evaluation/src/main/scala/TestReplica.scala

@@ -19,7 +19,7 @@ object TestReplica {
   val MAX_PENDING_REQUESTS = 50
 
   // If false, runs in CRDT mode. If true, uses leader-based replication.
-  val USE_LEADER = false
+  val USE_LEADER = true
 
   def startDaemon[T <: Runnable](runnable: T): T = {
     val thread = new Thread(runnable)
@@ -183,7 +183,8 @@ class FollowerThread(val leaderIp: String, replica: ReplicaThread, metrics: Metr
   def receive(bytes: Array[Byte]) {
     val move = Protocol.decodeMove(bytes)
     val requestId = Protocol.Ack(move.time, move.replica)
-    requests.remove(requestId).stop()
+    val timer = requests.remove(requestId)
+    if (timer != null) timer.stop()
     replica.request(move, null)
   }
 
@@ -199,6 +200,8 @@ class FollowerThread(val leaderIp: String, replica: ReplicaThread, metrics: Metr
 // sends either an Ack response (in CRDT mode) or a Move response (in leader mode)
 // back to the client.
 class ServerThread(replica: ReplicaThread, socket: Socket) extends Connection(socket, 4 * 8) {
+  replica.addServer(this)
+
   def send(ack: Protocol.Ack) {
     this.send(Protocol.encodeAck(ack))
   }
@@ -235,6 +238,7 @@ class ReplicaThread(replicaId: Long, metrics: MetricRegistry) extends Runnable {
   val random = new Random()
 
   var clients: List[ClientThread] = Nil
+  var servers: List[ServerThread] = Nil
 
   // For Lamport timestamps
   var counter: Long = 0
@@ -248,6 +252,10 @@ class ReplicaThread(replicaId: Long, metrics: MetricRegistry) extends Runnable {
     clients = client :: clients
   }
 
+  def addServer(server: ServerThread) {
+    servers = server :: servers
+  }
+
   // Incoming request from a ServerThread. The calling object is passed in so
   // that we know where to send the response once we've processed the operation.
   def request(move: Protocol.Move, sender: ServerThread) {
@@ -258,9 +266,10 @@ class ReplicaThread(replicaId: Long, metrics: MetricRegistry) extends Runnable {
   private[this] def processRequest(move: Protocol.Move, sender: ServerThread) {
     val timer = remoteTimer.time()
     try {
+      //println(s"Received: ${move}")
       applyMove(move)
       if (TestReplica.USE_LEADER) {
-        if (sender != null) sender.send(move)
[…]

### 2020-02-27 evaluation/run-test.sh
Avoid running out of memory

@@ -31,4 +31,4 @@ if [ -f "$LOGFILE" ]; then
     exit 1
 fi
 
-sbt "runMain TestReplica $REPLICA_ID $REMOTE1 $REMOTE2" | tee "$LOGFILE"
+sbt --mem 3072 "runMain TestReplica $REPLICA_ID $REMOTE1 $REMOTE2" 2>&1 | tee "$LOGFILE"

### 2020-02-27 evaluation/src/main/scala/TestReplica.scala

@@ -12,14 +12,14 @@ object TestReplica {
   // Time interval between generated operations (= 1 / operation rate)
   val OPERATION_INTERVAL = TimeUnit.MICROSECONDS.toNanos(1000000)
 
+  // If false, runs in CRDT mode. If true, uses leader-based replication.
+  val USE_LEADER = true
+
   // How long to run the test before shutting down
   val RUN_DURATION = TimeUnit.MINUTES.toNanos(10)
 
   // Backpressure kicks in if more than this number of requests in flight
-  val MAX_PENDING_REQUESTS = 50
-
-  // If false, runs in CRDT mode. If true, uses leader-based replication.
-  val USE_LEADER = true
+  val MAX_PENDING_REQUESTS = if (USE_LEADER) 5000 else 50
 
   def startDaemon[T <: Runnable](runnable: T): T = {
     val thread = new Thread(runnable)
@@ -309,6 +309,11 @@ class ReplicaThread(replicaId: Long, metrics: MetricRegistry) extends Runnable {
     }
     val operation = generated.Move(timestamp, BigInt(move.parent), "", BigInt(move.child))
     state = generated.integer_apply_op(operation)(state)
+
+    // Truncate the log from time to time
+    if (counter % 100000 == 0) {
+      state = (state._1.take(1000), state._2)
+    }
   }
 
   // The run loop does two things: it blocks waiting for incoming requests from

### 2020-02-28 evaluation/src/main/scala/TestReplica.scala
Add a metric to track in-flight requests

@@ -1,4 +1,4 @@
-import com.codahale.metrics.{ ConsoleReporter, MetricRegistry, Timer }
+import com.codahale.metrics.{ ConsoleReporter, MetricRegistry, Timer, Gauge }
 import java.io.{ ByteArrayInputStream, ByteArrayOutputStream, DataInputStream, DataOutputStream, InputStream, OutputStream }
 import java.net.{ ServerSocket, Socket }
 import java.util.concurrent.{ArrayBlockingQueue, ConcurrentHashMap, TimeUnit}
@@ -173,6 +173,10 @@ class FollowerThread(val leaderIp: String, replica: ReplicaThread, metrics: Metr
 
   val timer = metrics.timer(s"ClientThread($leaderIp).requests")
   val requests = new ConcurrentHashMap[Protocol.Ack, Timer.Context]()
+
+  metrics.register("RequestsMapSize", new Gauge[Int] {
+    def getValue: Int = requests.size
+  })
 
   def send(move: Protocol.Move) {
     val requestId = Protocol.Ack(move.time, move.replica)

### 2020-02-28 evaluation/src/main/scala/TestReplica.scala
Fix operation ID generation in leader mode

@@ -288,7 +288,8 @@ class ReplicaThread(replicaId: Long, metrics: MetricRegistry) extends Runnable {
   // otherwise generates a move operation and sends it to the leader.
   private[this] def generateMove() {
     if (TestReplica.USE_LEADER && replicaId == 0) return
-    val move = Protocol.Move(counter + 1, replicaId, random.nextInt(1000), random.nextInt(1000))
+    counter += 1
+    val move = Protocol.Move(counter, replicaId, random.nextInt(1000), random.nextInt(1000))
     //println(s"Generated: ${move}")
     if (!TestReplica.USE_LEADER) {
       val timer = localTimer.time()

### 2020-02-28 paper/move-op.tex
Reformat move-op paper

@@
-\documentclass[sigplan,anonymous]{acmart}
-
-% EuroSys recommended formatting
-\renewcommand\footnotetextcopyrightpermission[1]{}
-\pagestyle{plain}
-
+\documentclass{vldb}
+
+\usepackage[hidelinks]{hyperref}
+\usepackage{textcomp} % \textmu
+\usepackage{booktabs} % \toprule etc.
+\usepackage{setspace} % \setstretch
+\usepackage{balance}
 \usepackage{tikz}
 \usetikzlibrary{arrows.meta}
+
+\urlstyle{rm}
 
@@
 
-\setcopyright{rightsretained}
-\copyrightyear{2019}
-\acmYear{2019}
-\acmDOI{}
-\acmConference[]{}{}{}
-\acmBooktitle{}
-\acmPrice{}
-\acmISBN{}
+\vldbTitle{A highly-available move operation for replicated trees}
+\vldbAuthors{Martin Kleppmann, Dominic P. Mulligan, Victor B. F. Gomes, and Alastair R. Beresford}
+\vldbDOI{https://doi.org/10.14778/xxxxxxx.xxxxxxx}
+\vldbVolume{13}
+\vldbNumber{xxx}
+\vldbYear{2020}
 
@@
 
-\author{Martin Kleppmann}
+\numberofauthors{4}
+\author{
+\alignauthor
+    Martin Kleppmann\\
+    \affaddr{University of Cambridge}\\
+    \affaddr{Cambridge, UK}\\
     \email{mk428@cl.cam.ac.uk}
-\orcid{0000-0001-7252-6958}
-\affiliation{%
-  \institution{Department of Computer Science and Technology}
-  \streetaddress{15 JJ Thomson Avenue}
-  \city{Cambridge}
-  \state{}
-  \postcode{CB3 0FD}
-  \country{United Kingdom}
+\alignauthor
+    Dominic P.\ Mulligan\\
+    \affaddr{Arm Research}\\
+    \affaddr{Cambridge, UK}\\
+    \email{Dominic.Mulligan@arm.com}
+\alignauthor
+    Victor B.\ F.\ Gomes\\
+    \affaddr{University of Cambridge}\\
+    \affaddr{Cambridge, UK}\\
+    \email{vb358@cl.cam.ac.uk}
+\and\alignauthor
+    Alastair R.\ Beresford\\
+    \affaddr{University of Cambridge}\\
+    \affaddr{Cambridge, UK}\\
+    \email{arb33@cl.cam.ac.uk}
 }
-
-\author{Dominic P.\ Mulligan}
-\email{Dominic.Mulligan@arm.com}
-\orcid{0000-0003-4643-3541}
-\affiliation{%
-  \institution{Arm Research}
-  \streetaddress{110 Fulbourn Road}
-  \city{Cambridge}
-  \state{}
-  \postcode{CB1 9NJ}
-  \country{United Kingdom}
-}
-
-\author{Victor B.\ F.\ Gomes}
-\email{vb358@cl.cam.ac.uk}
-\orcid{0000-0002-2954-4648}
-\affiliation{%
-  \institution{Department of Computer Science and Technology}
-  \streetaddress{15 JJ Thomson Avenue}
-  \city{Cambridge}
-  \state{}
-  \postcode{CB3 0FD}
-  \country{United Kingdom}
-}
-
-\author{Alastair R.\ Beresford}
-\email{arb33@cl.cam.ac.uk}
-\orcid{0000-0003-0818-6535}
-\affiliation{%
-  \institution{Department of Computer Science and Technology}
-  \streetaddress{15 JJ Thomson Avenue}
-  \city{Cambridge}
-  \state{}
-  \postcode{CB3 0FD}
-  \country{United Kingdom}
-}
-
-%\renewcommand{\shortauthors}{Kleppmann, Mulligan, Gomes, and Beresford}
+\maketitle
 
@@
 \end{abstract}
-
-\settopmatter{printfolios=true} % Recommended EuroSys formatting
-\maketitle
 
@@
   \draw [network] (6.5,3) to [out=270,in=90] (7.2,0);
-  \node [rotate=90,blue,font=\footnotesize] at (7.5,1.5) {network communication};
+  \node [rotate=90,blue,font=\scriptsize] at (7.5,1.5) {network communication};
   \path [draw,dotted] (-3.1,1.5) -- (8.7,1.5);
   %%%
-  \node [fill=red!15] at (9.83,4.10) {(\emph{a})};
-  \node [fill=red!15] at (14.18,4.10) {(\emph{b})};
-  \node [fill=red!15] at (9.84,1.10) {(\emph{c})};
-  \node [fill=red!15] at (14.15,1.10) {(\emph{d})};
-  \node [rectangle,draw] at (10.9,3) {
+  \node [fill=red!15] at (9.63,4.10) {(\emph{a})};
+  \node [fill=red!15] at (13.98,4.10) {(\emph{b})};
+  \node [fill=red!15] at (9.64,1.10) {(\emph{c})};
+  \node [fill=red!15] at (13.95,1.10) {(\emph{d})};
+  \node [rectangle,draw] at (10.7,3) {
       \begin{tikzpicture}[level distance=7mm]
@@
   };
-  \node [rectangle,draw] at (13.5,3) {
+  \node [rectangle,draw] at (13.3,3) {
       \begin{tikzpicture}[level distance=7mm]
@@
   };
-  \node [rectangle,draw] at (10.7,0) {
+  \node [rectangle,draw] at (10.5,0) {
       \begin{tikzpicture}[level distance=7mm]
@@
   };
-  \node [rectangle,draw] at (13.3,0) {
+  \node [rectangle,draw] at (13.1,0) {
       \begin{tikzpicture}[level distance=7mm]
@@
 \label{fig:move-same-item}
-\Description{}
 \end{figure*}
@@
   \draw [network] (6.0,3) to [out=270,in=90] (6.7,0);
-  \node [rotate=90,blue,font=\footnotesize] at (7.0,1.5) {network communication};
+  \node [rotate=90,blue,font=\scriptsize] at (7.0,1.5) {network communication};
   \path [draw,dotted] (-2.5,1.6) -- (8.5,1.6);
   %%%%%
-  \node [rectangle,draw] at (10.95,3) {
+  \node [rectangle,draw] at (10.75,3) {
       \begin{tikzpicture}[level distance=7mm]
       \useasboundingbox (-1.3,-1.3) rectangle (1,1.2);
-      \node [fill=red!15] at (-1.10,1.03) {(\emph{a})};
+      \node [fill=red!15] at (-1.07,1.02) {(\emph{a})};
       \node at (0,1) {$\mathsf{root}$};
@@
   };
-  \node [rectangle,draw] at (13.8,3) {
+  \node [rectangle,draw] at (13.6,3) {
       \begin{tikzpicture}[level distance=7mm]
       \useasboundingbox (-1.15,-2.3) rectangle (1.15,0.2);
-      \node [fill=red!15] at (0.94,0.03) {(\emph{b})};
+      \node [fill=red!15] at (0.93,0.02) {(\emph{b})};
       \tikzstyle{level 1}=[sibling distance=12mm]
@@
   };
-  \node [rectangle,draw] (start) at (10.95,0) {
+  \node [rectangle,draw] (start) at (10.75,0) {
       \begin{tikzpicture}[level distance=7mm]
       \useasboundingbox (-1.17,-2.3) rectangle (1.15,0.2);
-      \node [fill=red!15] at (-0.98,0.03) {(\emph{c})};
+      \node [fill=red!15] at (-0.95,0.03) {(\emph{c})};
       \node {$\mathsf{root}$} child {node {$A$} child {node {$B$}} child {node {$C$}}};
@@
   };
-  \node [rectangle,draw] (right) at (13.8,0) {
+  \node [rectangle,draw] (right) at (13.6,0) {
       \begin{tikzpicture}[level distance=7mm]
       \useasboundingbox (-1.15,-2.3) rectangle (1.15,0.2);
-      \node [fill=red!15] at (0.93,0.03) {(\emph{d})};
+      \node [fill=red!15] at (0.91,0.03) {(\emph{d})};
       \node {$\mathsf{root}$} child {node {$B$} child {node {$A$} child {node {$C$}}}};
@@
 Boxes (\emph{a}) to (\emph{d}) show possible outcomes after the replicas have communicated and merged their states.}\label{fig:move-cycle}
-\Description{Visualisation of trees with concurrent move operations}
 \end{figure*}
@@
   \label{fig:gdrive-error}
-  \Description{Screenshot showing the errors ``Can't remove files with changes pending. Try again after sync is complete.'' and ``An unknown error occurred.''}
 \end{figure*}
@@
 \begin{isabellebody}
-\internallinenumbers
+\internallinenumbers\setstretch{1.2}
 \isacommand{datatype}\isamarkupfalse%
@@
 \label{fig:code}
-\Description{The move operation algorithm, implemented in the Isabelle/HOL language.}
 \end{figure*}
@@
 The \isa{apply\_op} function (lines 43--51) takes two arguments: a \isa{Move} operation to apply and the current replica state; and it returns the new replica state.
-The constraint \isa{'t::\{linorder\}} in the type signature indicates that timestamps \isa{'t} are instances of \isa{linorder} type class, and they can therefore be compared with the \isa{<} operator defining a linear (or total) order.
+The \isa{'t::\{linorder\}} constraint in the type signature indicates that timestamps \isa{'t} are instances of \isa{linorder} type class, and they can therefore be compared with the \isa{<} operator defining a linear (or total) order.
 This comparison occurs on line 49.
@@
 All theorems stated here have been formally proved and mechanically checked using Isabelle.
-For space reasons we give only a brief sketch of the proofs in this paper; the Isabelle files containing the full details are open source [URL omitted for blind review].
-
-% TODO \cite{SourceFiles}
+For space reasons we give only a brief sketch of the proofs in this paper; the Isabelle files containing the full details are open source~\cite{SourceFiles}.
 
@@
 
-\noindent Here the \isa{\isacharat} operator concatenates two lists.
+\noindent The \isa{\isacharat} operator concatenates two lists.
 Let \isa{t1 = move\_time oper1} and \isa{t2 = move\_time oper2} be the timestamps of the two operations \isa{oper1} and \isa{oper2}.
@@
 \begin{enumerate}
-    \item \isa{t3 < t1 < t2}.
+    \item \isa{t3 < t1 < t2}.\\
         Then neither applying \isa{oper1} nor applying \isa{oper2} will undo \isa{logop}.
         Thus, the two operations are applied in the same way as in the base case.
-    \item \isa{t1 < t3 < t2}.
-        Then \isa{apply\_op oper2 (apply\_op oper1 (logop {\isacharhash} ops, tree))} first undoes \isa{logop}, then applies \isa{oper1} using \isa{do\_op}, then reapplies \isa{logop} using \isa{redo\_op}, and finally applies \isa{oper2} using \isa{do\_op}.
-        On the other hand, \isa{apply\_op oper1 (apply\_op oper2 (logop {\isacharhash} ops, tree))} first applies \isa{oper2} using \isa{do\_op}, then undoes it again, then undoes \isa{logop}, applies \isa{oper1} and reapplies \isa{logop} as before, and finally reapplies \isa{oper2} using \isa{redo\_op}.
+    \item \isa{t1 < t3 < t2}. Then\\
+        \isa{apply\_op oper2 (apply\_op oper1 (logop {\isacharhash} ops, tree))}\\first undoes \isa{logop}, then applies \isa{oper1} using \isa{do\_op}, then reapplies \isa{logop} using \isa{redo\_op}, and finally applies \isa{oper2} using \isa{do\_op}.
+        On the other hand,\\\isa{apply\_op oper1 (apply\_op oper2 (logop {\isacharhash} ops, tree))}\\first applies \isa{oper2} using \isa{do\_op}, then undoes it again, then undoes \isa{logop}, applies \isa{oper1} and reapplies \isa{logop} as before, and finally reapplies \isa{oper2} using \isa{redo\_op}.
         The end result is the same.
-    \item \isa{t1 < t2 < t3}.
+    \item \isa{t1 < t2 < t3}.\\
         Then both applying \isa{oper1} and applying \isa{oper2} first undoes \isa{logop}, yielding \isa{tree' = undo\_op (logop, tree)}.
@@
 Using our commutativity proof above we integrate our tree datatype with this framework, and thus prove that our move operation on trees does indeed guarantee strong eventual consistency.
-The details appear in our Isabelle theory files [URL omitted for blind review]. % \cite{SourceFiles}.
+The details appear in our Isabelle theory files~\cite{SourceFiles}.
 
@@
 the trees are extensionally equivalent, despite having very different in-memory representations.
-We can also prove corollaries of \isa{apply\_ops\_commutes} and \isa{apply\_ops\_acyclic} for the hash-map-based implementation.
+We can also prove corollaries of the theorems \isa{apply\_ops\_commutes} and \isa{apply\_ops\_acyclic} for the hash-map-based implementation.
 %The analogue of \isa{apply\_ops\_unique\_parent} holds trivially since by construction, the hash-map contains at most one (metadata, parent) pair for a given child.
@@
 
-\subsection{Experimental setup and results}
+\subsection{Performance of move operation}
 
@@
   \label{tab:rpc-times}
+  \vspace{4pt}
   \begin{tabular}{r|rrr}
@@
 \begin{figure}
+  \vspace{20pt}
   \includegraphics[width=\columnwidth,keepaspectratio=true]{plot.pdf}
@@
   \label{fig:plots}
-  \Description{Plot in which the x axis goes from 0 to 600 move operations per second, and the y axis goes from 0 to 2 milliseconds.
-Local operations take a near-constant time around 50 microseconds; the time for remote operations increases linearly from 0.2ms to 1.4ms.}
 \end{figure}
 
-\subsection{Interpretation of results}
-
-A locally generated operation always has a timestamp greater than any existing operation at the generating replica at that time (by definition of Lamport timestamps).
-Applying that operation requires only a \isa{do\_op}, and no \isa{undo\_op} or \isa{redo\_op}, hence the near-constant execution time.
-
-However, for remote operations, a number of undos and redos must be performed, depending on the degree of concurrency.
+\begin{figure}
+  \vspace{20pt}
+  \includegraphics[width=\columnwidth,keepaspectratio=true]{leader-plot.pdf}
+  \caption{Median time to apply a local CRDT operation compared to the median time to perform a move operation using state machine replication, with a leader located in another region.
+Note the log scale on both axes.}
+  \label{fig:leader-plot}
+\end{figure}
+
+Local operations take constant time is because a locally generated operation always has a timestamp greater than any existing operation at the generating replica at that time (by definition of Lamport timestamps).
+Applying that operation requires only a \isa{do\_op}, and no \isa{undo\_op} or \isa{redo\_op}.
+
+On the other hand, for remote operations, a number of undos and redos must be performed, depending on the degree of concurrency.
 As the time interval between successive operations is smaller than the network delay between replicas, there are multiple operations ``in flight'' at the same time, and more calls to \isa{undo\_op} and \isa{redo\_op} are required to order operations by timestamp.
@@
 
-A hand-optimised implementation of our algorithm could achieve higher throughput than our 600 operations/sec, since code generated from HOL definitions contains inefficiencies.
-The generated code could serve as a reference implementation against which other implementations could be tested.
-
-Nevertheless, our evaluation sends a positive signal about the practical applicability of our algorithm.
-When a user is interacting with a system, they only have to wait for local operations to execute; any remote operations can be applied in the background without affecting user interaction.
+Note that when a user is interacting with the system, they only have to wait for local operations to execute; any remote operations can be applied in the background without affecting user interaction.
 Since our algorithm need not perform any undos or redos for local operations, these user interactions are consistently fast.
 
-The alternative to our optimistic replication approach would be to use a central server (master replica) or consensus protocol that processes all operations in a serialisable order.
-Performing an operation then requires waiting for communication with a server potentially located on a different continent, taking over 100~ms (Table~\ref{tab:rpc-times}); by contrast, a local operation taking around 50~{\textmu}s is at least three orders of magnitude faster.
-Moreover, a local operation requires no waiting for network communication, and thus it supports disconnected operation: mobile devices can read and modify the replicated tree even while disconnected from the Internet, which is not the case with centralised systems.
+\subsection{Comparison to state machine replication}\label{sec:smr}
+
+A simple alternative to our CRDT algorithm is to use state machine replication~\cite{Schneider:1990vy}: that is, we use a leader replica or consensus algorithm to impose a total order on all operations, and then execute operations in that same order on all replicas.
+For a move operation on trees, the state machine replication algorithm is much simpler than the CRDT: it still needs to check for cycles, but it never needs to undo or redo any operations because they are never applied out-of-order.
+
+To compare the performance of our algorithm to the state machine approach we ran another set of experiments on the same three replicas in California, Ireland, and Singapore.
+In this experiment, the Californian replica was designated leader; it totally ordered all operations it received, and sent them to all other replicas in the same order.
+The Irish and Singaporean replicas generated operations, sent them to the leader, and applied them to their local tree in the order they were received from the leader.
+In order to ensure a fair comparison to the CRDT algorithm, these experiments used the same Isabelle-generated Scala code.
+
+The results from this experiment are shown in Figure~\ref{fig:leader-plot}.
+The leader-based approach was able to sustain a throughput of 14,000 move operations per second, approximately 23 times the CRDT's throughput of 600 ops/sec (shown as an asymptote in Figure~\ref{fig:leader-plot}), due to the fact that it does not need to spend CPU cycles undoing and redoing operations.
+On the other hand, the latency of each operation is much higher in the state machine approach, because the generating replica needs to wait for a round trip through the leader.
+This latency is around 145~ms for operations generated in Ireland and around 176~ms for Singapore (cf.
+Table~\ref{tab:rpc-times}), compared with the CRDT's 50~{\textmu}s processing time for local operations.
+
+Therefore, we have a clear trade-off: in applications that need to prioritise throughput, a state machine replication approach is preferable, while in applications that need to minimise response times to user requests, our CRDT algorithm is preferable.
+Our algorithm also has the advantage that local user operations require no waiting for network communication, making them highly available in the face of network failures.
+It also supports offline operation: mobile devices can read and modify their local replica of the tree even while disconnected from the Internet, which is not the case with leader-based replication.
+
+We note that the throughput of 600 ops/sec is for a single tree, e.g.\ a filesystem belonging to one user.
+Operations on independent trees, belonging to different users, can be processed in parallel.
+We believe that for many applications, 600 ops/sec/user is plenty, and the ability to work offline is a valuable benefit.
 
@@
 
-Many replicated data systems use optimistic replication~\cite{Saito:2005jw}, which allows the state of replicas to temporarily diverge, in order to achieve better performance and availability in the presence of faults than strongly consistent systems~\cite{Bailis:2014th,Gilbert:2002il}.
+Many replicated data systems use optimistic replication \cite{Saito:2005jw}, which allows the state of replicas to temporarily diverge, in order to achieve better performance and availability in the presence of faults than strongly consistent systems~\cite{Bailis:2014th,Gilbert:2002il}.
 As a consequence, these systems require a mechanism for merging or reconciling conflicting updates that were made concurrently on different replicas.
 For example, version control systems such as Git~\cite{Chacon:2014kr} leave conflicts for the user to resolve manually.
-Systems such as Dynamo~\cite{DeCandia:2007ui} and Bayou~\cite{Terry:1995dn} rely on the application programmer to provide explicit conflict resolution logic; however, such logic is difficult to get right \cite{Bailis:2013jc,Burckhardt:2014hy,Gomes:2017gy}.
+Databases such as Dynamo~\cite{DeCandia:2007ui} and Bayou~\cite{Terry:1995dn} rely on the application programmer to provide explicit conflict resolution logic; however, such logic is difficult to get right \cite{Bailis:2013jc,Burckhardt:2014hy,Gomes:2017gy}.
 Hence, we focus on systems that automatically ensure that all replicas converge towards a consistent state, without requiring custom application logic---a consistency model known as \emph{strong eventual consistency}~\cite{Shapiro:2011un,Gomes:2017gy}.
@@
 
-Our algorithm is an example of an operation-based Conflict-free Replicated Data Type or CRDT~\cite{Shapiro:2011wy,Shapiro:2011un,Burckhardt:2014ft}.
+Our algorithm is an operation-based Conflict-free Replicated Data Type or CRDT~\cite{Shapiro:2011wy,Shapiro:2011un,Burckhardt:2014ft}.
 All CRDTs share the property that concurrent changes on different replicas can be merged in any order; any two replicas that have seen the same set of updates are guaranteed to be in the same state, regardless of the order in which they processed these updates.
@@
 \begin{itemize}
-    \item Martin et al.~\cite{Martin:2010ih} define a CRDT for XML data, and Kleppmann and Beresford~\cite{Kleppmann:2016ve} define a CRDT for JSON.
+    \item Martin et al.~\cite{Martin:2010ih} define a CRDT for XML data, and Kleppmann and Beresford~\cite{Kleppmann:2016ve} define a CRDT with a JSON data model.
         However, these algorithms only deal with insertion and deletion of tree nodes, and do not support moves.
@@
 Molli et al.~\cite{Molli:2003cd} define an OT tree structure with a move operation.
-However, it requires that all communication between replicas is performed via total order broadcast, which in turn requires the use of a sequencer server or a consensus algorithm~\cite{Chandra:1996cp}.
+However, it requires that all communication between replicas is performed via total order broadcast, which requires a leader replica or consensus algorithm, like in \S\ref{sec:smr}.
 Our algorithm has better availability characteristics in the presence of network partitions because it allows messages to be delivered in any order, e.g.\ via peer-to-peer protocols.
@@
 Many distributed filesystems, such as NFS, rely on synchronous interaction with a server.
-While this has the advantage of not requiring conflict resolution, it rules out users working while offline.
+This avoids the need for conflict resolution, but rules out users working offline.
 
@@
 It permits offline updates to both replicas.
-Like Rumor, Unison is a userspace process that works by comparing replica states.
+Like Rumor, Unison is a userspace process that compares replica states.
 Whenever it is run, Unison records a summary of the filesystem state on each replica, and it uses this summary to determine the changes made since the last synchronisation. 
@@
 
-In this paper we have defined a novel algorithm that handles arbitrary concurrent modifications of a tree data structure---adding, moving, and removing nodes---in a peer-to-peer replication setting.
+In this paper we have defined a novel algorithm that handles arbitrary concurrent modifications of a tree data structure -- adding, moving, and removing nodes -- in a peer-to-peer replication setting.
 It is applicable to distributed filesystems and many other applications that use a tree-structured data model.
@@
 We also extracted a formally verified Scala implementation from our Isabelle/HOL definitions, and evaluated its performance across three replicas in California, Ireland and Singapore.
-Our implementation applies local updates in approximately 50~{\textmu}s, orders of magnitude faster than is possible with a central server or consensus protocol operating over these distances.
-It is able to sustain 600 updates per second; a more scalable implementation could be created by optimising the automatically extracted Scala code.
+Our implementation applies local updates in approximately 50~{\textmu}s, orders of magnitude faster than is possible with a leader replica or consensus protocol operating over these distances, and it remains available in the face of network interruptions.
+However, compared to leader-based replication our algorithm has lower throughput of 600 updates per second.
+A hand-optimised implementation of our algorithm could achieve higher performance, since code generated from HOL definitions contains inefficiencies.
+The generated code could serve as a reference implementation against which other implementations could be tested.
 
 As mentioned in the introduction, our work is especially interesting due to the ubiquity of tree data models across many different computing systems.
-In future work we hope to build novel peer-to-peer distributed systems, such as a distributed filesystem, based on our algorithm.
-
-\begin{acks}
+In future work we hope to build novel applications, such as a distributed filesystem, using our algorithm.
+
+\section*{Acknowledgements}
+
 The authors wish to acknowledge the support of The Boeing Company and the EPSRC ``REMS: Rigorous Engineering for Mainstream Systems'' programme grant (EP/K008528).
-Martin Kleppmann is supported by a Leverhulme Trust Early Career Fellowship and by the Isaac Newton Trust.
+Martin Kleppmann is supported by a Leverhulme Early Career Fellowship and by the Isaac Newton Trust.
 Our evaluation was conducted using AWS credits from the AWS Educate program.
 Thank you to Marc Shapiro for feedback on a draft of this paper.
-\end{acks}
-
-\bibliographystyle{ACM-Reference-Format}
+
+%\balance
+\bibliographystyle{abbrv}
 \bibliography{references}{}

### 2020-03-01 paper/move-op.tex
Provide more databases context

@@
 \begin{abstract}
-    Replicated tree data structures are a fundamental building block of distributed filesystems, such as Google Drive and Dropbox, and collaborative applications with a JSON or XML data model.
+    Replicated tree data structures are data models supporting distributed filesystems, such as Google Drive and Dropbox, and XML or JSON databases.
     These systems need to support a \emph{move} operation that allows a subtree to be moved to a new location within the tree.
     However, such a move operation is difficult to implement correctly if different replicas can concurrently perform arbitrary move operations, and we demonstrate bugs in Google Drive and Dropbox that arise with concurrent moves.
-    In this paper we present a CRDT algorithm that handles arbitrary concurrent modifications on trees, while ensuring that the tree structure remains valid (in particular, no cycles are introduced), and guaranteeing that all replicas converge towards the same consistent state.
-    We formally prove the correctness of our algorithm using the Isabelle/HOL proof assistant.
-    Our formally verified implementation achieves a throughput of 600 operations/sec in a geo-replicated setting.
+    In this paper we present an algorithm that handles arbitrary concurrent modifications on trees, while ensuring that the tree structure remains valid (in particular, no cycles are introduced), and guaranteeing that all replicas converge towards the same consistent state.
+    Our algorithm requires no synchronous coordination between replicas, making it highly available in the face of network partitions.
+    We formally prove the correctness of our algorithm using the Isabelle/HOL proof assistant, and evaluate the performance of our formally verified implementation in a geo-replicated setting.
 \end{abstract}
@@
 
-Many systems use a tree-structured data model.
-For example:
-\begin{itemize}
-    \item The \textbf{filesystem} on most operating systems is a tree: directories are branch nodes, and files are leaf nodes.
-        In Unix systems the nodes of the tree are known as \emph{inodes}.
-        Strictly speaking, filesystems with hardlinks form a DAG; we will focus on trees for now and show in \S\ref{sec:extensions} how to support links.
-    \item \textbf{Rich text editors} maintain a HTML or XML structure: a tree of textual elements such as paragraphs, lists, figures, sections, and so on.
-        These can be nested: for example, a paragraph may appear inside a bulleted list item, which may in turn appear inside an enumeration.
-    \item \textbf{Vector graphics} and \textbf{presentation software} represent images using graphical objects such as text boxes, rectangles, lines, and so on~\cite{Wallace:2019vf}.
-        These objects are contained within nodes representing pages of a document, or slides of a presentation.
-        Objects may be combined into a \emph{group} and manipulated as a unit; this corresponds to making these objects children of a common parent node.
-        Multiple objects and groups may be combined further into higher-level groups, forming a tree.
-    \item \textbf{Note-taking} and \textbf{task-management tools} such as Org-mode for Emacs~\cite{OrgMode} or OmniOutliner~\cite{OmniOutliner} present the user with a tree structure that they can inspect and manipulate.
-\end{itemize}
-
-In this paper we consider applications that use such a tree data model, and that replicate this tree across multiple nodes.
-Moreover, we focus on \emph{optimistic replication}~\cite{Saito:2005jw}, that is, systems in which any replica can autonomously make changes to the data, without waiting for communication or coordination with any other replicas.
-Such systems have the advantage that they can continue processing read and write requests even in the presence of arbitrary network partitions; in other words, they are \emph{available} and \emph{partition-tolerant} in the sense of the CAP theorem~\cite{Gilbert:2002il}.
-This approach is desirable since it enables disconnected operation on mobile devices, and high availability in geo-replicated settings.
-
-As the tree structure is concurrently modified on different replicas, the state of these replicas temporarily diverges.
-In this paper we show how we can nevertheless achieve \emph{strong eventual consistency}~\cite{Shapiro:2011un,Gomes:2017gy}: as the replicas communicate, we guarantee that they converge to a consistent state.
-Our algorithm for achieving convergence is an example of a \emph{Conflict-free Replicated Data Type} or CRDT~\cite{Shapiro:2011wy,Shapiro:2011un}.
-
-In this paper we use an abstract model of a tree that can support any of the applications listed above, including distributed filesystems.
-We allow replicas to manipulate this tree in any way: by creating new nodes, deleting nodes, or moving subtrees to a new location within the tree.
-While there are many existing systems that support creating and deleting nodes (see \S\ref{sec:relwork}), the key innovation of our algorithm is the support for moving subtrees.
-We explain in \S\ref{sec:move-is-hard} why this move operation is so challenging.
-Our contributions are:
+Many data management applications use a tree-structured or hierarchical data model.
+For example, most file systems are trees (directories are branch nodes, and files are leaf nodes), and databases are used to store this directory structure and file metadata~\cite{Niazi:2017uf,Liu:2019zn}.\footnote{Strictly speaking, filesystems with hardlinks form a DAG; we will show in \S\ref{sec:extensions} how to support hardlinks.}
+Moreover, XML databases and JSON document databases also operate on trees, which might represent rich text (a tree of paragraphs, lists, figures, sections, etc.), vector graphics, CAD drawings, and many other types of data.
+Typically, a graphical user interface allows a user to edit this information interactively, resulting in updates to the underlying XML/JSON database: adding or deleting nodes in the tree, and moving nodes from one position in the tree to another.
+
+When updating these trees, standard concurrency control techniques can be used, such as two-phase locking~\cite{Bernstein:1987va} or optimistic concurrency control~\cite{Cahill:2008eg}.
+However, these techniques require synchronous coordination between replicas.
+If we want the system to continue processing read and write requests even in the presence of arbitrary network partitions (in other words, if we want \emph{availability} and \emph{partition-tolerance} in the sense of the CAP theorem~\cite{Gilbert:2002il}), then we must avoid such coordination~\cite{Davidson:1985hv}.
+Coordination avoidance has shown to be beneficial in several data systems~\cite{Bailis:2014th,AnnaKVS}, and network partitions are common~\cite{Alquraan:2018wj,Bailis:2014jx}.
+
+We can use \emph{optimistic replication}~\cite{Saito:2005jw}, which means that any replica can make changes to the data without waiting for communication with any other replicas; operations are sent to other replicas at some later time when a network connection is available.
+This approach has the advantage of enabling disconnected operation on mobile devices, and high availability in geo-replicated settings.
+It also improves performance for end users because the duration of a user request is independent of any network latency.
+
+Dropbox and Google Drive are widely-deployed examples of optimistically replicated filesystems: they run a daemon on the user's machine that watches a designated directory for changes, and locally maintains a database with the user's tree of files and directories.
+The user can read and arbitrarily modify the files on their local disk, even while their computer is offline.
+However, when the filesystem is concurrently updated on different computers, Google Drive and Dropbox exhibit bugs in their concurrency control, as we show in \S\ref{sec:move-is-hard}.
+
+In this paper we introduce a novel algorithm for handling concurrent updates to a replicated tree, such as a filesystem or XML database.
+It allows replicas to manipulate the tree by creating new nodes, deleting nodes, or moving subtrees to a new location within the tree.
+In order to rule out bugs like those in Google Drive we formally prove the correctness of our algorithm using the Isabelle/HOL proof assistant.
+
+Our algorithm supports optimistic replication, allowing replicas to temporarily diverge as they are updated, and ensuring that they always converge towards a consistent state.
+It is an example of a \emph{Conflict-free Replicated Data Type} or CRDT~\cite{Shapiro:2011wy,Shapiro:2011un}, and it guarantees a consistency model called \emph{strong eventual consistency}~\cite{Shapiro:2011un,Gomes:2017gy}.
+Our contributions in this paper are:
 \begin{itemize}
@@
   \includegraphics[width=\textwidth,keepaspectratio=true]{gdrive-error.png}
-  \caption{Error message produced by Google Drive Backup and Sync on Mac OS as a result of performing the operations shown in Figure~\ref{fig:move-cycle}, indicating a bug in the underlying sync implementation.}
+  \caption{Error message produced by Google Drive Backup and Sync on Mac OS as a result of performing the operations shown in Figure~\ref{fig:move-cycle}, indicating a bug in the underlying replication algorithm.}
   \label{fig:gdrive-error}
@@
 
-If a move operation is implemented by deleting the moved subtree from its old location, and then re-creating it at the new location, the merged state will be as shown in Figure~\ref{fig:move-same-item}a: the concurrent moves will duplicate the moved subtree, since each move independently recreates the subtree in each destination location.
+If a move operation is implemented by deleting the moved subtree from its old location, and then re-creating it at the new location, the merged state will be as shown in Figure \ref{fig:move-same-item}a: the concurrent moves will duplicate the moved subtree, since each move independently recreates the subtree in each destination location.
 We believe that this duplication is undesirable, since subsequent edits to nodes in the duplicated subtree will apply to only one of the copies.
@@
 This restriction is required because allowing this operation would introduce a cycle into the directory graph, and so the filesystem would no longer be a tree.
-The same restriction is required in any other tree structure that supports a move operation.
+This restriction is required in any other tree structure that supports a move operation.
 
@@
 Here, replica 1 moves $B$ to be a child of $A$, while concurrently replica 2 moves $A$ to be a child of $B$.
-As each replica propagates its operation to the other replica, a careless implementation might end up in the state shown in Figure~\ref{fig:move-cycle}a, in which $A$ and $B$ have formed a cycle, detached from the tree.
+As each replica propagates its operation to the other replica, a careless implementation might end up as shown in Figure~\ref{fig:move-cycle}a, in which $A$ and $B$ have formed a cycle, detached from the tree.
 
@@
 This error state persisted until the directories on the erroring replica were manually moved to match the state of the other replica.
+We have reported this bug to Google.
 On the other hand, Dropbox exhibited the duplication behaviour shown in Figure~\ref{fig:move-cycle}b.
 
-\subsection{A highly-available move operation: impossible?}\label{sec:impossibility}
+\subsection{Is a highly-available move operation impossible?}\label{sec:impossibility}
 
@@
 
-We define one kind of operation: \isa{Move t p m c} (line 1--5).
+We allow the tree to be updated in three ways: by creating a new child of any parent node, by deleting a node, or by moving a node to be a child of a new parent.
+However, all three types of update can be represented by a move operation.
+To create a node, we generate a fresh ID for that node, and issue an operation to move this new ID to be a child of its desired parent; the node is then implicitly created.
+We also designate as ``trash'' some node ID that does not exist in the tree; then we can delete a node by moving it to be a child of the trash.
+
+Thus, we define one kind of operation: \isa{Move t p m c} (Figure \ref{fig:code}, lines 1--5).
 A move operation is a 4-tuple consisting of a timestamp \isa{t} of type \isa{'t}, a parent node ID \isa{p} of type \isa{'n}, a metadata field \isa{m} of type \isa{'m}, and a child node ID \isa{c} of type \isa{'n}.
-Here, \isa{'t}, \isa{'n} and \isa{'m} are \emph{type variables} that can be replaced with arbitrary types; we only require that node identifiers \isa{'n} are globally unique (e.g.\ UUIDs~\cite{Leach:2005hm}); timestamps \isa{'t} need to be globally unique and totally ordered (e.g.
-Lamport timestamps~\cite{Lamport:1978jq}).
+Here, \isa{'t}, \isa{'n} and \isa{'m} are \emph{type variables} that can be replaced with arbitrary types; we only require that node identifiers \isa{'n} are globally unique (e.g.\ UUIDs); timestamps \isa{'t} need to be globally unique and totally ordered (e.g.\ Lamport timestamps~\cite{Lamport:1978jq}).
 
@@
 If \isa{c} does not currently exist in the tree, it is created as a child of \isa{p}.
-For this reason we do not need a separate operation type to create tree nodes: simply performing a move with a fresh child node ID \isa{c} is sufficient to create \isa{c}.
-To delete \isa{c}, we move \isa{c} to be a child of the ``trash'' (a designated node ID that does not exist in the tree), so \isa{c} is removed from the tree.
 
@@
 
+This resolution of a conflict between two operations can be generalised to any number of conflicting operations by applying the rules repeatedly pairwise.
+
 Note that the safety of an operation (whether or not it would introduce a cycle) may change as subsequent operations with lower timestamps are applied.
@@
 Another useful extension of the tree algorithm is to allow children of the same parent node to have an explicit ordering.
-For example, in a graphics application, the order of nodes determines visibility (objects that are ``further back'' may be obscured by objects that are ``closer to the front'').
-This can be implemented by maintaining an additional list CRDT for each branch node, e.g.\ using RGA~\cite{Roh:2011dw}, Treedoc~\cite{Preguica:2009fz}, Logoot~\cite{Weiss:2010hx}, or LSEQ~\cite{Nedelec:2013ky}.
+For example, in XML, the set of children of an element is ordered.
+This can be implemented by maintaining an additional list CRDT for each branch node, e.g.\ using RGA~\cite{Roh:2011dw} or Logoot~\cite{Weiss:2010hx}.
 These algorithms assign a unique ID to each element of the list, and this ID can be included in the metadata field of move operations in order to determine the order of sibling nodes.
 
-This approach to determining ordering also easily supports reordering of child nodes within a parent, e.g.\ to implement the ``bring to front'' and ``send to back'' feature of applications like Microsoft PowerPoint.
-To move a node to a different position in a list, we use the list CRDT to generate a new ID at the desired position in the sequence.
+This approach to determining ordering also easily supports reordering of child nodes within a parent: to move a node to a different position in a list, we use the list CRDT to generate a new ID at the desired position in the sequence.
 Then we perform a move operation in which the parent node is unchanged, but the metadata is changed to this new ID.
@@
 
+\subsubsection{Other correctness properties}
+
+There are further criteria we might use to determine if our algorithm is correct.
+For example, we might demonstrate that a single-replica system operates with the usual sequential semantics of a tree.
+In a system with multiple disjoint trees, we could prove that the trees don't get tangled together.
+We conjecture that those properties hold for our algorithm, but leave the proof out of scope for this paper.
+
 \subsection{Convergence}\label{sec:convergence}
@@
 
-To prove this theorem we must first show that \isa{undo\_op} performs the inverse of \isa{do\_op}, that is:
+To prove this theorem we must first show that \isa{undo\_op} performs the inverse of \isa{do\_op} and that every operation can be undone, that is:
 \begin{isabelle}
@@
 \end{enumerate}
+
+This completes the proof that any two operations commute, provided that their timestamps are distinct from each other and from the contents of the log.
+We can easily generalise this to show that any number of operations commute: given any sequence of operations, any permutation of that sequence will have the same effect, because we can produce the permutation by repeatedly swapping adjacent, commutative operations.
 
@@
 \begin{figure}
-  \vspace{20pt}
   \includegraphics[width=\columnwidth,keepaspectratio=true]{plot.pdf}
@@
 \begin{figure}
-  \vspace{20pt}
   \includegraphics[width=\columnwidth,keepaspectratio=true]{leader-plot.pdf}
@@
 All CRDTs share the property that concurrent changes on different replicas can be merged in any order; any two replicas that have seen the same set of updates are guaranteed to be in the same state, regardless of the order in which they processed these updates.
-CRDTs have been defined for various abstract datatypes, such as registers and counters \cite{Shapiro:2011wy,Shapiro:2011un}, maps \cite{Baquero:2016iv,Kleppmann:2016ve}, sets \cite{Bieniusa:2012wu,Bieniusa:2012gt}, lists (sequences) \cite{Roh:2011dw,Nedelec:2013ky}, and text \cite{Preguica:2009fz,Weiss:2010hx}.
-For trees, several approaches have been proposed:
+Several CRDTs for trees have been proposed:
 \begin{itemize}
@@
         However, these algorithms only deal with insertion and deletion of tree nodes, and do not support moves.
+        A move operation can be emulated by deleting and re-inserting the moved node, but this approach suffers from the duplication problem demonstrated in \S\ref{sec:move-same-item}.
     \item As discussed in \S\ref{sec:impossibility}, Najafzadeh et al.~\cite{Najafzadeh:2017vk,Najafzadeh:2018bw} propose two implementations for a replicated filesystem: a CRDT in which conflicting moves are handled by duplicating tree nodes (as in Figures~\ref{fig:move-same-item}b and~\ref{fig:move-cycle}b), and a centralised implementation in which move operations must obtain a lock before executing (not a CRDT since it relies on synchronous coordination).
@@
 
+Collaborative graphics software Figma uses an approach inspired by CRDTs, but prevents cycles in their object tree by relying on a central server; its replication protocol allows objects to temporarily disappear while syncing~\cite{Wallace:2019vf}.
+
 \subsection{Distributed filesystems}\label{sec:filesystems}
@@
 
-Hughes et al.~\cite{Hughes:2016fp} test Dropbox and Google Drive against a formal specification, but they do not consider moving files, and thus do not find the issue described in \S~\ref{sec:move-cycle}.
+Hughes et al.~\cite{Hughes:2016fp} test Dropbox and Google Drive against a formal specification, but they do not consider moving files, and thus they do not find the issue described in \S\ref{sec:move-cycle}.
 
@@
 
-Ordering operations by a timestamp, and undoing/redoing them as necessary so that they take effect in ascending timestamp order, is a well-known approach.
-It is used for example in systems like Bayou~\cite{Terry:1995dn}, Jefferson's \emph{Time Warp} mechanism~\cite{Jefferson:1985em}, Shapiro et al.'s \emph{capricious total order}~\cite{Shapiro:2016hp}, and Burckhardt's \emph{standard conflict resolution}~\cite[\S~4.3.3]{Burckhardt:2014hy}.
+Our approach of ordering operations by a timestamp, and undoing/redoing them as necessary so that they take effect in ascending timestamp order, is conceptually a very simple approach---although the details of the implementation are a bit subtle.
+Similar ideas appear in many other systems, including the Bayou database~\cite{Terry:1995dn}, Jefferson's \emph{Time Warp} mechanism~\cite{Jefferson:1985em}, Shapiro et al.'s \emph{capricious total order}~\cite{Shapiro:2016hp}, and Burckhardt's \emph{standard conflict resolution}~\cite[{\S}4.3.3]{Burckhardt:2014hy}.
+The concept of undo-redo is also well known from write-ahead logging~\cite{Mohan:1992fe,Mohan:1992hc}.
 
 However, to our knowledge, this approach has not previously been applied to the problem of replicated trees, nor are we aware of previous mechanised proofs (using Isabelle or other tools) that formalise this approach.
+The simplicity of the approach facilitates reasoning about it.
+We are exploring applying undo-do-redo to other replication problems as well, and we expect that our formal verification approach will also benefit other applications of this principle.
 
@@
 
-In this paper we have defined a novel algorithm that handles arbitrary concurrent modifications of a tree data structure -- adding, moving, and removing nodes -- in a peer-to-peer replication setting.
-It is applicable to distributed filesystems and many other applications that use a tree-structured data model.
-Our approach ensures that all replicas converge to the same consistent state without requiring any manual conflict resolution, and without needing application developers to implement conflict handling logic.
+We have defined a novel algorithm that handles arbitrary concurrent modifications of a tree data structure---adding, moving, and removing nodes---in a peer-to-peer replication setting.
+It is applicable to distributed filesystems, databases, and applications that use a tree-structured data model.
+Our approach ensures that all replicas converge to the same consistent state without needing any manual conflict resolution, and without requiring application developers to implement conflict handling logic.
 Updates made to a local replica take effect immediately, while operations from remote replicas can be propagated and applied in the background.
@@
 
-While existing systems exhibit anomalies (such as duplicating nodes or introducing cycles) or bugs and internal errors (like in Figure~\ref{fig:gdrive-error}) in the face of concurrent move operations, our algorithm is free from such problems.
-We formally verified the correctness of our algorithm using the Isabelle/HOL proof assistant; our theorems show that replicas can apply operations in any order, and that the result is always a valid tree (nodes have at most one parent, and the graph does not contain any cycles).
-Moreover, these results apply to unbounded executions and an arbitrary number of replicas---an inherent advantage of using a proof assistant rather than other formal approaches such as model checking.
+The principle behind our algorithm is easy to understand: undoing and redoing operations so that they are effectively executed in timestamp order.
+Nevertheless, it solves a real problem that has not been solved correctly in widely deployed software such as Google Drive and Dropbox, as we demonstrated in \S\ref{sec:move-is-hard}.
+To rule out such bugs, we formally verified the correctness of our algorithm using the Isabelle/HOL proof assistant; our theorems show that replicas can apply operations in any order, and that the result is always a valid tree (nodes have at most one parent, and the graph does not contain any cycles).
+Moreover, these results apply to unbounded executions and an arbitrary number of replicas---an advantage of using a proof assistant rather than other formal approaches such as model checking.
 
@@
 
-As mentioned in the introduction, our work is especially interesting due to the ubiquity of tree data models across many different computing systems.
+As highlighted in the introduction, our work is especially interesting due to the ubiquity of tree data models across many different types of applications and databases.
 In future work we hope to build novel applications, such as a distributed filesystem, using our algorithm.

### 2020-03-01 evaluation/data/leader_mode.sh
Data from experiments in leader mode

@@ -1,0 +1,12 @@
+#!/bin/bash
+
+echo  ",,us_west_1,,,,,,,ap_southeast_1" > leader_mode.csv
+echo -n "interval,total ops/sec," >> leader_mode.csv
+echo -n "generated ops/sec,RTT min,RTT median,RTT p95,apply min,apply median,apply p95," >> leader_mode.csv
+echo    "generated ops/sec,RTT min,RTT median,RTT p95,apply min,apply median,apply p95" >> leader_mode.csv
+
+for interval in 100000 50000 20000 10000 5000 2000 1000 500 200 100 50 20 10; do
+    eu_times="$(cat logs-leader/interval_${interval}_eu_west_1.log.gz      | gunzip | awk -f leader_mode.awk | tail -n 2 | head -n 1 | tr '\n' ',')"
+    ap_times="$(cat logs-leader/interval_${interval}_ap_southeast_1.log.gz | gunzip | awk -f leader_mode.awk | tail -n 2 | head -n 1 | tr -d '\n')"
+    echo "${interval},${eu_times}${ap_times#*,}" >> leader_mode.csv
+done

### 2020-03-01 paper/move-op.tex
Write up results of leader-based experiment

--- 2019-11-04 paper/move-op.tex
@@
 
-\subsection{Interpretation of results}
-
-A locally generated operation always has a timestamp greater than any existing operation at the generating replica at that time (by definition of Lamport timestamps).
-Applying that operation requires only a \isa{do\_op}, and no \isa{undo\_op} or \isa{redo\_op}, hence the near-constant execution time.
-
-However, for remote operations, a number of undos and redos must be performed, depending on the degree of concurrency.
+\begin{figure}
+  \vspace{20pt}
+  \includegraphics[width=\columnwidth,keepaspectratio=true]{leader-plot.pdf}
+  \caption{Median time to apply a local CRDT operation compared to the median time to perform a move operation using state machine replication, with a leader located in another region.
+Note the log scale on both axes.}
+  \label{fig:leader-plot}
+  \Description{Local CRDT operations take about 50 microseconds, and reach saturation at 600 ops/sec. Leader-based operations take about 150--180ms, and reach saturation at 14,000 ops/sec.}
+\end{figure}
+
+Local operations take constant time is because a locally generated operation always has a timestamp greater than any existing operation at the generating replica at that time (by definition of Lamport timestamps).
+Applying that operation requires only a \isa{do\_op}, and no \isa{undo\_op} or \isa{redo\_op}.
+
+On the other hand, for remote operations, a number of undos and redos must be performed, depending on the degree of concurrency.
 As the time interval between successive operations is smaller than the network delay between replicas, there are multiple operations ``in flight'' at the same time, and more calls to \isa{undo\_op} and \isa{redo\_op} are required to order operations by timestamp.
@@
 For example, version control systems such as Git~\cite{Chacon:2014kr} leave conflicts for the user to resolve manually.
-Systems such as Dynamo~\cite{DeCandia:2007ui} and Bayou~\cite{Terry:1995dn} rely on the application programmer to provide explicit conflict resolution logic; however, such logic is difficult to get right \cite{Bailis:2013jc,Burckhardt:2014hy,Gomes:2017gy}.
+Databases such as Dynamo~\cite{DeCandia:2007ui} and Bayou~\cite{Terry:1995dn} rely on the application programmer to provide explicit conflict resolution logic; however, such logic is difficult to get right \cite{Bailis:2013jc,Burckhardt:2014hy,Gomes:2017gy}.
 Hence, we focus on systems that automatically ensure that all replicas converge towards a consistent state, without requiring custom application logic---a consistency model known as \emph{strong eventual consistency}~\cite{Shapiro:2011un,Gomes:2017gy}.
@@
 We also extracted a formally verified Scala implementation from our Isabelle/HOL definitions, and evaluated its performance across three replicas in California, Ireland and Singapore.
-Our implementation applies local updates in approximately 50~{\textmu}s, orders of magnitude faster than is possible with a central server or consensus protocol operating over these distances.
-It is able to sustain 600 updates per second; a more scalable implementation could be created by optimising the automatically extracted Scala code.
+Our implementation applies local updates in approximately 50~{\textmu}s, orders of magnitude faster than is possible with a leader replica or consensus protocol operating over these distances, and it remains available in the face of network interruptions.
+However, compared to leader-based replication our algorithm has lower throughput of 600 updates per second.
+A hand-optimised implementation of our algorithm could achieve higher performance, since code generated from HOL definitions contains inefficiencies.
+The generated code could serve as a reference implementation against which other implementations could be tested.
 
 As mentioned in the introduction, our work is especially interesting due to the ubiquity of tree data models across many different computing systems.
-In future work we hope to build novel peer-to-peer distributed systems, such as a distributed filesystem, based on our algorithm.
+In future work we hope to build novel applications, such as a distributed filesystem, using our algorithm.
 

### 2020-03-03 paper/move-op.tex
Revert "Reformat move-op paper"
This reverts commit 6298d2abba36097ab8ba547704cfbe9cc04d5804.

--- 2020-03-01 paper/move-op.tex
@@
-\documentclass{vldb}
-
-\usepackage[hidelinks]{hyperref}
-\usepackage{textcomp} % \textmu
-\usepackage{booktabs} % \toprule etc.
-\usepackage{setspace} % \setstretch
-\usepackage{balance}
+\documentclass[sigplan,anonymous]{acmart}
+
+% EuroSys recommended formatting
+\renewcommand\footnotetextcopyrightpermission[1]{}
+\pagestyle{plain}
+
 \usepackage{tikz}
 \usetikzlibrary{arrows.meta}
-
-\urlstyle{rm}
 
@@
 
-\vldbTitle{A highly-available move operation for replicated trees}
-\vldbAuthors{Martin Kleppmann, Dominic P. Mulligan, Victor B. F. Gomes, and Alastair R. Beresford}
-\vldbDOI{https://doi.org/10.14778/xxxxxxx.xxxxxxx}
-\vldbVolume{13}
-\vldbNumber{xxx}
-\vldbYear{2020}
+\setcopyright{rightsretained}
+\copyrightyear{2019}
+\acmYear{2019}
+\acmDOI{}
+\acmConference[]{}{}{}
+\acmBooktitle{}
+\acmPrice{}
+\acmISBN{}
 
@@
 
-\numberofauthors{4}
-\author{
-\alignauthor
-    Martin Kleppmann\\
-    \affaddr{University of Cambridge}\\
-    \affaddr{Cambridge, UK}\\
+\author{Martin Kleppmann}
 \email{mk428@cl.cam.ac.uk}
-\alignauthor
-    Dominic P.\ Mulligan\\
-    \affaddr{Arm Research}\\
-    \affaddr{Cambridge, UK}\\
+\orcid{0000-0001-7252-6958}
+\affiliation{%
+  \institution{Department of Computer Science and Technology}
+  \streetaddress{15 JJ Thomson Avenue}
+  \city{Cambridge}
+  \state{}
+  \postcode{CB3 0FD}
+  \country{United Kingdom}
+}
+
+\author{Dominic P.\ Mulligan}
 \email{Dominic.Mulligan@arm.com}
-\alignauthor
-    Victor B.\ F.\ Gomes\\
-    \affaddr{University of Cambridge}\\
-    \affaddr{Cambridge, UK}\\
+\orcid{0000-0003-4643-3541}
+\affiliation{%
+  \institution{Arm Research}
+  \streetaddress{110 Fulbourn Road}
+  \city{Cambridge}
+  \state{}
+  \postcode{CB1 9NJ}
+  \country{United Kingdom}
+}
+
+\author{Victor B.\ F.\ Gomes}
 \email{vb358@cl.cam.ac.uk}
-\and\alignauthor
-    Alastair R.\ Beresford\\
-    \affaddr{University of Cambridge}\\
-    \affaddr{Cambridge, UK}\\
+\orcid{0000-0002-2954-4648}
+\affiliation{%
+  \institution{Department of Computer Science and Technology}
+  \streetaddress{15 JJ Thomson Avenue}
+  \city{Cambridge}
+  \state{}
+  \postcode{CB3 0FD}
+  \country{United Kingdom}
+}
+
+\author{Alastair R.\ Beresford}
 \email{arb33@cl.cam.ac.uk}
+\orcid{0000-0003-0818-6535}
+\affiliation{%
+  \institution{Department of Computer Science and Technology}
+  \streetaddress{15 JJ Thomson Avenue}
+  \city{Cambridge}
+  \state{}
+  \postcode{CB3 0FD}
+  \country{United Kingdom}
 }
-\maketitle
+
+%\renewcommand{\shortauthors}{Kleppmann, Mulligan, Gomes, and Beresford}
 
@@
 \end{abstract}
+
+\settopmatter{printfolios=true} % Recommended EuroSys formatting
+\maketitle
 
@@
   \draw [network] (6.5,3) to [out=270,in=90] (7.2,0);
-  \node [rotate=90,blue,font=\scriptsize] at (7.5,1.5) {network communication};
+  \node [rotate=90,blue,font=\footnotesize] at (7.5,1.5) {network communication};
   \path [draw,dotted] (-3.1,1.5) -- (8.7,1.5);
   %%%
-  \node [fill=red!15] at (9.63,4.10) {(\emph{a})};
-  \node [fill=red!15] at (13.98,4.10) {(\emph{b})};
-  \node [fill=red!15] at (9.64,1.10) {(\emph{c})};
-  \node [fill=red!15] at (13.95,1.10) {(\emph{d})};
-  \node [rectangle,draw] at (10.7,3) {
+  \node [fill=red!15] at (9.83,4.10) {(\emph{a})};
+  \node [fill=red!15] at (14.18,4.10) {(\emph{b})};
+  \node [fill=red!15] at (9.84,1.10) {(\emph{c})};
+  \node [fill=red!15] at (14.15,1.10) {(\emph{d})};
+  \node [rectangle,draw] at (10.9,3) {
       \begin{tikzpicture}[level distance=7mm]
@@
   };
-  \node [rectangle,draw] at (13.3,3) {
+  \node [rectangle,draw] at (13.5,3) {
       \begin{tikzpicture}[level distance=7mm]
@@
   };
-  \node [rectangle,draw] at (10.5,0) {
+  \node [rectangle,draw] at (10.7,0) {
       \begin{tikzpicture}[level distance=7mm]
@@
   };
-  \node [rectangle,draw] at (13.1,0) {
+  \node [rectangle,draw] at (13.3,0) {
       \begin{tikzpicture}[level distance=7mm]
@@
 \label{fig:move-same-item}
+\Description{}
 \end{figure*}
@@
   \draw [network] (6.0,3) to [out=270,in=90] (6.7,0);
-  \node [rotate=90,blue,font=\scriptsize] at (7.0,1.5) {network communication};
+  \node [rotate=90,blue,font=\footnotesize] at (7.0,1.5) {network communication};
   \path [draw,dotted] (-2.5,1.6) -- (8.5,1.6);
   %%%%%
-  \node [rectangle,draw] at (10.75,3) {
+  \node [rectangle,draw] at (10.95,3) {
       \begin{tikzpicture}[level distance=7mm]
       \useasboundingbox (-1.3,-1.3) rectangle (1,1.2);
-      \node [fill=red!15] at (-1.07,1.02) {(\emph{a})};
+      \node [fill=red!15] at (-1.10,1.03) {(\emph{a})};
       \node at (0,1) {$\mathsf{root}$};
@@
   };
-  \node [rectangle,draw] at (13.6,3) {
+  \node [rectangle,draw] at (13.8,3) {
       \begin{tikzpicture}[level distance=7mm]
       \useasboundingbox (-1.15,-2.3) rectangle (1.15,0.2);
-      \node [fill=red!15] at (0.93,0.02) {(\emph{b})};
+      \node [fill=red!15] at (0.94,0.03) {(\emph{b})};
       \tikzstyle{level 1}=[sibling distance=12mm]
@@
   };
-  \node [rectangle,draw] (start) at (10.75,0) {
+  \node [rectangle,draw] (start) at (10.95,0) {
       \begin{tikzpicture}[level distance=7mm]
       \useasboundingbox (-1.17,-2.3) rectangle (1.15,0.2);
-      \node [fill=red!15] at (-0.95,0.03) {(\emph{c})};
+      \node [fill=red!15] at (-0.98,0.03) {(\emph{c})};
       \node {$\mathsf{root}$} child {node {$A$} child {node {$B$}} child {node {$C$}}};
@@
   };
-  \node [rectangle,draw] (right) at (13.6,0) {
+  \node [rectangle,draw] (right) at (13.8,0) {
       \begin{tikzpicture}[level distance=7mm]
       \useasboundingbox (-1.15,-2.3) rectangle (1.15,0.2);
-      \node [fill=red!15] at (0.91,0.03) {(\emph{d})};
+      \node [fill=red!15] at (0.93,0.03) {(\emph{d})};
       \node {$\mathsf{root}$} child {node {$B$} child {node {$A$} child {node {$C$}}}};
@@
 Boxes (\emph{a}) to (\emph{d}) show possible outcomes after the replicas have communicated and merged their states.}\label{fig:move-cycle}
+\Description{Visualisation of trees with concurrent move operations}
 \end{figure*}
@@
   \label{fig:gdrive-error}
+  \Description{Screenshot showing the errors ``Can't remove files with changes pending. Try again after sync is complete.'' and ``An unknown error occurred.''}
 \end{figure*}
@@
 \begin{isabellebody}
-\internallinenumbers\setstretch{1.2}
+\internallinenumbers
 \isacommand{datatype}\isamarkupfalse%
@@
 The \isa{apply\_op} function (lines 43--51) takes two arguments: a \isa{Move} operation to apply and the current replica state; and it returns the new replica state.
-The \isa{'t::\{linorder\}} constraint in the type signature indicates that timestamps \isa{'t} are instances of \isa{linorder} type class, and they can therefore be compared with the \isa{<} operator defining a linear (or total) order.
+The constraint \isa{'t::\{linorder\}} in the type signature indicates that timestamps \isa{'t} are instances of \isa{linorder} type class, and they can therefore be compared with the \isa{<} operator defining a linear (or total) order.
 This comparison occurs on line 49.
@@
 All theorems stated here have been formally proved and mechanically checked using Isabelle.
-For space reasons we give only a brief sketch of the proofs in this paper; the Isabelle files containing the full details are open source~\cite{SourceFiles}.
+For space reasons we give only a brief sketch of the proofs in this paper; the Isabelle files containing the full details are open source [URL omitted for blind review].
+
+% TODO \cite{SourceFiles}
 
@@
 
-\noindent The \isa{\isacharat} operator concatenates two lists.
+\noindent Here the \isa{\isacharat} operator concatenates two lists.
 Let \isa{t1 = move\_time oper1} and \isa{t2 = move\_time oper2} be the timestamps of the two operations \isa{oper1} and \isa{oper2}.
@@
 \begin{enumerate}
-    \item \isa{t3 < t1 < t2}.\\
+    \item \isa{t3 < t1 < t2}.
         Then neither applying \isa{oper1} nor applying \isa{oper2} will undo \isa{logop}.
         Thus, the two operations are applied in the same way as in the base case.
-    \item \isa{t1 < t3 < t2}. Then\\
-        \isa{apply\_op oper2 (apply\_op oper1 (logop {\isacharhash} ops, tree))}\\first undoes \isa{logop}, then applies \isa{oper1} using \isa{do\_op}, then reapplies \isa{logop} using \isa{redo\_op}, and finally applies \isa{oper2} using \isa{do\_op}.
-        On the other hand,\\\isa{apply\_op oper1 (apply\_op oper2 (logop {\isacharhash} ops, tree))}\\first applies \isa{oper2} using \isa{do\_op}, then undoes it again, then undoes \isa{logop}, applies \isa{oper1} and reapplies \isa{logop} as before, and finally reapplies \isa{oper2} using \isa{redo\_op}.
+    \item \isa{t1 < t3 < t2}.
+        Then \isa{apply\_op oper2 (apply\_op oper1 (logop {\isacharhash} ops, tree))} first undoes \isa{logop}, then applies \isa{oper1} using \isa{do\_op}, then reapplies \isa{logop} using \isa{redo\_op}, and finally applies \isa{oper2} using \isa{do\_op}.
+        On the other hand, \isa{apply\_op oper1 (apply\_op oper2 (logop {\isacharhash} ops, tree))} first applies \isa{oper2} using \isa{do\_op}, then undoes it again, then undoes \isa{logop}, applies \isa{oper1} and reapplies \isa{logop} as before, and finally reapplies \isa{oper2} using \isa{redo\_op}.
         The end result is the same.
-    \item \isa{t1 < t2 < t3}.\\
+    \item \isa{t1 < t2 < t3}.
         Then both applying \isa{oper1} and applying \isa{oper2} first undoes \isa{logop}, yielding \isa{tree' = undo\_op (logop, tree)}.
@@
 the trees are extensionally equivalent, despite having very different in-memory representations.
-We can also prove corollaries of the theorems \isa{apply\_ops\_commutes} and \isa{apply\_ops\_acyclic} for the hash-map-based implementation.
+We can also prove corollaries of \isa{apply\_ops\_commutes} and \isa{apply\_ops\_acyclic} for the hash-map-based implementation.
 %The analogue of \isa{apply\_ops\_unique\_parent} holds trivially since by construction, the hash-map contains at most one (metadata, parent) pair for a given child.
@@
   \label{tab:rpc-times}
-  \vspace{4pt}
   \begin{tabular}{r|rrr}
@@
   \label{fig:plots}
+  \Description{Plot in which the x axis goes from 0 to 600 move operations per second, and the y axis goes from 0 to 2 milliseconds.
+Local operations take a near-constant time around 50 microseconds; the time for remote operations increases linearly from 0.2ms to 1.4ms.}
 \end{figure}
@@
   \label{fig:leader-plot}
+  \Description{Local CRDT operations take about 50 microseconds, and reach saturation at 600 ops/sec. Leader-based operations take about 150--180ms, and reach saturation at 14,000 ops/sec.}
 \end{figure}
@@
 
-Many replicated data systems use optimistic replication \cite{Saito:2005jw}, which allows the state of replicas to temporarily diverge, in order to achieve better performance and availability in the presence of faults than strongly consistent systems~\cite{Bailis:2014th,Gilbert:2002il}.
+Many replicated data systems use optimistic replication~\cite{Saito:2005jw}, which allows the state of replicas to temporarily diverge, in order to achieve better performance and availability in the presence of faults than strongly consistent systems~\cite{Bailis:2014th,Gilbert:2002il}.
 As a consequence, these systems require a mechanism for merging or reconciling conflicting updates that were made concurrently on different replicas.
@@
 
-Our algorithm is an operation-based Conflict-free Replicated Data Type or CRDT~\cite{Shapiro:2011wy,Shapiro:2011un,Burckhardt:2014ft}.
+Our algorithm is an example of an operation-based Conflict-free Replicated Data Type or CRDT~\cite{Shapiro:2011wy,Shapiro:2011un,Burckhardt:2014ft}.
 All CRDTs share the property that concurrent changes on different replicas can be merged in any order; any two replicas that have seen the same set of updates are guaranteed to be in the same state, regardless of the order in which they processed these updates.
@@
 \begin{itemize}
-    \item Martin et al.~\cite{Martin:2010ih} define a CRDT for XML data, and Kleppmann and Beresford~\cite{Kleppmann:2016ve} define a CRDT with a JSON data model.
+    \item Martin et al.~\cite{Martin:2010ih} define a CRDT for XML data, and Kleppmann and Beresford~\cite{Kleppmann:2016ve} define a CRDT for JSON.
         However, these algorithms only deal with insertion and deletion of tree nodes, and do not support moves.
@@
 
-\section*{Acknowledgements}
-
+\begin{acks}
 The authors wish to acknowledge the support of The Boeing Company and the EPSRC ``REMS: Rigorous Engineering for Mainstream Systems'' programme grant (EP/K008528).
-Martin Kleppmann is supported by a Leverhulme Early Career Fellowship and by the Isaac Newton Trust.
+Martin Kleppmann is supported by a Leverhulme Trust Early Career Fellowship and by the Isaac Newton Trust.
 Our evaluation was conducted using AWS credits from the AWS Educate program.
 Thank you to Marc Shapiro for feedback on a draft of this paper.
-
-%\balance
-\bibliographystyle{abbrv}
+\end{acks}
+
+\bibliographystyle{ACM-Reference-Format}
 \bibliography{references}{}

### 2020-03-03 paper/move-op.tex
Reformat in single-column layout

@@
-\documentclass[sigplan,anonymous]{acmart}
-
-% EuroSys recommended formatting
-\renewcommand\footnotetextcopyrightpermission[1]{}
-\pagestyle{plain}
+\documentclass[acmsmall,anonymous,review]{acmart}
+% Recommended options from acmart-pacmpl-template.tex
+\settopmatter{printfolios=true,printccs=false,printacmref=false}
 
@@
 
-\setcopyright{rightsretained}
-\copyrightyear{2019}
-\acmYear{2019}
+% Don't put a few lines of text on a page that is mostly occupied with figures
+\renewcommand{\floatpagefraction}{0.8}
+
+\setcopyright{none}
+\acmJournal{PACMPL}
+\acmVolume{}
+\acmNumber{OOPSLA}
+\acmArticle{}
+\acmYear{2020}
+\acmMonth{1}
 \acmDOI{}
-\acmConference[]{}{}{}
-\acmBooktitle{}
-\acmPrice{}
-\acmISBN{}
+\bibliographystyle{ACM-Reference-Format}
+\citestyle{acmauthoryear}
 
@@
 
-%\renewcommand{\shortauthors}{Kleppmann, Mulligan, Gomes, and Beresford}
-
 \begin{abstract}
@@
 
-\settopmatter{printfolios=true} % Recommended EuroSys formatting
 \maketitle
@@
   %%%
-  \node [fill=red!15] at (9.83,4.10) {(\emph{a})};
-  \node [fill=red!15] at (14.18,4.10) {(\emph{b})};
-  \node [fill=red!15] at (9.84,1.10) {(\emph{c})};
-  \node [fill=red!15] at (14.15,1.10) {(\emph{d})};
-  \node [rectangle,draw] at (10.9,3) {
+  \node [fill=red!15] at (-2.97,-2.4) {(\emph{a})};
+  \node [fill=red!15] at (0.53,-2.4) {(\emph{b})};
+  \node [fill=red!15] at (3.14,-2.4) {(\emph{c})};
+  \node [fill=red!15] at (7.85,-2.4) {(\emph{d})};
+  \node [rectangle,draw] at (-1.9,-3.5) {
       \begin{tikzpicture}[level distance=7mm]
@@
   };
-  \node [rectangle,draw] at (13.5,3) {
+  \node [rectangle,draw] at (1.2,-3.5) {
       \begin{tikzpicture}[level distance=7mm]
@@
   };
-  \node [rectangle,draw] at (10.7,0) {
+  \node [rectangle,draw] at (4,-3.5) {
       \begin{tikzpicture}[level distance=7mm]
@@
   };
-  \node [rectangle,draw] at (13.3,0) {
+  \node [rectangle,draw] at (7,-3.5) {
       \begin{tikzpicture}[level distance=7mm]
@@
   %%%%%
-  \node [rectangle,draw] at (10.95,3) {
+  \node [rectangle,draw] at (-1.5,-3.5) {
       \begin{tikzpicture}[level distance=7mm]
@@
   };
-  \node [rectangle,draw] at (13.8,3) {
+  \node [rectangle,draw] at (1.5,-3.5) {
       \begin{tikzpicture}[level distance=7mm]
       \useasboundingbox (-1.15,-2.3) rectangle (1.15,0.2);
-      \node [fill=red!15] at (0.94,0.03) {(\emph{b})};
+      \node [fill=red!15] at (-0.94,0.03) {(\emph{b})};
       \tikzstyle{level 1}=[sibling distance=12mm]
@@
   };
-  \node [rectangle,draw] (start) at (10.95,0) {
+  \node [rectangle,draw] (start) at (4.5,-3.5) {
       \begin{tikzpicture}[level distance=7mm]
@@
   };
-  \node [rectangle,draw] (right) at (13.8,0) {
+  \node [rectangle,draw] (right) at (7.5,-3.5) {
       \begin{tikzpicture}[level distance=7mm]
       \useasboundingbox (-1.15,-2.3) rectangle (1.15,0.2);
-      \node [fill=red!15] at (0.93,0.03) {(\emph{d})};
+      \node [fill=red!15] at (-0.93,0.03) {(\emph{d})};
       \node {$\mathsf{root}$} child {node {$B$} child {node {$A$} child {node {$C$}}}};
@@
 
-Najafzadeh et al.~\cite{Najafzadeh:2017vk,Najafzadeh:2018bw} previously implemented a replicated filesystem with a move operation, and analysed the case of concurrent move operations introducing a cycle.
+\citet{Najafzadeh:2017vk,Najafzadeh:2018bw} previously implemented a replicated filesystem with a move operation, and analysed the case of concurrent move operations introducing a cycle.
 Using the CISE proof tool~\cite{DBLP:conf/popl/GotsmanYFNS16,Najafzadeh:2016fi} the authors confirm that it is not sufficient for the replica that generates a move operation to check whether the operation introduces a cycle: like in Figure~\ref{fig:move-cycle}, two concurrent operations may be safe individually, but introduce a cycle when combined.
@@
 \ \ \ \ \ \ in\ {\isacharparenleft}op{\isadigit{2}}\ {\isacharhash}\ ops{\isacharcomma}\ tree{\isadigit{2}}{\isacharparenright}{\isacharparenright}{\isacartoucheclose}\isanewline
-\isanewline
+\end{isabellebody}
+\caption{The move operation algorithm, implemented in the Isabelle/HOL language (continued in Figure~\ref{fig:code2}).}
+\label{fig:code1}
+\Description{The move operation algorithm, implemented in the Isabelle/HOL language (continued in Figure~\ref{fig:code2}).}
+\end{figure*}
+
+\begin{figure*}
+\raggedright
+\begin{isabellebody}
+\internallinenumbers
 \isacommand{fun}\isamarkupfalse%
@@
 \isacommand{definition}\isamarkupfalse%
-\ apply{\isacharunderscore}ops\ {\isacharcolon}{\isacharcolon}\ {\isacartoucheopen}{\isacharparenleft}{\isacharprime}t{\isacharcolon}{\isacharcolon}{\isacharbraceleft}linorder{\isacharbraceright}{\isacharcomma}\ {\isacharprime}n{\isacharcomma}\ {\isacharprime}m{\isacharparenright}\ operation\ list\ {\isasymRightarrow}\ {\isacharparenleft}{\isacharprime}t{\isacharcomma}\ {\isacharprime}n{\isacharcomma}\ {\isacharprime}m{\isacharparenright}\ state{\isacartoucheclose}\ \isakeyword{where}\isanewline
-\ \ {\isacartoucheopen}apply{\isacharunderscore}ops\ ops\ {\isasymequiv}\ foldl\ {\isacharparenleft}{\isasymlambda}state\ oper{\isachardot}\ apply{\isacharunderscore}op\ oper\ state{\isacharparenright}\ {\isacharparenleft}{\isacharbrackleft}{\isacharbrackright}{\isacharcomma}\ {\isacharbraceleft}{\isacharbraceright}{\isacharparenright}\ ops{\isacartoucheclose}\isanewline
+\ apply{\isacharunderscore}ops\ {\isacharcolon}{\isacharcolon}\ {\isacartoucheopen}{\isacharparenleft}{\isacharprime}t{\isacharcolon}{\isacharcolon}{\isacharbraceleft}linorder{\isacharbraceright}{\isacharcomma}\ {\isacharprime}n{\isacharcomma}\ {\isacharprime}m{\isacharparenright}\ operation\ list\ {\isasymRightarrow}\ {\isacharparenleft}{\isacharprime}t{\isacharcomma}\ {\isacharprime}n{\isacharcomma}\ {\isacharprime}m{\isacharparenright}\ state{\isacartoucheclose}\isanewline
+\ \ \isakeyword{where}\ {\isacartoucheopen}apply{\isacharunderscore}ops\ ops\ {\isasymequiv}\ foldl\ {\isacharparenleft}{\isasymlambda}state\ oper{\isachardot}\ apply{\isacharunderscore}op\ oper\ state{\isacharparenright}\ {\isacharparenleft}{\isacharbrackleft}{\isacharbrackright}{\isacharcomma}\ {\isacharbraceleft}{\isacharbraceright}{\isacharparenright}\ ops{\isacartoucheclose}\isanewline
 \isanewline
@@
 \ unique{\isacharunderscore}parent\ {\isacharcolon}{\isacharcolon}\ {\isacartoucheopen}{\isacharparenleft}{\isacharprime}n\ {\isasymtimes}\ {\isacharprime}m\ {\isasymtimes}\ {\isacharprime}n{\isacharparenright}\ set\ {\isasymRightarrow}\ bool{\isacartoucheclose}\ \isakeyword{where}\isanewline
-\ \ {\isacartoucheopen}unique{\isacharunderscore}parent\ tree\ {\isasymequiv}\ {\isacharparenleft}{\isasymforall}p{\isadigit{1}}\ p{\isadigit{2}}\ m{\isadigit{1}}\ m{\isadigit{2}}\ c{\isachardot}\ {\isacharparenleft}p{\isadigit{1}}{\isacharcomma}\ m{\isadigit{1}}{\isacharcomma}\ c{\isacharparenright}\ {\isasymin}\ tree\ {\isasymand}\ {\isacharparenleft}p{\isadigit{2}}{\isacharcomma}\ m{\isadigit{2}}{\isacharcomma}\ c{\isacharparenright}\ {\isasymin}\ tree\ {\isasymlongrightarrow}\ p{\isadigit{1}}\ {\isacharequal}\ p{\isadigit{2}}\ {\isasymand}\ m{\isadigit{1}}\ {\isacharequal}\ m{\isadigit{2}}{\isacharparenright}{\isacartoucheclose}\isanewline%
+\ \ {\isacartoucheopen}unique{\isacharunderscore}parent\ tree\ {\isasymequiv}\isanewline
+\ \ \ \ {\isacharparenleft}{\isasymforall}p{\isadigit{1}}\ p{\isadigit{2}}\ m{\isadigit{1}}\ m{\isadigit{2}}\ c{\isachardot}\ {\isacharparenleft}p{\isadigit{1}}{\isacharcomma}\ m{\isadigit{1}}{\isacharcomma}\ c{\isacharparenright}\ {\isasymin}\ tree\ {\isasymand}\ {\isacharparenleft}p{\isadigit{2}}{\isacharcomma}\ m{\isadigit{2}}{\isacharcomma}\ c{\isacharparenright}\ {\isasymin}\ tree\ {\isasymlongrightarrow}\ p{\isadigit{1}}\ {\isacharequal}\ p{\isadigit{2}}\ {\isasymand}\ m{\isadigit{1}}\ {\isacharequal}\ m{\isadigit{2}}{\isacharparenright}{\isacartoucheclose}\isanewline%
 \isanewline
@@
 \end{isabellebody}
-\caption{The move operation algorithm, implemented in the Isabelle/HOL language.}
-\label{fig:code}
-\Description{The move operation algorithm, implemented in the Isabelle/HOL language.}
+\caption{The move operation algorithm (continuation of Figure~\ref{fig:code1}).}
+\label{fig:code2}
+\Description{The move operation algorithm (continuation of Figure~\ref{fig:code1}).}
 \end{figure*}
@@
 
-Figure~\ref{fig:code} gives the full source code for our algorithm in the Isabelle/HOL language~\cite{DBLP:conf/tphol/WenzelPN08}.
+Figures~\ref{fig:code1} and~\ref{fig:code2} give the full source code for our algorithm in the Isabelle/HOL language~\cite{DBLP:conf/tphol/WenzelPN08}.
 We choose this language because it combines the conciseness of pseudocode with the precision of mathematical notation.
@@
 
-In this section we walk through the code in Figure~\ref{fig:code} step by step, explaining the Isabelle/HOL syntax as we encounter it.
+In this section we walk through the code in Figures~\ref{fig:code1} and~\ref{fig:code2} step by step, explaining the Isabelle/HOL syntax as we encounter it.
 Additional background documentation is available~\cite{DBLP:books/sp/NipkowK14}.
@@
 
-Thus, we define one kind of operation: \isa{Move t p m c} (Figure \ref{fig:code}, lines 1--5).
+Thus, we define one kind of operation: \isa{Move t p m c} (Figure \ref{fig:code1}, lines 1--5).
 A move operation is a 4-tuple consisting of a timestamp \isa{t} of type \isa{'t}, a parent node ID \isa{p} of type \isa{'n}, a metadata field \isa{m} of type \isa{'m}, and a child node ID \isa{c} of type \isa{'n}.
@@
 That is, we remove any existing parent-child relationship for \isa{c} from the set \isa{tree}, and then add \isa{\{(p, m, c)\}} to represent the new parent-child relationship.
-This expression appears on lines 31 and 36 of Figure~\ref{fig:code}, as we shall explain shortly.
+This expression appears on lines 31 and 36 of Figure~\ref{fig:code1}, as we shall explain shortly.
 
@@
 
-The \isa{apply\_op} function (lines 43--51) takes two arguments: a \isa{Move} operation to apply and the current replica state; and it returns the new replica state.
+The \isa{apply\_op} function (Figure~\ref{fig:code2}, lines 43--51) takes two arguments: a \isa{Move} operation to apply and the current replica state; and it returns the new replica state.
 The constraint \isa{'t::\{linorder\}} in the type signature indicates that timestamps \isa{'t} are instances of \isa{linorder} type class, and they can therefore be compared with the \isa{<} operator defining a linear (or total) order.
@@
 \smallbreak\noindent\textbf{Log truncation.}
-The algorithm as specified in Figure~\ref{fig:code} retains operations in the log indefinitely, so the memory use grows without bound.
+The algorithm as specified in Figures~\ref{fig:code1} and~\ref{fig:code2} retains operations in the log indefinitely, so the memory use grows without bound.
 However, in practice it is easy to truncate the log, because \isa{apply\_op} only examines the log prefix of operations whose timestamp is greater than that of the operation being applied.
@@
 
-To reason about the state of a replica we first define the function \isa{apply\_ops} on lines 53--54 of Figure~\ref{fig:code}.
+To reason about the state of a replica we first define the function \isa{apply\_ops} on lines 53--54 of Figure~\ref{fig:code2}.
 It takes a list of operations \isa{ops} and returns the state of a replica after it has applied all the operations in \isa{ops}.
@@
 \noindent That is, we consider any list of operations \isa{ops} and define \isa{(log, tree)} to be the replica state after \isa{ops} have been applied.
-We then prove that \isa{unique\_parent tree} holds, where the \isa{unique\_parent} predicate is defined on lines 56--57 of Figure~\ref{fig:code}: whenever the tree contains a triple whose third element is the child node \isa{c}, then the first and second elements of the triple (the parent node and the metadata) are uniquely defined.
+We then prove that \isa{unique\_parent tree} holds, where the \isa{unique\_parent} predicate is defined on lines 56--58 of Figure~\ref{fig:code2}: whenever the tree contains a triple whose third element is the child node \isa{c}, then the first and second elements of the triple (the parent node and the metadata) are uniquely defined.
 
@@
 \end{isabelle}
-\noindent The \isa{acyclic} predicate is defined on lines 59--60 of Figure~\ref{fig:code}, using the \isa{ancestor} relation (the transitive closure of the graph's edges): a graph contains no cycles if no node is an ancestor of itself.
+\noindent The \isa{acyclic} predicate is defined on lines 60--61 of Figure~\ref{fig:code2}, using the \isa{ancestor} relation (the transitive closure of the graph's edges): a graph contains no cycles if no node is an ancestor of itself.
 
@@
 
-We therefore produce variants of the definitions in Figure~\ref{fig:code} that are designed for execution rather than theorem proving.
+We therefore produce variants of the definitions in Figures~\ref{fig:code1} and~\ref{fig:code2} that are designed for execution rather than theorem proving.
 Rather than representing the tree as a set of (parent, meta, child) triples, these definitions use a hash-map in which the keys are child nodes, and the values are (meta, parent) pairs, written in Isabelle/HOL as \isa{('n::\{hashable\}, 'm {\isasymtimes} 'n) hm}.
@@
 \ simulates\ \isakeyword{where}\ {\isacartoucheopen}simulates\ t\ T\ {\isasymequiv}\isanewline
-{\isacharparenleft}{\isasymforall}p\ m\ c{\isachardot}\ hm{\isachardot}lookup\ c\ t\ {\isacharequal}\ Some\ {\isacharparenleft}m{\isacharcomma}\ p{\isacharparenright}\ {\isasymlongleftrightarrow}\ {\isacharparenleft}p{\isacharcomma}\ m{\isacharcomma}\ c{\isacharparenright}\ {\isasymin}\ T{\isacharparenright}{\isacartoucheclose}
+\ \ {\isacharparenleft}{\isasymforall}p\ m\ c{\isachardot}\ hm{\isachardot}lookup\ c\ t\ {\isacharequal}\ Some\ {\isacharparenleft}m{\isacharcomma}\ p{\isacharparenright}\ {\isasymlongleftrightarrow}\ {\isacharparenleft}p{\isacharcomma}\ m{\isacharcomma}\ c{\isacharparenright}\ {\isasymin}\ T{\isacharparenright}{\isacartoucheclose}
 \end{isabelle}
@@
 \begin{figure}
-  \includegraphics[width=\columnwidth,keepaspectratio=true]{plot.pdf}
+  \includegraphics[width=10cm,keepaspectratio=true]{plot.pdf}
   \caption{Median execution time for applying an operation to the replica state, on an Amazon EC2 c5.large instance, at varying operation throughput rates.
@@
 \begin{figure}
-  \includegraphics[width=\columnwidth,keepaspectratio=true]{leader-plot.pdf}
+  \includegraphics[width=10cm,keepaspectratio=true]{leader-plot.pdf}
   \caption{Median time to apply a local CRDT operation compared to the median time to perform a move operation using state machine replication, with a leader located in another region.
@@
 
-Besides the 60 lines of definitions given in Figure~\ref{fig:code}, our Isabelle/HOL formalisation consists of a further 2,166 lines of code.
-Of this, we use 203 lines to prove that every node has a unique parent, 443 lines to prove that the tree contains no cycles, 450 lines to prove that move operations commute and replicas converge, 327 lines to prove the strong eventual consistency of our algorithm using the framework of Gomes et al.~\cite{Gomes:2017gy}, and 743 lines to define the executable variant of our algorithm (\S\ref{subsect.extracting}) and prove its equivalence to the abstract definitions of Figure~\ref{fig:code}.
+Besides the 61 lines of definitions given in Figures~\ref{fig:code1} and~\ref{fig:code2}, our Isabelle/HOL formalisation consists of a further 2,166 lines of code.
+Of this, we use 203 lines to prove that every node has a unique parent, 443 lines to prove that the tree contains no cycles, 450 lines to prove that move operations commute and replicas converge, 327 lines to prove the strong eventual consistency of our algorithm using the framework of Gomes et al.~\cite{Gomes:2017gy}, and 743 lines to define the executable variant of our algorithm (\S\ref{subsect.extracting}) and prove its equivalence to the abstract definitions of Figures~\ref{fig:code1} and~\ref{fig:code2}.
 
@@
 \begin{itemize}
-    \item Martin et al.~\cite{Martin:2010ih} define a CRDT for XML data, and Kleppmann and Beresford~\cite{Kleppmann:2016ve} define a CRDT for JSON.
+    \item \citet{Martin:2010ih} define a CRDT for XML data, and \citet{Kleppmann:2016ve} define a CRDT for JSON.
         However, these algorithms only deal with insertion and deletion of tree nodes, and do not support moves.
         A move operation can be emulated by deleting and re-inserting the moved node, but this approach suffers from the duplication problem demonstrated in \S\ref{sec:move-same-item}.
-    \item As discussed in \S\ref{sec:impossibility}, Najafzadeh et al.~\cite{Najafzadeh:2017vk,Najafzadeh:2018bw} propose two implementations for a replicated filesystem: a CRDT in which conflicting moves are handled by duplicating tree nodes (as in Figures~\ref{fig:move-same-item}b and~\ref{fig:move-cycle}b), and a centralised implementation in which move operations must obtain a lock before executing (not a CRDT since it relies on synchronous coordination).
-    \item Ahmed-Nacer et al.~\cite{AhmedNacer:2012us} outline approaches to handling conflicts on trees, but do not provide algorithms.
-    \item Tao et al.~\cite{Tao:2015gd} propose handling conflicting move operations by allowing the same object to appear in more than one location; thus, their datatype is strictly a DAG, not a tree.
+    \item As discussed in \S\ref{sec:impossibility}, \citet{Najafzadeh:2017vk,Najafzadeh:2018bw} propose two implementations for a replicated filesystem: a CRDT in which conflicting moves are handled by duplicating tree nodes (as in Figures~\ref{fig:move-same-item}b and~\ref{fig:move-cycle}b), and a centralised implementation in which move operations must obtain a lock before executing (not a CRDT since it relies on synchronous coordination).
+    \item \citet{AhmedNacer:2012us} outline approaches to handling conflicts on trees, but do not provide algorithms.
+    \item \citet{Tao:2015gd} propose handling conflicting move operations by allowing the same object to appear in more than one location; thus, their datatype is strictly a DAG, not a tree.
         Some conflicts are handled by duplicating tree nodes.
@@
 
-Molli et al.~\cite{Molli:2003cd} define an OT tree structure with a move operation.
+\citet{Molli:2003cd} define an OT tree structure with a move operation.
 However, it requires that all communication between replicas is performed via total order broadcast, which requires a leader replica or consensus algorithm, like in \S\ref{sec:smr}.
@@
 
-Hughes et al.~\cite{Hughes:2016fp} test Dropbox and Google Drive against a formal specification, but they do not consider moving files, and thus they do not find the issue described in \S\ref{sec:move-cycle}.
-
-Bj{\o}rner~\cite{Bjorner:2007hp} discusses the development of the Distributed File System Replication (DFS-R) component of Windows Server, during which a model checker found an issue with concurrent moves similar to Figure~\ref{fig:move-cycle}a.
+\citet{Hughes:2016fp} test Dropbox and Google Drive against a formal specification, but they do not consider moving files, and thus they do not find the issue described in \S\ref{sec:move-cycle}.
+
+\citet{Bjorner:2007hp} discusses the development of the Distributed File System Replication (DFS-R) component of Windows Server, during which a model checker found an issue with concurrent moves similar to Figure~\ref{fig:move-cycle}a.
 Bj{\o}rner outlines several possible solutions, but notes that model-checking their algorithm was not feasible due to state space explosion.
@@
 Our approach of ordering operations by a timestamp, and undoing/redoing them as necessary so that they take effect in ascending timestamp order, is conceptually a very simple approach---although the details of the implementation are a bit subtle.
-Similar ideas appear in many other systems, including the Bayou database~\cite{Terry:1995dn}, Jefferson's \emph{Time Warp} mechanism~\cite{Jefferson:1985em}, Shapiro et al.'s \emph{capricious total order}~\cite{Shapiro:2016hp}, and Burckhardt's \emph{standard conflict resolution}~\cite[{\S}4.3.3]{Burckhardt:2014hy}.
+Similar ideas appear in many other systems, including the Bayou database~\cite{Terry:1995dn}, \citet{Jefferson:1985em}'s \emph{Time Warp} mechanism, \citet{Shapiro:2016hp}'s \emph{capricious total order}, and Burckhardt's \emph{standard conflict resolution}~\cite[{\S}4.3.3]{Burckhardt:2014hy}.
 The concept of undo-redo is also well known from write-ahead logging~\cite{Mohan:1992fe,Mohan:1992hc}.
@@
 
-\bibliographystyle{ACM-Reference-Format}
 \bibliography{references}{}

### 2020-04-06 paper/list-move.tex
Wording tweak
List move paper: camera-ready

@@
 
-\setcopyright{rightsretained}
 \copyrightyear{2020} 
 \acmYear{2020} 
-\acmDOI{}
-\acmConference[]{}{}{}
-\acmBooktitle{}
-\acmPrice{}
-\acmISBN{}
+\setcopyright{acmlicensed}
+\acmConference[PaPoC '20]{7th Workshop on Principles and Practice of Consistency for Distributed Data}{April 27, 2020}{Heraklion, Greece}
+\acmBooktitle{7th Workshop on Principles and Practice of Consistency for Distributed Data (PaPoC '20), April 27, 2020, Heraklion, Greece}
+\acmPrice{15.00}
+\acmDOI{10.1145/3380787.3393677}
+\acmISBN{978-1-4503-7524-5/20/04}
 
@@
 Conflict-free Replicated Data Types (CRDTs) for lists allow multiple users to concurrently insert and delete elements in a shared list object.
-However, existing algorithms behave poorly when users concurrently move list elements to a new position (i.e.\ reorder the list elements).
+However, existing algorithms behave poorly when users concurrently move list elements to a new position (i.e.\ reorder the elements in the list).
 We demonstrate the need for such a move operation, and describe an algorithm that extends a list CRDT with an explicit move operation.
@@
 This is a surprising omission because moving is a commonly required operation: in many to-do list applications, a user can drag and drop list elements to reorder them, and graphics software allows users to reorder the object list with commands such as ``bring to front'' (which moves an object to the top of the list, so that it occludes other objects) and ``send to back'' (which moves an object to the bottom, so that it is occluded by other objects).
+In the example of Figure~\ref{fig:smiley}, imagine the user first creates the pupils and then the white circles of the eyes; if the software by default places new objects in front of existing objects, the pupils would then not be visible.
+The user thus needs to reorder the object list to bring the pupil objects to the front.
 
@@
 We have outlined an algorithm for single-element moves that can be retrofitted to any existing list CRDT that uses stable position IDs to refer to locations within the list (which is the case for all currently known list CRDTs).
-We have also demonstrated the need for moving a contiguous range of list elements to a new position; finding an algorithm that adds this feature to a CRDTs is at present an open problem.
+We have also demonstrated the need for moving a contiguous range of list elements to a new position; finding an algorithm that adds this feature to a CRDT is at present an open problem.
 We hope that future research will succeed in developing an algorithm for moving ranges of elements.

### 2020-05-04 paper/move-op.tex
Add new references

@@
 
-This approach to determining ordering also easily supports reordering of child nodes within a parent: to move a node to a different position in a list, we use the list CRDT to generate a new ID at the desired position in the sequence.
+This approach to determining ordering also easily supports reordering of child nodes within a parent: to move a node to a different position in a list, we use the list CRDT to generate a new ID at the desired position in the sequence~\cite{Kleppmann:2020jw}.
 Then we perform a move operation in which the parent node is unchanged, but the metadata is changed to this new ID.
@@
 
+After we performed the experiments described in Section~\ref{sec:move-is-hard}, the Dropbox engineering team published a blog post~\cite{Dropbox2020sync} acknowledging the problems of cycles and duplication due to concurrent moves.
+
 \subsection{Totally ordered operation log}

### 2020-05-04 paper/move-op.tex
Change to use lipics style

@@
-\documentclass[acmsmall,anonymous,review]{acmart}
-% Recommended options from acmart-pacmpl-template.tex
-\settopmatter{printfolios=true,printccs=false,printacmref=false}
-
+\documentclass[a4paper,anonymous,USenglish]{lipics-v2019}
+
+\usepackage{booktabs} % \toprule etc.
 \usepackage{tikz}
@@
 
-\setcopyright{none}
-\acmJournal{PACMPL}
-\acmVolume{}
-\acmNumber{OOPSLA}
-\acmArticle{}
-\acmYear{2020}
-\acmMonth{1}
-\acmDOI{}
-\bibliographystyle{ACM-Reference-Format}
-\citestyle{acmauthoryear}
-
 \hyphenation{da-ta-cen-ter da-ta-cen-ters time-stamp time-stamps time-stamped hard-links}
 
+\title{A highly-available move operation for replicated trees}
+
+\author{Martin Kleppmann}{University of Cambridge}{mk428@cl.cam.ac.uk}{https://orcid.org/0000-0001-7252-6958}{Supported by a Leverhulme Trust Early Career Fellowship and by the Isaac Newton Trust.}
+
+\author{Dominic P. Mulligan}{Arm Research, Cambridge}{Dominic.Mulligan@arm.com}{https://orcid.org/0000-0003-4643-3541}{}
+
+\author{Victor B. F. Gomes}{Google\footnote{Work done while at the University of Cambridge}}{vb358@cl.cam.ac.uk}{https://orcid.org/0000-0002-2954-4648}{}
+
+\author{Alastair R. Beresford}{University of Cambridge}{arb33@cl.cam.ac.uk}{https://orcid.org/0000-0003-0818-6535}{}
+
+\authorrunning{M. Kleppmann, D.\,P. Mulligan, V.\,B.\,F. Gomes, and A.\,R. Beresford}
+\Copyright{Martin Kleppmann, Dominic P. Mulligan, Victor B. F. Gomes, and Alastair R. Beresford}
+
+\funding{The authors wish to acknowledge the support of The Boeing Company and the EPSRC ``REMS: Rigorous Engineering for Mainstream Systems'' programme grant (EP/K008528).
+Our evaluation was conducted using AWS credits from the AWS Educate program.}
+\acknowledgements{Thank you to Marc Shapiro for feedback on a draft of this paper.}
+
+\begin{CCSXML}
+<ccs2012>
+   <concept>
+       <concept_id>10003752.10003809.10010172</concept_id>
+       <concept_desc>Theory of computation~Distributed algorithms</concept_desc>
+       <concept_significance>500</concept_significance>
+       </concept>
+   <concept>
+       <concept_id>10011007.10010940.10010992.10010998.10010999</concept_id>
+       <concept_desc>Software and its engineering~Software verification</concept_desc>
+       <concept_significance>500</concept_significance>
+       </concept>
+ </ccs2012>
+\end{CCSXML}
+
+\ccsdesc[500]{Theory of computation~Distributed algorithms}
+\ccsdesc[500]{Software and its engineering~Software verification}
+
+\keywords{Conflict-free Replicated Data Types (CRDTs), distributed filesystems, formal verification}
+
 \begin{document}
-\title{A highly-available move operation for replicated trees}
-
-\author{Martin Kleppmann}
-\email{mk428@cl.cam.ac.uk}
-\orcid{0000-0001-7252-6958}
-\affiliation{%
-  \institution{Department of Computer Science and Technology}
-  \streetaddress{15 JJ Thomson Avenue}
-  \city{Cambridge}
-  \state{}
-  \postcode{CB3 0FD}
-  \country{United Kingdom}
-}
-
-\author{Dominic P.\ Mulligan}
-\email{Dominic.Mulligan@arm.com}
-\orcid{0000-0003-4643-3541}
-\affiliation{%
-  \institution{Arm Research}
-  \streetaddress{110 Fulbourn Road}
-  \city{Cambridge}
-  \state{}
-  \postcode{CB1 9NJ}
-  \country{United Kingdom}
-}
-
-\author{Victor B.\ F.\ Gomes}
-\email{vb358@cl.cam.ac.uk}
-\orcid{0000-0002-2954-4648}
-\affiliation{%
-  \institution{Department of Computer Science and Technology}
-  \streetaddress{15 JJ Thomson Avenue}
-  \city{Cambridge}
-  \state{}
-  \postcode{CB3 0FD}
-  \country{United Kingdom}
-}
-
-\author{Alastair R.\ Beresford}
-\email{arb33@cl.cam.ac.uk}
-\orcid{0000-0003-0818-6535}
-\affiliation{%
-  \institution{Department of Computer Science and Technology}
-  \streetaddress{15 JJ Thomson Avenue}
-  \city{Cambridge}
-  \state{}
-  \postcode{CB3 0FD}
-  \country{United Kingdom}
-}
-
+\maketitle
 \begin{abstract}
@@
 \end{abstract}
-
-\maketitle
 
@@
 \label{fig:code1}
-\Description{The move operation algorithm, implemented in the Isabelle/HOL language (continued in Figure~\ref{fig:code2}).}
 \end{figure*}
@@
 \label{fig:code2}
-\Description{The move operation algorithm (continuation of Figure~\ref{fig:code1}).}
 \end{figure*}
@@
   \label{fig:plots}
-  \Description{Plot in which the x axis goes from 0 to 600 move operations per second, and the y axis goes from 0 to 2 milliseconds.
-Local operations take a near-constant time around 50 microseconds; the time for remote operations increases linearly from 0.2ms to 1.4ms.}
 \end{figure}
@@
   \label{fig:leader-plot}
-  \Description{Local CRDT operations take about 50 microseconds, and reach saturation at 600 ops/sec. Leader-based operations take about 150--180ms, and reach saturation at 14,000 ops/sec.}
 \end{figure}
@@
 
-\begin{acks}
-The authors wish to acknowledge the support of The Boeing Company and the EPSRC ``REMS: Rigorous Engineering for Mainstream Systems'' programme grant (EP/K008528).
-Martin Kleppmann is supported by a Leverhulme Trust Early Career Fellowship and by the Isaac Newton Trust.
-Our evaluation was conducted using AWS credits from the AWS Educate program.
-Thank you to Marc Shapiro for feedback on a draft of this paper.
-\end{acks}
-
+\bibliographystyle{plainurl}
 \bibliography{references}{}

### 2020-05-04 paper/move-op.tex
Page size adjustments

@@
-\documentclass[a4paper,anonymous,USenglish]{lipics-v2019}
-
+\documentclass[10pt,journal,compsoc]{IEEEtran}
+
+\usepackage[nocompress]{cite} % required by IEEE compsoc
 \usepackage{booktabs} % \toprule etc.
+\usepackage{amssymb} % \nexists
+\usepackage{textcomp} % \textmu
+\usepackage{url}
 \usepackage{tikz}
@@
 
+\begin{document}
 \title{A highly-available move operation for replicated trees}
 
-\author{Martin Kleppmann}{University of Cambridge}{mk428@cl.cam.ac.uk}{https://orcid.org/0000-0001-7252-6958}{Supported by a Leverhulme Trust Early Career Fellowship and by the Isaac Newton Trust.}
-
-\author{Dominic P. Mulligan}{Arm Research, Cambridge}{Dominic.Mulligan@arm.com}{https://orcid.org/0000-0003-4643-3541}{}
-
-\author{Victor B. F. Gomes}{Google\footnote{Work done while at the University of Cambridge}}{vb358@cl.cam.ac.uk}{https://orcid.org/0000-0002-2954-4648}{}
-
-\author{Alastair R. Beresford}{University of Cambridge}{arb33@cl.cam.ac.uk}{https://orcid.org/0000-0003-0818-6535}{}
-
-\authorrunning{M. Kleppmann, D.\,P. Mulligan, V.\,B.\,F. Gomes, and A.\,R. Beresford}
-\Copyright{Martin Kleppmann, Dominic P. Mulligan, Victor B. F. Gomes, and Alastair R. Beresford}
-
-\funding{The authors wish to acknowledge the support of The Boeing Company and the EPSRC ``REMS: Rigorous Engineering for Mainstream Systems'' programme grant (EP/K008528).
-Our evaluation was conducted using AWS credits from the AWS Educate program.}
-\acknowledgements{Thank you to Marc Shapiro for feedback on a draft of this paper.}
-
-\begin{CCSXML}
-<ccs2012>
-   <concept>
-       <concept_id>10003752.10003809.10010172</concept_id>
-       <concept_desc>Theory of computation~Distributed algorithms</concept_desc>
-       <concept_significance>500</concept_significance>
-       </concept>
-   <concept>
-       <concept_id>10011007.10010940.10010992.10010998.10010999</concept_id>
-       <concept_desc>Software and its engineering~Software verification</concept_desc>
-       <concept_significance>500</concept_significance>
-       </concept>
- </ccs2012>
-\end{CCSXML}
-
-\ccsdesc[500]{Theory of computation~Distributed algorithms}
-\ccsdesc[500]{Software and its engineering~Software verification}
-
-\keywords{Conflict-free Replicated Data Types (CRDTs), distributed filesystems, formal verification}
-
-\begin{document}
-\maketitle
+\author{Martin Kleppmann, Dominic P. Mulligan, Victor B. F. Gomes, and Alastair R. Beresford
+\IEEEcompsocitemizethanks{\IEEEcompsocthanksitem M. Kleppmann and A.R. Beresford are with the University of Cambridge.\protect\\
+\IEEEcompsocthanksitem D.P. Mulligan is with Arm Research, Cambridge, UK.\protect\\
+\IEEEcompsocthanksitem V.B.F. Gomes is with Google (this work was done while at the University of Cambridge).}}
+
+\IEEEtitleabstractindextext{%
 \begin{abstract}
@@
 
-\section{Introduction}\label{sec:intro}
-
-Many data management applications use a tree-structured or hierarchical data model.
+\begin{IEEEkeywords}
+Conflict-free Replicated Data Types (CRDTs), formal verification, distributed filesystems
+\end{IEEEkeywords}}
+\maketitle
+\IEEEdisplaynontitleabstractindextext
+
+\IEEEraisesectionheading{\section{Introduction}\label{sec:intro}}
+
+\IEEEPARstart{M}{any} data management applications use a tree-structured or hierarchical data model.
 For example, most file systems are trees (directories are branch nodes, and files are leaf nodes), and databases are used to store this directory structure and file metadata~\cite{Niazi:2017uf,Liu:2019zn}.\footnote{Strictly speaking, filesystems with hardlinks form a DAG; we will show in \S\ref{sec:extensions} how to support hardlinks.}
@@
   %%%
-  \node [fill=red!15] at (-2.97,-2.4) {(\emph{a})};
-  \node [fill=red!15] at (0.53,-2.4) {(\emph{b})};
-  \node [fill=red!15] at (3.14,-2.4) {(\emph{c})};
-  \node [fill=red!15] at (7.85,-2.4) {(\emph{d})};
-  \node [rectangle,draw] at (-1.9,-3.5) {
+  \node [fill=red!15] at (9.83,4.10) {(\emph{a})};
+  \node [fill=red!15] at (14.18,4.10) {(\emph{b})};
+  \node [fill=red!15] at (9.84,1.10) {(\emph{c})};
+  \node [fill=red!15] at (14.15,1.10) {(\emph{d})};
+  \node [rectangle,draw] at (10.9,3) {
       \begin{tikzpicture}[level distance=7mm]
@@
   };
-  \node [rectangle,draw] at (1.2,-3.5) {
+  \node [rectangle,draw] at (13.5,3) {
       \begin{tikzpicture}[level distance=7mm]
@@
   };
-  \node [rectangle,draw] at (4,-3.5) {
+  \node [rectangle,draw] at (10.7,0) {
       \begin{tikzpicture}[level distance=7mm]
@@
   };
-  \node [rectangle,draw] at (7,-3.5) {
+  \node [rectangle,draw] at (13.3,0) {
       \begin{tikzpicture}[level distance=7mm]
@@
   %%%%%
-  \node [rectangle,draw] at (-1.5,-3.5) {
+  \node [rectangle,draw] at (10.95,3) {
       \begin{tikzpicture}[level distance=7mm]
@@
   };
-  \node [rectangle,draw] at (1.5,-3.5) {
+  \node [rectangle,draw] at (13.8,3) {
       \begin{tikzpicture}[level distance=7mm]
       \useasboundingbox (-1.15,-2.3) rectangle (1.15,0.2);
-      \node [fill=red!15] at (-0.94,0.03) {(\emph{b})};
+      \node [fill=red!15] at (0.94,0.03) {(\emph{b})};
       \tikzstyle{level 1}=[sibling distance=12mm]
@@
   };
-  \node [rectangle,draw] (start) at (4.5,-3.5) {
+  \node [rectangle,draw] (start) at (10.95,0) {
       \begin{tikzpicture}[level distance=7mm]
@@
   };
-  \node [rectangle,draw] (right) at (7.5,-3.5) {
+  \node [rectangle,draw] (right) at (13.8,0) {
       \begin{tikzpicture}[level distance=7mm]
       \useasboundingbox (-1.15,-2.3) rectangle (1.15,0.2);
-      \node [fill=red!15] at (-0.93,0.03) {(\emph{d})};
+      \node [fill=red!15] at (0.93,0.03) {(\emph{d})};
       \node {$\mathsf{root}$} child {node {$B$} child {node {$A$} child {node {$C$}}}};
@@
 
-\citet{Najafzadeh:2017vk,Najafzadeh:2018bw} previously implemented a replicated filesystem with a move operation, and analysed the case of concurrent move operations introducing a cycle.
+Najafzadeh et al.~\cite{Najafzadeh:2017vk,Najafzadeh:2018bw} previously implemented a replicated filesystem with a move operation, and analysed the case of concurrent move operations introducing a cycle.
 Using the CISE proof tool~\cite{DBLP:conf/popl/GotsmanYFNS16,Najafzadeh:2016fi} the authors confirm that it is not sufficient for the replica that generates a move operation to check whether the operation introduces a cycle: like in Figure~\ref{fig:move-cycle}, two concurrent operations may be safe individually, but introduce a cycle when combined.
@@
 \ \ \ \ \ \ in\ {\isacharparenleft}op{\isadigit{2}}\ {\isacharhash}\ ops{\isacharcomma}\ tree{\isadigit{2}}{\isacharparenright}{\isacharparenright}{\isacartoucheclose}\isanewline
-\end{isabellebody}
-\caption{The move operation algorithm, implemented in the Isabelle/HOL language (continued in Figure~\ref{fig:code2}).}
-\label{fig:code1}
-\end{figure*}
-
-\begin{figure*}
-\raggedright
-\begin{isabellebody}
-\internallinenumbers
+\isanewline
 \isacommand{fun}\isamarkupfalse%
@@
 \isacommand{definition}\isamarkupfalse%
-\ apply{\isacharunderscore}ops\ {\isacharcolon}{\isacharcolon}\ {\isacartoucheopen}{\isacharparenleft}{\isacharprime}t{\isacharcolon}{\isacharcolon}{\isacharbraceleft}linorder{\isacharbraceright}{\isacharcomma}\ {\isacharprime}n{\isacharcomma}\ {\isacharprime}m{\isacharparenright}\ operation\ list\ {\isasymRightarrow}\ {\isacharparenleft}{\isacharprime}t{\isacharcomma}\ {\isacharprime}n{\isacharcomma}\ {\isacharprime}m{\isacharparenright}\ state{\isacartoucheclose}\isanewline
-\ \ \isakeyword{where}\ {\isacartoucheopen}apply{\isacharunderscore}ops\ ops\ {\isasymequiv}\ foldl\ {\isacharparenleft}{\isasymlambda}state\ oper{\isachardot}\ apply{\isacharunderscore}op\ oper\ state{\isacharparenright}\ {\isacharparenleft}{\isacharbrackleft}{\isacharbrackright}{\isacharcomma}\ {\isacharbraceleft}{\isacharbraceright}{\isacharparenright}\ ops{\isacartoucheclose}\isanewline
+\ apply{\isacharunderscore}ops\ {\isacharcolon}{\isacharcolon}\ {\isacartoucheopen}{\isacharparenleft}{\isacharprime}t{\isacharcolon}{\isacharcolon}{\isacharbraceleft}linorder{\isacharbraceright}{\isacharcomma}\ {\isacharprime}n{\isacharcomma}\ {\isacharprime}m{\isacharparenright}\ operation\ list\ {\isasymRightarrow}\ {\isacharparenleft}{\isacharprime}t{\isacharcomma}\ {\isacharprime}n{\isacharcomma}\ {\isacharprime}m{\isacharparenright}\ state{\isacartoucheclose}\ \isakeyword{where}\isanewline
+\ \ {\isacartoucheopen}apply{\isacharunderscore}ops\ ops\ {\isasymequiv}\ foldl\ {\isacharparenleft}{\isasymlambda}state\ oper{\isachardot}\ apply{\isacharunderscore}op\ oper\ state{\isacharparenright}\ {\isacharparenleft}{\isacharbrackleft}{\isacharbrackright}{\isacharcomma}\ {\isacharbraceleft}{\isacharbraceright}{\isacharparenright}\ ops{\isacartoucheclose}\isanewline
 \isanewline
@@
 \end{isabellebody}
-\caption{The move operation algorithm (continuation of Figure~\ref{fig:code1}).}
-\label{fig:code2}
+\caption{The move operation algorithm, implemented in the Isabelle/HOL language.}
+\label{fig:code}
 \end{figure*}
@@
 
-Figures~\ref{fig:code1} and~\ref{fig:code2} give the full source code for our algorithm in the Isabelle/HOL language~\cite{DBLP:conf/tphol/WenzelPN08}.
+Figure~\ref{fig:code} gives the full source code for our algorithm in the Isabelle/HOL language~\cite{DBLP:conf/tphol/WenzelPN08}.
 We choose this language because it combines the conciseness of pseudocode with the precision of mathematical notation.
@@
 
-In this section we walk through the code in Figures~\ref{fig:code1} and~\ref{fig:code2} step by step, explaining the Isabelle/HOL syntax as we encounter it.
+In this section we walk through the code in Figure~\ref{fig:code} step by step, explaining the Isabelle/HOL syntax as we encounter it.
 Additional background documentation is available~\cite{DBLP:books/sp/NipkowK14}.
@@
 
-Thus, we define one kind of operation: \isa{Move t p m c} (Figure \ref{fig:code1}, lines 1--5).
+Thus, we define one kind of operation: \isa{Move t p m c} (Figure \ref{fig:code}, lines 1--5).
 A move operation is a 4-tuple consisting of a timestamp \isa{t} of type \isa{'t}, a parent node ID \isa{p} of type \isa{'n}, a metadata field \isa{m} of type \isa{'m}, and a child node ID \isa{c} of type \isa{'n}.
@@
 That is, we remove any existing parent-child relationship for \isa{c} from the set \isa{tree}, and then add \isa{\{(p, m, c)\}} to represent the new parent-child relationship.
-This expression appears on lines 31 and 36 of Figure~\ref{fig:code1}, as we shall explain shortly.
+This expression appears on lines 31 and 36 of Figure~\ref{fig:code}, as we shall explain shortly.
 
@@
 
-The \isa{apply\_op} function (Figure~\ref{fig:code2}, lines 43--51) takes two arguments: a \isa{Move} operation to apply and the current replica state; and it returns the new replica state.
+The \isa{apply\_op} function (lines 43--51) takes two arguments: a \isa{Move} operation to apply and the current replica state; and it returns the new replica state.
 The constraint \isa{'t::\{linorder\}} in the type signature indicates that timestamps \isa{'t} are instances of \isa{linorder} type class, and they can therefore be compared with the \isa{<} operator defining a linear (or total) order.
@@
 \smallbreak\noindent\textbf{Log truncation.}
-The algorithm as specified in Figures~\ref{fig:code1} and~\ref{fig:code2} retains operations in the log indefinitely, so the memory use grows without bound.
+The algorithm as specified in Figure~\ref{fig:code} retains operations in the log indefinitely, so the memory use grows without bound.
 However, in practice it is easy to truncate the log, because \isa{apply\_op} only examines the log prefix of operations whose timestamp is greater than that of the operation being applied.
@@
 
-To reason about the state of a replica we first define the function \isa{apply\_ops} on lines 53--54 of Figure~\ref{fig:code2}.
+To reason about the state of a replica we first define the function \isa{apply\_ops} on lines 53--54 of Figure~\ref{fig:code}.
 It takes a list of operations \isa{ops} and returns the state of a replica after it has applied all the operations in \isa{ops}.
@@
 \noindent That is, we consider any list of operations \isa{ops} and define \isa{(log, tree)} to be the replica state after \isa{ops} have been applied.
-We then prove that \isa{unique\_parent tree} holds, where the \isa{unique\_parent} predicate is defined on lines 56--58 of Figure~\ref{fig:code2}: whenever the tree contains a triple whose third element is the child node \isa{c}, then the first and second elements of the triple (the parent node and the metadata) are uniquely defined.
+We then prove that \isa{unique\_parent tree} holds, where the \isa{unique\_parent} predicate is defined on lines 56--57 of Figure~\ref{fig:code}: whenever the tree contains a triple whose third element is the child node \isa{c}, then the first and second elements of the triple (the parent node and the metadata) are uniquely defined.
 
@@
 \end{isabelle}
-\noindent The \isa{acyclic} predicate is defined on lines 60--61 of Figure~\ref{fig:code2}, using the \isa{ancestor} relation (the transitive closure of the graph's edges): a graph contains no cycles if no node is an ancestor of itself.
+\noindent The \isa{acyclic} predicate is defined on lines 59--60 of Figure~\ref{fig:code}, using the \isa{ancestor} relation (the transitive closure of the graph's edges): a graph contains no cycles if no node is an ancestor of itself.
 
@@
 
-We therefore produce variants of the definitions in Figures~\ref{fig:code1} and~\ref{fig:code2} that are designed for execution rather than theorem proving.
+We therefore produce variants of the definitions in Figure~\ref{fig:code} that are designed for execution rather than theorem proving.
 Rather than representing the tree as a set of (parent, meta, child) triples, these definitions use a hash-map in which the keys are child nodes, and the values are (meta, parent) pairs, written in Isabelle/HOL as \isa{('n::\{hashable\}, 'm {\isasymtimes} 'n) hm}.
@@
 \begin{figure}
-  \includegraphics[width=10cm,keepaspectratio=true]{plot.pdf}
+  \includegraphics[width=\columnwidth,keepaspectratio=true]{plot.pdf}
   \caption{Median execution time for applying an operation to the replica state, on an Amazon EC2 c5.large instance, at varying operation throughput rates.
@@
 \begin{figure}
-  \includegraphics[width=10cm,keepaspectratio=true]{leader-plot.pdf}
+  \includegraphics[width=\columnwidth,keepaspectratio=true]{leader-plot.pdf}
   \caption{Median time to apply a local CRDT operation compared to the median time to perform a move operation using state machine replication, with a leader located in another region.
@@
 
-Besides the 61 lines of definitions given in Figures~\ref{fig:code1} and~\ref{fig:code2}, our Isabelle/HOL formalisation consists of a further 2,166 lines of code.
-Of this, we use 203 lines to prove that every node has a unique parent, 443 lines to prove that the tree contains no cycles, 450 lines to prove that move operations commute and replicas converge, 327 lines to prove the strong eventual consistency of our algorithm using the framework of Gomes et al.~\cite{Gomes:2017gy}, and 743 lines to define the executable variant of our algorithm (\S\ref{subsect.extracting}) and prove its equivalence to the abstract definitions of Figures~\ref{fig:code1} and~\ref{fig:code2}.
+Besides the 60 lines of definitions given in Figure~\ref{fig:code}, our Isabelle/HOL formalisation consists of a further 2,166 lines of code.
+Of this, we use 203 lines to prove that every node has a unique parent, 443 lines to prove that the tree contains no cycles, 450 lines to prove that move operations commute and replicas converge, 327 lines to prove the strong eventual consistency of our algorithm using the framework of Gomes et al.~\cite{Gomes:2017gy}, and 743 lines to define the executable variant of our algorithm (\S\ref{subsect.extracting}) and prove its equivalence to the abstract definitions of Figure~\ref{fig:code}.
 
@@
 \begin{itemize}
-    \item \citet{Martin:2010ih} define a CRDT for XML data, and \citet{Kleppmann:2016ve} define a CRDT for JSON.
+    \item Martin et al.~\cite{Martin:2010ih} define a CRDT for XML data, and Kleppmann and Beresford~\cite{Kleppmann:2016ve} define a CRDT for JSON.
         However, these algorithms only deal with insertion and deletion of tree nodes, and do not support moves.
         A move operation can be emulated by deleting and re-inserting the moved node, but this approach suffers from the duplication problem demonstrated in \S\ref{sec:move-same-item}.
-    \item As discussed in \S\ref{sec:impossibility}, \citet{Najafzadeh:2017vk,Najafzadeh:2018bw} propose two implementations for a replicated filesystem: a CRDT in which conflicting moves are handled by duplicating tree nodes (as in Figures~\ref{fig:move-same-item}b and~\ref{fig:move-cycle}b), and a centralised implementation in which move operations must obtain a lock before executing (not a CRDT since it relies on synchronous coordination).
-    \item \citet{AhmedNacer:2012us} outline approaches to handling conflicts on trees, but do not provide algorithms.
-    \item \citet{Tao:2015gd} propose handling conflicting move operations by allowing the same object to appear in more than one location; thus, their datatype is strictly a DAG, not a tree.
+    \item As discussed in \S\ref{sec:impossibility}, Najafzadeh et al.~\cite{Najafzadeh:2017vk,Najafzadeh:2018bw} propose two implementations for a replicated filesystem: a CRDT in which conflicting moves are handled by duplicating tree nodes (as in Figures~\ref{fig:move-same-item}b and~\ref{fig:move-cycle}b), and a centralised implementation in which move operations must obtain a lock before executing (not a CRDT since it relies on synchronous coordination).
+    \item Ahmed-Nacer et al.~\cite{AhmedNacer:2012us} outline approaches to handling conflicts on trees, but do not provide algorithms.
+    \item Tao et al.~\cite{Tao:2015gd} propose handling conflicting move operations by allowing the same object to appear in more than one location; thus, their datatype is strictly a DAG, not a tree.
         Some conflicts are handled by duplicating tree nodes.
@@
 
-\citet{Molli:2003cd} define an OT tree structure with a move operation.
+Molli et al.~\cite{Molli:2003cd} define an OT tree structure with a move operation.
 However, it requires that all communication between replicas is performed via total order broadcast, which requires a leader replica or consensus algorithm, like in \S\ref{sec:smr}.
@@
 
-\citet{Hughes:2016fp} test Dropbox and Google Drive against a formal specification, but they do not consider moving files, and thus they do not find the issue described in \S\ref{sec:move-cycle}.
-
-\citet{Bjorner:2007hp} discusses the development of the Distributed File System Replication (DFS-R) component of Windows Server, during which a model checker found an issue with concurrent moves similar to Figure~\ref{fig:move-cycle}a.
+Hughes et al.~\cite{Hughes:2016fp} test Dropbox and Google Drive against a formal specification, but they do not consider moving files, and thus they do not find the issue described in \S\ref{sec:move-cycle}.
+
+Bj{\o}rner~\cite{Bjorner:2007hp} discusses the development of the Distributed File System Replication (DFS-R) component of Windows Server, during which a model checker found an issue with concurrent moves similar to Figure~\ref{fig:move-cycle}a.
 Bj{\o}rner outlines several possible solutions, but notes that model-checking their algorithm was not feasible due to state space explosion.
@@
 Our approach of ordering operations by a timestamp, and undoing/redoing them as necessary so that they take effect in ascending timestamp order, is conceptually a very simple approach---although the details of the implementation are a bit subtle.
-Similar ideas appear in many other systems, including the Bayou database~\cite{Terry:1995dn}, \citet{Jefferson:1985em}'s \emph{Time Warp} mechanism, \citet{Shapiro:2016hp}'s \emph{capricious total order}, and Burckhardt's \emph{standard conflict resolution}~\cite[{\S}4.3.3]{Burckhardt:2014hy}.
+Similar ideas appear in many other systems, including the Bayou database~\cite{Terry:1995dn}, Jefferson's \emph{Time Warp} mechanism~\cite{Jefferson:1985em}, Shapiro et al.'s \emph{capricious total order}~\cite{Shapiro:2016hp}, and Burckhardt's \emph{standard conflict resolution}~\cite[{\S}4.3.3]{Burckhardt:2014hy}.
 The concept of undo-redo is also well known from write-ahead logging~\cite{Mohan:1992fe,Mohan:1992hc}.
@@
 
-\bibliographystyle{plainurl}
+\section*{Acknowledgements}
+
+The authors wish to acknowledge the support of The Boeing Company and the EPSRC ``REMS: Rigorous Engineering for Mainstream Systems'' programme grant (EP/K008528).
+Martin Kleppmann is supported by a Leverhulme Trust Early Career Fellowship and by the Isaac Newton Trust.
+Our evaluation was conducted using AWS credits from the AWS Educate program.
+Thank you to Marc Shapiro for feedback on a draft of this paper.
+
+\bibliographystyle{IEEEtran}
 \bibliography{references}{}

### 2020-05-22 paper/move-op.tex
Reformat Isabelle code to fit

--- 2020-05-04 paper/move-op.tex
@@
 \usepackage{isabelle,isabellesym}
-\isabellestyle{tt}
+\isabellestyle{rm}
 \renewenvironment{isabelle}{%
-  \medbreak\noindent%
+  \medbreak\indent%
   \renewcommand{\isanewline}{\\}%
@@
 \isacommand{fun}\isamarkupfalse%
-\ do{\isacharunderscore}op\ {\isacharcolon}{\isacharcolon}\ {\isacartoucheopen}{\isacharparenleft}{\isacharprime}t{\isacharcomma}\ {\isacharprime}n{\isacharcomma}\ {\isacharprime}m{\isacharparenright}\ operation\ {\isasymtimes}\ {\isacharparenleft}{\isacharprime}n\ {\isasymtimes}\ {\isacharprime}m\ {\isasymtimes}\ {\isacharprime}n{\isacharparenright}\ set\ {\isasymRightarrow}\isanewline
-\ \ \ \ \ \ \ \ \ \ \ \ \ \ {\isacharparenleft}{\isacharprime}t{\isacharcomma}\ {\isacharprime}n{\isacharcomma}\ {\isacharprime}m{\isacharparenright}\ log{\isacharunderscore}op\ {\isasymtimes}\ {\isacharparenleft}{\isacharprime}n\ {\isasymtimes}\ {\isacharprime}m\ {\isasymtimes}\ {\isacharprime}n{\isacharparenright}\ set{\isacartoucheclose}\ \isakeyword{where}\isanewline
+\ do{\isacharunderscore}op\ {\isacharcolon}{\isacharcolon}\ {\isacartoucheopen}{\isacharparenleft}{\isacharprime}t{\isacharcomma}\ {\isacharprime}n{\isacharcomma}\ {\isacharprime}m{\isacharparenright}\ operation\ {\isasymtimes}\ {\isacharparenleft}{\isacharprime}n\ {\isasymtimes}\ {\isacharprime}m\ {\isasymtimes}\ {\isacharprime}n{\isacharparenright}\ set\ {\isasymRightarrow}\ {\isacharparenleft}{\isacharprime}t{\isacharcomma}\ {\isacharprime}n{\isacharcomma}\ {\isacharprime}m{\isacharparenright}\ log{\isacharunderscore}op\ {\isasymtimes}\ {\isacharparenleft}{\isacharprime}n\ {\isasymtimes}\ {\isacharprime}m\ {\isasymtimes}\ {\isacharprime}n{\isacharparenright}\ set{\isacartoucheclose}\ \isakeyword{where}\isanewline
 \ \ {\isacartoucheopen}do{\isacharunderscore}op\ {\isacharparenleft}Move\ t\ newp\ m\ c{\isacharcomma}\ tree{\isacharparenright}\ {\isacharequal}\isanewline
@@
 \isacommand{fun}\isamarkupfalse%
-\ apply{\isacharunderscore}op\ {\isacharcolon}{\isacharcolon}\ {\isacartoucheopen}{\isacharparenleft}{\isacharprime}t{\isacharcolon}{\isacharcolon}{\isacharbraceleft}linorder{\isacharbraceright}{\isacharcomma}\ {\isacharprime}n{\isacharcomma}\ {\isacharprime}m{\isacharparenright}\ operation\ {\isasymRightarrow}\isanewline
-\ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ {\isacharparenleft}{\isacharprime}t{\isacharcomma}\ {\isacharprime}n{\isacharcomma}\ {\isacharprime}m{\isacharparenright}\ state\ {\isasymRightarrow}\ {\isacharparenleft}{\isacharprime}t{\isacharcomma}\ {\isacharprime}n{\isacharcomma}\ {\isacharprime}m{\isacharparenright}\ state{\isacartoucheclose}\ \isakeyword{where}\isanewline
+\ apply{\isacharunderscore}op\ {\isacharcolon}{\isacharcolon}\ {\isacartoucheopen}{\isacharparenleft}{\isacharprime}t{\isacharcolon}{\isacharcolon}{\isacharbraceleft}linorder{\isacharbraceright}{\isacharcomma}\ {\isacharprime}n{\isacharcomma}\ {\isacharprime}m{\isacharparenright}\ operation\ {\isasymRightarrow}\ {\isacharparenleft}{\isacharprime}t{\isacharcomma}\ {\isacharprime}n{\isacharcomma}\ {\isacharprime}m{\isacharparenright}\ state\ {\isasymRightarrow}\ {\isacharparenleft}{\isacharprime}t{\isacharcomma}\ {\isacharprime}n{\isacharcomma}\ {\isacharprime}m{\isacharparenright}\ state{\isacartoucheclose}\ \isakeyword{where}\isanewline
 \ \ {\isacartoucheopen}apply{\isacharunderscore}op\ op{\isadigit{1}}\ {\isacharparenleft}{\isacharbrackleft}{\isacharbrackright}{\isacharcomma}\ tree{\isadigit{1}}{\isacharparenright}\ {\isacharequal}\isanewline
@@
 That is, we remove any existing parent-child relationship for \isa{c} from the set \isa{tree}, and then add \isa{\{(p, m, c)\}} to represent the new parent-child relationship.
-This expression appears on lines 31 and 36 of Figure~\ref{fig:code}, as we shall explain shortly.
+This expression appears on lines 30 and 35 of Figure~\ref{fig:code}, as we shall explain shortly.
 
@@
 
-The \isa{do\_op} function (lines 26--31) now performs the actual work of applying a move operation.
+The \isa{do\_op} function (lines 26--30) now performs the actual work of applying a move operation.
 This function takes as argument a pair consisting of a \isa{Move} operation and the current tree, and it returns a pair consisting of a \isa{LogMove} operation (which will be added to the log) and an updated tree.
-In line 29, the \isa{LogMove} record is constructed as described in \S\ref{sec:state-log}, obtaining the prior parent and metadata of \isa{c} using the \isa{get\_parent} function.
-
-Line 30 performs the check that ensures no cycles are introduced: if \isa{ancestor tree c newp}, i.e.\ if the node \isa{c} is being moved, and \isa{c} is an ancestor of the new parent \isa{newp}, then the tree is returned unmodified---in other words, the operation is ignored.
+In line 28, the \isa{LogMove} record is constructed as described in \S\ref{sec:state-log}, obtaining the prior parent and metadata of \isa{c} using the \isa{get\_parent} function.
+
+Line 29 performs the check that ensures no cycles are introduced: if \isa{ancestor tree c newp}, i.e.\ if the node \isa{c} is being moved, and \isa{c} is an ancestor of the new parent \isa{newp}, then the tree is returned unmodified---in other words, the operation is ignored.
 Similarly, the operation is also ignored if \isa{c = newp}.
-Otherwise (line 31), the tree is updated by removing \isa{c} from its existing parent, if any, and adding the new parent-child relationship \isa{(newp, m, c)} to the tree.
+Otherwise (line 30), the tree is updated by removing \isa{c} from its existing parent, if any, and adding the new parent-child relationship \isa{(newp, m, c)} to the tree.
 
@@
 However, in an optimistic replication setting, each replica may apply the operations in a different order, and we need to ensure that the replica state nevertheless converges towards a consistent state.
-This goal is accomplished by the \isa{undo\_op}, \isa{redo\_op}, and \isa{apply\_op} functions (lines 33--51).
+This goal is accomplished by the \isa{undo\_op}, \isa{redo\_op}, and \isa{apply\_op} functions (lines 32--49).
 
@@
 
-The \isa{apply\_op} function (lines 43--51) takes two arguments: a \isa{Move} operation to apply and the current replica state; and it returns the new replica state.
-The constraint \isa{'t::\{linorder\}} in the type signature indicates that timestamps \isa{'t} are instances of \isa{linorder} type class, and they can therefore be compared with the \isa{<} operator defining a linear (or total) order.
-This comparison occurs on line 49.
+The \isa{apply\_op} function (lines 42--49) takes two arguments: a \isa{Move} operation to apply and the current replica state; and it returns the new replica state.
+The constraint \isa{'t::\{linorder\}} in the type signature indicates that timestamps \isa{'t} are instances of \isa{linorder} type class, and they can therefore be compared with the $<$ operator defining a linear (or total) order.
+This comparison occurs on line 47.
 
 Recall that the replica state includes the operation log (line 14), and we use this log to perform the undo-do-redo cycle.
-Lines 45--47 handle the case where the log is empty: in this case, we simply perform the operation using \isa{do\_op}, and return the new tree along with a log containing a single \isa{LogMove} record.
-If the log is nonempty (line 48), we take \isa{logop} to be the first element of the log, and \isa{ops} to be the rest.
+Lines 43--45 handle the case where the log is empty: in this case, we simply perform the operation using \isa{do\_op}, and return the new tree along with a log containing a single \isa{LogMove} record.
+If the log is nonempty (line 46), we take \isa{logop} to be the first element of the log, and \isa{ops} to be the rest.
 (The hash character in \isa{logop {\isacharhash} ops} is the \emph{list cons} operator that adds one element to the head of a list.)
-If the timestamp of \isa{logop} is greater than the timestamp of the new operation (line 49) we first undo \isa{logop} with \isa{undo\_op}, then recursively apply the new operation to the remaining log, and finally reapply \isa{logop} with \isa{redo\_op} (line 50).
-Otherwise we perform the operation using \isa{do\_op}, and add the corresponding \isa{LogMove} record as the head of the log (line 51).
+If the timestamp of \isa{logop} is greater than the timestamp of the new operation (line 47) we first undo \isa{logop} with \isa{undo\_op}, then recursively apply the new operation to the remaining log, and finally reapply \isa{logop} with \isa{redo\_op} (line 48).
+Otherwise we perform the operation using \isa{do\_op}, and add the corresponding \isa{LogMove} record as the head of the log (line 49).
 
 This logic ensures that the log is maintained in descending timestamp order, with the greatest timestamp at the head.
-\isa{undo\_op} (lines 33--36) inverts the effect of a previous move operation by restoring the prior parent and metadata that were recorded in the \isa{LogMove}'s additional field.
-\isa{redo\_op} (lines 38--41) uses \isa{do\_op} to perform an operation again and recomputes the \isa{LogMove} record (which might have changed due to the effect of the new operation).
+\isa{undo\_op} (lines 32--35) inverts the effect of a previous move operation by restoring the prior parent and metadata that were recorded in the \isa{LogMove}'s additional field.
+\isa{redo\_op} (lines 37--40) uses \isa{do\_op} to perform an operation again and recomputes the \isa{LogMove} record (which might have changed due to the effect of the new operation).
 
@@
 
-To reason about the state of a replica we first define the function \isa{apply\_ops} on lines 53--54 of Figure~\ref{fig:code}.
+To reason about the state of a replica we first define the function \isa{apply\_ops} on lines 51--52 of Figure~\ref{fig:code}.
 It takes a list of operations \isa{ops} and returns the state of a replica after it has applied all the operations in \isa{ops}.
@@
 \noindent That is, we consider any list of operations \isa{ops} and define \isa{(log, tree)} to be the replica state after \isa{ops} have been applied.
-We then prove that \isa{unique\_parent tree} holds, where the \isa{unique\_parent} predicate is defined on lines 56--57 of Figure~\ref{fig:code}: whenever the tree contains a triple whose third element is the child node \isa{c}, then the first and second elements of the triple (the parent node and the metadata) are uniquely defined.
+We then prove that \isa{unique\_parent tree} holds, where the \isa{unique\_parent} predicate is defined on lines 54--56 of Figure~\ref{fig:code}: whenever the tree contains a triple whose third element is the child node \isa{c}, then the first and second elements of the triple (the parent node and the metadata) are uniquely defined.
 
@@
 \end{isabelle}
-\noindent The \isa{acyclic} predicate is defined on lines 59--60 of Figure~\ref{fig:code}, using the \isa{ancestor} relation (the transitive closure of the graph's edges): a graph contains no cycles if no node is an ancestor of itself.
+\noindent The \isa{acyclic} predicate is defined on lines 58--59 of Figure~\ref{fig:code}, using the \isa{ancestor} relation (the transitive closure of the graph's edges): a graph contains no cycles if no node is an ancestor of itself.
 
@@
 Assume to the contrary that \isa{tree2} contains a cycle.
-If \isa{do\_op} left the tree unchanged (line 30), we immediately have a contradiction.
+If \isa{do\_op} left the tree unchanged (line 29), we immediately have a contradiction.
 Otherwise we can assume \isa{{\isasymnot}(ancestor tree c newp {\isasymor} c = newp)} and we have:
@@
 Moreover, since only one edge from \isa{newp} to \isa{c} was added, \isa{tree1} must contain a path from \isa{c} to \isa{newp} (which is then closed by the new edge to form a cycle), i.e.\ \isa{c} must be an ancestor of \isa{newp} in \isa{tree1}.
-However, in that case, the predicate \isa{ancestor tree1 c newp} on line 30 would have been true, which contradicts the earlier assumption.
+However, in that case, the predicate \isa{ancestor tree1 c newp} on line 29 would have been true, which contradicts the earlier assumption.
 Hence, the \isa{do\_op} function preserves the acyclicity invariant of the tree.
@@
 Let \isa{t1 = move\_time oper1} and \isa{t2 = move\_time oper2} be the timestamps of the two operations \isa{oper1} and \isa{oper2}.
-Assume \isa{t1 < t2} without loss of generality.
+Assume $\isa{t1} < \isa{t2}$ without loss of generality.
 We can now prove lemma \isa{apply\_op\_commute2} by induction over \isa{log}.
@@
 \begin{enumerate}
-    \item \isa{t3 < t1 < t2}.
+    \item $\isa{t3} < \isa{t1} < \isa{t2}$.
         Then neither applying \isa{oper1} nor applying \isa{oper2} will undo \isa{logop}.
         Thus, the two operations are applied in the same way as in the base case.
-    \item \isa{t1 < t3 < t2}.
+    \item $\isa{t1} < \isa{t3} < \isa{t2}$.
         Then \isa{apply\_op oper2 (apply\_op oper1 (logop {\isacharhash} ops, tree))} first undoes \isa{logop}, then applies \isa{oper1} using \isa{do\_op}, then reapplies \isa{logop} using \isa{redo\_op}, and finally applies \isa{oper2} using \isa{do\_op}.
@@
         The end result is the same.
-    \item \isa{t1 < t2 < t3}.
+    \item $\isa{t1} < \isa{t2} < \isa{t3}$.
         Then both applying \isa{oper1} and applying \isa{oper2} first undoes \isa{logop}, yielding \isa{tree' = undo\_op (logop, tree)}.
@@
 
-Besides the 60 lines of definitions given in Figure~\ref{fig:code}, our Isabelle/HOL formalisation consists of a further 2,166 lines of code.
+Besides the 59 lines of definitions given in Figure~\ref{fig:code}, our Isabelle/HOL formalisation consists of a further 2,166 lines of code.
 Of this, we use 203 lines to prove that every node has a unique parent, 443 lines to prove that the tree contains no cycles, 450 lines to prove that move operations commute and replicas converge, 327 lines to prove the strong eventual consistency of our algorithm using the framework of Gomes et al.~\cite{Gomes:2017gy}, and 743 lines to define the executable variant of our algorithm (\S\ref{subsect.extracting}) and prove its equivalence to the abstract definitions of Figure~\ref{fig:code}.

### 2020-05-22 paper/move-op.tex
Rewording and updates

@@
 \begin{document}
-\title{A highly-available move operation for replicated trees}
+\title{A highly-available move operation for replicated trees and distributed filesystems}
+
+% IEEE transactions review is single-blind by default, and double-blind only on special request
+% https://www.computer.org/publications/author-resources/peer-review/journals
 
@@
 \begin{abstract}
-    Replicated tree data structures are data models supporting distributed filesystems, such as Google Drive and Dropbox, and XML or JSON databases.
+    Replicated tree data structures are a fundamental building block of distributed filesystems, such as Google Drive and Dropbox, and collaborative applications with a JSON or XML data model.
     These systems need to support a \emph{move} operation that allows a subtree to be moved to a new location within the tree.
     However, such a move operation is difficult to implement correctly if different replicas can concurrently perform arbitrary move operations, and we demonstrate bugs in Google Drive and Dropbox that arise with concurrent moves.
-    In this paper we present an algorithm that handles arbitrary concurrent modifications on trees, while ensuring that the tree structure remains valid (in particular, no cycles are introduced), and guaranteeing that all replicas converge towards the same consistent state.
+    In this paper we present a CRDT algorithm that handles arbitrary concurrent modifications on trees, while ensuring that the tree structure remains valid (in particular, no cycles are introduced), and guaranteeing that all replicas converge towards the same consistent state.
     Our algorithm requires no synchronous coordination between replicas, making it highly available in the face of network partitions.
@@
 \begin{IEEEkeywords}
-Conflict-free Replicated Data Types (CRDTs), formal verification, distributed filesystems
+Conflict-free Replicated Data Types (CRDTs), formal verification, distributed filesystems, distributed collaboration
 \end{IEEEkeywords}}
@@
 
-\IEEEPARstart{M}{any} data management applications use a tree-structured or hierarchical data model.
-For example, most file systems are trees (directories are branch nodes, and files are leaf nodes), and databases are used to store this directory structure and file metadata~\cite{Niazi:2017uf,Liu:2019zn}.\footnote{Strictly speaking, filesystems with hardlinks form a DAG; we will show in \S\ref{sec:extensions} how to support hardlinks.}
-Moreover, XML databases and JSON document databases also operate on trees, which might represent rich text (a tree of paragraphs, lists, figures, sections, etc.), vector graphics, CAD drawings, and many other types of data.
-Typically, a graphical user interface allows a user to edit this information interactively, resulting in updates to the underlying XML/JSON database: adding or deleting nodes in the tree, and moving nodes from one position in the tree to another.
-
-When updating these trees, standard concurrency control techniques can be used, such as two-phase locking~\cite{Bernstein:1987va} or optimistic concurrency control~\cite{Cahill:2008eg}.
-However, these techniques require synchronous coordination between replicas.
-If we want the system to continue processing read and write requests even in the presence of arbitrary network partitions (in other words, if we want \emph{availability} and \emph{partition-tolerance} in the sense of the CAP theorem~\cite{Gilbert:2002il}), then we must avoid such coordination~\cite{Davidson:1985hv}.
-Coordination avoidance has shown to be beneficial in several data systems~\cite{Bailis:2014th,AnnaKVS}, and network partitions are common~\cite{Alquraan:2018wj,Bailis:2014jx}.
+\IEEEPARstart{M}{any} applications use a tree-structured data model.
+For example, most file systems are trees: directories are branch nodes, and files are leaf nodes.
+Moreover, XML and JSON documents are trees, and they are used in many applications to represent e.g.\ rich text (a tree of paragraphs, lists, figures, sections, etc.), vector graphics, CAD drawings, and many other types of data.
+Typically, a graphical user interface allows a user to edit this information interactively, resulting in updates to the underlying XML/JSON structure: adding or deleting nodes in the tree, and moving nodes from one position in the tree to another.
+
+In distributed filesystems and collaborative multi-user software, this tree is replicated across multiple nodes.
+If users attempt to update the tree concurrently on different replicas, concurrency control is required.
+However, standard techniques such as two-phase locking require synchronous coordination between replicas.
+If we want the system to continue processing read and write requests even in the presence of arbitrary network partitions (in other words, if we want \emph{availability} and \emph{partition-tolerance} in the sense of the CAP theorem~\cite{Gilbert:2002il}), then we must avoid such coordination.
+This is an important practical consideration since network partitions are common in many networks~\cite{Alquraan:2018wj,Bailis:2014jx}.
 
@@
 
-Dropbox and Google Drive are widely-deployed examples of optimistically replicated filesystems: they run a daemon on the user's machine that watches a designated directory for changes, and locally maintains a database with the user's tree of files and directories.
+Dropbox and Google Drive are widely-deployed examples of optimistically replicated filesystems: they run a daemon on the user's machine that watches a designated directory for changes.
 The user can read and arbitrarily modify the files on their local disk, even while their computer is offline.
@@
 
-In this paper we introduce a novel algorithm for handling concurrent updates to a replicated tree, such as a filesystem or XML database.
-It allows replicas to manipulate the tree by creating new nodes, deleting nodes, or moving subtrees to a new location within the tree.
-In order to rule out bugs like those in Google Drive we formally prove the correctness of our algorithm using the Isabelle/HOL proof assistant.
+In this paper we introduce a novel algorithm for handling concurrent updates to a replicated tree, such as an XML document or filesystem.
+It allows replicas to manipulate the tree by creating nodes, deleting nodes, or moving subtrees to a new location within the tree.
+We rule out bugs like those in Google Drive by formally proving our algorithm correct using the Isabelle/HOL proof assistant.
 
 Our algorithm supports optimistic replication, allowing replicas to temporarily diverge as they are updated, and ensuring that they always converge towards a consistent state.
-It is an example of a \emph{Conflict-free Replicated Data Type} or CRDT~\cite{Shapiro:2011wy,Shapiro:2011un}, and it guarantees a consistency model called \emph{strong eventual consistency}~\cite{Shapiro:2011un,Gomes:2017gy}.
-Our contributions in this paper are:
+It is an example of a \emph{Conflict-free Replicated Data Type} or CRDT~\cite{Shapiro:2011un}, and it guarantees a consistency model called \emph{strong eventual consistency}~\cite{Shapiro:2011un,Gomes:2017gy}.
+Our contributions are:
 \begin{itemize}
-    \item We define a Conflict-free Replicated Data Type for trees that supports a \emph{move} operation, without requiring any coordination between replicas such as locking or consensus.
+    \item We define a Conflict-free Replicated Data Type for trees that allow \emph{move} operations without any coordination between replicas such as locking or consensus.
         As discussed in \S\ref{sec:impossibility}, this has previously been thought to be impossible to achieve \cite{Najafzadeh:2017vk,Najafzadeh:2018bw}.
     \item We formalise the algorithm using Isabelle/HOL \cite{DBLP:conf/tphol/WenzelPN08}, a proof assistant based on higher-order logic, and obtain a computer-checked proof of correctness.
-        In particular, we prove that arbitrary concurrent modifications to the tree can be merged such that all replicas converge to a consistent state (strong eventual consistency) while keeping the data in a tree structure.
+        In particular, we prove that arbitrary concurrent modifications to the tree can be merged such that all replicas converge to a consistent state, while preserving the tree structure.
+        Our proof technique can also be applied to other distributed systems, making it of independent interest.
     \item To demonstrate the practical viability of our approach, we refine the algorithm to an executable implementation within Isabelle/HOL and prove the equivalence of the two.
-        We then extract a formally verified Scala implementation from Isabelle and evaluate its performance with replicas across three continents.
+        We extract a formally verified Scala implementation from Isabelle and evaluate its performance with replicas across three continents.
     \item We perform experiments with Dropbox and Google Drive, and show that they exhibit problems that would be prevented by our algorithm.
@@
   \node [rectangle,draw,inner sep=3mm] (merge2) at (8.2,0) {?};
-  \draw [time] (start1.east) -- node [above,text width=2.5cm,text centered,inner ysep=5pt] {Move $A$ to be a child of $B$} (change1.west);
+  \draw [time] (start1.east) -- node [above,text width=2.5cm,text centered,inner ysep=5pt,font=\footnotesize] {Move $A$ to be a child of $B$} (change1.west);
   \draw [time] (change1.east) -- (merge1.west);
-  \draw [time] (start2.east) -- node [above,text width=2.5cm,text centered,inner ysep=5pt] {Move $A$ to be a child of $C$} (change2.west);
+  \draw [time] (start2.east) -- node [above,text width=2.5cm,text centered,inner ysep=5pt,font=\footnotesize] {Move $A$ to be a child of $C$} (change2.west);
   \draw [time] (change2.east) -- (merge2.west);
@@
   \draw [network] (6.5,3) to [out=270,in=90] (7.2,0);
-  \node [rotate=90,blue,font=\footnotesize] at (7.5,1.5) {network communication};
+  \node [rotate=90,blue,font=\footnotesize,align=center] at (7.5,1.5) {network\\communication};
   \path [draw,dotted] (-3.1,1.5) -- (8.7,1.5);
@@
   \node [rectangle,draw,inner sep=3mm] (merge2) at (8,0) {?};
-  \draw [time] (start1.east) -- node [above,text width=2.5cm,text centered,inner ysep=5pt] {Move $B$ to be a child of $A$} (change1.west);
+  \draw [time] (start1.east) -- node [above,text width=2.5cm,text centered,inner ysep=5pt,font=\footnotesize] {Move $B$ to be a child of $A$} (change1.west);
   \draw [time] (change1.east) -- (merge1.west);
-  \draw [time] (start2.east) -- node [above,text width=2.5cm,text centered,inner ysep=5pt] {Move $A$ to be a child of $B$} (change2.west);
+  \draw [time] (start2.east) -- node [above,text width=2.5cm,text centered,inner ysep=5pt,font=\footnotesize] {Move $A$ to be a child of $B$} (change2.west);
   \draw [time] (change2.east) -- (merge2.west);
@@
   \draw [network] (6.0,3) to [out=270,in=90] (6.7,0);
-  \node [rotate=90,blue,font=\footnotesize] at (7.0,1.5) {network communication};
+  \node [rotate=90,blue,font=\footnotesize,align=center] at (7.0,1.5) {network\\communication};
   \path [draw,dotted] (-2.5,1.6) -- (8.5,1.6);
@@
         In the XML tree this corresponds to creating new list and bullet point nodes, and then moving the paragraph node inside the bullet point.
-    \item In presentation software, grouping two graphical objects corresponds to creating a new group node, and then moving the two objects into the new group node.
+    \item In graphics software, grouping two objects corresponds to creating a new group node, and then moving the two objects into the new group node.
 \end{itemize}
@@
 Two users who believe they are collaborating on the same file may in fact be editing two different copies, which will then become inconsistent with each other.
-In the rich text editor and presentation software examples, such duplication is also undesirable.
+In the rich text editor and graphics software examples, such duplication is also undesirable.
 
@@
 The ``winning'' location could be picked based on a timestamp in the operations, similarly to the ``last writer wins'' conflict resolution method of Thomas's write rule~\cite{Johnson:1975we}.
-(The \emph{timestamp} in this context need not come from a physical clock; it could also be logical, such as a Lamport timestamp~\cite{Lamport:1978jq}.)
+The \emph{timestamp} in this context need not come from a physical clock; it could also be logical, such as a Lamport timestamp~\cite{Lamport:1978jq}.
 
@@
 Here, replica 1 moves $B$ to be a child of $A$, while concurrently replica 2 moves $A$ to be a child of $B$.
-As each replica propagates its operation to the other replica, a careless implementation might end up as shown in Figure~\ref{fig:move-cycle}a, in which $A$ and $B$ have formed a cycle, detached from the tree.
+As each replica propagates its operation to the other replica, a careless implementation might end up in the state shown in Figure~\ref{fig:move-cycle}a: $A$ and $B$ have formed a cycle, detached from the tree.
 
@@
 
-In this section we walk through the code in Figure~\ref{fig:code} step by step, explaining the Isabelle/HOL syntax as we encounter it.
+In this section we walk through the code step by step, explaining the Isabelle/HOL syntax as we encounter it.
 Additional background documentation is available~\cite{DBLP:books/sp/NipkowK14}.
@@
 The metadata field \isa{m} in a move operation allows additional information to be associated with the parent-child relationship of \isa{p} and \isa{c}.
-For example, in a filesystem, the parent and child would be the inodes of a directory and a file within it, respectively, and the metadata would contain the filename of the child.
-Thus, a file with inode \isa{c} can be renamed by performing a \isa{Move t p m c}, where the new parent directory \isa{p} is the unchanged inode of the existing parent directory, but the metadata \isa{m} contains the new filename.
-
-When users want to make changes to the tree on their local replica, they generate new \isa{Move t p m c} operations for these changes, and apply these operations to their local replica using the algorithm described in the rest of this section.
+For example, in a filesystem, the parent and child are the inodes of a directory and a file within it, respectively, and the metadata contains the filename of the child.
+Thus, a file with inode \isa{c} can be renamed by performing a \isa{Move t p m c}, where the new parent directory \isa{p} is the inode of the existing parent (unchanged), but the metadata \isa{m} contains the new filename.
+
+When users want to make changes to the tree on their local replica, they generate new \isa{Move t p m c} operations for these changes, and apply these operations using the algorithm described in the rest of this section.
 
@@
 
-Our algorithm is an example of an operation-based Conflict-free Replicated Data Type or CRDT~\cite{Shapiro:2011wy,Shapiro:2011un,Burckhardt:2014ft}.
+Our algorithm is an example of an operation-based Conflict-free Replicated Data Type or CRDT~\cite{Shapiro:2011un,Burckhardt:2014ft}.
 All CRDTs share the property that concurrent changes on different replicas can be merged in any order; any two replicas that have seen the same set of updates are guaranteed to be in the same state, regardless of the order in which they processed these updates.
@@
 Similar ideas appear in many other systems, including the Bayou database~\cite{Terry:1995dn}, Jefferson's \emph{Time Warp} mechanism~\cite{Jefferson:1985em}, Shapiro et al.'s \emph{capricious total order}~\cite{Shapiro:2016hp}, and Burckhardt's \emph{standard conflict resolution}~\cite[{\S}4.3.3]{Burckhardt:2014hy}.
-The concept of undo-redo is also well known from write-ahead logging~\cite{Mohan:1992fe,Mohan:1992hc}.
+The concept of undo-redo is also well known from write-ahead logging~\cite{Mohan:1992fe}.
 

### 2020-05-22 paper/move-op.tex
Use italics for Isabelle code
Highlight "if/then/else" and "let/in" as keywords, and constructors in
roman font

@@
 \usepackage{isabelle,isabellesym}
-\isabellestyle{rm}
+\isabellestyle{it}
 \renewenvironment{isabelle}{%
@@
 \ {\isacharparenleft}{\isacharprime}t{\isacharcomma}\ {\isacharprime}n{\isacharcomma}\ {\isacharprime}m{\isacharparenright}\ operation\isanewline
-\ \ {\isacharequal}\ Move\ {\isacharparenleft}move{\isacharunderscore}time{\isacharcolon}\ {\isacharprime}t{\isacharparenright}\isanewline
+\ \ {\isacharequal}\ {\rm Move}\ {\isacharparenleft}move{\isacharunderscore}time{\isacharcolon}\ {\isacharprime}t{\isacharparenright}\isanewline
 \ \ \ \ \ \ \ \ \ {\isacharparenleft}move{\isacharunderscore}parent{\isacharcolon}\ {\isacharprime}n{\isacharparenright}\isanewline
@@
 \ {\isacharparenleft}{\isacharprime}t{\isacharcomma}\ {\isacharprime}n{\isacharcomma}\ {\isacharprime}m{\isacharparenright}\ log{\isacharunderscore}op\isanewline
-\ \ {\isacharequal}\ LogMove\ {\isacharparenleft}log{\isacharunderscore}time{\isacharcolon}\ {\isacharprime}t{\isacharparenright}\isanewline
+\ \ {\isacharequal}\ {\rm LogMove}\ {\isacharparenleft}log{\isacharunderscore}time{\isacharcolon}\ {\isacharprime}t{\isacharparenright}\isanewline
 \ \ \ \ \ \ \ \ \ \ \ \ {\isacharparenleft}old{\isacharunderscore}parent{\isacharcolon}\ {\isacartoucheopen}{\isacharparenleft}{\isacharprime}n\ {\isasymtimes}\ {\isacharprime}m{\isacharparenright}\ option{\isacartoucheclose}{\isacharparenright}\isanewline
@@
 \ \ {\isacartoucheopen}get{\isacharunderscore}parent\ tree\ child\ {\isasymequiv}\isanewline
-\ \ \ \ \ if\ {\isasymexists}{\isacharbang}parent{\isachardot}\ {\isasymexists}{\isacharbang}meta{\isachardot}\ {\isacharparenleft}parent{\isacharcomma}\ meta{\isacharcomma}\ child{\isacharparenright}\ {\isasymin}\ tree\ then\isanewline
-\ \ \ \ \ \ \ Some\ {\isacharparenleft}THE\ {\isacharparenleft}parent{\isacharcomma}\ meta{\isacharparenright}{\isachardot}\ {\isacharparenleft}parent{\isacharcomma}\ meta{\isacharcomma}\ child{\isacharparenright}\ {\isasymin}\ tree{\isacharparenright}\isanewline
-\ \ \ \ \ else\ None{\isacartoucheclose}\isanewline
+\ \ \ \ \ \isacommand{if}\ {\isasymexists}{\isacharbang}parent{\isachardot}\ {\isasymexists}{\isacharbang}meta{\isachardot}\ {\isacharparenleft}parent{\isacharcomma}\ meta{\isacharcomma}\ child{\isacharparenright}\ {\isasymin}\ tree\ \isacommand{then}\isanewline
+\ \ \ \ \ \ \ {\rm Some}\ {\isacharparenleft}{\rm THE}\ {\isacharparenleft}parent{\isacharcomma}\ meta{\isacharparenright}{\isachardot}\ {\isacharparenleft}parent{\isacharcomma}\ meta{\isacharcomma}\ child{\isacharparenright}\ {\isasymin}\ tree{\isacharparenright}\isanewline
+\ \ \ \ \ \isacommand{else}\ {\rm None}{\isacartoucheclose}\isanewline
 \isanewline
@@
 \ do{\isacharunderscore}op\ {\isacharcolon}{\isacharcolon}\ {\isacartoucheopen}{\isacharparenleft}{\isacharprime}t{\isacharcomma}\ {\isacharprime}n{\isacharcomma}\ {\isacharprime}m{\isacharparenright}\ operation\ {\isasymtimes}\ {\isacharparenleft}{\isacharprime}n\ {\isasymtimes}\ {\isacharprime}m\ {\isasymtimes}\ {\isacharprime}n{\isacharparenright}\ set\ {\isasymRightarrow}\ {\isacharparenleft}{\isacharprime}t{\isacharcomma}\ {\isacharprime}n{\isacharcomma}\ {\isacharprime}m{\isacharparenright}\ log{\isacharunderscore}op\ {\isasymtimes}\ {\isacharparenleft}{\isacharprime}n\ {\isasymtimes}\ {\isacharprime}m\ {\isasymtimes}\ {\isacharprime}n{\isacharparenright}\ set{\isacartoucheclose}\ \isakeyword{where}\isanewline
-\ \ {\isacartoucheopen}do{\isacharunderscore}op\ {\isacharparenleft}Move\ t\ newp\ m\ c{\isacharcomma}\ tree{\isacharparenright}\ {\isacharequal}\isanewline
-\ \ \ \ \ {\isacharparenleft}LogMove\ t\ {\isacharparenleft}get{\isacharunderscore}parent\ tree\ c{\isacharparenright}\ newp\ m\ c{\isacharcomma}\isanewline
-\ \ \ \ \ \ if\ ancestor\ tree\ c\ newp\ {\isasymor}\ c\ {\isacharequal}\ newp\ then\ tree\isanewline
-\ \ \ \ \ \ else\ {\isacharbraceleft}{\isacharparenleft}p{\isacharprime}{\isacharcomma}\ m{\isacharprime}{\isacharcomma}\ c{\isacharprime}{\isacharparenright}\ {\isasymin}\ tree{\isachardot}\ c{\isacharprime}\ {\isasymnoteq}\ c{\isacharbraceright}\ {\isasymunion}\ {\isacharbraceleft}{\isacharparenleft}newp{\isacharcomma}\ m{\isacharcomma}\ c{\isacharparenright}{\isacharbraceright}{\isacharparenright}{\isacartoucheclose}\isanewline
+\ \ {\isacartoucheopen}do{\isacharunderscore}op\ {\isacharparenleft}{\rm Move}\ t\ newp\ m\ c{\isacharcomma}\ tree{\isacharparenright}\ {\isacharequal}\isanewline
+\ \ \ \ \ {\isacharparenleft}{\rm LogMove}\ t\ {\isacharparenleft}get{\isacharunderscore}parent\ tree\ c{\isacharparenright}\ newp\ m\ c{\isacharcomma}\isanewline
+\ \ \ \ \ \ \isacommand{if}\ ancestor\ tree\ c\ newp\ {\isasymor}\ c\ {\isacharequal}\ newp\ \isacommand{then}\ tree\isanewline
+\ \ \ \ \ \ \isacommand{else}\ {\isacharbraceleft}{\isacharparenleft}p{\isacharprime}{\isacharcomma}\ m{\isacharprime}{\isacharcomma}\ c{\isacharprime}{\isacharparenright}\ {\isasymin}\ tree{\isachardot}\ c{\isacharprime}\ {\isasymnoteq}\ c{\isacharbraceright}\ {\isasymunion}\ {\isacharbraceleft}{\isacharparenleft}newp{\isacharcomma}\ m{\isacharcomma}\ c{\isacharparenright}{\isacharbraceright}{\isacharparenright}{\isacartoucheclose}\isanewline
 \isanewline
@@
 \ undo{\isacharunderscore}op\ {\isacharcolon}{\isacharcolon}\ {\isacartoucheopen}{\isacharparenleft}{\isacharprime}t{\isacharcomma}\ {\isacharprime}n{\isacharcomma}\ {\isacharprime}m{\isacharparenright}\ log{\isacharunderscore}op\ {\isasymtimes}\ {\isacharparenleft}{\isacharprime}n\ {\isasymtimes}\ {\isacharprime}m\ {\isasymtimes}\ {\isacharprime}n{\isacharparenright}\ set\ {\isasymRightarrow}\ {\isacharparenleft}{\isacharprime}n\ {\isasymtimes}\ {\isacharprime}m\ {\isasymtimes}\ {\isacharprime}n{\isacharparenright}\ set{\isacartoucheclose}\ \isakeyword{where}\isanewline
-\ \ {\isacartoucheopen}undo{\isacharunderscore}op\ {\isacharparenleft}LogMove\ t\ None\ newp\ m\ c{\isacharcomma}\ tree{\isacharparenright}\ {\isacharequal}\ {\isacharbraceleft}{\isacharparenleft}p{\isacharprime}{\isacharcomma}\ m{\isacharprime}{\isacharcomma}\ c{\isacharprime}{\isacharparenright}\ {\isasymin}\ tree{\isachardot}\ c{\isacharprime}\ {\isasymnoteq}\ c{\isacharbraceright}{\isacartoucheclose}\ {\isacharbar}\isanewline
-\ \ {\isacartoucheopen}undo{\isacharunderscore}op\ {\isacharparenleft}LogMove\ t\ {\isacharparenleft}Some\ {\isacharparenleft}oldp{\isacharcomma}\ oldm{\isacharparenright}{\isacharparenright}\ newp\ m\ c{\isacharcomma}\ tree{\isacharparenright}\ {\isacharequal}\isanewline
+    \ \ {\isacartoucheopen}undo{\isacharunderscore}op\ {\isacharparenleft}{\rm LogMove}\ t\ {\rm None}\ newp\ m\ c{\isacharcomma}\ tree{\isacharparenright}\ {\isacharequal}\ {\isacharbraceleft}{\isacharparenleft}p{\isacharprime}{\isacharcomma}\ m{\isacharprime}{\isacharcomma}\ c{\isacharprime}{\isacharparenright}\ {\isasymin}\ tree{\isachardot}\ c{\isacharprime}\ {\isasymnoteq}\ c{\isacharbraceright}{\isacartoucheclose}\ {\isacharbar}\isanewline
+\ \ {\isacartoucheopen}undo{\isacharunderscore}op\ {\isacharparenleft}{\rm LogMove}\ t\ {\isacharparenleft}{\rm Some}\ {\isacharparenleft}oldp{\isacharcomma}\ oldm{\isacharparenright}{\isacharparenright}\ newp\ m\ c{\isacharcomma}\ tree{\isacharparenright}\ {\isacharequal}\isanewline
 \ \ \ \ \ {\isacharbraceleft}{\isacharparenleft}p{\isacharprime}{\isacharcomma}\ m{\isacharprime}{\isacharcomma}\ c{\isacharprime}{\isacharparenright}\ {\isasymin}\ tree{\isachardot}\ c{\isacharprime}\ {\isasymnoteq}\ c{\isacharbraceright}\ {\isasymunion}\ {\isacharbraceleft}{\isacharparenleft}oldp{\isacharcomma}\ oldm{\isacharcomma}\ c{\isacharparenright}{\isacharbraceright}{\isacartoucheclose}\isanewline
@@
 \ redo{\isacharunderscore}op\ {\isacharcolon}{\isacharcolon}\ {\isacartoucheopen}{\isacharparenleft}{\isacharprime}t{\isacharcomma}\ {\isacharprime}n{\isacharcomma}\ {\isacharprime}m{\isacharparenright}\ log{\isacharunderscore}op\ {\isasymRightarrow}\ {\isacharparenleft}{\isacharprime}t{\isacharcomma}\ {\isacharprime}n{\isacharcomma}\ {\isacharprime}m{\isacharparenright}\ state\ {\isasymRightarrow}\ {\isacharparenleft}{\isacharprime}t{\isacharcomma}\ {\isacharprime}n{\isacharcomma}\ {\isacharprime}m{\isacharparenright}\ state{\isacartoucheclose}\ \isakeyword{where}\isanewline
-\ \ {\isacartoucheopen}redo{\isacharunderscore}op\ {\isacharparenleft}LogMove\ t\ {\isacharunderscore}\ p\ m\ c{\isacharparenright}\ {\isacharparenleft}ops{\isacharcomma}\ tree{\isacharparenright}\ {\isacharequal}\isanewline
-\ \ \ \ \ {\isacharparenleft}let\ {\isacharparenleft}op{\isadigit{2}}{\isacharcomma}\ tree{\isadigit{2}}{\isacharparenright}\ {\isacharequal}\ do{\isacharunderscore}op\ {\isacharparenleft}Move\ t\ p\ m\ c{\isacharcomma}\ tree{\isacharparenright}\isanewline
-\ \ \ \ \ \ in\ {\isacharparenleft}op{\isadigit{2}}\ {\isacharhash}\ ops{\isacharcomma}\ tree{\isadigit{2}}{\isacharparenright}{\isacharparenright}{\isacartoucheclose}\isanewline
+\ \ {\isacartoucheopen}redo{\isacharunderscore}op\ {\isacharparenleft}{\rm LogMove}\ t\ {\isacharunderscore}\ p\ m\ c{\isacharparenright}\ {\isacharparenleft}ops{\isacharcomma}\ tree{\isacharparenright}\ {\isacharequal}\isanewline
+\ \ \ \ \ {\isacharparenleft}\isacommand{let}\ {\isacharparenleft}op{\isadigit{2}}{\isacharcomma}\ tree{\isadigit{2}}{\isacharparenright}\ {\isacharequal}\ do{\isacharunderscore}op\ {\isacharparenleft}{\rm Move}\ t\ p\ m\ c{\isacharcomma}\ tree{\isacharparenright}\isanewline
+\ \ \ \ \ \ \isacommand{in}\ {\isacharparenleft}op{\isadigit{2}}\ {\isacharhash}\ ops{\isacharcomma}\ tree{\isadigit{2}}{\isacharparenright}{\isacharparenright}{\isacartoucheclose}\isanewline
 \isanewline
@@
 \ \ {\isacartoucheopen}apply{\isacharunderscore}op\ op{\isadigit{1}}\ {\isacharparenleft}{\isacharbrackleft}{\isacharbrackright}{\isacharcomma}\ tree{\isadigit{1}}{\isacharparenright}\ {\isacharequal}\isanewline
-\ \ \ \ \ {\isacharparenleft}let\ {\isacharparenleft}op{\isadigit{2}}{\isacharcomma}\ tree{\isadigit{2}}{\isacharparenright}\ {\isacharequal}\ do{\isacharunderscore}op\ {\isacharparenleft}op{\isadigit{1}}{\isacharcomma}\ tree{\isadigit{1}}{\isacharparenright}\isanewline
-\ \ \ \ \ \ in\ {\isacharparenleft}{\isacharbrackleft}op{\isadigit{2}}{\isacharbrackright}{\isacharcomma}\ tree{\isadigit{2}}{\isacharparenright}{\isacharparenright}{\isacartoucheclose}\ {\isacharbar}\isanewline
+\ \ \ \ \ {\isacharparenleft}\isacommand{let}\ {\isacharparenleft}op{\isadigit{2}}{\isacharcomma}\ tree{\isadigit{2}}{\isacharparenright}\ {\isacharequal}\ do{\isacharunderscore}op\ {\isacharparenleft}op{\isadigit{1}}{\isacharcomma}\ tree{\isadigit{1}}{\isacharparenright}\isanewline
+\ \ \ \ \ \ \isacommand{in}\ {\isacharparenleft}{\isacharbrackleft}op{\isadigit{2}}{\isacharbrackright}{\isacharcomma}\ tree{\isadigit{2}}{\isacharparenright}{\isacharparenright}{\isacartoucheclose}\ {\isacharbar}\isanewline
 \ \ {\isacartoucheopen}apply{\isacharunderscore}op\ op{\isadigit{1}}\ {\isacharparenleft}logop\ {\isacharhash}\ ops{\isacharcomma}\ tree{\isadigit{1}}{\isacharparenright}\ {\isacharequal}\isanewline
-\ \ \ \ \ {\isacharparenleft}if\ move{\isacharunderscore}time\ op{\isadigit{1}}\ {\isacharless}\ log{\isacharunderscore}time\ logop\isanewline
-\ \ \ \ \ \ then\ redo{\isacharunderscore}op\ logop\ {\isacharparenleft}apply{\isacharunderscore}op\ op{\isadigit{1}}\ {\isacharparenleft}ops{\isacharcomma}\ undo{\isacharunderscore}op\ {\isacharparenleft}logop{\isacharcomma}\ tree{\isadigit{1}}{\isacharparenright}{\isacharparenright}{\isacharparenright}\isanewline
-\ \ \ \ \ \ else\ let\ {\isacharparenleft}op{\isadigit{2}}{\isacharcomma}\ tree{\isadigit{2}}{\isacharparenright}\ {\isacharequal}\ do{\isacharunderscore}op\ {\isacharparenleft}op{\isadigit{1}}{\isacharcomma}\ tree{\isadigit{1}}{\isacharparenright}\ in\ {\isacharparenleft}op{\isadigit{2}}\ {\isacharhash}\ logop\ {\isacharhash}\ ops{\isacharcomma}\ tree{\isadigit{2}}{\isacharparenright}{\isacharparenright}{\isacartoucheclose}\isanewline
+\ \ \ \ \ {\isacharparenleft}\isacommand{if}\ move{\isacharunderscore}time\ op{\isadigit{1}}\ {\isacharless}\ log{\isacharunderscore}time\ logop\isanewline
+\ \ \ \ \ \ \isacommand{then}\ redo{\isacharunderscore}op\ logop\ {\isacharparenleft}apply{\isacharunderscore}op\ op{\isadigit{1}}\ {\isacharparenleft}ops{\isacharcomma}\ undo{\isacharunderscore}op\ {\isacharparenleft}logop{\isacharcomma}\ tree{\isadigit{1}}{\isacharparenright}{\isacharparenright}{\isacharparenright}\isanewline
+\ \ \ \ \ \ \isacommand{else}\ \isacommand{let}\ {\isacharparenleft}op{\isadigit{2}}{\isacharcomma}\ tree{\isadigit{2}}{\isacharparenright}\ {\isacharequal}\ do{\isacharunderscore}op\ {\isacharparenleft}op{\isadigit{1}}{\isacharcomma}\ tree{\isadigit{1}}{\isacharparenright}\ \isacommand{in}\ {\isacharparenleft}op{\isadigit{2}}\ {\isacharhash}\ logop\ {\isacharhash}\ ops{\isacharcomma}\ tree{\isadigit{2}}{\isacharparenright}{\isacharparenright}{\isacartoucheclose}\isanewline
 \isanewline

### 2020-05-22 paper/move-op.tex
Squeeze into 12 pages

@@
 
-\begin{figure*}
-  \centering
-  \includegraphics[width=\textwidth,keepaspectratio=true]{gdrive-error.png}
-  \caption{Error message produced by Google Drive Backup and Sync on Mac OS as a result of performing the operations shown in Figure~\ref{fig:move-cycle}, indicating a bug in the underlying replication algorithm.}
-  \label{fig:gdrive-error}
-\end{figure*}
-
 \section{Why a move operation is hard}\label{sec:move-is-hard}
@@
 As before, we tested this scenario with Google Drive and Dropbox.
-In Google Drive, one replica was able to successfully sync with the server, while the other replica displayed the ``unknown error'' message shown in Figure~\ref{fig:gdrive-error}.
+In Google Drive, one replica was able to successfully sync with the server, while the other replica displayed an ``unknown error'' message.
 The replica in an error state refused to sync the conflicting directory, and its filesystem state remained permanently inconsistent with the other replica.
@@
 All theorems stated here have been formally proved and mechanically checked using Isabelle.
-For space reasons we give only a brief sketch of the proofs in this paper; the Isabelle files containing the full details are open source~\cite{SourceFiles}.
+For space reasons this paper gives only the statements that were proved, but elides a discussion of the reasoning steps.
+The Isabelle files containing the full details are open source~\cite{SourceFiles}, and are included as supplementary material with this paper.
 
@@
 A tree is an acyclic graph in which every node has exactly one parent, except for the root, which has no parent.
-To prove that our algorithm maintains a tree structure, no matter which operations are applied, we prove that these invariants hold.
 In fact, we slightly generalise this property and allow more than one root to exist, so the graph represents a forest, allowing an application to move nodes between different trees, if desired.
 For example, the ``trash'' node used for deletion (\S\ref{sec:ops-trees}) can be separate from the main tree.
+To prove that our algorithm maintains a forest structure, no matter which operations are applied, we demonstrate several invariants.
 
-\subsubsection{Each node's parent is unique}\label{sec:unique-parent}
-
+\smallbreak\noindent\textbf{Each node's parent is unique.}
 The first invariant we prove is that each tree node has either no parent (if it is the root of a tree) or exactly one parent (if it is a non-root node).
-We state this theorem in Isabelle/HOL as follows, where \isa{apply\_ops\_unique\_parent} is the name we give to this theorem:
+We state this theorem in Isabelle/HOL as follows, where \isa{apply\_ops\_unique\_parent} is the name of this theorem:
 \begin{isabelle}
@@
 \noindent That is, we consider any list of operations \isa{ops} and define \isa{(log, tree)} to be the replica state after \isa{ops} have been applied.
-We then prove that \isa{unique\_parent tree} holds, where the \isa{unique\_parent} predicate is defined on lines 54--56 of Figure~\ref{fig:code}: whenever the tree contains a triple whose third element is the child node \isa{c}, then the first and second elements of the triple (the parent node and the metadata) are uniquely defined.
+We then prove that ``\isa{unique\_parent tree}'' holds, where the \isa{unique\_parent} predicate is defined on lines 54--56 of Figure~\ref{fig:code}: whenever the tree contains a triple whose third element is the child node \isa{c}, then the first and second elements of the triple (the parent node and the metadata) are uniquely defined.
+As we make no assumptions about \isa{ops}, this theorem holds for any replica state that can be reached by applying any number of operations.
 
-As we make no assumptions about \isa{ops}, this theorem holds for any replica state that can be reached by applying any number of operations.
-Proving it is easy because the \isa{do\_op}, \isa{undo\_op} and \isa{redo\_op} functions individually maintain the invariant \isa{unique\_parent tree}, and thus \isa{apply\_op} also maintains the invariant.
-
-\subsubsection{The graph contains no cycles}\label{sec:no-cycles}
-
-The second invariant we require is that the graph must not contain any cycles.
-In Isabelle/HOL:
+\smallbreak\noindent\textbf{The graph contains no cycles.}
+This second invariant is expressed as follows in Isabelle/HOL:
 \begin{isabelle}
@@
 
-\emph{Proof sketch}: Assume \isa{do\_op (op1, tree1) = (op2, tree2)} for some \isa{op1}, \isa{op2}, \isa{tree1} and \isa{tree2}.
-Further assume \isa{acyclic tree1}.
-We can then prove by contradiction that \isa{tree2} is also acyclic.
-Assume to the contrary that \isa{tree2} contains a cycle.
-If \isa{do\_op} left the tree unchanged (line 29), we immediately have a contradiction.
-Otherwise we can assume \isa{{\isasymnot}(ancestor tree c newp {\isasymor} c = newp)} and we have:
-\begin{isabelle}
-tree2 = \{(p', m', c') {\isasymin} tree1. c' {\isasymnoteq} c\} {\isasymunion} \{(newp, m, c)\}
-\end{isabelle}
-That is, \isa{do\_op} adds one edge \isa{\{(newp, m, c)\}} to the graph, and perhaps removes another edge.
-Since \isa{tree1} is acyclic, the added edge must lie on the cycle in \isa{tree2}.
-Moreover, since only one edge from \isa{newp} to \isa{c} was added, \isa{tree1} must contain a path from \isa{c} to \isa{newp} (which is then closed by the new edge to form a cycle), i.e.\ \isa{c} must be an ancestor of \isa{newp} in \isa{tree1}.
-However, in that case, the predicate \isa{ancestor tree1 c newp} on line 29 would have been true, which contradicts the earlier assumption.
-Hence, the \isa{do\_op} function preserves the acyclicity invariant of the tree.
-
-Next, we show that for any tree produced by \isa{apply\_ops}, there exists a sequence of operations such that applying those operations using \isa{do\_op} invocations alone (not using \isa{undo\_op} or \isa{redo\_op}) results in the same tree.
-(This is the case when operations are applied in order of increasing timestamp.)
-Then, from the result of the previous paragraph, any tree produced by \isa{apply\_ops} is acyclic.
-
-\subsubsection{Other correctness properties}
-
+\smallbreak\noindent\textbf{Other correctness properties}
 There are further criteria we might use to determine if our algorithm is correct.
@@
 The predicate \isa{distinct} takes a list as argument, and returns true if all elements of the list are distinct (i.e.\ no value occurs more than once in the list).
-The assumption \isa{distinct (map move\_time ops1)} therefore states that in the list of operations \isa{ops1}, there are no two operations with the same timestamp.
+Therefore, the assumption \isa{distinct (map move\_time ops1)} states that in the list \isa{ops1}, there are no two operations with the same timestamp.
 
@@
 
-\subsubsection{Commutativity of apply\_op}\label{sec:commutativity}
-
-To prove this theorem we must first show that \isa{undo\_op} performs the inverse of \isa{do\_op} and that every operation can be undone, that is:
-\begin{isabelle}
-\isacommand{lemma}\isamarkupfalse%
-\ do{\isacharunderscore}undo{\isacharunderscore}op{\isacharunderscore}inv{\isacharcolon}\isanewline
-\ \ \isakeyword{assumes}\ {\isacartoucheopen}unique{\isacharunderscore}parent\ tree{\isacartoucheclose}\isanewline
-\ \ \isakeyword{shows}\ {\isacartoucheopen}undo{\isacharunderscore}op\ {\isacharparenleft}do{\isacharunderscore}op\ {\isacharparenleft}oper{\isacharcomma}\ tree{\isacharparenright}{\isacharparenright}\ {\isacharequal}\ tree{\isacartoucheclose}
-\end{isabelle}
-
-\noindent \isa{apply\_ops\_commutes} then follows from the following lemma, which proves that two calls of \isa{apply\_op} commute, provided that the timestamps in the operations and in the log are all distinct:
-\begin{isabelle}
-\isacommand{lemma}\isamarkupfalse%
-\ apply{\isacharunderscore}op{\isacharunderscore}commute{\isadigit{2}}{\isacharcolon}\isanewline
-\ \ \isakeyword{assumes}\ {\isacartoucheopen}unique{\isacharunderscore}parent\ tree{\isacartoucheclose}\isanewline
-\ \ \ \ \isakeyword{and}\ {\isacartoucheopen}distinct\ {\isacharparenleft}{\isacharparenleft}map\ log{\isacharunderscore}time\ log{\isacharparenright}\ {\isacharat}\isanewline
-\ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ {\isacharparenleft}map\ move{\isacharunderscore}time\ {\isacharbrackleft}oper{\isadigit{1}}{\isacharcomma}\ oper{\isadigit{2}}{\isacharbrackright}{\isacharparenright}{\isacharparenright}{\isacartoucheclose}\isanewline
-\ \ \isakeyword{shows}\ {\isacartoucheopen}apply{\isacharunderscore}op\ oper{\isadigit{2}}\ {\isacharparenleft}apply{\isacharunderscore}op\ oper{\isadigit{1}}\ {\isacharparenleft}log{\isacharcomma}\ tree{\isacharparenright}{\isacharparenright}\ {\isacharequal}\isanewline
-\ \ \ \ \ \ \ \ \ apply{\isacharunderscore}op\ oper{\isadigit{1}}\ {\isacharparenleft}apply{\isacharunderscore}op\ oper{\isadigit{2}}\ {\isacharparenleft}log{\isacharcomma}\ tree{\isacharparenright}{\isacharparenright}{\isacartoucheclose}
-\end{isabelle}
-
-\noindent Here the \isa{\isacharat} operator concatenates two lists.
-Let \isa{t1 = move\_time oper1} and \isa{t2 = move\_time oper2} be the timestamps of the two operations \isa{oper1} and \isa{oper2}.
-Assume $\isa{t1} < \isa{t2}$ without loss of generality.
-We can now prove lemma \isa{apply\_op\_commute2} by induction over \isa{log}.
-
-For the base case, where \isa{log} is empty, \isa{apply\_op oper2 (apply\_op oper1 ([], tree))} evaluates to first applying \isa{oper1} using \isa{do\_op}, then applying \isa{oper2} using \isa{do\_op}.
-On the other hand, \isa{apply\_op oper1 (apply\_op oper2 ([], tree))} first applies \isa{oper2} using \isa{do\_op}, then undoes \isa{oper2} again to yield the original \isa{tree} (per \isa{do\_undo\_op\_inv}), then applies \isa{oper1} using \isa{do\_op}, and finally reapplies \isa{oper2} using \isa{redo\_op}.
-The final states are the same on both sides of the equality.
-
-For the inductive step, let the log be \isa{logop {\isacharhash} ops} (where \isa{\isacharhash} is the \emph{list cons} operator).
-We now need to show that:
-\begin{isabelle}
-apply\_op oper2 (apply\_op oper1 (logop {\isacharhash} ops, tree)) =\\
-apply\_op oper1 (apply\_op oper2 (logop {\isacharhash} ops, tree)).
-\end{isabelle}
-Let \isa{t3 = log\_time logop} be the timestamp of the operation at the head of the log.
-We now consider three cases:
-\begin{enumerate}
-    \item $\isa{t3} < \isa{t1} < \isa{t2}$.
-        Then neither applying \isa{oper1} nor applying \isa{oper2} will undo \isa{logop}.
-        Thus, the two operations are applied in the same way as in the base case.
-    \item $\isa{t1} < \isa{t3} < \isa{t2}$.
-        Then \isa{apply\_op oper2 (apply\_op oper1 (logop {\isacharhash} ops, tree))} first undoes \isa{logop}, then applies \isa{oper1} using \isa{do\_op}, then reapplies \isa{logop} using \isa{redo\_op}, and finally applies \isa{oper2} using \isa{do\_op}.
-        On the other hand, \isa{apply\_op oper1 (apply\_op oper2 (logop {\isacharhash} ops, tree))} first applies \isa{oper2} using \isa{do\_op}, then undoes it again, then undoes \isa{logop}, applies \isa{oper1} and reapplies \isa{logop} as before, and finally reapplies \isa{oper2} using \isa{redo\_op}.
-        The end result is the same.
-    \item $\isa{t1} < \isa{t2} < \isa{t3}$.
-        Then both applying \isa{oper1} and applying \isa{oper2} first undoes \isa{logop}, yielding \isa{tree' = undo\_op (logop, tree)}.
-        From the inductive hypothesis we have:
-        \begin{isabelle}
-        apply\_op oper2 (apply\_op oper1 (ops, tree')) =\\
-        apply\_op oper1 (apply\_op oper2 (ops, tree'))
-        \end{isabelle}
-        We now apply \isa{oper1} and \isa{oper2} as in the base case, and finally reapply \isa{logop} to both sides using \isa{redo\_op}, resulting in the same state on both sides.
-\end{enumerate}
-
-This completes the proof that any two operations commute, provided that their timestamps are distinct from each other and from the contents of the log.
-We can easily generalise this to show that any number of operations commute: given any sequence of operations, any permutation of that sequence will have the same effect, because we can produce the permutation by repeatedly swapping adjacent, commutative operations.
-
-\subsubsection{Strong eventual consistency}
-
-Gomes et al.~\cite{Gomes:2017gy} define a framework in Isabelle/HOL for proving the strong eventual consistency properties of CRDTs above an axiomatic network model.
-Using our commutativity proof above we integrate our tree datatype with this framework, and thus prove that our move operation on trees does indeed guarantee strong eventual consistency.
+\smallbreak\noindent\textbf{Strong eventual consistency.}
+Gomes et al.~\cite{Gomes:2017gy} define a framework in Isabelle/HOL for proving the strong eventual consistency properties of CRDTs.
+Using our convergence proof above we integrate our tree datatype with this framework, and thus demonstrate that our move operation on trees does indeed guarantee strong eventual consistency.
 The details appear in our Isabelle theory files~\cite{SourceFiles}.
@@
 The \isa{hashable} type class means that keys must have a hashing function.
-In effect, this hash-map is an index over the set of triples, using the fact that the parent and metadata for a given child are unique (\S\ref{sec:unique-parent}).
+In effect, this hash-map is an index over the set of triples, using the fact that the parent and metadata for a given child are unique (\S\ref{sec:tree-invariants}).
 The hash-map implementation is from the Isabelle Collections Framework~\cite{DBLP:conf/itp/LammichL10}.
@@
 \noindent where \isa{hm.lookup c t} looks up the key \isa{c} in the hash-map \isa{t}, returning \isa{Some x} if \isa{c} maps to the value \isa{x}, and returning \isa{None} if \isa{c} does not appear in the hash-map.
-We can now prove the equivalence of the set-based and the hash-map-based versions of each function.
-For example, \isa{executable\_do\_op} is the equivalent of the \isa{do\_op} function based on a hash-map:
-\begin{isabelle}
-\isacommand{lemma}\isamarkupfalse%
-\ executable{\isacharunderscore}do{\isacharunderscore}op{\isacharunderscore}simulates{\isacharcolon}\isanewline
-\ \ \isakeyword{assumes}\ {\isacartoucheopen}simulates\ t\ T{\isacartoucheclose}\isanewline
-\ \ \ \ \isakeyword{and}\ {\isacartoucheopen}executable{\isacharunderscore}do{\isacharunderscore}op\ {\isacharparenleft}oper{\isacharcomma}\ t{\isacharparenright}\ {\isacharequal}\ {\isacharparenleft}log{\isadigit{1}}{\isacharcomma}\ u{\isacharparenright}{\isacartoucheclose}\isanewline
-\ \ \ \ \isakeyword{and}\ {\isacartoucheopen}do{\isacharunderscore}op\ {\isacharparenleft}oper{\isacharcomma}\ T{\isacharparenright}\ {\isacharequal}\ {\isacharparenleft}log{\isadigit{2}}{\isacharcomma}\ U{\isacharparenright}{\isacartoucheclose}\isanewline
-\ \ \isakeyword{shows}\ {\isacartoucheopen}log{\isadigit{1}}\ {\isacharequal}\ log{\isadigit{2}}\ {\isasymand}\ simulates\ u\ U{\isacartoucheclose}
-\end{isabelle}
-
-Similarly, we can show this simulation relation holds for each function in our algorithm, all the way to \isa{apply\_ops}:
+We can now prove the equivalence of the set-based and the hash-map-based implementations:
 \begin{isabelle}
@@
 We deployed three replicas of this service on Amazon EC2 \texttt{c5.large} instances in different regions worldwide: in Northern California (region \texttt{us-west-1}), Ireland (\texttt{eu-west-1}), and Singapore (\texttt{ap-southeast-1}).
-The network delays between these regions are substantial, as shown in Table~\ref{tab:rpc-times}.
+The network delays between these regions are substantial: in our experiments, a request-response round-trip from Ireland to California takes a median of 145~ms, while a request from Singapore to California takes 176~ms.
 
@@
 At around 600 operations/sec each single-threaded replica is fully utilising one CPU core, and increasing the rate at which operations are generated does not increase throughput any further.
-
-\begin{table}
-  \caption{Median response times of an RPC request between replicas in different regions.}
-  \label{tab:rpc-times}
-  \begin{tabular}{r|rrr}
-    \toprule
-                    & to California & to Ireland & to Singapore \\
-    \midrule
-    from California & ---           & 139 ms     & 175 ms       \\
-    from Ireland    & 145 ms        & ---        & 182 ms       \\
-    from Singapore  & 176 ms        & 178 ms     & ---          \\
-    \bottomrule
-\end{tabular}
-\end{table}
 
@@
 On the other hand, for remote operations, a number of undos and redos must be performed, depending on the degree of concurrency.
-As the time interval between successive operations is smaller than the network delay between replicas, there are multiple operations ``in flight'' at the same time, and more calls to \isa{undo\_op} and \isa{redo\_op} are required to order operations by timestamp.
+As the time interval between successive operations decreases relative to the network delay, there are more operations ``in flight'' at the same time, and more calls to \isa{undo\_op} and \isa{redo\_op} are required to order operations by timestamp.
 The time taken to apply a remote operation is proportional to the number of undos and redos required, and hence proportional to the operation rate.
@@
 The results from this experiment are shown in Figure~\ref{fig:leader-plot}.
-The leader-based approach was able to sustain a throughput of 14,000 move operations per second, approximately 23 times the CRDT's throughput of 600 ops/sec (shown as an asymptote in Figure~\ref{fig:leader-plot}), due to the fact that it does not need to spend CPU cycles undoing and redoing operations.
-On the other hand, the latency of each operation is much higher in the state machine approach, because the generating replica needs to wait for a round trip through the leader.
-This latency is around 145~ms for operations generated in Ireland and around 176~ms for Singapore (cf.
-Table~\ref{tab:rpc-times}), compared with the CRDT's 50~{\textmu}s processing time for local operations.
+The leader-based approach was able to sustain a throughput of 14,000 move operations per second, approximately 23 times the CRDT's throughput of 600 ops/sec, due to the fact that it does not need to spend CPU cycles undoing and redoing operations.
+On the other hand, the latency of each operation is around 3,000 times greater in the state machine approach, due to the network delay to the leader.
 
 Therefore, we have a clear trade-off: in applications that need to prioritise throughput, a state machine replication approach is preferable, while in applications that need to minimise response times to user requests, our CRDT algorithm is preferable.
-Our algorithm also has the advantage that local user operations require no waiting for network communication, making them highly available in the face of network failures.
-It also supports offline operation: mobile devices can read and modify their local replica of the tree even while disconnected from the Internet, which is not the case with leader-based replication.
+Our algorithm also has the advantage that local user operations require no waiting for network communication, so users can still modify the tree while disconnected from the Internet.
+In the leader-based approach, clients cannot make updates while offline.
 
@@
 Besides the 59 lines of definitions given in Figure~\ref{fig:code}, our Isabelle/HOL formalisation consists of a further 2,166 lines of code.
-Of this, we use 203 lines to prove that every node has a unique parent, 443 lines to prove that the tree contains no cycles, 450 lines to prove that move operations commute and replicas converge, 327 lines to prove the strong eventual consistency of our algorithm using the framework of Gomes et al.~\cite{Gomes:2017gy}, and 743 lines to define the executable variant of our algorithm (\S\ref{subsect.extracting}) and prove its equivalence to the abstract definitions of Figure~\ref{fig:code}.
+Of this, we use 203 lines to prove that every node has a unique parent, 443 lines to prove that the tree contains no cycles, 450 lines to prove that move operations commute and replicas converge, 327 lines to prove the strong eventual consistency property, and 743 lines to define the executable variant of our algorithm and prove its equivalence to the definitions of Figure~\ref{fig:code}.
 
@@
 Databases such as Dynamo~\cite{DeCandia:2007ui} and Bayou~\cite{Terry:1995dn} rely on the application programmer to provide explicit conflict resolution logic; however, such logic is difficult to get right \cite{Bailis:2013jc,Burckhardt:2014hy,Gomes:2017gy}.
-Hence, we focus on systems that automatically ensure that all replicas converge towards a consistent state, without requiring custom application logic---a consistency model known as \emph{strong eventual consistency}~\cite{Shapiro:2011un,Gomes:2017gy}.
+Hence, we want to automatically ensure that all replicas converge towards a consistent state, without requiring custom application logic---a consistency model known as \emph{strong eventual consistency}~\cite{Shapiro:2011un,Gomes:2017gy}.
 
@@
     \item As discussed in \S\ref{sec:impossibility}, Najafzadeh et al.~\cite{Najafzadeh:2017vk,Najafzadeh:2018bw} propose two implementations for a replicated filesystem: a CRDT in which conflicting moves are handled by duplicating tree nodes (as in Figures~\ref{fig:move-same-item}b and~\ref{fig:move-cycle}b), and a centralised implementation in which move operations must obtain a lock before executing (not a CRDT since it relies on synchronous coordination).
-    \item Ahmed-Nacer et al.~\cite{AhmedNacer:2012us} outline approaches to handling conflicts on trees, but do not provide algorithms.
+    \item Ahmed-Nacer et al.~\cite{AhmedNacer:2012us} outline approaches to handling conflicts on trees, but provide no algorithms.
     \item Tao et al.~\cite{Tao:2015gd} propose handling conflicting move operations by allowing the same object to appear in more than one location; thus, their datatype is strictly a DAG, not a tree.
@@
 Besides CRDTs, another family of algorithms for concurrent modification of data structures is \emph{Operational Transformation} (OT)~\cite{Sun:1998vf}.
-Several authors have defined concurrent tree structures using OT \cite{Jungnickel:2016cb,Ignat:2003jy,Davis:2002iv}, but they only consider insertion and deletion of nodes, and do not support moves.
+Several authors have defined concurrent tree structures using OT \cite{Jungnickel:2016cb,Ignat:2003jy,Davis:2002iv}, but they only handle insertion and deletion of nodes, and do not support moves.
 
@@
 Like Rumor, Unison is a userspace process that compares replica states.
-Whenever it is run, Unison records a summary of the filesystem state on each replica, and it uses this summary to determine the changes made since the last synchronisation. 
+Whenever it is run, Unison determines the changes made since the last synchronisation. 
 When presented with the move operations described in Figure~\ref{fig:move-same-item}, Unison duplicates the files, resulting in the outcome shown in Figure~\ref{fig:move-same-item}a. 
@@
 
-Our approach of ordering operations by a timestamp, and undoing/redoing them as necessary so that they take effect in ascending timestamp order, is conceptually a very simple approach---although the details of the implementation are a bit subtle.
+Our approach of ordering operations by a timestamp, and undoing/redoing them as necessary so that they take effect in ascending timestamp order, is conceptually very simple.
 Similar ideas appear in many other systems, including the Bayou database~\cite{Terry:1995dn}, Jefferson's \emph{Time Warp} mechanism~\cite{Jefferson:1985em}, Shapiro et al.'s \emph{capricious total order}~\cite{Shapiro:2016hp}, and Burckhardt's \emph{standard conflict resolution}~\cite[{\S}4.3.3]{Burckhardt:2014hy}.
@@
 To rule out such bugs, we formally verified the correctness of our algorithm using the Isabelle/HOL proof assistant; our theorems show that replicas can apply operations in any order, and that the result is always a valid tree (nodes have at most one parent, and the graph does not contain any cycles).
-Moreover, these results apply to unbounded executions and an arbitrary number of replicas---an advantage of using a proof assistant rather than other formal approaches such as model checking.
+Moreover, these results apply to unbounded executions and an arbitrary number of replicas---an advantage of using a proof assistant rather than model checking.
 

### 2020-05-22 paper/move-op.tex
Additions to conclusions to address review comments

@@
 The simplicity of the approach facilitates reasoning about it.
-We are exploring applying undo-do-redo to other replication problems as well, and we expect that our formal verification approach will also benefit other applications of this principle.
 
@@
 
+One might wonder whether our algorithm's consistency model is strong enough for practical use.
+On this point, we note that this model is what Google Drive and Dropbox use today~\cite{Hughes:2016fp} (apart from the aforementioned bugs).
+
 We also extracted a formally verified Scala implementation from our Isabelle/HOL definitions, and evaluated its performance across three replicas in California, Ireland and Singapore.
-Our implementation applies local updates in approximately 50~{\textmu}s, orders of magnitude faster than is possible with a leader replica or consensus protocol operating over these distances, and it remains available in the face of network interruptions.
+Our implementation applies local updates in approximately 50~{\textmu}s, three orders of magnitude faster than is possible with a leader replica or consensus protocol operating over these distances, and it remains available in the face of network interruptions.
 However, compared to leader-based replication our algorithm has lower throughput of 600 updates per second.
@@
 In future work we hope to build novel applications, such as a distributed filesystem, using our algorithm.
+
+We are also exploring whether the undo-do-redo approach can be used in other concurrent data structures; for example, there is an open problem in collaborative text editing~\cite{Kleppmann:2020jw} that might be solved by this approach.
+We expect that our Isabelle/HOL formalisation will be a valuable resource for future work in this area, which can use our definitions and proofs as starting point.
+The formalisation is included as supplementary data of this paper.
 

### 2020-06-08 paper/move-op.tex
Add photos and biographies

@@
 \bibliography{references}{}
+\newpage
+
+\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{photo-martin.jpg}}]{Martin Kleppmann}
+is a Senior Research Associate at the University of Cambridge.
+His research spans distributed systems, databases, security, and formal verification, with a particular focus on decentralised collaboration software and CRDTs.
+His book \emph{Designing Data-Intensive Applications} was published in 2017 and has been translated into six languages.
+Previously, he worked as a software engineer and entrepreneur at several internet companies, including Rapportive and LinkedIn.
+\end{IEEEbiography}
+\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{photo-dominic.jpg}}]{Dominic P.\ Mulligan}
+is a Staff Research Engineer in the Security Group at Arm Research, UK.
+Prior to moving to Arm, he worked as a postdoctoral researcher at the Universities of Cambridge and Bologna, Italy, investigating the formal specification and verification of systems software such as C compilers and linkers.
+His interests include formal verification, distributed systems, and privacy and security.
+\end{IEEEbiography}
+\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{photo-victor.jpg}}]{Victor B.\ F.\ Gomes}
+received a diplôme d'ingénieur from INSA Lyon and a PhD degree from the University of Sheffield.
+He was a research associate at the University of Cambridge from 2016 to 2019 and he is currently working at Google.
+His main research interests are semantics of programming languages, formal verification via theorem prover assistants and algebraic approaches for program verification.
+\end{IEEEbiography}
+\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{photo-alastair.jpg}}]{Alastair R.\ Beresford}
+is Professor of Computer Security in the Department of Computer Science at the University of Cambridge.
+His research work explores the security and privacy of large-scale distributed systems, with a particular focus on networked mobile devices such as smartphones, tablets and laptops.
+He looks at the security and privacy of the devices themselves, as well as the security and privacy problems induced by the interaction between mobile devices and cloud-based Internet services.
+\end{IEEEbiography}
+\vfill
 \end{document}

### 2020-06-08 paper/move-op.tex
Reinstate the error message figure

@@
 
+\begin{figure*}
+  \centering
+  \includegraphics[width=\textwidth,keepaspectratio=true]{gdrive-error.png}
+  \caption{Error message produced by Google Drive Backup and Sync on Mac OS as a result of performing the operations shown in Figure~\ref{fig:move-cycle}, indicating a bug in the underlying replication algorithm.}
+  \label{fig:gdrive-error}
+\end{figure*}
+
 \section{Why a move operation is hard}\label{sec:move-is-hard}
@@
 As before, we tested this scenario with Google Drive and Dropbox.
-In Google Drive, one replica was able to successfully sync with the server, while the other replica displayed an ``unknown error'' message.
+In Google Drive, one replica was able to successfully sync with the server, while the other replica displayed the ``unknown error'' message shown in Figure~\ref{fig:gdrive-error}.
 The replica in an error state refused to sync the conflicting directory, and its filesystem state remained permanently inconsistent with the other replica.
@@
 Like Rumor, Unison is a userspace process that compares replica states.
-Whenever it is run, Unison determines the changes made since the last synchronisation. 
+Whenever it is run, Unison records a summary of the filesystem state on each replica, and it uses this summary to determine the changes made since the last synchronisation. 
 When presented with the move operations described in Figure~\ref{fig:move-same-item}, Unison duplicates the files, resulting in the outcome shown in Figure~\ref{fig:move-same-item}a. 
@@
 To rule out such bugs, we formally verified the correctness of our algorithm using the Isabelle/HOL proof assistant; our theorems show that replicas can apply operations in any order, and that the result is always a valid tree (nodes have at most one parent, and the graph does not contain any cycles).
-Moreover, these results apply to unbounded executions and an arbitrary number of replicas---an advantage of using a proof assistant rather than model checking.
+Moreover, these results apply to unbounded executions and an arbitrary number of replicas---an advantage of using a proof assistant rather than other formal approaches such as model checking.
 

### 2021-02-10 paper/review-response.tex
Add TPDS reviews

\documentclass[10pt]{article}
\usepackage[a4paper,margin=1in]{geometry}
\usepackage{spverbatim}
\sloppy

\usepackage{xr}
\externaldocument{move-op}

\newcommand{\todo}[1]{\textcolor{red}{TODO: #1}}
\newcommand{\authorcomment}[1]{\begin{quote}\textbf{Author comment:} #1\end{quote}}

\begin{document}
\title{Summary of changes of ``A highly-available move operation for replicated trees''}
\author{Martin Kleppmann, Dominic P.\ Mulligan, Victor B.F.\ Gomes, Alastair R.\ Beresford}
\date{}
\maketitle

We thank the reviewers very much for their comments, which have improved the paper.
In this document we explain how we have addressed the issues highlighted in the reviews.
The most important changes are as follows:
\begin{itemize}
\item We have changed the title (previously \emph{``A highly-available move operation for replicated trees and distributed filesystems''}) to remove \emph{``and distributed filesystems''}, as suggested by Reviewer 2.
\end{itemize}

\section{Reviewer 1}

\begin{spverbatim}
Recommendation: Author Should Prepare A Major Revision For A Second Review

Comments:
Introduction section offers very brief and simplified coverage of common topics and don't really offer anything new.
It would be even more valuable if some critical comparisons were made among the systems and approaches. There is no discussion that could help the reader understand that, given a particular system/approach configuration or set of performance or availability goals, approach X would be preferable to approach Y.
In Section 3.5, please explain all cases with examples.
My main concern is whether the comparison is fair. It shows that some results are achievable. How could one trustfully decide to apply the operations on the basis of these results for another scenario?

Additional Questions:
1.  Please explain how this manuscript advances this field of research and/or contributes something new to the literature.: It presents move operation for distributed filesystems.

2. Is the manuscript technically sound? Please explain your answer under Public Comments below.: Yes

1. Which category describes this manuscript?: Practice / Application / Case Study / Experience Report

2. How relevant is this manuscript to the readers of this periodical? Please explain your rating under Public Comments below.: Relevant

1. Are the title, abstract, and keywords appropriate? Please explain under Public Comments below.: Yes

2. Does the manuscript contain sufficient and appropriate references? Please explain under Public Comments below.: Important references are missing; more references are needed

3. Does the introduction state the objectives of the manuscript in terms that encourage the reader to read on? Please explain your answer under Public Comments below.: Could be improved

4. How would you rate the organization of the manuscript? Is it focused? Is the length appropriate for the topic? Please explain under Public Comments below.: Could be improved

5. Please rate the readability of the manuscript. Explain your rating under Public Comments below.: Readable - but requires some effort to understand

6. Should the supplemental material be included? (Click on the Supplementary Files icon to view files): Yes, as part of the digital library for this submission if accepted

7. If yes to 6, should it be accepted: After revisions.  Please include explanation under Public Comments below.

8. Would you recommend adding the code/data associated with this paper to help address your concerns and/or strengthen the paper?: Yes

Please rate the manuscript. Please explain your choice.: Good
\end{spverbatim}

\section{Reviewer 2}

\begin{spverbatim}
Recommendation: Author Should Prepare A Major Revision For A Second Review

Comments:
## Is the manuscript technically sound?
Yes. As stated before the paper has a clear contribution, which includes a mechanised proof of the algorithm correctness. Moreover, the proof code is presented in the appendix and is available online with instructions on how to check proofs locally.

## How relevant is this manuscript to the readers of this periodical?
The topic is relevant for TPDS community as it proposes a distributed algorithm to support optimistic replication of tree data structures.

## Are the title, abstract, and keywords appropriate?
Yes. However, it is unclear why is necessary to include "distributed filesystems" in the title. The work is concerned with a tree replicated datatype and distributed filesystems is an application.

## Does the manuscript contain sufficient and appropriate references?
The following reference should be included and is missing:

De Porre K, Myter F, De Troyer C, Scholliers C, De Meuter W, Boix EG. Putting order in strong eventual consistency. In IFIP International Conference on Distributed Applications and Interoperable Systems 2019 Jun 17 (pp. 36-56). Springer, Cham.

De Porre et al. work proposes an optimistic replication algorithm that can transform a general-purpose data type into a replicated data type. Similarly to the submitted paper, strong eventual consistency is ensured by maintaining a global total order of the operations. Since the case study in De Porre's paper is an AVL tree, a comparison should be included in the related work.

## Does the introduction state the objectives of the manuscript in terms that encourage the reader to read on?
I found the introduction enticing and, although brief, it does a great job in defining the context, illustrating the problem, and clearly defining the contributions.

## How would you rate the organization of the manuscript? Is it focused? Is the length appropriate for the topic?
The paper is well structured. I particularly liked section 2 that clearly illustrates the main issues with concurrent non-coordinated move operations. Also, by showing that the replicated filesystems of Dropbox and Google drive both have bugs demonstrates how difficult is to devise a correct resolution policy for the move operation. Overall, section 2 nicely shows the work relevance and potential impact without getting too technical.
The replication algorithm is thoroughly described and the proof of correctness is presented at the right level of detail - although the proofs are an essential component a more detailed presentation would become tedious.

## Please rate the readability of the manuscript.
The paper is well written and easy to follow.

## Detailed comments

C1: The system model and network requirements should be explicitly defined.

C2: Is the metadata parameter necessary? It seems that this parameter could be abstracted away which would simplify the algorithm presentation.

C3: Further justification is needed for choosing to encode add and remove operations as move operations. My intuition, base on previously published results as CISE and Hamsaz, is that not all operations need to follow a total ordered across replicas. However, because all operations are defined as moves, conflicts are always possible. I was left wondering if by having explicit add and remove operations some undo-do-redo cycles could be avoided.

C4: Instead of having an undo-do-redo cycle would it be possible to have state checkpoints? As operations might arrive in any order, the number of operations to undo can be high, in that case, it might be useful to have state checkpoints and combine these with undo operations.

C5: Notation is not consistent. It is stated that types are primed letters (2nd paragraph of section 3.1), however, this naming convention is dropped later on.

C6: The evaluation proves the point that the tree CRDT is the best option for fast response time and disconnected mode (in this case it is the only option).

C7: The choice of using the extracted code for the evaluation is questionable. Stating that the code could be optimised, does not provide useful information. It is unclear what the impact would be if optimisations were applied. It's reasonable to conclude that the comparison between the proposed CRDT and state machine replication would be maintained, but the optimised code would give a clear evaluation of the tree CRDT.

## Overall evaluation

I really enjoyed reading this paper, it's well written and easy to follow and addresses a relevant topic to the TPDS community.
It has a novel contribution as it defines an optimistic replication algorithm for trees that ensures convergence and the structural properties of trees. The idea behind the tree CRDT is simple, but it provides a solution to an existing problem. The fact that authors found problems in widely used applications that could be addressed with the tree CRDT proves its usefulness.
The fact the CRDT was formally verified for any number of replicas and a certified implementation was extracted from the formal definitions, strengths considerably the submission.
There are some choices that need to be further discussed and the evaluation should consider an optimised implementation of the algorithm.

Additional Questions:
1.  Please explain how this manuscript advances this field of research and/or contributes something new to the literature.: The paper presents the first tree replicated data type that allows concurrent move operations without any coordination between replicas. The tree data type relies on an optimistic replication algorithm that ensures state convergence and that the tree structure properties are preserved at all times. Given the complex (inductive) structure of trees, a strong point of the work is the mechanised proof that provides certification for the correctness of the algorithm.

2. Is the manuscript technically sound? Please explain your answer under Public Comments below.: Yes

1. Which category describes this manuscript?: Research/Technology

2. How relevant is this manuscript to the readers of this periodical? Please explain your rating under Public Comments below.: Relevant

1. Are the title, abstract, and keywords appropriate? Please explain under Public Comments below.: Yes

2. Does the manuscript contain sufficient and appropriate references? Please explain under Public Comments below.: Important references are missing; more references are needed

3. Does the introduction state the objectives of the manuscript in terms that encourage the reader to read on? Please explain your answer under Public Comments below.: Yes

4. How would you rate the organization of the manuscript? Is it focused? Is the length appropriate for the topic? Please explain under Public Comments below.: Satisfactory

5. Please rate the readability of the manuscript. Explain your rating under Public Comments below.: Easy to read

6. Should the supplemental material be included? (Click on the Supplementary Files icon to view files): Yes, as part of the digital library for this submission if accepted

7. If yes to 6, should it be accepted: After revisions.  Please include explanation under Public Comments below.

8. Would you recommend adding the code/data associated with this paper to help address your concerns and/or strengthen the paper?: No

Please rate the manuscript. Please explain your choice.: Excellent
\end{spverbatim}

\section{Reviewer 3}

\begin{spverbatim}
Recommendation: Accept With No Changes

Comments:
The paper presents an implementation of a replicated tree data type with operations for adding, removing, and moving tree nodes. Replicated tree data types are essential basic blocks of a wide range of distributed software like distributed file systems, editors, applications manipulating XML/JSON structures, etc. While replication is needed to ensure resilience to network/machine failures, an important and difficult challenge in this context is ensuring some form of convergence, e.g., that all the nodes storing a replica of the data structure (tree) reach the same state when all the updates have been propagated in the network. This is an instance of strong eventual consistency. The paper presents a solution to this challenge that provides high availability as well (clients do not need to wait for the replicas to synchronize), and that is also formally verified in the Isabelle/HOL theorem prover.

The paper starts with a very clear description of the motivation, noting that current solutions for the above challenge, deployed in Google Drive or Dropbox, are not satisfactory. The Google Drive solution may lead to unexpected errors while the Dropbox solution may duplicate nodes of the tree which may impact the user experience. It identifies the main challenge in designing replicated trees which is related to the move operation: moving a tree node so that it becomes the child of a different parent. In fact, node insertion and removal can be reduced to move operations.

Then, it describes an implementation that in particular, solves the issues in Google Drive or Dropbox, which is based on a very neat and simple idea. Essentially, every operation is associated with a timestamp (one can use Lamport timestamps [11] and avoid synchronized clocks), and every replica maintains a tree and a log of operations (issued locally or remote operations received from other replicas), ordered by timestamps. Applying a remote operation O with timestamp T consists in undoing operations with a timestamp bigger than T, applying O, and finally, re-applying the undone operations. This solution, which is quite generic and could be applied for other data types as well, makes it possible to handle concurrent move operations that try to move the same node (taking into account only the one with a bigger timestamp like in Last Writer Wins conflict resolution [10]), and avoid introducing cycles due to concurrent move operations. It also ensures strong eventual consistency because when all the updates are received, any two replicas apply them in the same order (according to their timestamps). The fact that the implementation satisfies strong eventual consistency and that every replica stores a tree (without cycles) is proved in the Isabelle/HOL theorem prover. The implementation in Isabelle is compiled to executable Scala code which is evaluated in a rather realistic deployment (with three replicas on different continents) and compared to a standard master-slave implementation where the order between operations is managed by a single master node. This evaluation shows that indeed, the current solution has very low latency compared to the master-slave implementation, although a smaller throughput.

The related work is quite extensive, covering the literature on Conflict-Free Data Types (the implementation in this paper being an instance of such a datatype), distributed file systems (the main application of this work), and works that use similar implementation ideas, based on timestamps.

Overall, I think that the paper is very well written. The motivation and the description of the challenges are very convincing, the flow of the ideas being very easy to follow. Technically, the implementation relies on an idea which is simple in hindsight, but which has not been used in previous work (up to my knowledge), and which seems to have a very important practical impact. The formalization and proofs in Isabelle/HOL are a very important and non-trivial contribution which is meant to increase the confidence we have in this solution.

Additional Questions:
1.  Please explain how this manuscript advances this field of research and/or contributes something new to the literature.: The paper presents a new formally-verified implementation of a replicated tree data type, which is an essential ingredient of a wide range of distributed software like distributed file systems, editors, applications manipulating XML/JSON structures, etc. This implementation is highly-available (the latency of applying an operation is independent of the network's latency) and eventually consistent (all replicas reach the same state when all updates have been propagated). These properties are essential in modern distributed services. The implementation solves existing issues in industrial applications like Google Drive and Dropbox, and contradicts claims about the existence of such an implementation in previous research articles.

2. Is the manuscript technically sound? Please explain your answer under Public Comments below.: Yes

1. Which category describes this manuscript?: Research/Technology

2. How relevant is this manuscript to the readers of this periodical? Please explain your rating under Public Comments below.: Very Relevant

1. Are the title, abstract, and keywords appropriate? Please explain under Public Comments below.: Yes

2. Does the manuscript contain sufficient and appropriate references? Please explain under Public Comments below.: References are sufficient and appropriate

3. Does the introduction state the objectives of the manuscript in terms that encourage the reader to read on? Please explain your answer under Public Comments below.: Yes

4. How would you rate the organization of the manuscript? Is it focused? Is the length appropriate for the topic? Please explain under Public Comments below.: Satisfactory

5. Please rate the readability of the manuscript. Explain your rating under Public Comments below.: Easy to read

6. Should the supplemental material be included? (Click on the Supplementary Files icon to view files): Yes, as part of the digital library for this submission if accepted

7. If yes to 6, should it be accepted: As is

8. Would you recommend adding the code/data associated with this paper to help address your concerns and/or strengthen the paper?: Yes

Please rate the manuscript. Please explain your choice.: Excellent
\end{spverbatim}
\end{document}

### 2021-02-24 evaluation/data/leader_mode.sh
New data processing scripts and graphs

@@ -1,12 +1,0 @@
-#!/bin/bash
-
-echo  ",,us_west_1,,,,,,,ap_southeast_1" > leader_mode.csv
-echo -n "interval,total ops/sec," >> leader_mode.csv
-echo -n "generated ops/sec,RTT min,RTT median,RTT p95,apply min,apply median,apply p95," >> leader_mode.csv
-echo    "generated ops/sec,RTT min,RTT median,RTT p95,apply min,apply median,apply p95" >> leader_mode.csv
-
-for interval in 100000 50000 20000 10000 5000 2000 1000 500 200 100 50 20 10; do
-    eu_times="$(cat logs-leader/interval_${interval}_eu_west_1.log.gz      | gunzip | awk -f leader_mode.awk | tail -n 2 | head -n 1 | tr '\n' ',')"
-    ap_times="$(cat logs-leader/interval_${interval}_ap_southeast_1.log.gz | gunzip | awk -f leader_mode.awk | tail -n 2 | head -n 1 | tr -d '\n')"
-    echo "${interval},${eu_times}${ap_times#*,}" >> leader_mode.csv
-done

### 2021-02-24 evaluation/data/process-crdt.sh

@@ -1,0 +1,40 @@
+#!/bin/bash
+set -euo pipefail
+
+LOGDIR="${1?Please specify directory containing log files}"
+
+# Processes the log files produced by runs of the experiment in CRDT mode
+# (USE_LEADER = false), and extracts the numbers we're interested in, in
+# a format ready for Gnuplot. The output file contains the following columns:
+#
+#  1. Interval between generated ops (microseconds)
+#  2. Ops/sec generated by the us-west-1 replica
+#  3. Minimum    time to apply a local  operation in us-west-1 (microseconds)
+#  4. Median     time to apply a local  operation in us-west-1 (microseconds)
+#  5. 95th perc. time to apply a local  operation in us-west-1 (microseconds)
+#  6. Minimum    time to apply a remote operation in us-west-1 (microseconds)
+#  7. Median     time to apply a remote operation in us-west-1 (microseconds)
+#  8. 95th perc. time to apply a remote operation in us-west-1 (microseconds)
+#  9. Ops/sec generated by the eu-west-1 replica
+# 10. Minimum    time to apply a local  operation in eu-west-1 (microseconds)
+# 11. Median     time to apply a local  operation in eu-west-1 (microseconds)
+# 12. 95th perc. time to apply a local  operation in eu-west-1 (microseconds)
+# 13. Minimum    time to apply a remote operation in eu-west-1 (microseconds)
+# 14. Median     time to apply a remote operation in eu-west-1 (microseconds)
+# 15. 95th perc. time to apply a remote operation in eu-west-1 (microseconds)
+# 16. Ops/sec generated by the ap-southeast-1 replica
+# 17. Minimum    time to apply a local  operation in ap-southeast-1 (microseconds)
+# 18. Median     time to apply a local  operation in ap-southeast-1 (microseconds)
+# 19. 95th perc. time to apply a local  operation in ap-southeast-1 (microseconds)
+# 20. Minimum    time to apply a remote operation in ap-southeast-1 (microseconds)
+# 21. Median     time to apply a remote operation in ap-southeast-1 (microseconds)
+# 22. 95th perc. time to apply a remote operation in ap-southeast-1 (microseconds)
+
+rm -f "$LOGDIR/summary.data"
+
[…]

### 2021-02-24 evaluation/data/process-leader.sh

@@ -1,0 +1,28 @@
+#!/bin/bash
+set -euo pipefail
+
+LOGDIR="${1?Please specify directory containing log files}"
+
+# Processes the log files produced by runs of the experiment in leader mode
+# (state machine replication, USE_LEADER = true), and extracts the numbers
+# we're interested in, in a format ready for Gnuplot. The output file contains
+# the following columns:
+#
+#  1. Interval between generated ops (microseconds)
+#  2. Operations/sec processed by all replicas
+#  3. Requests/sec made by eu-west-1 replica to the leader (us-west-1)
+#  4. Minimum    round trip time from eu-west-1 to the leader (microseconds)
+#  5. Median     round trip time from eu-west-1 to the leader (microseconds)
+#  6. 95th perc. round trip time from eu-west-1 to the leader (microseconds)
+#  7. Requests/sec made by ap-southeast-1 replica to the leader (us-west-1)
+#  8. Minimum    round trip time from ap-southeast-1 to the leader (microseconds)
+#  9. Median     round trip time from ap-southeast-1 to the leader (microseconds)
+# 10. 95th perc. round trip time from ap-southeast-1 to the leader (microseconds)
+
+rm -f "$LOGDIR/summary.data"
+
+for interval in $(ls "$LOGDIR"/*.log.gz | sed -e 's/.*interval_//' -e 's/_.*//' | uniq | sort -rn); do
+    eu_times="$(cat "${LOGDIR}/interval_${interval}_eu_west_1.log.gz"      | gunzip | awk -f process-leader.awk | tail -n 2 | head -n 1 | tr '\n' ',')"
+    ap_times="$(cat "${LOGDIR}/interval_${interval}_ap_southeast_1.log.gz" | gunzip | awk -f process-leader.awk | tail -n 2 | head -n 1 | tr '\n' ',')"
+    echo "${interval},${eu_times}${ap_times#*,}" | tr ',' '\t' >> "$LOGDIR/summary.data"
+done

### 2021-02-24 evaluation/data/processing_times.sh

@@ -1,15 +1,0 @@
-#!/bin/bash
-
-echo ",,us_west_1,,,,,,,eu_west_1,,,,,,,ap_southeast_1" >> processing_times.csv
-echo -n "interval,ops/sec," >> processing_times.csv
-echo -n "local min,local median,local p95,remote min,remote median,remote p95," >> processing_times.csv
-echo -n "local min,local median,local p95,remote min,remote median,remote p95," >> processing_times.csv
-echo    "local min,local median,local p95,remote min,remote median,remote p95"  >> processing_times.csv
-
-for interval in 100000 50000 20000 14000 10000 7000 6000 5000 2000; do
-    us_times="$(cat logs/interval_${interval}_us_west_1.log.gz      | gunzip | awk -f processing_times.awk | tail -n 2 | head -n 1 | tr '\n' ',')"
-    eu_times="$(cat logs/interval_${interval}_eu_west_1.log.gz      | gunzip | awk -f processing_times.awk | tail -n 2 | head -n 1 | tr '\n' ',')"
-    ap_times="$(cat logs/interval_${interval}_ap_southeast_1.log.gz | gunzip | awk -f processing_times.awk | tail -n 2 | head -n 1 | tr '\n' ',')"
-    op_rate="$(echo "${us_times%%,*} + ${eu_times%%,*} + ${ap_times%%,*}" | bc)"
-    echo "${interval},${op_rate},${us_times#*,}${eu_times#*,}${ap_times#*,}" >> processing_times.csv
-done

### 2021-02-24 evaluation/run-all.sh

@@ -6,7 +6,7 @@ set -euo pipefail
 INTERVAL="${1?Please specify interval between operations (in microseconds)}"
 
 # If false, runs in CRDT mode. If true, runs in leader-based mode.
-USE_LEADER=true
+USE_LEADER=false
 
 # If true, uses algorithm implementation extracted from Isabelle.
 # If false, uses a hand-written (not formally verified) implementation.
@@ -39,11 +39,11 @@ if [ "$USE_LEADER" = "true" ]; then
     LOGDIR="data/logs-leader"
     # Define us-west-1 to be the leader
     ssh -i ~/.ec2/martin-aws-us-west-1.pem ubuntu@$US_WEST_1 \
-        /home/ubuntu/move-op/evaluation/run-test.sh $INTERVAL $USE_LEADER $USE_GENERATED_CODE 0 &
+        /home/ubuntu/move-op/evaluation/run-test.sh $INTERVAL $USE_LEADER $USE_GENERATED_CODE 0 '' '' &
     ssh -i ~/.ec2/martin-aws-eu-west-1.pem ubuntu@$EU_WEST_1 \
-        /home/ubuntu/move-op/evaluation/run-test.sh $INTERVAL $USE_LEADER $USE_GENERATED_CODE 1 $US_WEST_1 &
+        /home/ubuntu/move-op/evaluation/run-test.sh $INTERVAL $USE_LEADER $USE_GENERATED_CODE 1 $US_WEST_1 '' &
     ssh -i ~/.ec2/martin-aws-ap-southeast-1.pem ubuntu@$AP_SOUTHEAST_1 \
-        /home/ubuntu/move-op/evaluation/run-test.sh $INTERVAL $USE_LEADER $USE_GENERATED_CODE 2 $US_WEST_1 &
+        /home/ubuntu/move-op/evaluation/run-test.sh $INTERVAL $USE_LEADER $USE_GENERATED_CODE 2 $US_WEST_1 '' &
     wait
 else
     LOGDIR="data/logs"

### 2021-02-24 evaluation/run-test.sh

@@ -1,5 +1,5 @@
 #!/bin/bash
-set -euo pipefail
+set -eo pipefail
 
 # This script executes on an EC2 VM. It runs a replica that generates
 # operations with a given inter-operation interval.

### 2021-02-26 paper/review-response.tex
Address reviewer comments

@@
 \begin{document}
-\title{Summary of changes of ``A highly-available move operation for replicated trees''}
+\title{Summary of changes to ``A highly-available move operation for replicated trees''}
 \author{Martin Kleppmann, Dominic P.\ Mulligan, Victor B.F.\ Gomes, Alastair R.\ Beresford}
@@
 \item We have changed the title (previously \emph{``A highly-available move operation for replicated trees and distributed filesystems''}) to remove \emph{``and distributed filesystems''}, as suggested by Reviewer 2.
+\item We have added an optimisation for operations that create new tree nodes, allowing them to be applied without any undo or redo, and proved using Isabelle/HOL that this optimisation is safe.
+We have also added an explanation of why the same optimisation does not work for removing tree nodes, addressing point C3 by Reviewer 2.
+\item We have created an optimised alternative implementation of our algorithm, and added it to the performance evaluation.
+The optimised implementation is not formally verified, but it is approximately 10 times faster than the verified code generated by Isabelle/HOL.
+This implementation significantly narrows the performance gap between our algorithm and state machine replication.
+This addresses point C7 by Reviewer 2.
+\item We have expanded the discussion of related work as suggested by the reviewers, re-run all of the experiments, and made various other improvements to clarity.
 \end{itemize}
@@
 There is no discussion that could help the reader understand that, given a particular system/approach configuration or set of performance or availability goals, approach X would be preferable to approach Y.
+\end{spverbatim}
+\authorcomment{Section \ref{sec:smr} contains a discussion of the trade-offs between our algorithm, locking, and state machine replication, including an explanation of the situations in which a particular approach would be preferable.
+These are all the approaches that we currently know of.}
+\begin{spverbatim}
 In Section 3.5, please explain all cases with examples.
+\end{spverbatim}
+\authorcomment{Done.}
+\begin{spverbatim}
 My main concern is whether the comparison is fair.
@@
 How could one trustfully decide to apply the operations on the basis of these results for another scenario?
+\end{spverbatim}
+\authorcomment{We requested clarification from the reviewer (via the editor) as to which specific aspects of our evaluation they believed to be unfair, but we did not receive any reply.
+Without a concrete request it is impossible for us to address this point.}
+\begin{spverbatim}
 
@@
 2. Does the manuscript contain sufficient and appropriate references? Please explain under Public Comments below.: Important references are missing; more references are needed
-
+\end{spverbatim}
+\authorcomment{We asked the reviewer to name the references they wanted us to add (again via the editor), but we did not receive a response.
+The editor advised us to prepare a revision based on the other reviews.}
+\begin{spverbatim}
 3. Does the introduction state the objectives of the manuscript in terms that encourage the reader to read on? Please explain your answer under Public Comments below.: Could be improved
@@
 Yes. However, it is unclear why is necessary to include "distributed filesystems" in the title. The work is concerned with a tree replicated datatype and distributed filesystems is an application.
-
+\end{spverbatim}
+\authorcomment{We have removed this part of the title as suggested.}
+\begin{spverbatim}
 ## Does the manuscript contain sufficient and appropriate references?
@@
 Since the case study in De Porre's paper is an AVL tree, a comparison should be included in the related work.
-
+\end{spverbatim}
+\authorcomment{Thank you for pointing out this paper, which we had missed. We now discuss it in Section~\ref{sec:oplogs}.}
+\begin{spverbatim}
 ## Does the introduction state the objectives of the manuscript in terms that encourage the reader to read on?
@@
 C1: The system model and network requirements should be explicitly defined.
-
+\end{spverbatim}
+\authorcomment{We have improved the description of the system model at the beginning of Section~\ref{sec:algorithm}.}
+\begin{spverbatim}
 C2: Is the metadata parameter necessary? It seems that this parameter could be abstracted away which would simplify the algorithm presentation.
-
+\end{spverbatim}
+\authorcomment{The metadata is not needed for the functioning of our algorithm per se, but it is needed for several applications of our algorithm, such as filesystems (metadata is a filename) or JSON/XML (metadata indicates the order of children of the same parent).
+Since it does not significantly complicate the algorithm, we would prever to keep it as-is.}
+\begin{spverbatim}
 C3: Further justification is needed for choosing to encode add and remove operations as move operations.
@@
 I was left wondering if by having explicit add and remove operations some undo-do-redo cycles could be avoided.
-
+\end{spverbatim}
+\authorcomment{We have done substantial further work on this question, and added Section~\ref{sec:create} to discuss it.
+In particular, we have proved in Isabelle that creation operations can skip the undo/redo process, provided that some additional assumptions are met.
+We now also explain why the same optimisation does not work for remove operations.}
+\begin{spverbatim}
 C4: Instead of having an undo-do-redo cycle would it be possible to have state checkpoints?
 As operations might arrive in any order, the number of operations to undo can be high, in that case, it might be useful to have state checkpoints and combine these with undo operations.
+\end{spverbatim}
+\authorcomment{Periodically checkpointing the state would indeed reduce the number of operations to undo, since a replica could jump to the checkpoint nearest to the log position where a new operation needs to be inserted.
+However, we believe that this would not bring about a large performance improvement, because we would still have to perform the redo step for all operations with timestamps greater than the new operation, which still has a cost that is proportional to the number of operations we have to undo if there are no checkpoints.
 
+This redo is necessary because the new operation may cause any subsequent operation to flip from safe to unsafe or vice versa, and so we have to recompute that safety status for all operations with timestamps greater than the new operation.}
+\begin{spverbatim}
 C5: Notation is not consistent. It is stated that types are primed letters (2nd paragraph of section 3.1), however, this naming convention is dropped later on.
+\end{spverbatim}
+\authorcomment{We have carefully checked the paper and confirmed that the notation is correct.
 
+Primed letters are \emph{type variables}, not concrete types.
+For example, the tree has type $('n, '\!m, '\!n)\;\mathit{set}$, which means a set of triples where the first and third element of each triple have the same type (since they both represent a tree node, namely the parent and the child).
+The type variable $'n$ can then be instantiated with any concrete type; for example, in our evaluation it is a 64-bit integer.}
+\begin{spverbatim}
 C6: The evaluation proves the point that the tree CRDT is the best option for fast response time and disconnected mode (in this case it is the only option).
@@
 It's reasonable to conclude that the comparison between the proposed CRDT and state machine replication would be maintained, but the optimised code would give a clear evaluation of the tree CRDT.
+\end{spverbatim}
+\authorcomment{We have implemented an alternative version of the algorithm directly in Scala, and added it to the evaluation.
+It is almost 10 times faster than the Isabelle-generated code in terms of the CRDT operation throughput it can sustain.
+This work allows us to distinguish between the costs that are incidental inefficiencies of Isabelle's generated code, and the costs that are inherent to our algorithm.
+We thank the reviewer for prompting this work, which we believe has made the paper stronger.}
+\begin{spverbatim}
 
@@
 There are some choices that need to be further discussed and the evaluation should consider an optimised implementation of the algorithm.
-
+\end{spverbatim}
+\authorcomment{Done.}
+\begin{spverbatim}
 Additional Questions:
@@
 \end{spverbatim}
+\authorcomment{Thank you for the very positive review!}
 \end{document}

### 2021-02-26 paper/move-op.tex
New data processing scripts and graphs
Update paper title
Address reviewer comments

@@
 \begin{document}
-\title{A highly-available move operation for replicated trees and distributed filesystems}
+\title{A highly-available move operation for replicated trees}
 
@@
 However, standard techniques such as two-phase locking require synchronous coordination between replicas.
-If we want the system to continue processing read and write requests even in the presence of arbitrary network partitions (in other words, if we want \emph{availability} and \emph{partition-tolerance} in the sense of the CAP theorem~\cite{Gilbert:2002il}), then we must avoid such coordination.
-This is an important practical consideration since network partitions are common in many networks~\cite{Alquraan:2018wj,Bailis:2014jx}.
+If the software is running on mobile devices with unreliable network connectivity, an application based on synchronous coordination becomes unresponsive during network interruptions, leaving users unable to work while they are offline.
 
-We can use \emph{optimistic replication}~\cite{Saito:2005jw}, which means that any replica can make changes to the data without waiting for communication with any other replicas; operations are sent to other replicas at some later time when a network connection is available.
-This approach has the advantage of enabling disconnected operation on mobile devices, and high availability in geo-replicated settings.
-It also improves performance for end users because the duration of a user request is independent of any network latency.
+If we want to allow offline work, we must allow the system to continue processing read and write requests even in the presence of arbitrary network partitions; in other words, we require high \emph{availability} and \emph{partition-tolerance} in the sense of the CAP theorem~\cite{Gilbert:2002il}.
+We can achieve this goal by using \emph{optimistic replication}~\cite{Saito:2005jw}, which means that any replica can make changes to the data without waiting for communication with any other replicas; updates made while disconnected are sent to other replicas later when a network connection is available.
+Besides allowing the system to tolerate network partitions, this approach can also improve performance for end users because the response time of a user request is independent of any network latency.
 
@@
         Our proof technique can also be applied to other distributed systems, making it of independent interest.
-    \item To demonstrate the practical viability of our approach, we refine the algorithm to an executable implementation within Isabelle/HOL and prove the equivalence of the two.
-        We extract a formally verified Scala implementation from Isabelle and evaluate its performance with replicas across three continents.
+    \item To demonstrate its practical viability, we extract a formally verified Scala implementation of our algorithm from Isabelle.
+        We compare its performance to a hand-optimised (not formally verified) implementation and to classic state machine replication.
+        In a geo-replicated setup across three continents, our algorithm has approximately half the throughput of state machine replication, but up to five orders of magnitude lower latency.
     \item We perform experiments with Dropbox and Google Drive, and show that they exhibit problems that would be prevented by our algorithm.
@@
 This restriction is required because allowing this operation would introduce a cycle into the directory graph, and so the filesystem would no longer be a tree.
-This restriction is required in any other tree structure that supports a move operation.
+Any tree data structure that supports a move operation must handle this case.
 
@@
 
-Another possible resolution is shown in Figure~\ref{fig:move-cycle}b: the nodes involved in the concurrent moves (and their children) could be duplicated, so that both ``$A$ as a child of $B$'' and ``$B$ as a child of $A$'' can exist in the tree.
+Another possible outcome is shown in Figure~\ref{fig:move-cycle}b: the nodes involved in the concurrent moves (and their children) could be duplicated, so that both ``$A$ as a child of $B$'' and ``$B$ as a child of $A$'' can exist in the tree.
 However, such duplication is undesirable for the same reasons as in \S\ref{sec:move-same-item}.
@@
 
-\section{The replicated move operation}\label{sec:algorithm}
+\section{The replicated tree algorithm}\label{sec:algorithm}
 
@@
 When the user wants to make a change to the tree, they generate an operation and apply it to their local replica.
-Every operation is also asynchronously sent over the network to all other replicas, and applied to every remote replica using the same algorithm as for local operations.
-The network may deliver operations in any order; thus, the communication can be performed peer-to-peer, and it does not require any central server or consensus protocol.
-We only assume that eventually every operation is applied on every replica, provided that network partitions are of a finite duration.
+Every operation is also asynchronously sent over the network to all other replicas, and applied by every remote replica using the same algorithm as for local operations.
+A replica may communicate directly with any other replica.
+The network may arbitrarily delay or reorder messages, but we assume that an underlying network protocol detects and retransmits lost messages, and suppresses duplicates.
+We do not assume any central server or consensus protocol.
+Any number of replicas may fail by crashing, and any non-crashed subset of replicas can continue executing operations.
 
 The key consistency property of our algorithm is \emph{convergence}: that is, whenever any two replicas have applied the same set of operations, then they must be in the same state---even if the operations were applied in a different order on different replicas.
-We prove this in \S\ref{sec:proof} by showing that applying operations is commutative.
+We prove this in \S\ref{sec:proof} by showing that applying operations using our algorithm is commutative.
 
@@
 We also designate as ``trash'' some node ID that does not exist in the tree; then we can delete a node by moving it to be a child of the trash.
+Node creation and deletion are discussed further in \S\ref{sec:create}.
 
@@
 Otherwise, \isa{oldp} records the previous parent and metadata of \isa{c}: if there exist \isa{p'} and \isa{m'} such that \isa{(p', m', c) {\isasymin} tree}, then \isa{oldp} is set to \isa{Some (p', m')}.
-The \isa{get\_parent} function implements this (lines 16--20).
 
+The \isa{get\_parent} function (lines 16--20) implements this.
 In the first line of \isa{get\_parent}, the expression between \isa{::} and \textbf{where} is the type signature of the function, in this case:
@@
 The \isa{apply\_op} function (lines 42--49) takes two arguments: a \isa{Move} operation to apply and the current replica state; and it returns the new replica state.
-The constraint \isa{'t::\{linorder\}} in the type signature indicates that timestamps \isa{'t} are instances of \isa{linorder} type class, and they can therefore be compared with the $<$ operator defining a linear (or total) order.
+The constraint \isa{'t::\{linorder\}} in the type signature indicates that timestamps \isa{'t} are instances of the \isa{linorder} type class, and they can therefore be compared with the $<$ operator defining a linear (or total) order.
 This comparison occurs on line 47.
@@
         Since move operations do not specify the old location of a node, but only the new location, this sequential execution of concurrent operations is well-defined.
-    \item If two operations would introduce a cycle when combined, as in \S\ref{sec:move-cycle}, then the operation with the greater timestamp is ignored.
-        This is the case because \isa{do\_op} checks for cycles based on the tree created by all operations with a lower timestamp.
+        In Figure~\ref{fig:move-same-item}, the outcome is (c) if the operation from replica 1 has the greater timestamp, or (d) if the operation from replica 2 has the greater timestamp.
+    \item If two operations would introduce a cycle when combined, as in \S\ref{sec:move-cycle}, then the operation with the greater timestamp is ignored because \isa{do\_op} checks for cycles based on the tree created by all operations with a lower timestamp.
         The lower of two conflicting operations will take effect, since that operation by itself is safe.
         When the higher-timestamped operation is applied, \isa{do\_op} detects that it would introduce a cycle, and therefore ignores the operation.
+        In the example of Figure~\ref{fig:move-cycle}, the outcome is (c) if the operation from replica 1 has the lower timestamp, or (d) if the operation from replica 2 has the lower timestamp.
 \end{itemize}
 
-This resolution of a conflict between two operations can be generalised to any number of conflicting operations by applying the rules repeatedly pairwise.
+This resolution of a conflict between two operations can be generalised to any number of conflicting operations by repeatedly applying the rules pairwise.
 
@@
 In practice, the collision would be resolved by making the filenames distinct, e.g.\ by appending a replica identifier to the filenames.
+
+\subsection{Node creation and deletion}\label{sec:create}
+
+As discussed previously, no separate operations for node creation and deletion are needed, since a node is implicitly created when it is first moved, and a node can be deleted by moving it to a designated trash node outside of the tree.
+By avoiding a distinction between different operation types we simplify the algorithm and proofs of correctness, but we potentially sacrifice performance in applications that mostly create and delete nodes, and only occasionally move nodes.
+Ideally, we would want to incur the costs of the undo-do-redo process only for genuine move operations, and not for node creation and deletion.
+
+This optimisation is possible for operations that create new tree nodes: such operations can be directly applied to the tree without any undo/redo by using \isa{do\_op} instead of \isa{apply\_op}, and they do not need to recorded in the log.
+We have proved in Isabelle that this optimisation is safe, under the following additional assumptions: 1.~the parent node in any move/create operation must exist in the tree; 2.~the creation operation for a given node must be unique (i.e.\ there cannot be two operations that both create the same node); 3.~a node creation operation is applied before any operation that moves the created node.
+These assumptions are easily satisfied in practice.
+
+For node deletion we have not been able to find a similar optimisation.
+Deletion operations must go through the undo-do-redo process because deleting a node may cause a previously unsafe move operation with a greater timestamp to become safe by breaking a particular ancestor relationship; and as the previously ignored operation takes effect, we must re-evaluate all operations with greater timestamps to determine whether they are safe or not.
+
+When a node is deleted, any children of the deleted node remain in the tree, but since they are no longer reachable from the root, they effectively become invisible to the application.
+We do not recursively remove children, since a move operation concurrent to the deletion operation might move the deleted subtree back out of the trash and into the visible tree; in this scenario we want all children of the moved subtree to be preserved unmodified.
 
@@
 This approach to determining ordering also easily supports reordering of child nodes within a parent: to move a node to a different position in a list, we use the list CRDT to generate a new ID at the desired position in the sequence~\cite{Kleppmann:2020jw}.
-Then we perform a move operation in which the parent node is unchanged, but the metadata is changed to this new ID.
+Then we perform a move operation in which the parent node is unchanged, and this new ID is used as metadata.
 
-\smallbreak\noindent\textbf{File merging.}
-In a distributed filesystem, replication and conflict resolution is required not only for the directory structure, but also for the contents of individual files.
-This can be accomplished by using CRDTs for file contents as well.
-We discuss distributed filesystems further in \S\ref{sec:filesystems}.
+%\smallbreak\noindent\textbf{File merging.}
+%In a distributed filesystem, replication and conflict resolution is required not only for the directory structure, but also for the contents of individual files.
+%This can be accomplished by using CRDTs for file contents as well.
+%We discuss distributed filesystems further in \S\ref{sec:filesystems}.
 
@@
 A tree is an acyclic graph in which every node has exactly one parent, except for the root, which has no parent.
-In fact, we slightly generalise this property and allow more than one root to exist, so the graph represents a forest, allowing an application to move nodes between different trees, if desired.
-For example, the ``trash'' node used for deletion (\S\ref{sec:ops-trees}) can be separate from the main tree.
+In fact, we slightly generalise this property and allow more than one root to exist, so the graph represents a forest, allowing an application to move nodes between different trees if desired.
+For example, the trash node used for deletion can be separate from the main tree.
 To prove that our algorithm maintains a forest structure, no matter which operations are applied, we demonstrate several invariants.
@@
 
-\smallbreak\noindent\textbf{Other correctness properties}
+\smallbreak\noindent\textbf{Other correctness properties.}
 There are further criteria we might use to determine if our algorithm is correct.
@@
 
+\begin{figure}
+  \begin{center}
+  \includegraphics{../evaluation/data/crdt-generated.pdf}
+  \end{center}
+  \caption{Median execution time to apply an operation to the replica state, using Scala code generated by Isabelle. Error bars indicate the minimum and 95th percentile.}
+  \label{fig:plot-generated}
+\end{figure}
+
+\begin{figure}
+  \begin{center}
+  \includegraphics{../evaluation/data/crdt-optimised.pdf}
+  \end{center}
+  \caption{Median execution time to apply an operation to the replica state, using a hand-optimised (not formally verified) Scala implementation. Error bars indicate the minimum and 95th percentile.}
+  \label{fig:plot-optimised}
+\end{figure}
+
 \section{Evaluation}\label{sec:evaluation}
@@
 
-We used Isabelle/HOL's code generation mechanism to extract executable Scala code from our HOL definitions, as discussed in \S\ref{subsect.extracting}, and wrapped this implementation in a simple network service.
-We deployed three replicas of this service on Amazon EC2 \texttt{c5.large} instances in different regions worldwide: in Northern California (region \texttt{us-west-1}), Ireland (\texttt{eu-west-1}), and Singapore (\texttt{ap-southeast-1}).
+Our algorithm can be computationally expensive if a large number of undos and redos are required.
+To demonstrate that this cost is acceptable in practice, we evaluated the performance of two different implementations of our algorithm.
+The first implementation is Scala code extracted from our formally verified HOL definitions using Isabelle/HOL's code generation mechanism, as discussed in \S\ref{subsect.extracting}.
+The second implementation is handwritten Scala code that we believe to be functionally equivalent, but which we have not formally verified.
+By comparing these implementations we can distinguish between inefficiencies introduced by code generation and costs that are inherent to our algorithm.
+
+We wrapped both implementations in a simple network service and deployed three replicas of this service on Amazon EC2 \texttt{c5.large} instances in different regions worldwide: in Northern California (region \texttt{us-west-1}), Ireland (\texttt{eu-west-1}), and Singapore (\texttt{ap-southeast-1}).
 The network delays between these regions are substantial: in our experiments, a request-response round-trip from Ireland to California takes a median of 145~ms, while a request from Singapore to California takes 176~ms.
 
-We use a synthetic workload in which each replica starts with an empty tree, and then generates move operations at a fixed rate; for each move operation the generating replica chooses parent and child nodes uniformly at random from a set of 1,000 tree nodes.
-A tree node is created by the first operation that refers to it.
-We use Lamport timestamps~\cite{Lamport:1978jq} as operation timestamps, and 64-bit integers as tree node identifiers.
-When a replica generates an operation, it immediately applies that operation to its local state, and asynchronously sends the operation to the other two replicas via TCP connections.
+\begin{figure*}
+  \begin{center}
+  \includegraphics{../evaluation/data/leader-vs-crdt.pdf}
+  \end{center}
+  \caption{Median time to apply a local CRDT operation compared to the median time to perform a move operation using state machine replication, with a leader located in another region.
+Note the log scale on both axes.}
+  \label{fig:leader-plot}
+\end{figure*}
+
+We use a synthetic workload in which each replica starts with an empty tree and generates move operations at a fixed rate; for each move operation the generating replica chooses parent and child nodes uniformly at random from a set of 1,000 tree nodes.
+A tree node is created by the first operation that refers to it, and for simplicity we do not use the optimisation discussed in \S~\ref{sec:create}.
+We use Lamport clocks~\cite{Lamport:1978jq} as operation timestamps, and 64-bit integers to identify tree nodes.
+When a replica generates an operation, it immediately applies that operation to its local state, and it asynchronously sends the operation to the other two replicas via TCP connections.
 When a replica receives an operation from a remote replica, it also applies that operation to its own replica state.
 Thus, of the operations applied by each replica, approximately one third are locally generated, and two thirds are received from the other two replicas.
+A replica does not wait for a response for the previous operation before generating and sending the next, so there may be many operations ``in flight'' at the same time.
 
@@
 We repeated the experiment with different operation rates, and measured the execution time of our algorithm for applying each operation.
-Figure~\ref{fig:plots} shows the median execution times for local and remote operations respectively.
-The execution time for local operations is near-constant (median times between 50~{\textmu}s and 70~{\textmu}s), while the execution time for remote operations increases approximately linearly from 0.14~ms to 1.5~ms over the range of 30 to 600 operations/sec.
-At around 600 operations/sec each single-threaded replica is fully utilising one CPU core, and increasing the rate at which operations are generated does not increase throughput any further.
+Figures~\ref{fig:plot-generated} and~\ref{fig:plot-optimised} show the results for the Isabelle-generated code and the handwritten code respectively.
+In each figure, the upper plot shows the distribution of execution times to apply one operation received from another replica, and the lower plot shows the time to apply one locally generated operation.
 
-\begin{figure}
-  \includegraphics[width=\columnwidth,keepaspectratio=true]{plot.pdf}
-  \caption{Median execution time for applying an operation to the replica state, on an Amazon EC2 c5.large instance, at varying operation throughput rates.
-Error bars indicate the minimum and 95th percentile execution times.}
-  \label{fig:plots}
-\end{figure}
+Both implementations exhibit qualitatively similar behaviour, but the absolute numbers differ significantly.
+A local operation takes near-constant time to apply because its timestamp is always greater than any existing operation at the generating replica (by definition of Lamport clocks), so it does not require any undo or redo.
+The median time to apply a local operation is around 50~{\textmu}s for the Isabelle-generated code and around 1--2~{\textmu}s for the handwritten code.
+The reduced performance at low operation rates can be explained by the JVM delaying JIT optimisations until a method has been executed a certain number of times.
 
-\begin{figure}
-  \includegraphics[width=\columnwidth,keepaspectratio=true]{leader-plot.pdf}
-  \caption{Median time to apply a local CRDT operation compared to the median time to perform a move operation using state machine replication, with a leader located in another region.
-Note the log scale on both axes.}
-  \label{fig:leader-plot}
-\end{figure}
+For remote operations, the execution time increases proportionally to the rate at which operations are generated: as the time interval between successive operations decreases relative to the network delay, there are more operations in flight at the same time, and more undos and redos are required to put operations in timestamp order.
+At some point, each single-threaded replica is fully utilising one CPU core, and increasing the rate at which operations are generated does not increase throughput any further.
+The Isabelle-generated code is saturated at a rate of 600 operations/sec (with a median remote operation taking around 1~ms), while the optimised code is saturated at 5,700 operations/sec.
+This factor of 9.5 performance difference is solely a result of inefficient code generation; functionally both implementations are performing the same work.
 
-Local operations take constant time is because a locally generated operation always has a timestamp greater than any existing operation at the generating replica at that time (by definition of Lamport timestamps).
-Applying that operation requires only a \isa{do\_op}, and no \isa{undo\_op} or \isa{redo\_op}.
+At peak, the optimised implementation is performing on average 200 undos and redos per remote operation.
+The throughput could be increased by applying a batch of operations at once, which would allow the cost of the undos and redos to be amortised over the size of the batch.
+We leave an evaluation of this optimisation for future work.
 
-On the other hand, for remote operations, a number of undos and redos must be performed, depending on the degree of concurrency.
-As the time interval between successive operations decreases relative to the network delay, there are more operations ``in flight'' at the same time, and more calls to \isa{undo\_op} and \isa{redo\_op} are required to order operations by timestamp.
-The time taken to apply a remote operation is proportional to the number of undos and redos required, and hence proportional to the operation rate.
-
-Note that when a user is interacting with the system, they only have to wait for local operations to execute; any remote operations can be applied in the background without affecting user interaction.
-Since our algorithm need not perform any undos or redos for local operations, these user interactions are consistently fast.
+% Note that when a user is interacting with the system, they only have to wait for local operations to execute; any remote operations can be applied in the background without affecting user interaction.
+% Since our algorithm need not perform any undos or redos for local operations, these user interactions are consistently fast.
 
@@
 
-A simple alternative to our CRDT algorithm is to use state machine replication~\cite{Schneider:1990vy}: that is, we use a leader replica or consensus algorithm to impose a total order on all operations, and then execute operations in that same order on all replicas.
+An alternative to our CRDT algorithm is to use a locking protocol (\S~\ref{sec:impossibility}).
+For example, a replica can acquire a lock from a lock server, perform an update, and then release the lock again.
+The minimum time for which a lock will be held after it is acquired is then the round-trip time.
+If the lock server is in California and the replica is in Singapore, and if one move operation is performed per lock acquisition, this system could only perform $1/176\,\mathrm{ms} = 5.7$ operations/sec.
+
+A better alternative to our CRDT algorithm is to use state machine replication~\cite{Schneider:1990vy}: that is, we use a leader replica or consensus algorithm to impose a total order on all operations, and then execute operations in that same order on all replicas.
 For a move operation on trees, the state machine replication algorithm is much simpler than the CRDT: it still needs to check for cycles, but it never needs to undo or redo any operations because they are never applied out-of-order.
@@
 The Irish and Singaporean replicas generated operations, sent them to the leader, and applied them to their local tree in the order they were received from the leader.
-In order to ensure a fair comparison to the CRDT algorithm, these experiments used the same Isabelle-generated Scala code.
+In order to ensure a fair comparison to the CRDT algorithm, we ran these experiments using the same two implementations (Isabelle-generated and handwritten) and the same networking code.
 
-The results from this experiment are shown in Figure~\ref{fig:leader-plot}.
-The leader-based approach was able to sustain a throughput of 14,000 move operations per second, approximately 23 times the CRDT's throughput of 600 ops/sec, due to the fact that it does not need to spend CPU cycles undoing and redoing operations.
-On the other hand, the latency of each operation is around 3,000 times greater in the state machine approach, due to the network delay to the leader.
+The results are shown in Figure~\ref{fig:leader-plot}.
+Using the Isabelle-generated code, the leader-based approach is able to sustain 14,000 move operations per second (23 times the CRDT's throughput of 600 ops/sec).
+Using the handwritten code, the leader-based throughput is 22,000 ops/sec (4 times the CRDT's throughput of 5,700 ops/sec).
+The downside of state machine replication is that performing an operation requires waiting for a round-trip to the leader, which implies around five orders of magnitude higher latency (145--176~ms) than the 1--2~{\textmu}s it takes to execute local CRDT operations.
 
-Therefore, we have a clear trade-off: in applications that need to prioritise throughput, a state machine replication approach is preferable, while in applications that need to minimise response times to user requests, our CRDT algorithm is preferable.
-Our algorithm also has the advantage that local user operations require no waiting for network communication, so users can still modify the tree while disconnected from the Internet.
+Therefore, we have a clear trade-off: in applications that need to maximise throughput, a state machine replication approach is preferable; in applications that need to minimise response times to user requests or that need to continue to be available during network interruptions, our CRDT algorithm is preferable.
 In the leader-based approach, clients cannot make updates while offline.
-
-We note that the throughput of 600 ops/sec is for a single tree, e.g.\ a filesystem belonging to one user.
-Operations on independent trees, belonging to different users, can be processed in parallel.
-We believe that for many applications, 600 ops/sec/user is plenty, and the ability to work offline is a valuable benefit.
 
@@
 Our proofs contain no unproven assumptions (i.e. no occurrences of the \textbf{sorry} keyword).
-Checking all of the proofs takes 3.5 minutes on a 2018 MacBook Pro.
+Checking all of the proofs takes 3 minutes on a 2018 MacBook Pro.
 
-Besides the 59 lines of definitions given in Figure~\ref{fig:code}, our Isabelle/HOL formalisation consists of a further 2,166 lines of code.
-Of this, we use 203 lines to prove that every node has a unique parent, 443 lines to prove that the tree contains no cycles, 450 lines to prove that move operations commute and replicas converge, 327 lines to prove the strong eventual consistency property, and 743 lines to define the executable variant of our algorithm and prove its equivalence to the definitions of Figure~\ref{fig:code}.
+Besides the 59 lines of definitions given in Figure~\ref{fig:code}, our Isabelle/HOL formalisation consists of a further 2,495 lines of code.
+Of this, we use 203 lines to prove that every node has a unique parent, 443 lines to prove that the tree contains no cycles, 450 lines to prove that move operations commute and replicas converge, 327 lines to prove the strong eventual consistency property, 743 lines to define the executable variant of our algorithm and prove its equivalence to the definitions of Figure~\ref{fig:code}, and 779 lines to prove the safety of the optimisation in \S\ref{sec:create}.
 
@@
         Tao et al.\ also perform experiments with Dropbox, Google Drive, and OneDrive, similar to our experiments discussed in \S\ref{sec:move-is-hard}.
+    \item Nair et al.~\cite{Nair:2021} develop a CRDT tree with move operation.
+        This work is concurrent to ours and it is unpublished at the time of writing, so we have not been able to conduct a detailed comparison.
 \end{itemize}
+
+Several other CRDTs, such as Treedoc~\cite{Preguica:2009fz} and LSEQ~\cite{Nedelec:2013ky}, use a tree structure internally.
+However, their data model is a linear sequence; the tree structure is not accessible to the application (for example, an application cannot freely choose the parent node of a new tree node), and they do not provide a move operation.
 
@@
 
-After we performed the experiments described in Section~\ref{sec:move-is-hard}, the Dropbox engineering team published a blog post~\cite{Dropbox2020sync} acknowledging the problems of cycles and duplication due to concurrent moves.
+After we performed the experiments described in \S\ref{sec:move-is-hard}, the Dropbox engineering team published a blog post~\cite{Dropbox2020sync} acknowledging the problems of cycles and duplication due to concurrent moves.
 
-\subsection{Totally ordered operation log}
+\subsection{Totally ordered operation log}\label{sec:oplogs}
 
 Our approach of ordering operations by a timestamp, and undoing/redoing them as necessary so that they take effect in ascending timestamp order, is conceptually very simple.
-Similar ideas appear in many other systems, including the Bayou database~\cite{Terry:1995dn}, Jefferson's \emph{Time Warp} mechanism~\cite{Jefferson:1985em}, Shapiro et al.'s \emph{capricious total order}~\cite{Shapiro:2016hp}, and Burckhardt's \emph{standard conflict resolution}~\cite[{\S}4.3.3]{Burckhardt:2014hy}.
+Similar ideas appear in many other systems, including the Bayou database~\cite{Terry:1995dn}, Jefferson's \emph{Time Warp} mechanism~\cite{Jefferson:1985em}, Shapiro et al.'s \emph{capricious total order}~\cite{Shapiro:2016hp}, Burckhardt's \emph{standard conflict resolution}~\cite[{\S}4.3.3]{Burckhardt:2014hy}.
 The concept of undo-redo is also well known from write-ahead logging~\cite{Mohan:1992fe}.
 
-However, to our knowledge, this approach has not previously been applied to the problem of replicated trees, nor are we aware of previous mechanised proofs (using Isabelle or other tools) that formalise this approach.
-The simplicity of the approach facilitates reasoning about it.
+The SECRO approach~\cite{DePorre:2019} allows arbitrary CRDTs to be defined by having each replica execute deterministic update functions in the same total order, and replaying operation history if necessary.
+In principle, our algorithm could be expressed as a SECRO, but we provide optimisations that go beyond the SECRO model.
+SECRO always processes operations in forward direction, whereas our use of undo/redo is more efficient on long operation histories if replicas have most of the operation log in common.
+(SECRO allows long operation histories to be truncated, but this mechanism may result in discarded operations if replicas are disconnected for extended periods of time.)
+Moreover, our optimisation of node creation operations (\S\ref{sec:create}) is specific to our algorithm, and is not possible in the general SECRO model.
+
+As an example of the SECRO approach, De Porre et al.\ present an AVL tree CRDT~\cite{DePorre:2019}.
+Even though this data structure is internally a tree, the interface it exposes is an ordered set datatype, not a tree in which the user can choose to create, delete, and move arbitrary nodes.
+
+To our knowledge, the approach of ordering operations by timestamp has not previously been applied to the problem of replicated trees, and in particular not a move operation.
+We are not aware of previous mechanised proofs (using Isabelle or other tools) that formalise this approach.
 
@@
 Our approach ensures that all replicas converge to the same consistent state without needing any manual conflict resolution, and without requiring application developers to implement conflict handling logic.
-Updates made to a local replica take effect immediately, while operations from remote replicas can be propagated and applied in the background.
+Updates made to a local replica take effect immediately, while operations from remote replicas can be propagated and applied asynchronously.
 This approach means that user interaction is consistently fast, even in the face of unbounded communication delays or during disconnected operation of mobile devices.
@@
 To rule out such bugs, we formally verified the correctness of our algorithm using the Isabelle/HOL proof assistant; our theorems show that replicas can apply operations in any order, and that the result is always a valid tree (nodes have at most one parent, and the graph does not contain any cycles).
-Moreover, these results apply to unbounded executions and an arbitrary number of replicas---an advantage of using a proof assistant rather than other formal approaches such as model checking.
+Moreover, these results apply to unbounded executions and an arbitrary number of replicas---an advantage of using a proof assistant over other formal approaches such as model checking.
 
@@
 
-We also extracted a formally verified Scala implementation from our Isabelle/HOL definitions, and evaluated its performance across three replicas in California, Ireland and Singapore.
-Our implementation applies local updates in approximately 50~{\textmu}s, three orders of magnitude faster than is possible with a leader replica or consensus protocol operating over these distances, and it remains available in the face of network interruptions.
-However, compared to leader-based replication our algorithm has lower throughput of 600 updates per second.
-A hand-optimised implementation of our algorithm could achieve higher performance, since code generated from HOL definitions contains inefficiencies.
-The generated code could serve as a reference implementation against which other implementations could be tested.
+We also evaluated the performance of two implementations of our algorithm (one formally verified, the other optimised for performance) across three replicas in California, Ireland and Singapore.
+Our optimised implementation applies local updates in 1--2~{\textmu}s, five orders of magnitude faster than is possible with a leader replica or consensus protocol operating over these distances, and our algorithm remains available in the face of network interruptions.
+However, compared to leader-based replication our algorithm has four times lower throughput.
 
 As highlighted in the introduction, our work is especially interesting due to the ubiquity of tree data models across many different types of applications and databases.
-In future work we hope to build novel applications, such as a distributed filesystem, using our algorithm.
+In future work we hope to integrate our algorithm into CRDT libraries such as Automerge,\footnote{\url{https://github.com/automerge/automerge}} and use it to build novel collaborative applications.
 
@@
 We expect that our Isabelle/HOL formalisation will be a valuable resource for future work in this area, which can use our definitions and proofs as starting point.
-The formalisation is included as supplementary data of this paper.
+The formalisation is open source~\cite{SourceFiles} and included as supplementary data of this paper.
 
@@
 The authors wish to acknowledge the support of The Boeing Company and the EPSRC ``REMS: Rigorous Engineering for Mainstream Systems'' programme grant (EP/K008528).
-Martin Kleppmann is supported by a Leverhulme Trust Early Career Fellowship and by the Isaac Newton Trust.
+Martin Kleppmann is supported by a Leverhulme Trust Early Career Fellowship, the Isaac Newton Trust, and Nokia Bell Labs.
 Our evaluation was conducted using AWS credits from the AWS Educate program.
@@
 \bibliography{references}{}
-\newpage
 

### 2021-02-26 paper/move-op.tex
Add Patreon supporters

@@
 The authors wish to acknowledge the support of The Boeing Company and the EPSRC ``REMS: Rigorous Engineering for Mainstream Systems'' programme grant (EP/K008528).
-Martin Kleppmann is supported by a Leverhulme Trust Early Career Fellowship, the Isaac Newton Trust, and Nokia Bell Labs.
+Martin Kleppmann is supported by a Leverhulme Trust Early Career Fellowship, the Isaac Newton Trust, Nokia Bell Labs, and crowdfunding supporters including Molham Aref, Henry Caushi, Chet Corcos, Gabriel Herrero-Beaumont, Bogdan Kanivets, David Pollak, SoftwareMill, Talent Formation Network, and Adam Wiggins.
 Our evaluation was conducted using AWS credits from the AWS Educate program.

### 2021-03-02 paper/review-response.tex
Revision submitted to TPDS

@@
 \authorcomment{The metadata is not needed for the functioning of our algorithm per se, but it is needed for several applications of our algorithm, such as filesystems (metadata is a filename) or JSON/XML (metadata indicates the order of children of the same parent).
-Since it does not significantly complicate the algorithm, we would prever to keep it as-is.}
+Since it does not significantly complicate the algorithm, we would prefer to keep it as-is.}
 \begin{spverbatim}
@@
 Primed letters are \emph{type variables}, not concrete types.
-For example, the tree has type $('n, '\!m, '\!n)\;\mathit{set}$, which means a set of triples where the first and third element of each triple have the same type (since they both represent a tree node, namely the parent and the child).
+For example, the type of our tree is written in Isabelle as $('n \times '\!m \times '\!n)\;\mathit{set}$, which means a set of triples where the first and third element of each triple have the same type (since they both represent a tree node, namely the parent and the child).
 The type variable $'n$ can then be instantiated with any concrete type; for example, in our evaluation it is a 64-bit integer.}

### 2021-03-02 paper/move-op.tex
Revision submitted to TPDS

--- 2021-02-26 paper/move-op.tex
@@
 \IEEEPARstart{M}{any} applications use a tree-structured data model.
-For example, most file systems are trees: directories are branch nodes, and files are leaf nodes.
-Moreover, XML and JSON documents are trees, and they are used in many applications to represent e.g.\ rich text (a tree of paragraphs, lists, figures, sections, etc.), vector graphics, CAD drawings, and many other types of data.
+Most filesystems are trees: directories are branch nodes, files are leaf nodes (we discuss hardlinks in \S\ref{sec:extensions}).
+XML and JSON documents are also trees, and they are used in many applications to represent e.g.\ rich text (a tree of paragraphs, lists, figures, sections, etc.), vector graphics, CAD drawings, and many other types of data.
 Typically, a graphical user interface allows a user to edit this information interactively, resulting in updates to the underlying XML/JSON structure: adding or deleting nodes in the tree, and moving nodes from one position in the tree to another.
@@
 The authors wish to acknowledge the support of The Boeing Company and the EPSRC ``REMS: Rigorous Engineering for Mainstream Systems'' programme grant (EP/K008528).
-Martin Kleppmann is supported by a Leverhulme Trust Early Career Fellowship, the Isaac Newton Trust, and Nokia Bell Labs.
+Martin Kleppmann is supported by a Leverhulme Trust Early Career Fellowship, the Isaac Newton Trust, Nokia Bell Labs, and crowdfunding supporters including Molham Aref, Chet Corcos, Mintter.com, David Pollak, SoftwareMill, Talent Formation Network, and Adam Wiggins.
 Our evaluation was conducted using AWS credits from the AWS Educate program.

### 2021-08-18 paper/move-op.tex
Add note about O(hd) complexity; edit to fit in 14 pages

@@
         We compare its performance to a hand-optimised (not formally verified) implementation and to classic state machine replication.
-        In a geo-replicated setup across three continents, our algorithm has approximately half the throughput of state machine replication, but up to five orders of magnitude lower latency.
+        In a geo-replicated setup across three continents, state machine replication has approximately four times higher throughput than our algorithm, but our algorithm has up to 100,000 times lower latency.
     \item We perform experiments with Dropbox and Google Drive, and show that they exhibit problems that would be prevented by our algorithm.
@@
 
-Our algorithm can be computationally expensive if a large number of undos and redos are required.
-To demonstrate that this cost is acceptable in practice, we evaluated the performance of two different implementations of our algorithm.
-The first implementation is Scala code extracted from our formally verified HOL definitions using Isabelle/HOL's code generation mechanism, as discussed in \S\ref{subsect.extracting}.
+With our algorithm, the worst-case cost of applying a move operation is $O(hd)$, where $h$ is the number of operations in the history that need to be undone and redone, and $d$ is the depth of the tree (the number of parent relationships that need to be traversed to check whether the operation would introduce a cycle).
+To demonstrate that this cost is acceptable in practice, we evaluated the performance of two implementations of our algorithm.
+The first implementation is Scala code extracted from our formally verified HOL definitions using Isabelle/HOL's code generation mechanism.
 The second implementation is handwritten Scala code that we believe to be functionally equivalent, but which we have not formally verified.
@@
 
-We wrapped both implementations in a simple network service and deployed three replicas of this service on Amazon EC2 \texttt{c5.large} instances in different regions worldwide: in Northern California (region \texttt{us-west-1}), Ireland (\texttt{eu-west-1}), and Singapore (\texttt{ap-southeast-1}).
-The network delays between these regions are substantial: in our experiments, a request-response round-trip from Ireland to California takes a median of 145~ms, while a request from Singapore to California takes 176~ms.
+We wrapped both implementations in a simple network service and deployed three replicas on Amazon EC2 \texttt{c5.large} instances in Northern California (\texttt{us-west-1}), Ireland (\texttt{eu-west-1}), and Singapore (\texttt{ap-southeast-1}).
+The network delays between these regions are substantial: a round trip from Ireland to California takes a median of 145~ms, and from Singapore to California takes 176~ms.
 
@@
 
-For remote operations, the execution time increases proportionally to the rate at which operations are generated: as the time interval between successive operations decreases relative to the network delay, there are more operations in flight at the same time, and more undos and redos are required to put operations in timestamp order.
+For remote operations, the execution time increases proportionally to the rate at which operations are generated: as the time interval between successive operations decreases relative to the network delay, more operations are in flight at the same time, and more undos/redos are required to put operations in timestamp order.
 At some point, each single-threaded replica is fully utilising one CPU core, and increasing the rate at which operations are generated does not increase throughput any further.
-The Isabelle-generated code is saturated at a rate of 600 operations/sec (with a median remote operation taking around 1~ms), while the optimised code is saturated at 5,700 operations/sec.
-This factor of 9.5 performance difference is solely a result of inefficient code generation; functionally both implementations are performing the same work.
+The Isabelle-generated code is saturated at a rate of 600 operations/sec (median remote operations taking $\approx$1~ms), while the optimised code is saturated at 5,700 operations/sec.
+This factor of 9.5 performance difference is solely due to inefficient code generation; functionally both implementations perform the same work.
 
@@
 
-\subsection{Comparison to state machine replication}\label{sec:smr}
+\subsection{Comparison to locking/state machine replication}\label{sec:smr}
 
@@
 If the lock server is in California and the replica is in Singapore, and if one move operation is performed per lock acquisition, this system could only perform $1/176\,\mathrm{ms} = 5.7$ operations/sec.
+The throughput of this locking-based scheme is therefore three orders of magnitude lower than our optimised algorithm's throughput of 5,700 operations/sec.
 
@@
 
-Besides the 59 lines of definitions given in Figure~\ref{fig:code}, our Isabelle/HOL formalisation consists of a further 2,495 lines of code.
+Besides the 59 lines of definitions given in Figure~\ref{fig:code}, our Isabelle/HOL formalisation consists of a further 2,495 lines of proof code.
 Of this, we use 203 lines to prove that every node has a unique parent, 443 lines to prove that the tree contains no cycles, 450 lines to prove that move operations commute and replicas converge, 327 lines to prove the strong eventual consistency property, 743 lines to define the executable variant of our algorithm and prove its equivalence to the definitions of Figure~\ref{fig:code}, and 779 lines to prove the safety of the optimisation in \S\ref{sec:create}.
@@
 \begin{itemize}
-    \item Martin et al.~\cite{Martin:2010ih} define a CRDT for XML data, and Kleppmann and Beresford~\cite{Kleppmann:2016ve} define a CRDT for JSON.
+    \item Martin et al.~\cite{Martin:2010ih,Martin:2011} define a CRDT for XML data, and Kleppmann and Beresford~\cite{Kleppmann:2016ve} define a CRDT for JSON.
         However, these algorithms only deal with insertion and deletion of tree nodes, and do not support moves.
@@
     \item As discussed in \S\ref{sec:impossibility}, Najafzadeh et al.~\cite{Najafzadeh:2017vk,Najafzadeh:2018bw} propose two implementations for a replicated filesystem: a CRDT in which conflicting moves are handled by duplicating tree nodes (as in Figures~\ref{fig:move-same-item}b and~\ref{fig:move-cycle}b), and a centralised implementation in which move operations must obtain a lock before executing (not a CRDT since it relies on synchronous coordination).
+        We discuss the performance costs of locking in \S\ref{sec:smr}.
     \item Ahmed-Nacer et al.~\cite{AhmedNacer:2012us} outline approaches to handling conflicts on trees, but provide no algorithms.
@@
         This work is concurrent to ours and it is unpublished at the time of writing, so we have not been able to conduct a detailed comparison.
+        % TODO now available on arxiv: https://arxiv.org/abs/2103.04828
 \end{itemize}
@@
 Our approach of ordering operations by a timestamp, and undoing/redoing them as necessary so that they take effect in ascending timestamp order, is conceptually very simple.
-Similar ideas appear in many other systems, including the Bayou database~\cite{Terry:1995dn}, Jefferson's \emph{Time Warp} mechanism~\cite{Jefferson:1985em}, Shapiro et al.'s \emph{capricious total order}~\cite{Shapiro:2016hp}, Burckhardt's \emph{standard conflict resolution}~\cite[{\S}4.3.3]{Burckhardt:2014hy}.
+Similar ideas appear in many other systems, including the Bayou database~\cite{Terry:1995dn}, Jefferson's \emph{Time Warp} mechanism~\cite{Jefferson:1985em}, and Burckhardt's \emph{standard conflict resolution}~\cite[{\S}4.3.3]{Burckhardt:2014hy}.
 The concept of undo-redo is also well known from write-ahead logging~\cite{Mohan:1992fe}.
@@
 
-As highlighted in the introduction, our work is especially interesting due to the ubiquity of tree data models across many different types of applications and databases.
+Our work is especially interesting due to the ubiquity of tree data models across many different types of applications and databases.
 In future work we hope to integrate our algorithm into CRDT libraries such as Automerge,\footnote{\url{https://github.com/automerge/automerge}} and use it to build novel collaborative applications.
@@
 We are also exploring whether the undo-do-redo approach can be used in other concurrent data structures; for example, there is an open problem in collaborative text editing~\cite{Kleppmann:2020jw} that might be solved by this approach.
-We expect that our Isabelle/HOL formalisation will be a valuable resource for future work in this area, which can use our definitions and proofs as starting point.
-The formalisation is open source~\cite{SourceFiles} and included as supplementary data of this paper.
+Our Isabelle/HOL formalisation, which is open source~\cite{SourceFiles}, can be used for future work in this area.
 
@@
 The authors wish to acknowledge the support of The Boeing Company and the EPSRC ``REMS: Rigorous Engineering for Mainstream Systems'' programme grant (EP/K008528).
-Martin Kleppmann is supported by a Leverhulme Trust Early Career Fellowship, the Isaac Newton Trust, Nokia Bell Labs, and crowdfunding supporters including Molham Aref, Chet Corcos, Mintter.com, David Pollak, SoftwareMill, Talent Formation Network, and Adam Wiggins.
+Martin Kleppmann is supported by a Leverhulme Trust Early Career Fellowship, the Isaac Newton Trust, Nokia Bell Labs, and crowdfunding supporters including Ably, Adrià Arcarons, Chet Corcos, Macrometa, Mintter, David Pollak, RelationalAI, SoftwareMill, Talent Formation Network, and Adam Wiggins.
 Our evaluation was conducted using AWS credits from the AWS Educate program.
@@
 is a Senior Research Associate at the University of Cambridge.
-His research spans distributed systems, databases, security, and formal verification, with a particular focus on decentralised collaboration software and CRDTs.
+His research spans distributed systems, databases, security, and formal verification, with a focus on decentralised collaboration software and CRDTs.
 His book \emph{Designing Data-Intensive Applications} was published in 2017 and has been translated into six languages.
-Previously, he worked as a software engineer and entrepreneur at several internet companies, including Rapportive and LinkedIn.
+Previously, he was a software engineer and entrepreneur at Internet companies including Rapportive and LinkedIn.
 \end{IEEEbiography}

### 2021-08-18 paper/review-response2.tex
Another round of review responses

--- 2021-02-10 paper/review-response.tex
@@
 
-We thank the reviewers very much for their comments, which have improved the paper.
-In this document we explain how we have addressed the issues highlighted in the reviews.
-The most important changes are as follows:
-\begin{itemize}
-\item We have changed the title (previously \emph{``A highly-available move operation for replicated trees and distributed filesystems''}) to remove \emph{``and distributed filesystems''}, as suggested by Reviewer 2.
-\end{itemize}
+Thank you for this second round of reviews.
+We are happy to note that reviewers 2 and 3 would like to see the paper accepted with no further changes.
+In this revision we have therefore only made minor changes to address comments by reviewer 1, and added some minor clarifications.
 
@@
 Comments:
-Introduction section offers very brief and simplified coverage of common topics and don't really offer anything new.
-It would be even more valuable if some critical comparisons were made among the systems and approaches.
-There is no discussion that could help the reader understand that, given a particular system/approach configuration or set of performance or availability goals, approach X would be preferable to approach Y.
-In Section 3.5, please explain all cases with examples.
-My main concern is whether the comparison is fair.
-It shows that some results are achievable.
-How could one trustfully decide to apply the operations on the basis of these results for another scenario?
+What is the difference between the proposed algorithm and the data replication algorithms in the cloud? Please state their challenges in the introduction by mentioning references.
+A review of data replication based on meta-heuristics approach in cloud computing and data grid. doi.org/10.1007/s00500-020-04802-1.
+\end{spverbatim}
+\authorcomment{The cited paper is about using optimisation algorithms to decide what to replicate, how many replicas to create, where to locate them, when to access them, and so on.
+Those are important questions, but they are unrelated to our work.
+Our paper deals with the specific problem of ensuring that replicas converge to a consistent state.
+It can be deployed in any system that uses replicated trees, regardless of whether it is cloud-based, client-server, peer-to-peer, or uses any other topology.
+There is no need for us to distinguish between cloud systems and other types of replicated system because from the point of view of our algorithm they are the same.}
+\begin{spverbatim}
+A brief survey on replica consistency in cloud environments. doi.org/10.1186/s13174-020-0122-y
+\end{spverbatim}
+\authorcomment{As we already explain in {\S}1, {\S}4.2, and {\S}6, our algorithm provides a consistency model called ``strong eventual consistency'' [Shapiro et al., 2011].
+The survey referenced above unfortunately overlooks this important consistency model.
+Strong eventual consistency is a strengthening of eventual consistency, and it is often combined with causal consistency.
+It is one of the strongest consistency models that is always available in the sense of the CAP theorem.
+We believe our paper already adequately explains the consistency model, and references [2,3,4] of our paper explain it in further detail.}
+\begin{spverbatim}
+Explain the time complexity of the proposed method.
+\end{spverbatim}
+\authorcomment{We have added a discussion of the time complexity of our algorithm to the beginning of {\S}5.1.}
+\begin{spverbatim}
+Unfortunately, there are no "quantitative" performance comparisons among related strategies.
+It’s also important if the author can summarize the advantages, and disadvantages of each strategy according to the performance metrics.
+\end{spverbatim}
+\authorcomment{Our evaluation ({\S}5) performs a quantitative performance comparison between two implementations of our algorithm and state machine replication.
+We explain the pros and cons of these approaches in detail in this section.
+In \S~5.2 we also compare our algorithm to a locking-based approach, and show that it would have three orders of magnitude lower throughput than our algorithm in our test setup (5.7 versus 5,700 ops/sec).
+
+We therefore believe that our paper already contains the quantitative comparison that the reviewer is asking for.
+If there is a particular performance comparison that is missing in the reviewer's opinion, we request that they specify precisely what should be added.}
+\begin{spverbatim}
 
 Additional Questions:
-1.  Please explain how this manuscript advances this field of research and/or contributes something new to the literature.: It presents move operation for distributed filesystems.
+1.
+Please explain how this manuscript advances this field of research and/or contributes something new to the literature.: Present Conflict-free Replicated Data Type for trees that allow move operations without any coordination between replicas.
 
-2. Is the manuscript technically sound? Please explain your answer under Public Comments below.: Yes
+2. Is the manuscript technically sound? Please explain your answer under Public Comments below.: Appears to be - but didn't check completely
 
-1. Which category describes this manuscript?: Practice / Application / Case Study / Experience Report
+1. Which category describes this manuscript?: Research/Technology
 
@@
 2. Does the manuscript contain sufficient and appropriate references? Please explain under Public Comments below.: Important references are missing; more references are needed
+\end{spverbatim}
+\authorcomment{If references are missing, please specify what they are.}
+\begin{spverbatim}
 
@@
 
-4. How would you rate the organization of the manuscript? Is it focused? Is the length appropriate for the topic? Please explain under Public Comments below.: Could be improved
+4. How would you rate the organization of the manuscript? Is it focused? Is the length appropriate for the topic? Please explain under Public Comments below.: Satisfactory
 
@@
 
-6. Should the supplemental material be included? (Click on the Supplementary Files icon to view files): Yes, as part of the digital library for this submission if accepted
+6. Should the supplemental material be included? (Click on the Supplementary Files icon to view files): Yes, as part of the main paper if accepted (cannot exceed the strict page limit)
 
@@
 
-8. Would you recommend adding the code/data associated with this paper to help address your concerns and/or strengthen the paper?: Yes
+8. Would you recommend adding the code/data associated with this paper to help address your concerns and/or strengthen the paper?: No
 
@@
 \begin{spverbatim}
-Recommendation: Author Should Prepare A Major Revision For A Second Review
+Recommendation: Accept With No Changes
 
 Comments:
-## Is the manuscript technically sound?
-Yes.
-As stated before the paper has a clear contribution, which includes a mechanised proof of the algorithm correctness.
-Moreover, the proof code is presented in the appendix and is available online with instructions on how to check proofs locally.
-
-## How relevant is this manuscript to the readers of this periodical?
-The topic is relevant for TPDS community as it proposes a distributed algorithm to support optimistic replication of tree data structures.
-
-## Are the title, abstract, and keywords appropriate?
-Yes. However, it is unclear why is necessary to include "distributed filesystems" in the title. The work is concerned with a tree replicated datatype and distributed filesystems is an application.
-
-## Does the manuscript contain sufficient and appropriate references?
-The following reference should be included and is missing:
-
-De Porre K, Myter F, De Troyer C, Scholliers C, De Meuter W, Boix EG.
-Putting order in strong eventual consistency.
-In IFIP International Conference on Distributed Applications and Interoperable Systems 2019 Jun 17 (pp.
-36-56).
-Springer, Cham.
-
-De Porre et al.
-work proposes an optimistic replication algorithm that can transform a general-purpose data type into a replicated data type.
-Similarly to the submitted paper, strong eventual consistency is ensured by maintaining a global total order of the operations.
-Since the case study in De Porre's paper is an AVL tree, a comparison should be included in the related work.
-
-## Does the introduction state the objectives of the manuscript in terms that encourage the reader to read on?
-I found the introduction enticing and, although brief, it does a great job in defining the context, illustrating the problem, and clearly defining the contributions.
-
-## How would you rate the organization of the manuscript? Is it focused? Is the length appropriate for the topic?
-The paper is well structured.
-I particularly liked section 2 that clearly illustrates the main issues with concurrent non-coordinated move operations.
-Also, by showing that the replicated filesystems of Dropbox and Google drive both have bugs demonstrates how difficult is to devise a correct resolution policy for the move operation.
-Overall, section 2 nicely shows the work relevance and potential impact without getting too technical.
-The replication algorithm is thoroughly described and the proof of correctness is presented at the right level of detail - although the proofs are an essential component a more detailed presentation would become tedious.
-
-## Please rate the readability of the manuscript.
-The paper is well written and easy to follow.
-
-## Detailed comments
-
-C1: The system model and network requirements should be explicitly defined.
-
-C2: Is the metadata parameter necessary? It seems that this parameter could be abstracted away which would simplify the algorithm presentation.
-
-C3: Further justification is needed for choosing to encode add and remove operations as move operations.
-My intuition, base on previously published results as CISE and Hamsaz, is that not all operations need to follow a total ordered across replicas.
-However, because all operations are defined as moves, conflicts are always possible.
-I was left wondering if by having explicit add and remove operations some undo-do-redo cycles could be avoided.
-
-C4: Instead of having an undo-do-redo cycle would it be possible to have state checkpoints?
-As operations might arrive in any order, the number of operations to undo can be high, in that case, it might be useful to have state checkpoints and combine these with undo operations.
-
-C5: Notation is not consistent. It is stated that types are primed letters (2nd paragraph of section 3.1), however, this naming convention is dropped later on.
-
-C6: The evaluation proves the point that the tree CRDT is the best option for fast response time and disconnected mode (in this case it is the only option).
-
-C7: The choice of using the extracted code for the evaluation is questionable.
-Stating that the code could be optimised, does not provide useful information.
-It is unclear what the impact would be if optimisations were applied.
-It's reasonable to conclude that the comparison between the proposed CRDT and state machine replication would be maintained, but the optimised code would give a clear evaluation of the tree CRDT.
-
-## Overall evaluation
-
-I really enjoyed reading this paper, it's well written and easy to follow and addresses a relevant topic to the TPDS community.
-It has a novel contribution as it defines an optimistic replication algorithm for trees that ensures convergence and the structural properties of trees.
-The idea behind the tree CRDT is simple, but it provides a solution to an existing problem.
-The fact that authors found problems in widely used applications that could be addressed with the tree CRDT proves its usefulness.
-The fact the CRDT was formally verified for any number of replicas and a certified implementation was extracted from the formal definitions, strengths considerably the submission.
-There are some choices that need to be further discussed and the evaluation should consider an optimised implementation of the algorithm.
+The authors have addressed adequately my comments.
 
@@
 1.
-Please explain how this manuscript advances this field of research and/or contributes something new to the literature.: The paper presents the first tree replicated data type that allows concurrent move operations without any coordination between replicas.
+Please explain how this manuscript advances this field of research and/or contributes something new to the literature.: The paper presents a tree replicated data type that allows concurrent move operations without any coordination between replicas.
 The tree data type relies on an optimistic replication algorithm that ensures state convergence and that the tree structure properties are preserved at all times.
-Given the complex (inductive) structure of trees, a strong point of the work is the mechanised proof that provides certification for the correctness of the algorithm.
+Given the complex (inductive) structure of trees, a strong point of the work is the mechanized proof that provides certification for the correctness of the algorithm.
 
@@
 
-2. Does the manuscript contain sufficient and appropriate references? Please explain under Public Comments below.: Important references are missing; more references are needed
+2. Does the manuscript contain sufficient and appropriate references? Please explain under Public Comments below.: References are sufficient and appropriate
 
@@
 
-7. If yes to 6, should it be accepted: After revisions.  Please include explanation under Public Comments below.
+7. If yes to 6, should it be accepted: As is
 
@@
 
-Please rate the manuscript. Please explain your choice.: Excellent
+Please rate the manuscript. Please explain your choice.: Good
 \end{spverbatim}
@@
 Comments:
-The paper presents an implementation of a replicated tree data type with operations for adding, removing, and moving tree nodes.
-Replicated tree data types are essential basic blocks of a wide range of distributed software like distributed file systems, editors, applications manipulating XML/JSON structures, etc.
-While replication is needed to ensure resilience to network/machine failures, an important and difficult challenge in this context is ensuring some form of convergence, e.g., that all the nodes storing a replica of the data structure (tree) reach the same state when all the updates have been propagated in the network.
-This is an instance of strong eventual consistency.
-The paper presents a solution to this challenge that provides high availability as well (clients do not need to wait for the replicas to synchronize), and that is also formally verified in the Isabelle/HOL theorem prover.
-
-The paper starts with a very clear description of the motivation, noting that current solutions for the above challenge, deployed in Google Drive or Dropbox, are not satisfactory.
-The Google Drive solution may lead to unexpected errors while the Dropbox solution may duplicate nodes of the tree which may impact the user experience.
-It identifies the main challenge in designing replicated trees which is related to the move operation: moving a tree node so that it becomes the child of a different parent.
-In fact, node insertion and removal can be reduced to move operations.
-
-Then, it describes an implementation that in particular, solves the issues in Google Drive or Dropbox, which is based on a very neat and simple idea.
-Essentially, every operation is associated with a timestamp (one can use Lamport timestamps [11] and avoid synchronized clocks), and every replica maintains a tree and a log of operations (issued locally or remote operations received from other replicas), ordered by timestamps.
-Applying a remote operation O with timestamp T consists in undoing operations with a timestamp bigger than T, applying O, and finally, re-applying the undone operations.
-This solution, which is quite generic and could be applied for other data types as well, makes it possible to handle concurrent move operations that try to move the same node (taking into account only the one with a bigger timestamp like in Last Writer Wins conflict resolution [10]), and avoid introducing cycles due to concurrent move operations.
-It also ensures strong eventual consistency because when all the updates are received, any two replicas apply them in the same order (according to their timestamps).
-The fact that the implementation satisfies strong eventual consistency and that every replica stores a tree (without cycles) is proved in the Isabelle/HOL theorem prover.
-The implementation in Isabelle is compiled to executable Scala code which is evaluated in a rather realistic deployment (with three replicas on different continents) and compared to a standard master-slave implementation where the order between operations is managed by a single master node.
-This evaluation shows that indeed, the current solution has very low latency compared to the master-slave implementation, although a smaller throughput.
-
-The related work is quite extensive, covering the literature on Conflict-Free Data Types (the implementation in this paper being an instance of such a datatype), distributed file systems (the main application of this work), and works that use similar implementation ideas, based on timestamps.
-
-Overall, I think that the paper is very well written.
-The motivation and the description of the challenges are very convincing, the flow of the ideas being very easy to follow.
-Technically, the implementation relies on an idea which is simple in hindsight, but which has not been used in previous work (up to my knowledge), and which seems to have a very important practical impact.
-The formalization and proofs in Isabelle/HOL are a very important and non-trivial contribution which is meant to increase the confidence we have in this solution.
+I followed the list of revisions and I think that the authors did a great job in addressing the comments of the other reviewers.
+I think that the paper is great and should be accepted in the current form.
 
@@
 Please explain how this manuscript advances this field of research and/or contributes something new to the literature.: The paper presents a new formally-verified implementation of a replicated tree data type, which is an essential ingredient of a wide range of distributed software like distributed file systems, editors, applications manipulating XML/JSON structures, etc.
-This implementation is highly-available (the latency of applying an operation is independent of the network's latency) and eventually consistent (all replicas reach the same state when all updates have been propagated).
+This implementation is highly-available (the latency of applying an operation is independent of the network’s latency) and eventually consistent (all replicas reach the same state when all updates have been propagated).
 These properties are essential in modern distributed services.
@@
 \end{spverbatim}
+
 \end{document}

### 2021-08-18 paper/move-op.tex
paper: updated my job title and affiliation

@@
 \begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{photo-dominic.jpg}}]{Dominic P.\ Mulligan}
-is a Staff Research Engineer in the Security Group at Arm Research, UK.
+is a Principal Research Engineer in the Systems Group at Arm Research, UK.
 Prior to moving to Arm, he worked as a postdoctoral researcher at the Universities of Cambridge and Bologna, Italy, investigating the formal specification and verification of systems software such as C compilers and linkers.

### 2021-08-18 paper/review-response2.tex
review-response: made some minor edits to review-response2.tex

@@
 
-Thank you for this second round of reviews.
-We are happy to note that reviewers 2 and 3 would like to see the paper accepted with no further changes.
+We thank all reviewers for this second round of reviews and are happy to note that reviewers 2 and 3 both would like to see the paper accepted as-is with no further changes.
 In this revision we have therefore only made minor changes to address comments by reviewer 1, and added some minor clarifications.
@@
 \end{spverbatim}
-\authorcomment{The cited paper is about using optimisation algorithms to decide what to replicate, how many replicas to create, where to locate them, when to access them, and so on.
-Those are important questions, but they are unrelated to our work.
-Our paper deals with the specific problem of ensuring that replicas converge to a consistent state.
-It can be deployed in any system that uses replicated trees, regardless of whether it is cloud-based, client-server, peer-to-peer, or uses any other topology.
-There is no need for us to distinguish between cloud systems and other types of replicated system because from the point of view of our algorithm they are the same.}
+\authorcomment{The cited paper is about using optimisation algorithms to decide what to replicate, how many replicas to create, where to locate them, when to access them, and other similar questions.
+Whilst we agree that those questions are indeed important when deciding how to deploy real-world systems we think that they are unrelated to our work, which is a largely theoretical work introducing (and thereafter proving correct) a novel distributed algorithm.
+In particular, our paper introduces techniques that ensure that all replicas converge to a consistent ``view'' of a distributed tree data structure.
+Our work can be deployed in any system that requires the use of a replicated tree, regardless of whether it is cloud-based, client-server, peer-to-peer, or uses any other topology---from our point-of-view we are completely ambivalent about such deployment questions as they exist at a level of abstraction \emph{below} that in which we are working.
+Simply put: there is no need for us to distinguish between cloud systems and other types of replicated system because from the point of view of our algorithm they are the same.}
 \begin{spverbatim}
@@
 \end{spverbatim}
-\authorcomment{As we already explain in {\S}1, {\S}4.2, and {\S}6, our algorithm provides a consistency model called ``strong eventual consistency'' [Shapiro et al., 2011].
+\authorcomment{As we explain in {\S}1, {\S}4.2, and {\S}6, our algorithm provides a consistency model called ``strong eventual consistency'' [Shapiro et al., 2011].
 The survey referenced above unfortunately overlooks this important consistency model.
-Strong eventual consistency is a strengthening of eventual consistency, and it is often combined with causal consistency.
-It is one of the strongest consistency models that is always available in the sense of the CAP theorem.
-We believe our paper already adequately explains the consistency model, and references [2,3,4] of our paper explain it in further detail.}
+Strong eventual consistency is a strengthening of eventual consistency, and it is often combined with causal consistency, and it is one of the strongest consistency models that is always provides ``availability'' in the sense of the CAP theorem.
+We believe our paper already adequately explains the consistency model, and the prior works [2,3,4] referenced in our paper explain it in further detail.}
 \begin{spverbatim}
@@
 \end{spverbatim}
-\authorcomment{We have added a discussion of the time complexity of our algorithm to the beginning of {\S}5.1.}
+\authorcomment{At the request of the reviewer we have added a discussion of the time complexity of our algorithm to the beginning of {\S}5.1.}
 \begin{spverbatim}
@@
 \authorcomment{Our evaluation ({\S}5) performs a quantitative performance comparison between two implementations of our algorithm and state machine replication.
-We explain the pros and cons of these approaches in detail in this section.
+We explain the advantages and disadvantgaes of these approaches in detail in this section.
 In \S~5.2 we also compare our algorithm to a locking-based approach, and show that it would have three orders of magnitude lower throughput than our algorithm in our test setup (5.7 versus 5,700 ops/sec).
@@
 We therefore believe that our paper already contains the quantitative comparison that the reviewer is asking for.
-If there is a particular performance comparison that is missing in the reviewer's opinion, we request that they specify precisely what should be added.}
+If there is a particular performance comparison that is missing in the reviewer's opinion, we kindly request that they specify precisely what should be added.}
 \begin{spverbatim}

### 2021-08-18 paper/review-response2.tex
review-response: a little more on the first point

@@
 Our work can be deployed in any system that requires the use of a replicated tree, regardless of whether it is cloud-based, client-server, peer-to-peer, or uses any other topology---from our point-of-view we are completely ambivalent about such deployment questions as they exist at a level of abstraction \emph{below} that in which we are working.
+Note that, whilst we provide an empirical assessment of our algorithm, this assessment is merely intended to demonstrate that our algorithm is not \emph{ipso facto} unworkably inefficient.
+
 Simply put: there is no need for us to distinguish between cloud systems and other types of replicated system because from the point of view of our algorithm they are the same.}
@@
 The survey referenced above unfortunately overlooks this important consistency model.
-Strong eventual consistency is a strengthening of eventual consistency, and it is often combined with causal consistency, and it is one of the strongest consistency models that is always provides ``availability'' in the sense of the CAP theorem.
+Strong eventual consistency is a strengthening of eventual consistency, is often combined with causal consistency, and it is one of the strongest consistency models that is always provides ``availability'' in the sense of the CAP theorem.
 We believe our paper already adequately explains the consistency model, and the prior works [2,3,4] referenced in our paper explain it in further detail.}

### 2021-10-04 paper/move-op.tex
Terminology tweak

@@
 
-With our algorithm, the worst-case cost of applying a move operation is $O(hd)$, where $h$ is the number of operations in the history that need to be undone and redone, and $d$ is the depth of the tree (the number of parent relationships that need to be traversed to check whether the operation would introduce a cycle).
+With our algorithm, the worst-case cost of applying a move operation is $O(nd)$, where $n$ is the number of operations in the log that need to be undone and redone, and $d$ is the depth of the tree (the number of parent relationships that need to be traversed to check whether the operation would introduce a cycle).
 To demonstrate that this cost is acceptable in practice, we evaluated the performance of two implementations of our algorithm.

### 2021-10-07 proof/document/root.tex
Move proof files to subdirectory

\documentclass[10pt]{article}
\usepackage{isabelle,isabellesym}
\usepackage[a4paper,portrait,margin=1in]{geometry}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{pdfsetup}

\urlstyle{rm}
\isabellestyle{tt}

\begin{document}
\title{A highly-available move operation for replicated trees (Proof Document)}
\author{Martin Kleppmann, Dominic P. Mulligan, Victor B. F. Gomes, and Alastair R. Beresford}
\date{}
\maketitle
\tableofcontents
\parindent 0pt\parskip 0.5ex

\input{session} % generated text of all theories
\end{document}

### 2021-10-08 evaluation/src/main/scala/Move_Code.scala
Scala code generated by Isabelle2021

@@ -31,19 +31,19 @@ object DiffArray {
   import scala.collection.mutable.ArraySeq
 
   protected abstract sealed class DiffArray_D[A]
-  final case class Current[A] (a:ArraySeq[A]) extends DiffArray_D[A]
+  final case class Current[A] (a:ArraySeq[AnyRef]) extends DiffArray_D[A]
   final case class Upd[A] (i:Int, v:A, n:DiffArray_D[A]) extends DiffArray_D[A]
 
   object DiffArray_Realizer {
-    def realize[A](a:DiffArray_D[A]) : ArraySeq[A] = a match {
+    def realize[A](a:DiffArray_D[A]) : ArraySeq[AnyRef] = a match {
       case Current(a) => ArraySeq.empty ++ a
-      case Upd(j,v,n) => {val a = realize(n); a.update(j, v); a}
+      case Upd(j,v,n) => {val a = realize(n); a.update(j, v.asInstanceOf[AnyRef]); a}
     }
   }
 
   class T[A] (var d:DiffArray_D[A]) {
 
-    def realize () = { val a=DiffArray_Realizer.realize(d); d = Current(a); a }
+    def realize (): ArraySeq[AnyRef] = { val a=DiffArray_Realizer.realize(d); d = Current(a); a }
     override def toString() = realize().toSeq.toString
 
     override def equals(obj:Any) =
@@ -53,10 +53,10 @@ object DiffArray {
   }
 
 
-  def array_of_list[A](l : List[A]) : T[A] = new T(Current(ArraySeq.empty ++ l))
-    def new_array[A](v:A, sz : BigInt) = new T[A](Current[A](ArraySeq.fill[A](sz.intValue)(v)))
+  def array_of_list[A](l : List[A]) : T[A] = new T(Current(ArraySeq.empty ++ l.asInstanceOf[List[AnyRef]]))
+  def new_array[A](v:A, sz : BigInt) = new T[A](Current[A](ArraySeq.fill[AnyRef](sz.intValue)(v.asInstanceOf[AnyRef])))
 
-    private def length[A](a:DiffArray_D[A]) : BigInt = a match {
+  private def length[A](a:DiffArray_D[A]) : BigInt = a match {
     case Current(a) => a.length
     case Upd(_,_,n) => length(n)
   }
@@ -64,20 +64,20 @@ object DiffArray {
   def length[A](a : T[A]) : BigInt = length(a.d)
 
   private def sub[A](a:DiffArray_D[A], i:Int) : A = a match {
[…]
